### 2012-03-20 papers/ecoop2012/article.tex
Added paper submitted to ECOOP.
This is a version that we submitted
for review. It doesn't (yet) incorporate
the feedback we got.

%%%%%%%%%%%%%%%%%%%%%%% file typeinst.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is the LaTeX source for the instructions to authors using
% the LaTeX document class 'llncs.cls' for contributions to
% the Lecture Notes in Computer Sciences series.
% http://www.springer.com/lncs       Springer Heidelberg 2006/05/04
%
% It may be used as a template for your own input - copy it
% to a new file with a new name and use it as the basis
% for your article.
%
% NB: the document class 'llncs' has its own and detailed documentation, see
% ftp://ftp.springer.de/data/pubftp/pub/tex/latex/llncs/latex2e/llncsdoc.pdf
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[runningheads,a4paper]{llncs}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{chngpage}

\usepackage{amssymb}
\setcounter{tocdepth}{3}
\usepackage{graphicx}
\usepackage{wrapfig}

\usepackage{url}
\newcommand{\keywords}[1]{\par\addvspace\baselineskip
\noindent\keywordname\enspace\ignorespaces#1}

\input{macros}

\begin{document}

\mainmatter  % start of an individual contribution

\title{JavaScript as an Embedded DSL}
% a short form should be given in case it is too long for the running head
%\titlerunning{}

% the name(s) of the author(s) follow(s) next
%
% NB: Chinese authors should write their first names(s) in front of
% their surnames. This ensures that the names appear correctly in
% the running heads and the author index.
%
\author{Grzegorz Kossakowski \and Nada Amin \and Tiark Rompf \and Martin Odersky}
%
\authorrunning{JavaScript as an Embedded DSL}
% (feature abused for this document to repeat the title also on left hand pages)

% the affiliations are given next; don't give your e-mail address
% unless you accept that it will be published
\institute{Ecole Polytechnique F\'ed\'erale de Lausanne (EPFL)\\
\email{\texttt{\{first.last\}@epfl.ch}}}
%\mailsa\\
%\mailsb\\
%\url{http://lamp.epfl.ch}}

%
% NB: a more complex sample for affiliations and the mapping to the
% corresponding authors can be found in the file "llncs.dem"
% (search for the string "\mainmatter" where a contribution starts).
% "llncs.dem" accompanies the document class "llncs.cls".
%

%\toctitle{}
%\tocauthor{}
\maketitle

\begin{abstract}
% =>70 & <=150 words

Developing rich web applications requires mastering different
environments on the client and server sides. While there is
considerable choice on the server-side, the client-side is tied to
JavaScript, which poses considerable software engineering challenges,
such as moving or sharing pieces of code between the environments. We
embed JavaScript as a DSL in Scala, using Lightweight Modular Staging.
DSL code can be compiled to JavaScript or executed as part of the
server application. We use features of the host language to make
client-side programming safer and more convenient. We use gradual
typing to interface typed DSL programs with existing JavaScript APIs.
We exploit a selective CPS transform already available in the host
language to provide a compelling abstraction over asynchronous
callback-driven programming in our DSL.

\keywords{JavaScript, Scala, DSL, programming languages}
\end{abstract}

\section{Introduction}

Developing rich web applications requires mastering a heterogeneous environment: though the server-side can be implemented in any language, on the client-side, the choice is limited to JavaScript. The trend towards alternative approaches to client-side programming (as embodied by CoffeeScript~\cite{coffeescript}, Dart~\cite{dart} \& GWT~\cite{gwt}) shows the need for more options on the client-side. How do we bring advances in programming languages to client-side programming?

One challenge in developing a large code base in JavaScript is the lack of static typing, as types are helpful for maintenance, refactoring, and reasoning about correctness. Furthermore, there is a need for more abstraction and modularity. ``Inversion of control'' in asynchronous callback-driven programming leads to code with control structures that are difficult to reason about. A challenge is to introduce helpful abstractions without a big hit on performance and/or code size. Communication between the server side and the client side aggravates the impedance mismatch: in particular, data validation logic needs to be duplicated on the client-side for interactivity and on the server-side for security.

% Motivation for our approach vs other approaches

There are three widely known approaches for addressing the challenges outlined above. One is to create a standalone language or DSL that is compiled to JavaScript and provides different abstractions compared to JavaScript. Examples include WebDSL~\cite{webdsl}, Links~\cite{links,links-form} and Dart~\cite{dart}. However, this approach usually requires a lot effort in terms of language and compiler design, and tooling support, although WebDSL leverages Spoofax~\cite{spoofax} to alleviate this effort. Furthermore, it is not always clear how these languages interact with other languages on the server-side or with the existing JavaScript ecosystem on the client-side.

Another approach is to start with an existing language like Java, Scala or Clojure and compile it to JavaScript. Examples include GWT~\cite{gwt}, Scala+GWT~\cite{scala-gwt} and Clojurescript~\cite{clojurescript}. This approach addresses the problem of impedance mismatch between client and server programming but comes with its own set of challenges. In particular, compiling Scala code to JavaScript requires compiling Scala's standard library to JavaScript as any non-trivial Scala program uses Scala collections. This leads to not taking full advantage of libraries and abstractions provided by the target platform which results in big code size and suboptimal performance of Scala applications compiled to JavaScript. For example, a map keyed by strings would be implemented natively in JavaScript as an object literal, while, in Scala, one would likely use the hash map from the standard library, causing it to be compiled to and emulated in JavaScript. Moreover, both approaches tend to not accommodate very well to different API design and programming styles seen in many existing JavaScript libraries.

A third approach is to design a language than is a thin layer on top of JavaScript but provides some new features. A prime example of this idea is CoffeeScript~\cite{coffeescript}. This approach makes it easy to integrate with existing JavaScript libraries but does not solve the impedance mismatch problem. In addition, it typically does not give rise to new abstractions addressing problems seen in callback-driven programming style, though some JavaScript libraries such as Flapjax~\cite{flapjax} and Arrowlets~\cite{arrowlets} are specifically designed for this purpose.

We present a different approach, based on Lightweight Modular Staging
(LMS)~\cite{lms}, that aims to incorporate good ideas from all the
approaches presented above but at the same time tries to avoid their
described shortcomings. LMS is a technique for embedding DSLs as
libraries into a host language such as Scala, while enabling
domain-specific compilation / code-generation. The program is split
into two stages: the first stage is a program generator that, when
run, produces the second stage program. Whether an expression belongs
to the first or second stage is decided by its type. Expressions
belonging to the second stage, also called ``staged expressions'',
have type \code{Rep[T]} in the first stage when yielding a computation
of type \code{T} in the second stage. Expressions evaluated in the
first stage become constants at the second stage. Other approaches to
staging include MetaML~\cite{metaml}, LISP quasiquotations, and
binding-time analysis in partial evaluation. Previous work has
established LMS as a pragmatic approach to runtime code generation and
compiled DSLs. In particular, the Delite
framework~\cite{brown11dsl,rompf11dsl,ppopp11delite} uses this
approach to provide an extensible suite of high-performance DSLs
targeting heterogeneous parallel platforms (with options to generate
code to Scala, C and Cuda)~\cite{lee11dsl}, for domains such as
machine learning~\cite{icml11optiml}, numeric array
processing~\cite{ureche12} and mesh-based partial differential
equation solvers~\cite{hassan10virtualization}. LMS has also been used
to generate SQL queries~\cite{siq}.

%Two widely used approaches to tackle these challenges is to create a new language which targets JavaScript (such as WebDSL) or to compile an existing language to JavaScript (such as GWT). These approaches make it difficult to inter-operate with existing JavaScript libraries. One could think of compiling Scala to JavaScript, but programming in Scala involves using the standard library. A naive translation of the Scala standard library to JavaScript might not Translating existing libraries in the source language to JavaScript

% WebDSL -> issue with integrating existing JS libraries
% compiling an existing language to JS ->

% programmable code generator

We propose to embed JavaScript as a DSL in a host language. Through LMS, we tackle the challenges outlined above with minimal effort, as most of the work is off-loaded to the host language. In particular, we make the following contributions:
\begin{itemize}
\item Our DSL is statically typed through the host language, yet supports gradual typing notably for incorporating external JavaScript libraries and APIs (section~\ref{sec:typing}).
\item In addition to generating JavaScript code, our DSL can be executed directly in the host language, allowing code to be shared between client and server (section~\ref{sec:trivial-embedding}).
\item We use advanced object-oriented techniques to achieve modularity in our DSL: each language primitive and API is defined in a separate module (section~\ref{sec:modularity}).
\item Our DSL supports typed object literals and class-based objects. The translations to JavaScript are lightweight and intuitive: the object literals translate to JSON-like object literals and the class-based objects to JavaScript constructor-based objects (section~\ref{sec:reification}).
\item With minimal effort, we exploit the selective CPS transform already existing in the host language to provide a compelling abstraction over asynchronous callback-driven programming in our DSL (section~\ref{sec:cps}). This case-study demonstrates the fruitfulness of re-using host language features to enhance our embedded DSL.
\end{itemize}

In section~\ref{sec:eval}, we describe our experience in using the DSL, and conclude in section~\ref{sec:conclusion}.

\section{Introduction to LMS}

In LMS, a DSL is split into two parts, its
interface and its implementation. Both parts can be assembled from
components in the form of Scala traits. DSL programs are written in
terms of the DSL interface only, without knowledge of the
implementation.

Part of each DSL interface is an abstract type constructor
\code{Rep[_]} that is used to wrap types in the DSL programs. The DSL
implementation provides a concrete instantiation of \code{Rep} as IR
nodes. When the DSL program is staged, it produces an intermediate
representation (IR), from which the final code can be generated. In
the DSL program, wrapped types such as \code{Rep[Int]} represent
staged computations while expressions of plain unwrapped types
(\code{Int}, \code{Bool}, etc.) are evaluated at staging time as
in~\cite{finally-tagless,polymorphic-embedding}.

Consider the difference between these two programs:
\begin{lstlisting}
def prog1(b: Bool, x: Rep[Int]) = if (b) x else x+1
def prog2(b: Rep[Bool], x: Rep[Int]) = if (b) x else x+1
\end{lstlisting}

The only difference in these two programs is the type of the parameter
\code{b}, illustrating that staging is purely type-driven with no
syntactic overhead as the body of the programs are identical.

In \code{prog1}, \code{b} is a simple boolean, so it must be provided
at staging time, and the \code{if} is evaluated at staging time. For
example, \code{prog1(true, x)} evaluates to \code{x}. In \code{prog2},
\code{b} is a staged value, representing a computation which yields a
boolean. So \code{prog2(b, x)} evaluates to an IR node for the \code{if}:
{\tt\small If(b, x, Plus(x, Const(1)))}.

For \code{prog2}, notice that the \code{if} got transformed into an IR
node. To achieve this, LMS uses an extension of Scala,
Scala-Virtualized~\cite{scala-virtualized}, in which control
structures such as \code{if} can be reified into method calls, so that
alternative implementation can be provided. In our case, we provide an
implementation of \code{if} that constructs an IR node instead of
acting as a conditional. In addition, the \code{+} operation is
overloaded to act on both staged and unstaged expressions. This is
achieved by an implicit conversion from \code{Rep[Int]} to a class
\code{IntOps}, which defines a \code{+} method that creates an IR node
\code{Plus} when executed. Both of \code{Plus}'s arguments must be
staged. We use an implicit conversion to stage constants when needed
by creating a \code{Const} IR node.

\subsection{Example: a DSL program and its generated JavaScript code}

The following DSL snippet creates an array representing a table of
multiplications:
\begin{lstlisting}
def test(n: Rep[Int]): Rep[Array[Int]] =
  for (i <- range(0, n); j <- range(0, n)) yield i*j
\end{lstlisting}

Here is the JavaScript code generated for this snippet:
\begin{lstlisting}[language=JavaScript]
function test(x0) {
  var x6 = []
  for(var x1=0;x1<x0;x1++){
    var x4 = []
    for(var x2=0;x2<x0;x2++){
      var x3 = x1 * x2
      x4[x2]=x3
    }
    x6.splice.apply(x6, [x6.length,0].concat(x4))
  }
  return x6
}
\end{lstlisting}

\subsection{Walkthrough: defining a DSL component}\label{sec:defineDslComponent}

To conclude the introduction to LMS, we show how to add a component
for logging in a DSL, generating JavaScript code which calls
\code{console.log}.

We start by defining the interface:
\begin{lstlisting}
trait Debug extends Base {
  def log(msg: Rep[String]): Rep[Unit]
}
\end{lstlisting}

The \code{Base} trait is part of the core LMS framework and provides
the abstract type constructor \code{Rep}.

Now, we define the implementation:
\begin{lstlisting}
trait DebugExp extends Debug with EffectExp {
  case class Log(msg: Exp[String]) extends Def[Unit]
  def log(msg: Exp[String]): Exp[Unit] = reflectEffect(Log(s))
}
\end{lstlisting}

The \code{EffectExp} trait is part of the core LMS framework. It
inherits from \code{BaseExp} which instantiates \code{Rep} as
\code{Exp}. \code{Exp} represents an IR via two subclasses:
\code{Const} for constants, \code{Sym} for named values defining a
\code{Def}. \code{Def} is the base class for all IR nodes. In our
\code{DebugExp} trait, we extend \code{Def} to support a new IR node:
\code{Log}.

IR nodes are defined as \code{Def}s but they are never referenced
explicitly as such. Instead each \code{Def} has a corresponding symbol
(an instance of \code{Sym}). IR nodes refer to each other using their
symbols. This is why, in the code shown, the \code{msg} parameter is
of type \code{Exp} (not \code{Def}). The method \code{log} 
returns an \code{Exp}. Calling \code{reflectEffect} is what creates
this symbol from the \code{Def}.

In general, the framework provides an implicit conversion from
\code{Def} to \code{Exp}, which performs common subexpression
elimination by re-using the same symbol for identical definitions. We
don't use it here, because \code{log} is a side-effecting operation,
and we don't want to (re)move any such calls even if their message
is the same.

The framework schedules the code generation from the graph of
\code{Exp}s and their dependencies through \code{Def}s. It chooses
which \code{Sym}/\code{Def} pairs to emit and in which order. To
implement code generation to JavaScript for our logging IR node, we
simply override \code{emitNode} to handle \code{Log}:
\begin{lstlisting}
trait JSGenDebug extends JSGenEffect {
  val IR: DebugExp
  import IR._
  override def emitNode(sym: Sym[Any], rhs: Def[Any])(
    implicit stream: PrintWriter) = rhs match {
      case Log(s) => emitValDef(sym, "console.log(" + quote(s) + ")")
      case _ => super.emitNode(sym, rhs)
    }
}
\end{lstlisting}
Notice that in order to compose nicely with other traits, the
overridden method just handles the case it knows and delegates to
other traits, via \code{super}, the emitting of nodes it doesn't know
about.

%- staging virtualization rep types IR construction code generation
%- advantage: can have abstractions that only exist at staging-time

\section{Gradual Typing for Interfacing with Existing APIs}\label{sec:typing}
Since our DSL is embedded in Scala, it inherits its static type
system. However, the generated JavaScript code doesn't need the static
types. Therefore, to help integrate external JavaScript libraries and
APIs (for example, the browser's DOM API), we support a form of
gradual typing. This has proved especially useful for
rapid-prototyping, where external libraries are first incorporated
dynamically, and later declared as typed APIs. Various practical and
theoretical aspects of gradual typing have been studied
by~\cite{wadler-blame,blame-for-all,abadi91,siek06,siek07}.

\subsection{Typed APIs}\label{sec:typedApis}
First, we show how to incorporate an external JavaScript API in a
fully-typed way into our DSL. As an example, consider the following
DSL snippet, which gets the context of an HTML5 canvas element
selected by id:
\begin{lstlisting}
val context = document.getElementById("canvas").as[Canvas].geContext()
\end{lstlisting}

At the DSL interface level, we declare our typed APIs as abstract
Scala traits:
\begin{lstlisting}
trait Dom {
  val document: Rep[Element]
  trait Element
  trait ElementOps {
    def getElementById(id: Rep[String]): Rep[Element]
  }
  trait Canvas extends Element
  trait CanvasOps extends ElementOps {
    def getContext(context: Rep[String]): Rep[Context]
  }
  trait Context
  trait ContextOps {
    def lineTo(x: Rep[Int], y: Rep[Int]): Rep[Unit]
    // etc.
  }
}
\end{lstlisting}

Notice that \code{document} has type \code{Rep[Element]}, and needs to
implement the interface of \code{ElementOps}, so that
\code{document.getElementById("canvas")} is well-typed. We achieve
this using an implicit conversion from \code{Rep[Element]} to
\code{ElementOps}.  At the DSL implementation level, the \code{ElementOps}
returned by this implicit conversion needs to generate an IR node for
each method call, as shown in the walkthrough
in~\ref{sec:defineDslComponent}. For example,
\code{document.getElementById("canvas")} becomes the IR node {\tt\small
  MethodCall(document, "getElementById", List("canvas"))}. This is a
mechanical transformation, implemented by \code{repProxy} using
reflection to intercept method calls and generate IR nodes based on
the method name and the arguments of the invocation. Note that this
use of reflection is purely at staging time, so there is no overhead
in the generated code.
\begin{lstlisting}
trait DomLift extends Dom with JSProxyBase {
  implicit def repToElementOps(x: Rep[Element]): ElementOps =
    repProxy[Element,ElementOps](x)
  implicit def repToCanvasOps(x: Rep[Canvas]): CanvasOps =
    repProxy[Canvas,CanvasOps](x)
  implicit def repToContextOps(x: Rep[Context]): ContextOps =
    repProxy[Context,ContextOps](x)
}
\end{lstlisting}

Note also that since \code{getElementById} returns an arbitrary DOM
element, we need to cast it to a \code{Canvas} using
\code{as[Canvas]}. \code{as} is implemented simply as a cast in the
host language (no IR node is created):
\begin{lstlisting}
  trait AsRep {
    def as[T]: Rep[T]
  }
  implicit def asRep(x: Rep[_]): AsRep = new AsRep {
    def as[T]: Rep[T] = x.asInstanceOf[Rep[T]]
  }
\end{lstlisting}
Instead of this no-op implementation, it is possible to insert
run-time check-cast assertions in the generated JavaScript code.

\subsection{Casting and Optional Runtime Type Checks}
The need for casting arises in a few contexts. One of them is the
boundary between typed and untyped portions of a
program~\cite{wadler-blame,blame-for-all}. Passing a value from an
untyped portion to a typed one usually requires a cast. Another
situation where casts are needed is interaction with external
services. For example, to process data from an external service such
as Twitter, we cast it to its expected type:
\begin{lstlisting}
type TwitterResponse = Array[JSLiteral {val text: String}]
def fetchTweets(username: Rep[String]) = {
 val raw = ajax.get { ... }
 raw.as[TwitterResponse]
}
\end{lstlisting}
A more complete example is provided in section~\ref{sec:cps}. In this
situation, it is useful to generate runtime checks either as an aid
during development and debugging time or as a security mechanism that
validates data coming from an external source. In the example above,
if runtime checks are enabled, by failing early, we obtain a guarantee
that all data returned from \code{fetchTweets} conforms to type
\code{TwitterResponse} which means that any later access to the
\code{text} field of any element of the data array will never fail, and
always return a string.

Notice that when a typed API is defined for an external library, there
are implicit casts introduced for argument and return types of the
defined methods. These casts can also be checked at runtime to
ensure compliance.

We've implemented a component that generates JavaScript code that
asserts casts at runtime. It was fairly straightforward as the host
language allows us to easily inspect types involved in casting. Since
this component just provides a different implementation of the same
casting method \code{as}, it can be enabled selectively for
performance reasons.

%For example, once testing of typed APIs is done, runtime checks for
%implicit casts can be disabled. On the other hand, casts related to
%external service calls should be always enabled for security reasons.

\subsection{Scala Dynamic}
Since our DSL compiles to JavaScript, which is dynamically typed, it
is appealing to allow expressions and APIs in our DSL to also,
selectively, be dynamically typed. This is especially useful for
rapid-prototyping.

We provide a component, \code{JSDynamic}, which allows any expression to become
dynamically typed, by wrapping it in a \code{dynamic} call. The
\code{dynamic} wrapper returns a \code{DynamicRep}, on which any
method call, field access and field update is possible. A dynamic
method call and field access returns another \code{DynamicRep}
expression, so dynamic expressions can be chained.

\code{DynamicRep} exploits a new Scala feature, based on a
marker trait \code{Dynamic}: An expression whose type \code{T} is a
subtype of \code{Dynamic} is subject to the following rewrites:
\begin{itemize}
\item \code{x.f} rewrites to \code{x.selectDynamic("f")},
\item \code{x.m(a, ..., z)} rewrites to \code{x.applyDynamic("m")(a, ..., z)},
\item \code{x.f = v} rewrites to \code{x.updateDynamic("f")(v)}.
\end{itemize}
These rewrites take place when the type \code{T} doesn't statically
have field \code{f} and method \code{m}.

At the implementation level, as these rewriting take place, we
generate IR nodes which allow us to then generate straightforward
JavaScript for the original expressions. For example, for the
expression \code{dynamic(x).foo(1, 2)}, we would generate an IR node
like {\tt\small MethodCall(x, "foo", List(Const(1), Const(2)))}. From this
IR node, it is easy to generate the JavaScript code \code{x.foo(1,
  2)}. Note the similarity with the IR nodes generated for typed APIs.

\subsection{From Dynamic to Static}

This possibility to escape into dynamic typing is particularly useful
in simplifying the incorporation of external JavaScript APIs and
libraries. Sometimes, the user might not want to build a statically
typed API for each external JavaScript library. In addition, for
some library, it might awkward to come up with such a statically-typed
interface. In general, we expect users to start with a dynamic API for
an external library, and progressively migrate it to a typed API as
the code matures. Consider again the example introduced in the typed
API section:
\begin{lstlisting}
val context = document.getElementById("canvas").as[Canvas].getContext()
\end{lstlisting}

In a fully dynamic scenario, we declare the DOM API simply as:
\begin{lstlisting}
trait Dom extends JSDynamic {
  val document: DynamicRep
}
\end{lstlisting}
The \code{as[Canvas]} cast is not necessary in this dynamically typed
setting:
\begin{lstlisting}[escapechar=+]
val context = document.getElementById("canvas").geContext()
\end{lstlisting}

As a first step towards statically typing the API, we declare the type \code{Element}:
\begin{lstlisting}
trait Dom extends JSProxyBase with JSDynamic {
  val document: Rep[Element]
  trait Element
  trait ElementOps {
    def getElementById(id: Rep[String]): DynamicRep
  }
  implicit def repToElementOps(x: Rep[Element]): ElementOps =
    repProxy[Element,ElementOps](x)
}
\end{lstlisting}
Since the method \code{getElementById} returns a \code{DynamicRep},
only the emphasized part of the expression is statically typed:
\begin{lstlisting}[escapechar=+]
val context = +{\it document.getElementById("canvas")}+.geContext()
\end{lstlisting}

We can then complete the static typing by declaring types for
\code{Canvas} and \code{Context} as seen in the typed API section.

In our gradual typing scheme, an expression is either completely
statically typed or completely dynamically typed. Once \code{document}
is declared as \code{Rep[Element]} instead of \code{DynamicRep}, it is
a type error to call an arbitrary method which is not part of its
declared \code{ElementOps} interface. If needed, it is always possible to
explicitly move from a statically-typed expression to a
dynamically-typed one by wrapping it in a \code{dynamic} call.

\section{Sharing Code between Client and Server}\label{sec:trivial-embedding} % adriaan: rename without trivial
In addition to generating JavaScript / client-side code, we want to be able to re-use our DSL code on the Scala / server-side. In the LMS approach, the DSL uses the abstract type constructor \code{Rep}~\cite{higher-kind}. When generating JavaScript, this abstract type constructor is defined by IR nodes. Another definition, which we dub ``trivial embedding'', is to use the identity type constructor: \code{Rep[T] = T}. By stripping out \code{Rep}s in this way, our DSL can operate on concrete Scala types, replacing staging with direct evaluation. Even in the trivial embedding, when the DSL code operates on concrete Scala types, virtualization still occurs because the usage layer of the DSL is still in terms of abstract \code{Rep}s.

As an example, consider the following DSL snippet, which computes the absolute value:
\begin{lstlisting}
trait Ex extends JS {
  def abs(x: Rep[Int]) = if (x < 0) -x else x
}
\end{lstlisting}
We can use this DSL snippet to generate JavaScript code:
\begin{lstlisting}
    new Ex with JSExp { self =>
      val codegen = new JSGen { ... }
      codegen.emitSource(abs _, "abs", ...)
    }
\end{lstlisting}

We can also use this DSL snippet directly in Scala via the trivial embedding (defined by \code{JSInScala}):
\begin{lstlisting}
    new Ex with JSInScala { self =>
      println(abs(-3))
    }
\end{lstlisting}

In the JavaScript example, when \code{abs} gets evaluated, through mixing in \code{JSExp}, it results in an IR tree roughly equivalent to {\tt\small If(LessThan(Sym("x"), Const(0)), Neg(Sym("x")), Sym("x"))}. In the trivial embedding, when \code{abs(-3)} gets called, it evaluates to \code{3} by executing the virtualized \code{if} as a normal condition. In short, in the trivial embedding, the virtualized method calls are evaluated in-place without constructing IR nodes.

In the previous section, we showed how to define typed APIs to represent JavaScript external libraries or dependencies. In the trivial embedding, we need to give an interpretation to these APIs. For example, we can implement a Canvas context in Scala by using native graphics to draw on a desktop widget instead of a web page. We translated David Flanagan's Canvas example from the book ``JavaScript: The Definitive Guide''~\cite{flanagan}, which draws Koch snowflakes on a canvas~\cite{koch-code}. First, the translation from JavaScript to our DSL is straightforward: the code looks the same except for some minor declarations. Then, from our DSL code, we can generate JavaScript code to draw the snowflakes on a canvas as in the original code. In addition, via the trivial embedding, we can execute the DSL code in Scala to draw snowflakes on a desktop widget. Screenshot presenting snowflakes rendered in a browser using HTML5 Canvas and Java's 2D are presented in figure \ref{fig:snowflakes}.

\begin{figure}
\centering
\includegraphics[scale=0.4]{koch-browser}
\includegraphics[scale=0.4]{koch-java2d}
\caption{Snowflakes rendered using HTML5 Canvas and Java's 2D}
\label{fig:snowflakes}
\end{figure}

HTML5 Canvas is a standard that is not implemented by all browsers yet so a fall-back mechanism is needed to support users of older browsers. This can be achieved through the trivial embedding by drawing using Java's 2D API, saving the result as an image and sending it to the browser. The decision to either generate a JavaScript snippet that draws on canvas and send it to the browser, or render the image on the server can be made at runtime (e.g. after inspecting information about the client's browser). In the case of rendering on the server-side, one can store computation that renders an image using Java's graphics 2D in a hash map and send back to the client the key as an url for an image. When a browser makes a second request, computation can be restored from the hash map, executed and the result sent back to the browser. All of that is possible because the computation itself is expressed against an abstract DSL API so we can swap implementations to either generate JavaScript or run the computation at the server side. Moreover, our DSL is just a library and computations are expressed as first-class values in a host language so they can be passed around the server program, stored in a hash map, etc.

A drawback of the trivial embedding is that the virtualization overhead is incurred each time the program is evaluated. To avoid this, we could generate Scala code in the same way as we now generate JavaScript code, relying on a mechanism to incorporate the generated Scala code into the rest of the program. In fact, this approach is taken by the Delite framework.

%- DOM Canvas Koch example: generate both client-side code or desktop widget

\section{Modularity Interlude}\label{sec:modularity}

The design of our DSL supports modularity at many levels. We use
Scala's traits heavily to allow our DSL to be assembled from and
extended with components~\cite{components-scala}.

For example, the feature to escape into dynamic typing is implemented
as an independent component that can be mixed and matched with
others. Similarly, users specify external APIs as components. The
separation between the interface level and the implementation level is
also done by having distinct components for each level. This allows
the same DSL program to be interpreted in multiple ways, as has been
shown with the trivial embedding to Scala.

As the features available in a DSL program are specified by composing
components, it is possible to use this mechanism to enforce that a
subprogram only uses a restricted set of features. For example, worker
threads in JavaScript (``WebWorkers'') are not allowed to manipulate
the DOM. This can be enforced by not mixing in the DOM component in
the subprogram for a worker thread.

The code generation level is assembled from components as in the
interface and implementation levels. Furthermore, optimizations in
code generation can be implemented as optional components to be mixed
in.

\section{Reification of Objects}\label{sec:reification}

By exploiting staging, the generated code can remain simple and
relatively unstructured as many of the high-level constructs can be
evaluated away at staging time. However, it is sometimes useful to be
able to reify more complex structures. For example, APIs sometimes
expect arguments or return results as object literals. Therefore, we
support a few type-safe ways to create more complex staged structures,
which we explain below.

\subsection{Functions}\label{sec:functions}

A function in the host language acting on reified types has type:
\code{Rep[A] => Rep[B]}. For example, the function \code{inc} has type
\code{Rep[Int] => Rep[Int]}:
\begin{lstlisting}
val inc = (x: Rep[Int]) => x + 1
\end{lstlisting}

Invoking such a function at staging time simply inlines the call:
\code{inc(4*x)} results in \code{4*x + 1}. This is useful and nice,
because it removes abstraction overhead from the generated
code.

We also want the ability to treat functions as first-class staged
values, since JavaScript supports them. In order to do this, we
provide a higher-order function \code{fun} which takes a function of
type \code{Rep[A] => Rep[B]} and converts\footnote{We refer the reader
  to ~\cite{lms} for implementation details} it to a staged function
of type \code{Rep[A => B]}. For example, if we define \code{inc} in
the following way, its type is \code{Rep[Int => Int]}:
\begin{lstlisting}
val inc = fun { (x: Rep[Int]) => x + 1 }
\end{lstlisting}
Calling \code{inc(4*x)} results in an \code{Apply} IR node. We
actually generate JavaScript code for the staged \code{inc} function,
while we didn't for the unstaged one, since it's inlined at every call
site during staging. The generated code looks roughly like the
following:
\begin{lstlisting}[language=JavaScript]
var inc = function(a) {
  return a+1
}
inc(4*x)
\end{lstlisting}

First-class functions are widely used in JavaScript. One particular
common case is for callback-driven programming. Therefore, staged
functions are important to interface with existingn libraries. They
will also play a crucial role in section~\ref{sec:cps}, where we
abstract over callback-driven programming.

\subsection{Typed Object Literals}\label{sec:object-literals}

Our DSL provides typed immutable object literals to represent
JavaScript object literals. As an example:
\begin{lstlisting}
val o = new JSLiteral {
  val a = 1
  val b = a + 1
}
\end{lstlisting}
\code{o} has type \code{Rep[JSLiteral \{val a: Int; val b: Int\}]}. All
the fields of \code{o} are \code{Rep}s, so \code{o.a} has type
\code{Rep[Int]}. The translation to JavaScript is straightforward:
\begin{lstlisting}
var o = { a : 1, b : 2 }
\end{lstlisting}
This straightforward translation makes it possible to pass the typed
object literals of our DSL to JavaScript functions which expects
object literals, such as \code{JSON.stringify} or jQuery's \code{css}.

As for implementation, notice that the type of a \code{new JSLiteral
  \{...\}} expression is not \code{JSLiteral \{...\}} but
\code{Rep[JSLiteral \{...\}]}. This is achieved with support from the
Scala-Virtualized compiler. JSLiteral inherits from a special marker
trait, which indicates its \code{new} expressions should be
reified. So the \code{new} expression is turned into a method call,
with information about all the field definitions. A complication is
that a field definition might reference another field being defined
(such as \code{b} being defined in terms of \code{a} in the example
above). So each definition is represented by its name and a {\it
  function} which takes a self type for the object literal. These
definition functions are evaluated in an order which allows the self
references to be resolved.

\subsection{Classes}
Our DSL also supports reified classes. For convenience, these are
defined as traits, use the \code{repProxy} mechanism underlying typed
APIs and are also implemented using reflection. The translation to
JavaScript, based on constructors and prototypes, is
straightforward. Through the trivial embedding, classes implemented in
our DSL can be used on both the server and client sides.

% contribution to LMS, completeness
%- JS Literals: intuitive translation to JS as useful JSON-like objects, static typing
%- mice example: shows use of JSLiterals + API gluing (create abstractions that you need, not necessarily exactly matching the underlying JS API)
%- reification of traits: mutability, inheritance

\section{CPS Transformation for Asynchronous Code Patterns}\label{sec:cps}
  A callback-driven programming style is pervasive in JavaScript programs. Because of lack of thread support, callbacks are used for I/O, scheduling and event-handling. For example, in an Ajax call, one has to provide a callback that will be called once the requested data arrives from the server. This style of programming is known to be unwieldy in more complicated scenarios. To give a specific example, let's consider a scenario where we have an array of Twitter account names and we want to ask Twitter for the latest tweets of each account. Once we obtain the tweets of all users, we would like to log that fact in a console.

We'll implement this program both in JavaScript and in our DSL. Let's start by implementing logic that fetches tweets for a single user by using the jQuery library for Ajax calls (listings~\ref{code:js:twitter_fetch}~\&~\ref{code:twitter_fetch}).

\lstset{basicstyle=\ttfamily\scriptsize}
\begin{figure}
\setlength{\columnseprule}{0.5pt}
\begin{changemargin}{-1.5cm}{-1.5cm}
\begin{multicols}{2}
\lstinputlisting[language=JavaScript,caption=Fetching tweets in JavaScript,label=code:js:twitter_fetch]{fetchTweets.js}
\lstinputlisting[escapechar=|,caption=Fetching tweets in DSL,label=code:twitter_fetch]{fetchTweets.scala}
\end{multicols}
\end{changemargin}
\end{figure}

Note that JavaScript version takes a callback as second argument that will be used to process the fetched tweets. We provide the rest of the logic that iterates over array of users and makes Ajax requests (listings~\ref{code:js:twitter_example}~\&~\ref{code:twitter_example}).

\begin{figure}
\begin{changemargin}{-0.7cm}{-0.7cm}
\setlength{\columnseprule}{0.5pt}
\begin{multicols}{2}
\lstinputlisting[language=JavaScript,caption=Twitter example in JavaScript,label=code:js:twitter_example]{twitter.js}
\lstinputlisting[escapechar=|,caption=Twitter example in DSL,label=code:twitter_example]{twitter.scala}
\end{multicols}
\end{changemargin}
\end{figure}

\lstset{basicstyle=\ttfamily}

Because of the inverted control flow of callbacks, synchronization between callbacks has to be handled manually. Also, the inverted control flow leads to a code structure that is distant from the programmer's intent. Notice that the in JavaScript version, the call to console that prints ``done" is put inside of the foreach loop. If it was put it after the loop, we would get ``done'' printed \emph{before} any Ajax call has been made leading to counterintuitive behaviour.

As an alternative to the callback-driven programming style, one can use an explicit monadic style, possibly sugared by a Haskell-like ``do''-notation. However, this requires rewriting possibly large parts of a program into monadic style when a single async operation is added. Another possibility is to automatically transform the program into continuation passing style (CPS), enabling the programmer to express the algorithm in a straightforward, sequential way and creating all the necessary callbacks and book-keeping code automatically. Links~\cite{links} uses this approach. However, a whole-program CPS transformation can cause performance degradation, code size blow-up, and stack overflows. In addition, it is not clear how to interact with existing non-CPS libraries as the whole program needs to adhere to the CPS style. We suggest using a {\it selective} CPS transformation, which precisely identifies what needs to be CPS transformed.

In fact, the Scala compiler already does selective, \code{@suspendable} type-annotation-driven CPS transformations of Scala programs~\cite{cps-scala,danvy90,danvy92}. We show how this mechanism can be used for transforming our DSL code before staging and stripping out most CPS abstractions at staging time. The generated JavaScript code doesn't have any CPS-specific code but is written in CPS-style by use of JavaScript anonymous functions.

\subsection{CPS in Scala}
Before presenting how CPS transformations are used in our DSL, let's consider a typical situation where CPS rewrites act on Scala programs.

As an example, we'll consider a \code{sleep} method implemented in non-blocking, asynchronous style. This is useful, for example, when using \code{ThreadPool}s as no thread is being blocked during the sleep period. Let's see how our \code{sleep} method written in CPS can be used:

\begin{lstlisting}
def foo() = {
  sleep(1000)
  println("Called foo")
}
reset {
  println("look, Ma ...")
  foo()
  sleep(1000)
  println(" no threads!")
}
\end{lstlisting}

The \code{reset} delimits the scope of CPS rewrite. Let's see how the rewrite itself looks like:

\begin{lstlisting}
def foo(): ControlContext = {
  sleep(1000).map((tmp1: Unit) => println("Called foo"))
}
reset {
  println("look, Ma ...")
  foo().flatMap((tmp2: Unit) =>
    sleep(1000).map((tmp3: Unit) => println(" no threads!"))
  )
}
\end{lstlisting}

There are a few things worth noting here. First, the return type of \code{foo} method is rewritten to be \code{ControlContext}\footnote{The \code{ControlContext} class implements the continuation monad and is provided by Scala's standard library. \code{ControlContext} is similar to C\#'s \code{Task<T>}, but more general.}. This is due to the fact that \code{sleep} is used in the body of \code{foo}. Also, note that the code to be executed after sleeping is captured as a continuation (anynomous function) and passed to the \code{ControlContext} through a call of the \code{map} method. The body of the \code{reset} block is rewritten in a similar vein.

Now, let's have a closer look how \code{sleep} itself is implemented:

\begin{lstlisting}[label=code:cps_scala]
import java.util.{Timer,TimerTask}
val timer = new Timer
def sleep(delay: Int): Unit @suspendable = shift { k =>
  val task = new TimerTask { def run() = k() }
  timer.schedule(task, delay.toLong)
}
\end{lstlisting}

Notice the \code{@suspendable} type annotation attached to \code{sleep}'s return type \code{Unit}. The \code{@suspendable} annotation means that the \code{sleep} method can be used in a side-effecting continuation context. In the definition of the \code{sleep} method, we use Java's \code{Timer} and \code{TimerTask} abstractions for asynchronous, delayed task execution. The \code{TimerTask} interface has one method, \code{run}, that will be executed after a specified period of time. The \code{shift} control abstraction allows us to capture the continuation as a first-class value and then use it in the body of the \code{run} method. Both the \code{reset} and \code{shift} control abstractions are described in detail in~\cite{danvy90}.

After the CPS transformation, the code presented above becomes
\begin{lstlisting}
def sleep(time: Int): ControlContext =
  shiftR { k =>
    val task = new TimerTask { def run() = k() }
    timer.schedule(task, delay.toLong)
  }
\end{lstlisting}

After the rewriting, all CPS-related type annotations are dropped and use of the \code{ControlContext} class that supports continuation passing is introduced. The \code{shiftR} method is an internal method that takes a block (which itself is a function) and wraps it in \code{ControlContext} structure.

\subsection{CPS and Staging}
Let's write the example from listing \ref{code:cps_scala} in our DSL. We need to define \code{sleep} to use JavaScript's \code{setTimeout} as a replacement for the \code{Timer} abstraction.
\begin{lstlisting}
def sleep(delay: Rep[Int]) = shift { k: (Rep[Unit]=>Rep[Unit]) =>
  window.setTimeout(fun(k), delay)
}
\end{lstlisting}

The \code{setTimeout} method expects an argument of type \code{Rep[Unit=>Unit]} which denotes a \emph{representation} of a function of type \code{Unit=>Unit}. The \code{shift} method offers us a function of type \code{Rep[Unit] => Rep[Unit]}, so we need to reify it to obtain the desired representation. The reification is achieved by the \code{fun} function (described in \ref{sec:functions}) provided by our framework and performed at staging time.

It's important to note that reification preserves function composition. Specifically, let \code{f: Rep[A] => Rep[B]} and \code{g: Rep[B] => Rep[C]} then {\tt\small fun(g compose f) == (fun(g) compose fun(f))} where we consider two reified functions to be equal if they yield the same observable effects at runtime. That property of function reification is at the core of reusing the continuation monad in our DSL. Thanks to the fact that the continuation monad composes functions, we can just insert reification at some places (like in a \code{sleep}) and make sure that we reify \emph{effects} of the continuation monad without the need to reify the monad itself.

\subsection{CPS for Interruptible Traversals}
We need to be able to interrupt our execution while traversing an array in order to implement functionality from listing \ref{code:twitter_example}. Let's consider a simplified example where we want to iterate over an array and sleep during each iteration. We present both code written in JavaScript and our DSL that achieves that (listings~\ref{code:js:sleep_iter}~\&~\ref{code:sleep_iter}).
\begin{figure}
\lstset{basicstyle=\ttfamily\scriptsize}
\begin{changemargin}{-0.7cm}{-0.7cm}
\setlength{\columnseprule}{0.5pt}
\begin{multicols}{2}
\begin{lstlisting}[caption=JavaScript,label=code:js:sleep_iter]
var xs = [1, 2, 3]
var i = 0
var msg = null
function f1() {
  if (i < xs.length) {
    window.setTimeout(f2, xs[i]*1000)
    msg = xs[i]
    i++
  }
}
function f2() {
  console.log(msg)
  f1()
}
f1()
\end{lstlisting}
\begin{lstlisting}[escapechar=|,caption=DSL,label=code:sleep_iter]
val xs = array(1, 2, 3)
// shorthand for xs.suspendable.foreach
for (x <- xs.suspendable) {
  sleep(x * 1000)
  console.log(String.valueOf(x))
}
log("done")

|$\quad$|
\end{lstlisting}
\end{multicols}
\end{changemargin}
\lstset{basicstyle=\ttfamily}
\end{figure}

Both programs, when executed, will print following to the JavaScript console:
\begin{lstlisting}
//pause for 1s
1
//pause for 2s
2
//pause for 3s
3
done
\end{lstlisting}

In the DSL code, we use a \code{suspendable} variant of arrays, which is achieved through an implicit conversion from regular arrays:
\begin{lstlisting}
implicit def richArray(xs: Rep[Array[A]]) = new {
  def suspendable: SArrayOps[A] = new SArrayOps[A](xs)
}
\end{lstlisting}
The idea behind \code{suspendable} arrays is that iteration over them can be interrupted. We'll have a closer look at how to achieve that with the help of CPS. The \code{suspendable} method returns a new instance of the \code{SArrayOps} class defined here:

\begin{lstlisting}[caption=Suspendable foreach,label=code:suspendable_foreach]
class SArrayOps(xs: Rep[Array[A]]) {
  def foreach(yld: Rep[A] => Rep[Unit] @suspendable):
    Rep[Unit] @suspendable = {
      var i = 0
      suspendableWhile(i < xs.length) { yld(xs(i)); i += 1 }
    }
}
\end{lstlisting}
Note that one cannot use while loops in CPS but we can simulate them with recursive functions. Let's see how regular while loop can be simulated with a recursive function:
\begin{lstlisting}
def recursiveWhile(cond: => Boolean)(body: => Unit): Unit = {
  def rec = () => if (cond) { body; rec() } else {}
  rec()
}
\end{lstlisting}

By adding CPS-related declarations and control abstractions, we implement \code{suspendableWhile}:

\begin{lstlisting}
def suspendableWhile(cond: => Rep[Boolean])(
  body: => Rep[Unit] @suspendable): Rep[Unit] @suspendable =
  shift { k =>
    def rec = fun { () =>
      if (cond) reset { body; rec() } else { k() }
    }
    rec()
  }
\end{lstlisting}

\subsection{Defining the Ajax API}

With the abstractions for interruptible loops and traversals at hand,
what remains to complete the Twitter example from the beginning of the section
is the actual Ajax request/response cycle.

The Ajax interface component provides a type \code{Request} that captures the 
request structure expected by the underlying JavaScript/jQuery implementation and the
necessary object and method definitions to enable the use of \code{ajax.get} in user
code:
\begin{lstlisting}
trait Ajax extends JS with CPS {
  type Request = JSLiteral {
    val url: String
    val `type`: String
    val dataType: String
    val data: JSLiteral
  }
  type Response = Any
  object ajax {
    def get(request: Rep[Request]) = ajax_get(request)
  }
  def ajax_get(request: Rep[Request]): Rep[Response] @suspendable
}
\end{lstlisting}

Notice that the \code{Request} type is flexible enough to support an arbitrary object literal type for the \code{data} field through subtyping. The \code{Response} type alias points at \code{Any} which means that it is the user's responsibility to either use \code{dynamic} or perform an explicit cast to the expected data type.

The corresponding implementation component implements \code{ajax_get} to capture a continuation, reify it as a staged function using \code{fun} and store it in an \code{AjaxGet} IR node. 
\begin{lstlisting}
trait AjaxExp extends JSExp with Ajax {
  case class AjaxGet(request: Rep[Request],
    success: Rep[Response => Unit]) extends Def[Unit]
  def ajax_get(request: Rep[Request]): Rep[Response] @suspendable = 
    shift { k =>
      reflectEffect(AjaxGet(request, fun(k)))
    }
}
\end{lstlisting}

During code generation, we emit code to attach the captured continuation as a callback function in the \code{success} field of the request object:
\begin{lstlisting}
trait GenAjax extends JSGenBase {
  val IR: AjaxExp
  import IR._
  override def emitNode(sym: Sym[Any], rhs: Def[Any])(
    implicit stream: PrintWriter) = rhs match {
      case AjaxGet(req, succ) => 
        stream.println(quote(req) + ".success = " + quote(succ)) 
        emitValDef(sym, "jQuery.ajax(" + quote(req) + ")")
    case _ => super.emitNode(sym, rhs)
  }
}
\end{lstlisting}

It's interesting to note that, since the request already has the right structure for the \code{jQuery.ajax} function, we can simply pass it to the framework-provided \code{quote} method, which knows how to generate JavaScript representations of any \code{JSLiteral}.

The Ajax component completes the functionality required to run the Twitter example with one caveat:
The outer loop in listing \ref{code:twitter_example} uses \code{parSuspendable} to traverse arrays instead of the \code{suspendable} traversal variant we have defined in listing \ref{code:suspendable_foreach}.

In fact, if we change the code to use \code{suspendable} instead of \code{parSuspendable} and run the generated JavaScript program, we'll get following output printed to the JavaScript console:
\begin{lstlisting}
fetching gkossakowski
finished fetching gkossakowski
fetched [...]
fetched [...]
fetching odersky
finished fetching odersky
fetched [...]
fetched [...]
fetching adriaanm
finished fetching adriaanm
fetched [...]
fetched [...]
done
\end{lstlisting}

Notice that all Ajax requests were done sequentially. Specifically,
there was just one Ajax request active at a given time; when the
callback to process one request is called, it would resume the
continuation to start another request, and so on. In many cases this
is exactly the desired behavior, however, we will most likely
want to perform our Ajax request in parallel.

\subsection{CPS for Parallelism}
The goal of this section is to implement a parallel variant of the
\code{foreach} method from listing~\ref{code:suspendable_foreach}. We'll start 
with defining a few primitives like futures and dataflow cells. 
Let's start with cells, which we
decide to define in JavaScript, as another example of
integrating external libraries with our DSL:

\begin{lstlisting}[language=JavaScript,caption=JavaScript-based implementation of a non-blocking Cell]
function Cell() {
  this.value = undefined
  this.isDefined = false
  this.queue = []
  this.get = function (k) {
    if (this.isDefined) {
      k(this.value)
    } else {
      this.queue.push(k)
    }
  }
  this.set = function (v) {
    if (this.isDefined) {
      throw "can't set value twice"
    } else {
      this.value = v
      this.isDefined = true
      this.queue.forEach(function (f) { 
        f(v) //non-trivial spawn could be used here
      })
    }
  }
}
\end{lstlisting}

A cell object allows us to track dependencies between values. Whenever the \code{get} method is called and the value is not in the cell yet, the continuation will be added to a queue so it can be suspended until the value arrives. The \code{set} method takes care of resuming queued continuations. We expose \code{Cell} as external library using our typed API mechanism and we use it for implementing an abstraction for futures.

\begin{lstlisting}
def createCell(): Rep[Cell[A]]
trait Cell[A]
trait CellOps[A] {
  def get(k: Rep[A => Unit]): Rep[Unit]
  def set(v: Rep[A]): Rep[Unit]
}
implicit def repToCellOps(x: Rep[Cell[A]]): CellOps[A] =
  repProxy[Cell[A],CellOps[A]](x)

def spawn(body: => Rep[Unit] @suspendable): Rep[Unit] = {
  reset(body) //non-trivial implementation uses
              //trampolining to prevent stack overflows
}
def future(body: => Rep[A] @suspendable) = {
  val cell = createCell[A]()
  spawn { cell.set(body) }
  cell
}
\end{lstlisting}

The last bit of general functionality we need is \code{RichCellOps} that ties \code{Cell}s and continuations together inside of our DSL.

\begin{lstlisting}
class RichCellOps(cell: Rep[Cell[A]]) {
  def apply() = shift { k: (Rep[A] => Rep[Unit]) =>
    cell.get(fun(k))
  }
}
implicit def richCellOps(x: Rep[Cell[A]]): RichCell[A] =
  new RichCellOps(x)
\end{lstlisting}

It's worth noting that \code{RichCellOps} is not reified so it will be dropped at staging time and its method will get inlined whenever used. Also, it contains CPS-specific code that allows us to capture the continuation. The \code{fun} function reifies the captured continuation.

We are ready to present the parallel version of \code{foreach} defined in listing \ref{code:suspendable_foreach}.

\begin{lstlisting}
def foreach(yld: Rep[A] => Rep[Unit] @suspendable):
  Rep[Unit] @suspendable = {
    val futures = xs.map(x => future(yld(x)))
    futures.suspendable.foreach(_.apply())
  }
\end{lstlisting}

We instantiate each future separately so they can be executed in parallel. As a second step we make sure that all futures are evaluated before we leave the \code{foreach} method by forcing evaluation of each future and ``waiting'' for its completion. Thanks to CPS transformations, all of that will be implemented in a non-blocking style.

The only difference between the parallel and serial versions of the Twitter example~\ref{code:twitter_example} is the use of \code{parSuspendable} instead of \code{suspendable} so the parallel implementation of the \code{foreach} method is used. The rest of the code stays the same. It's easy to switch between both versions, and users are free to make their choice according to their needs and performance requirements.

%Parallelism doesn't come for free as it involves more CPS rewritings and involves indirections introduced by our abstractions. However, since it's easy to switch between both versions, users are free to make those choices according to their needs and performance requirements.

\section{Evaluation}\label{sec:eval}

We've implemented our DSL in Scala~\footnote{The code can be found at
  \url{http://github.com/namin/lms-sandbox}.}. We used our DSL to
develop a few web applications, which are simple but not
trivial. First, we implemented a few drawing examples like the
snowflakes of figure~\ref{fig:snowflakes}. We extended the Twitter
example from section~\ref{sec:cps}, which presents the latest tweets
from selected users in an interactive way. In order to do so, we
incorporated a useful subset of the DOM API and jQuery library using our
typed APIs.

\begin{wrapfigure}{r}{5cm}
\includegraphics[scale=0.2]{drawing}
\end{wrapfigure}
Finally, we developed a collaborative drawing application, which
includes a server-side component (implemented using the Jetty web
server). We use web sockets to communicate between the server and
clients. Each client transmits drawing commands to the server, which
broadcasts them to all the clients. When a new client joins, the
server sends the complete drawing history to the new client, and the
client reconstructs the image by playing the commands. A very
simple improvement is to make the server execute the drawing
commands as well, and keep an up-to-date bitmap of the drawing -- this
can easily be achieved by using the trivial embedding described in
section~\ref{sec:trivial-embedding}. New clients then just obtain the
bitmap instead of replaying the history, which is potentially be large.

\section{Conclusion}\label{sec:conclusion}

In this paper, we have shown how to embed a JavaScript DSL in Scala
using LMS. A recurring theme of our approach is to exploit the
features of the host language to enhance the DSL with minimal
effort. Moreover, through staging, we can use many abstractions at the
code-generation stage, without complexity and performance overhead in
the generated code. We believe this approach addresses some important
challenges of developing rich web applications.

\section{Acknowledgments}
We thank Adriaan Moors for maintaining the Scala-Virtualized compiler, adding our feature requests, fixing our reported bugs, and for insightful discussions.

\bibliographystyle{abbrv}
\bibliography{refs,ppl}

\end{document}

### 2012-03-20 papers/ecoop2012/fetchTweets.js

@@ -1,0 +1,15 @@
+function fetchTweets(username, callback) {
+  jQuery.ajax({
+    url: 
+"http://api.twitter.com/1/statuses/user_timeline.json/",
+    type: "GET",
+    dataType: "jsonp",
+    data: {
+      screen_name : username,
+      include_rts : true,
+      count : 5,
+      include_entities : true
+    },
+    success: callback
+  })
+}

### 2012-03-20 papers/ecoop2012/fetchTweets.scala

@@ -1,0 +1,16 @@
+def fetchTweets(username: Rep[String]) = (ajax.get {
+  new JSLiteral {
+    val url = 
+"http://api.twitter.com/1/statuses/user_timeline.json"
+    val `type` = "GET"
+    val dataType = "jsonp"
+    val data = new JSLiteral {
+      val screen_name = user
+      val include_rts = true
+      val count = 5
+      val include_entities = true
+    }
+  }
+}).as[TwitterResponse]
+
+type TwitterResponse = Array[JSLiteral {val text: String}]

### 2012-03-20 papers/ecoop2012/twitter.js

@@ -1,0 +1,15 @@
+var processed = 0
+var users = ["gkossakowski", "odersky", "adriaanm"]
+users.forEach(function (user) {
+  console.log("fetching " + user)
+  fetchTweets(user, function(data) {
+    console.log("finished fetching " + user)
+    data.forEach(function (tweet) {
+      console.log("fetched " + tweet.text)
+    })
+    processed += 1
+    if (processed == users.length) {
+      console.log("done")
+    }
+  })
+})

### 2012-03-20 papers/ecoop2012/twitter.scala

@@ -1,0 +1,16 @@
+val users = array("gkossakowski", "odersky", "adriaanm")
+for (user <- users.parSuspendable) {
+  console.log("fetching " + user)
+  val tweets = fetchTweets(user)
+  console.log("finished fetching " + user)
+  for (t <- tweets)
+    console.log("fetched " + t.text)
+}
+console.log("done")
+
+
+
+
+
+
+|$\quad$|

### 2012-03-21 papers/ecoop2012/article.tex
minor typos from official feedback

@@
 Examples include WebDSL~\cite{webdsl}, Links~\cite{links,links-form} and Dart~\cite{dart}.
-However, this approach usually requires a lot effort in terms of language and compiler design, and tooling support, although WebDSL leverages Spoofax~\cite{spoofax} to alleviate this effort.
+However, this approach usually requires a lot of effort in terms of language and compiler design, and tooling support, although WebDSL leverages Spoofax~\cite{spoofax} to alleviate this effort.
 Furthermore, it is not always clear how these languages interact with other languages on the server-side or with the existing JavaScript ecosystem on the client-side.
@@
 structures such as \code{if} can be reified into method calls, so that
-alternative implementation can be provided. In our case, we provide an
+alternative implementations can be provided. In our case, we provide an
 implementation of \code{if} that constructs an IR node instead of
@@
 \code{Exp}. \code{Exp} represents an IR via two subclasses:
-\code{Const} for constants, \code{Sym} for named values defining a
+\code{Const} for constants and \code{Sym} for named values defining a
 \code{Def}. \code{Def} is the base class for all IR nodes. In our
@@
 \begin{lstlisting}
-val context = document.getElementById("canvas").as[Canvas].geContext()
+val context = document.getElementById("canvas").as[Canvas].getContext()
 \end{lstlisting}
@@
 each method call, as shown in the walkthrough
-in~\ref{sec:defineDslComponent}. For example,
+in section~\ref{sec:defineDslComponent}. For example,
 \code{document.getElementById("canvas")} becomes the IR node {\tt\small
@@
 element, we need to cast it to a \code{Canvas} using
-\code{as[Canvas]}. \code{as} is implemented simply as a cast in the
+\code{as[Canvas]}. The \code{as} operation is implemented simply as a cast in the
 host language (no IR node is created):
@@
 typed API for each external JavaScript library. In addition, for
-some library, it might awkward to come up with such a statically-typed
+some library, it might be awkward to come up with such a statically-typed
 interface. In general, we expect users to start with a dynamic API for
@@
 \begin{lstlisting}[escapechar=+]
-val context = document.getElementById("canvas").geContext()
+val context = document.getElementById("canvas").getContext()
 \end{lstlisting}
@@
 \begin{lstlisting}[escapechar=+]
-val context = +{\it document.getElementById("canvas")}+.geContext()
+val context = +{\it document.getElementById("canvas")}+.getContext()
 \end{lstlisting}
@@
 common case is for callback-driven programming. Therefore, staged
-functions are important to interface with existingn libraries. They
+functions are important to interface with existing libraries. They
 will also play a crucial role in section~\ref{sec:cps}, where we
@@
 \code{Rep[JSLiteral \{...\}]}. This is achieved with support from the
-Scala-Virtualized compiler. JSLiteral inherits from a special marker
+Scala-Virtualized compiler. \code{JSLiteral} inherits from a special marker
 trait, which indicates its \code{new} expressions should be
@@
 Note that JavaScript version takes a callback as second argument that will be used to process the fetched tweets.
-We provide the rest of the logic that iterates over array of users and makes Ajax requests (listings~\ref{code:js:twitter_example}~\&~\ref{code:twitter_example}).
+We provide the rest of the logic that iterates over an array of users and makes Ajax requests (listings~\ref{code:js:twitter_example}~\&~\ref{code:twitter_example}).
 
@@
 \subsection{CPS and Staging}
-Let's write the example from listing \ref{code:cps_scala} in our DSL. We need to define \code{sleep} to use JavaScript's \code{setTimeout} as a replacement for the \code{Timer} abstraction.
+Let's write the example from listing~\ref{code:cps_scala} in our DSL. We need to define \code{sleep} to use JavaScript's \code{setTimeout} as a replacement for the \code{Timer} abstraction.
 \begin{lstlisting}
@@
 \subsection{CPS for Interruptible Traversals}
-We need to be able to interrupt our execution while traversing an array in order to implement functionality from listing \ref{code:twitter_example}.
+We need to be able to suspend our execution while traversing an array in order to implement functionality from listing~\ref{code:twitter_example}.
 Let's consider a simplified example where we want to iterate over an array and sleep during each iteration.
@@
 
-Both programs, when executed, will print following to the JavaScript console:
+Both programs, when executed, will print the following to the JavaScript console:
 \begin{lstlisting}
@@
 The Ajax component completes the functionality required to run the Twitter example with one caveat:
-The outer loop in listing \ref{code:twitter_example} uses \code{parSuspendable} to traverse arrays instead of the \code{suspendable} traversal variant we have defined in listing \ref{code:suspendable_foreach}.
+The outer loop in listing~\ref{code:twitter_example} uses \code{parSuspendable} to traverse arrays instead of the \code{suspendable} traversal variant we have defined in listing~\ref{code:suspendable_foreach}.
 
@@
 The \code{set} method takes care of resuming queued continuations.
-We expose \code{Cell} as external library using our typed API mechanism and we use it for implementing an abstraction for futures.
+We expose \code{Cell} as an external library using our typed API mechanism and we use it for implementing an abstraction for futures.
 
@@
 
-We've implemented our DSL in Scala~\footnote{The code can be found at
-  \url{http://github.com/namin/lms-sandbox}.}. We used our DSL to
+We've implemented our DSL in Scala.\footnote{The code can be found at
+  \url{http://github.com/namin/lms-sandbox} We used our DSL to
 develop a few web applications, which are simple but not
@@
 section~\ref{sec:trivial-embedding}. New clients then just obtain the
-bitmap instead of replaying the history, which is potentially be large.
+bitmap instead of replaying the history, which is potentially large.
 

### 2012-03-21 papers/ecoop2012/fetchTweets.scala

@@ -5,7 +5,7 @@ def fetchTweets(username: Rep[String]) = (ajax.get {
     val `type` = "GET"
     val dataType = "jsonp"
     val data = new JSLiteral {
-      val screen_name = user
+      val screen_name = username
       val include_rts = true
       val count = 5
       val include_entities = true

### 2012-03-21 papers/ecoop2012/article.tex
minor code fixes

@@
   case class Log(msg: Exp[String]) extends Def[Unit]
-  def log(msg: Exp[String]): Exp[Unit] = reflectEffect(Log(s))
+  def log(msg: Exp[String]): Exp[Unit] = reflectEffect(Log(msg))
 }
@@
     i++
+  } else {
+    console.log("done")
   }
@@
 We've implemented our DSL in Scala.\footnote{The code can be found at
-  \url{http://github.com/namin/lms-sandbox} We used our DSL to
+  \url{http://github.com/namin/lms-sandbox}} We used our DSL to
 develop a few web applications, which are simple but not

### 2012-03-21 papers/ecoop2012/article.tex
contributions from rebuttal

@@
 This case-study demonstrates the fruitfulness of re-using host language features to enhance our embedded DSL.
+A novel insight is that CPS transforming a program generator allows it to generate code that is in CPS.
+\item The present work extends both the range of virtualizable language features and the LMS framework, which is beneficial to future DSL efforts taking the JavaScript work as a case study.
+Previous LMS embeddings had to define each operation explicitly (like in section~\ref{sec:defineDslComponent}).
+This paper contributes lifting of whole traits or classes (through the \code{repProxy} mechanism described and used in sections~\ref{sec:typedApis}\&~\ref{sec:reified-classes}), untyped or optionally typed operations (sections~\ref{sec:casting},~\ref{sec:dynamic}\&~\ref{sec:dynamic2static}), and typed object literals (section~\ref{sec:object-literals}) including the necessary Scala-Virtualized support.
 \end{itemize}
@@
 
-\subsection{Casting and Optional Runtime Type Checks}
+\subsection{Casting and Optional Runtime Type Checks}\label{sec:casting}
 The need for casting arises in a few contexts. One of them is the
@@
 
-\subsection{Scala Dynamic}
+\subsection{Scala Dynamic}\label{sec:dynamic}
 Since our DSL compiles to JavaScript, which is dynamically typed, it
@@
 
-\subsection{From Dynamic to Static}
+\subsection{From Dynamic to Static}\label{sec:dynamic2static}
 
@@
 
-\subsection{Classes}
+\subsection{Classes}\label{sec:reified-classes}
 Our DSL also supports reified classes. For convenience, these are

### 2012-03-21 papers/ecoop2012/article.tex
most minor comments done

@@
 }
+
+The generated code ressembles single-assignment form. The nested
+\code{for}-loop is desugared into a \code{flatMap} which generates the
+nested \code{for}-loop and the \code{splice} pattern concatenating the
+inner \code{x4} arrays into one \code{x6} array in the JavaScript
+code.\footnote{Obviously, the generated code can be optimized further.}
+
 \end{lstlisting}
@@
   MethodCall(document, "getElementById", List("canvas"))}. This is a
-mechanical transformation, implemented by \code{repProxy} using
+mechanical transformation, implemented by \code{repProxy}, a library method using
 reflection to intercept method calls and generate IR nodes based on
@@
 services. For example, to process data from an external service such
-as Twitter, we cast it to its expected type:
+as Twitter, we cast it to its expected type (an array of JSON objects, each with a field called \code{text}):
 \begin{lstlisting}
@@
 \code{DynamicRep} exploits a new Scala feature, based on a
-marker trait \code{Dynamic}: An expression whose type \code{T} is a
+special trait \code{Dynamic}: An expression whose type \code{T} is a
 subtype of \code{Dynamic} is subject to the following rewrites:
@@
 \code{Rep[JSLiteral \{...\}]}. This is achieved with support from the
-Scala-Virtualized compiler. \code{JSLiteral} inherits from a special marker
+Scala-Virtualized compiler. \code{JSLiteral} inherits from a special
 trait, which indicates its \code{new} expressions should be
@@
 
-There are a few things worth noting here.
-First, the return type of \code{foo} method is rewritten to be \code{ControlContext}\footnote{The \code{ControlContext} class implements the continuation monad and is provided by Scala's standard library.
-\code{ControlContext} is similar to C\#'s \code{Task<T>}, but more general.}.
-This is due to the fact that \code{sleep} is used in the body of \code{foo}.
-Also, note that the code to be executed after sleeping is captured as a continuation (anynomous function) and passed to the \code{ControlContext} through a call of the \code{map} method.
-The body of the \code{reset} block is rewritten in a similar vein.
+There are a few things worth noting here. First, the return type of
+\code{foo} method is rewritten to be
+\code{ControlContext}.\footnote{The \code{ControlContext} class
+  implements the continuation monad and is provided by Scala's
+  standard library. \code{ControlContext[A,B,C]} is similar to C\#'s
+  \code{Task<T>}, but more general. In our discussion, we omit the
+  type parameters for simplicity.} This is due to the fact that
+\code{sleep} is used in the body of \code{foo}. Also, note that the
+code to be executed after sleeping is captured as a continuation
+(anynomous function) and passed to the \code{ControlContext} through a
+call of the \code{map} method. The body of the \code{reset} block is
+rewritten in a similar vein.
 
@@
 
-\subsection{CPS for Interruptible Traversals}
+\subsection{CPS for Suspendable Traversals}
 We need to be able to suspend our execution while traversing an array in order to implement functionality from listing~\ref{code:twitter_example}.
@@
 \end{lstlisting}
-The idea behind \code{suspendable} arrays is that iteration over them can be interrupted.
+The idea behind \code{suspendable} arrays is that iteration over them can be suspended.
 We'll have a closer look at how to achieve that with the help of CPS.
@@
 \end{lstlisting}
-Note that one cannot use while loops in CPS but we can simulate them with recursive functions. Let's see how regular while loop can be simulated with a recursive function:
+Note that one cannot use while loops in CPS but we can simulate them
+with recursive functions. Let's see how a regular while loop can be
+simulated with a recursive function with call-by-name parameters:
 \begin{lstlisting}
@@
 
-With the abstractions for interruptible loops and traversals at hand,
+With the abstractions for suspendable loops and traversals at hand,
 what remains to complete the Twitter example from the beginning of the section

### 2012-03-26 papers/ecoop2012/article.tex
Add footnote explaining setTimeout.

@@
 \subsection{CPS and Staging}
-Let's write the example from listing~\ref{code:cps_scala} in our DSL. We need to define \code{sleep} to use JavaScript's \code{setTimeout} as a replacement for the \code{Timer} abstraction.
+Let's write the example from listing~\ref{code:cps_scala} in our DSL. We need to define \code{sleep} to use JavaScript's \code{setTimeout}
+\footnote{The \code{setTimeout} function executes asynchronously a function passed as an argument after specified delay.}
+as a replacement for the \code{Timer} abstraction.
 \begin{lstlisting}

### 2012-03-26 papers/ecoop2012/article.tex
Added a paragraph clarifying how CPS is achieved.
Added a paragraph that explains how we
implemented continuations in our dsl.

@@
 
+Our implementation of continuations support is an example of an interesting technique of applying selective cps
+transformation to embedded DSLs by means of a deep linguistic reuse, exploiting a host language feature to
+implement the corresponding DSL feature without (much) additional work. Interesting insight is that CPS
+transforming a program generator allows it to generate code that is in CPS.
+
 \subsection{CPS in Scala}

### 2012-03-26 papers/ecoop2012/article.tex
grammar

--- 2012-03-26 papers/ecoop2012/article.tex
@@
 
+Our implementation of continuations is an example of an interesting technique of applying selective cps
+transformation to embedded DSLs by means of a deep linguistic reuse, exploiting a host language feature to
+implement the corresponding DSL feature without (much) additional work. An interesting insight is that CPS
+transforming a program generator allows it to generate code that is in CPS.
+
 \subsection{CPS in Scala}

### 2012-03-26 papers/ecoop2012/article.tex
kept more precise formulation

--- 2012-03-26 papers/ecoop2012/article.tex
@@
 
+Our implementation to support continuations is an example of an interesting technique of applying selective cps
+transformation to embedded DSLs by means of a deep linguistic reuse, exploiting a host language feature to
+implement the corresponding DSL feature without (much) additional work. An interesting insight is that CPS
+transforming a program generator allows it to generate code that is in CPS.
+
 \subsection{CPS in Scala}

### 2012-03-26 papers/ecoop2012/article.tex
grammar

@@
 Let's write the example from listing~\ref{code:cps_scala} in our DSL. We need to define \code{sleep} to use JavaScript's \code{setTimeout}
-\footnote{The \code{setTimeout} function executes asynchronously a function passed as an argument after specified delay.}
+\footnote{The \code{setTimeout} function asynchronously executes a function passed as an argument after a specified delay.}
 as a replacement for the \code{Timer} abstraction.

### 2012-03-26 papers/ecoop2012/article.tex
no need for article. wow.

--- 2012-03-26 papers/ecoop2012/article.tex
@@
 Let's write the example from listing~\ref{code:cps_scala} in our DSL. We need to define \code{sleep} to use JavaScript's \code{setTimeout}
-\footnote{The \code{setTimeout} function executes asynchronously a function passed as an argument after specified delay.}
+\footnote{The \code{setTimeout} function asynchronously executes a function passed as argument after a specified delay.}
 as a replacement for the \code{Timer} abstraction.

### 2012-03-26 papers/ecoop2012/article.tex
Fixed problem with listing numbering.
Note: if you happen to have colon
in a label for your listing you need
to put it in curly braces, otherwise
latex gets confused.

--- 2012-03-26 papers/ecoop2012/article.tex
@@
 
-Our implementation to support continuations is an example of an interesting technique of applying selective cps
+Our implementation of continuations support is an example of an interesting technique of applying selective cps
 transformation to embedded DSLs by means of a deep linguistic reuse, exploiting a host language feature to
-implement the corresponding DSL feature without (much) additional work. An interesting insight is that CPS
+implement the corresponding DSL feature without (much) additional work. Interesting insight is that CPS
 transforming a program generator allows it to generate code that is in CPS.
@@
 
-\begin{lstlisting}[label=code:cps_scala]
+\begin{lstlisting}[label={code:cps_scala},caption={\code{sleep} implementation}]
 import java.util.{Timer,TimerTask}

### 2012-03-29 papers/ecoop2012/article.tex
change code url
Fix messed up listing.
Added a mention of play validators in evaluation.

@@
 }
+\end{lstlisting}
 
@@
 code.\footnote{Obviously, the generated code can be optimized further.}
-
-\end{lstlisting}
 
@@
 We've implemented our DSL in Scala.\footnote{The code can be found at
-  \url{http://github.com/namin/lms-sandbox}} We used our DSL to
+  \url{http://github.com/js-scala}} We used our DSL to
 develop a few web applications, which are simple but not
@@
 typed APIs.
+
+Also, we integrated our DSL with an existing web framework, Play 2.0.\footnote{http://www.playframework.org/}
+Our DSL can be used to define Play form validators that are executed both at the client and the server side.
+We managed to achieve that without deep changes to how the framework works internally proving that our
+DSL can be used as a library.
 

### 2012-03-29 papers/ecoop2012/article.tex
wrap play link in url

@@
 
-Also, we integrated our DSL with an existing web framework, Play 2.0.\footnote{http://www.playframework.org/}
+Also, we integrated our DSL with an existing web framework, Play 2.0.\footnote{\url{http://www.playframework.org}}
 Our DSL can be used to define Play form validators that are executed both at the client and the server side.

### 2012-03-29 papers/ecoop2012/article.tex
added a footnote about incorrectly calling our DSL JavaScript

@@
 
-We propose to embed JavaScript as a DSL in a host language.
-Through LMS, we tackle the challenges outlined above with minimal effort, as most of the work is off-loaded to the host language.
+We propose to embed JavaScript as a DSL in a host language.\footnote{Surely, the embedded language is not exactly JavaScript: as an embedded language, it naturally is a subset of Scala.
+However it is quite close to JavaScript: often one can take snippets of JavaScript code and use them in the DSL with minor syntactic tweaking (as demonstrated by the Snowflake example).} Through LMS, we tackle the challenges outlined above with minimal effort, as most of the work is off-loaded to the host language.
 In particular, we make the following contributions:

### 2012-03-29 papers/ecoop2012/article.tex
move the discussion of our contributions to LMS to the end of the intro

@@
 A novel insight is that CPS transforming a program generator allows it to generate code that is in CPS.
-\item The present work extends both the range of virtualizable language features and the LMS framework, which is beneficial to future DSL efforts taking the JavaScript work as a case study.
+\end{itemize}
+
+In section~\ref{sec:eval}, we describe our experience in using the DSL, and conclude in section~\ref{sec:conclusion}.
+
+The present work extends both the range of virtualizable language features and the LMS framework, which is beneficial to future DSL efforts taking the JavaScript work as a case study.
 Previous LMS embeddings had to define each operation explicitly (like in section~\ref{sec:defineDslComponent}).
 This paper contributes lifting of whole traits or classes (through the \code{repProxy} mechanism described and used in sections~\ref{sec:typedApis}\&~\ref{sec:reified-classes}), untyped or optionally typed operations (sections~\ref{sec:casting},~\ref{sec:dynamic}\&~\ref{sec:dynamic2static}), and typed object literals (section~\ref{sec:object-literals}) including the necessary Scala-Virtualized support.
-\end{itemize}
-
-In section~\ref{sec:eval}, we describe our experience in using the DSL, and conclude in section~\ref{sec:conclusion}.
 

### 2012-03-29 papers/ecoop2012/article.tex
citing csharp dynamic

@@
 
-We propose to embed JavaScript as a DSL in a host language.\footnote{Surely, the embedded language is not exactly JavaScript: as an embedded language, it naturally is a subset of Scala.
-However it is quite close to JavaScript: often one can take snippets of JavaScript code and use them in the DSL with minor syntactic tweaking (as demonstrated by the Snowflake example).} Through LMS, we tackle the challenges outlined above with minimal effort, as most of the work is off-loaded to the host language.
+We propose to embed JavaScript as a DSL in a host language.\footnote{Surely, the embedded language is not exactly JavaScript: it naturally is a subset of Scala, the host language.
+However, it is quite close to JavaScript: often one can take snippets of JavaScript code and use them in the DSL with minor syntactic tweaking (as demonstrated by the Snowflake example).} Through LMS, we tackle the challenges outlined above with minimal effort, as most of the work is off-loaded to the host language.
 In particular, we make the following contributions:
@@
 
-\code{DynamicRep} exploits a new Scala feature, based on a
+\code{DynamicRep} exploits a new Scala feature\footnote{C\# 4.0 type ``dynamic''~\cite{csharp-dynamic} can serve the same purpose.}, based on a
 special trait \code{Dynamic}: An expression whose type \code{T} is a

### 2012-03-29 papers/ecoop2012/article.tex
incorporating fsharp work in intro

@@
 Moreover, both approaches tend to not accommodate very well to different API design and programming styles seen in many existing JavaScript libraries.
+The F\# Web Tools~\cite{tomasp-fsharp} is another example of starting with an existing language, F\#, and compiling it to both the client and the server sides.
+This work has similar goals to ours but the underlying techniques are very different (explicit quotations vs LMS, monads vs cps).
+A drawback of this second approach is that the whole starting language needs to be translated to JavaScript: there is no easy or modular way to limit the scope of the source language.
 

### 2012-03-29 papers/ecoop2012/article.tex
minor grammar

@@
 Also, we integrated our DSL with an existing web framework, Play 2.0.\footnote{\url{http://www.playframework.org}}
-Our DSL can be used to define Play form validators that are executed both at the client and the server side.
+Our DSL can be used to define Play form validators that are executed on both the client and server sides.
 We managed to achieve that without deep changes to how the framework works internally proving that our

### 2012-03-30 papers/ecoop2012/article.tex
Add ref to section where snowflake is described.

@@
 We propose to embed JavaScript as a DSL in a host language.\footnote{Surely, the embedded language is not exactly JavaScript: it naturally is a subset of Scala, the host language.
-However, it is quite close to JavaScript: often one can take snippets of JavaScript code and use them in the DSL with minor syntactic tweaking (as demonstrated by the Snowflake example).} Through LMS, we tackle the challenges outlined above with minimal effort, as most of the work is off-loaded to the host language.
+However, it is quite close to JavaScript: often one can take snippets of JavaScript code and use them in the DSL with minor syntactic tweaking (as demonstrated by the Snowflake example described in section \ref{sec:trivial-embedding}).} Through LMS, we tackle the challenges outlined above with minimal effort, as most of the work is off-loaded to the host language.
 In particular, we make the following contributions:

### 2012-03-30 papers/ecoop2012/article.tex
Removed old and outdated comment.

@@
 
-\section{Sharing Code between Client and Server}\label{sec:trivial-embedding} % adriaan: rename without trivial
+\section{Sharing Code between Client and Server}\label{sec:trivial-embedding}
 In addition to generating JavaScript / client-side code, we want to be able to re-use our DSL code on the Scala / server-side.

### 2012-03-30 papers/ecoop2012/article.tex
Added footnote to functions reification equeation.

@@
 
-It's important to note that reification preserves function composition.
-Specifically, let \code{f: Rep[A] => Rep[B]} and \code{g: Rep[B] => Rep[C]} then {\tt\small fun(g compose f) == (fun(g) compose fun(f))} where we consider two reified functions to be equal if they yield the same observable effects at runtime.
+It's important to note that reification preserves function composition (roughly speaking it's homomorphism between DSL code and generated JavaScript code).
+Specifically, let \code{f: Rep[A] => Rep[B]} and \code{g: Rep[B] => Rep[C]} then {\tt\small fun(g compose f) == (fun(g) compose fun(f))}
+where we consider two reified functions to be equal if they yield the same observable effects at runtime
+\footnote{Note that composition on the left and the right hand side of the equation is not the same operation but they have the same observable runtime effect.}.
 That property of function reification is at the core of reusing the continuation monad in our DSL.

### 2012-03-30 papers/ecoop2012/article.tex
global pass, intro and minor changes in different places

@@
 considerable choice on the server-side, the client-side is tied to
-JavaScript, which poses considerable software engineering challenges,
+JavaScript, which poses substantial software engineering challenges,
 such as moving or sharing pieces of code between the environments. We
@@
 ``Inversion of control'' in asynchronous callback-driven programming leads to code with control structures that are difficult to reason about.
-A challenge is to introduce helpful abstractions without a big hit on performance and/or code size.
+Another challenge is to introduce helpful abstractions without a big hit on performance and/or code size.
 Communication between the server side and the client side aggravates the impedance mismatch: in particular, data validation logic needs to be duplicated on the client-side for interactivity and on the server-side for security.
@@
 Moreover, both approaches tend to not accommodate very well to different API design and programming styles seen in many existing JavaScript libraries.
-The F\# Web Tools~\cite{tomasp-fsharp} is another example of starting with an existing language, F\#, and compiling it to both the client and the server sides.
+%The F\# Web Tools~\cite{tomasp-fsharp} is another example of starting with an existing language, F\#, and compiling it to both the client and the server sides.
 This work has similar goals to ours but the underlying techniques are very different (explicit quotations vs LMS, monads vs cps).
 A drawback of this second approach is that the whole starting language needs to be translated to JavaScript: there is no easy or modular way to limit the scope of the source language.
-
-A third approach is to design a language than is a thin layer on top of JavaScript but provides some new features.
+The F\# Web Tools~\cite{tomasp-fsharp} are an interesting variation of this approach that employ F\# quotations to translate only
+parts of a program and allow programmers to define custom mappings for individual data types.
+
+A third approach is to design a language that is a thin layer on top of JavaScript but provides some new features.
 A prime example of this idea is CoffeeScript~\cite{coffeescript}.
@@
 
-We propose to embed JavaScript as a DSL in a host language.\footnote{Surely, the embedded language is not exactly JavaScript: it naturally is a subset of Scala, the host language.
-However, it is quite close to JavaScript: often one can take snippets of JavaScript code and use them in the DSL with minor syntactic tweaking (as demonstrated by the Snowflake example described in section \ref{sec:trivial-embedding}).} Through LMS, we tackle the challenges outlined above with minimal effort, as most of the work is off-loaded to the host language.
+We propose to embed JavaScript as a DSL in a host language.
+\footnote{Surely, the embedded language is not \emph{exactly} JavaScript: it naturally is a subset of Scala, the host language.
+However, it is quite close to JavaScript.
+Often one can take snippets of JavaScript code and use them in the DSL with minor syntactic tweaking, as demonstrated by the Snowflake example described in section \ref{sec:trivial-embedding}.}
+Through LMS (reviewed in section~\ref{sec:lms}), we tackle the challenges outlined above with minimal effort, as most of the work is off-loaded to the host language.
 In particular, we make the following contributions:
@@
 The translations to JavaScript are lightweight and intuitive: the object literals translate to JSON-like object literals and the class-based objects to JavaScript constructor-based objects (section~\ref{sec:reification}).
-\item With minimal effort, we exploit the selective CPS transform already existing in the host language to provide a compelling abstraction over asynchronous callback-driven programming in our DSL (section~\ref{sec:cps}).
-This case-study demonstrates the fruitfulness of re-using host language features to enhance our embedded DSL.
-A novel insight is that CPS transforming a program generator allows it to generate code that is in CPS.
+\item On top of the straightforward embedding, we implement advanced abstractions in the host language.
+With minimal effort, we exploit the selective CPS transform already existing in Scala to provide a compelling abstraction over asynchronous callback-driven programming in our DSL (section~\ref{sec:cps}).
+The new insight here is that CPS transforming a program generator allows it to generate code that is in CPS.
+This case-study demonstrates the fruitfulness of re-using existing host language features to enhance our embedded DSL.
 \end{itemize}
@@
 
-The present work extends both the range of virtualizable language features and the LMS framework, which is beneficial to future DSL efforts taking the JavaScript work as a case study.
-Previous LMS embeddings had to define each operation explicitly (like in section~\ref{sec:defineDslComponent}).
-This paper contributes lifting of whole traits or classes (through the \code{repProxy} mechanism described and used in sections~\ref{sec:typedApis}\&~\ref{sec:reified-classes}), untyped or optionally typed operations (sections~\ref{sec:casting},~\ref{sec:dynamic}\&~\ref{sec:dynamic2static}), and typed object literals (section~\ref{sec:object-literals}) including the necessary Scala-Virtualized support.
+In addition to the contributions above, the present work significantly extends the LMS framework, which is beneficial to future DSL efforts taking the JavaScript work as a case study.
+Previous LMS embeddings had to define each staged operation explicitly (like in section~\ref{sec:defineDslComponent}).
+This paper contributes lifting of whole traits or classes (through the \code{repProxy} mechanism described and used in sections~\ref{sec:typedApis}\&~\ref{sec:reified-classes}), untyped or optionally typed operations (sections~\ref{sec:casting},~\ref{sec:dynamic}\&~\ref{sec:dynamic2static}), and typed object
+literals (section~\ref{sec:object-literals}) including necessary language support of the 
+Scala-Virtualized~\cite{scala-virtualized} compiler.
 
 \section{Introduction to LMS}
+\label{sec:lms}
 
@@
 For \code{prog2}, notice that the \code{if} got transformed into an IR
-node. To achieve this, LMS uses an extension of Scala,
-Scala-Virtualized~\cite{scala-virtualized}, in which control
+node. To achieve this, LMS uses
+Scala-Virtualized~\cite{scala-virtualized}, a suite of minimal
+extensions to the regular Scala compiler, in which control
 structures such as \code{if} can be reified into method calls, so that
@@
 
-The generated code ressembles single-assignment form. The nested
+The generated code resembles single-assignment form. The nested
 \code{for}-loop is desugared into a \code{flatMap} which generates the
@@
 elimination by re-using the same symbol for identical definitions. We
-don't use it here, because \code{log} is a side-effecting operation,
-and we don't want to (re)move any such calls even if their message
+do not use the automatic conversion here, because \code{log} is a 
+side-effecting operation, and we do not want 
+to (re)move any such calls even if their message
 is the same.
@@
 
-We've implemented a component that generates JavaScript code that
+We have implemented a component that generates JavaScript code that
 asserts casts at runtime. It was fairly straightforward as the host
@@
 
-\code{DynamicRep} exploits a new Scala feature\footnote{C\# 4.0 type ``dynamic''~\cite{csharp-dynamic} can serve the same purpose.}, based on a
-special trait \code{Dynamic}: An expression whose type \code{T} is a
+\code{DynamicRep} exploits a new Scala feature\footnote{C\#'s type ``dynamic''~\cite{csharp-dynamic} can serve the same purpose.}, 
+based on a special trait \code{Dynamic}: An expression whose type \code{T} is a
 subtype of \code{Dynamic} is subject to the following rewrites:
@@
 
-In the JavaScript example, when \code{abs} gets evaluated, through mixing in \code{JSExp}, it results in an IR tree roughly equivalent to {\tt\small If(LessThan(Sym("x"), Const(0)), Neg(Sym("x")), Sym("x"))}.
+In the JavaScript example (when mixing in \code{JSExp}) evaluating \code{abs(x)} results in an IR tree roughly equivalent to {\tt\small If(LessThan(Sym("x"), Const(0)), Neg(Sym("x")), Sym("x"))}.
 In the trivial embedding, when \code{abs(-3)} gets called, it evaluates to \code{3} by executing the virtualized \code{if} as a normal condition.
@@
 
-A drawback of the trivial embedding is that the virtualization overhead is incurred each time the program is evaluated.
+A drawback of the trivial embedding is that although lightweight, virtualization is not completely free, and a small virtualization overhead is incurred each time the program is evaluated.
 To avoid this, we could generate Scala code in the same way as we now generate JavaScript code, relying on a mechanism to incorporate the generated Scala code into the rest of the program.
@@
 type \code{Rep[A] => Rep[B]} and converts\footnote{We refer the reader
-  to ~\cite{lms} for implementation details} it to a staged function
+  to~\cite{lms} for implementation details.} it to a staged function
 of type \code{Rep[A => B]}. For example, if we define \code{inc} in
@@
 actually generate JavaScript code for the staged \code{inc} function,
-while we didn't for the unstaged one, since it's inlined at every call
+while we did not for the unstaged one, since it is inlined at every call
 site during staging. The generated code looks roughly like the
@@
 As an alternative to the callback-driven programming style, one can use an explicit monadic style, possibly sugared by a Haskell-like ``do''-notation.
+This is the approach taken by F\# Web Tools~\cite{tomasp-fsharp}.
 However, this requires rewriting possibly large parts of a program into monadic style when a single async operation is added.
@@
 
-In fact, the Scala compiler already does selective, \code{@suspendable} type-annotation-driven CPS transformations of Scala programs~\cite{cps-scala,danvy90,danvy92}.
+In fact, the Scala compiler already does selective CPS transformations of Scala programs, driven by \code{@suspendable} type annotations ~\cite{cps-scala,danvy90,danvy92}.
 We show how this mechanism can be used for transforming our DSL code before staging and stripping out most CPS abstractions at staging time.
-The generated JavaScript code doesn't have any CPS-specific code but is written in CPS-style by use of JavaScript anonymous functions.
+The generated JavaScript code does not contain any CPS-specific code but is written in CPS-style by use of JavaScript anonymous functions.
 
@@
 transformation to embedded DSLs by means of a deep linguistic reuse, exploiting a host language feature to
-implement the corresponding DSL feature without (much) additional work. An interesting insight is that CPS
-transforming a program generator allows it to generate code that is in CPS.
+implement the corresponding DSL feature without (much) additional work. An interesting insight is that our method of
+CPS transforming a direct-style program generator allows it to generate code that is in CPS.
 
@@
 
-As an example, we'll consider a \code{sleep} method implemented in non-blocking, asynchronous style.
+As an example, we consider a \code{sleep} method implemented in non-blocking, asynchronous style.
 This is useful, for example, when using \code{ThreadPool}s as no thread is being blocked during the sleep period.
@@
 
-It's important to note that reification preserves function composition (roughly speaking it's homomorphism between DSL code and generated JavaScript code).
+It is important to note that reification preserves function composition (roughly speaking it is a homomorphism between DSL code and generated JavaScript code).
 Specifically, let \code{f: Rep[A] => Rep[B]} and \code{g: Rep[B] => Rep[C]} then {\tt\small fun(g compose f) == (fun(g) compose fun(f))}
@@
 
-It's interesting to note that, since the request already has the right structure for the \code{jQuery.ajax} function, we can simply pass it to the framework-provided \code{quote} method, which knows how to generate JavaScript representations of any \code{JSLiteral}.
+It is interesting to note that, since the request already has the right structure for the \code{jQuery.ajax} function, we can simply pass it to the framework-provided \code{quote} method, which knows how to generate JavaScript representations of any \code{JSLiteral}.
 
@@
 
-In fact, if we change the code to use \code{suspendable} instead of \code{parSuspendable} and run the generated JavaScript program, we'll get following output printed to the JavaScript console:
+In fact, if we change the code to use \code{suspendable} instead of \code{parSuspendable} and run the generated JavaScript program, we will get following output printed to the JavaScript console:
 \begin{lstlisting}
@@
 
-It's worth noting that \code{RichCellOps} is not reified so it will be dropped at staging time and its method will get inlined whenever used.
+It is worth noting that \code{RichCellOps} is not reified so it will be dropped at staging time and its method will get inlined whenever used.
 Also, it contains CPS-specific code that allows us to capture the continuation.
@@
 The rest of the code stays the same.
-It's easy to switch between both versions, and users are free to make their choice according to their needs and performance requirements.
+It is easy to switch between both versions, and users are free to make their choice according to their needs and performance requirements.
 
 %Parallelism doesn't come for free as it involves more CPS rewritings and involves indirections introduced by our abstractions.
-However, since it's easy to switch between both versions, users are free to make those choices according to their needs and performance requirements.
+However, since it is easy to switch between both versions, users are free to make those choices according to their needs and performance requirements.
 
@@
 
-We've implemented our DSL in Scala.\footnote{The code can be found at
+We have implemented our DSL in Scala.\footnote{The code can be found at
   \url{http://github.com/js-scala}} We used our DSL to
@@
 
-Also, we integrated our DSL with an existing web framework, Play 2.0.\footnote{\url{http://www.playframework.org}}
+Moreover, we integrated our DSL with an existing web framework, Play 2.0.\footnote{\url{http://www.playframework.org}}
 Our DSL can be used to define Play form validators that are executed on both the client and server sides.
-We managed to achieve that without deep changes to how the framework works internally proving that our
+We managed to achieve that without deep changes to the framework, proving that our
 DSL can be used as a library.
@@
 server sends the complete drawing history to the new client, and the
-client reconstructs the image by playing the commands. A very
+client reconstructs the image by playing back the commands. A very
 simple improvement is to make the server execute the drawing
@@
 section~\ref{sec:trivial-embedding}. New clients then just obtain the
-bitmap instead of replaying the history, which is potentially large.
+bitmap instead of replaying the history, which can be large and grow
+unboundedly.
 
@@
 \section{Acknowledgments}
-We thank Adriaan Moors for maintaining the Scala-Virtualized compiler, adding our feature requests, fixing our reported bugs, and for insightful discussions.
-
+We thank Adriaan Moors for maintaining the Scala-Virtualized compiler, adding our feature requests, 
+fixing our reported bugs, and for insightful discussions.
+
+{
+%\clearpage
+%\linespread{0.99}
 \bibliographystyle{abbrv}
 \bibliography{refs,ppl}
+}
 

### 2012-04-30 papers/ecoop2012/fetchTweets.js
Removed changemargin trick.
We introduced changemargin trick to fit
two-column listings by making them wider
than regular text. This turned out to be
a bad idea for publishing.

Removed changemargin and reformatted
listing so they fit into more tight
columns.

@@ -1,7 +1,7 @@
 function fetchTweets(username, callback) {
   jQuery.ajax({
-    url: 
-"http://api.twitter.com/1/statuses/user_timeline.json/",
+    url: "http://api.twitter.com/1/
+      statuses/user_timeline.json/",
     type: "GET",
     dataType: "jsonp",
     data: {

### 2012-04-30 papers/ecoop2012/fetchTweets.scala

@@ -1,16 +1,18 @@
-def fetchTweets(username: Rep[String]) = (ajax.get {
-  new JSLiteral {
-    val url = 
-"http://api.twitter.com/1/statuses/user_timeline.json"
-    val `type` = "GET"
-    val dataType = "jsonp"
-    val data = new JSLiteral {
-      val screen_name = username
-      val include_rts = true
-      val count = 5
-      val include_entities = true
+def fetchTweets(username: Rep[String]) =
+  (ajax.get {
+    new JSLiteral {
+      val url = "http://api.twitter.com/1/
+        statuses/user_timeline.json"
+      val `type` = "GET"
+      val dataType = "jsonp"
+      val data = new JSLiteral {
+        val screen_name = username
+        val include_rts = true
+        val count = 5
+        val include_entities = true
+      }
     }
-  }
-}).as[TwitterResponse]
+  }).as[TwitterResponse]
 
-type TwitterResponse = Array[JSLiteral {val text: String}]
+type TwitterResponse =
+  Array[JSLiteral {val text: String}]

### 2012-04-30 papers/ecoop2012/twitter.js

@@ -1,5 +1,6 @@
 var processed = 0
-var users = ["gkossakowski", "odersky", "adriaanm"]
+var users = ["gkossakowski", "odersky",
+  "adriaanm"]
 users.forEach(function (user) {
   console.log("fetching " + user)
   fetchTweets(user, function(data) {

### 2012-04-30 papers/ecoop2012/twitter.scala

@@ -1,4 +1,5 @@
-val users = array("gkossakowski", "odersky", "adriaanm")
+val users = array("gkossakowski", "odersky",
+  "adriaanm")
 for (user <- users.parSuspendable) {
   console.log("fetching " + user)
   val tweets = fetchTweets(user)

### 2012-07-12 src/main/scala/scala/js/JSEqual.scala
Updated to delite-develop.

@@ -7,7 +7,7 @@ trait JSGenEqual extends JSGenBase {
   val IR: EqualExp
   import IR._
   
-  override def emitNode(sym: Sym[Any], rhs: Def[Any])(implicit stream: PrintWriter) = rhs match {
+  override def emitNode(sym: Sym[Any], rhs: Def[Any]) = rhs match {
     case Equal(a,b) =>  emitValDef(sym, "" + quote(a) + "==" + quote(b))
     case NotEqual(a,b) =>  emitValDef(sym, "" + quote(a) + "!=" + quote(b))
     case _ => super.emitNode(sym, rhs)

### 2012-07-12 src/main/scala/scala/js/JSProxy.scala

@@ -82,7 +82,7 @@ trait JSGenProxy extends JSGenBase with JSGenEffect {
   val IR: JSProxyExp
   import IR._
 
-  override def emitNode(sym: Sym[Any], rhs: Def[Any])(implicit stream: PrintWriter) = rhs match {
+  override def emitNode(sym: Sym[Any], rhs: Def[Any]) = rhs match {
     case MethodCall(receiver, method, args) =>  emitValDef(sym,
       quote(receiver) + "." + method + args.map(quote).mkString("(", ",", ")"))
     case SuperMethodCall(receiver, parentConstructor, method, args) =>  emitValDef(sym,

### 2012-07-12 src/main/scala/scala/js/JSRegExps.scala

@@ -30,7 +30,7 @@ trait JSGenRegExps extends JSGenBase {
   val IR: JSRegExpsExp
   import IR._
 
-  override def emitNode(sym: Sym[Any], rhs: Def[Any])(implicit stream: PrintWriter) = rhs match {
+  override def emitNode(sym: Sym[Any], rhs: Def[Any]) = rhs match {
     case StringRegExp(r) => emitValDef(sym, "new RegExp(" + quote(r) + ")")
     case RegExpTest(re, str) => emitValDef(sym, quote(re) + ".test(" + quote(str) + ")")
     case StringSearch(str, re) => emitValDef(sym, quote(str) + ".search(" + quote(re) + ")")

### 2012-07-12 src/main/scala/scala/js/JSTraits.scala

@@ -14,7 +14,7 @@ trait JSTraits extends JSProxyBase {
 trait JSTraitsExp extends JSTraits with JSProxyExp {
   trait Constructor[+T]
 
-  case class MethodTemplate(name: String, params: List[Sym[Any]], body: Exp[Any])
+  case class MethodTemplate(name: String, params: List[Sym[Any]], body: Block[Any])
   case class ParentTemplate(constructor: Exp[Any], instance: Exp[Any])
 
   case class This[+T:Manifest]() extends Exp[T]
@@ -85,7 +85,7 @@ trait JSGenTraits extends JSGenBase with JSGenProxy {
   val IR: JSTraitsExp
   import IR._
 
-  override def emitNode(sym: Sym[Any], rhs: Def[Any])(implicit stream: PrintWriter) = rhs match {
+  override def emitNode(sym: Sym[Any], rhs: Def[Any]) = rhs match {
     case ClassTemplate(parentTemplate, methodTemplates) =>
       stream.println("var " + quote(sym) + " = function() {")
       stream.println("this.$init$()")

### 2012-07-12 src/main/scala/scala/js/Koch.scala

@@ -57,7 +57,7 @@ object Koch {
 
   def codegen(pw: PrintWriter) {
     new KochProg with JSExp with DomsExp { self =>
-      val codegen = new JSGenOpt with GenDoms { val IR: self.type = self }
+      val codegen = new JSGen with GenDoms { val IR: self.type = self }
       codegen.emitSource0(draw _, "draw", pw)
     }
   }

### 2012-07-12 src/test/scala/scala/js/TestTuple.scala

@@ -39,7 +39,7 @@ class TestTuple extends FileDiffSuite {
   def testTwoArgs = {
     withOutFile(prefix+"tuple") {
       new TwoArgsProg with TupleOpsExp with NumericOpsExpOpt { self =>
-        val codegen = new JSTupledCodegen with JSGenNumericOps { val IR: self.type = self }
+        val codegen = new JSNestedCodegen with JSTupledCodegen with JSGenNumericOps { val IR: self.type = self }
         codegen.emitSource(test _, "main", new PrintWriter(System.out))
       }
     }

### 2012-07-13 src/main/java/scala/js/actor/OnTextMessage.java
cherry-pick from drawing branch

@@ -1,7 +1,0 @@
-package scala.js.actor;
-
-/** Dummy Java interface that serves as a work-around for illegal cyclic
-    dependency problem in Scala compiler. */
-interface OnTextMessage extends org.eclipse.jetty.websocket.WebSocket.OnTextMessage {
-
-}

### 2012-07-13 src/main/scala/scala/js/Main.scala

@@ -71,5 +71,4 @@ object Main extends App {
     Koch.run()
   Birds.writeJs("examples/birds/Bird_.js")
   Twitter.writeJs("examples/ajax/twitter_.js")
-  Mice.writeJs("src/main/webapp/mice_.js")
 }

### 2012-07-13 src/main/scala/scala/js/Mice.scala

@@ -1,154 +1,0 @@
-package scala.js
-
-import scala.virtualization.lms.common._
-
-import java.io.PrintWriter
-
-trait MiceApi extends JSProxyBase with JSLiteral {
-  val window: Rep[Any]
-
-  type DataLiteral = JSLiteral {val data : String}
-  def initWebSocket(): Rep[WebSocket]
-  trait WebSocket {
-    var onmessage: Rep[DataLiteral => Any]
-    def send(data: Rep[String]): Rep[Unit]
-  }
-  implicit def repToSocket(x: Rep[WebSocket]): WebSocket = repProxy[WebSocket](x)
-
-  type ActionLiteral = JSLiteral {val action: String; val id: String}
-  type MoveLiteral = JSLiteral {val id: String; val cx: Int; val cy: Int; val w: Int; val h: Int; val color: String}
-
-  def currentTime(): Rep[Int]
-
-  val json: Rep[JSON]
-  trait JSON {
-    def stringify(literal: Rep[JSLiteral]): Rep[String]
-    def parse[T <: JSLiteral](data: Rep[String]): Rep[T]
-  }
-  implicit def repToJSON(x: Rep[JSON]): JSON = repProxy[JSON](x)
-
-  def jQuery(x: Rep[Any]): Rep[JQuery]
-  trait JQuery {
-    var length: Rep[Int]
-
-    def width(): Rep[Int]
-    def height(): Rep[Int]
-
-    def append(s: Rep[String]): Rep[JQuery]
-    def css(o: Rep[JSLiteral]): Rep[JQuery]
-    def click(fn: Rep[JQueryEvent => Any]): Rep[JQuery]
-    def mousemove(fn: Rep[JQueryEvent => Any]): Rep[JQuery]
-    def remove(): Rep[JQuery]
-  }
-  implicit def repToJQuery(x: Rep[JQuery]): JQuery = repProxy[JQuery](x)
-  type JQueryEvent = JSLiteral {val pageX: Int; val pageY: Int}
-}
-
-trait MiceApiExp extends MiceApi with JSProxyExp with JSLiteralExp with DomsExp {
-
-  case object WindowVar extends Exp[Any]
-  val window = WindowVar
-
-  case object InitWebSocket extends Def[WebSocket]
-  def initWebSocket() = reflectEffect(InitWebSocket)
-
-  case object CurrentTime extends Def[Int]
-  def currentTime() = reflectEffect(CurrentTime)
-
[…]

### 2012-07-13 src/main/scala/scala/js/actor/JettyChat.scala

@@ -1,58 +1,0 @@
-package scala.js.actor
-
-import java.util.concurrent.CopyOnWriteArraySet
-
-import javax.servlet.http.HttpServletRequest
-import javax.servlet.http.HttpServletResponse
-
-import org.eclipse.jetty.websocket.WebSocket
-import org.eclipse.jetty.websocket.WebSocket.Connection
-import org.eclipse.jetty.websocket.WebSocketServlet
-
-import scala.collection.JavaConversions._
-
-class WebSocketChatServlet extends WebSocketServlet {
-  private val colors = Array("red", "blue", "yellow", "green", "purple", "orange")
-  private var _count = 0;
-  private val _members = new CopyOnWriteArraySet[ChatWebSocket]();
-
-  override protected def doGet(request: HttpServletRequest, response: HttpServletResponse) {
-    getServletContext().getNamedDispatcher("default").forward(request, response)
-  }
-
-  override def doWebSocketConnect(request: HttpServletRequest, protocol: String) : WebSocket = new ChatWebSocket()
-
-  class ChatWebSocket extends OnTextMessage {
-    private var _connection : Connection = null;
-    private var _id: Int = 0;
-
-    def onOpen(connection: Connection) {
-      // Log.info(this+" onConnect");
-      _connection=connection;
-      _count += 1;
-      _id = _count;
-      _members.add(this);
-    }
-
-    def onMessage(frame: Byte, data: Array[Byte], offset: Int, length: Int) {
-      // Log.info(this+" onMessage: "+TypeUtil.toHexString(data,offset,length));
-    }
-
-    private def broadcast(data: String) {
-      val xdata = data.replace("{", "{\"id\":" + _id + "," + "\"color\":\"" + colors(_id % colors.length) + "\",")
-      for (member <- _members) {
-        member._connection.sendMessage(xdata);
-      }
-    }
-
-    def onMessage(data: String) {
-      broadcast(data)
-    }
-
-    def onClose(code: Int, message: String) {
-      // Log.info(this+" onDisconnect");
-      _members.remove(this);
-      broadcast("{\"action\":\"close\"}")
-    }
-  }
-}

### 2012-07-13 src/main/webapp/mice.js

@@ -1,43 +1,0 @@
-var socket = new WebSocket(document.location.toString().replace('http://','ws://').replace('https://','wss://'));
-socket.onmessage=function(m){
-    data = JSON.parse(m.data);
-    if(data['action'] == 'close'){
-	$('#mouse_'+data['id']).remove();
-    } else if(data['action'] == 'move'){
-	move(data);
-    };
-}
-
-function ratelimit(fn, ms) {
-    var last = (new Date()).getTime();
-    return (function() {
-	var now = (new Date()).getTime();
-	if (now - last > ms) {
-	    last = now;
-	    fn.apply(null, arguments);
-	}
-    });
-}
-
-function move(mouse){
-    if($('#mouse_'+mouse['id']).length == 0) {
-	$('body').append('<span class="mouse" id="mouse_'+mouse['id']+'"><span style="display:none;" class="chat"/></span>');
-    }
-    $('#mouse_'+mouse['id']).css({
-	'left' : (($(window).width() - mouse['w']) / 2 + mouse['cx']) + 'px',
-	'top' : mouse['cy'] + 'px',
-        'background-color' : mouse['color']
-    })
-};
-
-$(document).mousemove(
-    ratelimit(function(e){
-	socket.send(JSON.stringify({
-	    action: 'move',
-	    cx: e.pageX,
-	    cy: e.pageY,
-	    w: $(window).width(),
-	    h: $(window).height()
-	}))
-    }, 40)
-);

### 2012-07-13 examples/actors/jsworker.js
cleanup main

@@ -1,9 +1,0 @@
-var connections = 0; // count active connections
-self.addEventListener("connect", function (e) {
-    var port = e.ports[0];
-    connections++;
-    port.addEventListener("message", function (e) {
-	port.postMessage("Hello " + e.data + " (port #" + connections + ")");
-    }, false);
-    port.start();
-}, false);

### 2012-07-13 examples/koch.js

@@ -1,39 +1,0 @@
-var deg = Math.PI/180;  // For converting degrees to radians
-
-// Draw a level-n Koch Snowflake fractal on the canvas context c,
-// with lower-left corner at (x,y) and side length len.
-function snowflake(c, n, x, y, len) {
-    c.save();           // Save current transformation
-    c.translate(x,y);   // Translate origin to starting point
-    c.moveTo(0,0);      // Begin a new subpath at the new origin
-    leg(n);             // Draw the first leg of the snowflake
-    c.rotate(-120*deg); // Now rotate 120 degrees counterclockwise
-    leg(n);             // Draw the second leg
-    c.rotate(-120*deg); // Rotate again
-    leg(n);             // Draw the final leg
-    c.closePath();      // Close the subpath
-    c.restore();        // And restore original transformation
-
-    // Draw a single leg of a level-n Koch snowflake.
-    // This function leaves the current point at the end of the leg it has
-    // drawn and translates the coordinate system so the current point is (0,0).
-    // This means you can easily call rotate() after drawing a leg.
-    function leg(n) {
-        c.save();               // Save the current transformation
-        if (n == 0) {           // Nonrecursive case:
-            c.lineTo(len, 0);   //   Just draw a horizontal line
-        }                       //                                       _  _
-        else {                  // Recursive case: draw 4 sub-legs like:  \/
-            c.scale(1/3,1/3);   // Sub-legs are 1/3rd the size of this leg
-            leg(n-1);           // Recurse for the first sub-leg
-            c.rotate(60*deg);   // Turn 60 degrees clockwise
-            leg(n-1);           // Second sub-leg
-            c.rotate(-120*deg); // Rotate 120 degrees back
-            leg(n-1);           // Third sub-leg
-            c.rotate(60*deg);   // Rotate back to our original heading
-            leg(n-1);           // Final sub-leg
-        }
-        c.restore();            // Restore the transformation
-        c.translate(len, 0);    // But translate to make end of leg (0,0)
-    }
-}

### 2012-07-13 src/main/scala/scala/js/Main.scala

@@ -2,75 +2,12 @@ package scala.js
 
 import scala.virtualization.lms.common._
 
-import java.io.PrintWriter
-import java.io.FileOutputStream
-
-trait Print extends Base {
-  implicit def unit(s: String): Rep[String]
-  def print(s: Rep[Any]): Rep[Unit]
-}
-
-trait PrintExp extends Print with EffectExp {
-  implicit def unit(s: String): Rep[String] = Const(s)
-  case class Print(s: Rep[Any]) extends Def[Unit]
-  def print(s: Rep[Any]) = reflectEffect(Print(s))
-}
-
-trait ScalaGenPrint extends ScalaGenEffect {
-  val IR: PrintExp
-  import IR._
-  
-  override def emitNode(sym: Sym[Any], rhs: Def[Any]) = rhs match {
-    case Print(s) =>  emitValDef(sym, "println(" + quote(s) + ")")
-    case _ => super.emitNode(sym, rhs)
-  }
-}
-
-trait JSGenPrint extends JSGenEffect {
-  val IR: PrintExp
-  import IR._
-  
-  // TODO: should have a function for this
-  override def emitNode(sym: Sym[Any], rhs: Def[Any]) = rhs match {
-    case Print(s) =>  emitValDef(sym, "document.body.appendChild(document.createElement(\"div\"))"+
-        ".appendChild(document.createTextNode("+quote(s)+"))")
-    case _ => super.emitNode(sym, rhs)
-  }
-}
-
-trait FunProg { this: JSFunctions =>
-  def test(x: Rep[Any]): Rep[Any] = {
-    val id = fun { x : Rep[Any] => x }
-    id.apply(x)
-  }
-}
-
-trait SomeProg { this: JS =>
-  def test(x: Rep[Any]): Rep[Any] = {
-    lazy val id: Rep[Any => Any] = fun { x : Rep[Any] =>
-      dynamic(x).foo.bar()
-      id.apply(x)
-    }
-    id.apply(x)
-  }
-}
-
 object Main extends App {
-  new FunProg with JSFunctionsExp { self =>
-    val codegen = new JSGenFunctions { val IR: self.type = self }
-    codegen.emitSource(test _, "main", new PrintWriter(System.out))
-  }
-
-  new SomeProg with JSExp { self =>
-    val codegen = new JSGen { val IR: self.type = self }
[…]

### 2012-07-13 src/main/scala/scala/js/examples/Birds.scala
moving examples to own package

@@ -1,5 +1,6 @@
-package scala.js
+package scala.js.examples
 
+import scala.js._
 import java.io.PrintWriter
 
 object Birds {

### 2012-07-13 src/main/scala/scala/js/examples/D3.scala

@@ -1,5 +1,6 @@
-package scala.js
+package scala.js.examples
 
+import scala.js._
 import java.io.PrintWriter
 
 object D3 {

### 2012-07-13 src/main/scala/scala/js/examples/Koch.scala

@@ -1,5 +1,6 @@
-package scala.js
+package scala.js.examples
 
+import scala.js._
 import java.io.PrintWriter
 
 object Koch {

### 2012-07-13 src/main/scala/scala/js/examples/Main.scala

@@ -1,6 +1,4 @@
-package scala.js
-
-import scala.virtualization.lms.common._
+package scala.js.examples
 
 object Main extends App {
   Koch.writeHtml("examples/canvas/koch_.html")

### 2012-07-13 src/main/scala/scala/js/examples/Twitter.scala

@@ -1,5 +1,6 @@
-package scala.js
+package scala.js.examples
 
+import scala.js._
 import scala.virtualization.lms.common._
 import scala.util.continuations._
 

### 2012-07-13 papers/ecoop2012/template.tex

%%%%%%%%%%%%%%%%%%%%%%% file typeinst.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is the LaTeX source for the instructions to authors using
% the LaTeX document class 'llncs.cls' for contributions to
% the Lecture Notes in Computer Sciences series.
% http://www.springer.com/lncs       Springer Heidelberg 2006/05/04
%
% It may be used as a template for your own input - copy it
% to a new file with a new name and use it as the basis
% for your article.
%
% NB: the document class 'llncs' has its own and detailed documentation, see
% ftp://ftp.springer.de/data/pubftp/pub/tex/latex/llncs/latex2e/llncsdoc.pdf
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[runningheads,a4paper]{llncs}

\usepackage{amssymb}
\setcounter{tocdepth}{3}
\usepackage{graphicx}

\usepackage{url}
\newcommand{\keywords}[1]{\par\addvspace\baselineskip
\noindent\keywordname\enspace\ignorespaces#1}

\input{macros}

\begin{document}

\mainmatter  % start of an individual contribution

\title{JavaScript as an embedded DSL in Scala}
% a short form should be given in case it is too long for the running head
%\titlerunning{}

% the name(s) of the author(s) follow(s) next
%
% NB: Chinese authors should write their first names(s) in front of
% their surnames. This ensures that the names appear correctly in
% the running heads and the author index.
%
\author{Grzegorz Kossakowski \and Nada Amin \and Tiark Rompf \and Martin Odersky}
%
\authorrunning{JavaScript as an embedded DSL in Scala}
% (feature abused for this document to repeat the title also on left hand pages)

% the affiliations are given next; don't give your e-mail address
% unless you accept that it will be published
\institute{Ecole Polytechnique F\'ed\'erale de Lausanne,\\
\email{\{first.last\}@epfl.ch}}
%\mailsa\\
%\mailsb\\
%\url{http://lamp.epfl.ch}}

%
% NB: a more complex sample for affiliations and the mapping to the
% corresponding authors can be found in the file "llncs.dem"
% (search for the string "\mainmatter" where a contribution starts).
% "llncs.dem" accompanies the document class "llncs.cls".
%

%\toctitle{}
%\tocauthor{}
\maketitle

\begin{abstract}
% =>70 & <=150 words
\note{TODO}

\keywords{JavaScript, Scala, DSL, programming languages}
\end{abstract}

We use Lightweight Modular Staging (LMS) to create a DSL in Scala which generates JavaScript code.

Our approach addresses many problems encountered in web development in an elegant, coherent and lightweight way. Programming directly in JavaScript include the following problems:
\begin{itemize}
\item lack of static typing, which makes it difficult to maintain, refactor and develop large web applications,
\item impedance mismatch between client and server code,
\item lack of modularity and abstractions,
\item difficult to reason about control structures that arise from high-performance asynchronous callback-driven programming style.
\end{itemize}
Frameworks such as GWT and Links which generates JavaScript solve some of the issues, but introduce new problems, including:
\begin{itemize}
\item lack of integration with existing JavaScript libraries,
\item difficult to extend with new APIs, let alone new language primitives,
\item little control over generated code which makes performance-tuning very difficult,
\item no first-class status for generated code, which makes it impossible to compose and abstract over snippets.
\end{itemize}

Our approach is lightweight: By embedding our DSL as a library in Scala, we exploit the features of the host language such as expressive static typing, easy creation of new control structures, and even partial CPS transformation.

\section{Introduction to LMS}
- virtualization
- rep types
- IR construction
- code generation
advantage: can have abstractions that only exist at staging-time

\section{Gradual Typing}
- Scala Dynamic
- use of external APIs
- snippets with a simple example showing evolution from dynamic to typed (Canvas example which will become Koch example in the next section)

\section{Trivial Embedding}
- DOM Canvas Koch example: generate both client-side code or desktop widget

\section{Modularity Interlude}
- modularity at many levels
   e.g. dynamic feature is just a trait
   users specify external APIs with traits
   interpretation of rep also done via traits (generating IR nodes vs trivial embedding)
- example of user adding a language primitive: e.g. for-comprehension
- mix and match features by composing traits (e.g. everything available on main thread 'JS trait', vs no-DOM manipualtion in worker threads 'WorkerJS trait)

\section{Reification of Objects}
- JS Literals: intuitive translation to JS as useful JSON-like objects, static typing
- mice example: shows use of JSLiterals + API gluing (create abstractions that you need, not necessarily exactly matching the underlying JS API)
- reification of traits: mutability, inheritance

\section{Case Study: CPS transformation to simply asynchronous web development}

\clearpage
\section{Introduction}

You are strongly encouraged to use \LaTeXe{} for the
preparation of your camera-ready manuscript together with the
corresponding Springer class file \verb+llncs.cls+. Only if you use
\LaTeXe{} can hyperlinks be generated in the online version
of your manuscript.

The \LaTeX{} source of this instruction file for \LaTeX{} users may be
used as a template. This is
located in the ``authors'' subdirectory in
\url{ftp://ftp.springer.de/pub/tex/latex/llncs/latex2e/instruct/} and
entitled \texttt{typeinst.tex}. There is a separate package for Word 
users. Kindly send the final and checked source
and PDF files of your paper to the Contact Volume Editor. This is
usually one of the organizers of the conference. You should make sure
that the \LaTeX{} and the PDF files are identical and correct and that
only one version of your paper is sent. It is not possible to update
files at a later stage. Please note that we do not need the printed
paper.

We would like to draw your attention to the fact that it is not possible
to modify a paper in any way, once it has been published. This applies
to both the printed book and the online version of the publication.
Every detail, including the order of the names of the authors, should
be checked before the paper is sent to the Volume Editors.

\subsection{Checking the PDF File}

Kindly assure that the Contact Volume Editor is given the name and email
address of the contact author for your paper. The Contact Volume Editor
uses these details to compile a list for our production department at
SPS in India. Once the files have been worked upon, SPS sends a copy of
the final pdf of each paper to its contact author. The contact author is
asked to check through the final pdf to make sure that no errors have
crept in during the transfer or preparation of the files. This should
not be seen as an opportunity to update or copyedit the papers, which is
not possible due to time constraints. Only errors introduced during the
preparation of the files will be corrected.

This round of checking takes place about two weeks after the files have
been sent to the Editorial by the Contact Volume Editor, i.e., roughly
seven weeks before the start of the conference for conference
proceedings, or seven weeks before the volume leaves the printer's, for
post-proceedings. If SPS does not receive a reply from a particular
contact author, within the timeframe given, then it is presumed that the
author has found no errors in the paper. The tight publication schedule
of LNCS does not allow SPS to send reminders or search for alternative
email addresses on the Internet.

In some cases, it is the Contact Volume Editor that checks all the final
pdfs. In such cases, the authors are not involved in the checking phase.

\subsection{Additional Information Required by the Volume Editor}

If you have more than one surname, please make sure that the Volume Editor
knows how you are to be listed in the author index.

\subsection{Copyright Forms}

The copyright form may be downloaded from the ``For Authors"
(Information for LNCS Authors) section of the LNCS Website:
\texttt{www.springer.com/lncs}. Please send your signed copyright form
to the Contact Volume Editor, either as a scanned pdf or by fax or by
courier. One author may sign on behalf of all of the other authors of a
particular paper. Digital signatures are acceptable.

\section{Paper Preparation}

Springer provides you with a complete integrated \LaTeX{} document class
(\texttt{llncs.cls}) for multi-author books such as those in the LNCS
series. Papers not complying with the LNCS style will be reformatted.
This can lead to an increase in the overall number of pages. We would
therefore urge you not to squash your paper.

Please always cancel any superfluous definitions that are
not actually used in your text. If you do not, these may conflict with
the definitions of the macro package, causing changes in the structure
of the text and leading to numerous mistakes in the proofs.

If you wonder what \LaTeX{} is and where it can be obtained, see the
``\textit{LaTeX project site}'' (\url{http://www.latex-project.org})
and especially the webpage ``\textit{How to get it}''
(\url{http://www.latex-project.org/ftp.html}) respectively.

When you use \LaTeX\ together with our document class file,
\texttt{llncs.cls},
your text is typeset automatically in Computer Modern Roman (CM) fonts.
Please do
\emph{not} change the preset fonts. If you have to use fonts other
than the preset fonts, kindly submit these with your files.

Please use the commands \verb+\label+ and \verb+\ref+ for
cross-references and the commands \verb+\bibitem+ and \verb+\cite+ for
references to the bibliography, to enable us to create hyperlinks at
these places.

For preparing your figures electronically and integrating them into
your source file we recommend using the standard \LaTeX{} \verb+graphics+ or
\verb+graphicx+ package. These provide the \verb+\includegraphics+ command.
In general, please refrain from using the \verb+\special+ command.

Remember to submit any further style files and
fonts you have used together with your source files.

\subsubsection{Headings.}

Headings should be capitalized
(i.e., nouns, verbs, and all other words
except articles, prepositions, and conjunctions should be set with an
initial capital) and should,
with the exception of the title, be aligned to the left.
Words joined by a hyphen are subject to a special rule. If the first
word can stand alone, the second word should be capitalized.

Here are some examples of headings: ``Criteria to Disprove
Context-Freeness of Collage Language", ``On Correcting the Intrusion of
Tracing Non-deterministic Programs by Software", ``A User-Friendly and
Extendable Data Distribution System", ``Multi-flip Networks:
Parallelizing GenSAT", ``Self-determinations of Man".

\subsubsection{Lemmas, Propositions, and Theorems.}

The numbers accorded to lemmas, propositions, and theorems, etc. should
appear in consecutive order, starting with Lemma 1, and not, for
example, with Lemma 11.

\subsection{Figures}

For \LaTeX\ users, we recommend using the \emph{graphics} or \emph{graphicx}
package and the \verb+\includegraphics+ command.

Please check that the lines in line drawings are not
interrupted and are of a constant width. Grids and details within the
figures must be clearly legible and may not be written one on top of
the other. Line drawings should have a resolution of at least 800 dpi
(preferably 1200 dpi). The lettering in figures should have a height of
2~mm (10-point type). Figures should be numbered and should have a
caption which should always be positioned \emph{under} the figures, in
contrast to the caption belonging to a table, which should always appear
\emph{above} the table; this is simply achieved as matter of sequence in
your source.

Please center the figures or your tabular material by using the \verb+\centering+
declaration. Short captions are centered by default between the margins
and typeset in 9-point type (Fig.~\ref{fig:example} shows an example).
The distance between text and figure is preset to be about 8~mm, the
distance between figure and caption about 6~mm.

To ensure that the reproduction of your illustrations is of a reasonable
quality, we advise against the use of shading. The contrast should be as
pronounced as possible.

If screenshots are necessary, please make sure that you are happy with
the print quality before you send the files.
\begin{figure}
\centering
%\includegraphics[height=6.2cm]{eijkel2}
\caption{One kernel at $x_s$ (\emph{dotted kernel}) or two kernels at
$x_i$ and $x_j$ (\textit{left and right}) lead to the same summed estimate
at $x_s$. This shows a figure consisting of different types of
lines. Elements of the figure described in the caption should be set in
italics, in parentheses, as shown in this sample caption.}
\label{fig:example}
\end{figure}

Please define figures (and tables) as floating objects. Please avoid
using optional location parameters like ``\verb+[h]+" for ``here".

\paragraph{Remark 1.}

In the printed volumes, illustrations are generally black and white
(halftones), and only in exceptional cases, and if the author is
prepared to cover the extra cost for color reproduction, are colored
pictures accepted. Colored pictures are welcome in the electronic
version free of charge. If you send colored figures that are to be
printed in black and white, please make sure that they really are
legible in black and white. Some colors as well as the contrast of
converted colors show up very poorly when printed in black and white.

\subsection{Formulas}

Displayed equations or formulas are centered and set on a separate
line (with an extra line or halfline space above and below). Displayed
expressions should be numbered for reference. The numbers should be
consecutive within each section or within the contribution,
with numbers enclosed in parentheses and set on the right margin --
which is the default if you use the \emph{equation} environment, e.g.,
\begin{equation}
  \psi (u) = \int_{o}^{T} \left[\frac{1}{2}
  \left(\Lambda_{o}^{-1} u,u\right) + N^{\ast} (-u)\right] dt \;  .
\end{equation}

Equations should be punctuated in the same way as ordinary
text but with a small space before the end punctuation mark.

\subsection{Footnotes}

The superscript numeral used to refer to a footnote appears in the text
either directly after the word to be discussed or -- in relation to a
phrase or a sentence -- following the punctuation sign (comma,
semicolon, or period). Footnotes should appear at the bottom of
the
normal text area, with a line of about 2~cm set
immediately above them.\footnote{The footnote numeral is set flush left
and the text follows with the usual word spacing.}

\subsection{Program Code}

Program listings or program commands in the text are normally set in
typewriter font, e.g., CMTT10 or Courier.

\medskip

\noindent
{\it Example of a Computer Program}
\begin{verbatim}
program Inflation (Output)
  {Assuming annual inflation rates of 7%, 8%, and 10%,...
   years};
   const
     MaxYears = 10;
   var
     Year: 0..MaxYears;
     Factor1, Factor2, Factor3: Real;
   begin
     Year := 0;
     Factor1 := 1.0; Factor2 := 1.0; Factor3 := 1.0;
     WriteLn('Year  7% 8% 10%'); WriteLn;
     repeat
       Year := Year + 1;
       Factor1 := Factor1 * 1.07;
       Factor2 := Factor2 * 1.08;
       Factor3 := Factor3 * 1.10;
       WriteLn(Year:5,Factor1:7:3,Factor2:7:3,Factor3:7:3)
     until Year = MaxYears
end.
\end{verbatim}
%
\noindent
{\small (Example from Jensen K., Wirth N. (1991) Pascal user manual and
report. Springer, New York)}

\subsection{Citations}

For citations in the text please use
square brackets and consecutive numbers: \cite{jour}, \cite{lncschap},
\cite{proceeding1} -- provided automatically
by \LaTeX 's \verb|\cite| \dots\verb|\bibitem| mechanism.

\subsection{Page Numbering and Running Heads}

There is no need to include page numbers. If your paper title is too
long to serve as a running head, it will be shortened. Your suggestion
as to how to shorten it would be most welcome.

\section{LNCS Online}

The online version of the volume will be available in LNCS Online.
Members of institutes subscribing to the Lecture Notes in Computer
Science series have access to all the pdfs of all the online
publications. Non-subscribers can only read as far as the abstracts. If
they try to go beyond this point, they are automatically asked, whether
they would like to order the pdf, and are given instructions as to how
to do so.

Please note that, if your email address is given in your paper,
it will also be included in the meta data of the online version.

\section{BibTeX Entries}

The correct BibTeX entries for the Lecture Notes in Computer Science
volumes can be found at the following Website shortly after the
publication of the book:
\url{http://www.informatik.uni-trier.de/~ley/db/journals/lncs.html}

\subsubsection*{Acknowledgments.} The heading should be treated as a
subsubsection heading and should not be assigned a number.

\section{The References Section}\label{references}

In order to permit cross referencing within LNCS-Online, and eventually
between different publishers and their online databases, LNCS will,
from now on, be standardizing the format of the references. This new
feature will increase the visibility of publications and facilitate
academic research considerably. Please base your references on the
examples below. References that don't adhere to this style will be
reformatted by Springer. You should therefore check your references
thoroughly when you receive the final pdf of your paper.
The reference section must be complete. You may not omit references.
Instructions as to where to find a fuller version of the references are
not permissible.

We only accept references written using the latin alphabet. If the title
of the book you are referring to is in Russian or Chinese, then please write
(in Russian) or (in Chinese) at the end of the transcript or translation
of the title.

The following section shows a sample reference list with entries for
journal articles \cite{jour}, an LNCS chapter \cite{lncschap}, a book
\cite{book}, proceedings without editors \cite{proceeding1} and
\cite{proceeding2}, as well as a URL \cite{url}.
Please note that proceedings published in LNCS are not cited with their
full titles, but with their acronyms!

\begin{thebibliography}{4}

\bibitem{jour} Smith, T.F., Waterman, M.S.: Identification of Common Molecular
Subsequences. J. Mol. Biol. 147, 195--197 (1981)

\bibitem{lncschap} May, P., Ehrlich, H.C., Steinke, T.: ZIB Structure Prediction Pipeline:
Composing a Complex Biological Workflow through Web Services. In: Nagel,
W.E., Walter, W.V., Lehner, W. (eds.) Euro-Par 2006. LNCS, vol. 4128,
pp. 1148--1158. Springer, Heidelberg (2006)

\bibitem{book} Foster, I., Kesselman, C.: The Grid: Blueprint for a New Computing
Infrastructure. Morgan Kaufmann, San Francisco (1999)

\bibitem{proceeding1} Czajkowski, K., Fitzgerald, S., Foster, I., Kesselman, C.: Grid
Information Services for Distributed Resource Sharing. In: 10th IEEE
International Symposium on High Performance Distributed Computing, pp.
181--184. IEEE Press, New York (2001)

\bibitem{proceeding2} Foster, I., Kesselman, C., Nick, J., Tuecke, S.: The Physiology of the
Grid: an Open Grid Services Architecture for Distributed Systems
Integration. Technical report, Global Grid Forum (2002)

\bibitem{url} National Center for Biotechnology Information, \url{http://www.ncbi.nlm.nih.gov}

\end{thebibliography}

\section*{Appendix: Springer-Author Discount}

LNCS authors are entitled to a 33.3\% discount off all Springer
publications. Before placing an order, the author should send an email, 
giving full details of his or her Springer publication,
to \url{orders-HD-individuals@springer.com} to obtain a so-called token. This token is a
number, which must be entered when placing an order via the Internet, in
order to obtain the discount.

\section{Checklist of Items to be Sent to Volume Editors}
Here is a checklist of everything the volume editor requires from you:

\begin{itemize}
\settowidth{\leftmargin}{{\Large$\square$}}\advance\leftmargin\labelsep
\itemsep8pt\relax
\renewcommand\labelitemi{{\lower1.5pt\hbox{\Large$\square$}}}

\item The final \LaTeX{} source files
\item A final PDF file
\item A copyright form, signed by one author on behalf of all of the
authors of the paper.
\item A readme giving the name and email address of the
corresponding author.
\end{itemize}
\end{document}

### 2012-07-14 examples/canvas/koch.js
add original koch example

@@ -1,0 +1,39 @@
+var deg = Math.PI/180;  // For converting degrees to radians
+
+// Draw a level-n Koch Snowflake fractal on the canvas context c,
+// with lower-left corner at (x,y) and side length len.
+function snowflake(c, n, x, y, len) {
+    c.save();           // Save current transformation
+    c.translate(x,y);   // Translate origin to starting point
+    c.moveTo(0,0);      // Begin a new subpath at the new origin
+    leg(n);             // Draw the first leg of the snowflake
+    c.rotate(-120*deg); // Now rotate 120 degrees counterclockwise
+    leg(n);             // Draw the second leg
+    c.rotate(-120*deg); // Rotate again
+    leg(n);             // Draw the final leg
+    c.closePath();      // Close the subpath
+    c.restore();        // And restore original transformation
+
+    // Draw a single leg of a level-n Koch snowflake.
+    // This function leaves the current point at the end of the leg it has
+    // drawn and translates the coordinate system so the current point is (0,0).
+    // This means you can easily call rotate() after drawing a leg.
+    function leg(n) {
+        c.save();               // Save the current transformation
+        if (n == 0) {           // Nonrecursive case:
+            c.lineTo(len, 0);   //   Just draw a horizontal line
+        }                       //                                       _  _
+        else {                  // Recursive case: draw 4 sub-legs like:  \/
+            c.scale(1/3,1/3);   // Sub-legs are 1/3rd the size of this leg
+            leg(n-1);           // Recurse for the first sub-leg
+            c.rotate(60*deg);   // Turn 60 degrees clockwise
+            leg(n-1);           // Second sub-leg
+            c.rotate(-120*deg); // Rotate 120 degrees back
+            leg(n-1);           // Third sub-leg
+            c.rotate(60*deg);   // Rotate back to our original heading
+            leg(n-1);           // Final sub-leg
+        }
+        c.restore();            // Restore the transformation
+        c.translate(len, 0);    // But translate to make end of leg (0,0)
+    }
+}

### 2012-11-21 papers/ecoop2013/repairing.tex
ecoop2013 draft

\documentclass[american,english,runningheads]{llncs}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{url}
\usepackage[pdftex]{graphicx}
\usepackage[unicode=true,hidelinks]{hyperref}
\usepackage{times}
\usepackage{comment}
\setcounter{tocdepth}{2}

\clubpenalty=10000
\widowpenalty=10000
\tolerance=1000

\newcommand{\ie}{\emph{i.e.}}
\newcommand{\eg}{\emph{e.g.}}
\newcommand{\cf}{\emph{cf.~}}
\newcommand{\etal}{\emph{et al.~}}
\newcommand{\etc}{\emph{etc.}}
\newcommand{\aka}{\emph{a.k.a.}}

\newcommand{\noun}[1]{\textsc{#1}}

\newtheorem{mydef}{Definition}

\input{macros}

\begin{document}

\title{Statically typed Web programming with JavaScript}

\author{Julien \noun{Richard-Foy} \and Olivier \noun{Barais} \and Jean-Marc \noun{Jézéquel}\\
\texttt{\{first\}.\{last\}@irisa.fr}}
\institute{INRIA}

\maketitle

\begin{abstract}
Writing large applications on the Web is challenging for at least two reasons : (1) their distributed nature may lead to code duplication between server and client sides and add distance between code parts, and (2) the JavaScript language, typically used on client-side, has several inadequacies making it hardly suitable to maintain large code base. Several initiatives attempt to solve these problems by providing a way to use a single language for both client and server sides. One of these approaches is to define JavaScript as a compiled embedded DSL in Scala. We followed this path to improve the safety of native APIs for DOM manipulation, event handling and asynchronous programming. We used advanced type constructs of the host language to make the embedded DSL more expressive and used staging to generate efficient code.
\end{abstract}

{\bf Keywords:} Web, programming languages, embedded DSL, type-level programming, Scala

\section{Introduction}

The Web is an appealing platform on which to write applications (\aka{} Rich Internet Applications, RIA) because it makes them easy to deploy on clients and allows large scale innovative collaborative experiences~\cite{Farrell07_RIA,Mikkonen08_SpaghettiJs}. RIAs are characterized by partial updates of the user interface (as opposed to refreshing the whole page with the classic hyperlink navigation), so a large part of the client-side code listens to user interface events (such as mouse clicks), triggers the appropriate action asynchronously on the server using AJAX~\cite{Garrett05_Ajax} and updates the application’s state and the DOM content with a new DOM fragment computed from the response data~\cite{Farrell07_RIA,Busch09_StateOfArtRIA}.

Writing large Web applications is known to be challenging~\cite{Mikkonen08_SpaghettiJs,Preciado05_RIAMethodologyNecessity}, mainly because of the heterogeneous nature of the client-side and server-side environments~\cite{Echeverria09_RIA,Kuuskeri09_PartitioningClientServer}: writing distributed code requires its parts to be consistent together and leads to duplication unless you use the same language in both sides, which, in the case of the Web, means using JavaScript to write the whole application. However this language has several inadequacies making it hardly suitable for large code base (\eg{} no static typing, no module system, no support of \emph{adhoc} polymorphism, \etc). Some other in-browser execution environments give the opportunity to write the client-side code in another language than JavaScript, \eg{} Java applets~\cite{Anuff96_JavaApplets}, Adobe Flash~\cite{Curtis00_Flash} and Microsoft Silverlight\footnote{\href{http://www.microsoft.com/silverlight/}{http://www.microsoft.com/silverlight/}}. However these technologies have several drawbacks: they require an additional browser plugin to be installed (which may not be available on all devices having a Web browser: for instance there is no way to execute Flash objects within an Apple smart phone), the page content can't naturally be referenced by search engines, and the content is not structured in URLs that users can bookmark or share.

An increasing number of initiatives attempt to allow developers to write the client-side code in a language different of JavaScript but that can be compiled to JavaScript (\eg{} GWT~\cite{Chaganti07_GWT}, Dart~\cite{Griffith11_Dart}, TypeScript\footnote{\href{http://www.typescriptlang.org/}{http://www.typescriptlang.org/}}, Roy~\cite{McKenna_Roy}). Some of these languages can also be compiled to another runtime environment usable on server-side (\eg{} GWT, Dart, Kotlin\footnote{\href{http://kotlin.jetbrains.org/}{http://kotlin.jetbrains.org/}}, ClojureScript~\cite{McGranaghan11_ClojureScript}, Fay\footnote{\href{http://fay-lang.org/}{http://fay-lang.org/}}, Haxe~\cite{Cannasse08_HaXe} or Opa\footnote{\href{http://opalang.org/}{http://opalang.org/}}). These languages are usually more suitable to write large applications either because they provide more constructs to build abstractions (such as object orientation), or because they are statically typed, or because they add a bunch of useful concepts that are missing in JavaScript (such as \emph{adhoc} polymorphism or namespaces). Moreover, their ability to compile for both server and client sides usually allows developers to share some parts of code between server and client sides for free. However this shared code is restricted to use exclusively language constructs: concepts provided by external APIs can't be shared between server and client sides because the bindings with these environments are different on the server and client sides.

In this paper we investigate further a path already introduced by Kossakowski \etal{}: defining JavaScript as a compiled embedded DSL in Scala~\cite{Kossakowski12_JsDESL}. This approach is based on Lightweight Modular Staging~\cite{Rompf12_LMSThesis}: a Scala program written using the embedded DSL evaluates to an intermediate representation that can be further processed to perform domain specific optimizations and then to generate a JavaScript program (for the client-side) and a Scala program (for the server-side). This approach has two main advantages: (1) the embedding reduces the effort needed to define the language since we can reuse the Scala infrastructure and tooling, and (2) staging gives more knowledge to the compiler about how a given abstraction should be efficiently translated into the target environment. In other words any library-level abstraction can have the efficiency of a language-level abstraction.

The previous work showed how to write client-side code in Scala using the JavaScript embedded DSL and how to share code between server-side and client-side. The authors also showed that the code was safer and more convenient to write thanks to Scala's static typing and class system. Finally, they showed how to \emph{un-invert} the control of callback-based asynchronous APIs by using continuations but they did not address other concerns of RIAs development such as user interface events handling, DOM creation and manipulation, and application’s state management.

This paper continues their work and presents and discusses the implementation of the following contributions:

\begin{itemize}

% JSDom, Forest
\item We expose native JavaScript APIs for DOM manipulation and event handling as statically typed APIs while reducing their syntactic noise ;

% State monad, FutureOps, OptionOps
\item We define a high-level abstraction to handle program's state changes and an abstraction turning asynchronous computations into first-class values. Then we show how to exploit staging to translate these abstractions into JavaScript with a minimal overhead ;

% NumericOps, Typeclasses
\item We use Scala's advanced type-level constructs to elegantly express type coercion constraints on operands in arithmetic operations and to achieve \emph{adhoc} polymorphism with no overhead at all in the generated code ;

% Forest server-side, cross-browser incompatibilities
\item Our code generation system is flexible enough to handle cross-browsers incompatibilities and, unlike other approaches, to allow the sharing of concepts between client-side and server-side even if these concepts are translated into code using external APIs.

\end{itemize}

The next section reviews the framework used to define compiled embedded DSLs in Scala, sections \ref{improving} to \ref{lms-benefits} present our contributions and section \ref{discussion} concludes.

%\section{Web programming challenges?}

\section{Introduction to Lightweight Modular Staging}
\label{intro-lms}

Lightweight Modular Staging (LMS) is a Scala framework for defining compiled embedded DSLs. The main idea is that a program written using a DSL is evaluated in two stages (or steps): the first stage builds an intermediate representation of the program and the second stage transforms this intermediate representation into executable code. The bindings between stages is type-directed: a value of type \code{Rep[Int]} in the first stage will yield a value of type \code{Int} in the second stage. If you consider the following code:
\begin{lstlisting}
val inc: Rep[Int] => Rep[Int] =
  x => x + 1
\end{lstlisting}
The \code{inc} function returns the intermediate representation of a computation yielding the number following the value of the parameter \code{x} (that itself is the intermediate representation of a number). The function looks like a regular Scala function excepted that its parameter type and its return type are wrapped in the \code{Rep[T]} type constructor that denotes intermediate representations. You can get a \code{T} value from a \code{Rep[T]} value by generating code from the intermediate representation and compiling it:
\begin{lstlisting}
val compiledInc: Int => Int =
  compile(inc)
\end{lstlisting}
The \code{compile} function takes a staged program of type \code{Rep[A] => Rep[B]} and returns a final program of type \code{A => B}.

The intermediate representation implementation is hidden for users but DSLs authors have to provide the corresponding intermediate representation of each construction of their language. For that purpose, LMS comes with an extensible intermediate representation implementation defining computations as a graph of statements. Then, the code generation process consists in sorting this graph according to expressions dependencies and to emit the code corresponding to each node. The listings \ref{codegen-js} and \ref{codegen-scala} show the JavaScript and Scala generated code for the \code{inc} function:
\begin{multicols}{2}
\lstinputlisting[language=JavaScript, caption=JavaScript code generation, label=codegen-js]{inc.js}
\lstinputlisting[caption=Scala code generation, label=codegen-scala]{inc.scala}
\end{multicols}

Defining a DSL consists in three steps, each defining:

\begin{itemize}
\item The concrete syntax: an abstract API manipulating \code{Rep[_]} values~;
\item The intermediate representation: an implementation of the concrete syntax in terms of statement nodes~;
\item A code generator for the intermediate representation: a \emph{pretty-printer} for each DSL statement node.
\end{itemize}

\section{Improving native JavaScript APIs}
\label{improving}

This section shows how we built DSLs improving either the expressivity or the safety or both of the equivalent native JavaScript API.

\subsection{Events}

To handle events using the native JavaScript API you have to attach an event handler for a type of events on an object. An event handler is just a function taking the event as a parameter. Consider the following code printing keys hits and mouse clicks in the console:

\begin{lstlisting}[language=JavaScript,label=event-js,caption=Native JavaScript API to handle events]
window.addEventListener("keyup", function (e) {
  console.log(e.key);
});
window.addEventListener("click", function (e) {
  console.log(e.button);
});
\end{lstlisting}

We want to “translate” the above API in Scala in a statically typed way without adding verbosity or complexity. The difficulty comes from the fact that the type of the event passed to the handler varies with the event name. A value of type \code{KeyboardEvent} contains a \code{key} field returning the value of the key involved in the event, and a value of type \code{MouseEvent} contains a \code{button} field returning the value of the mouse button involved in the event. For instance in the above listing, the first handler processs \code{KeyboardEvent} values because it listens to \code{keyup} events.

A possible solution could be to define a distinct method for each event instead of the single \code{addEventListener} method, so each method takes a handler with the according event type:

\begin{lstlisting}
window.onKeyUp { e: KeyboardEvent =>
  println(e.key)
}
window.onClick { e: MouseEvent =>
  println(e.button)
}
\end{lstlisting}

However, implementing this solution requires a high effort because we have to define as many methods as there are events. We want to provide a single polymorphic method similar to the native API. To achieve that we need to encode at the type-level the dependency relation between an event name and its corresponding event type. This dependency can naturally and elegantly be encoded using dependent method types:

\begin{lstlisting}
def on(ev: EventDef)(handler: Rep[ev.Type] => Rep[Unit]): Rep[Unit]
\end{lstlisting}

(We renamed \code{addEventListener} to \code{on}, for the sake of brevity). An \code{EventDef} value represents an event that carries its corresponding event type in its \code{Type} member. We can then define the \code{keyup} and \code{click} events as follows:

\begin{lstlisting}
object KeyUp extends EventDef { type Type = KeyboardEvent }
object Click extends EventDef { type Type = MouseEvent }
\end{lstlisting}

The following Scala listing is equivalent to the JavaScript listing \ref{event-js} and is completely typesafe: if the user misspells the event name or tries to use an undefined field on an event his program won’t compile.

\begin{lstlisting}
window.on(KeyUp) { e =>
  println(e.key)
}
window.on(Click) { e =>
  println(e.button)
}
\end{lstlisting}

\subsection{DOM manipulation}
\label{forest}

Updating the user interface usually means replacing a part of the DOM with another DOM fragment computed from data fetched by an AJAX request. This requires to write how to compute the new DOM fragment in JavaScript, but the native API for creating DOM is very verbose, making the code hard to reason about. For instance the following listing shows how to build a \code{div} element containing a greeting message:

\begin{lstlisting}[language=JavaScript,label=dom-api,caption=DOM fragment creation using the native API]
var hello = function (name) {
  var root = document.createElement("div");
  root.setAttribute("class", "greeting");
  var helloT = document.createTextNode("Hello, ");
  var recipient = document.createElement("strong");
  var recipientT = document.createTextNode(name.toUpperCase());
  recipient.appendChild(recipientT);
  root.appendChild(helloT);
  root.appendChild(recipient);
  return root
};
\end{lstlisting}

Calling \code{hello("Obama")} produces a DOM fragment equivalent to the following HTML:

\begin{lstlisting}[language=HTML]
<div class="greeting">
  Hello, <strong>OBAMA</strong>
</div>
\end{lstlisting}

Actually, very few developers use the native API for creating DOM, another way to build DOM is to build a String containing the desired markup and then to ask the browser to parse it as HTML:

\begin{lstlisting}[language=JavaScript]
var hello = function (name) {
  return '<div class="greeting">' +
           'Hello, <strong>'+name.toUpperCase()+'</strong>' +
         '</div>'
};
\end{lstlisting}

This implementation may be more readable, however the code is wrong: if we call \code{hello("<me>")} the brackets won’t be escaped and will produce a \code{ME} tag, which is not the expected behavior. So, although this way reads better it’s not really less error prone.

HTML template engines like Mustache\footnote{\href{http://mustache.github.com/}{http://mustache.github.com/}} or Closure Templates\footnote{\href{https://developers.google.com/closure/templates/}{https://developers.google.com/closure/templates/}} aim to solve these problems by providing a convenient syntax to describe the HTML structure and allowing the insertion of dynamic expressions in safe way. However, template engines require templates to be compiled, so they add an extra step in the build chain. Some template engines can be used on both client-side and server-side but they have severe limitations (\eg{} Closure Template has no expression language) and are not statically typed. We’ll show in section \ref{lms-benefits} how we solved this problem.

In this section we show how we can define, as an embedded DSL, with a small effort, a template engine that is statically typed, able to insert dynamic content in a safe way, that provides a powerful expression language and requires no extra compilation step.

Since the template engine is defined as en embedded DSL, we can reuse Scala’s constructs:

\begin{itemize}
\item a function taking some parameters and returning a DOM fragment directly models a template taking parameters and returning a DOM fragment~;
\item the type system typechecks template definitions and template calls~;
\item the Scala language itself is the expression language~;
\item compiling a template is the same as compiling user code.
\end{itemize}

So the only remaining work consists in defining the DSL vocabulary to define DOM nodes. We provide a \code{tag} function to define a tag and a \code{text} function to define a text node. The following listing uses our DSL and generates a code equivalent to the listing \ref{dom-api}:

\begin{lstlisting}[label=forest-hello,caption=DOM definition DSL]
def hello(name: Rep[String]) =
  tag("div", "class"->"greeting") {
    text("Hello, "),
    tag("strong") { text(name.toUpperCase) }
  }
\end{lstlisting}

The readability has been highly improved: nesting tags is just like nesting code blocks, HTML entities are automatically escaped in text nodes and developers have the full computational power of Scala to inject dynamic data, DOM fragments definitions are written using functions so they compose just as functions compose.

\section{High-level abstractions without performance penalty}

This section shows how we used staging to remove the overhead of high-level DSLs.

\subsection{Null values}

Null values are a known source of problems in programming languages~\cite{Hoare09_Null,Nanda09_Null}. For example, consider the following typical code finding a particular widget in the page and a then particular button in the widget:

\begin{lstlisting}[language=JavaScript,label=null-unsafe,caption=Unsafe code]
var loginWidget = document.querySelector("div.login");
var loginButton = loginWidget.querySelector("button.submit");
loginButton.addEventListener("click", function (e) { ... });
\end{lstlisting}

The native \code{querySelector} method returns \code{null} if no node matched the given selector in the document. If we run the above code in a page where the widget is not present, it will throw an error and stop further JavaScript execution. We can write defensive code to handle \code{null} cases, but it leads to very cumbersome code:

\begin{lstlisting}[language=JavaScript,label=efficient-null,caption=Defensive programming to handle null references]
var loginWidget = document.querySelector("div.login");
if (loginWidget !== null) {
  var loginButton = loginWidget.querySelector("button.submit");
  if (loginButton !== null) {
    loginButton.addEventListener("click", function (e) { ... });
  }
}
\end{lstlisting}

The most popular JavaScript library, jQuery~\cite{Bibeault08_jQuery}, used by more than 40\% of the top million sites\footnote{\href{http://trends.builtwith.com/javascript}{http://trends.builtwith.com/javascript}} handles this problem at runtime by wrapping each query result in a container so before each further method call it tests the emptiness of the container and applies effectively the operation only if the container is not empty. Rewriting the above code using jQuery gives the following listing:

\begin{lstlisting}
var loginWidget = $("div.login");
var loginButton = loginWidget.find("button.submit");
loginButton.on("click", function (e) { ... });
\end{lstlisting}

We can see that it has as few noise as in the listing \ref{null-unsafe} but it has the safety of the listing \ref{efficient-null}. However, the checking involves a slight performance penalty since each result is wrapped and then unwrapped for the subsequent method invocation.

Our solution has both the safety and performance of the listing \ref{efficient-null} and the expressiveness of the listing \ref{null-unsafe}. We encode the potential emptiness of a value of type \code{Rep[A]} using the \code{Rep[Option[A]]} type at the stage level but we don’t generate code wrapping values in a container, instead we just check if a value is \code{null} or not. Here is a listing using our DSL:

\begin{lstlisting}
for {
  loginWidget <- document.find("div.login")
  loginButton <- loginWidget.find("submit.button")
} loginButton.on(Click) { e => ... }
\end{lstlisting}

\subsection{State monad}

Managing the state of an application is cumbersome and fragile in JavaScript due to the lack of encapsulation constructs. For example, consider the following function incrementing a counter and returning its previous value:
\begin{lstlisting}[language=JavaScript,label=state-fragile,caption=Fragile state handling]
var x;
var inc = function () {
  var prev = x;
  x = x + 1;
  return prev
};

// Usage
x = 42;
inc();
var y = inc();
alert(y);
\end{lstlisting}
This code suffers from two problems: the variable \code{x} is global and if the user calls the \code{inc} function without initializing first the \code{x} variable the behavior will be unexpected.

To workaround these issues, developers can wrap the variable declaration and the \code{inc} function code in a factory function (cf listing~\ref{state-encapsulated}). Thus the state would be hidden and developers could not forget the initialization.

\begin{lstlisting}[language=JavaScript,label=state-encapsulated,caption=State encapsulation within a function]
var incFactory = function (init) {
  var x = init;
  return function () {
    var prev = x;
    x = x + 1;
    return prev
  }
};

// Usage
var inc = incFactory(42);
inc();
var x = inc();
alert(x);
\end{lstlisting}

The code is now safer, but it is also bigger, and it still uses a mutable state. It’s known that mutable objects in the code make this one harder to reason about~\cite{Grogono94_Immutability}. Getting rid of mutable objects is simple: instead of modifying a state, a function takes its initial value as a parameter and returns its new value in its result. However, explicitly threading the state in all functions parameters is not desired because of the syntactic and performance cost of adding an extra parameter to each function and of returning two results instead of one:

\begin{lstlisting}[language=JavaScript,label=state-explicit,caption=Explicit state threading]
var inc = function (state) {
  return {
    state: state + 1,
    value: state
  }
};

// Usage
var x1 = inc(42);
var x2 = inc(x1.state);
alert(x2.value);
\end{lstlisting}

The State monad~\cite{Wadler92_StateM} is a functional programming pattern that avoids to explicitly thread the state in functions using it or modifying it. In Scala, the \code{for} notation provides an elegant way to sequence monadic computations, and thus to implicitly thread a state across functions, as shown in the following listing:

\begin{lstlisting}[label=state-monad,caption=Implicitly threaded state using the State Monad]
val inc = for {
  x <- get[Int]
  _ <- put(x + 1)
} yield x

// Usage
val usage = for {
  _ <- inc
  x <- inc
} yield alert(x)
usage.init(42)
\end{lstlisting}

The \code{get} and \code{put} primitives are part of the State monad DSL, they allow to retrieve the current state and to change it. The \code{usage} value is a reusable piece of program that calls two times the \code{inc} function and prints the result of the second call. The last line of the listing runs the \code{usage} program on an initial state.

The State monad programming model has the advantages of both listings \ref{state-encapsulated} and \ref{state-explicit} because there is no global variable and everything is immutable, but it is less verbose than these listings.

Our DSL implementation produces an intermediate representation semantically equivalent to the monadic sequences but using regular variables, so that there is no abstraction penalty: the listing \ref{state-monad} compiles to code equivalent to the listing \ref{state-fragile}.

\subsection{Future monad}

RIAs perform Ajax requests to asynchronously fetch data and then process them in a callback (inversion of control). Dealing with asynchronous code is hard because of the well known “callback hell”. Kossakowski \etal{} proposed a way to \emph{un-invert} the control of callback-based APIs using continuations so developers don’t really see that their code is asynchronous. For example, consider the following code:
\begin{lstlisting}[label=async-continuation,caption=Asynchronous computations using continuations]
sleep(1000)
println("Hello")
sleep(1000)
println(", World!")
\end{lstlisting}

The code looks synchronous however the \code{sleep} calls are asynchronous: they translate to the following JavaScript:
\begin{lstlisting}[language=JavaScript]
window.setTimeout(function () {
  console.log("Hello");
  window.setTimeout(function () {
    console.log(", World!");
  }, 1000);
}, 1000);
\end{lstlisting}

Actually, the listing \ref{async-continuation} does not compile: at some point developers have to think of the asynchronous translation of their code to delimit the end of the callback generated by the code translation. Consider these two variants:

\begin{lstlisting}[label=async-v1,caption=Version 1]
reset {
  sleep(1000)
  println("Hello")
  sleep(1000)
  println(", World!")
  println(42)
}
\end{lstlisting}
\begin{lstlisting}[label=async-v2,caption=Version 2]
reset {
  sleep(1000)
  println("Hello")
  sleep(1000)
  println(", World!")
}
println(42)
\end{lstlisting}

The two above version behave differently: in listing \ref{async-v1} the last \code{println} statement is included in the callback while it is not the case in listing \ref{async-v2}. So this last listing will print \code{42} immediately. Sometimes the enclosing \code{reset} may be hidden in a function call:

\begin{lstlisting}[label=async-ambiguous,caption=Ambiguous reading]
reset {
  foo {
    sleep(1000)
    println("A")
  }
  println("B")
}
\end{lstlisting}

If \code{foo} already contains a call to \code{reset}, the above code will print \code{BA} instead of \code{AB} as we may expect.

Because asynchronous computations may not be reflected explicitly in the types and because developers have to look backward to the enclosing \code{reset} and then forward to see what part of their code block will be captured by the continuation, we think this programming model may be confusing.

Instead, we propose a monadic DSL that does not attempt to hide its asynchronous nature but provides a way to lift asynchronous values as first-class citizen and, thanks to the \code{for} notation, a convenient syntax to avoid the callback-hell.

An asynchronous value of type \code{Rep[A]} is modelled by a value of type \code{Rep[Future[A]]}. This type has the following methods:
\begin{itemize}
\item \code{foreach(f: Rep[A] => Rep[Unit]): Rep[Unit]}, registering a callback to eventually do something with the value when available~;
\item \code{map(f: Rep[A] => Rep[B]): Rep[Future[B]]}, to eventually transform the value when available~;
\item \code{flatMap(f: Rep[A] => Rep[Future[B]]): Rep[Future[B]]} to eventually transform the value when available~;
\end{itemize}

The \code{map} and \code{flatMap} methods allow asynchronous values to be transformed and passed around, thus turning them into first-class citizen.

Because callbacks are explicitly passed to the \code{foreach}, \code{map} and \code{flatMap} methods, there is no risk of ambiguity such as the one of listing \ref{async-ambiguous}. The Scala \code{for} notation handles the callback-hell problem.

The listing \ref{async-v2} can be re-written as follows with our DSL:

\begin{lstlisting}
for {
  _ <- sleep(1000)
  _ <- future(println("Hello"))
  _ <- sleep(1000)
  _ <- future(println(", World!"))
} ()
println(42)
\end{lstlisting}

The \code{for} notation can intuitively be thought of as a sequencing notation: whenever the first \code{sleep} call ends, the next statement will be called, and so on. The \code{println(42)} statement which is outside of the sequence won’t be part of it, as expected.

The value of the \code{for} notation is more evident when asynchronous computations return values:

\begin{lstlisting}
val fooF = Ajax.get("/foo")
val barF = Ajax.get("/bar")
for {
  foo <- fooF
  bar <- barF
} println(foo + bar)
\end{lstlisting}

% TODO mentionner le traitement de cas d’erreurs ?

\section{Type-level abstractions}
\label{type-level}

This section shows how we encoded some abstractions at the type level to increase the expressivity of our DSL with no overhead at all in the generated code.

\subsection{Arithmetic operation operands coercion}

JavaScript comes with only one numeric type, \code{Number}, that represents double-precision 64-bit values. If developers want to use another numeric data type, \eg{} representing integer values or arbitrary length values, they can’t use the native arithmetic operators (\code{+}, \code{-}, \etc) on them, making the code less readable because the same concepts (\eg{} adding two numeric values) are described by different operators or methods. Furthermore, the code is also harder to write when it comes to mix different types of numeric values. Indeed, this requires to manually take care of their compatibility: before adding an arbitrary length value with a \code{Number} value, developers have to convert the latter in the former’s type.

Our DSL solves these problems by defining numeric operations in a safe and extensible way: each operation is a polymorphic method and operand coercion is expressed using functional dependencies~\cite{Jones00_FunDeps}. It allows developers to use a homogeneous syntax for arithmetic operations with different numeric types and to mix heterogeneous types in a same operation. For example, the add operation is defined as follows:

\begin{lstlisting}
def numeric_plus[A : Numeric](lhs: Rep[A], rhs: Rep[A]): Rep[A]
\end{lstlisting}

The \code{A : Numeric} syntax means that the function is polymorphic on any type \code{A}, as long as there is an available \code{Numeric[A]} value.

Then we add a \code{+} method to any \code{Rep[_]} value, using an implicit class:

\begin{lstlisting}
implicit class NumericOps(lhs: Rep[A]) {
  def + (rhs: Rep[A]) = numeric_plus(lhs, rhs)
}
\end{lstlisting}

However, the above code does not handle operands coercion yet. To do so, we want to express that an expression \code{a + b} is valid either if \code{a} and \code{b} have the same numeric type, or if one of them has a numeric type and the other can be converted to this numeric type. The result type of such an expression depends on the types of the operands, \eg{} an integer and a double value produce a double value, two integer values produce another integer value. We express these constraints using functional dependencies:

\begin{lstlisting}
implicit class NumericOps(lhs: Rep[A]) {
  def + [B, C](rhs: Rep[B])(implicit ev: (A ~ B) ~> C): Rep[C] =
    numeric_plus(ev.lhs(lhs), ev.rhs(rhs))
}
\end{lstlisting}

The \code{ev} parameter has type \code{(A ~ B) ~> C}, which means \emph{combining a \code{A} and a \code{B} gives a \code{C}}. Then we need to define which combinations of \code{A}, \code{B} and \code{C} are valid:

\begin{lstlisting}
implicit def sameType[A : Numeric] =
  new ((A ~ A) ~> A)(identity, identity)
implicit def promoteLhs[A, B : Numeric](implicit aToB: A => B) =
  new ((A ~ B) ~> B)(a => convert[B](a), identity)
implicit def promoteRhs[A, B : Numeric](implicit aToB: A => B) =
  new ((B ~ A) ~> B)(identity, a => convert[B](a))
\end{lstlisting}

The \code{sameType} value says that two values of type \code{A} give another value of type \code{A}. The \code{promoteLhs} value says that a value of type \code{A} and a value of a numeric type \code{B} can be used as operands as long as there is a way to convert the value of type \code{A} to a value of type \code{B} (so the left hand side is promoted to the type \code{B}). The \code{promoteRhs} does the same for the right hand side.

The \code{\~\>} type takes two parameters, saying how to obtain a value of the target type from each operand. Here is the definition of the \code{\~\>} type:

\begin{lstlisting}
trait Args { type Lhs; type Rhs }
class ~>[A <: Args, B : Numeric](
  val lhs: Rep[A#Lhs] => Rep[B],
  val rhs: Rep[A#Rhs] => Rep[B]
)
\end{lstlisting}

\subsection{\emph{Adhoc} polymorphism}

Consider the following listing showing how to define a \code{show} method, returning a text describing the object on which the method is called, on a \code{User} and a \code{Point} data types:

\begin{lstlisting}[language=JavaScript]
// Type `User`
var User = function (name, age) {
  this.name = name;
  this.age = age;
};

// Type `Point`
var Point = function (x, y) {
  this.x = x;
  this.y = y;
};

// Implementation of `show` for a `User`
User.prototype.show = function () {
  return this.name + " (" + this.age + ")"
};

// Implementation of `show` for a `Point`
Point.prototype.show = function () {
  return "(" + this.x + ", " + this.y + ")"
};

// Main program
var main = function (user, point) {
  console.log(user.show());
  console.log(point.show());
};
\end{lstlisting}

We define the \code{show} method on the prototypes of the target data types \code{User} and \code{Point}, so when the main program calls the \code{show} method on an object it is automatically dispatched to the right implementation. However, modifying object prototypes is considered to be a bad practice~\cite{Zakas12_MaintainableJs}. Another way could consist in manually code the dispatch logic, by registering supported data types at the beginning of the program execution, as described in section 2.4.3 of~\cite{Abelson83_SICP}, but this solution is painful for developers and incurs a performance overhead.

Kossakowski \etal{} showed how to use Scala classes to have \emph{adhoc} polymorphism without having to play directly with JavaScript prototypes. However their solution suffers from some limitations: it doesn’t support open extension, it looses information on the side-effects performed by a method call (so it prevents some optimizations to be applied), it is limited to single inheritance (and even single interface inheritance) and does not support constructor parameters.

We propose another way to achieve \emph{adhoc} polymorphism that is entirely type-directed (the dispatch happens at compile-time rather than at runtime), with no impact on the generated code and supporting open extension, using typeclasses~\cite{Wadler89_AdhocPolymorphism,Odersky06_Typeclasses,Oliveira10_Typeclasses}.

\begin{lstlisting}
// Interface
case class Show[A](show: Rep[A => String])

// Polymorphic function
def show[A : Show](a: Rep[A]) = implicitly[Show[A]].show(a)

// Type `User`
type User = Record { val name: String; val age: Int }
// Implementation of Show for a User
implicit val showUser =
  Show[User](user => user.name + " (" + user.age + " years)")

// Type `Point`
type Point = Record { val x: Double; val y: Double }
// Implementation of Show for a Point
implicit val showPint =
  Show[Point](point => "(" + point.x + ", " + point.y + ")")

// Main program
def main(user: Rep[User], point: Rep[Point]) = {
  println(show(user))
  println(show(point))
}
\end{lstlisting}
The generated code looks like the following:
\begin{lstlisting}[language=JavaScript]
var main = function (user, point) {
  var x0 = showUser(user);
  console.log(x0);
  var x1 = showPoint(point);
  console.log(x1);
};
var showUser = function (user) {
  var x2 = user.name + "(";
  var x3 = x2 + user.age;
  var x4 = x3 + " years)";
  return x4
};
var showPoint = function (point) {
  var x5 = "(" + point.x;
  var x6 = x5 + ", ";
  var x7 = x6 + point.y;
  var x8 = x7 + ")";
  return x8
};
\end{lstlisting}
Because the dispatch is performed at compile-time, the JavaScript code calls the right function according to its parameters, even though there is no way, in JavaScript, to get back the type of these parameters.

\section{Modular code generation and aspects composition}
\label{lms-benefits}

\subsection{Reuse the DOM creation DSL on server-side}

Markup generation is a task that often needs to be performed from both server and client sides. From the server-side it produces HTML pages that search engines can crawl and from the client-side it produces DOM fragments that can be used to update the user interface. We want to share HTML templates on both server and client sides to get consistency in the rendering and to avoid to duplicate the markup structure.

A few existing template engines can be used on both server and client sides. They provide a convenient syntax to describe a HTML structure in which you can inject dynamic values. However, these template engines are limited because they’re not statically typed and they don’t provide a convenient expression language for the dynamic values. These two problems are in fact two aspects of a same problem: if we write the expression \code{xs.size == 0} in a template, how should it be translated in JavaScript and in Scala? Several solutions are valid, depending on the meaning of the expression: in JavaScript we could translate it as \code{xs.size === 0} or \code{xs.length === 0} if \code{xs} refers to a collection (in JavaScript the property to get the size of a collection is named \code{length}). So we need to define a proper expression language and to define how it should be translated into JavaScript and Scala, but such a task requires a high effort and is hard to achieve in an extensible way (so users can still use their own data types in expressions instead of being restricted to a set of supported types). That’s probably for this reason that existing template engines usable on server and client sides have no expression language at all: the dynamic content can only be provided as a map of key-value pairs and no operation can be done on them.

Our DSL presented in section \ref{forest} is equivalent as a template engine with Scala as the expression language. Making it usable on both server and client sides is as simple as defining another code generator for the DSL, producing Scala code. The result is a lightweight statically typed template engine, seamlessly integrated with the other code parts (because it is defined as an embedded DSL) and with a convenient expression language. For instance, the template written in the listing \ref{forest-hello} produces the following Scala code usable on server-side (the generated code for client-side is roughly equivalent to the listing \ref{dom-api}):

\begin{lstlisting}
def hello(name: String) = {
  val x0 = {xml.Text("Hello, ")}
  val x1 = name.toUpperCase
  val x2 = {xml.Text(x1)}
  val x3 = <strong>{x2}</strong>
  val x4 = <div class="greeting">{List(x0, x3)}</div>
  x4
}
\end{lstlisting}

The only system, which we are aware of, that allows developers to share their HTML templates on both server and client sides in a statically typed way and with a convenient expression language is Opa. In Opa they made this possible because HTML templating is part of the Opa language itself. If they wanted to support the sharing of another concept such as form field validation rules they would need to introduce this concept as a part of their Opa language.

On the other hand, in our case introducing the support of form field validation as a DSL usable on both server and client sides would be as simple as defining an API and its translation into both Scala and JavaScript. We wouldn’t need to change the Scala language itself.

\subsection{Custom code generation to handle browsers incompatibilities}

Each component of a DSL is defined in a trait and each corresponding code generator for a given target is also defined in a trait. The Scala language allows developers to mix several traits together. Developers can build their language by picking the DSL traits they want to use.

Furthermore, a code generator can be tweaked by defining a sub-trait specializing some of its methods. For example, the \code{forEach} method on JavaScripts arrays was not supported by Internet Explorer before version 9. As some users may still have old versions of Internet Explorer we want to send them a specialized version of JavaScript files to handle this incompatibility. We can do that by extending the code generator for the \code{List} operations to specialize the generation of the \code{ListForeach} node to generate a JavaScript \code{for} loop instead of a \code{forEeach} call:

\begin{lstlisting}
trait JSGenListIE extends JSGenList {
 import IR._
 override def emitNode(sym: Sym[Any], rhs: Def[Any]) =
  rhs match {
   case ListForeach(l, x, b) =>
    var i = fresh[Int]
    stream.println("for (var "+quote(i)+" = 0 ; "+
                        quote(i)+" < "+quote(l)+".length ; "+
                        quote(i)+"++) {")
    emitValDef(x, quote(l)+"["+quote(i)+"]")
    emitBlock(b)
    stream.println("}")
   case _ => super.emitNode(sym, rhs)
  }
}
\end{lstlisting}

Internet Explorer suffers from some others incompatibilities, \eg{} its events API is different of the standard. On the same way we can define a \code{JSGenDomIE} trait generating code targeting the Internet Explorer’s events API for the \code{Dom} DSL.

Then, we can pack together all the traits handling Internet Explorer’s incompatibilities in a \code{IESupport} trait:

\begin{lstlisting}
trait IESupport extends JSGenListIE with JSGenDomIE {
  val IR: ListOpsExp with DomExp
}
\end{lstlisting}

And, finally, we can mix this \code{IESupport} trait with any JavaScript code generator to target Internet Explorer:

\begin{lstlisting}
val jsGen = new JSGen with IESupport {
  val IR: prog.type = prog
}
\end{lstlisting}

\section{Discussion}
\label{discussion}

What did we do?

\bibliographystyle{ieeetr}
\bibliography{biblio}

\end{document}

### 2012-11-21 papers/ecoop2013/inc.scala

@@ -1,0 +1,4 @@
+def inc(x0: Int): Int = {
+  val x1 = x0 + 1
+  x1
+}

### 2012-11-23 papers/ecoop2013/repairing.tex
Update (but still not finished yet)

@@
 
-Actually, very few developers use the native API for creating DOM, another way to build DOM is to build a String containing the desired markup and then to ask the browser to parse it as HTML:
+Another way to build DOM is to build a String containing the desired markup and then to ask the browser to parse it as HTML:
 
@@
 This implementation may be more readable, however the code is wrong: if we call \code{hello("<me>")} the brackets won’t be escaped and will produce a \code{ME} tag, which is not the expected behavior.
-So, although this way reads better it’s not really less error prone.
+So, although this way reads better it’s not less error prone.
 
@@
 
-The readability has been highly improved: nesting tags is just like nesting code blocks, HTML entities are automatically escaped in text nodes and developers have the full computational power of Scala to inject dynamic data, DOM fragments definitions are written using functions so they compose just as functions compose.
+The readability has been highly improved: nesting tags is just like nesting code blocks, HTML entities are automatically escaped in text nodes, developers have the full computational power of Scala to inject dynamic data and DOM fragments definitions are written using functions so they compose just as functions compose.
 
@@
 
-\begin{lstlisting}[language=JavaScript,label=efficient-null,caption=Defensive programming to handle null references]
+\begin{lstlisting}[language=JavaScript,label=null-defensive,caption=Defensive programming to handle null references]
 var loginWidget = document.querySelector("div.login");
@@
 
-We can see that it has as few noise as in the listing \ref{null-unsafe} but it has the safety of the listing \ref{efficient-null}.
+We can see that it has as few noise as in the listing \ref{null-unsafe} but it has the safety of the listing \ref{null-defensive}.
 However, the checking involves a slight performance penalty since each result is wrapped and then unwrapped for the subsequent method invocation.
 
-Our solution has both the safety and performance of the listing \ref{efficient-null} and the expressiveness of the listing \ref{null-unsafe}.
+Our solution has both the safety and performance of the listing \ref{null-defensive} and the expressiveness of the listing \ref{null-unsafe}.
 We encode the potential emptiness of a value of type \code{Rep[A]} using the \code{Rep[Option[A]]} type at the stage level but we don’t generate code wrapping values in a container, instead we just check if a value is \code{null} or not.
-Here is a listing using our DSL:
+Here is a listing that uses our DSL and compiles to code equivalent to the listing \ref{null-defensive}:
 
@@
 
-To workaround these issues, developers can wrap the variable declaration and the \code{inc} function code in a factory function (cf listing~\ref{state-encapsulated}).
-Thus the state would be hidden and developers could not forget the initialization.
+To workaround these issues, developers can wrap the variable declaration and the \code{inc} function code in a factory function.
+Thus the state is hidden and developers may not forget the initialization:
 
@@
 The code is now safer, but it is also bigger, and it still uses a mutable state.
-It’s known that mutable objects in the code make this one harder to reason about~\cite{Grogono94_Immutability}.
-Getting rid of mutable objects is simple: instead of modifying a state, a function takes its initial value as a parameter and returns its new value in its result.
-However, explicitly threading the state in all functions parameters is not desired because of the syntactic and performance cost of adding an extra parameter to each function and of returning two results instead of one:
+It’s known that mutable objects in the code make this one harder to reason about~\cite{Grogono94_Immutability,Kjolstad11_Immutability}.
+Getting rid of mutable objects is simple: instead of modifying a state, a function takes its initial value as a parameter and returns its new value in its result:
 
@@
 
+However, explicitly threading the state in all functions parameters is not desired because of the syntactic and performance cost of adding an extra parameter to each function and of returning two results instead of one.
+Furthermore, manually passing the result of each previous call to the next one is error prone.
+
 The State monad~\cite{Wadler92_StateM} is a functional programming pattern that avoids to explicitly thread the state in functions using it or modifying it.
-In Scala, the \code{for} notation provides an elegant way to sequence monadic computations, and thus to implicitly thread a state across functions, as shown in the following listing:
+In Scala, the \code{for} notation provides an elegant way to sequence monadic computations, and thus to implicitly thread a state across computations, as shown in the following listing:
 
@@
 
-The State monad programming model has the advantages of both listings \ref{state-encapsulated} and \ref{state-explicit} because there is no global variable and everything is immutable, but it is less verbose than these listings.
+The State monad programming model has the advantages but not the shortcomings of both listings \ref{state-encapsulated} and \ref{state-explicit} because there is no global variable (the state is implicitly threaded thanks to the State monad but state modifications explicitly use the \code{get} and \code{put} primitives) and everything is immutable.
 
-Our DSL implementation produces an intermediate representation semantically equivalent to the monadic sequences but using regular variables, so that there is no abstraction penalty: the listing \ref{state-monad} compiles to code equivalent to the listing \ref{state-fragile}.
+Furthermore, our DSL implementation produces an intermediate representation semantically equivalent to the monadic sequences but using regular variables, so that there is no abstraction penalty: the listing \ref{state-monad} compiles to code equivalent to the listing \ref{state-fragile}.
 
@@
 The code looks synchronous however the \code{sleep} calls are asynchronous: they translate to the following JavaScript:
-\begin{lstlisting}[language=JavaScript]
+
+\begin{lstlisting}[language=JavaScript,label=async-js,caption=Generated code for listing \ref{async-continuation}]
 window.setTimeout(function () {
@@
 
-Instead, we propose a monadic DSL that does not attempt to hide its asynchronous nature but provides a way to lift asynchronous values as first-class citizen and, thanks to the \code{for} notation, a convenient syntax to avoid the callback-hell.
+In this paper we propose a DSL that explicitly reflects the asynchronous nature of computations in their type and provides methods turning them into first-class citizen.
+Our DSL is monadic so we can solve the callback-hell problem thanks to the Scala \code{for} notation.
 
@@
 Because callbacks are explicitly passed to the \code{foreach}, \code{map} and \code{flatMap} methods, there is no risk of ambiguity such as the one of listing \ref{async-ambiguous}.
-The Scala \code{for} notation handles the callback-hell problem.
 
@@
 
-% TODO mentionner le traitement de cas d’erreurs ?
+The above listing produces the following code (we renamed some identifiers for the sake of readability):
+
+\begin{lstlisting}[language=JavaScript]
+var x1 = new Promise();
+AjaxGet("/foo", x1);
+var x2 = new Promise();
+AjaxGet("/bar", x2);
+x1.foreach(function (foo) {
+  x2.foreach(function (bar) {
+    var x3 = foo + bar;
+    console.log(x3);
+  });
+});
+\end{lstlisting}
+
+The \code{Promise} constructor code has been omitted, it creates an object holding a list of callbacks to call when the asynchronous value is completed.
+The \code{AjaxGet} function code has also been omitted, it creates a \code{XMLHttpRequest} object that sends an HTTP request and completes its promise parameter with the response, when received.
+
+\begin{lstlisting}
+def f(aF: Rep[Future[String]]) =
+  for (a <- aF) yield a.toUpperCase
+def g(aF: Rep[Future[String]]) =
+  for (a <- aF) yield a.toLowerCase
+
+def fAndG(aF: Rep[Future[String]]) = for {
+  u <- f(aF)
+  l <- g(aF)
+} yield u + l
+\end{lstlisting}
+
+(TODO Explain the value of the above listing...)
 
@@
 
+The following listing demonstrates the homogeneous syntax to perform an addition on different data types and the operand coercion between heterogeneous types:
+
+\begin{lstlisting}
+def main(i: Rep[Int], j: Rep[Int],
+         x: Rep[Double], y: Rep[Double]) = {
+  println(i + j)
+  println(x + y)
+  // i is promoted to Rep[Double], n has type Rep[Double]
+  val n = i + y
+  // j is promoted to Rep[Double], m has type Rep[Double]
+  val m = x + j
+  println(n + m)
+}
+\end{lstlisting}
+
 \subsection{\emph{Adhoc} polymorphism}
@@
 
-Kossakowski \etal{} showed how to use Scala classes to have \emph{adhoc} polymorphism without having to play directly with JavaScript prototypes.
-However their solution suffers from some limitations: it doesn’t support open extension, it looses information on the side-effects performed by a method call (so it prevents some optimizations to be applied), it is limited to single inheritance (and even single interface inheritance) and does not support constructor parameters.
+Kossakowski \etal{} showed how to use Scala traits to have \emph{adhoc} polymorphism without having to play directly with JavaScript prototypes.
+However their solution suffers from some limitations: it doesn’t support retroactive extension, it does not track the side-effects potentially performed by a method call (so it prevents some optimizations to be applied by LMS), it is limited to single inheritance (and even single interface inheritance) and does not support constructor parameters.
 
-We propose another way to achieve \emph{adhoc} polymorphism that is entirely type-directed (the dispatch happens at compile-time rather than at runtime), with no impact on the generated code and supporting open extension, using typeclasses~\cite{Wadler89_AdhocPolymorphism,Odersky06_Typeclasses,Oliveira10_Typeclasses}.
+We propose another way to achieve \emph{adhoc} polymorphism that is entirely type-directed (the dispatch happens at compile-time rather than at runtime), with no impact on the generated code and supporting retroactive extension, using typeclasses~\cite{Wadler89_AdhocPolymorphism,Odersky06_Typeclasses,Oliveira10_Typeclasses}.
 
@@
 \end{lstlisting}
-Because the dispatch is performed at compile-time, the JavaScript code calls the right function according to its parameters, even though there is no way, in JavaScript, to get back the type of these parameters.
+
+(TODO Conclusion)
 
@@
 
-What did we do?
+We exposed native JavaScript APIs as statically typed APIs while decreasing their verbosity.
 

### 2012-11-23 papers/ecoop2013/repairing.tex
Another abstract

@@
 \begin{abstract}
-Writing large applications on the Web is challenging for at least two reasons : (1) their distributed nature may lead to code duplication between server and client sides and add distance between code parts, and (2) the JavaScript language, typically used on client-side, has several inadequacies making it hardly suitable to maintain large code base.
-Several initiatives attempt to solve these problems by providing a way to use a single language for both client and server sides.
-One of these approaches is to define JavaScript as a compiled embedded DSL in Scala.
-We followed this path to improve the safety of native APIs for DOM manipulation, event handling and asynchronous programming.
-We used advanced type constructs of the host language to make the embedded DSL more expressive and used staging to generate efficient code.
+Rich Internet Applications involve more code on the client-side, dealing with DOM manipulation, event handling and asynchronous calls to the server.
+Writing and maintaining large JavaScript code base is challenging because this language has several inadequacies (no static typing, no module system, no \emph{adhoc} polymorphism support, verbose syntax, \etc{}).
+We investigate further the definition of JavaScript as a Scala compiled embedded DSL using Lightweight Modular Staging.
+We define safe and expressive abstractions for DOM manipulation, event handling and asynchronous programming that translate to efficient JavaScript code using native APIs.
+We use both type-level information and staging to introduce high-level abstractions with a few or no overhead on the generated JavaScript code.
 \end{abstract}
 
-{\bf Keywords:} Web, programming languages, embedded DSL, type-level programming, Scala
+{\bf Keywords:} Web, programming languages, embedded DSL, type-level programming, staging, Scala
 

### 2012-11-29 src/main/scala/scala/js/OptionOps.scala
Existential types FTW ©

@@ -3,42 +3,43 @@ package scala.js
 import virtualization.lms.common._
 
 trait OptionOps { this: Base =>
-  implicit class OptionOpsCls[+A : Manifest](o: Rep[Option[A]]) {
-    def foreach(f: Rep[A] => Rep[Unit]) = option_foreach(o, f)
-    def map[B : Manifest](f: Rep[A] => Rep[B]) = option_map(o, f)
-    def flatMap[B : Manifest](f: Rep[A] => Rep[Option[B]]) = option_flatMap(o, f)
-    def isEmpty = option_isEmpty(o)
-    def fold[B : Manifest](none: => Rep[B], some: Rep[A] => Rep[B]) = option_fold(o, none, some)
+
+  implicit def OptionOpsCls[A : Manifest](o: Rep[Option[A]]): OptionOpsCls[A]
+  type OptionOpsCls[+A] <: OptionOpsBase[A]
+
+  trait OptionOpsBase[+A] {
+    def foreach(f: Rep[A] => Rep[Unit]): Rep[Unit]
+    def map[B : Manifest](f: Rep[A] => Rep[B]): Rep[Option[B]]
+    def flatMap[B : Manifest](f: Rep[A] => Rep[Option[B]]): Rep[Option[B]]
+    def isEmpty: Rep[Boolean]
+    def fold[B : Manifest](none: => Rep[B], some: Rep[A] => Rep[B]): Rep[B]
   }
-  def option_foreach[A : Manifest](o: Rep[Option[A]], f: Rep[A] => Rep[Unit]): Rep[Unit]
-  def option_map[A : Manifest, B : Manifest](o: Rep[Option[A]], f: Rep[A] => Rep[B]): Rep[Option[B]]
-  def option_flatMap[A : Manifest, B : Manifest](o: Rep[Option[A]], f: Rep[A] => Rep[Option[B]]): Rep[Option[B]]
-  def option_isEmpty[A : Manifest](o: Rep[Option[A]]): Rep[Boolean]
-  def option_fold[A : Manifest, B : Manifest](o: Rep[Option[A]], none: => Rep[B], some: Rep[A] => Rep[B]): Rep[B]
 }
 
 trait OptionOpsExp extends OptionOps with EffectExp {
-  def option_foreach[A : Manifest](o: Exp[Option[A]], f: Exp[A] => Exp[Unit]) = {
-    val a = fresh[A]
-    val block = reifyEffects(f(a))
[…]

### 2012-11-30 papers/ecoop2013/repairing.tex
Layout refactoring

@@
 Some of these languages can also be compiled to another runtime environment usable on server-side (\eg{} GWT, Dart, Kotlin\footnote{\href{http://kotlin.jetbrains.org/}{http://kotlin.jetbrains.org/}}, ClojureScript~\cite{McGranaghan11_ClojureScript}, Fay\footnote{\href{http://fay-lang.org/}{http://fay-lang.org/}}, Haxe~\cite{Cannasse08_HaXe} or Opa\footnote{\href{http://opalang.org/}{http://opalang.org/}}).
-These languages are usually more suitable to write large applications either because they provide more constructs to build abstractions (such as object orientation), or because they are statically typed, or because they add a bunch of useful concepts that are missing in JavaScript (such as \emph{adhoc} polymorphism or namespaces).
+These languages are usually more suitable to write large applications either because they provide more constructs to build abstractions (such as object orientation), or because they are statically typed, or because they add a bunch of useful concepts that are
+missing in JavaScript (such as \emph{adhoc} polymorphism or namespaces).
 Moreover, their ability to compile for both server and client sides usually allows developers to share some parts of code between server and client sides for free.
@@
 
-\section{Introduction to Lightweight Modular Staging}
+\section{Background and motivating example}
+
+To illustrate the challenges of writing a RIA, we introduce a simple “mind mapping” application and show how the client-side code would be typically written in pure JavaScript.
+
+\subsection{Example description}
+
+In our application, a mind map is represented as a graph where ideas or concepts are represented by vertices, and the relation between two ideas is represented by an edge.
+Users can visualize a mind map, zoom in and out and move it.
+They can also add ideas and link them together.
+The figure \ref{mindmap} shows the visualization of a simple map consisting of three ideas linked together, using our application.
+
+\begin{figure}
+ \caption{A simple mind map as visualized with our application}
+ \centering
+ \includegraphics[width=8cm]{mindmap.png}
+ \label{mindmap}
+\end{figure}
+
+\subsection{Implementation in JavaScript}
+
+\subsubsection{Events handling}
+
+RIAs distinguish from classic Web pages by a higher degree of interactivity.
+In our example we want to let users zoom in and out a map using their mouse wheel.
+To implement this feature we need to attach an event handler for the \code{mousewheel} event on the DOM element containing the map.
+The following listing shows how to attach such an event handler using the native JavaScript API:
+
+\begin{lstlisting}[language=JavaScript,label=event-js,caption=Native JavaScript API to handle events]
+mapElement.addEventListener("mousewheel", function (e) {
+  scale = Math.round(scale + e.wheelDeltaY / 16);
+  // ... update the user interface with the new scale
+});
+\end{lstlisting}
+
+The event handler is simply a function taking the event data as a parameter. In the above example we use the \code{wheelDeltaY} property of the event to change the scale of the map.
+
+We think this code is fragile for two reasons.
+(1) The name of the event is passed as a string, so it is easy to mispell it.
+(2) The callback passed as a second parameter takes a parameter \code{e} whose fields vary according to the listened event but developers have no way to check that the fields they’re using are indeed defined on the event they’re listening to.
+
+\subsubsection{Asynchronous programming}
+
+Another characteristic of RIAs is that requests to the server are often performed asynchronously: instead of asking the browser to reload the whole page, the JavaScript code has to perform a request and to process the response, when available, to update a part of the user interface.
+The following listing shows how to send a request creating a vertex on the server and to insert it to the user interface:
+
+\begin{lstlisting}
+var createVertex = function (text) {
+  Ajax.post("/create", { content: text }, function (vertex) {
+    addVertex(vertex);
+  });
+}
+\end{lstlisting}
+
+\code{Ajax.post} is a helper function that sends a HTTP request to the server and calls its last parameter (that is a \emph{callback}) when the response data is available.
+The inverted control makes modularization harder to achieve: in the above listing the function creating the vertex is also responsible to updating the user interface.
+If we want to relax this coupling the only option is to add a callback parameter to the \code{createVertex} function:
+
+\begin{lstlisting}[language=JavaScript,label=async-js]
+var createVertex = function (text, callback) {
+  Ajax.get("/create", { content: text }, function (data) {
+    callback(data);
+  });
+}
+
+createVertex("Hello, World!", function (vertex) {
+  addVertex(vertex);
+});
+\end{lstlisting}
+
+However increasing the number of callbacks makes the code flow harder to follow and adds distance to the programmer initial intent.
+
+Another issue with callback-driven programming arises when several dependent asynchronous computations are executed sequentially.
+Consider for example the following listing performing three consecutive Ajax requests:
+
+\begin{lstlisting}[language=JavaScript]
+Ajax.get(fooUrl(), function (foo) {
+  Ajax.get(barUrl(foo), function (bar) {
+    Ajax.get(bazUrl(bar), function (baz) {
+      console.log(foo + bar + baz);
+    });
+    console.log("bar = " + bar)
+  });
+});
+\end{lstlisting}
+
+Notice that the code is getting deeper toward the right, this problem is often named “callback-hell”.
+This code is also hard to reason about: when will the second \code{console.log} statement be executed?
+Before or after \code{baz} has been fetched?
+
+\subsubsection{DOM manipulation}
+\label{forest}
+
+Updating the user interface usually means replacing a part of the DOM with another DOM fragment computed from data fetched by an AJAX request.
+This requires to write how to compute the new DOM fragment in JavaScript, but the native API for creating DOM is very verbose, making the code hard to reason about.
+For instance the following listing shows how to build a DOM tree representing a vertex in the mind map:
+
+\begin{lstlisting}[language=JavaScript,label=dom-api,caption=DOM fragment creation using the native API]
+var vertexDom = function (v) {
+  var root = document.createElement("g");
+  root.setAttribute("class", "vertex");
+  root.setAttribute("transform",
+    "translate(" + v.x + "," + v.y + ")");
+  var rect = document.createElement("rect");
+  rect.setAttribute("width", v.width);
+  rect.setAttribute("height", v.height);
+  var text = document.createElement("text");
+  text.setAttribute("width", v.width);
+  text.setAttribute("height", v.height);
+  text.appendChild(document.createTextNode(v.content));
+  root.appendChild(rect);
+  root.appendChild(text);
+  return root
+};
+\end{lstlisting}
+
+For instance the following call:
+\begin{lstlisting}
+vertexDom({
+  x: 10, y: 10,
+  width: 100, height: 60,
+  content: "Hello, World!"
+});
+\end{lstlisting}
+produces a DOM tree equivalent to the following HTML:
+\begin{lstlisting}[language=HTML]
+<g class=vertex transform="translate(10,10)">
+  <rect width=100 height=60 />
+  <text width=100 height=60>Hello, World!</text>
+</g>
+\end{lstlisting}
+
+Not only the \code{vertexDom} function is very verbose, but it does not help to see the markup structure.
+
+Another way to build a DOM tree is to build a String containing the desired markup and then to ask the browser to parse it as HTML:
+
+\begin{lstlisting}[language=JavaScript]
+var vertexDom = function (v) {
+  return '<g class=vertex ' +
+             'transform="translate('+v.x+','+v.y+')">' +
+           '<rect width='+v.width+' height='+v.height+' />' +
+           '<text width='+v.width+' height='+v.height+'>' +
+              v.content +
+           '</text>' +
+         '</g>'
+};
+\end{lstlisting}
+
+This implementation may be more readable, however the code is wrong: if we call it with a content containing brackets (\eg{} \code{"<foo>"}) they won’t be escaped and will produce a \code{foo} tag nested in the \code{text} tag, which is not the intended behavior.
+So, although this way reads better it’s not less error prone.
+
+By the way, markup generation is a task that often needs to be performed from both server and client sides.
+From the server-side it produces HTML pages that search engines can crawl and from the client-side it produces DOM fragments that can be used to update the user interface.
+We want to share HTML templates on both server and client sides to get consistency in the rendering and to avoid to duplicate the markup structure.
+
+HTML template engines like Mustache\footnote{\href{http://mustache.github.com/}{http://mustache.github.com/}} or Closure Templates\footnote{\href{https://developers.google.com/closure/templates/}{https://developers.google.com/closure/templates/}} aim to make simpler the definition of DOM fragments by providing a convenient syntax to describe the HTML structure and allowing the insertion of dynamic expressions in safe way.
+However, template engines require templates to be compiled, so they add an extra step in the build chain.
+Some template engines can be used on both client-side and server-side but they have severe limitations (\eg{} Closure Template has no expression language) and are not statically typed.
+
+\subsection{Assessment}
+
+- Few static checks
+
+- Code hard to reason about
+
+- Code too much verbose
+
+A few existing template engines can be used on both server and client sides.
+They provide a convenient syntax to describe a HTML structure in which you can inject dynamic values.
+However, these template engines are limited because they’re not statically typed and they don’t provide a convenient expression language for the dynamic values.
+These two problems are in fact two aspects of a same problem: if we write the expression \code{xs.size == 0} in a template, how should it be translated in JavaScript and in Scala?
+Several solutions are valid, depending on the meaning of the expression: in JavaScript we could translate it as \code{xs.size === 0} or \code{xs.length === 0} if \code{xs} refers to a collection (in JavaScript the property to get the size of a collection is named \code{length}).
+So we need to define a proper expression language and to define how it should be translated into JavaScript and Scala, but such a task requires a high effort and is hard to achieve in an extensible way (so users can still
+use their own data types in expressions instead of being restricted to a set of supported types).
+That’s probably for this reason that existing template engines usable on server and client sides have no expression language at all: the dynamic content can only be provided as a map of key-value pairs and no operation can be done on them.
+
+... so we need better ways to handle DOM definition, user interaction events and asynchronous programming.
+
+\section{DSL design and implementation}
+
+This section presents both the design and the implementation of a compiled embedded DSL addressing the problems introduced in the preceding section.
+
+\subsection{Introduction to Lightweight Modular Staging}
 \label{intro-lms}
 
+We chose to implement our DSL as a compiled embedded DSL in Scala using Lightweight Modular Staging.
+The embedding lets us reuse the host language infrastructure (compiler, development tools) and type system, and staging gives the opportunity to generate efficient code from expressive code by applying domain specific optimizations.
+
 Lightweight Modular Staging (LMS) is a Scala framework for defining compiled embedded DSLs.
-The main idea is that a program written using a DSL is evaluated in two stages (or steps): the first stage builds an intermediate representation of the program and the second stage transforms this intermediate representation into executable code.
-The bindings between stages is type-directed: a value of type \code{Rep[Int]} in the first stage will yield a value of type \code{Int} in the second stage.
-If you consider the following code:
+The main idea is that a program written using a DSL is evaluated in two stages (or steps): the first stage builds an intermediate representation of the program and the second stage transforms this intermediate representation into executable code (figure \ref{lms-diagram}).
+
+\begin{figure}
+  \centering
+  \includegraphics[width=7cm]{lms.pdf}
+  \caption{Compilation of a program using LMS}
+  \label{lms-diagram}
+\end{figure}
+
+The bindings between stages is type-directed: a value of type \code{Rep[Int]} in the first stage will yield a value of type \code{Int} in the second stage. If you consider the following code:
 \begin{lstlisting}
@@
 
-\section{Improving native JavaScript APIs}
-\label{improving}
+\subsection{Events handling}
 
-This section shows how we built DSLs improving either the expressivity or the safety or both of the equivalent native JavaScript API.
-
-\subsection{Events}
-
-To handle events using the native JavaScript API you have to attach an event handler for a type of events on an object.
-An event handler is just a function taking the event as a parameter.
-Consider the following code printing keys hits and mouse clicks in the console:
-
-\begin{lstlisting}[language=JavaScript,label=event-js,caption=Native JavaScript API to handle events]
-window.addEventListener("keyup", function (e) {
-  console.log(e.key);
-});
-window.addEventListener("click", function (e) {
-  console.log(e.button);
-});
-\end{lstlisting}
-
-We want to “translate” the above API in Scala in a statically typed way without adding verbosity or complexity.
+We want to bring safety to the events handling API so that developers can’t mispell an event name and can’t attempt to read a property that is not defined on the type of the handled event.
 The difficulty comes from the fact that the type of the event passed to the handler varies with the event name.
-A value of type \code{KeyboardEvent} contains a \code{key} field returning the value of the key involved in the event, and a value of type \code{MouseEvent} contains a \code{button} field returning the value of the mouse button involved in the event.
-For instance in the above listing, the first handler processs \code{KeyboardEvent} values because it listens to \code{keyup} events.
+A value of type \code{KeyboardEvent} contains a \code{key} field returning the value of the key involved in the event, and a value of type \code{MouseWheelEvent} contains a \code{wheelDeltaY} field returning the value of the mouse button involved in the event.
+For instance in the listing \ref{event-js}, the handler processses \code{MouseWheelEvent} values because it listens to \code{mousewheel} events.
 
@@
 }
-window.onClick { e: MouseEvent =>
-  println(e.button)
+window.onMouseWheel { e: MouseWheelEvent =>
+  println(e.wheelDeltaY)
 }
@@
 To achieve that we need to encode at the type-level the dependency relation between an event name and its corresponding event type.
-This dependency can naturally and elegantly be encoded using dependent method types:
+This dependency can naturally and elegantly be encoded using dependent method types~\cite{Oliveira10_Typeclasses}:
 
 \begin{lstlisting}
+class EventDef {
+  type Type
+}
+
 def on(ev: EventDef)(handler: Rep[ev.Type] => Rep[Unit]): Rep[Unit]
@@
 (We renamed \code{addEventListener} to \code{on}, for the sake of brevity).
-An \code{EventDef} value represents an event that carries its corresponding event type in its \code{Type} member.
-We can then define the \code{keyup} and \code{click} events as follows:
+An \code{EventDef} value represents an event that carries its corresponding event type in its \code{Type} type member.
+The \code{on} method takes a parameter \code{ev} of type \code{EventDef} and a \code{handler} parameter whose type refers to the \code{Type} member of the \code{ev} parameter, so the type of the handler depends on the \code{ev} value.
+We can then define the \code{keyup} and \code{mousewheel} events as follows:
 
@@
 object KeyUp extends EventDef { type Type = KeyboardEvent }
-object Click extends EventDef { type Type = MouseEvent }
+object MouseWheel extends EventDef { type Type = MouseWheelEvent }
 \end{lstlisting}
 
-The following Scala listing is equivalent to the JavaScript listing \ref{event-js} and is completely typesafe: if the user misspells the event name or tries to use an undefined field on an event his program won’t compile.
+The following Scala listing is equivalent to the JavaScript listing \ref{event-js} and is completely typesafe: if the user misspells the event name or tries to use an undefined property on an event his program won’t compile.
 
 \begin{lstlisting}
-window.on(KeyUp) { e =>
-  println(e.key)
-}
-window.on(Click) { e =>
-  println(e.button)
+window.on(MouseWheel) { e =>
+  scale = Math.round(scale + e.wheelDeltaY / 16)
 }
@@
 
-\subsection{DOM manipulation}
-\label{forest}
+(FIXME Should I show, for the sake of example, the traits defining the user facing DSL, its implementation and the code generator?)
 
-Updating the user interface usually means replacing a part of the DOM with another DOM fragment computed from data fetched by an AJAX request.
-This requires to write how to compute the new DOM fragment in JavaScript, but the native API for creating DOM is very verbose, making the code hard to reason about.
-For instance the following listing shows how to build a \code{div} element containing a greeting message:
+\subsection{Asynchronous programming}
 
-\begin{lstlisting}[language=JavaScript,label=dom-api,caption=DOM fragment creation using the native API]
-var hello = function (name) {
-  var root = document.createElement("div");
-  root.setAttribute("class", "greeting");
-  var helloT = document.createTextNode("Hello, ");
-  var recipient = document.createElement("strong");
-  var recipientT = document.createTextNode(name.toUpperCase());
-  recipient.appendChild(recipientT);
-  root.appendChild(helloT);
-  root.appendChild(recipient);
-  return root
-};
+We propose a DSL that explicitly reflects the asynchronous nature of computations in their type and provides methods turning them into first-class citizen.
+Our DSL is monadic so we can solve the callback-hell problem thanks to the Scala \code{for} notation.
+
+An asynchronous value of type \code{Rep[A]} is modelled by a value of type \code{Rep[Future[A]]}. This type has the following methods:
+\begin{itemize}
+\item \code{foreach(f: Rep[A] => Rep[Unit]): Rep[Unit]}, registering a callback to eventually do something with the value when available~;
+\item \code{map(f: Rep[A] => Rep[B]): Rep[Future[B]]}, to eventually transform the value when available~;
+\item \code{flatMap(f: Rep[A] => Rep[Future[B]]): Rep[Future[B]]} to eventually transform the value when available~;
+\end{itemize}
+
+The \code{map} and \code{flatMap} methods allow asynchronous values to be transformed and passed around, thus turning them into first-class citizen.
+
+The listing \ref{async-js} can be re-written as follows with our DSL:
+
+\begin{lstlisting}
+def createVertex(text: Rep[String]): Rep[Future[Vertex]] =
+  Ajax.post[Vertex]("/create", new Record { val content = text })
+
+// Usage
+val vertexAsync = createVertex("Hello, World!")
+for (vertex <- vertexAsync) {
+  addVertex(vertex)
+}
 \end{lstlisting}
 
-Calling \code{hello("Obama")} produces a DOM fragment equivalent to the following HTML:
+The \code{createVertex} now returns an asynchronous value instead of taking a callback as parameter.
 
-\begin{lstlisting}[language=HTML]
-<div class="greeting">
-  Hello, <strong>OBAMA</strong>
-</div>
+And the “callback-hell” example can be re-written as follows:
+
+\begin{lstlisting}
+for {
+  foo <- Ajax.get(fooUrl())
+  bar <- Ajax.get(barUrl(foo))
+  _ <- future(println("bar = " + bar))
+  baz <- Ajax.get(bazUrl(bar))
+} yield println(foo + bar + baz)
 \end{lstlisting}
 
-Another way to build DOM is to build a String containing the desired markup and then to ask the browser to parse it as HTML:
+The \code{for} notation can intuitively be thought of as a sequencing notation: whenever the response of the first Ajax request is available, the next statement will be executed, and so on.
+
+(FIXME Update the following or skip?) The above listing produces the following code (we renamed some identifiers for the sake of readability):
 
 \begin{lstlisting}[language=JavaScript]
-var hello = function (name) {
-  return '<div class="greeting">' +
-           'Hello, <strong>'+name.toUpperCase()+'</strong>' +
-         '</div>'
-};
+var x1 = new Promise();
+AjaxGet("/foo", x1);
+var x2 = new Promise();
+AjaxGet("/bar", x2);
+x1.foreach(function (foo) {
+  x2.foreach(function (bar) {
+    var x3 = foo + bar;
+    console.log(x3);
+  });
+});
 \end{lstlisting}
 
-This implementation may be more readable, however the code is wrong: if we call \code{hello("<me>")} the brackets won’t be escaped and will produce a \code{ME} tag, which is not the expected behavior.
-So, although this way reads better it’s not less error prone.
+The \code{Promise} constructor code has been omitted, it creates an object holding a list of callbacks to call when the asynchronous value is completed.
+The \code{AjaxGet} function code has also been omitted, it creates a \code{XMLHttpRequest} object that sends an HTTP request and completes its promise parameter with the response, when received.
 
-HTML template engines like Mustache\footnote{\href{http://mustache.github.com/}{http://mustache.github.com/}} or Closure Templates\footnote{\href{https://developers.google.com/closure/templates/}{https://developers.google.com/closure/templates/}} aim to solve these problems by providing a convenient syntax to describe the HTML structure and allowing the insertion of dynamic expressions in safe way.
-However, template engines require templates to be compiled, so they add an extra step in the build chain.
-Some template engines can be used on both client-side and server-side but they have severe limitations (\eg{} Closure Template has no expression language) and are not statically typed.
-We’ll show in section \ref{lms-benefits} how we solved this problem.
+% \begin{lstlisting}[caption=Concrete syntax of the asynchronous programming DSL]
+% trait FutureOps { this: Base =>
+% 
+%   implicit def FutureOpsCls[A](f: Rep[Future[A]]): FutureOpsCls[A]
+%   type FutureOpsCls[+A] <: FutureOpsBase[A]
+% 
+%   trait FutureOpsBase[+A] {
+%     def map[B](f: Rep[A] => Rep[B]): Rep[Future[B]]
+%     def flatMap[B](f: Rep[A] => Rep[Future[B]]): Rep[Future[B]]
+%     def foreach(f: Rep[A] => Rep[Unit]): Rep[Unit]
+%     def withFilter(f: Rep[A] => Rep[Boolean]): Rep[Future[A]]
+%   }
+% }
+% \end{lstlisting}
+% 
+% The \code{FutureOps} module depends on the \code{Base} module defined by LMS.
+The \code{Base} module essentially defines the abstract type constructor \code{Rep[_]}.
+Values of type \code{Rep[Future[A]]} have no methods because the \code{Rep[_]} type is abstract and unbound, so we use an implicit conversion from \code{Rep[Future[A]]} to \code{FutureOpsCls[A]} to add methods to \code{Rep[Future[A]]} values.
+The type \code{FutureOpsCls[A]} is bound by the \code{FutureOpsBase[A]} trait that defines the signature of the methods added to \code{Rep[Future[A]]} values.
+Everything in the \code{FutureOps} module is abstract, these definitions only provide the vocabulary of the DSL.
+% 
+% The listing \ref{futureopsexp} shows the definition of the \code{FutureOpsExp} module that implements the \code{FutureOps} module and defines how statement nodes are built from the DSL constructs.
+% 
+% \begin{figure}
+% \begin{lstlisting}[label=futureopsexp,caption=Definition of the intermediate representation of the asynchronous programming DSL]
+% trait FutureOpsExp extends FutureOps { this: EffectExp =>
+% 
+%   implicit class FutureOpsCls[+A](a: Exp[Future[A]])
+%       extends FutureOpsBase[A] {
+%     def map[B](f: Exp[A] => Exp[B]) = {
+%       val x = fresh[A]
+%       val b = reifyEffects(f(x))
+%       reflectEffect(FutureMap(a, x, b), summarizeEffects(b).star)
+%     }
+% 
+%     // ... same for other methods (flatMap, foreach and withFilter)
+%   }
+%   
+%   case class FutureMap[A, B](a: Exp[Future[A]], x: Sym[A], b: Block[B])
+%       extends Def[Future[B]]
+%   // ... same for other operations
+% }
+% \end{lstlisting}
+% \end{figure}
+% 
+% The \code{FutureOpsExp} module depends on the \code{EffectExp} module that is a module refining the \code{Base} module, defining the type \code{Rep[A]} as \code{Exp[A]}, which is a base class defining intermediate representations as a graph of statements.
+The \code{EffectExp} module also provides tools to handle effectful statements.
+The implicit class \code{FutureOpsCls} implements both the abstract type and the implicit conversion of the same name defined in the \code{FutureOps} trait.
+Each method of the \code{FutureOpsCls} class is reified to a statement node, for instance the \code{map} method returns a \code{FutureMap} node containing a reference to the node on which the \code{map} operation is applied and the code block that was passed as a parameter to the \code{map} operation.
+We propagate the information about the effects performed by the statements in the code block by calling \code{reflectEffect} on our statement node.
+% 
+% The listing \ref{jsgenfutureops} shows the code of the JavaScript code generator for the DSL.
+The code generation process basically consists in traversing the graph of statement nodes and in emitting the code corresponding to each node.
+To implement a code generator for a given DSL we just need to say how to generate code for this DSL statement nodes.
+% 
+% \begin{figure}
+%  \begin{lstlisting}[label=jsgenfutureops,caption=JavaScript code generator for the asynchronous programming DSL]
+% trait JSGenFutureOps extends JSNestedCodegen {
+%   val IR: EffectExp with FutureOpsExp
+%   import IR._
+% 
+%   override def emitNode(sym: Sym[Any], rhs: Def[Any]) = rhs match {
+%     case FutureMap(a, x, b) =>
+%       emitValDef(sym, "new Promise()")
+%       stream.println(quote(a) + ".onComplete(function (" + quote(x) + ") {")
+%       emitBlock(b)
+%       stream.println(quote(sym) + ".complete(" + quote(getBlockResult(b)) + ");")
+%       stream.println("});")
+%     // ... same for each node
+%     case _ => super.emitNode(sym, rhs)
+%   }
+% }
+%  \end{lstlisting}
+% \end{figure}
+% 
+% The \code{JSGenFutureOps} code generator inherits from the \code{JSNestedCodegen} code generator that defines the building blocks of the JavaScript code generation.
+The main method is \code{emitNode}, it is called by LMS for each traversed node of the statement graph.
+In the listing we only showed the code generation for the \code{FutureMap} node: when the asynchronous value on which the \code{map} method has been applied is available, we apply the function passed to the \code{map} method on the value and complete a newly created asynchronous value with the result.
+The \code{Promise} constructor is provided by a JavaScript preamble and creates an asynchronous value with two methods: \code{onComplete}, registering a callback to be called when the value is available, and \code{complete}, fulfilling the promise with a value.
 
-In this section we show how we can define, as an embedded DSL, with a small effort, a template engine that is statically typed, able to insert dynamic content in a safe way, that provides a powerful expression language and requires no extra compilation step.
+\subsection{DOM definition}
 
-Since the template engine is defined as en embedded DSL, we can reuse Scala’s constructs:
+In this section we show how we can define, as an embedded DSL, with a small effort, a template engine that is statically typed, able to insert dynamic content in a safe way, that provides a powerful expression language, requires no extra compilation step and that can be used on both client-side and server-side.
+
+Because the template engine is defined as en embedded DSL, we can reuse Scala’s constructs:
 
@@
 \begin{lstlisting}[label=forest-hello,caption=DOM definition DSL]
-def hello(name: Rep[String]) =
-  tag("div", "class"->"greeting") {
-    text("Hello, "),
-    tag("strong") { text(name.toUpperCase) }
-  }
+def vertexDom(v: Rep[Vertex]) =
+  tag("g", "class"->"vertex",
+           "transform"->("translate("+v.x+","+v.y+")"))(
+    tag("rect", "width"->v.width, "height"->v.height)(),
+    tag("text", "width"->v.width, "height"->v.height)(
+      text(v.content)
+    )
+  )
 \end{lstlisting}
@@
 
-\section{High-level abstractions without performance penalty}
+\subsubsection{Reuse the DOM definition DSL from server-side}
 
-This section shows how we used staging to remove the overhead of high-level DSLs.
+(TODO More on the exploit of sharing typesafe expressions between client and server sides)
 
-\subsection{Null values}
+Our DSL is equivalent as a template engine with Scala as the expression language.
+Making it usable on both server and client sides is as simple as defining another code generator for the DSL, producing Scala code.
+The result is a lightweight statically typed template engine, seamlessly integrated with the other code parts (because it is defined as an embedded DSL) and with a convenient expression language.
+For instance, the template written in the listing \ref{forest-hello} produces the following Scala code usable on server-side (the generated code for client-side is roughly equivalent to the listing \ref{dom-api}):
 
-Null values are a known source of problems in programming languages~\cite{Hoare09_Null,Nanda09_Null}.
+\begin{lstlisting}
+def vertexDom(v: Vertex) = {
+  val x0 =
+    <text width="{v.width}" height="{v.height}">
+      {v.content}
+    </text>
+  val x1 = <rect width="{v.width}" height="{v.height}" />
+  val x2 =
+    <g class="vertex" transform="translate({v.x},{v.y})">
+      {List(x0, x1)}
+    </g>
+  x2
+}
+\end{lstlisting}
+
+The only system, which we are aware of, that allows developers to share their HTML templates on both server and client sides in a statically typed way and with a convenient expression language is Opa.
+In Opa they made this possible because HTML templating is part of the Opa language itself.
+If they wanted to support the sharing of another concept such as form field validation rules they would need to introduce this concept as a part of their Opa language.
+
+On the other hand, in our case introducing the support of form field validation as a DSL usable on both server and client sides would be as simple as defining an API and its translation into both Scala and JavaScript.
+We wouldn’t need to change the Scala language itself.
+
+\subsection{More general purpose improvements of JavaScript}
+
+This section presents some more general purpose improvements of JavaScript.
+
+\subsubsection{Null references}
+
+Null references are a known source of problems in programming languages~\cite{Hoare09_Null,Nanda09_Null}.
 For example, consider the following typical code finding a particular widget in the page and a then particular button in the widget:
@@
 
-\subsection{State monad}
+\subsubsection{State monad}
 
@@
 
-\subsection{Future monad}
+\subsubsection{\emph{Adhoc} polymorphism}
 
-RIAs perform Ajax requests to asynchronously fetch data and then process them in a callback (inversion of control).
-Dealing with asynchronous code is hard because of the well known “callback hell”.
-Kossakowski \etal{} proposed a way to \emph{un-invert} the control of callback-based APIs using continuations so developers don’t really see that their code is asynchronous.
-For example, consider the following code:
-\begin{lstlisting}[label=async-continuation,caption=Asynchronous computations using continuations]
-sleep(1000)
-println("Hello")
-sleep(1000)
-println(", World!")
-\end{lstlisting}
+Because of the dynamically typed nature of JavaScript, when calling a function there is no proper way to call a specialized implementation according to the function’s parameters types.
+JavaScript is only able to dispatch according to a method receiver prototype, \eg{} if one writes \code{foo.bar()} the JavaScript runtime will look into the prototype of the \code{foo} object for a property named \code{bar} and will call it.
+So, the only way to achieve \emph{adhoc} polymorphism on JavaScript objects consists in defining the polymorphic function on the prototypes of the objects.
+However, modifying existing object prototypes is considered to be a bad practice~\cite{Zakas12_MaintainableJs}.
+Another way could consist in manually code the dispatch logic, by registering supported data types at the beginning of the program execution, as described in section 2.4.3 of~\cite{Abelson83_SICP}, but this solution is painful for developers and incurs a performance overhead.
 
-The code looks synchronous however the \code{sleep} calls are asynchronous: they translate to the following JavaScript:
+We propose another way to achieve \emph{adhoc} polymorphism that is entirely type-directed (the dispatch happens at compile-time rather than at runtime), with no impact on the generated code and supporting retroactive extension, using typeclasses~\cite{Wadler89_AdhocPolymorphism,Odersky06_Typeclasses,Oliveira10_Typeclasses}.
 
-\begin{lstlisting}[language=JavaScript,label=async-js,caption=Generated code for listing \ref{async-continuation}]
-window.setTimeout(function () {
-  console.log("Hello");
-  window.setTimeout(function () {
-    console.log(", World!");
-  }, 1000);
-}, 1000);
-\end{lstlisting}
+\begin{figure}
+\begin{lstlisting}[label=polymorphism,caption=Adhoc polymorphism using typeclasses]
+// Interface
+case class Show[A](show: Rep[A => Node])
 
-Actually, the listing \ref{async-continuation} does not compile: at some point developers have to think of the asynchronous translation of their code to delimit the end of the callback generated by the code translation.
-Consider these two variants:
+// Polymorphic function
+def listWidget[A : Show](items: Rep[List[A]]): Rep[Node] = {
+  tag("ul")(
+    for (item <- items) yield {
+      tag("li")(implicitly[Show[A]].show(item))
+    }
+  )
+}
 
-\begin{lstlisting}[label=async-v1,caption=Version 1]
-reset {
-  sleep(1000)
-  println("Hello")
-  sleep(1000)
-  println(", World!")
-  println(42)
+// Type `User`
+type User = Record { val name: String; val age: Int }
+// Implementation of Show for a User
+implicit val showUser = Show[User] { user =>
+  tag("span", "class"->"user")(
+    text(user.name + "(" + user.age + " years)")
+  )
+}
+
+// Type `Article`
+type Article = Record { val name: String; val price: Double }
+// Implementation of Show for an Article
+implicit val showArticle = Show[Article] { article =>
+  tag("span")(
+    text(article.name),
+    tag("strong")(text(article.price + " Euros"))
+  )
+}
+
+// Main program
+def main(users: Rep[List[User]], articles: Rep[List[Article]]) = {
+  document.body.append(listWidget(users))
+  document.body.append(listWidget(articles))
 }
 \end{lstlisting}
-\begin{lstlisting}[label=async-v2,caption=Version 2]
-reset {
-  sleep(1000)
-  println("Hello")
-  sleep(1000)
-  println(", World!")
-}
-println(42)
-\end{lstlisting}
+\end{figure}
 
-The two above version behave differently: in listing \ref{async-v1} the last \code{println} statement is included in the callback while it is not the case in listing \ref{async-v2}.
-So this last listing will print \code{42} immediately.
-Sometimes the enclosing \code{reset} may be hidden in a function call:
+The listing \ref{polymorphism} demonstrates how to define a polymorphic \code{listWidget} function that returns a DOM tree containing the representation of a list of items.
+The \code{Show[A]} typeclass defines how to produce a DOM tree for a value of type \code{A}.
 
-\begin{lstlisting}[label=async-ambiguous,caption=Ambiguous reading]
-reset {
-  foo {
-    sleep(1000)
-    println("A")
-  }
-  println("B")
-}
-\end{lstlisting}
-
-If \code{foo} already contains a call to \code{reset}, the above code will print \code{BA} instead of \code{AB} as we may expect.
-
-Because asynchronous computations may not be reflected explicitly in the types and because developers have to look backward to the enclosing \code{reset} and then forward to see what part of their code block will be captured by the continuation, we think this programming model may be confusing.
-
-In this paper we propose a DSL that explicitly reflects the asynchronous nature of computations in their type and provides methods turning them into first-class citizen.
-Our DSL is monadic so we can solve the callback-hell problem thanks to the Scala \code{for} notation.
-
-An asynchronous value of type \code{Rep[A]} is modelled by a value of type \code{Rep[Future[A]]}. This type has the following methods:
-\begin{itemize}
-\item \code{foreach(f: Rep[A] => Rep[Unit]): Rep[Unit]}, registering a callback to eventually do something with the value when available~;
-\item \code{map(f: Rep[A] => Rep[B]): Rep[Future[B]]}, to eventually transform the value when available~;
-\item \code{flatMap(f: Rep[A] => Rep[Future[B]]): Rep[Future[B]]} to eventually transform the value when available~;
-\end{itemize}
-
-The \code{map} and \code{flatMap} methods allow asynchronous values to be transformed and passed around, thus turning them into first-class citizen.
-
-Because callbacks are explicitly passed to the \code{foreach}, \code{map} and \code{flatMap} methods, there is no risk of ambiguity such as the one of listing \ref{async-ambiguous}.
-
-The listing \ref{async-v2} can be re-written as follows with our DSL:
-
-\begin{lstlisting}
-for {
-  _ <- sleep(1000)
-  _ <- future(println("Hello"))
-  _ <- sleep(1000)
-  _ <- future(println(", World!"))
-} ()
-println(42)
-\end{lstlisting}
-
-The \code{for} notation can intuitively be thought of as a sequencing notation: whenever the first \code{sleep} call ends, the next statement will be called, and so on.
-The \code{println(42)} statement which is outside of the sequence won’t be part of it, as expected.
-
-The value of the \code{for} notation is more evident when asynchronous computations return values:
-
-\begin{lstlisting}
-val fooF = Ajax.get("/foo")
-val barF = Ajax.get("/bar")
-for {
-  foo <- fooF
-  bar <- barF
-} println(foo + bar)
-\end{lstlisting}
-
-The above listing produces the following code (we renamed some identifiers for the sake of readability):
-
-\begin{lstlisting}[language=JavaScript]
-var x1 = new Promise();
-AjaxGet("/foo", x1);
-var x2 = new Promise();
-AjaxGet("/bar", x2);
-x1.foreach(function (foo) {
-  x2.foreach(function (bar) {
-    var x3 = foo + bar;
-    console.log(x3);
-  });
-});
-\end{lstlisting}
-
-The \code{Promise} constructor code has been omitted, it creates an object holding a list of callbacks to call when the asynchronous value is completed.
-The \code{AjaxGet} function code has also been omitted, it creates a \code{XMLHttpRequest} object that sends an HTTP request and completes its promise parameter with the response, when received.
-
-\begin{lstlisting}
-def f(aF: Rep[Future[String]]) =
-  for (a <- aF) yield a.toUpperCase
-def g(aF: Rep[Future[String]]) =
-  for (a <- aF) yield a.toLowerCase
-
-def fAndG(aF: Rep[Future[String]]) = for {
-  u <- f(aF)
-  l <- g(aF)
-} yield u + l
-\end{lstlisting}
-
-(TODO Explain the value of the above listing...)
-
-\section{Type-level abstractions}
-\label{type-level}
-
-This section shows how we encoded some abstractions at the type level to increase the expressivity of our DSL with no overhead at all in the generated code.
-
-\subsection{Arithmetic operation operands coercion}
+\subsubsection{Type coercion}
 
@@
 
-\subsection{\emph{Adhoc} polymorphism}
+\subsection{Discussion}
 
-Consider the following listing showing how to define a \code{show} method, returning a text describing the object on which the method is called, on a \code{User} and a \code{Point} data types:
-
-\begin{lstlisting}[language=JavaScript]
-// Type `User`
-var User = function (name, age) {
-  this.name = name;
-  this.age = age;
-};
-
-// Type `Point`
-var Point = function (x, y) {
-  this.x = x;
-  this.y = y;
-};
-
-// Implementation of `show` for a `User`
-User.prototype.show = function () {
-  return this.name + " (" + this.age + ")"
-};
-
-// Implementation of `show` for a `Point`
-Point.prototype.show = function () {
-  return "(" + this.x + ", " + this.y + ")"
-};
-
-// Main program
-var main = function (user, point) {
-  console.log(user.show());
-  console.log(point.show());
-};
-\end{lstlisting}
-
-We define the \code{show} method on the prototypes of the target data types \code{User} and \code{Point}, so when the main program calls the \code{show} method on an object it is automatically dispatched to the right implementation.
-However, modifying object prototypes is considered to be a bad practice~\cite{Zakas12_MaintainableJs}.
-Another way could consist in manually code the dispatch logic, by registering supported data types at the beginning of the program execution, as described in section 2.4.3 of~\cite{Abelson83_SICP}, but this solution is painful for developers and incurs a performance overhead.
-
-Kossakowski \etal{} showed how to use Scala traits to have \emph{adhoc} polymorphism without having to play directly with JavaScript prototypes.
-However their solution suffers from some limitations: it doesn’t support retroactive extension, it does not track the side-effects potentially performed by a method call (so it prevents some optimizations to be applied by LMS), it is limited to single inheritance (and even single interface inheritance) and does not support constructor parameters.
-
-We propose another way to achieve \emph{adhoc} polymorphism that is entirely type-directed (the dispatch happens at compile-time rather than at runtime), with no impact on the generated code and supporting retroactive extension, using typeclasses~\cite{Wadler89_AdhocPolymorphism,Odersky06_Typeclasses,Oliveira10_Typeclasses}.
-
-\begin{lstlisting}
-// Interface
-case class Show[A](show: Rep[A => String])
-
-// Polymorphic function
-def show[A : Show](a: Rep[A]) = implicitly[Show[A]].show(a)
-
-// Type `User`
-type User = Record { val name: String; val age: Int }
-// Implementation of Show for a User
-implicit val showUser =
-  Show[User](user => user.name + " (" + user.age + " years)")
-
-// Type `Point`
-type Point = Record { val x: Double; val y: Double }
-// Implementation of Show for a Point
-implicit val showPint =
-  Show[Point](point => "(" + point.x + ", " + point.y + ")")
-
-// Main program
-def main(user: Rep[User], point: Rep[Point]) = {
-  println(show(user))
-  println(show(point))
-}
-\end{lstlisting}
-The generated code looks like the following:
-\begin{lstlisting}[language=JavaScript]
-var main = function (user, point) {
-  var x0 = showUser(user);
-  console.log(x0);
-  var x1 = showPoint(point);
-  console.log(x1);
-};
-var showUser = function (user) {
-  var x2 = user.name + "(";
-  var x3 = x2 + user.age;
-  var x4 = x3 + " years)";
-  return x4
-};
-var showPoint = function (point) {
-  var x5 = "(" + point.x;
-  var x6 = x5 + ", ";
-  var x7 = x6 + point.y;
-  var x8 = x7 + ")";
-  return x8
-};
-\end{lstlisting}
-
-(TODO Conclusion)
-
-\section{Modular code generation and aspects composition}
-\label{lms-benefits}
-
-\subsection{Reuse the DOM creation DSL on server-side}
-
-Markup generation is a task that often needs to be performed from both server and client sides.
-From the server-side it produces HTML pages that search engines can crawl and from the client-side it produces DOM fragments that can be used to update the user interface.
-We want to share HTML templates on both server and client sides to get consistency in the rendering and to avoid to duplicate the markup structure.
-
-A few existing template engines can be used on both server and client sides.
-They provide a convenient syntax to describe a HTML structure in which you can inject dynamic values.
-However, these template engines are limited because they’re not statically typed and they don’t provide a convenient expression language for the dynamic values.
-These two problems are in fact two aspects of a same problem: if we write the expression \code{xs.size == 0} in a template, how should it be translated in JavaScript and in Scala?
-Several solutions are valid, depending on the meaning of the expression: in JavaScript we could translate it as \code{xs.size === 0} or \code{xs.length === 0} if \code{xs} refers to a collection (in JavaScript the property to get the size of a collection is named \code{length}).
-So we need to define a proper expression language and to define how it should be translated into JavaScript and Scala, but such a task requires a high effort and is hard to achieve in an extensible way (so users can still use their own data types in expressions instead of being restricted to a set of supported types).
-That’s probably for this reason that existing template engines usable on server and client sides have no expression language at all: the dynamic content can only be provided as a map of key-value pairs and no operation can be done on them.
-
-Our DSL presented in section \ref{forest} is equivalent as a template engine with Scala as the expression language.
-Making it usable on both server and client sides is as simple as defining another code generator for the DSL, producing Scala code.
-The result is a lightweight statically typed template engine, seamlessly integrated with the other code parts (because it is defined as an embedded DSL) and with a convenient expression language.
-For instance, the template written in the listing \ref{forest-hello} produces the following Scala code usable on server-side (the generated code for client-side is roughly equivalent to the listing \ref{dom-api}):
-
-\begin{lstlisting}
-def hello(name: String) = {
-  val x0 = {xml.Text("Hello, ")}
-  val x1 = name.toUpperCase
-  val x2 = {xml.Text(x1)}
-  val x3 = <strong>{x2}</strong>
-  val x4 = <div class="greeting">{List(x0, x3)}</div>
-  x4
-}
-\end{lstlisting}
-
-The only system, which we are aware of, that allows developers to share their HTML templates on both server and client sides in a statically typed way and with a convenient expression language is Opa.
-In Opa they made this possible because HTML templating is part of the Opa language itself.
-If they wanted to support the sharing of another concept such as form field validation rules they would need to introduce this concept as a part of their Opa language.
-
-On the other hand, in our case introducing the support of form field validation as a DSL usable on both server and client sides would be as simple as defining an API and its translation into both Scala and JavaScript.
-We wouldn’t need to change the Scala language itself.
-
-\subsection{Custom code generation to handle browsers incompatibilities}
+\subsubsection{Custom code generation to handle browsers incompatibilities}
 
@@
 
-\section{Discussion}
-\label{discussion}
+\section{Comparison with other approaches}
+
+\subsection{DOM definition}
+
+Powerful typesafe expression language shared by client-side and server-side.
+
+\subsection{Events}
+
+Type safety. GWT?
+
+\subsection{Asynchronous programming}
+
+jQuery deferred, Q.js. Previous scala-js paper.
+
+\section{Conclusion and perspectives}
 
 We exposed native JavaScript APIs as statically typed APIs while decreasing their verbosity.
+More code sharing.
 

### 2012-12-03 papers/ecoop2013/repairing.tex
On the way

@@
 \institute{INRIA}
+
+\authorrunning{Julien \noun{Richard-Foy} \etal{}}
 
@@
 
-This paper continues their work and presents and discusses the implementation of the following contributions:
+This paper continues their work and presents safe and expressive DSL constructs for RIAs development.
 
-\begin{itemize}
+% This paper continues their work and presents and discusses the implementation of the following contributions:
+% 
+% \begin{itemize}
+% 
+% % JSDom, Forest
+% \item We expose native JavaScript APIs for DOM manipulation and event handling as statically typed APIs while reducing their syntactic noise ;
+% 
+% % State monad, FutureOps, OptionOps
+% \item We define a high-level abstraction to handle program's state changes and an abstraction turning asynchronous computations into first-class values.
+Then we show how to exploit staging to translate these abstractions into JavaScript with a minimal overhead ;
+% 
+% % NumericOps, Typeclasses
+% \item We use Scala's advanced type-level constructs to elegantly express type coercion constraints on operands in arithmetic operations and to achieve \emph{adhoc} polymorphism with no overhead at all in the generated code ;
+% 
+% % Forest server-side, cross-browser incompatibilities
+% \item Our code generation system is flexible enough to handle cross-browsers incompatibilities and, unlike other approaches, to allow the sharing of concepts between client-side and server-side even if these concepts are translated into code using external APIs.
+% 
+% \end{itemize}
 
-% JSDom, Forest
-\item We expose native JavaScript APIs for DOM manipulation and event handling as statically typed APIs while reducing their syntactic noise ;
-
-% State monad, FutureOps, OptionOps
-\item We define a high-level abstraction to handle program's state changes and an abstraction turning asynchronous computations into first-class values.
-Then we show how to exploit staging to translate these abstractions into JavaScript with a minimal overhead ;
-
-% NumericOps, Typeclasses
-\item We use Scala's advanced type-level constructs to elegantly express type coercion constraints on operands in arithmetic operations and to achieve \emph{adhoc} polymorphism with no overhead at all in the generated code ;
-
-% Forest server-side, cross-browser incompatibilities
-\item Our code generation system is flexible enough to handle cross-browsers incompatibilities and, unlike other approaches, to allow the sharing of concepts between client-side and server-side even if these concepts are translated into code using external APIs.
-
-\end{itemize}
-
-The next section reviews the framework used to define compiled embedded DSLs in Scala, sections \ref{improving} to \ref{lms-benefits} present our contributions and section \ref{discussion} concludes.
+The next section introduces a simple application example showing challenges of RIA development, section \ref{solution} presents our DSL tackling these challenges, section \ref{validation} validates our solution and compares it to related work and section \ref{discussion} concludes.
 
@@
 \section{Background and motivating example}
+\label{problem}
 
@@
 (2) The callback passed as a second parameter takes a parameter \code{e} whose fields vary according to the listened event but developers have no way to check that the fields they’re using are indeed defined on the event they’re listening to.
+For instance, in our case we use the \code{wheelDeltaY} event field that is defined only on the \code{mousewheel} event.
 
@@
   });
-}
+};
 \end{lstlisting}
@@
   });
-}
+};
 
@@
 
-However increasing the number of callbacks makes the code flow harder to follow and adds distance to the programmer initial intent.
+However increasing the number of callbacks makes the code flow harder to follow and adds distance to the programmer's initial intent.
 
@@
 
-Notice that the code is getting deeper toward the right, this problem is often named “callback-hell”.
+Notice that the code is getting deeper toward the right, this problem is often referred to as the “callback-hell”.
 This code is also hard to reason about: when will the second \code{console.log} statement be executed?
@@
 
-Not only the \code{vertexDom} function is very verbose, but it does not help to see the markup structure.
+Not only the \code{vertexDom} function is very verbose, but it does not reflect the markup nested structure, making it hard to read and reason about.
 
@@
 
-This implementation may be more readable, however the code is wrong: if we call it with a content containing brackets (\eg{} \code{"<foo>"}) they won’t be escaped and will produce a \code{foo} tag nested in the \code{text} tag, which is not the intended behavior.
-So, although this way reads better it’s not less error prone.
+This implementation may be more readable, however the above implementation is wrong: if we call it with a content containing brackets (\eg{} \code{"<foo>"}) they won’t be escaped and will produce a \code{foo} tag nested in the \code{text} tag, which is not the intended behavior.
+So, although this way reads slightly better it’s not less error prone.
 
-By the way, markup generation is a task that often needs to be performed from both server and client sides.
+By the way, unlike the previous points markup generation is a task that often needs to be performed from both server and client sides.
 From the server-side it produces HTML pages that search engines can crawl and from the client-side it produces DOM fragments that can be used to update the user interface.
-We want to share HTML templates on both server and client sides to get consistency in the rendering and to avoid to duplicate the markup structure.
+So we want to share HTML fragments definition between both server and client sides to get consistency in the rendering and to avoid duplication.
 
 HTML template engines like Mustache\footnote{\href{http://mustache.github.com/}{http://mustache.github.com/}} or Closure Templates\footnote{\href{https://developers.google.com/closure/templates/}{https://developers.google.com/closure/templates/}} aim to make simpler the definition of DOM fragments by providing a convenient syntax to describe the HTML structure and allowing the insertion of dynamic expressions in safe way.
-However, template engines require templates to be compiled, so they add an extra step in the build chain.
-Some template engines can be used on both client-side and server-side but they have severe limitations (\eg{} Closure Template has no expression language) and are not statically typed.
+Some template engines can be used on both client-side and server-side, however none has a practical expression language (dynamic content can only be provided as a map of key-value pairs and no operation can be done on values) or is statically typed.
 
@@
 
-- Few static checks
+In the previous section we showed the three main tasks performed by the JavaScript code in RIAs: events handling, asynchronous programming and DOM manipulation. We noticed the following issues:
 
-- Code hard to reason about
+\subsubsection{Lack of static checks}
 
-- Code too much verbose
+Dynamic typing may give programmers more flexibility but can make it harder for little script to grow into mature and robust code bases and to perform refactorings.
 
-A few existing template engines can be used on both server and client sides.
-They provide a convenient syntax to describe a HTML structure in which you can inject dynamic values.
-However, these template engines are limited because they’re not statically typed and they don’t provide a convenient expression language for the dynamic values.
-These two problems are in fact two aspects of a same problem: if we write the expression \code{xs.size == 0} in a template, how should it be translated in JavaScript and in Scala?
+\subsubsection{Code hard to reason about}
+
+Callback-driven APIs are very common in JavaScript but their inverted control makes the code flow hard to follow.
+
+\subsubsection{Verbose native APIs}
+
+Native APIs are too much verbose, making the code longer than it should be.
+
+\subsubsection{Limited sharing of code between client and server sides}
+
+The lack of static checks is even harder to tackle in the case of code shared between client and server sides.
+If we write the expression \code{xs.size == 0} in a template, how should it be translated in JavaScript and in Scala?
 Several solutions are valid, depending on the meaning of the expression: in JavaScript we could translate it as \code{xs.size === 0} or \code{xs.length === 0} if \code{xs} refers to a collection (in JavaScript the property to get the size of a collection is named \code{length}).
-So we need to define a proper expression language and to define how it should be translated into JavaScript and Scala, but such a task requires a high effort and is hard to achieve in an extensible way (so users can still
-use their own data types in expressions instead of being restricted to a set of supported types).
-That’s probably for this reason that existing template engines usable on server and client sides have no expression language at all: the dynamic content can only be provided as a map of key-value pairs and no operation can be done on them.
-
-... so we need better ways to handle DOM definition, user interaction events and asynchronous programming.
+We need to define a proper expression language and to define how it should be translated into JavaScript and Scala, but such a task requires a high effort and is hard to achieve in an extensible way (so users can still use their own data types in expressions instead of being restricted to a set of supported types).
 
 \section{DSL design and implementation}
+\label{solution}
 
-This section presents both the design and the implementation of a compiled embedded DSL addressing the problems introduced in the preceding section.
+This section presents both the design and the implementation of a compiled embedded DSL addressing the problems introduced in the previous section.
 
@@
 We propose a DSL that explicitly reflects the asynchronous nature of computations in their type and provides methods turning them into first-class citizen.
-Our DSL is monadic so we can solve the callback-hell problem thanks to the Scala \code{for} notation.
+The DSL is monadic so we can solve the callback-hell problem thanks to the Scala \code{for} notation.
 
@@
 
-The \code{map} and \code{flatMap} methods allow asynchronous values to be transformed and passed around, thus turning them into first-class citizen.
+Asynchronous values can be transformed using the \code{map} and \code{flatMap} methods, turning them into first-class citizen: functions can take as parameters and return asynchronous values.
 
@@
   baz <- Ajax.get(bazUrl(bar))
-} yield println(foo + bar + baz)
+} println(foo + bar + baz)
 \end{lstlisting}
@@
 The \code{for} notation can intuitively be thought of as a sequencing notation: whenever the response of the first Ajax request is available, the next statement will be executed, and so on.
+There is no nested callbacks and the order of execution is directly reflected by the order of statements.
 
-(FIXME Update the following or skip?) The above listing produces the following code (we renamed some identifiers for the sake of readability):
+The following listing shows the generated JavaScript code for an example where two asynchronous computations run in parallel (we renamed some identifiers for the sake of readability):
 
+\begin{multicols}{2}
+\begin{lstlisting}
+val fooAsync = Ajax.get("/foo")
+val barAsync = Ajax.get("/bar")
+for {
+  foo <- fooAsync
+  bar <- barAsync
+} println(foo + bar)
+\end{lstlisting}
+\vfill
+\columnbreak
 \begin{lstlisting}[language=JavaScript]
@@
 AjaxGet("/bar", x2);
-x1.foreach(function (foo) {
-  x2.foreach(function (bar) {
+x1.onComplete(function (foo) {
+ x2.onComplete(function (bar) {
   var x3 = foo + bar;
@@
 \end{lstlisting}
+\end{multicols}
 
@@
 \end{lstlisting}
-
-The only system, which we are aware of, that allows developers to share their HTML templates on both server and client sides in a statically typed way and with a convenient expression language is Opa.
-In Opa they made this possible because HTML templating is part of the Opa language itself.
-If they wanted to support the sharing of another concept such as form field validation rules they would need to introduce this concept as a part of their Opa language.
-
-On the other hand, in our case introducing the support of form field validation as a DSL usable on both server and client sides would be as simple as defining an API and its translation into both Scala and JavaScript.
-We wouldn’t need to change the Scala language itself.
 
@@
 
+(TODO More details about this optimization)
+
 \subsubsection{\emph{Adhoc} polymorphism}
@@
 
+(TODO More general problem description)
+
 JavaScript comes with only one numeric type, \code{Number}, that represents double-precision 64-bit values.
@@
 
-Then we add a \code{+} method to any \code{Rep[_]} value, using an implicit class:
+Then we add a \code{+} method to any \code{Rep[_]} value, using an implicit class, so we can use an homogeneous syntax to write arithmetic operations whatever the numeric type of values this operation is applied on:
 
 \begin{lstlisting}
-implicit class NumericOps(lhs: Rep[A]) {
+implicit class NumericOps[A : Numeric](lhs: Rep[A]) {
   def + (rhs: Rep[A]) = numeric_plus(lhs, rhs)
@@
   println(i + j)
-  println(x + y)
-  // i is promoted to Rep[Double], n has type Rep[Double]
-  val n = i + y
-  // j is promoted to Rep[Double], m has type Rep[Double]
-  val m = x + j
-  println(n + m)
+  // i is promoted to Rep[Double], m has type Rep[Double]
+  val m = i + y
+  // j is promoted to Rep[Double], n has type Rep[Double]
+  val n = x + j
+  println(m + n)
 }
@@
 
-\subsection{Discussion}
+\subsection{Implementation discussion}
+
+We implemented our language as a compiled embedded DSL in Scala.
+The embedding gave us the opportunity to reuse Scala’s infrastructure: we didn’t have to write a parser and, more importantly, we could reuse Scala’s type system.
+Furthermore, staging gave us the opportunity to generate efficient code.
+
+Code generation using LMS is usually described as a two step process: a program first evaluates to an intermediate representation and then the final program’s code is generated from the intermediate representation.
+We think we can consider a third step, occurring before the program evaluation: the compilation.
+
+Indeed, the code generation process consists in three steps giving DSL authors as many opportunities to implement their language features.
+First, features like type coercion and \emph{adhoc} polymorphism only rely on the type system, so they operate during the compilation.
+Then, during the evaluation, the DSL constructs build an intermediate representation of the program, for instance the State monad DSL builds an imperative sequence of assignments on a variable.
+Last, the code generator produces the final program using the domain specific information available in the intermediate representation to generate efficient code, for example the Option monad code generator produces code checking against nullity before dereferencing a value.
+
+Each of these steps takes out some information on the program.
 
@@
 
-\section{Comparison with other approaches}
+(FIXME Simplify this subsubsection)
+
+\subsubsection{Lack of a data type definition DSL}
+
+Defining a language as an embedded DSL gives benefits but on the other side of the coin some features are more difficult to achieve.
+For instance there is no simple way to implement a DSL construct for defining data types.
+
+(TODO Explain briefly why)
+
+\section{Validation and comparison with other approaches}
+\label{validation}
+
+Our DSL aims to give developers a safer and more expressive way to write the client-side code.
+We focused on the main tasks of RIAs development: handling events, asynchronous programming and DOM definition.
+
+\subsection{Rewrite of the application using the DSL?}
 
@@
 Powerful typesafe expression language shared by client-side and server-side.
+
+The only system, which we are aware of, that allows developers to share their HTML templates on both server and client sides in a statically typed way and with a convenient expression language is Opa.
+In Opa they made this possible because HTML templating is part of the Opa language itself.
+If they wanted to support the sharing of another concept such as form field validation rules they would need to introduce this concept as a part of their Opa language.
+
+On the other hand, in our case introducing the support of form field validation as a DSL usable on both server and client sides would be as simple as defining an API and its translation into both Scala and JavaScript.
+We wouldn’t need to change the Scala language itself.
 
@@
 
-jQuery deferred, Q.js. Previous scala-js paper.
+Several JavaScript libraries turn asynchronous values into first-class citizen (\eg{} jQuery.Deferred\footnote{\href{http://api.jquery.com/category/deferred-object/}{http://api.jquery.com/category/deferred-object/}}, Q.js\footnote{\href{https://github.com/kriskowal/q}{https://github.com/kriskowal/q}}).
+They allow developers to register callbacks in order to be notified when the value is available.
+
+Roy solves the callback-hell problem.
 
 \section{Conclusion and perspectives}
+\label{discussion}
 
 We exposed native JavaScript APIs as statically typed APIs while decreasing their verbosity.
+
 More code sharing.
+
+\section{Acknowledgements}
+
+We thank Tiark Rompf for his useful comments.
 

### 2012-12-04 papers/ecoop2013/repairing.tex
Time to go to the party

@@
 
-\title{Statically typed Web programming with JavaScript}
+\title{Statically typed Web programming}
 
@@
 Rich Internet Applications involve more code on the client-side, dealing with DOM manipulation, event handling and asynchronous calls to the server.
-Writing and maintaining large JavaScript code base is challenging because this language has several inadequacies (no static typing, no module system, no \emph{adhoc} polymorphism support, verbose syntax, \etc{}).
+Writing and maintaining large JavaScript code bases is challenging because this language has several inadequacies (no static typing, no module system, verbose syntax, \etc{}).
 We investigate further the definition of JavaScript as a Scala compiled embedded DSL using Lightweight Modular Staging.
@@
 Writing large Web applications is known to be challenging~\cite{Mikkonen08_SpaghettiJs,Preciado05_RIAMethodologyNecessity}, mainly because of the heterogeneous nature of the client-side and server-side environments~\cite{Echeverria09_RIA,Kuuskeri09_PartitioningClientServer}: writing distributed code requires its parts to be consistent together and leads to duplication unless you use the same language in both sides, which, in the case of the Web, means using JavaScript to write the whole application.
-However this language has several inadequacies making it hardly suitable for large code base (\eg{} no static typing, no module system, no support of \emph{adhoc} polymorphism, \etc).
+However this language has several inadequacies making it hardly suitable for large code base (\eg{} no static typing, no module system, verbose syntax, \etc).
 Some other in-browser execution environments give the opportunity to write the client-side code in another language than JavaScript, \eg{} Java applets~\cite{Anuff96_JavaApplets}, Adobe Flash~\cite{Curtis00_Flash} and Microsoft Silverlight\footnote{\href{http://www.microsoft.com/silverlight/}{http://www.microsoft.com/silverlight/}}.
@@
 
-\begin{lstlisting}[language=JavaScript,label=async-js]
+\begin{lstlisting}[language=JavaScript,label=async-js,caption=Callback-driven JavaScript APIs]
 var createVertex = function (text, callback) {
@@
 
-\begin{lstlisting}[language=JavaScript]
+\begin{lstlisting}[language=JavaScript,label=async-callback-hell,caption=Sequential asynchronous calls]
 Ajax.get(fooUrl(), function (foo) {
@@
 Updating the user interface usually means replacing a part of the DOM with another DOM fragment computed from data fetched by an AJAX request.
-This requires to write how to compute the new DOM fragment in JavaScript, but the native API for creating DOM is very verbose, making the code hard to reason about.
+This requires to write how to compute the new DOM fragment in JavaScript.
 For instance the following listing shows how to build a DOM tree representing a vertex in the mind map:
@@
     <rect width=100 height=60 />
-  <text width=100 height=60>Hello, World!</text>
+    <text width=100 height=60>
+        Hello, World!
+    </text>
 </g>
@@
 
-This implementation may be more readable, however the above implementation is wrong: if we call it with a content containing brackets (\eg{} \code{"<foo>"}) they won’t be escaped and will produce a \code{foo} tag nested in the \code{text} tag, which is not the intended behavior.
+The above code may be more readable, however this implementation is wrong: if we call it with a content containing brackets (\eg{} \code{"<foo>"}) they won’t be escaped and will produce a \code{foo} tag nested in the \code{text} tag, which is not the intended behavior.
 So, although this way reads slightly better it’s not less error prone.
 
-By the way, unlike the previous points markup generation is a task that often needs to be performed from both server and client sides.
+By the way, unlike the previous points, markup generation is a task that often needs to be performed from both server and client sides.
 From the server-side it produces HTML pages that search engines can crawl and from the client-side it produces DOM fragments that can be used to update the user interface.
@@
 
-HTML template engines like Mustache\footnote{\href{http://mustache.github.com/}{http://mustache.github.com/}} or Closure Templates\footnote{\href{https://developers.google.com/closure/templates/}{https://developers.google.com/closure/templates/}} aim to make simpler the definition of DOM fragments by providing a convenient syntax to describe the HTML structure and allowing the insertion of dynamic expressions in safe way.
+HTML template engines like Mustache\footnote{\href{http://mustache.github.com/}{http://mustache.github.com/}} or Closure Templates\footnote{\href{https://developers.google.com/closure/templates/}{https://developers.google.com/closure/templates/}} aim to make simpler the definition of DOM fragments by providing a convenient syntax to describe the HTML structure and allowing the insertion of dynamic expressions in a safe way.
 Some template engines can be used on both client-side and server-side, however none has a practical expression language (dynamic content can only be provided as a map of key-value pairs and no operation can be done on values) or is statically typed.
@@
 
-Dynamic typing may give programmers more flexibility but can make it harder for little script to grow into mature and robust code bases and to perform refactorings.
+Dynamic typing may give programmers more flexibility but can make it harder for little scripts to grow into mature and robust code bases and to perform refactorings.
+For instance, in the case of RIAs, a mispell in an event name may break the program behavior without getting a sensible error message.
 
-\subsubsection{Code hard to reason about}
+\subsubsection{Code hard to reason about and to modularize}
 
-Callback-driven APIs are very common in JavaScript but their inverted control makes the code flow hard to follow.
+Callback-driven APIs are very common in JavaScript but their inverted control makes the code flow hard to follow and bothers modularization.
+Furthermore, native APIs are too much verbose, making the code hard to read.
 
-\subsubsection{Verbose native APIs}
-
-Native APIs are too much verbose, making the code longer than it should be.
-
-\subsubsection{Limited sharing of code between client and server sides}
+\subsubsection{Sharing type safe code between clients and servers is hard to achieve}
 
@@
 
-This section presents both the design and the implementation of a compiled embedded DSL addressing the problems introduced in the previous section.
+This section presents both the design and the implementation of a compiled embedded DSL addressing the problems presented in the previous section.
 
@@
 
-The bindings between stages is type-directed: a value of type \code{Rep[Int]} in the first stage will yield a value of type \code{Int} in the second stage. If you consider the following code:
+The bindings between stages are type-directed: a value of type \code{Rep[Int]} in the first stage will yield a value of type \code{Int} in the second stage. If you consider the following code:
 \begin{lstlisting}
@@
 \begin{lstlisting}
-window.onKeyUp { e: KeyboardEvent =>
+window.onKeyUp { e: Rep[KeyboardEvent] =>
   println(e.key)
 }
-window.onMouseWheel { e: MouseWheelEvent =>
+window.onMouseWheel { e: Rep[MouseWheelEvent] =>
   println(e.wheelDeltaY)
@@
 
-The following Scala listing is equivalent to the JavaScript listing \ref{event-js} and is completely typesafe: if the user misspells the event name or tries to use an undefined property on an event his program won’t compile.
+The following Scala listing is equivalent to the JavaScript listing \ref{event-js} and is completely type safe: if the user misspells the event name or tries to use an undefined property on an event his program won’t compile.
 
@@
 \end{lstlisting}
-
-(FIXME Should I show, for the sake of example, the traits defining the user facing DSL, its implementation and the code generator?)
 
@@
 
-And the “callback-hell” example can be re-written as follows:
+The “callback-hell” example (listing \ref{async-callback-hell}) can be re-written as follows:
 
@@
 \begin{multicols}{2}
-\begin{lstlisting}
+\begin{lstlisting}[caption=Parallel computations in Scala]
 val fooAsync = Ajax.get("/foo")
@@
 \columnbreak
-\begin{lstlisting}[language=JavaScript]
+\begin{lstlisting}[language=JavaScript,caption=Generated JavaScript code]
 var x1 = new Promise();
@@
 
-The \code{Promise} constructor code has been omitted, it creates an object holding a list of callbacks to call when the asynchronous value is completed.
+The generated \code{Promise} constructor code has been omitted, it creates an object holding a list of callbacks to call when the asynchronous value is completed.
 The \code{AjaxGet} function code has also been omitted, it creates a \code{XMLHttpRequest} object that sends an HTTP request and completes its promise parameter with the response, when received.
@@
 
-(TODO More on the exploit of sharing typesafe expressions between client and server sides)
-
 Our DSL is equivalent as a template engine with Scala as the expression language.
@@
 \end{lstlisting}
-
-The most popular JavaScript library, jQuery~\cite{Bibeault08_jQuery}, used by more than 40\% of the top million sites\footnote{\href{http://trends.builtwith.com/javascript}{http://trends.builtwith.com/javascript}} handles this problem at runtime by wrapping each query result in a container so before each further method call it tests the emptiness of the container and applies effectively the operation only if the container is not empty.
-Rewriting the above code using jQuery gives the following listing:
-
-\begin{lstlisting}
-var loginWidget = $("div.login");
-var loginButton = loginWidget.find("button.submit");
-loginButton.on("click", function (e) { ... });
-\end{lstlisting}
-
-We can see that it has as few noise as in the listing \ref{null-unsafe} but it has the safety of the listing \ref{null-defensive}.
-However, the checking involves a slight performance penalty since each result is wrapped and then unwrapped for the subsequent method invocation.
 
@@
 Each component of a DSL is defined in a trait and each corresponding code generator for a given target is also defined in a trait.
-The Scala language allows developers to mix several traits together.
-Developers can build their language by picking the DSL traits they want to use.
+The Scala language allows developers to mix several traits together, so they can build their language by picking the DSL traits they want to use.
 
 Furthermore, a code generator can be tweaked by defining a sub-trait specializing some of its methods.
-For example, the \code{forEach} method on JavaScripts arrays was not supported by Internet Explorer before version 9.
+This is useful to handle cross-browser incompatibilities.
+For instance we have written code generators specializing the output for some statement nodes in order to handle Internet Explorer incompatibilities.
+
+% For example, the \code{forEach} method on JavaScripts arrays was not supported by Internet Explorer before version 9.
 As some users may still have old versions of Internet Explorer we want to send them a specialized version of JavaScript files to handle this incompatibility.
 We can do that by extending the code generator for the \code{List} operations to specialize the generation of the \code{ListForeach} node to generate a JavaScript \code{for} loop instead of a \code{forEeach} call:
-
-\begin{lstlisting}
-trait JSGenListIE extends JSGenList {
- import IR._
- override def emitNode(sym: Sym[Any], rhs: Def[Any]) =
-  rhs match {
-   case ListForeach(l, x, b) =>
-    var i = fresh[Int]
-    stream.println("for (var "+quote(i)+" = 0 ; "+
-                        quote(i)+" < "+quote(l)+".length ; "+
-                        quote(i)+"++) {")
-    emitValDef(x, quote(l)+"["+quote(i)+"]")
-    emitBlock(b)
-    stream.println("}")
-   case _ => super.emitNode(sym, rhs)
-  }
-}
-\end{lstlisting}
-
-Internet Explorer suffers from some others incompatibilities, \eg{} its events API is different of the standard.
+% 
+% \begin{lstlisting}
+% trait JSGenListIE extends JSGenList {
+%  import IR._
+%  override def emitNode(sym: Sym[Any], rhs: Def[Any]) =
+%   rhs match {
+%    case ListForeach(l, x, b) =>
+%     var i = fresh[Int]
+%     stream.println("for (var "+quote(i)+" = 0 ; "+
+%                         quote(i)+" < "+quote(l)+".length ; "+
+%                         quote(i)+"++) {")
+%     emitValDef(x, quote(l)+"["+quote(i)+"]")
+%     emitBlock(b)
+%     stream.println("}")
+%    case _ => super.emitNode(sym, rhs)
+%   }
+% }
+% \end{lstlisting}
+% 
+% Internet Explorer suffers from some others incompatibilities, \eg{} its events API is different of the standard.
 On the same way we can define a \code{JSGenDomIE} trait generating code targeting the Internet Explorer’s events API for the \code{Dom} DSL.
-
-Then, we can pack together all the traits handling Internet Explorer’s incompatibilities in a \code{IESupport} trait:
-
-\begin{lstlisting}
-trait IESupport extends JSGenListIE with JSGenDomIE {
-  val IR: ListOpsExp with DomExp
-}
-\end{lstlisting}
-
-And, finally, we can mix this \code{IESupport} trait with any JavaScript code generator to target Internet Explorer:
-
-\begin{lstlisting}
-val jsGen = new JSGen with IESupport {
-  val IR: prog.type = prog
-}
-\end{lstlisting}
-
-(FIXME Simplify this subsubsection)
-
-\subsubsection{Lack of a data type definition DSL}
-
-Defining a language as an embedded DSL gives benefits but on the other side of the coin some features are more difficult to achieve.
-For instance there is no simple way to implement a DSL construct for defining data types.
-
-(TODO Explain briefly why)
+% 
+% Then, we can pack together all the traits handling Internet Explorer’s incompatibilities in a \code{IESupport} trait:
+% 
+% \begin{lstlisting}
+% trait IESupport extends JSGenListIE with JSGenDomIE {
+%   val IR: ListOpsExp with DomExp
+% }
+% \end{lstlisting}
+% 
+% And, finally, we can mix this \code{IESupport} trait with any JavaScript code generator to target Internet Explorer:
+% 
+% \begin{lstlisting}
+% val jsGen = new JSGen with IESupport {
+%   val IR: prog.type = prog
+% }
+% \end{lstlisting}
 
@@
 
-\subsection{Rewrite of the application using the DSL?}
+\subsection{Observed benefits of the use of our DSL}
 
-\subsection{DOM definition}
+Using our DSL, we were able to divide by two the length of some parts of code (DOM definition and \code{null} reference handling).
+The size of code parts dealing with events did not change but they gain type safety.
+Finally, code handling asynchronous computations were easier to modularize.
 
-Powerful typesafe expression language shared by client-side and server-side.
+All these benefits came with no runtime overhead, excepted for the asynchronous programming DSL that creates, for each asynchronous computation, a \code{Promise} value registering the callbacks to call when the value is be available.
 
+\subsection{JavaScript libraries}
+
+The most popular JavaScript library, jQuery~\cite{Bibeault08_jQuery}, used by more than 40\% of the top million sites\footnote{\href{http://trends.builtwith.com/javascript}{http://trends.builtwith.com/javascript}}, and Q.js\footnote{\href{https://github.com/kriskowal/q}{https://github.com/kriskowal/q}} provide an API turning asynchronous computations into first-class citizen.
+However, the callback-hell problem can’t always be avoided due to the absence of a special syntactic sugar for continuation passing style.
+
+jQuery handles the \code{null} references problem by wrapping each query result in a container so before each further method call it tests the emptiness of the container and applies effectively the operation only if the container is not empty.
+It provides an expressive API but the emptiness checking involves a slight performance penalty since each result is wrapped and then unwrapped for the subsequent method invocation.
+
+\subsection{GWT, Roy, Kotlin}
+
+GWT also has a type safe API to handle events.
+
+Roy solves the callback-hell problem.
+
+\subsection{Template engines}
+
+Templates written using our HTML templating DSL are about two times shorter than templates written using the native JavaScript API.
+Compared to usual template engines our solution has no runtime overhead, does not require a separated compilation step and provides a flexible type safe expression language.
+
+Furthermore, templates can be shared between client and server sides.
 The only system, which we are aware of, that allows developers to share their HTML templates on both server and client sides in a statically typed way and with a convenient expression language is Opa.
@@
 
-\subsection{Events}
+(TODO Make this result more prominent)
 
-Type safety. GWT?
-
-\subsection{Asynchronous programming}
-
-Several JavaScript libraries turn asynchronous values into first-class citizen (\eg{} jQuery.Deferred\footnote{\href{http://api.jquery.com/category/deferred-object/}{http://api.jquery.com/category/deferred-object/}}, Q.js\footnote{\href{https://github.com/kriskowal/q}{https://github.com/kriskowal/q}}).
-They allow developers to register callbacks in order to be notified when the value is available.
-
-Roy solves the callback-hell problem.
+% \subsection{Lack of a data type definition DSL}
+% 
+% Defining a language as an embedded DSL gives benefits but on the other side of the coin some features are more difficult to achieve.
+For instance there is no simple way to implement a DSL construct for defining data types.
+This feature comes almost for free with approaches like GWT, Fay or Kotlin because they derive the client-side program and the server-side program from the initial program source.
+In our case, the client-side program and the server-side program are derived from the intermediate representation produced by the evaluation of the initial program.
+So, only concepts reified in an intermediate representation can be part of the final programs.
+For instance, it means defining a type \code{A} whose constructor yields a value of type \code{Rep[A]}.
+Scala-virtualized~\cite{Moors12_Virtualized} provides such a mechanism but is limited to product types and provides only projection functions.
+% 
+% In summary, the compiled embedded approach may require more work to support custom data types but gives control on the way this data type will be encoded for each target platform (\eg{} the JavaScript code generator uses native arrays to encode the lists of the list manipulation DSL).
+% 
+% However, the next release of Scala will have support for macros making it possible to generate reified data types from a concise syntax without relying on a compiler extension.
 
@@
 
-We exposed native JavaScript APIs as statically typed APIs while decreasing their verbosity.
+Writing RIAs requires large JavaScript code bases, however this language has several inadequacies for this purpose.
+
+We provide a safer and more expressive language than JavaScript and its native APIs to write RIAs.
 
@@
 
-We thank Tiark Rompf for his useful comments.
+We thank Tiark Rompf for his precious feedback.
 

### 2012-12-06 papers/ecoop2013/repairing.tex
Fin de journée

@@
 This paper continues their work and presents safe and expressive DSL constructs for RIAs development.
+We increase the expressivity of JavaScript native APIs with no runtime overhead and expose them through statically typed APIs.
+We make asynchronous programming simpler by raising asynchronous computations to first-class citizen.
+We leverage the Scala’s type system to bring type-directed language features such as \emph{adhoc} polymorphism and type coercion to the client-side.
 
@@
 
-To illustrate the challenges of writing a RIA, we introduce a simple “mind mapping” application and show how the client-side code would be typically written in pure JavaScript.
+To illustrate the challenges of writing a RIA and the need for a DSL for writing them, we introduce a simple “mind mapping” application and show how the client-side code would be typically written in pure JavaScript.
 
@@
 
-Notice that the code is getting deeper toward the right, this problem is often referred to as the “callback-hell”.
+Notice that the code is getting deeper toward the right, this problem is often referred to as the “callback-hell”~\cite{McKenna_Roy}.
 This code is also hard to reason about: when will the second \code{console.log} statement be executed?
@@
 Several solutions are valid, depending on the meaning of the expression: in JavaScript we could translate it as \code{xs.size === 0} or \code{xs.length === 0} if \code{xs} refers to a collection (in JavaScript the property to get the size of a collection is named \code{length}).
-We need to define a proper expression language and to define how it should be translated into JavaScript and Scala, but such a task requires a high effort and is hard to achieve in an extensible way (so users can still use their own data types in expressions instead of being restricted to a set of supported types).
+We need to define a proper expression language and to define how it should be translated for the client-side environment (JavaScript) and the server-side environment (\eg{} the Java Virtual Machine), but such a task requires a high effort and is hard to achieve in an extensible way (so users can still use their own data types in expressions instead of being restricted to a set of supported types).
 
@@
 
-This section presents both the design and the implementation of a compiled embedded DSL addressing the problems presented in the previous section.
+JavaScript libraries could address some of the challenges presented above, but addressing type safety and asynchronous programming issues require language modifications (\eg{} to bring a type system and a sequencing notation).
+Defining a language requires a high effort, especially to build development tools for the language so we chose to define our language as a compiled embedded DSL is Scala\footnote{The code is available at \href{http://github.com/js-scala}{http://github.com/js-scala}}.
 
@@
 
-We chose to implement our DSL as a compiled embedded DSL in Scala using Lightweight Modular Staging.
+We use the Lightweight Modular Staging framework (LMS) to define our compiled embedded DSL.
+The main idea is that a program written using a DSL is evaluated in two stages (or steps): the first stage builds an intermediate representation of the program and the second stage transforms this intermediate representation into executable code (figure \ref{lms-diagram}).
 The embedding lets us reuse the host language infrastructure (compiler, development tools) and type system, and staging gives the opportunity to generate efficient code from expressive code by applying domain specific optimizations.
-
-Lightweight Modular Staging (LMS) is a Scala framework for defining compiled embedded DSLs.
-The main idea is that a program written using a DSL is evaluated in two stages (or steps): the first stage builds an intermediate representation of the program and the second stage transforms this intermediate representation into executable code (figure \ref{lms-diagram}).
 
@@
   \includegraphics[width=7cm]{lms.pdf}
-  \caption{Compilation of a program using LMS}
+  \caption{Compilation of a program using LMS.
+An initial Scala program using embedded DSLs (on the left) evaluates to an intermediate representation from which the final program’s code is generated (on the right).}
   \label{lms-diagram}
@@
 
-We propose a DSL that explicitly reflects the asynchronous nature of computations in their type and provides methods turning them into first-class citizen.
-The DSL is monadic so we can solve the callback-hell problem thanks to the Scala \code{for} notation.
-
-An asynchronous value of type \code{Rep[A]} is modelled by a value of type \code{Rep[Future[A]]}. This type has the following methods:
-\begin{itemize}
-\item \code{foreach(f: Rep[A] => Rep[Unit]): Rep[Unit]}, registering a callback to eventually do something with the value when available~;
-\item \code{map(f: Rep[A] => Rep[B]): Rep[Future[B]]}, to eventually transform the value when available~;
-\item \code{flatMap(f: Rep[A] => Rep[Future[B]]): Rep[Future[B]]} to eventually transform the value when available~;
-\end{itemize}
-
-Asynchronous values can be transformed using the \code{map} and \code{flatMap} methods, turning them into first-class citizen: functions can take as parameters and return asynchronous values.
-
-The listing \ref{async-js} can be re-written as follows with our DSL:
-
-\begin{lstlisting}
+\begin{figure}
+\centering
+\begin{lstlisting}[numbers=left,caption=Asynchronous values are first class citizen,label=async-first-class]
 def createVertex(text: Rep[String]): Rep[Future[Vertex]] =
@@
 \end{lstlisting}
+\end{figure}
 
+We propose a DSL that explicitly reflects the asynchronous nature of computations in their type and provides methods turning them into first-class citizen.
+The DSL is monadic so we can solve the callback-hell problem thanks to the Scala \code{for} notation.
+For instance, listing \ref{async-first-class} shows how the listing \ref{async-js} can be re-written with our DSL.
 The \code{createVertex} now returns an asynchronous value instead of taking a callback as parameter.
+At the usage site, the retrieval of the vertex (line 5) is separated from its insertion in the user interface (lines 6-8).
 
-The “callback-hell” example (listing \ref{async-callback-hell}) can be re-written as follows:
-
-\begin{lstlisting}
+\begin{figure}
+\centering
+\begin{lstlisting}[caption=No callback hell,label=async-no-callback]
 for {
@@
 \end{lstlisting}
+\end{figure}
 
+The listing \ref{async-no-callback} translates the “callback-hell” example (listing \ref{async-callback-hell}) using our DSL.
 The \code{for} notation can intuitively be thought of as a sequencing notation: whenever the response of the first Ajax request is available, the next statement will be executed, and so on.
@@
 
-The following listing shows the generated JavaScript code for an example where two asynchronous computations run in parallel (we renamed some identifiers for the sake of readability):
+An asynchronous value of type \code{Rep[A]} is modelled by a value of type \code{Rep[Future[A]]}.
+Because Scala’s \code{for} notation is just a syntactic sugar for methods named \code{foreach}, \code{map} and \code{flatMap}, we only need to define these methods on \code{Rep[Future[A]]} values, with the following semantic:
+\begin{itemize}
+\item \code{foreach(f: Rep[A] => Rep[Unit]): Rep[Unit]}, eventually does something with the value when available~;
+\item \code{map(f: Rep[A] => Rep[B]): Rep[Future[B]]}, eventually transforms the value when available~;
+\item \code{flatMap(f: Rep[A] => Rep[Future[B]]): Rep[Future[B]]} eventually transforms the value when available~;
+\end{itemize}
 
+Asynchronous values can be transformed using the \code{map} and \code{flatMap} methods, turning them into first-class citizen: functions can take as parameters and return asynchronous values.
+
+\begin{figure}
 \begin{multicols}{2}
-\begin{lstlisting}[caption=Parallel computations in Scala]
+\begin{lstlisting}[caption=Parallel computations in Scala,label=async-parallel-1]
 val fooAsync = Ajax.get("/foo")
@@
 \columnbreak
-\begin{lstlisting}[language=JavaScript,caption=Generated JavaScript code]
+\begin{lstlisting}[language=JavaScript,caption=Generated JavaScript code,label=async-parallel-2]
 var x1 = new Promise();
@@
 \end{multicols}
+\end{figure}
 
+Listing \ref{async-parallel-1} and \ref{async-parallel-2} define two asynchronous computations running in parallel and the generated JavaScript code (we renamed some identifiers for the sake of readability).
 The generated \code{Promise} constructor code has been omitted, it creates an object holding a list of callbacks to call when the asynchronous value is completed.
@@
 
-Furthermore, our DSL implementation produces an intermediate representation semantically equivalent to the monadic sequences but using regular variables, so that there is no abstraction penalty: the listing \ref{state-monad} compiles to code equivalent to the listing \ref{state-fragile}.
-
-(TODO More details about this optimization)
+Furthermore, our implementation produces efficient imperative code using variables instead of functions passing the state through.
+To achieve that, we transform the intermediate representation of the stack of state modifications produced by the \code{for} sequences into an equivalent intermediate representation using regular variables.
+Finally, the listing \ref{state-monad} compiles to code equivalent to the listing \ref{state-fragile} so that there is no abstraction penalty.
 
@@
 
-\subsection{GWT, Roy, Kotlin}
+\subsection{Languages usable on both server-side and client-side}
 
@@
 Furthermore, templates can be shared between client and server sides.
-The only system, which we are aware of, that allows developers to share their HTML templates on both server and client sides in a statically typed way and with a convenient expression language is Opa.
-In Opa they made this possible because HTML templating is part of the Opa language itself.
-If they wanted to support the sharing of another concept such as form field validation rules they would need to introduce this concept as a part of their Opa language.
+Opa is also a language making it possible to share statically typed templates with a convenient expression language between client and server sides.
+In Opa they made this possible because HTML templating is part of the language itself.
+However, creating a brand new language for each problem requires a high effort, especially since the effort to define common language constructs is duplicated.
+To reduce the effort, the solution consists in implementing abstractions specific to a given problem as libraries on top of a general purpose language.
+For instance, Kotlin provides an embedded DSL similar to ours to define HTML fragments.
+Because the Kotlin language can be compiled to JavaScript and to JVM bytecode, HTML templates can be shared between client and server sides.
+However, their DSL does not produce code building DOM fragments on client-side.
+Actually, on both server-side and client-side the output is the same: an abstract tree containing the HTML fragment.
+So, users need to transform this AST into a proper DOM fragment using a function specific to the client-side.
+Kotlin and similar approaches (Fay, ClojureScript, GWT) can’t generate specialized code for a given abstraction because their translation scheme is based on the initial program source code: if an abstraction is defined as a library, they will generate code reimplementing the library on each backend (client-side and server-side).
+There is no way to specialize the generated code for a given library.
 
-On the other hand, in our case introducing the support of form field validation as a DSL usable on both server and client sides would be as simple as defining an API and its translation into both Scala and JavaScript.
-We wouldn’t need to change the Scala language itself.
+On one hand defining a new language (or modifying an existing language) to integrate a concept specific to a problem is not desired either because it requires a high-effort to define a full-featured language or because several small languages are hard to integrate seamlessly together.
+On the other hand, defining a concept specific to a problem as a library allows developers to concentrate their effort on the concept itself and to reuse the host language infrastructure, but they have no way to specialize, for each backend, the code generated by the compiler for their concept.
 
-(TODO Make this result more prominent)
+Compiled embedded DSLs provide the best of both worlds because DSLs are defined as libraries and, because they are compiled, developers can specialize the generated code, for each backend, for their DSL concepts.
+This opportunity made it possible to define, as a library, the HTML templating DSL generating code building DOM fragments on client-side and code building XML trees on server-side.
 
-% \subsection{Lack of a data type definition DSL}
-% 
-% Defining a language as an embedded DSL gives benefits but on the other side of the coin some features are more difficult to achieve.
-For instance there is no simple way to implement a DSL construct for defining data types.
-This feature comes almost for free with approaches like GWT, Fay or Kotlin because they derive the client-side program and the server-side program from the initial program source.
+\subsection{Lack of a data type definition DSL}
+
+On the other side of the coin, because that is not possible to define a library without defining how to translate it on both server-side and client-side, there is no simple way to define a library for defining libraries.
+This feature comes almost for free with approaches like GWT, Fay or Kotlin because they derive the client-side program and the server-side program from the initial program source (that contains data type definitions such as classes definitions).
 In our case, the client-side program and the server-side program are derived from the intermediate representation produced by the evaluation of the initial program.
-So, only concepts reified in an intermediate representation can be part of the final programs.
-For instance, it means defining a type \code{A} whose constructor yields a value of type \code{Rep[A]}.
-Scala-virtualized~\cite{Moors12_Virtualized} provides such a mechanism but is limited to product types and provides only projection functions.
-% 
-% In summary, the compiled embedded approach may require more work to support custom data types but gives control on the way this data type will be encoded for each target platform (\eg{} the JavaScript code generator uses native arrays to encode the lists of the list manipulation DSL).
-% 
-% However, the next release of Scala will have support for macros making it possible to generate reified data types from a concise syntax without relying on a compiler extension.
+So, only concepts reified in an intermediate representation during this evaluation can be part of the final programs.
+However regular Scala’s classes definition are not reified in an intermediate representation (it would mean that the constructor of a type \code{A} would yield a value of type \code{Rep[A]}).
+Scala-virtualized~\cite{Moors12_Virtualized} provides such a mechanism but is limited to product types and provides only projection functions, so for more complex cases users have to write some repetitive and boilerplate code.
+
+In summary, the compiled embedded approach may require more work to support custom data types but gives control on the way this data type will be encoded for each target platform (\eg{} the JavaScript code generator uses native arrays to encode the lists of the list manipulation DSL).
+
+However, the next release of Scala will have support for macros making it possible to generate reified data types from a concise syntax without relying on a compiler extension.
 

### 2012-12-07 papers/ecoop2013/repairing.tex
Word wrap

@@
 \begin{abstract}
-Rich Internet Applications involve more code on the client-side, dealing with DOM manipulation, event handling and asynchronous calls to the server.
-Writing and maintaining large JavaScript code bases is challenging because this language has several inadequacies (no static typing, no module system, verbose syntax, \etc{}).
-We investigate further the definition of JavaScript as a Scala compiled embedded DSL using Lightweight Modular Staging.
-We define safe and expressive abstractions for DOM manipulation, event handling and asynchronous programming that translate to efficient JavaScript code using native APIs.
-We use both type-level information and staging to introduce high-level abstractions with a few or no overhead on the generated JavaScript code.
+Rich Internet Applications involve more code on the client-side, dealing with DOM manipulation, event handling and
+asynchronous calls to the server. Writing and maintaining large JavaScript code bases is challenging because this
+language has several inadequacies (no static typing, no module system, verbose syntax, \etc{}). We investigate
+further the definition of JavaScript as a Scala compiled embedded DSL using Lightweight Modular Staging. We define
+safe and expressive abstractions for DOM manipulation, event handling and asynchronous programming that translate to
+efficient JavaScript code using native APIs. We use both type-level information and staging to introduce high-level
+abstractions with a few or no overhead on the generated JavaScript code.
 \end{abstract}
@@
 
-The Web is an appealing platform on which to write applications (\aka{} Rich Internet Applications, RIA) because it makes them easy to deploy on clients and allows large scale innovative collaborative experiences~\cite{Farrell07_RIA,Mikkonen08_SpaghettiJs}.
-RIAs are characterized by partial updates of the user interface (as opposed to refreshing the whole page with the classic hyperlink navigation), so a large part of the client-side code listens to user interface events (such as mouse clicks), triggers the appropriate action asynchronously on the server using AJAX~\cite{Garrett05_Ajax} and updates the application’s state and the DOM content with a new DOM fragment computed from the response data~\cite{Farrell07_RIA,Busch09_StateOfArtRIA}.
-
-Writing large Web applications is known to be challenging~\cite{Mikkonen08_SpaghettiJs,Preciado05_RIAMethodologyNecessity}, mainly because of the heterogeneous nature of the client-side and server-side environments~\cite{Echeverria09_RIA,Kuuskeri09_PartitioningClientServer}: writing distributed code requires its parts to be consistent together and leads to duplication unless you use the same language in both sides, which, in the case of the Web, means using JavaScript to write the whole application.
-However this language has several inadequacies making it hardly suitable for large code base (\eg{} no static typing, no module system, verbose syntax, \etc).
-Some other in-browser execution environments give the opportunity to write the client-side code in another language than JavaScript, \eg{} Java applets~\cite{Anuff96_JavaApplets}, Adobe Flash~\cite{Curtis00_Flash} and Microsoft Silverlight\footnote{\href{http://www.microsoft.com/silverlight/}{http://www.microsoft.com/silverlight/}}.
-However these technologies have several drawbacks: they require an additional browser plugin to be installed (which may not be available on all devices having a Web browser: for instance there is no way to execute Flash objects within an Apple smart phone), the page content can't naturally be referenced by search engines, and the content is not structured in URLs that users can bookmark or share.
-
-An increasing number of initiatives attempt to allow developers to write the client-side code in a language different of JavaScript but that can be compiled to JavaScript (\eg{} GWT~\cite{Chaganti07_GWT}, Dart~\cite{Griffith11_Dart}, TypeScript\footnote{\href{http://www.typescriptlang.org/}{http://www.typescriptlang.org/}}, Roy~\cite{McKenna_Roy}).
-Some of these languages can also be compiled to another runtime environment usable on server-side (\eg{} GWT, Dart, Kotlin\footnote{\href{http://kotlin.jetbrains.org/}{http://kotlin.jetbrains.org/}}, ClojureScript~\cite{McGranaghan11_ClojureScript}, Fay\footnote{\href{http://fay-lang.org/}{http://fay-lang.org/}}, Haxe~\cite{Cannasse08_HaXe} or Opa\footnote{\href{http://opalang.org/}{http://opalang.org/}}).
-These languages are usually more suitable to write large applications either because they provide more constructs to build abstractions (such as object orientation), or because they are statically typed, or because they add a bunch of useful concepts that are
-missing in JavaScript (such as \emph{adhoc} polymorphism or namespaces).
-Moreover, their ability to compile for both server and client sides usually allows developers to share some parts of code between server and client sides for free.
-However this shared code is restricted to use exclusively language constructs: concepts provided by external APIs can't be shared between server and client sides because the bindings with these environments are different on the server and client sides.
-
-In this paper we investigate further a path already introduced by Kossakowski \etal{}: defining JavaScript as a compiled embedded DSL in Scala~\cite{Kossakowski12_JsDESL}.
-This approach is based on Lightweight Modular Staging~\cite{Rompf12_LMSThesis}: a Scala program written using the embedded DSL evaluates to an intermediate representation that can be further processed to perform domain specific optimizations and then to generate a JavaScript program (for the client-side) and a Scala program (for the server-side).
-This approach has two main advantages: (1) the embedding reduces the effort needed to define the language since we can reuse the Scala infrastructure and tooling, and (2) staging gives more knowledge to the compiler about how a given abstraction should be efficiently translated into the target environment.
-In other words any library-level abstraction can have the efficiency of a language-level abstraction.
-
-The previous work showed how to write client-side code in Scala using the JavaScript embedded DSL and how to share code between server-side and client-side.
-The authors also showed that the code was safer and more convenient to write thanks to Scala's static typing and class system.
-Finally, they showed how to \emph{un-invert} the control of callback-based asynchronous APIs by using continuations but they did not address other concerns of RIAs development such as user interface events handling, DOM creation and manipulation, and application’s state management.
-
-This paper continues their work and presents safe and expressive DSL constructs for RIAs development.
-We increase the expressivity of JavaScript native APIs with no runtime overhead and expose them through statically typed APIs.
-We make asynchronous programming simpler by raising asynchronous computations to first-class citizen.
-We leverage the Scala’s type system to bring type-directed language features such as \emph{adhoc} polymorphism and type coercion to the client-side.
+The Web is an appealing platform on which to write applications (\aka{} Rich Internet Applications, RIA) because it
+makes them easy to deploy on clients and allows large scale innovative collaborative
+experiences~\cite{Farrell07_RIA,Mikkonen08_SpaghettiJs}. RIAs are characterized by partial updates of the user
+interface (as opposed to refreshing the whole page with the classic hyperlink navigation), so a large part of the
+client-side code listens to user interface events (such as mouse clicks), triggers the appropriate action
+asynchronously on the server using AJAX~\cite{Garrett05_Ajax} and updates the application’s state and the DOM content
+with a new DOM fragment computed from the response data~\cite{Farrell07_RIA,Busch09_StateOfArtRIA}.
+
+Writing large Web applications is known to be
+challenging~\cite{Mikkonen08_SpaghettiJs,Preciado05_RIAMethodologyNecessity}, mainly because of the heterogeneous
+nature of the client-side and server-side environments~\cite{Echeverria09_RIA,Kuuskeri09_PartitioningClientServer}:
+writing distributed code requires its parts to be consistent together and leads to duplication unless you use the
+same language in both sides, which, in the case of the Web, means using JavaScript to write the whole application.
+However this language has several inadequacies making it hardly suitable for large code base (\eg{} no static typing,
+no module system, verbose syntax, \etc). Some other in-browser execution environments give the opportunity to write
+the client-side code in another language than JavaScript, \eg{} Java applets~\cite{Anuff96_JavaApplets}, Adobe
+Flash~\cite{Curtis00_Flash} and Microsoft
+Silverlight\footnote{\href{http://www.microsoft.com/silverlight/}{http://www.microsoft.com/silverlight/}}. However
+these technologies have several drawbacks: they require an additional browser plugin to be installed (which may not
+be available on all devices having a Web browser: for instance there is no way to execute Flash objects within an
+Apple smart phone), the page content can't naturally be referenced by search engines, and the content is not
+structured in URLs that users can bookmark or share.
+
+An increasing number of initiatives attempt to allow developers to write the client-side code in a language different
+of JavaScript but that can be compiled to JavaScript (\eg{} GWT~\cite{Chaganti07_GWT}, Dart~\cite{Griffith11_Dart},
+TypeScript\footnote{\href{http://www.typescriptlang.org/}{http://www.typescriptlang.org/}}, Roy~\cite{McKenna_Roy}).
+Some of these languages can also be compiled to another runtime environment usable on server-side (\eg{} GWT, Dart,
+Kotlin\footnote{\href{http://kotlin.jetbrains.org/}{http://kotlin.jetbrains.org/}},
+ClojureScript~\cite{McGranaghan11_ClojureScript}, Fay\footnote{\href{http://fay-lang.org/}{http://fay-lang.org/}},
+Haxe~\cite{Cannasse08_HaXe} or Opa\footnote{\href{http://opalang.org/}{http://opalang.org/}}). These languages are
+usually more suitable to write large applications either because they provide more constructs to build abstractions
+(such as object orientation), or because they are statically typed, or because they add a bunch of useful concepts
+that are missing in JavaScript (such as \emph{adhoc} polymorphism or namespaces). Moreover, their ability to compile
+for both server and client sides usually allows developers to share some parts of code between server and client
+sides for free. However this shared code is restricted to use exclusively language constructs: concepts provided by
+external APIs can't be shared between server and client sides because the bindings with these environments are
+different on the server and client sides.
+
+In this paper we investigate further a path already introduced by Kossakowski \etal{}: defining JavaScript as a
+compiled embedded DSL in Scala~\cite{Kossakowski12_JsDESL}. This approach is based on Lightweight Modular
+Staging~\cite{Rompf12_LMSThesis}: a Scala program written using the embedded DSL evaluates to an intermediate
+representation that can be further processed to perform domain specific optimizations and then to generate a
+JavaScript program (for the client-side) and a Scala program (for the server-side). This approach has two main
+advantages: (1) the embedding reduces the effort needed to define the language since we can reuse the Scala
+infrastructure and tooling, and (2) staging gives more knowledge to the compiler about how a given abstraction
+should be efficiently translated into the target environment. In other words any library-level abstraction can have
+the efficiency of a language-level abstraction.
+
+The previous work showed how to write client-side code in Scala using the JavaScript embedded DSL and how to share
+code between server-side and client-side. The authors also showed that the code was safer and more convenient to
+write thanks to Scala's static typing and class system. Finally, they showed how to \emph{un-invert} the control of
+callback-based asynchronous APIs by using continuations but they did not address other concerns of RIAs development
+such as user interface events handling, DOM creation and manipulation, and application’s state management.
+
+This paper continues their work and presents safe and expressive DSL constructs for RIAs development. We increase the
+expressivity of JavaScript native APIs with no runtime overhead and expose them through statically typed APIs. We
+make asynchronous programming simpler by raising asynchronous computations to first-class citizen. We leverage the
+Scala’s type system to bring type-directed language features such as \emph{adhoc} polymorphism and type coercion to
+the client-side.
 
@@
 
-The next section introduces a simple application example showing challenges of RIA development, section \ref{solution} presents our DSL tackling these challenges, section \ref{validation} validates our solution and compares it to related work and section \ref{discussion} concludes.
+The next section introduces a simple application example showing challenges of RIA development, section
+\ref{solution} presents our DSL tackling these challenges, section \ref{validation} validates our solution and
+compares it to related work and section \ref{discussion} concludes.
 
@@
 
-To illustrate the challenges of writing a RIA and the need for a DSL for writing them, we introduce a simple “mind mapping” application and show how the client-side code would be typically written in pure JavaScript.
+To illustrate the challenges of writing a RIA and the need for a DSL for writing them, we introduce a simple “mind
+mapping” application and show how the client-side code would be typically written in pure JavaScript.
 
@@
 
-In our application, a mind map is represented as a graph where ideas or concepts are represented by vertices, and the relation between two ideas is represented by an edge.
-Users can visualize a mind map, zoom in and out and move it.
-They can also add ideas and link them together.
-The figure \ref{mindmap} shows the visualization of a simple map consisting of three ideas linked together, using our application.
+In our application, a mind map is represented as a graph where ideas or concepts are represented by vertices, and the
+relation between two ideas is represented by an edge. Users can visualize a mind map, zoom in and out and move it.
+They can also add ideas and link them together. The figure \ref{mindmap} shows the visualization of a simple map
+consisting of three ideas linked together, using our application.
 
@@
 
-RIAs distinguish from classic Web pages by a higher degree of interactivity.
-In our example we want to let users zoom in and out a map using their mouse wheel.
-To implement this feature we need to attach an event handler for the \code{mousewheel} event on the DOM element containing the map.
-The following listing shows how to attach such an event handler using the native JavaScript API:
+RIAs distinguish from classic Web pages by a higher degree of interactivity. In our example we want to let users zoom
+in and out a map using their mouse wheel. To implement this feature we need to attach an event handler for the
+\code{mousewheel} event on the DOM element containing the map. The following listing shows how to attach such an
+event handler using the native JavaScript API:
 
@@
 
-The event handler is simply a function taking the event data as a parameter. In the above example we use the \code{wheelDeltaY} property of the event to change the scale of the map.
-
-We think this code is fragile for two reasons.
-(1) The name of the event is passed as a string, so it is easy to mispell it.
-(2) The callback passed as a second parameter takes a parameter \code{e} whose fields vary according to the listened event but developers have no way to check that the fields they’re using are indeed defined on the event they’re listening to.
-For instance, in our case we use the \code{wheelDeltaY} event field that is defined only on the \code{mousewheel} event.
+The event handler is simply a function taking the event data as a parameter. In the above example we use the
+\code{wheelDeltaY} property of the event to change the scale of the map.
+
+We think this code is fragile for two reasons. (1) The name of the event is passed as a string, so it is easy to
+mispell it. (2) The callback passed as a second parameter takes a parameter \code{e} whose fields vary according to
+the listened event but developers have no way to check that the fields they’re using are indeed defined on the event
+they’re listening to. For instance, in our case we use the \code{wheelDeltaY} event field that is defined only on the
+\code{mousewheel} event.
 
@@
 
-Another characteristic of RIAs is that requests to the server are often performed asynchronously: instead of asking the browser to reload the whole page, the JavaScript code has to perform a request and to process the response, when available, to update a part of the user interface.
-The following listing shows how to send a request creating a vertex on the server and to insert it to the user interface:
+Another characteristic of RIAs is that requests to the server are often performed asynchronously: instead of asking
+the browser to reload the whole page, the JavaScript code has to perform a request and to process the response, when
+available, to update a part of the user interface. The following listing shows how to send a request creating a
+vertex on the server and to insert it to the user interface:
 
@@
 
-\code{Ajax.post} is a helper function that sends a HTTP request to the server and calls its last parameter (that is a \emph{callback}) when the response data is available.
-The inverted control makes modularization harder to achieve: in the above listing the function creating the vertex is also responsible to updating the user interface.
+\code{Ajax.post} is a helper function that sends a HTTP request to the server and calls its last parameter
+(that is a \emph{callback}) when the response data is available. The inverted control makes modularization harder to
+achieve: in the above listing the function creating the vertex is also responsible to updating the user interface.
 If we want to relax this coupling the only option is to add a callback parameter to the \code{createVertex} function:
@@
 
-However increasing the number of callbacks makes the code flow harder to follow and adds distance to the programmer's initial intent.
-
-Another issue with callback-driven programming arises when several dependent asynchronous computations are executed sequentially.
-Consider for example the following listing performing three consecutive Ajax requests:
+However increasing the number of callbacks makes the code flow harder to follow and adds distance to the programmer's
+initial intent.
+
+Another issue with callback-driven programming arises when several dependent asynchronous computations are executed
+sequentially. Consider for example the following listing performing three consecutive Ajax requests:
 
@@
 
-Notice that the code is getting deeper toward the right, this problem is often referred to as the “callback-hell”~\cite{McKenna_Roy}.
-This code is also hard to reason about: when will the second \code{console.log} statement be executed?
-Before or after \code{baz} has been fetched?
+Notice that the code is getting deeper toward the right, this problem is often referred to as the
+“callback-hell”~\cite{McKenna_Roy}. This code is also hard to reason about: when will the second \code{console.log}
+statement be executed? Before or after \code{baz} has been fetched?
 
@@
 
-Updating the user interface usually means replacing a part of the DOM with another DOM fragment computed from data fetched by an AJAX request.
-This requires to write how to compute the new DOM fragment in JavaScript.
-For instance the following listing shows how to build a DOM tree representing a vertex in the mind map:
+Updating the user interface usually means replacing a part of the DOM with another DOM fragment computed from data
+fetched by an AJAX request. This requires to write how to compute the new DOM fragment in JavaScript. For instance
+the following listing shows how to build a DOM tree representing a vertex in the mind map:
 
@@
 
-Not only the \code{vertexDom} function is very verbose, but it does not reflect the markup nested structure, making it hard to read and reason about.
-
-Another way to build a DOM tree is to build a String containing the desired markup and then to ask the browser to parse it as HTML:
+Not only the \code{vertexDom} function is very verbose, but it does not reflect the markup nested structure, making
+it hard to read and reason about.
+
+Another way to build a DOM tree is to build a String containing the desired markup and then to ask the browser to
+parse it as HTML:
 
@@
 
-The above code may be more readable, however this implementation is wrong: if we call it with a content containing brackets (\eg{} \code{"<foo>"}) they won’t be escaped and will produce a \code{foo} tag nested in the \code{text} tag, which is not the intended behavior.
-So, although this way reads slightly better it’s not less error prone.
-
-By the way, unlike the previous points, markup generation is a task that often needs to be performed from both server and client sides.
-From the server-side it produces HTML pages that search engines can crawl and from the client-side it produces DOM fragments that can be used to update the user interface.
-So we want to share HTML fragments definition between both server and client sides to get consistency in the rendering and to avoid duplication.
-
-HTML template engines like Mustache\footnote{\href{http://mustache.github.com/}{http://mustache.github.com/}} or Closure Templates\footnote{\href{https://developers.google.com/closure/templates/}{https://developers.google.com/closure/templates/}} aim to make simpler the definition of DOM fragments by providing a convenient syntax to describe the HTML structure and allowing the insertion of dynamic expressions in a safe way.
-Some template engines can be used on both client-side and server-side, however none has a practical expression language (dynamic content can only be provided as a map of key-value pairs and no operation can be done on values) or is statically typed.
+The above code may be more readable, however this implementation is wrong: if we call it with a content containing
+brackets (\eg{} \code{"<foo>"}) they won’t be escaped and will produce a \code{foo} tag nested in the \code{text} tag,
+which is not the intended behavior. So, although this way reads slightly better it’s not less error prone.
+
+By the way, unlike the previous points, markup generation is a task that often needs to be performed from both server
+and client sides. From the server-side it produces HTML pages that search engines can crawl and from the client-side
+it produces DOM fragments that can be used to update the user interface. So we want to share HTML fragments definition
+between both server and client sides to get consistency in the rendering and to avoid duplication.
+
+HTML template engines like Mustache\footnote{\href{http://mustache.github.com/}{http://mustache.github.com/}} or
+Closure Templates\footnote{\href{https://developers.google.com/closure/templates/}{https://developers.google.com/closure/templates/}}
+aim to make simpler the definition of DOM fragments by providing a convenient syntax to describe the HTML structure
+and allowing the insertion of dynamic expressions in a safe way. Some template engines can be used on both
+client-side and server-side, however none has a practical expression language (dynamic content can only be provided
+as a map of key-value pairs and no operation can be done on values) or is statically typed.
 
@@
 
-In the previous section we showed the three main tasks performed by the JavaScript code in RIAs: events handling, asynchronous programming and DOM manipulation. We noticed the following issues:
+In the previous section we showed the three main tasks performed by the JavaScript code in RIAs: events handling,
+asynchronous programming and DOM manipulation. We noticed the following issues:
 
@@
 
-Dynamic typing may give programmers more flexibility but can make it harder for little scripts to grow into mature and robust code bases and to perform refactorings.
-For instance, in the case of RIAs, a mispell in an event name may break the program behavior without getting a sensible error message.
+Dynamic typing may give programmers more flexibility but can make it harder for little scripts to grow into mature
+and robust code bases and to perform refactorings. For instance, in the case of RIAs, a mispell in an event name may
+break the program behavior without getting a sensible error message.
 
@@
 
-Callback-driven APIs are very common in JavaScript but their inverted control makes the code flow hard to follow and bothers modularization.
-Furthermore, native APIs are too much verbose, making the code hard to read.
+Callback-driven APIs are very common in JavaScript but their inverted control makes the code flow hard to follow and
+bothers modularization. Furthermore, native APIs are too much verbose, making the code hard to read.
 
@@
 
-The lack of static checks is even harder to tackle in the case of code shared between client and server sides.
-If we write the expression \code{xs.size == 0} in a template, how should it be translated in JavaScript and in Scala?
-Several solutions are valid, depending on the meaning of the expression: in JavaScript we could translate it as \code{xs.size === 0} or \code{xs.length === 0} if \code{xs} refers to a collection (in JavaScript the property to get the size of a collection is named \code{length}).
-We need to define a proper expression language and to define how it should be translated for the client-side environment (JavaScript) and the server-side environment (\eg{} the Java Virtual Machine), but such a task requires a high effort and is hard to achieve in an extensible way (so users can still use their own data types in expressions instead of being restricted to a set of supported types).
+The lack of static checks is even harder to tackle in the case of code shared between client and server sides. If we
+write the expression \code{xs.size == 0} in a template, how should it be translated in JavaScript and in Scala?
+Several solutions are valid, depending on the meaning of the expression: in JavaScript we could translate it as
+\code{xs.size === 0} or \code{xs.length === 0} if \code{xs} refers to a collection (in JavaScript the property to get
+the size of a collection is named \code{length}). We need to define a proper expression language and to define how it
+should be translated for the client-side environment (JavaScript) and the server-side environment (\eg{} the Java
+Virtual Machine), but such a task requires a high effort and is hard to achieve in an extensible way (so users can
+still use their own data types in expressions instead of being restricted to a set of supported types).
 
@@
 
-JavaScript libraries could address some of the challenges presented above, but addressing type safety and asynchronous programming issues require language modifications (\eg{} to bring a type system and a sequencing notation).
-Defining a language requires a high effort, especially to build development tools for the language so we chose to define our language as a compiled embedded DSL is Scala\footnote{The code is available at \href{http://github.com/js-scala}{http://github.com/js-scala}}.
+JavaScript libraries could address some of the challenges presented above, but addressing type safety and
+asynchronous programming issues require language modifications (\eg{} to bring a type system and a sequencing
+notation). Defining a language requires a high effort, especially to build development tools for the language so we
+chose to define our language as a compiled embedded DSL is Scala\footnote{The code is available at
+\href{http://github.com/js-scala}{http://github.com/js-scala}}.
 
@@
 
-We use the Lightweight Modular Staging framework (LMS) to define our compiled embedded DSL.
-The main idea is that a program written using a DSL is evaluated in two stages (or steps): the first stage builds an intermediate representation of the program and the second stage transforms this intermediate representation into executable code (figure \ref{lms-diagram}).
-The embedding lets us reuse the host language infrastructure (compiler, development tools) and type system, and staging gives the opportunity to generate efficient code from expressive code by applying domain specific optimizations.
+We use the Lightweight Modular Staging framework (LMS) to define our compiled embedded DSL. The main idea is that a
+program written using a DSL is evaluated in two stages (or steps): the first stage builds an intermediate
+representation of the program and the second stage transforms this intermediate representation into executable code
+(figure \ref{lms-diagram}). The embedding lets us reuse the host language infrastructure (compiler, development
+tools) and type system, and staging gives the opportunity to generate efficient code from expressive code by applying
+domain specific optimizations.
 
@@
   \includegraphics[width=7cm]{lms.pdf}
-  \caption{Compilation of a program using LMS.
-An initial Scala program using embedded DSLs (on the left) evaluates to an intermediate representation from which the final program’s code is generated (on the right).}
+  \caption{Compilation of a program using LMS. An initial Scala program using embedded DSLs (on the left) evaluates
+  to an intermediate representation from which the final program’s code is generated (on the right).}
   \label{lms-diagram}
@@
 
-The bindings between stages are type-directed: a value of type \code{Rep[Int]} in the first stage will yield a value of type \code{Int} in the second stage. If you consider the following code:
+The bindings between stages are type-directed: a value of type \code{Rep[Int]} in the first stage will yield a value
+of type \code{Int} in the second stage. If you consider the following code:
 \begin{lstlisting}
@@
 \end{lstlisting}
-The \code{inc} function returns the intermediate representation of a computation yielding the number following the value of the parameter \code{x} (that itself is the intermediate representation of a number).
-The function looks like a regular Scala function excepted that its parameter type and its return type are wrapped in the \code{Rep[T]} type constructor that denotes intermediate representations.
-You can get a \code{T} value from a \code{Rep[T]} value by generating code from the intermediate representation and compiling it:
+The \code{inc} function returns the intermediate representation of a computation yielding the number following the
+value of the parameter \code{x} (that itself is the intermediate representation of a number). The function looks like
+a regular Scala function excepted that its parameter type and its return type are wrapped in the \code{Rep[T]} type
+constructor that denotes intermediate representations. You can get a \code{T} value from a \code{Rep[T]} value by
+generating code from the intermediate representation and compiling it:
 \begin{lstlisting}
@@
 \end{lstlisting}
-The \code{compile} function takes a staged program of type \code{Rep[A] => Rep[B]} and returns a final program of type \code{A => B}.
-
-The intermediate representation implementation is hidden for users but DSLs authors have to provide the corresponding intermediate representation of each construction of their language.
-For that purpose, LMS comes with an extensible intermediate representation implementation defining computations as a graph of statements.
-Then, the code generation process consists in sorting this graph according to expressions dependencies and to emit the code corresponding to each node.
-The listings \ref{codegen-js} and \ref{codegen-scala} show the JavaScript and Scala generated code for the \code{inc} function:
+The \code{compile} function takes a staged program of type \code{Rep[A] => Rep[B]} and returns a final program of
+type \code{A => B}.
+
+The intermediate representation implementation is hidden for users but DSLs authors have to provide the corresponding
+intermediate representation of each construction of their language. For that purpose, LMS comes with an extensible
+intermediate representation implementation defining computations as a graph of statements. Then, the code generation
+process consists in sorting this graph according to expressions dependencies and to emit the code corresponding to
+each node. The listings \ref{codegen-js} and \ref{codegen-scala} show the JavaScript and Scala generated code for the
+\code{inc} function:
 \begin{multicols}{2}
@@
 
-We want to bring safety to the events handling API so that developers can’t mispell an event name and can’t attempt to read a property that is not defined on the type of the handled event.
-The difficulty comes from the fact that the type of the event passed to the handler varies with the event name.
-A value of type \code{KeyboardEvent} contains a \code{key} field returning the value of the key involved in the event, and a value of type \code{MouseWheelEvent} contains a \code{wheelDeltaY} field returning the value of the mouse button involved in the event.
-For instance in the listing \ref{event-js}, the handler processses \code{MouseWheelEvent} values because it listens to \code{mousewheel} events.
-
-A possible solution could be to define a distinct method for each event instead of the single \code{addEventListener} method, so each method takes a handler with the according event type:
+We want to bring safety to the events handling API so that developers can’t mispell an event name and can’t attempt
+to read a property that is not defined on the type of the handled event. The difficulty comes from the fact that the
+type of the event passed to the handler varies with the event name. A value of type \code{KeyboardEvent} contains a
+\code{key} field returning the value of the key involved in the event, and a value of type \code{MouseWheelEvent}
+contains a \code{wheelDeltaY} field returning the value of the mouse button involved in the event. For instance in
+the listing \ref{event-js}, the handler processses \code{MouseWheelEvent} values because it listens to
+\code{mousewheel} events.
+
+A possible solution could be to define a distinct method for each event instead of the single \code{addEventListener}
+method, so each method takes a handler with the according event type:
 
@@
 
-However, implementing this solution requires a high effort because we have to define as many methods as there are events.
-We want to provide a single polymorphic method similar to the native API.
-To achieve that we need to encode at the type-level the dependency relation between an event name and its corresponding event type.
-This dependency can naturally and elegantly be encoded using dependent method types~\cite{Oliveira10_Typeclasses}:
+However, implementing this solution requires a high effort because we have to define as many methods as there are
+events. We want to provide a single polymorphic method similar to the native API. To achieve that we need to encode
+at the type-level the dependency relation between an event name and its corresponding event type. This dependency can
+naturally and elegantly be encoded using dependent method types~\cite{Oliveira10_Typeclasses}:
 
@@
 
-(We renamed \code{addEventListener} to \code{on}, for the sake of brevity).
-An \code{EventDef} value represents an event that carries its corresponding event type in its \code{Type} type member.
-The \code{on} method takes a parameter \code{ev} of type \code{EventDef} and a \code{handler} parameter whose type refers to the \code{Type} member of the \code{ev} parameter, so the type of the handler depends on the \code{ev} value.
-We can then define the \code{keyup} and \code{mousewheel} events as follows:
+(We renamed \code{addEventListener} to \code{on}, for the sake of brevity). An \code{EventDef} value represents an
+event that carries its corresponding event type in its \code{Type} type member. The \code{on} method takes a
+parameter \code{ev} of type \code{EventDef} and a \code{handler} parameter whose type refers to the \code{Type}
+member of the \code{ev} parameter, so the type of the handler depends on the \code{ev} value. We can then define the
+\code{keyup} and \code{mousewheel} events as follows:
 
@@
 
-The following Scala listing is equivalent to the JavaScript listing \ref{event-js} and is completely type safe: if the user misspells the event name or tries to use an undefined property on an event his program won’t compile.
+The following Scala listing is equivalent to the JavaScript listing \ref{event-js} and is completely type safe: if
+the user misspells the event name or tries to use an undefined property on an event his program won’t compile.
 
@@
 
-We propose a DSL that explicitly reflects the asynchronous nature of computations in their type and provides methods turning them into first-class citizen.
-The DSL is monadic so we can solve the callback-hell problem thanks to the Scala \code{for} notation.
-For instance, listing \ref{async-first-class} shows how the listing \ref{async-js} can be re-written with our DSL.
-The \code{createVertex} now returns an asynchronous value instead of taking a callback as parameter.
-At the usage site, the retrieval of the vertex (line 5) is separated from its insertion in the user interface (lines 6-8).
+We propose a DSL that explicitly reflects the asynchronous nature of computations in their type and provides methods
+turning them into first-class citizen. The DSL is monadic so we can solve the callback-hell problem thanks to the
+Scala \code{for} notation. For instance, listing \ref{async-first-class} shows how the listing \ref{async-js} can be
+re-written with our DSL. The \code{createVertex} now returns an asynchronous value instead of taking a callback as
+parameter. At the usage site, the retrieval of the vertex (line 5) is separated from its insertion in the user
+interface (lines 6-8).
 
@@
 
-The listing \ref{async-no-callback} translates the “callback-hell” example (listing \ref{async-callback-hell}) using our DSL.
-The \code{for} notation can intuitively be thought of as a sequencing notation: whenever the response of the first Ajax request is available, the next statement will be executed, and so on.
-There is no nested callbacks and the order of execution is directly reflected by the order of statements.
-
-An asynchronous value of type \code{Rep[A]} is modelled by a value of type \code{Rep[Future[A]]}.
-Because Scala’s \code{for} notation is just a syntactic sugar for methods named \code{foreach}, \code{map} and \code{flatMap}, we only need to define these methods on \code{Rep[Future[A]]} values, with the following semantic:
+The listing \ref{async-no-callback} translates the “callback-hell” example (listing \ref{async-callback-hell}) using
+our DSL. The \code{for} notation can intuitively be thought of as a sequencing notation: whenever the response of the
+first Ajax request is available, the next statement will be executed, and so on. There is no nested callbacks and the
+order of execution is directly reflected by the order of statements.
+
+An asynchronous value of type \code{Rep[A]} is modelled by a value of type \code{Rep[Future[A]]}. Because Scala’s
+\code{for} notation is just a syntactic sugar for methods named \code{foreach}, \code{map} and \code{flatMap}, we
+only need to define these methods on \code{Rep[Future[A]]} values, with the following semantic:
 \begin{itemize}
@@
 
-Asynchronous values can be transformed using the \code{map} and \code{flatMap} methods, turning them into first-class citizen: functions can take as parameters and return asynchronous values.
+Asynchronous values can be transformed using the \code{map} and \code{flatMap} methods, turning them into first-class
+citizen: functions can take as parameters and return asynchronous values.
 
@@
 
-Listing \ref{async-parallel-1} and \ref{async-parallel-2} define two asynchronous computations running in parallel and the generated JavaScript code (we renamed some identifiers for the sake of readability).
-The generated \code{Promise} constructor code has been omitted, it creates an object holding a list of callbacks to call when the asynchronous value is completed.
-The \code{AjaxGet} function code has also been omitted, it creates a \code{XMLHttpRequest} object that sends an HTTP request and completes its promise parameter with the response, when received.
+Listing \ref{async-parallel-1} and \ref{async-parallel-2} define two asynchronous computations running in parallel
+and the generated JavaScript code (we renamed some identifiers for the sake of readability). The generated
+\code{Promise} constructor code has been omitted, it creates an object holding a list of callbacks to call when the
+asynchronous value is completed. The \code{AjaxGet} function code has also been omitted, it creates a
+\code{XMLHttpRequest} object that sends an HTTP request and completes its promise parameter with the response, when
+received.
 
@@
 
-In this section we show how we can define, as an embedded DSL, with a small effort, a template engine that is statically typed, able to insert dynamic content in a safe way, that provides a powerful expression language, requires no extra compilation step and that can be used on both client-side and server-side.
+In this section we show how we can define, as an embedded DSL, with a small effort, a template engine that is
+statically typed, able to insert dynamic content in a safe way, that provides a powerful expression language,
+requires no extra compilation step and that can be used on both client-side and server-side.
 
@@
 \begin{itemize}
-\item a function taking some parameters and returning a DOM fragment directly models a template taking parameters and returning a DOM fragment~;
+\item a function taking some parameters and returning a DOM fragment directly models a template taking parameters and
+returning a DOM fragment~;
 \item the type system typechecks template definitions and template calls~;
@@
 
-So the only remaining work consists in defining the DSL vocabulary to define DOM nodes.
-We provide a \code{tag} function to define a tag and a \code{text} function to define a text node.
-The following listing uses our DSL and generates a code equivalent to the listing \ref{dom-api}:
+So the only remaining work consists in defining the DSL vocabulary to define DOM nodes. We provide a \code{tag}
+function to define a tag and a \code{text} function to define a text node. The following listing uses our DSL and
+generates a code equivalent to the listing \ref{dom-api}:
 
@@
 
-The readability has been highly improved: nesting tags is just like nesting code blocks, HTML entities are automatically escaped in text nodes, developers have the full computational power of Scala to inject dynamic data and DOM fragments definitions are written using functions so they compose just as functions compose.
+The readability has been highly improved: nesting tags is just like nesting code blocks, HTML entities are
+automatically escaped in text nodes, developers have the full computational power of Scala to inject dynamic data and
+DOM fragments definitions are written using functions so they compose just as functions compose.
 
@@
 
-Our DSL is equivalent as a template engine with Scala as the expression language.
-Making it usable on both server and client sides is as simple as defining another code generator for the DSL, producing Scala code.
-The result is a lightweight statically typed template engine, seamlessly integrated with the other code parts (because it is defined as an embedded DSL) and with a convenient expression language.
-For instance, the template written in the listing \ref{forest-hello} produces the following Scala code usable on server-side (the generated code for client-side is roughly equivalent to the listing \ref{dom-api}):
+Our DSL is equivalent as a template engine with Scala as the expression language. Making it usable on both server and
+client sides is as simple as defining another code generator for the DSL, producing Scala code. The result is a
+lightweight statically typed template engine, seamlessly integrated with the other code parts (because it is defined
+as an embedded DSL) and with a convenient expression language. For instance, the template written in the listing
+\ref{forest-hello} produces the following Scala code usable on server-side (the generated code for client-side is
+roughly equivalent to the listing \ref{dom-api}):
 
@@
 
-Null references are a known source of problems in programming languages~\cite{Hoare09_Null,Nanda09_Null}.
-For example, consider the following typical code finding a particular widget in the page and a then particular button in the widget:
+Null references are a known source of problems in programming languages~\cite{Hoare09_Null,Nanda09_Null}. For
+example, consider the following typical code finding a particular widget in the page and a then particular button in
+the widget:
 
@@
 
-The native \code{querySelector} method returns \code{null} if no node matched the given selector in the document.
-If we run the above code in a page where the widget is not present, it will throw an error and stop further JavaScript execution.
-We can write defensive code to handle \code{null} cases, but it leads to very cumbersome code:
+The native \code{querySelector} method returns \code{null} if no node matched the given selector in the document. If
+we run the above code in a page where the widget is not present, it will throw an error and stop further JavaScript
+execution. We can write defensive code to handle \code{null} cases, but it leads to very cumbersome code:
 
@@
 
-Our solution has both the safety and performance of the listing \ref{null-defensive} and the expressiveness of the listing \ref{null-unsafe}.
-We encode the potential emptiness of a value of type \code{Rep[A]} using the \code{Rep[Option[A]]} type at the stage level but we don’t generate code wrapping values in a container, instead we just check if a value is \code{null} or not.
-Here is a listing that uses our DSL and compiles to code equivalent to the listing \ref{null-defensive}:
+Our solution has both the safety and performance of the listing \ref{null-defensive} and the expressiveness of the
+listing \ref{null-unsafe}. We encode the potential emptiness of a value of type \code{Rep[A]} using the
+\code{Rep[Option[A]]} type at the stage level but we don’t generate code wrapping values in a container, instead we
+just check if a value is \code{null} or not. Here is a listing that uses our DSL and compiles to code equivalent to
+the listing \ref{null-defensive}:
 
@@
 
-Managing the state of an application is cumbersome and fragile in JavaScript due to the lack of encapsulation constructs.
-For example, consider the following function incrementing a counter and returning its previous value:
+Managing the state of an application is cumbersome and fragile in JavaScript due to the lack of encapsulation
+constructs. For example, consider the following function incrementing a counter and returning its previous value:
 \begin{lstlisting}[language=JavaScript,label=state-fragile,caption=Fragile state handling]
@@
 \end{lstlisting}
-This code suffers from two problems: the variable \code{x} is global and if the user calls the \code{inc} function without initializing first the \code{x} variable the behavior will be unexpected.
-
-To workaround these issues, developers can wrap the variable declaration and the \code{inc} function code in a factory function.
-Thus the state is hidden and developers may not forget the initialization:
+This code suffers from two problems: the variable \code{x} is global and if the user calls the \code{inc} function
+without initializing first the \code{x} variable the behavior will be unexpected.
+
+To workaround these issues, developers can wrap the variable declaration and the \code{inc} function code in a
+factory function. Thus the state is hidden and developers may not forget the initialization:
 
@@
 
-The code is now safer, but it is also bigger, and it still uses a mutable state.
-It’s known that mutable objects in the code make this one harder to reason about~\cite{Grogono94_Immutability,Kjolstad11_Immutability}.
-Getting rid of mutable objects is simple: instead of modifying a state, a function takes its initial value as a parameter and returns its new value in its result:
+The code is now safer, but it is also bigger, and it still uses a mutable state. It’s known that mutable objects in
+the code make this one harder to reason about~\cite{Grogono94_Immutability,Kjolstad11_Immutability}. Getting rid of
+mutable objects is simple: instead of modifying a state, a function takes its initial value as a parameter and
+returns its new value in its result:
 
@@
 
-However, explicitly threading the state in all functions parameters is not desired because of the syntactic and performance cost of adding an extra parameter to each function and of returning two results instead of one.
+However, explicitly threading the state in all functions parameters is not desired because of the syntactic and
+performance cost of adding an extra parameter to each function and of returning two results instead of one.
 Furthermore, manually passing the result of each previous call to the next one is error prone.
 
-The State monad~\cite{Wadler92_StateM} is a functional programming pattern that avoids to explicitly thread the state in functions using it or modifying it.
-In Scala, the \code{for} notation provides an elegant way to sequence monadic computations, and thus to implicitly thread a state across computations, as shown in the following listing:
+The State monad~\cite{Wadler92_StateM} is a functional programming pattern that avoids to explicitly thread the state
+in functions using it or modifying it. In Scala, the \code{for} notation provides an elegant way to sequence monadic
+computations, and thus to implicitly thread a state across computations, as shown in the following listing:
 
@@
 
-The \code{get} and \code{put} primitives are part of the State monad DSL, they allow to retrieve the current state and to change it.
-The \code{usage} value is a reusable piece of program that calls two times the \code{inc} function and prints the result of the second call.
-The last line of the listing runs the \code{usage} program on an initial state.
-
-The State monad programming model has the advantages but not the shortcomings of both listings \ref{state-encapsulated} and \ref{state-explicit} because there is no global variable (the state is implicitly threaded thanks to the State monad but state modifications explicitly use the \code{get} and \code{put} primitives) and everything is immutable.
-
-Furthermore, our implementation produces efficient imperative code using variables instead of functions passing the state through.
-To achieve that, we transform the intermediate representation of the stack of state modifications produced by the \code{for} sequences into an equivalent intermediate representation using regular variables.
-Finally, the listing \ref{state-monad} compiles to code equivalent to the listing \ref{state-fragile} so that there is no abstraction penalty.
+The \code{get} and \code{put} primitives are part of the State monad DSL, they allow to retrieve the current state
+and to change it. The \code{usage} value is a reusable piece of program that calls two times the \code{inc} function
+and prints the result of the second call. The last line of the listing runs the \code{usage} program on an initial
+state.
+
+The State monad programming model has the advantages but not the shortcomings of both listings
+\ref{state-encapsulated} and \ref{state-explicit} because there is no global variable (the state is implicitly
+threaded thanks to the State monad but state modifications explicitly use the \code{get} and \code{put} primitives)
+and everything is immutable.
+
+Furthermore, our implementation produces efficient imperative code using variables instead of functions passing the
+state through. To achieve that, we transform the intermediate representation of the stack of state modifications
+produced by the \code{for} sequences into an equivalent intermediate representation using regular variables. Finally,
+the listing \ref{state-monad} compiles to code equivalent to the listing \ref{state-fragile} so that there is no
+abstraction penalty.
 
@@
 
-Because of the dynamically typed nature of JavaScript, when calling a function there is no proper way to call a specialized implementation according to the function’s parameters types.
-JavaScript is only able to dispatch according to a method receiver prototype, \eg{} if one writes \code{foo.bar()} the JavaScript runtime will look into the prototype of the \code{foo} object for a property named \code{bar} and will call it.
-So, the only way to achieve \emph{adhoc} polymorphism on JavaScript objects consists in defining the polymorphic function on the prototypes of the objects.
-However, modifying existing object prototypes is considered to be a bad practice~\cite{Zakas12_MaintainableJs}.
-Another way could consist in manually code the dispatch logic, by registering supported data types at the beginning of the program execution, as described in section 2.4.3 of~\cite{Abelson83_SICP}, but this solution is painful for developers and incurs a performance overhead.
-
-We propose another way to achieve \emph{adhoc} polymorphism that is entirely type-directed (the dispatch happens at compile-time rather than at runtime), with no impact on the generated code and supporting retroactive extension, using typeclasses~\cite{Wadler89_AdhocPolymorphism,Odersky06_Typeclasses,Oliveira10_Typeclasses}.
+Because of the dynamically typed nature of JavaScript, when calling a function there is no proper way to call a
+specialized implementation according to the function’s parameters types. JavaScript is only able to dispatch
+according to a method receiver prototype, \eg{} if one writes \code{foo.bar()} the JavaScript runtime will look into
+the prototype of the \code{foo} object for a property named \code{bar} and will call it. So, the only way to achieve
+\emph{adhoc} polymorphism on JavaScript objects consists in defining the polymorphic function on the prototypes of
+the objects. However, modifying existing object prototypes is considered to be a bad
+practice~\cite{Zakas12_MaintainableJs}. Another way could consist in manually code the dispatch logic, by registering
+supported data types at the beginning of the program execution, as described in section 2.4.3
+of~\cite{Abelson83_SICP}, but this solution is painful for developers and incurs a performance overhead.
+
+We propose another way to achieve \emph{adhoc} polymorphism that is entirely type-directed (the dispatch happens at
+compile-time rather than at runtime), with no impact on the generated code and supporting retroactive extension,
+using typeclasses~\cite{Wadler89_AdhocPolymorphism,Odersky06_Typeclasses,Oliveira10_Typeclasses}.
 
@@
 
-The listing \ref{polymorphism} demonstrates how to define a polymorphic \code{listWidget} function that returns a DOM tree containing the representation of a list of items.
-The \code{Show[A]} typeclass defines how to produce a DOM tree for a value of type \code{A}.
+The listing \ref{polymorphism} demonstrates how to define a polymorphic \code{listWidget} function that returns a DOM
+tree containing the representation of a list of items. The \code{Show[A]} typeclass defines how to produce a DOM tree
+for a value of type \code{A}.
 
@@
 
-JavaScript comes with only one numeric type, \code{Number}, that represents double-precision 64-bit values.
-If developers want to use another numeric data type, \eg{} representing integer values or arbitrary length values, they can’t use the native arithmetic operators (\code{+}, \code{-}, \etc) on them, making the code less readable because the same concepts (\eg{} adding two numeric values) are described by different operators or methods.
-Furthermore, the code is also harder to write when it comes to mix different types of numeric values.
-Indeed, this requires to manually take care of their compatibility: before adding an arbitrary length value with a \code{Number} value, developers have to convert the latter in the former’s type.
-
-Our DSL solves these problems by defining numeric operations in a safe and extensible way: each operation is a polymorphic method and operand coercion is expressed using functional dependencies~\cite{Jones00_FunDeps}.
-It allows developers to use a homogeneous syntax for arithmetic operations with different numeric types and to mix heterogeneous types in a same operation.
-For example, the add operation is defined as follows:
+JavaScript comes with only one numeric type, \code{Number}, that represents double-precision 64-bit values. If
+developers want to use another numeric data type, \eg{} representing integer values or arbitrary length values, they
+can’t use the native arithmetic operators (\code{+}, \code{-}, \etc) on them, making the code less readable because
+the same concepts (\eg{} adding two numeric values) are described by different operators or methods. Furthermore, the
+code is also harder to write when it comes to mix different types of numeric values. Indeed, this requires to
+manually take care of their compatibility: before adding an arbitrary length value with a \code{Number} value,
+developers have to convert the latter in the former’s type.
+
+Our DSL solves these problems by defining numeric operations in a safe and extensible way: each operation is a
+polymorphic method and operand coercion is expressed using functional dependencies~\cite{Jones00_FunDeps}. It allows
+developers to use a homogeneous syntax for arithmetic operations with different numeric types and to mix
+heterogeneous types in a same operation. For example, the add operation is defined as follows:
 
@@
 
-The \code{A : Numeric} syntax means that the function is polymorphic on any type \code{A}, as long as there is an available \code{Numeric[A]} value.
-
-Then we add a \code{+} method to any \code{Rep[_]} value, using an implicit class, so we can use an homogeneous syntax to write arithmetic operations whatever the numeric type of values this operation is applied on:
+The \code{A : Numeric} syntax means that the function is polymorphic on any type \code{A}, as long as there is an
+available \code{Numeric[A]} value.
+
+Then we add a \code{+} method to any \code{Rep[_]} value, using an implicit class, so we can use an homogeneous
+syntax to write arithmetic operations whatever the numeric type of values this operation is applied on:
 
@@
 
-However, the above code does not handle operands coercion yet.
-To do so, we want to express that an expression \code{a + b} is valid either if \code{a} and \code{b} have the same numeric type, or if one of them has a numeric type and the other can be converted to this numeric type.
-The result type of such an expression depends on the types of the operands, \eg{} an integer and a double value produce a double value, two integer values produce another integer value.
-We express these constraints using functional dependencies:
+However, the above code does not handle operands coercion yet. To do so, we want to express that an expression
+\code{a + b} is valid either if \code{a} and \code{b} have the same numeric type, or if one of them has a numeric
+type and the other can be converted to this numeric type. The result type of such an expression depends on the types
+of the operands, \eg{} an integer and a double value produce a double value, two integer values produce another
+integer value. We express these constraints using functional dependencies:
 
@@
 
-The \code{ev} parameter has type \code{(A ~ B) ~> C}, which means \emph{combining a \code{A} and a \code{B} gives a \code{C}}.
-Then we need to define which combinations of \code{A}, \code{B} and \code{C} are valid:
+The \code{ev} parameter has type \code{(A ~ B) ~> C}, which means \emph{combining a \code{A} and a \code{B} gives a
+\code{C}}. Then we need to define which combinations of \code{A}, \code{B} and \code{C} are valid:
 
@@
 
-The \code{sameType} value says that two values of type \code{A} give another value of type \code{A}.
-The \code{promoteLhs} value says that a value of type \code{A} and a value of a numeric type \code{B} can be used as operands as long as there is a way to convert the value of type \code{A} to a value of type \code{B} (so the left hand side is promoted to the type \code{B}).
-The \code{promoteRhs} does the same for the right hand side.
-
-The \code{\~\>} type takes two parameters, saying how to obtain a value of the target type from each operand. Here is the definition of the \code{\~\>} type:
+The \code{sameType} value says that two values of type \code{A} give another value of type \code{A}. The
+\code{promoteLhs} value says that a value of type \code{A} and a value of a numeric type \code{B} can be used as
+operands as long as there is a way to convert the value of type \code{A} to a value of type \code{B} (so the left
+hand side is promoted to the type \code{B}). The \code{promoteRhs} does the same for the right hand side.
+
+The \code{\~\>} type takes two parameters, saying how to obtain a value of the target type from each operand. Here is
+the definition of the \code{\~\>} type:
 
@@
 
-The following listing demonstrates the homogeneous syntax to perform an addition on different data types and the operand coercion between heterogeneous types:
+The following listing demonstrates the homogeneous syntax to perform an addition on different data types and the
+operand coercion between heterogeneous types:
 
@@
 
-We implemented our language as a compiled embedded DSL in Scala.
-The embedding gave us the opportunity to reuse Scala’s infrastructure: we didn’t have to write a parser and, more importantly, we could reuse Scala’s type system.
+We implemented our language as a compiled embedded DSL in Scala. The embedding gave us the opportunity to reuse
+Scala’s infrastructure: we didn’t have to write a parser and, more importantly, we could reuse Scala’s type system.
 Furthermore, staging gave us the opportunity to generate efficient code.
 
-Code generation using LMS is usually described as a two step process: a program first evaluates to an intermediate representation and then the final program’s code is generated from the intermediate representation.
-We think we can consider a third step, occurring before the program evaluation: the compilation.
-
-Indeed, the code generation process consists in three steps giving DSL authors as many opportunities to implement their language features.
-First, features like type coercion and \emph{adhoc} polymorphism only rely on the type system, so they operate during the compilation.
-Then, during the evaluation, the DSL constructs build an intermediate representation of the program, for instance the State monad DSL builds an imperative sequence of assignments on a variable.
-Last, the code generator produces the final program using the domain specific information available in the intermediate representation to generate efficient code, for example the Option monad code generator produces code checking against nullity before dereferencing a value.
+Code generation using LMS is usually described as a two step process: a program first evaluates to an intermediate
+representation and then the final program’s code is generated from the intermediate representation. We think we can
+consider a third step, occurring before the program evaluation: the compilation.
+
+Indeed, the code generation process consists in three steps giving DSL authors as many opportunities to implement
+their language features. First, features like type coercion and \emph{adhoc} polymorphism only rely on the type
+system, so they operate during the compilation. Then, during the evaluation, the DSL constructs build an intermediate
+representation of the program, for instance the State monad DSL builds an imperative sequence of assignments on a
+variable. Last, the code generator produces the final program using the domain specific information available in the
+intermediate representation to generate efficient code, for example the Option monad code generator produces code
+checking against nullity before dereferencing a value.
 
@@
 
-Each component of a DSL is defined in a trait and each corresponding code generator for a given target is also defined in a trait.
-The Scala language allows developers to mix several traits together, so they can build their language by picking the DSL traits they want to use.
-
-Furthermore, a code generator can be tweaked by defining a sub-trait specializing some of its methods.
-This is useful to handle cross-browser incompatibilities.
-For instance we have written code generators specializing the output for some statement nodes in order to handle Internet Explorer incompatibilities.
+Each component of a DSL is defined in a trait and each corresponding code generator for a given target is also
+defined in a trait. The Scala language allows developers to mix several traits together, so they can build their
+language by picking the DSL traits they want to use.
+
+Furthermore, a code generator can be tweaked by defining a sub-trait specializing some of its methods. This is useful
+to handle cross-browser incompatibilities. For instance we have written code generators specializing the output for
+some statement nodes in order to handle Internet Explorer incompatibilities.
 
@@
 
-Our DSL aims to give developers a safer and more expressive way to write the client-side code.
-We focused on the main tasks of RIAs development: handling events, asynchronous programming and DOM definition.
+Our DSL aims to give developers a safer and more expressive way to write the client-side code. We focused on the main
+tasks of RIAs development: handling events, asynchronous programming and DOM definition.
 
@@
 
-Using our DSL, we were able to divide by two the length of some parts of code (DOM definition and \code{null} reference handling).
-The size of code parts dealing with events did not change but they gain type safety.
-Finally, code handling asynchronous computations were easier to modularize.
-
-All these benefits came with no runtime overhead, excepted for the asynchronous programming DSL that creates, for each asynchronous computation, a \code{Promise} value registering the callbacks to call when the value is be available.
+Using our DSL, we were able to divide by two the length of some parts of code (DOM definition and \code{null}
+reference handling). The size of code parts dealing with events did not change but they gain type safety. Finally,
+code handling asynchronous computations were easier to modularize.
+
+All these benefits came with no runtime overhead, excepted for the asynchronous programming DSL that creates, for
+each asynchronous computation, a \code{Promise} value registering the callbacks to call when the value is be
+available.
 
@@
 
-The most popular JavaScript library, jQuery~\cite{Bibeault08_jQuery}, used by more than 40\% of the top million sites\footnote{\href{http://trends.builtwith.com/javascript}{http://trends.builtwith.com/javascript}}, and Q.js\footnote{\href{https://github.com/kriskowal/q}{https://github.com/kriskowal/q}} provide an API turning asynchronous computations into first-class citizen.
-However, the callback-hell problem can’t always be avoided due to the absence of a special syntactic sugar for continuation passing style.
-
-jQuery handles the \code{null} references problem by wrapping each query result in a container so before each further method call it tests the emptiness of the container and applies effectively the operation only if the container is not empty.
-It provides an expressive API but the emptiness checking involves a slight performance penalty since each result is wrapped and then unwrapped for the subsequent method invocation.
+The most popular JavaScript library, jQuery~\cite{Bibeault08_jQuery}, used by more than 40\% of the top million
+sites\footnote{\href{http://trends.builtwith.com/javascript}{http://trends.builtwith.com/javascript}}, and
+Q.js\footnote{\href{https://github.com/kriskowal/q}{https://github.com/kriskowal/q}} provide an API turning
+asynchronous computations into first-class citizen. However, the callback-hell problem can’t always be avoided due to
+the absence of a special syntactic sugar for continuation passing style.
+
+jQuery handles the \code{null} references problem by wrapping each query result in a container so before each further
+method call it tests the emptiness of the container and applies effectively the operation only if the container is
+not empty. It provides an expressive API but the emptiness checking involves a slight performance penalty since each
+result is wrapped and then unwrapped for the subsequent method invocation.
 
@@
 
-Templates written using our HTML templating DSL are about two times shorter than templates written using the native JavaScript API.
-Compared to usual template engines our solution has no runtime overhead, does not require a separated compilation step and provides a flexible type safe expression language.
-
-Furthermore, templates can be shared between client and server sides.
-Opa is also a language making it possible to share statically typed templates with a convenient expression language between client and server sides.
-In Opa they made this possible because HTML templating is part of the language itself.
-However, creating a brand new language for each problem requires a high effort, especially since the effort to define common language constructs is duplicated.
-To reduce the effort, the solution consists in implementing abstractions specific to a given problem as libraries on top of a general purpose language.
-For instance, Kotlin provides an embedded DSL similar to ours to define HTML fragments.
-Because the Kotlin language can be compiled to JavaScript and to JVM bytecode, HTML templates can be shared between client and server sides.
-However, their DSL does not produce code building DOM fragments on client-side.
-Actually, on both server-side and client-side the output is the same: an abstract tree containing the HTML fragment.
-So, users need to transform this AST into a proper DOM fragment using a function specific to the client-side.
-Kotlin and similar approaches (Fay, ClojureScript, GWT) can’t generate specialized code for a given abstraction because their translation scheme is based on the initial program source code: if an abstraction is defined as a library, they will generate code reimplementing the library on each backend (client-side and server-side).
-There is no way to specialize the generated code for a given library.
-
-On one hand defining a new language (or modifying an existing language) to integrate a concept specific to a problem is not desired either because it requires a high-effort to define a full-featured language or because several small languages are hard to integrate seamlessly together.
-On the other hand, defining a concept specific to a problem as a library allows developers to concentrate their effort on the concept itself and to reuse the host language infrastructure, but they have no way to specialize, for each backend, the code generated by the compiler for their concept.
-
-Compiled embedded DSLs provide the best of both worlds because DSLs are defined as libraries and, because they are compiled, developers can specialize the generated code, for each backend, for their DSL concepts.
-This opportunity made it possible to define, as a library, the HTML templating DSL generating code building DOM fragments on client-side and code building XML trees on server-side.
+Templates written using our HTML templating DSL are about two times shorter than templates written using the native
+JavaScript API. Compared to usual template engines our solution has no runtime overhead, does not require a separated
+compilation step and provides a flexible type safe expression language.
+
+Furthermore, templates can be shared between client and server sides. Opa is also a language making it possible to
+share statically typed templates with a convenient expression language between client and server sides. In Opa they
+made this possible because HTML templating is part of the language itself. However, creating a brand new language for
+each problem requires a high effort, especially since the effort to define common language constructs is duplicated.
+To reduce the effort, the solution consists in implementing abstractions specific to a given problem as libraries on
+top of a general purpose language. For instance, Kotlin provides an embedded DSL similar to ours to define HTML
+fragments. Because the Kotlin language can be compiled to JavaScript and to JVM bytecode, HTML templates can be
+shared between client and server sides. However, their DSL does not produce code building DOM fragments on
+client-side. Actually, on both server-side and client-side the output is the same: an abstract tree containing the
+HTML fragment. So, users need to transform this AST into a proper DOM fragment using a function specific to the
+client-side. Kotlin and similar approaches (Fay, ClojureScript, GWT) can’t generate specialized code for a given
+abstraction because their translation scheme is based on the initial program source code: if an abstraction is
+defined as a library, they will generate code reimplementing the library on each backend (client-side and
+server-side). There is no way to specialize the generated code for a given library.
+
+On one hand defining a new language (or modifying an existing language) to integrate a concept specific to a problem
+is not desired either because it requires a high-effort to define a full-featured language or because several small
+languages are hard to integrate seamlessly together. On the other hand, defining a concept specific to a problem as a
+library allows developers to concentrate their effort on the concept itself and to reuse the host language
+infrastructure, but they have no way to specialize, for each backend, the code generated by the compiler for their
+concept.
+
+Compiled embedded DSLs provide the best of both worlds because DSLs are defined as libraries and, because they are
+compiled, developers can specialize the generated code, for each backend, for their DSL concepts. This opportunity
+made it possible to define, as a library, the HTML templating DSL generating code building DOM fragments on
+client-side and code building XML trees on server-side.
 
@@
 
-On the other side of the coin, because that is not possible to define a library without defining how to translate it on both server-side and client-side, there is no simple way to define a library for defining libraries.
-This feature comes almost for free with approaches like GWT, Fay or Kotlin because they derive the client-side program and the server-side program from the initial program source (that contains data type definitions such as classes definitions).
-In our case, the client-side program and the server-side program are derived from the intermediate representation produced by the evaluation of the initial program.
-So, only concepts reified in an intermediate representation during this evaluation can be part of the final programs.
-However regular Scala’s classes definition are not reified in an intermediate representation (it would mean that the constructor of a type \code{A} would yield a value of type \code{Rep[A]}).
-Scala-virtualized~\cite{Moors12_Virtualized} provides such a mechanism but is limited to product types and provides only projection functions, so for more complex cases users have to write some repetitive and boilerplate code.
-
-In summary, the compiled embedded approach may require more work to support custom data types but gives control on the way this data type will be encoded for each target platform (\eg{} the JavaScript code generator uses native arrays to encode the lists of the list manipulation DSL).
-
-However, the next release of Scala will have support for macros making it possible to generate reified data types from a concise syntax without relying on a compiler extension.
+On the other side of the coin, because that is not possible to define a library without defining how to translate it
+on both server-side and client-side, there is no simple way to define a library for defining libraries. This feature
+comes almost for free with approaches like GWT, Fay or Kotlin because they derive the client-side program and the
+server-side program from the initial program source (that contains data type definitions such as classes
+definitions). In our case, the client-side program and the server-side program are derived from the intermediate
+representation produced by the evaluation of the initial program. So, only concepts reified in an intermediate
+representation during this evaluation can be part of the final programs. However regular Scala’s classes definition
+are not reified in an intermediate representation (it would mean that the constructor of a type \code{A} would yield
+a value of type \code{Rep[A]}). Scala-virtualized~\cite{Moors12_Virtualized} provides such a mechanism but is 
+limited to product types and provides only projection functions, so for more complex cases users have to write some
+repetitive and boilerplate code.
+
+In summary, the compiled embedded approach may require more work to support custom data types but gives control on
+the way this data type will be encoded for each target platform (\eg{} the JavaScript code generator uses native
+arrays to encode the lists of the list manipulation DSL).
+
+However, the next release of Scala will have support for macros making it possible to generate reified data types
+from a concise syntax without relying on a compiler extension.
 

### 2012-12-07 papers/ecoop2013/repairing.tex
That’s enough for today

@@
+\def\thelstlisting{}
 \documentclass[american,english,runningheads]{llncs}
@@
 
+\setcounter{lstlisting}{0}
+\renewcommand{\thelstlisting}{\arabic{lstlisting}}
+
 \title{Statically typed Web programming}
 
-\author{Julien \noun{Richard-Foy} \and Olivier \noun{Barais} \and Jean-Marc \noun{Jézéquel}\\
-\texttt{\{first\}.\{last\}@irisa.fr}}
-\institute{INRIA}
+\author{Julien \noun{Richard-Foy}\textsuperscript{\textasteriskcentered} \and Olivier \noun{Barais}\textsuperscript{\textasteriskcentered} \and
+Tiark \noun{Rompf}\textsuperscript{\textdagger} \and Jean-Marc \noun{Jézéquel}\textsuperscript{\textasteriskcentered}}
+\institute{\textsuperscript{\textasteriskcentered}IRISA, Université de Rennes 1, France: \texttt{\{first\}.\{last\}@irisa.fr}\\
+\textsuperscript{\textdagger}EPFL, Lausanne, Switzerland: \texttt{\{first\}.\{last\}@epfl.ch}}
 
@@
 To illustrate the challenges of writing a RIA and the need for a DSL for writing them, we introduce a simple “mind
-mapping” application and show how the client-side code would be typically written in pure JavaScript.
+mapping” application and show how the client-side code would be typically written in JavaScript.
 
@@
 The event handler is simply a function taking the event data as a parameter. In the above example we use the
-\code{wheelDeltaY} property of the event to change the scale of the map.
+\code{wheelDeltaY} property of the event to update the scale of the map.
 
@@
 (that is a \emph{callback}) when the response data is available. The inverted control makes modularization harder to
-achieve: in the above listing the function creating the vertex is also responsible to updating the user interface.
-If we want to relax this coupling the only option is to add a callback parameter to the \code{createVertex} function:
+achieve: in the above listing the function creating the vertex is also responsible of updating the user interface.
+To relax this coupling the only option is to add a callback parameter to the \code{createVertex} function:
 
@@
 \subsection{Assessment}
+\label{assessment}
 
@@
 The lack of static checks is even harder to tackle in the case of code shared between client and server sides. If we
-write the expression \code{xs.size == 0} in a template, how should it be translated in JavaScript and in Scala?
-Several solutions are valid, depending on the meaning of the expression: in JavaScript we could translate it as
-\code{xs.size === 0} or \code{xs.length === 0} if \code{xs} refers to a collection (in JavaScript the property to get
-the size of a collection is named \code{length}). We need to define a proper expression language and to define how it
-should be translated for the client-side environment (JavaScript) and the server-side environment (\eg{} the Java
-Virtual Machine), but such a task requires a high effort and is hard to achieve in an extensible way (so users can
-still use their own data types in expressions instead of being restricted to a set of supported types).
+write the expression \code{xs.size == 0} in a template, how should it be translated to the client-side environment
+(JavaScript) and to the server-side environment (\eg{} the Java Virtual Machine)? Several solutions are valid,
+depending on the meaning of the expression: in JavaScript we could translate it as \code{xs.size === 0} or
+\code{xs.length === 0} if \code{xs} refers to a collection (in JavaScript the property to get the size of a
+collection is named \code{length}). We need to define a proper expression language and to define how it should be
+translated for the client-side environment and the server-side environment, but such a task requires a high effort
+and is hard to achieve in an extensible way (so users can still integrate their own data types in expressions instead
+of being restricted to a set of supported types).
 
@@
 notation). Defining a language requires a high effort, especially to build development tools for the language so we
-chose to define our language as a compiled embedded DSL is Scala\footnote{The code is available at
+chose to define our language as a compiled embedded DSL in Scala\footnote{The code is available at
 \href{http://github.com/js-scala}{http://github.com/js-scala}}.
@@
 \end{lstlisting}
-The \code{inc} function returns the intermediate representation of a computation yielding the number following the
-value of the parameter \code{x} (that itself is the intermediate representation of a number). The function looks like
+The function looks like
 a regular Scala function excepted that its parameter type and its return type are wrapped in the \code{Rep[T]} type
-constructor that denotes intermediate representations. You can get a \code{T} value from a \code{Rep[T]} value by
-generating code from the intermediate representation and compiling it:
+constructor that denotes intermediate representations. The \code{+} operator has been defined on \code{Rep[Int]}
+values and returns the intermediate representation of an addition. Finally, the \code{inc} function returns the
+intermediate representation of a computation yielding the number following the value of the parameter \code{x}. You
+can get a \code{T} value from a \code{Rep[T]} value by generating code from the intermediate representation and
+compiling it:
 \begin{lstlisting}
@@
 The intermediate representation implementation is hidden for users but DSLs authors have to provide the corresponding
-intermediate representation of each construction of their language. For that purpose, LMS comes with an extensible
-intermediate representation implementation defining computations as a graph of statements. Then, the code generation
-process consists in sorting this graph according to expressions dependencies and to emit the code corresponding to
-each node. The listings \ref{codegen-js} and \ref{codegen-scala} show the JavaScript and Scala generated code for the
-\code{inc} function:
+intermediate representation of each construct of their language. For that purpose, LMS comes with an extensible
+intermediate representation implementation defining computations as a graph of statements. In the case of the
+\code{inc} function, this graph contains a \code{Plus} node applied on a \code{Sym} node (the \code{x} parameter) and
+a \code{Const} node (the literal value \code{1}).
+
+Then, the code generation process consists in sorting this graph according to expressions dependencies and to emit
+the code corresponding to each node. The listings \ref{codegen-js} and \ref{codegen-scala} show the JavaScript and
+Scala generated code for the \code{inc} function:
 \begin{multicols}{2}
@@
 
+The remaining of this section describe the design and the implementation of DSLs for Web programming using LMS.
+
 \subsection{Events handling}
@@
 to read a property that is not defined on the type of the handled event. The difficulty comes from the fact that the
-type of the event passed to the handler varies with the event name. A value of type \code{KeyboardEvent} contains a
-\code{key} field returning the value of the key involved in the event, and a value of type \code{MouseWheelEvent}
-contains a \code{wheelDeltaY} field returning the value of the mouse button involved in the event. For instance in
-the listing \ref{event-js}, the handler processses \code{MouseWheelEvent} values because it listens to
-\code{mousewheel} events.
+type of the event passed to the handler varies with the event name. For instance in the listing \ref{event-js}, the
+handler processses \code{MouseWheelEvent} values because it listens to \code{mousewheel} events. Values of type
+\code{MouseWheelEvent} have a property \code{wheelDeltaY} but that’s not the case of an event data of type
+\code{KeyboardEvent}, for example.
 
@@
 \end{lstlisting}
+
+In the above listing, the \code{onKeyUp} method attaches an event handler for the \code{keyup} event that uses values
+of type \code{KeyboardEvent}, and the \code{onMouseWheel} method does the same for \code{mousewheel} events that use
+values of type \code{MouseWheelEvent}.
 
@@
 (We renamed \code{addEventListener} to \code{on}, for the sake of brevity). An \code{EventDef} value represents an
-event that carries its corresponding event type in its \code{Type} type member. The \code{on} method takes a
+event that carries its corresponding event data type in its \code{Type} type member. The \code{on} method takes a
 parameter \code{ev} of type \code{EventDef} and a \code{handler} parameter whose type refers to the \code{Type}
@@
 \centering
-\begin{lstlisting}[numbers=left,caption=Asynchronous values are first class citizen,label=async-first-class]
+\begin{lstlisting}[caption=Asynchronous values are first class citizen,label=async-first-class]
 def createVertex(text: Rep[String]): Rep[Future[Vertex]] =
@@
 
-// Usage
-val vertexAsync = createVertex("Hello, World!")
-for (vertex <- vertexAsync) {
+for (vertex <- createVertex("Hello, World!")) {
   addVertex(vertex)
@@
 turning them into first-class citizen. The DSL is monadic so we can solve the callback-hell problem thanks to the
-Scala \code{for} notation. For instance, listing \ref{async-first-class} shows how the listing \ref{async-js} can be
+Scala \code{for} notation.
+
+For instance, listing \ref{async-first-class} shows how the listing \ref{async-js} can be
 re-written with our DSL. The \code{createVertex} now returns an asynchronous value instead of taking a callback as
-parameter. At the usage site, the retrieval of the vertex (line 5) is separated from its insertion in the user
-interface (lines 6-8).
+parameter. Then, the \code{for} expression allows us to use the vertex, when available, and to insert it on the user
+interface. By making the \code{createVertex} function returning an asynchronous value instead of taking a callback as
+a parameter, the code is easier to modularize into loosely coupled parts.
 
@@
 An asynchronous value of type \code{Rep[A]} is modelled by a value of type \code{Rep[Future[A]]}. Because Scala’s
-\code{for} notation is just a syntactic sugar for methods named \code{foreach}, \code{map} and \code{flatMap}, we
-only need to define these methods on \code{Rep[Future[A]]} values, with the following semantic:
+\code{for} notation is just a syntactic sugar for methods \code{foreach}, \code{map} and \code{flatMap}, we are
+able to define our DSL by just defining these methods on \code{Rep[Future[A]]} values, with the following semantic:
 \begin{itemize}
@@
 automatically escaped in text nodes, developers have the full computational power of Scala to inject dynamic data and
-DOM fragments definitions are written using functions so they compose just as functions compose.
-
-\subsubsection{Reuse the DOM definition DSL from server-side}
+DOM fragments definitions are written using functions so they compose just as functions compose. These benefits come
+with no performance loss because the DSL generates code building DOM fragments using the native JavaScript API.
+
+\subsubsection{Reuse the DOM definition DSL on server-side}
 
 Our DSL is equivalent as a template engine with Scala as the expression language. Making it usable on both server and
-client sides is as simple as defining another code generator for the DSL, producing Scala code. The result is a
-lightweight statically typed template engine, seamlessly integrated with the other code parts (because it is defined
-as an embedded DSL) and with a convenient expression language. For instance, the template written in the listing
-\ref{forest-hello} produces the following Scala code usable on server-side (the generated code for client-side is
-roughly equivalent to the listing \ref{dom-api}):
+client sides is surprisingly as simple as defining another code generator for the DSL, producing Scala code.
+
+For instance, the template written in the listing \ref{forest-hello} produces the following Scala code usable on
+server-side (the generated code for client-side is roughly equivalent to the listing \ref{dom-api}):
 
@@
 
+We are able to tackle the code sharing issues described in section \ref{assessment} because of the embdedded nature
+of our DSLs: dynamic content of templates is written using embdedded DSLs too, so their translation into JavaScript
+and Scala is managed by their respective code generators.
+
 \subsection{More general purpose improvements of JavaScript}
@@
 listing \ref{null-unsafe}. We encode the potential emptiness of a value of type \code{Rep[A]} using the
-\code{Rep[Option[A]]} type at the stage level but we don’t generate code wrapping values in a container, instead we
-just check if a value is \code{null} or not. Here is a listing that uses our DSL and compiles to code equivalent to
-the listing \ref{null-defensive}:
+\code{Rep[Option[A]]} type at the stage level but we don’t generate code wrapping values in a container, instead the
+generated code just checks if a value is \code{null} or not. Here is a listing that uses our DSL and compiles to code
+equivalent to the listing \ref{null-defensive}:
 
@@
 
-We propose another way to achieve \emph{adhoc} polymorphism that is entirely type-directed (the dispatch happens at
-compile-time rather than at runtime), with no impact on the generated code and supporting retroactive extension,
-using typeclasses~\cite{Wadler89_AdhocPolymorphism,Odersky06_Typeclasses,Oliveira10_Typeclasses}.
+We propose to achieve \emph{adhoc} polymorphism using
+typeclasses~\cite{Wadler89_AdhocPolymorphism,Odersky06_Typeclasses,Oliveira10_Typeclasses} so that it supports
+retroactive extension without modifying objects prototypes because it is type-directed: the dispatch happens at
+compile-time rather than at runtime.
 
@@
 
-(TODO More general problem description)
-
 JavaScript comes with only one numeric type, \code{Number}, that represents double-precision 64-bit values. If
@@
 
-We implemented our language as a compiled embedded DSL in Scala. The embedding gave us the opportunity to reuse
-Scala’s infrastructure: we didn’t have to write a parser and, more importantly, we could reuse Scala’s type system.
-Furthermore, staging gave us the opportunity to generate efficient code.
-
-Code generation using LMS is usually described as a two step process: a program first evaluates to an intermediate
-representation and then the final program’s code is generated from the intermediate representation. We think we can
-consider a third step, occurring before the program evaluation: the compilation.
-
-Indeed, the code generation process consists in three steps giving DSL authors as many opportunities to implement
-their language features. First, features like type coercion and \emph{adhoc} polymorphism only rely on the type
-system, so they operate during the compilation. Then, during the evaluation, the DSL constructs build an intermediate
-representation of the program, for instance the State monad DSL builds an imperative sequence of assignments on a
-variable. Last, the code generator produces the final program using the domain specific information available in the
-intermediate representation to generate efficient code, for example the Option monad code generator produces code
-checking against nullity before dereferencing a value.
-
-Each of these steps takes out some information on the program.
+\subsubsection{Benefits of the compiled embedded approach}
+
+We implemented our language as a compiled embedded DSL in Scala. The process of generating code from a DSL is usually
+described as a two steps process: a program first evaluates to an intermediate representation and then the final
+program’s code is generated from this intermediate representation, with the opportunity to generate efficient code
+by applying domain specific optimizations. We think it’s worth considering a third step, which occurs before the
+program evaluation: its compilation. This step affects DSLs design because their embedded nature let them reuse the
+host language features (\eg{} the type system or syntactic sugars) to implement their own features. The following
+paragraph gives an overview of the different mechanisms used to implement our DSLs.
+
+First, features like type coercion and \emph{adhoc} polymorphism only rely on the type system, so they operate during
+the compilation. Then, during the evaluation, the DSL constructs build an intermediate representation of the program,
+for instance the State monad DSL builds an imperative sequence of assignments on a variable. Last, the code generator
+produces the final program using the domain specific information available in the intermediate representation to
+generate efficient code, for example the Option monad code generator produces code checking against nullity before
+dereferencing a value.
+
+An important consequence of the implementation as compiled embedded DSLs is the low effort required to be able to
+share code between server and client sides. The other ways to define DSLs compiling to multiple backends are either
+to define a brand new language so the compilation generates appropriate code for each target (\eg{} Opa), or to
+define a library within a host language that already compiles to several targets (\eg{} GWT, Kotlin). However, the
+first way requires a high effort or leads to defining several small languages that are hard to integrate seamlessly,
+and the second way does not give the opportunity to specialize the generated code according to the targets (so you
+can’t generate, from the same library, code building DOM fragments on the client-side and code building XML trees on
+the server-side).
+
+Compiled embedded DSLs are a good tradeoff because they are simply defined as libraries but they let developers
+specialize the generated code according to each target. This opportunity made it possible to define, as a library,
+the HTML templating DSL generating code building DOM fragments on client-side and code building XML trees on
+server-side.
+
+% Furthermore, templates can be shared between client and server sides. Opa is also a language making it possible to
+% share statically typed templates with a convenient expression language between client and server sides. In Opa they
+% made this possible because HTML templating is part of the language itself. However, creating a brand new language for
+% each problem requires a high effort, especially since the effort to define common language constructs is duplicated.
+% To reduce the effort, the solution consists in implementing abstractions specific to a given problem as libraries on
+% top of a general purpose language. For instance, Kotlin provides an embedded DSL similar to ours to define HTML
+% fragments. Because the Kotlin language can be compiled to JavaScript and to JVM bytecode, HTML templates can be
+% shared between client and server sides. However, their DSL does not produce code building DOM fragments on
+% client-side. Actually, on both server-side and client-side the output is the same: an abstract tree containing the
+% HTML fragment. So, users need to transform this AST into a proper DOM fragment using a function specific to the
+% client-side. Kotlin and similar approaches (Fay, ClojureScript, GWT) can’t generate specialized code for a given
+% abstraction because their translation scheme is based on the initial program source code: if an abstraction is
+% defined as a library, they will generate code reimplementing the library on each backend (client-side and
+% server-side). There is no way to specialize the generated code for a given library.
+% 
+% On one hand defining a new language (or modifying an existing language) to integrate a concept specific to a problem
+% is not desired either because it requires a high-effort to define a full-featured language or because several small
+% languages are hard to integrate seamlessly together. On the other hand, defining a concept specific to a problem as a
+% library allows developers to concentrate their effort on the concept itself and to reuse the host language
+% infrastructure, but they have no way to specialize, for each backend, the code generated by the compiler for their
+% concept.
+
+\subsection{Lack of a data type definition DSL}
+
+(TODO Example)
+
+On the other side of the coin, because that is not possible to define a library without defining how to translate it
+on both server and client sides, there is no simple way to let DSLs users define their own libraries. This feature
+comes almost for free with approaches like GWT, Fay or Kotlin because they derive the client-side program and the
+server-side program from the initial program source (that contains data type definitions such as classes
+definitions). In our case, the client-side program and the server-side program are derived from the intermediate
+representation produced by the evaluation of the initial program. So, only concepts reified in an intermediate
+representation during this evaluation can be part of the final programs. However regular Scala’s classes definition
+are not reified in an intermediate representation (it would mean that the constructor of a type \code{A} would yield
+a value of type \code{Rep[A]}). Scala-virtualized~\cite{Moors12_Virtualized} provides such a mechanism but is 
+limited to record types and provides only projection functions, so for more complex cases users have to write some
+repetitive and boilerplate code.
+
+In summary, the compiled embedded approach may require more work to support custom data types but gives control on
+the way this data type will be encoded for each target platform (\eg{} the JavaScript code generator uses native
+arrays to encode the lists of the list manipulation DSL).
+
+However, the next release of Scala will have support for macros making it possible to generate reified data types
+from a concise syntax without relying on a compiler extension.
 
@@
 
-\section{Validation and comparison with other approaches}
+\section{Evaluation}
 \label{validation}
@@
 Our DSL aims to give developers a safer and more expressive way to write the client-side code. We focused on the main
-tasks of RIAs development: handling events, asynchronous programming and DOM definition.
+tasks of RIAs development: handling events, asynchronous programming and DOM fragments definition, as well as on
+more general programming concerns: \code{null} references handling and type safety.
 
@@
 
-\subsection{JavaScript libraries}
+\subsection{Comparison with other approaches}
+
+Table \ref{comparison} gives a synthetic view of the comparison of our work with other approaches. Lines give
+features desired to write large Web applications and colums list famous tools for Web programming. Cells show the
+support level of each feature for each tool.
+
+\begin{table}
+\centering
+\begin{tabular}{| l | c | c | c | c | c | c | c | c |}
+\hline
+& jQuery & TypeScript & Roy & GWT & Kotlin & Opa & Dart & js-scala \\
+\hline
+Event handling & * & & & *** & & ? & & *** \\
+\hline
+DOM definition & * & & & *** & ** & *** & * & *** \\
+\hline
+Asynchronous programming & ** & & *** & & & *** & * & *** \\
+\hline
+Null references & ** & & *** & & * & ? & & *** \\
+\hline
+Type coercion & & & & * & * & ? & * & *** \\
+\hline
+General programming language features & * & ** & ** & * & *** & ? & * & *** \\
+\hline
+Static typing & & * & ** & ** & ** & ** & * & ** \\
+\hline
+Easiness for defining libraries & * & *** & *** & *** & *** & *** & *** & * \\
+\hline
+Client-server code sharing & & * & * & * & * & ** & * & *** \\
+\hline
+\end{tabular}
+\caption{Support level of Web programming features by famous tools. Zero star means that the feature is not supported
+at all, three stars mean a strong support.}
+\label{comparison}
+\end{table}
 
 The most popular JavaScript library, jQuery~\cite{Bibeault08_jQuery}, used by more than 40\% of the top million
-sites\footnote{\href{http://trends.builtwith.com/javascript}{http://trends.builtwith.com/javascript}}, and
-Q.js\footnote{\href{https://github.com/kriskowal/q}{https://github.com/kriskowal/q}} provide an API turning
-asynchronous computations into first-class citizen. However, the callback-hell problem can’t always be avoided due to
-the absence of a special syntactic sugar for continuation passing style.
-
-jQuery handles the \code{null} references problem by wrapping each query result in a container so before each further
-method call it tests the emptiness of the container and applies effectively the operation only if the container is
-not empty. It provides an expressive API but the emptiness checking involves a slight performance penalty since each
-result is wrapped and then unwrapped for the subsequent method invocation.
-
-\subsection{Languages usable on both server-side and client-side}
-
-GWT also has a type safe API to handle events.
-
-Roy solves the callback-hell problem.
-
-\subsection{Template engines}
-
-Templates written using our HTML templating DSL are about two times shorter than templates written using the native
-JavaScript API. Compared to usual template engines our solution has no runtime overhead, does not require a separated
-compilation step and provides a flexible type safe expression language.
-
-Furthermore, templates can be shared between client and server sides. Opa is also a language making it possible to
-share statically typed templates with a convenient expression language between client and server sides. In Opa they
-made this possible because HTML templating is part of the language itself. However, creating a brand new language for
-each problem requires a high effort, especially since the effort to define common language constructs is duplicated.
-To reduce the effort, the solution consists in implementing abstractions specific to a given problem as libraries on
-top of a general purpose language. For instance, Kotlin provides an embedded DSL similar to ours to define HTML
-fragments. Because the Kotlin language can be compiled to JavaScript and to JVM bytecode, HTML templates can be
-shared between client and server sides. However, their DSL does not produce code building DOM fragments on
-client-side. Actually, on both server-side and client-side the output is the same: an abstract tree containing the
-HTML fragment. So, users need to transform this AST into a proper DOM fragment using a function specific to the
-client-side. Kotlin and similar approaches (Fay, ClojureScript, GWT) can’t generate specialized code for a given
-abstraction because their translation scheme is based on the initial program source code: if an abstraction is
-defined as a library, they will generate code reimplementing the library on each backend (client-side and
-server-side). There is no way to specialize the generated code for a given library.
-
-On one hand defining a new language (or modifying an existing language) to integrate a concept specific to a problem
-is not desired either because it requires a high-effort to define a full-featured language or because several small
-languages are hard to integrate seamlessly together. On the other hand, defining a concept specific to a problem as a
-library allows developers to concentrate their effort on the concept itself and to reuse the host language
-infrastructure, but they have no way to specialize, for each backend, the code generated by the compiler for their
-concept.
-
-Compiled embedded DSLs provide the best of both worlds because DSLs are defined as libraries and, because they are
-compiled, developers can specialize the generated code, for each backend, for their DSL concepts. This opportunity
-made it possible to define, as a library, the HTML templating DSL generating code building DOM fragments on
-client-side and code building XML trees on server-side.
-
-\subsection{Lack of a data type definition DSL}
-
-On the other side of the coin, because that is not possible to define a library without defining how to translate it
-on both server-side and client-side, there is no simple way to define a library for defining libraries. This feature
-comes almost for free with approaches like GWT, Fay or Kotlin because they derive the client-side program and the
-server-side program from the initial program source (that contains data type definitions such as classes
-definitions). In our case, the client-side program and the server-side program are derived from the intermediate
-representation produced by the evaluation of the initial program. So, only concepts reified in an intermediate
-representation during this evaluation can be part of the final programs. However regular Scala’s classes definition
-are not reified in an intermediate representation (it would mean that the constructor of a type \code{A} would yield
-a value of type \code{Rep[A]}). Scala-virtualized~\cite{Moors12_Virtualized} provides such a mechanism but is 
-limited to product types and provides only projection functions, so for more complex cases users have to write some
-repetitive and boilerplate code.
-
-In summary, the compiled embedded approach may require more work to support custom data types but gives control on
-the way this data type will be encoded for each target platform (\eg{} the JavaScript code generator uses native
-arrays to encode the lists of the list manipulation DSL).
-
-However, the next release of Scala will have support for macros making it possible to generate reified data types
-from a concise syntax without relying on a compiler extension.
+sites\footnote{\href{http://trends.builtwith.com/javascript}{http://trends.builtwith.com/javascript}} aims to
+simplify RIA development. Unsurprisingly, it supports, at least partially, the RIA specific features of the table but
+suffers from the JavaScript language limitations. jQuery handles the \code{null} references problem by wrapping each
+query result in a container so before each further method call it tests the emptiness of the container and applies
+effectively the operation only if the container is not empty. It provides an expressive API but the emptiness
+checking involves a slight performance penalty since each result is wrapped and then unwrapped for the subsequent
+method invocation.
+
+TypeScript is a language compiling to JavaScript supporting classes, modules and soft typing. Because TypeScript is
+a superset of JavaScript, any valid JavaScript program is a valid TypeScript program. Developers can progressively
+leverage TypeScript’s features to turn their JavaScript sources into large, modular and more robust code bases.
+However, TypeScript does not address problems specific to RIAs development like DOM fragments definition or
+asynchronous programming.
+
+The Roy programming language is a statically typed functional programming language compiling to JavaScript. It
+features global type inference, algebraic data types, pattern matching and a sequencing notation similar to Scala’s
+\code{for} notation.
+
+GWT exposes the browser’s APIs through type safe Java APIs and simplifies DOM fragments definition. However it still
+uses a callback-driven programming models for asynchronous computations.
+
+Kotlin is a recent language written by JetBrains, that target both the JVM and JavaScript. It has libraries
+addressing, at least partially, Web programming concerns and supports advanced programming language features such as
+mixins, type inference, variance annotations for type parameters and pattern matching.
+
+Opa is a language specifically designed for Web programming.
+
+Dart is another language designed specifically for Web programming.
 
@@
 
-Writing RIAs requires large JavaScript code bases, however this language has several inadequacies for this purpose.
+Writing RIAs is challenging because the JavaScript language and the Web browser’s native APIs have several
+inadequacies for this purpose.
+
+We provide a way to write statically typed client-side code, we simplify asynchronous programming
 

### 2012-12-08 papers/ecoop2013/repairing.tex
Remove acknowledgments
Proofreading. Just minor grammatical fixes.

@@
 efficient JavaScript code using native APIs. We use both type-level information and staging to introduce high-level
-abstractions with a few or no overhead on the generated JavaScript code.
+abstractions with few or no overhead on the generated JavaScript code.
 \end{abstract}
@@
 nature of the client-side and server-side environments~\cite{Echeverria09_RIA,Kuuskeri09_PartitioningClientServer}:
-writing distributed code requires its parts to be consistent together and leads to duplication unless you use the
+writing distributed code requires its parts to be consistent together and leads to duplication unless one uses the
 same language in both sides, which, in the case of the Web, means using JavaScript to write the whole application.
@@
 An increasing number of initiatives attempt to allow developers to write the client-side code in a language different
-of JavaScript but that can be compiled to JavaScript (\eg{} GWT~\cite{Chaganti07_GWT}, Dart~\cite{Griffith11_Dart},
+than JavaScript but that can be compiled to JavaScript (\eg{} GWT~\cite{Chaganti07_GWT}, Dart~\cite{Griffith11_Dart},
 TypeScript\footnote{\href{http://www.typescriptlang.org/}{http://www.typescriptlang.org/}}, Roy~\cite{McKenna_Roy}).
@@
 usually more suitable to write large applications either because they provide more constructs to build abstractions
-(such as object orientation), or because they are statically typed, or because they add a bunch of useful concepts
+(such as object orientation), or because they are statically typed, or because they add some useful concepts
 that are missing in JavaScript (such as \emph{adhoc} polymorphism or namespaces). Moreover, their ability to compile
@@
 
-RIAs distinguish from classic Web pages by a higher degree of interactivity. In our example we want to let users zoom
+RIAs distinguish themselves from classic Web pages by a higher degree of interactivity. In our example we want to let users zoom
 in and out a map using their mouse wheel. To implement this feature we need to attach an event handler for the
@@
 Updating the user interface usually means replacing a part of the DOM with another DOM fragment computed from data
-fetched by an AJAX request. This requires to write how to compute the new DOM fragment in JavaScript. For instance
+fetched by an AJAX request. This requires writing how to compute the new DOM fragment in JavaScript. For instance
 the following listing shows how to build a DOM tree representing a vertex in the mind map:
@@
 Callback-driven APIs are very common in JavaScript but their inverted control makes the code flow hard to follow and
-bothers modularization. Furthermore, native APIs are too much verbose, making the code hard to read.
+bothers modularization. Furthermore, native APIs are too verbose, making the code hard to read.
 
@@
 parameter. Then, the \code{for} expression allows us to use the vertex, when available, and to insert it on the user
-interface. By making the \code{createVertex} function returning an asynchronous value instead of taking a callback as
+interface. By making the \code{createVertex} function return an asynchronous value instead of taking a callback as
 a parameter, the code is easier to modularize into loosely coupled parts.
@@
 An asynchronous value of type \code{Rep[A]} is modelled by a value of type \code{Rep[Future[A]]}. Because Scala’s
-\code{for} notation is just a syntactic sugar for methods \code{foreach}, \code{map} and \code{flatMap}, we are
+\code{for} notation is just syntactic sugar for methods \code{foreach}, \code{map} and \code{flatMap}, we are
 able to define our DSL by just defining these methods on \code{Rep[Future[A]]} values, with the following semantic:
@@
 
-In this section we show how we can define, as an embedded DSL, with a small effort, a template engine that is
+In this section we show how we can define, as an embedded DSL, with minimal effort, a template engine that is
 statically typed, able to insert dynamic content in a safe way, that provides a powerful expression language,
@@
 DOM fragments definitions are written using functions so they compose just as functions compose. These benefits come
-with no performance loss because the DSL generates code building DOM fragments using the native JavaScript API.
+with no performance loss because the DSL generates code building DOM fragments by using the native JavaScript API.
 
@@
 
-Our DSL is equivalent as a template engine with Scala as the expression language. Making it usable on both server and
+Our DSL is equivalent to a template engine with Scala as the expression language. Making it usable on both server and
 client sides is surprisingly as simple as defining another code generator for the DSL, producing Scala code.
@@
 This code suffers from two problems: the variable \code{x} is global and if the user calls the \code{inc} function
-without initializing first the \code{x} variable the behavior will be unexpected.
+without first initializing the \code{x} variable the behavior will be unexpected.
 
@@
 
-The State monad~\cite{Wadler92_StateM} is a functional programming pattern that avoids to explicitly thread the state
+The State monad~\cite{Wadler92_StateM} is a functional programming pattern that avoids explicitly threading the state
 in functions using it or modifying it. In Scala, the \code{for} notation provides an elegant way to sequence monadic
@@
 \emph{adhoc} polymorphism on JavaScript objects consists in defining the polymorphic function on the prototypes of
-the objects. However, modifying existing object prototypes is considered to be a bad
-practice~\cite{Zakas12_MaintainableJs}. Another way could consist in manually code the dispatch logic, by registering
+the objects. However, modifying existing object prototypes is considered bad
+practice~\cite{Zakas12_MaintainableJs}. Another way could consist in manually coding the dispatch logic, by registering
 supported data types at the beginning of the program execution, as described in section 2.4.3
@@
 
-Kotlin is a recent language written by JetBrains, that target both the JVM and JavaScript. It has libraries
+Kotlin is a recent language written by JetBrains, that targets both the JVM and JavaScript. It has libraries
 addressing, at least partially, Web programming concerns and supports advanced programming language features such as
@@
 
-\section{Acknowledgements}
-
-We thank Tiark Rompf for his precious feedback.
-
 \bibliographystyle{ieeetr}

### 2012-12-10 papers/ecoop2013/repairing.tex
Ready for j-m

@@
 JavaScript comes with only one numeric type, \code{Number}, that represents double-precision 64-bit values. If
-developers want to use another numeric data type, \eg{} representing integer values or arbitrary length values, they
-can’t use the native arithmetic operators (\code{+}, \code{-}, \etc) on them, making the code less readable because
-the same concepts (\eg{} adding two numeric values) are described by different operators or methods. Furthermore, the
-code is also harder to write when it comes to mix different types of numeric values. Indeed, this requires to
-manually take care of their compatibility: before adding an arbitrary length value with a \code{Number} value,
-developers have to convert the latter in the former’s type.
+developers want to use another numeric data type, \eg{} representing rational numbers, they can’t use the native
+arithmetic operators (\code{+}, \code{-}, \etc) on them, making the code less readable because the same concepts
+(\eg{} adding two numeric values) are described by different operators or methods. Furthermore, the code is also
+harder to write when it comes to mix different types of numeric values. Indeed, this requires to manually take care
+of their compatibility: before adding a rational number with a \code{Number} value, developers have to convert the
+former in the latter’s type. Consider as an example the listing \ref{type-coercion} mixing rational numbers and
+native numbers.
+
+\begin{lstlisting}[language=JavaScript,label=type-coercion,caption=Mixing integer and floating-point arithmetic in
+JavaScript.]
+// i and j are rational numbers, x and y are native numbers
+var main = function (i, j, x, y) {
+  console.log(rat_plus(i, j));
+  var m = rat_toNumber(i) + y;
+  var n = x + rat_toNumber(j);
+  console.log(m + n);
+};
+\end{lstlisting}
 
@@
 
-The following listing demonstrates the homogeneous syntax to perform an addition on different data types and the
-operand coercion between heterogeneous types:
-
-\begin{lstlisting}
-def main(i: Rep[Int], j: Rep[Int],
+The listing \ref{type-coercion} can be rewritten as shown in listing \ref{type-coercion-solution} with our DSL.
+The \code{+} operator can be safely used to add rational and floating-point numbers and type promotion is
+automatically handled by the type system. It is worth noting that the \code{Rational} data type can be written
+independently of the type coercion system.
+
+\begin{lstlisting}[label=type-coercion-solution,caption=Type-coercion automatically handled by our DSL]
+def main(i: Rep[Rational], j: Rep[Rational],
          x: Rep[Double], y: Rep[Double]) = {
@@
 
-(TODO Example)
-
 On the other side of the coin, because that is not possible to define a library without defining how to translate it
@@
 limited to record types and provides only projection functions, so for more complex cases users have to write some
-repetitive and boilerplate code.
+repetitive and boilerplate code. (FIXME Example?)
 
@@
 However, the next release of Scala will have support for macros making it possible to generate reified data types
-from a concise syntax without relying on a compiler extension.
+from a concise syntax without even relying on a compiler extension such as Scala-virtualized.
 
@@
 
-Our DSL aims to give developers a safer and more expressive way to write the client-side code. We focused on the main
-tasks of RIAs development: handling events, asynchronous programming and DOM fragments definition, as well as on
-more general programming concerns: \code{null} references handling and type safety.
+Our goal was to define, with a minimal effort, a language for RIAs development. The main challange was to provide
+language constructs both safer and more expressive without abstraction penalty. We focused on the main concerns of
+RIA development: handling events, asynchronous programming and DOM fragments definition, as well as on more general
+programming concerns: \code{null} references handling and type safety.
 
@@
 
-Using our DSL, we were able to divide by two the length of some parts of code (DOM definition and \code{null}
-reference handling). The size of code parts dealing with events did not change but they gain type safety. Finally,
-code handling asynchronous computations were easier to modularize.
+Using our DSL, we were able to completely rewrite the client-side part of our application, and divided by two the
+length of some parts of code (DOM definition and \code{null} reference handling). The size of code parts dealing with
+events did not change but gained type safety. Finally, code handling asynchronous computations were easier to
+modularize. An interesting side-effect of the design of our DSL as a compiled embedded DSL is the ability to reuse
+some code between both client and server sides (DOM definition).
 
@@
 \hline
-Event handling & * & & & *** & & ? & & *** \\
+Type safe event handling & & & & *** & & & & *** \\
 \hline
@@
 \hline
-Asynchronous programming & ** & & *** & & & *** & * & *** \\
+Asynchronous programming & ** & & *** & & & *** & ** & *** \\
 \hline
-Null references & ** & & *** & & * & ? & & *** \\
+Null references & ** & & *** & & * & ** & & *** \\
 \hline
-Type coercion & & & & * & * & ? & * & *** \\
+Type coercion & & & & * & * & * & * & *** \\
 \hline
-General programming language features & * & ** & ** & * & *** & ? & * & *** \\
-\hline
+%General programming language features & * & ** & ** & * & *** & ? & * & *** \\
+%\hline
 Static typing & & * & ** & ** & ** & ** & * & ** \\
@@
 \hline
-Client-server code sharing & & * & * & * & * & ** & * & *** \\
+Client-server code sharing & & * & * & * & * & *  & * & *** \\
 \hline
@@
 
-GWT exposes the browser’s APIs through type safe Java APIs and simplifies DOM fragments definition. However it still
-uses a callback-driven programming models for asynchronous computations.
+GWT compiles Java code to JavaScript. It exposes the browser’s APIs through type safe Java APIs and simplifies DOM
+fragments definition. However, it still uses a callback-driven programming models for asynchronous computations.
 
@@
 
-Opa is a language specifically designed for Web programming.
-
-Dart is another language designed specifically for Web programming.
+Opa is a language and a full stack designed for Web programming. It features static typing with global
+type inference, modules and pattern matching. The language provides a syntactic sugar for writing HTML fragments in
+the code. The server and client parts of an application can be written in the Opa language, the system can decide
+whether a code fragment will run on server-side, client-side or both.
+
+Dart is a language designed specifically for Web programming. It has been designed to have a familiar syntax for
+JavaScript developers and supports soft typing, classes and a module system. It also provides a high-level concurrent
+programming system similar to an actor system. Finally, Dart code can be executed on both client and server sides.
+
+Concerning the client-server code sharing feature, because our solution is the only one that gives full control on
+the generated code on both client and server for a given piece of shared code, we considered that our support was
+better than other approaches.
 
@@
 Writing RIAs is challenging because the JavaScript language and the Web browser’s native APIs have several
-inadequacies for this purpose.
-
-We provide a way to write statically typed client-side code, we simplify asynchronous programming
-
-We provide a safer and more expressive language than JavaScript and its native APIs to write RIAs.
-
-More code sharing.
+inadequacies, making the code fragile, hard to reason about and to maintain.
+
+We improved the js-scala DSL by adding first-class support for RIA development concerns such as events handling,
+DOM fragments definition and asynchronous programming. Our work provides a safer and more expressive language than
+JavaScript and its native APIs to write RIAs.
+
+Our approach’s distinguishing support of code sharing between server and clients let us consider the exploration of
+this feature more deeply in our future work, in order to factor out an increasing part of code between client and
+servers in RIAs.
 

### 2012-12-13 papers/ecoop2013/repairing.tex
Implementation section draft

@@
 
-\title{Statically typed Web programming}
-
-\author{Julien \noun{Richard-Foy}\textsuperscript{\textasteriskcentered} \and Olivier \noun{Barais}\textsuperscript{\textasteriskcentered} \and
-Tiark \noun{Rompf}\textsuperscript{\textdagger} \and Jean-Marc \noun{Jézéquel}\textsuperscript{\textasteriskcentered}}
-\institute{\textsuperscript{\textasteriskcentered}IRISA, Université de Rennes 1, France: \texttt{\{first\}.\{last\}@irisa.fr}\\
+\title{Statically Typed Web Programming using Scala}
+
+\author{Julien \noun{Richard-Foy}\textsuperscript{\textasteriskcentered}\textsuperscript{\textdaggerdbl} \and Olivier
+\noun{Barais}\textsuperscript{\textasteriskcentered} \and Tiark \noun{Rompf}\textsuperscript{\textdagger} \and
+Jean-Marc \noun{Jézéquel}\textsuperscript{\textasteriskcentered}}
+\institute{\textsuperscript{\textasteriskcentered}IRISA, Université de Rennes 1, France:
+\texttt{\{first\}.\{last\}@irisa.fr}\\
+\textsuperscript{\textdaggerdbl}Zenexity, Paris, France: \texttt{jrf@zenexity.com}\\
 \textsuperscript{\textdagger}EPFL, Lausanne, Switzerland: \texttt{\{first\}.\{last\}@epfl.ch}}
@@
 \begin{abstract}
-Rich Internet Applications involve more code on the client-side, dealing with DOM manipulation, event handling and
-asynchronous calls to the server. Writing and maintaining large JavaScript code bases is challenging because this
-language has several inadequacies (no static typing, no module system, verbose syntax, \etc{}). We investigate
-further the definition of JavaScript as a Scala compiled embedded DSL using Lightweight Modular Staging. We define
-safe and expressive abstractions for DOM manipulation, event handling and asynchronous programming that translate to
-efficient JavaScript code using native APIs. We use both type-level information and staging to introduce high-level
-abstractions with few or no overhead on the generated JavaScript code.
+Rich Internet Applications involve even more code on the client-side, dealing with DOM manipulation, event handling
+and asynchronous calls to the server. Writing and maintaining large JavaScript code bases is challenging because this
+language has no static typing and no module system. We investigate further the definition of JavaScript as a Scala
+compiled embedded DSL using Lightweight Modular Staging. We define safe and expressive abstractions for DOM
+manipulation, event handling and asynchronous programming that translate to efficient JavaScript code using native
+APIs. We use both type-level information and staging to introduce high-level abstractions with few or no overhead on
+the generated JavaScript code.
 \end{abstract}
@@
 nature of the client-side and server-side environments~\cite{Echeverria09_RIA,Kuuskeri09_PartitioningClientServer}:
-writing distributed code requires its parts to be consistent together and leads to duplication unless one uses the
-same language in both sides, which, in the case of the Web, means using JavaScript to write the whole application.
-However this language has several inadequacies making it hardly suitable for large code base (\eg{} no static typing,
-no module system, verbose syntax, \etc). Some other in-browser execution environments give the opportunity to write
-the client-side code in another language than JavaScript, \eg{} Java applets~\cite{Anuff96_JavaApplets}, Adobe
-Flash~\cite{Curtis00_Flash} and Microsoft
+writing distributed code requires its different parts to be consistent and leads to code duplication unless one uses
+the same language in both sides, which, in the case of the Web, means using JavaScript to write the whole
+application. However this language has several drawbacks making it hardly suitable for large code base (\eg{} no
+static typing, no module system, verbose syntax, \etc). Some other in-browser execution environments give the
+opportunity to write the client-side code in another language that JavaScript, \eg{} Java
+applets~\cite{Anuff96_JavaApplets}, Adobe Flash~\cite{Curtis00_Flash} and Microsoft
 Silverlight\footnote{\href{http://www.microsoft.com/silverlight/}{http://www.microsoft.com/silverlight/}}. However
-these technologies have several drawbacks: they require an additional browser plugin to be installed (which may not
-be available on all devices having a Web browser: for instance there is no way to execute Flash objects within an
+these technologies also have several drawbacks: they require an additional browser plugin to be installed (which may
+not be available on all devices having a Web browser: for instance there is no way to execute Flash objects within an
 Apple smart phone), the page content can't naturally be referenced by search engines, and the content is not
@@
 for both server and client sides usually allows developers to share some parts of code between server and client
-sides for free. However this shared code is restricted to use exclusively language constructs: concepts provided by
+sides for free. However this shared code is restricted to exclusively use language constructs: concepts provided by
 external APIs can't be shared between server and client sides because the bindings with these environments are
@@
 
-In this paper we investigate further a path already introduced by Kossakowski \etal{}: defining JavaScript as a
-compiled embedded DSL in Scala~\cite{Kossakowski12_JsDESL}. This approach is based on Lightweight Modular
-Staging~\cite{Rompf12_LMSThesis}: a Scala program written using the embedded DSL evaluates to an intermediate
+In this paper we investigate further \emph{js-scala}, a path already introduced by Kossakowski \etal{}: defining
+JavaScript as a compiled embedded DSL in Scala~\cite{Kossakowski12_JsDESL}. This approach is based on Lightweight
+Modular Staging~\cite{Rompf12_LMSThesis}: a Scala program written using the embedded DSL evaluates to an intermediate
 representation that can be further processed to perform domain specific optimizations and then to generate a
@@
 
-The previous work showed how to write client-side code in Scala using the JavaScript embedded DSL and how to share
-code between server-side and client-side. The authors also showed that the code was safer and more convenient to
-write thanks to Scala's static typing and class system. Finally, they showed how to \emph{un-invert} the control of
-callback-based asynchronous APIs by using continuations but they did not address other concerns of RIAs development
-such as user interface events handling, DOM creation and manipulation, and application’s state management.
+This previous work on js-scala showed how to write client-side code in Scala using the JavaScript embedded DSL and
+how to share code between server-side and client-side. The authors also showed that the code was safer and more
+convenient to write thanks to Scala's static typing and class system. Finally, they showed how to \emph{un-invert}
+the control of callback-based asynchronous APIs by using continuations but they did not address other concerns of
+RIAs development such as user interface events handling and DOM creation and manipulation.
 
@@
 make asynchronous programming simpler by raising asynchronous computations to first-class citizen. We leverage the
-Scala’s type system to bring type-directed language features such as \emph{adhoc} polymorphism and type coercion to
+Scala type system to bring type-directed language features such as \emph{ad hoc} polymorphism and type coercion to
 the client-side.
@@
 The next section introduces a simple application example showing challenges of RIA development, section
-\ref{solution} presents our DSL tackling these challenges, section \ref{validation} validates our solution and
-compares it to related work and section \ref{discussion} concludes.
+\ref{solution} presents our DSL tackling these challenges, section \ref{validation} validates our solution, section
+\ref{related} presents related works and compares it to related work and section \ref{discussion} concludes.
 
@@
 
-\section{Background and motivating example}
+\section{Background and Motivating Example}
 \label{problem}
@@
 To illustrate the challenges of writing a RIA and the need for a DSL for writing them, we introduce a simple “mind
-mapping” application and show how the client-side code would be typically written in JavaScript.
-
-\subsection{Example description}
+mapping” application and show how the client-side code would typically be written in JavaScript.
+
+\subsection{Example Description}
 
@@
 relation between two ideas is represented by an edge. Users can visualize a mind map, zoom in and out and move it.
-They can also add ideas and link them together. The figure \ref{mindmap} shows the visualization of a simple map
+They can also add ideas and link them together. Figure \ref{mindmap} shows the visualization of a simple map
 consisting of three ideas linked together, using our application.
@@
 
-\subsubsection{Events handling}
+\subsubsection{Event handling}
 
@@
 
-We think this code is fragile for two reasons. (1) The name of the event is passed as a string, so it is easy to
+We claim this code is fragile for two reasons. (1) The name of the event is passed as a string, so it is easy to
 mispell it. (2) The callback passed as a second parameter takes a parameter \code{e} whose fields vary according to
@@
 Closure Templates\footnote{\href{https://developers.google.com/closure/templates/}{https://developers.google.com/closure/templates/}}
-aim to make simpler the definition of DOM fragments by providing a convenient syntax to describe the HTML structure
-and allowing the insertion of dynamic expressions in a safe way. Some template engines can be used on both
+aim to make it simpler the definition of DOM fragments by providing a convenient syntax to describe the HTML
+structure and allowing the insertion of dynamic expressions in a safe way. Some template engines can be used on both
 client-side and server-side, however none has a practical expression language (dynamic content can only be provided
@@
 
-\subsubsection{Lack of static checks}
+\subsubsection{Lack of static hecks}
 
@@
 and robust code bases and to perform refactorings. For instance, in the case of RIAs, a mispell in an event name may
-break the program behavior without getting a sensible error message.
+break the program behavior without anyone getting a sensible error message.
 
@@
 Callback-driven APIs are very common in JavaScript but their inverted control makes the code flow hard to follow and
-bothers modularization. Furthermore, native APIs are too verbose, making the code hard to read.
+hampers modularization. Furthermore, native APIs are too verbose, making the code hard to read.
 
@@
 
-\section{DSL design and implementation}
+\section{A DSL for Web Programming}
 \label{solution}
@@
 \label{intro-lms}
+
+\subsubsection{Overview}
 
@@
 
-The remaining of this section describe the design and the implementation of DSLs for Web programming using LMS.
-
-\subsection{Events handling}
-
-We want to bring safety to the events handling API so that developers can’t mispell an event name and can’t attempt
+\subsubsection{Example: null references}
+
+As an illustration of the staging mechanism, we present a simple DSL to handle null references. This DSL provides an
+abstraction at the stage-level that is removed by optimization during the code generation.
+
+Null references are a known source of problems in programming languages~\cite{Hoare09_Null,Nanda09_Null}. For
+example, consider the following typical JavaScript code finding a particular widget in the page and a then particular
+button in the widget:
+
+\begin{lstlisting}[language=JavaScript,label=null-unsafe,caption=Unsafe code]
+var loginWidget = document.querySelector("div.login");
+var loginButton = loginWidget.querySelector("button.submit");
+loginButton.addEventListener("click", function (e) { ... });
+\end{lstlisting}
+
+The native \code{querySelector} method returns \code{null} if no node matched the given selector in the document. If
+we run the above code in a page where the widget is not present, it will throw an error and stop further JavaScript
+execution. We can write defensive code to handle \code{null} cases, but it leads to very cumbersome code:
+
+\begin{lstlisting}[language=JavaScript,label=null-defensive,caption=Defensive programming to handle null references]
+var loginWidget = document.querySelector("div.login");
+if (loginWidget !== null) {
+  var loginButton = loginWidget.querySelector("button.submit");
+  if (loginButton !== null) {
+    loginButton.addEventListener("click", function (e) { ... });
+  }
+}
+\end{lstlisting}
+
+We want to define a DSL that has both the safety and performance of listing \ref{null-defensive} but the
+expressiveness of listing \ref{null-unsafe}. We can get safety by wrapping potentially null values of type
+\code{Rep[A]} in a container of type \code{Rep[Option[A]]} requiring explicit dereferencing, we can get
+expressiveness by using the Scala \code{for} notation for dereferencing, and finally we can get performance by
+generating code that does not actually wraps values in a container but instead checks if they are \code{null} or not
+when dereferenced. The wrapping container exists only at the stage-level and is removed during the code generation.
+Here is a Scala listing that uses our DSL (implementation details are given in section \ref{implementation}):
+
+\begin{lstlisting}
+for {
+  loginWidget <- document.find("div.login")
+  loginButton <- loginWidget.find("submit.button")
+} loginButton.on(Click) { e => ... }
+\end{lstlisting}
+
+The evaluation of the above listing produces a graph of statements from which JavaScript code equivalent to
+listing \ref{null-defensive} is generated.
+
+The remaining of this section describes the design and the implementation of DSLs for Web programming using LMS.
+
+\subsection{Event Handling}
+
+We want to bring safety to the event handling API so that developers cannot mispell an event name and cannot attempt
 to read a property that is not defined on the type of the handled event. The difficulty comes from the fact that the
@@
 However, implementing this solution requires a high effort because we have to define as many methods as there are
-events. We want to provide a single polymorphic method similar to the native API. To achieve that we need to encode
-at the type-level the dependency relation between an event name and its corresponding event type. This dependency can
+events. We want to provide a single polymorphic method similar to the native API. To achieve that, we need to encode
+the dependency relation between an event name and its corresponding event type at the type-level. This dependency can
 naturally and elegantly be encoded using dependent method types~\cite{Oliveira10_Typeclasses}:
@@
 
-\subsection{Asynchronous programming}
+\subsection{Asynchronous Programming}
+
+We propose a DSL that explicitly reflects the asynchronous nature of computations in their types and provides methods
+turning them into first-class citizens. The DSL is monadic so we can solve the callback-hell problem thanks to the
+Scala \code{for} notation.
+
+For instance, listing \ref{async-first-class} shows how the listing \ref{async-js} can be
+re-written with our DSL. The \code{createVertex} now returns an asynchronous value instead of taking a callback as
+parameter. Then, the \code{for} expression allows us to use the vertex, when available, and to insert it on the user
+interface. By making the \code{createVertex} function return an asynchronous value instead of taking a callback as
+a parameter, the code is easier to modularize into loosely coupled parts.
 
@@
 \end{figure}
-
-We propose a DSL that explicitly reflects the asynchronous nature of computations in their type and provides methods
-turning them into first-class citizen. The DSL is monadic so we can solve the callback-hell problem thanks to the
-Scala \code{for} notation.
-
-For instance, listing \ref{async-first-class} shows how the listing \ref{async-js} can be
-re-written with our DSL. The \code{createVertex} now returns an asynchronous value instead of taking a callback as
-parameter. Then, the \code{for} expression allows us to use the vertex, when available, and to insert it on the user
-interface. By making the \code{createVertex} function return an asynchronous value instead of taking a callback as
-a parameter, the code is easier to modularize into loosely coupled parts.
 
@@
 \item \code{map(f: Rep[A] => Rep[B]): Rep[Future[B]]}, eventually transforms the value when available~;
-\item \code{flatMap(f: Rep[A] => Rep[Future[B]]): Rep[Future[B]]} eventually transforms the value when available~;
+\item \code{flatMap(f: Rep[A] => Rep[Future[B]]): Rep[Future[B]]} also eventually transforms the value when available~;
 \end{itemize}
@@
 Asynchronous values can be transformed using the \code{map} and \code{flatMap} methods, turning them into first-class
-citizen: functions can take as parameters and return asynchronous values.
+citizens: functions can take them as parameters and return them.
 
@@
 
-Listing \ref{async-parallel-1} and \ref{async-parallel-2} define two asynchronous computations running in parallel
-and the generated JavaScript code (we renamed some identifiers for the sake of readability). The generated
-\code{Promise} constructor code has been omitted, it creates an object holding a list of callbacks to call when the
-asynchronous value is completed. The \code{AjaxGet} function code has also been omitted, it creates a
+Listing \ref{async-parallel-1} defines two asynchronous computations running in parallel and listing
+\ref{async-parallel-2} shows the generated JavaScript code (we renamed some identifiers for the sake of readability).
+The generated \code{Promise} constructor code has been omitted, it creates an object holding a list of callbacks to
+call when the asynchronous value is completed. The \code{AjaxGet} function code has also been omitted, it creates a
 \code{XMLHttpRequest} object that sends an HTTP request and completes its promise parameter with the response, when
@@
 
-% \begin{lstlisting}[caption=Concrete syntax of the asynchronous programming DSL]
-% trait FutureOps { this: Base =>
-% 
-%   implicit def FutureOpsCls[A](f: Rep[Future[A]]): FutureOpsCls[A]
-%   type FutureOpsCls[+A] <: FutureOpsBase[A]
-% 
-%   trait FutureOpsBase[+A] {
-%     def map[B](f: Rep[A] => Rep[B]): Rep[Future[B]]
-%     def flatMap[B](f: Rep[A] => Rep[Future[B]]): Rep[Future[B]]
-%     def foreach(f: Rep[A] => Rep[Unit]): Rep[Unit]
-%     def withFilter(f: Rep[A] => Rep[Boolean]): Rep[Future[A]]
-%   }
-% }
-% \end{lstlisting}
-% 
-% The \code{FutureOps} module depends on the \code{Base} module defined by LMS.
-The \code{Base} module essentially defines the abstract type constructor \code{Rep[_]}.
-Values of type \code{Rep[Future[A]]} have no methods because the \code{Rep[_]} type is abstract and unbound, so we use an implicit conversion from \code{Rep[Future[A]]} to \code{FutureOpsCls[A]} to add methods to \code{Rep[Future[A]]} values.
-The type \code{FutureOpsCls[A]} is bound by the \code{FutureOpsBase[A]} trait that defines the signature of the methods added to \code{Rep[Future[A]]} values.
-Everything in the \code{FutureOps} module is abstract, these definitions only provide the vocabulary of the DSL.
-% 
-% The listing \ref{futureopsexp} shows the definition of the \code{FutureOpsExp} module that implements the \code{FutureOps} module and defines how statement nodes are built from the DSL constructs.
-% 
-% \begin{figure}
-% \begin{lstlisting}[label=futureopsexp,caption=Definition of the intermediate representation of the asynchronous programming DSL]
-% trait FutureOpsExp extends FutureOps { this: EffectExp =>
-% 
-%   implicit class FutureOpsCls[+A](a: Exp[Future[A]])
-%       extends FutureOpsBase[A] {
-%     def map[B](f: Exp[A] => Exp[B]) = {
-%       val x = fresh[A]
-%       val b = reifyEffects(f(x))
-%       reflectEffect(FutureMap(a, x, b), summarizeEffects(b).star)
-%     }
-% 
-%     // ... same for other methods (flatMap, foreach and withFilter)
-%   }
-%   
-%   case class FutureMap[A, B](a: Exp[Future[A]], x: Sym[A], b: Block[B])
-%       extends Def[Future[B]]
-%   // ... same for other operations
-% }
-% \end{lstlisting}
-% \end{figure}
-% 
-% The \code{FutureOpsExp} module depends on the \code{EffectExp} module that is a module refining the \code{Base} module, defining the type \code{Rep[A]} as \code{Exp[A]}, which is a base class defining intermediate representations as a graph of statements.
-The \code{EffectExp} module also provides tools to handle effectful statements.
-The implicit class \code{FutureOpsCls} implements both the abstract type and the implicit conversion of the same name defined in the \code{FutureOps} trait.
-Each method of the \code{FutureOpsCls} class is reified to a statement node, for instance the \code{map} method returns a \code{FutureMap} node containing a reference to the node on which the \code{map} operation is applied and the code block that was passed as a parameter to the \code{map} operation.
-We propagate the information about the effects performed by the statements in the code block by calling \code{reflectEffect} on our statement node.
-% 
-% The listing \ref{jsgenfutureops} shows the code of the JavaScript code generator for the DSL.
-The code generation process basically consists in traversing the graph of statement nodes and in emitting the code corresponding to each node.
-To implement a code generator for a given DSL we just need to say how to generate code for this DSL statement nodes.
-% 
-% \begin{figure}
-%  \begin{lstlisting}[label=jsgenfutureops,caption=JavaScript code generator for the asynchronous programming DSL]
-% trait JSGenFutureOps extends JSNestedCodegen {
-%   val IR: EffectExp with FutureOpsExp
-%   import IR._
-% 
-%   override def emitNode(sym: Sym[Any], rhs: Def[Any]) = rhs match {
-%     case FutureMap(a, x, b) =>
-%       emitValDef(sym, "new Promise()")
-%       stream.println(quote(a) + ".onComplete(function (" + quote(x) + ") {")
-%       emitBlock(b)
-%       stream.println(quote(sym) + ".complete(" + quote(getBlockResult(b)) + ");")
-%       stream.println("});")
-%     // ... same for each node
-%     case _ => super.emitNode(sym, rhs)
-%   }
-% }
-%  \end{lstlisting}
-% \end{figure}
-% 
-% The \code{JSGenFutureOps} code generator inherits from the \code{JSNestedCodegen} code generator that defines the building blocks of the JavaScript code generation.
-The main method is \code{emitNode}, it is called by LMS for each traversed node of the statement graph.
-In the listing we only showed the code generation for the \code{FutureMap} node: when the asynchronous value on which the \code{map} method has been applied is available, we apply the function passed to the \code{map} method on the value and complete a newly created asynchronous value with the result.
-The \code{Promise} constructor is provided by a JavaScript preamble and creates an asynchronous value with two methods: \code{onComplete}, registering a callback to be called when the value is available, and \code{complete}, fulfilling the promise with a value.
-
-\subsection{DOM definition}
-
-In this section we show how we can define, as an embedded DSL, with minimal effort, a template engine that is
-statically typed, able to insert dynamic content in a safe way, that provides a powerful expression language,
-requires no extra compilation step and that can be used on both client-side and server-side.
+\subsection{DOM Definition}
+
+In this section we show how we can define a template engine as an embedded DSL, with minimal effort. This template
+engine is statically typed and able to insert dynamic content in a safe way. It provides a powerful expression
+language, requires no extra compilation step and can be used on both client-side and server-side.
 
@@
 function to define a tag and a \code{text} function to define a text node. The following listing uses our DSL and
-generates a code equivalent to the listing \ref{dom-api}:
+generates a code equivalent to listing \ref{dom-api}:
 
@@
 
-For instance, the template written in the listing \ref{forest-hello} produces the following Scala code usable on
-server-side (the generated code for client-side is roughly equivalent to the listing \ref{dom-api}):
+For instance, the template written in listing \ref{forest-hello} produces the following Scala code usable on
+server-side (the generated code for client-side is roughly equivalent to listing \ref{dom-api}):
 
@@
 We are able to tackle the code sharing issues described in section \ref{assessment} because of the embdedded nature
-of our DSLs: dynamic content of templates is written using embdedded DSLs too, so their translation into JavaScript
+of our DSLs: dynamic content of templates is written using embedded DSLs too, so their translation into JavaScript
 and Scala is managed by their respective code generators.
 
-\subsection{More general purpose improvements of JavaScript}
-
-This section presents some more general purpose improvements of JavaScript.
-
-\subsubsection{Null references}
-
-Null references are a known source of problems in programming languages~\cite{Hoare09_Null,Nanda09_Null}. For
-example, consider the following typical code finding a particular widget in the page and a then particular button in
-the widget:
-
-\begin{lstlisting}[language=JavaScript,label=null-unsafe,caption=Unsafe code]
-var loginWidget = document.querySelector("div.login");
-var loginButton = loginWidget.querySelector("button.submit");
-loginButton.addEventListener("click", function (e) { ... });
-\end{lstlisting}
-
-The native \code{querySelector} method returns \code{null} if no node matched the given selector in the document. If
-we run the above code in a page where the widget is not present, it will throw an error and stop further JavaScript
-execution. We can write defensive code to handle \code{null} cases, but it leads to very cumbersome code:
-
-\begin{lstlisting}[language=JavaScript,label=null-defensive,caption=Defensive programming to handle null references]
-var loginWidget = document.querySelector("div.login");
-if (loginWidget !== null) {
-  var loginButton = loginWidget.querySelector("button.submit");
-  if (loginButton !== null) {
-    loginButton.addEventListener("click", function (e) { ... });
-  }
-}
-\end{lstlisting}
-
-Our solution has both the safety and performance of the listing \ref{null-defensive} and the expressiveness of the
-listing \ref{null-unsafe}. We encode the potential emptiness of a value of type \code{Rep[A]} using the
-\code{Rep[Option[A]]} type at the stage level but we don’t generate code wrapping values in a container, instead the
-generated code just checks if a value is \code{null} or not. Here is a listing that uses our DSL and compiles to code
-equivalent to the listing \ref{null-defensive}:
-
-\begin{lstlisting}
-for {
-  loginWidget <- document.find("div.login")
-  loginButton <- loginWidget.find("submit.button")
-} loginButton.on(Click) { e => ... }
-\end{lstlisting}
-
-\subsubsection{State monad}
-
-Managing the state of an application is cumbersome and fragile in JavaScript due to the lack of encapsulation
-constructs. For example, consider the following function incrementing a counter and returning its previous value:
-\begin{lstlisting}[language=JavaScript,label=state-fragile,caption=Fragile state handling]
-var x;
-var inc = function () {
-  var prev = x;
-  x = x + 1;
-  return prev
-};
-
-// Usage
-x = 42;
-inc();
-var y = inc();
-alert(y);
-\end{lstlisting}
-This code suffers from two problems: the variable \code{x} is global and if the user calls the \code{inc} function
-without first initializing the \code{x} variable the behavior will be unexpected.
-
-To workaround these issues, developers can wrap the variable declaration and the \code{inc} function code in a
-factory function. Thus the state is hidden and developers may not forget the initialization:
-
-\begin{lstlisting}[language=JavaScript,label=state-encapsulated,caption=State encapsulation within a function]
-var incFactory = function (init) {
-  var x = init;
-  return function () {
-    var prev = x;
-    x = x + 1;
-    return prev
-  }
-};
-
-// Usage
-var inc = incFactory(42);
-inc();
-var x = inc();
-alert(x);
-\end{lstlisting}
-
-The code is now safer, but it is also bigger, and it still uses a mutable state. It’s known that mutable objects in
-the code make this one harder to reason about~\cite{Grogono94_Immutability,Kjolstad11_Immutability}. Getting rid of
-mutable objects is simple: instead of modifying a state, a function takes its initial value as a parameter and
-returns its new value in its result:
-
-\begin{lstlisting}[language=JavaScript,label=state-explicit,caption=Explicit state threading]
-var inc = function (state) {
-  return {
-    state: state + 1,
-    value: state
-  }
-};
-
-// Usage
-var x1 = inc(42);
-var x2 = inc(x1.state);
-alert(x2.value);
-\end{lstlisting}
-
-However, explicitly threading the state in all functions parameters is not desired because of the syntactic and
-performance cost of adding an extra parameter to each function and of returning two results instead of one.
-Furthermore, manually passing the result of each previous call to the next one is error prone.
-
-The State monad~\cite{Wadler92_StateM} is a functional programming pattern that avoids explicitly threading the state
-in functions using it or modifying it. In Scala, the \code{for} notation provides an elegant way to sequence monadic
-computations, and thus to implicitly thread a state across computations, as shown in the following listing:
-
-\begin{lstlisting}[label=state-monad,caption=Implicitly threaded state using the State Monad]
-val inc = for {
-  x <- get[Int]
-  _ <- put(x + 1)
-} yield x
-
-// Usage
-val usage = for {
-  _ <- inc
-  x <- inc
-} yield alert(x)
-usage.init(42)
-\end{lstlisting}
-
-The \code{get} and \code{put} primitives are part of the State monad DSL, they allow to retrieve the current state
-and to change it. The \code{usage} value is a reusable piece of program that calls two times the \code{inc} function
-and prints the result of the second call. The last line of the listing runs the \code{usage} program on an initial
-state.
-
-The State monad programming model has the advantages but not the shortcomings of both listings
-\ref{state-encapsulated} and \ref{state-explicit} because there is no global variable (the state is implicitly
-threaded thanks to the State monad but state modifications explicitly use the \code{get} and \code{put} primitives)
-and everything is immutable.
-
-Furthermore, our implementation produces efficient imperative code using variables instead of functions passing the
-state through. To achieve that, we transform the intermediate representation of the stack of state modifications
-produced by the \code{for} sequences into an equivalent intermediate representation using regular variables. Finally,
-the listing \ref{state-monad} compiles to code equivalent to the listing \ref{state-fragile} so that there is no
-abstraction penalty.
-
-\subsubsection{\emph{Adhoc} polymorphism}
+\subsection{Leveraging the Scala Type System on the Client-Side}
+
+This section presents some more general purpose improvements of JavaScript based on type-directed mechanisms of
+Scala.
+
+% \subsubsection{State monad}
+% 
+% Managing the state of an application is cumbersome and fragile in JavaScript due to the lack of encapsulation
+% constructs. For example, consider the following function incrementing a counter and returning its previous value:
+% \begin{lstlisting}[language=JavaScript,label=state-fragile,caption=Fragile state handling]
+% var x;
+% var inc = function () {
+%   var prev = x;
+%   x = x + 1;
+%   return prev
+% };
+% 
+% // Usage
+% x = 42;
+% inc();
+% var y = inc();
+% alert(y);
+% \end{lstlisting}
+% This code suffers from two problems: the variable \code{x} is global and if the user calls the \code{inc} function
+% without first initializing the \code{x} variable the behavior will be unexpected.
+% 
+% To workaround these issues, developers can wrap the variable declaration and the \code{inc} function code in a
+% factory function. Thus the state is hidden and developers may not forget the initialization:
+% 
+% \begin{lstlisting}[language=JavaScript,label=state-encapsulated,caption=State encapsulation within a function]
+% var incFactory = function (init) {
+%   var x = init;
+%   return function () {
+%     var prev = x;
+%     x = x + 1;
+%     return prev
+%   }
+% };
+% 
+% // Usage
+% var inc = incFactory(42);
+% inc();
+% var x = inc();
+% alert(x);
+% \end{lstlisting}
+% 
+% The code is now safer, but it is also bigger, and it still uses a mutable state. It’s known that mutable objects in
+% the code make this one harder to reason about~\cite{Grogono94_Immutability,Kjolstad11_Immutability}. Getting rid of
+% mutable objects is simple: instead of modifying a state, a function takes its initial value as a parameter and
+% returns its new value in its result:
+% 
+% \begin{lstlisting}[language=JavaScript,label=state-explicit,caption=Explicit state threading]
+% var inc = function (state) {
+%   return {
+%     state: state + 1,
+%     value: state
+%   }
+% };
+% 
+% // Usage
+% var x1 = inc(42);
+% var x2 = inc(x1.state);
+% alert(x2.value);
+% \end{lstlisting}
+% 
+% However, explicitly threading the state in all functions parameters is not desired because of the syntactic and
+% performance cost of adding an extra parameter to each function and of returning two results instead of one.
+% Furthermore, manually passing the result of each previous call to the next one is error prone.
+% 
+% The State monad~\cite{Wadler92_StateM} is a functional programming pattern that avoids explicitly threading the state
+% in functions using it or modifying it. In Scala, the \code{for} notation provides an elegant way to sequence monadic
+% computations, and thus to implicitly thread a state across computations, as shown in the following listing:
+% 
+% \begin{lstlisting}[label=state-monad,caption=Implicitly threaded state using the State Monad]
+% val inc = for {
+%   x <- get[Int]
+%   _ <- put(x + 1)
+% } yield x
+% 
+% // Usage
+% val usage = for {
+%   _ <- inc
+%   x <- inc
+% } yield alert(x)
+% usage.init(42)
+% \end{lstlisting}
+% 
+% The \code{get} and \code{put} primitives are part of the State monad DSL, they allow to retrieve the current state
+% and to change it. The \code{usage} value is a reusable piece of program that calls twice the \code{inc} function
+% and prints the result of the second call. The last line of the listing runs the \code{usage} program on an initial
+% state.
+% 
+% The State monad programming model has the advantages but not the shortcomings of both listings
+% \ref{state-encapsulated} and \ref{state-explicit} because there is no global variable (the state is implicitly
+% threaded thanks to the State monad but state modifications explicitly use the \code{get} and \code{put} primitives)
+% and everything is immutable.
+% 
+% Furthermore, our implementation produces efficient imperative code using variables instead of functions passing the
+% state through. To achieve that, we transform the intermediate representation of the stack of state modifications
+% produced by the \code{for} sequences into an equivalent intermediate representation using regular variables. Finally,
+% the listing \ref{state-monad} compiles to code equivalent to the listing \ref{state-fragile} so that there is no
+% abstraction penalty.
+
+\subsubsection{\emph{Ad hoc} polymorphism}
 
@@
 the prototype of the \code{foo} object for a property named \code{bar} and will call it. So, the only way to achieve
-\emph{adhoc} polymorphism on JavaScript objects consists in defining the polymorphic function on the prototypes of
+\emph{ad hoc} polymorphism on JavaScript objects consists in defining the polymorphic function on the prototypes of
 the objects. However, modifying existing object prototypes is considered bad
-practice~\cite{Zakas12_MaintainableJs}. Another way could consist in manually coding the dispatch logic, by registering
-supported data types at the beginning of the program execution, as described in section 2.4.3
+practice~\cite{Zakas12_MaintainableJs}. Another way could consist in manually coding the dispatch logic, by
+registering supported data types at the beginning of the program execution, as described in section 2.4.3
 of~\cite{Abelson83_SICP}, but this solution is painful for developers and incurs a performance overhead.
 
-We propose to achieve \emph{adhoc} polymorphism using
+We propose to achieve \emph{ad hoc} polymorphism using
 typeclasses~\cite{Wadler89_AdhocPolymorphism,Odersky06_Typeclasses,Oliveira10_Typeclasses} so that it supports
@@
 \begin{figure}
-\begin{lstlisting}[label=polymorphism,caption=Adhoc polymorphism using typeclasses]
+\begin{lstlisting}[label=polymorphism,caption=Ad hoc polymorphism using typeclasses]
 // Interface
@@
 
-The listing \ref{polymorphism} demonstrates how to define a polymorphic \code{listWidget} function that returns a DOM
+Listing \ref{polymorphism} demonstrates how to define a polymorphic \code{listWidget} function that returns a DOM
 tree containing the representation of a list of items. The \code{Show[A]} typeclass defines how to produce a DOM tree
-for a value of type \code{A}.
+for a value of type \code{A}. It is used by the \code{listWidget} function to get the DOM fragments of the list
+items. The listing shows how to reuse the same \code{listWidget} function to show a list of users and a list of
+articles.
 
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[caption=Numeric operands constraints,label=type-coercion-constraints]
 implicit def sameType[A : Numeric] =
@@
 hand side is promoted to the type \code{B}). The \code{promoteRhs} does the same for the right hand side.
-
-The \code{\~\>} type takes two parameters, saying how to obtain a value of the target type from each operand. Here is
-the definition of the \code{\~\>} type:
-
-\begin{lstlisting}
-trait Args { type Lhs; type Rhs }
-class ~>[A <: Args, B : Numeric](
-  val lhs: Rep[A#Lhs] => Rep[B],
-  val rhs: Rep[A#Rhs] => Rep[B]
-)
-\end{lstlisting}
 
@@
 
-\subsection{Implementation discussion}
+\subsection{Discussion}
 
@@
 
-First, features like type coercion and \emph{adhoc} polymorphism only rely on the type system, so they operate during
+First, features like type coercion and \emph{ad hoc} polymorphism only rely on the type system, so they operate during
 the compilation. Then, during the evaluation, the DSL constructs build an intermediate representation of the program,
-for instance the State monad DSL builds an imperative sequence of assignments on a variable. Last, the code generator
-produces the final program using the domain specific information available in the intermediate representation to
-generate efficient code, for example the Option monad code generator produces code checking against nullity before
-dereferencing a value.
+for instance the DOM definition DSL builds a tree of DOM nodes. Last, the code generator produces the final program
+using the domain specific information available in the intermediate representation to generate efficient code, for
+example the Option monad code generator produces code checking against nullity before dereferencing a value.
 
@@
 
-Compiled embedded DSLs are a good tradeoff because they are simply defined as libraries but they let developers
-specialize the generated code according to each target. This opportunity made it possible to define, as a library,
-the HTML templating DSL generating code building DOM fragments on client-side and code building XML trees on
-server-side.
+Compiled embedded DSLs have the advantages of both approaches because they are simply defined as libraries but they
+let developers specialize the generated code according to each target. This opportunity made it possible to define,
+as a library, the HTML templating DSL generating code building DOM fragments on client-side and code building XML
+trees on server-side.
 
@@
 
-\subsection{Lack of a data type definition DSL}
+\subsubsection{Lack of a data type definition DSL}
 
@@
 representation produced by the evaluation of the initial program. So, only concepts reified in an intermediate
-representation during this evaluation can be part of the final programs. However regular Scala’s classes definition
+representation during this evaluation can be part of the final programs. However regular Scala’s classes definitions
 are not reified in an intermediate representation (it would mean that the constructor of a type \code{A} would yield
@@
 limited to record types and provides only projection functions, so for more complex cases users have to write some
-repetitive and boilerplate code. (FIXME Example?)
+repetitive and boilerplate code.
 
@@
 
-\section{Evaluation}
+\section{Implementation and Evaluation}
 \label{validation}
 
+\subsection{Implementation}
+\label{implementation}
+
+This section presents relevant implementation details of the js-scala DSL.
+
+\subsubsection{Type-directed DSL constructs}
+
+The \emph{ad hoc} polymorphism feature of our DSL is directly borrowed from Scala’s typeclasses but the type coercion
+feature required to define a type-level DSL to express constraints. Numeric operations take an implicit value of
+type \code{(A \~ B) ~> C}. This type is actually a syntactic sugar for the type \code{\~>[\~[A, B], C]} (Scala
+supports an infix notation for type constructors of arity two). The \code{\~\>} type also takes two value parameters,
+saying how to obtain a value of the target type for each operand. Here is the definition of the \code{\~\>} type:
+
+\begin{lstlisting}
+class ~>[A <: Args, B : Numeric](
+  val lhs: Rep[A#Lhs] => Rep[B],
+  val rhs: Rep[A#Rhs] => Rep[B]
+)
+trait Args { type Lhs; type Rhs }
+type ~[A, B] = Args { type Lhs = A; type Rhs = B }
+\end{lstlisting}
+
+The \code{\~} type is just an alias to the \code{Args} type to get the infix notation. The \code{Args} type
+aggregates the types of the operands (\code{Lhs} and \code{Rhs}) involved in a binary operation.
+
+When a numeric operation is used, the Scala compiler fills the required implicit parameter of type
+\code{(A ~ B) ~> C} by looking for an implicit value of this type that satisfies the operands types and fixes the
+\code{C} type parameter value. The defined implicit values are given in lisiting \ref{type-coercion-constraints}.
+For example, for a numeric operation involving a \code{Rep[Int]} value and a \code{Rep[Double]} value, the
+\code{sameType} value is not applicable because its \code{A} type parameter cannot match both \code{Int} and
+\code{Double}, the \code{promoteRhs} value is not applicable because there is no implicit conversion from
+\code{Double} to \code{Int} but the \code{promoteLhs} value is applicable and fixes the \code{C} type parameter to
+\code{Double}. So, such an operation would promote the \code{Rep[Int]} parameter into a \code{Rep[Double]} value and
+would return a \code{Rep[Double]} value.
+
+\subsubsection{Intermediate representation construction}
+
+Every DSL construct producing code in the final program is first reified in an intermediate representation. Listing
+\ref{option-impl} shows extracts of both the abstract DSL interface of the \code{null} reference handling DSL and its
+implementation building an intermediate representation of values dereferencing.
+
+\begin{figure}
+\begin{lstlisting}[caption=Null reference handling DSL,label=option-impl]
+// DSL interface
+trait OptionOps { this: Base =>
+  implicit def repOptionToOps[A](o: Rep[Option[A]]) = new OptionOpsCls(o)
+  class OptionOpsCls[+A](o: Rep[Option[A]]) {
+    def isEmpty = option_isEmpty(o)
+    def foreach(f: Rep[A] => Rep[Unit]) = option_foreach(o, f)
+  }
+
+  def option_isEmpty[A](o: Rep[Option[A]]): Rep[Boolean]
+  def option_foreach[A](o: Rep[Option[A]], f: Rep[A] => Rep[Unit]): Rep[Unit]
+}
+// Intermediate representation
+trait OptionOpsExp extends OptionOps { this: EffectExp =>
+  def option_isEmpty[A](o: Exp[Option[A]]) = OptionIsEmpty(o)
+  def option_foreach[A](o: Exp[Option[A]], f: Exp[A] => Exp[Unit]) = {
+    val a = fresh[A]
+    val block = reifyEffects(f(a))
+    reflectEffect(OptionForeach(o, a, block), summarizeEffects(block).star)
+  }
+
+  case class OptionIsEmpty[A](o: Exp[Option[A]]) extends Def[Boolean]
+  case class OptionForeach[A](o: Exp[Option[A]], a: Sym[A], block: Block[Unit]) extends Def[Unit]
+}
+\end{lstlisting}
+\end{figure}
+
+We limited the DSL to two operations for the sake of brevity. The \code{OptionOps} trait defines the user facing DSL,
+it enriches \code{Rep[Option[_]]} values by providing an implicit conversion to \code{OptionOpsCls} that defines the
+\code{isEmpty} and \code{foreach} methods that respectively tests the emptiness of a value and derefences a value if
+not empty. These methods provide a convenient syntax to use \code{Rep[Option[_]]} values and delegate to the
+\code{option_isEmpty} and \code{option_foreach} operations, respectively, that are left abstract. The
+\code{OptionOpsExp} trait implements the \code{option_isEmpty} and \code{option_foreach} methods.
+The \code{option_isEmpty} implementation simply returns a statement node of type \code{ForeachIsEmpty}, but the
+\code{option_foreach} implementation is more subtle because it has to track the dependencies of its function
+parameter \code{f} (for instance, side-effects performed within the \code{f} function must be reflected by the
+returned \code{OptionForeach} node).
+
+The principle is the same for all DSLs: first, provide a trait with the DSL vocabulary, then implement each DSL
+construct to return a reified intermediate representation of itself. However, sometimes you have the opportunity to
+perform some optimizations at the evaluation level such as constants folding. For instance, the DOM definition DSL
+builds trees consisting of a root node and eventually several children. If these children are known at evaluation
+time we can build a node having a \code{List[Rep[Node]]} value as children instead of a \code{Rep[List[Node]]} so the
+generated code will be able to inline the loop adding the children to the parent node. Listing \ref{forest-impl}
+shows the relevant part of code checking if the \code{children} parameter of the \code{forest_tag} method is a
+constant list (the \code{ListNew} statement node represents constant lists) and building different intermediate
+representations according to this test.
+
+\begin{figure}
+\begin{lstlisting}[caption=Optimized intermediate representation for the DOM definition DSL,label=forest-impl]
+trait ForestExp extends Forest with EffectExp { this: ListOpsExp =>
+  def forest_tag(name: String, attrs: Map[String, Exp[String]], children: Exp[List[Node]]) = {
+    reflectEffect {
+      children match {
+        case Def(ListNew(children)) =>
+          Tag(name, Left(children.toList), attrs)
+        case _ =>
+          Tag(name, Right(children), attrs)
+      }
+    }
+  }
+  case class Tag(name: String, children: Either[List[Exp[Node]], Exp[List[Node]]], attrs: Map[String, Exp[String]]) extends Def[Node]
+}
+\end{lstlisting}
+\end{figure}
+
+\subsubsection{Code generation}
+
+Code generation consists in traversing the statement nodes procudes by the program evaluation according to their
+dependencies and to emit the code corresponding to each statement. LMS already sorts the statements graph so DSL
+authors just need to say how to emit code for each statement node of their DSL. Listing \ref{option-codegen} shows
+such a code generator for the \code{null} reference handling DSL. The \code{emitNode} method handles
+\code{OptionIsEmpty} and \code{OptionForeach} nodes. In the case of the \code{OptionIsEmpty} node, it simply
+generates an expression testing if the value is \code{null}, in the case of the \code{OptionForeach} node, it wraps
+the code block using the value within a \code{if} checking that the value is not \code{null}.
+
+\begin{figure}
+\begin{lstlisting}[caption=Null reference handling DSL code generator,label=option-codegen]
+trait JSGenOptionOps extends JSGenEffect {
+  val IR: OptionOpsExp
+  import IR._
+  override def emitNode(sym: Sym[Any], rhs: Def[Any]) = rhs match {
+    case OptionIsEmpty(o) =>
+      emitValDef(sym, quote(o) + " === null")
+    case OptionForeach(o, a, b) =>
+      stream.println("if (" + quote(o) + " !== null) {")
+      emitValDef(a, quote(o))
+      emitBlock(b)
+      stream.println("}")
+    case _ super.emitNode(sym, rhs)
+  }
+}
+\end{lstlisting}
+\end{figure}
+
+\subsection{Evaluation}
+
 Our goal was to define, with a minimal effort, a language for RIAs development. The main challange was to provide
-language constructs both safer and more expressive without abstraction penalty. We focused on the main concerns of
-RIA development: handling events, asynchronous programming and DOM fragments definition, as well as on more general
-programming concerns: \code{null} references handling and type safety.
-
-\subsection{Observed benefits of the use of our DSL}
-
-Using our DSL, we were able to completely rewrite the client-side part of our application, and divided by two the
-length of some parts of code (DOM definition and \code{null} reference handling). The size of code parts dealing with
+language constructs both safer and more expressive programming without abstraction penalty. We focused on the main
+concerns of RIA development: event handling, asynchronous programming and DOM fragments definition, as well as on
+more general programming concerns: \code{null} reference handling and type safety.
+
+Using our DSL, we were able to completely rewrite the client-side part of our mind mapping application, and divided
+by two the length of some parts of code (DOM definition and \code{null} reference handling). The size of code parts dealing with
 events did not change but gained type safety. Finally, code handling asynchronous computations were easier to
 modularize. An interesting side-effect of the design of our DSL as a compiled embedded DSL is the ability to reuse
-some code between both client and server sides (DOM definition).
+some code between both client and server sides (DOM definition). (TODO More on this consequence)
 
@@
 
-\subsection{Comparison with other approaches}
+\section{Related Works}
+\label{related}
+
+The most popular JavaScript library, jQuery~\cite{Bibeault08_jQuery}, used by more than 40\% of the top million
+sites\footnote{\href{http://trends.builtwith.com/javascript}{http://trends.builtwith.com/javascript}} aims to
+simplify RIA development. Unsurprisingly, it supports, at least partially, the RIA specific features of the table but
+suffers from the JavaScript language limitations. jQuery handles the \code{null} references problem by wrapping each
+query result in a container so before each further method call it tests the emptiness of the container and applies
+effectively the operation only if the container is not empty. It provides an expressive API but the emptiness
+checking involves a slight performance penalty since each result is wrapped and then unwrapped for the subsequent
+method invocation.
+
+TypeScript is a language compiling to JavaScript supporting classes, modules and soft typing. Because TypeScript is
+a superset of JavaScript, any valid JavaScript program is a valid TypeScript program. Developers can progressively
+leverage TypeScript’s features to turn their JavaScript sources into large, modular and more robust code bases.
+However, TypeScript does not address problems specific to RIAs development like DOM fragments definition or
+asynchronous programming.
+
+The Roy programming language is a statically typed functional programming language compiling to JavaScript. It
+features global type inference, algebraic data types, pattern matching and a sequencing notation similar to Scala’s
+\code{for} notation.
+
+GWT compiles Java code to JavaScript. It exposes the browser’s APIs through type safe Java APIs and simplifies DOM
+fragments definition. However, it still uses a callback-driven programming models for asynchronous computations.
+
+Kotlin is a recent language written by JetBrains, that targets both the JVM and JavaScript. It has libraries
+addressing, at least partially, Web programming concerns and supports advanced programming language features such as
+mixins, type inference, variance annotations for type parameters and pattern matching.
+
+Opa is a language and a full stack designed for Web programming. It features static typing with global
+type inference, modules and pattern matching. The language provides a syntactic sugar for writing HTML fragments in
+the code. The server and client parts of an application can be written in the Opa language, the system can decide
+whether a code fragment will run on server-side, client-side or both.
+
+Dart is a language designed specifically for Web programming. It has been designed to have a familiar syntax for
+JavaScript developers and supports soft typing, classes and a module system. It also provides a high-level concurrent
+programming system similar to an actor system. Finally, Dart code can be executed on both client and server sides.
+
+(TODO Some words about the CPS-based DSL for asynchronous programming of the previous paper)
+
+Concerning the client-server code sharing feature, our solution is the only one that gives full control on the
+generated code on both client and server for a given piece of shared code.
 
 Table \ref{comparison} gives a synthetic view of the comparison of our work with other approaches. Lines give
-features desired to write large Web applications and colums list famous tools for Web programming. Cells show the
+features desired to write large Web applications and colums list mainstream tools for Web programming. Cells show the
 support level of each feature for each tool.
@@
 \end{tabular}
-\caption{Support level of Web programming features by famous tools. Zero star means that the feature is not supported
+\caption{Support level of Web programming features by mainstream tools. Zero star means that the feature is not supported
 at all, three stars mean a strong support.}
@@
 
-The most popular JavaScript library, jQuery~\cite{Bibeault08_jQuery}, used by more than 40\% of the top million
-sites\footnote{\href{http://trends.builtwith.com/javascript}{http://trends.builtwith.com/javascript}} aims to
-simplify RIA development. Unsurprisingly, it supports, at least partially, the RIA specific features of the table but
-suffers from the JavaScript language limitations. jQuery handles the \code{null} references problem by wrapping each
-query result in a container so before each further method call it tests the emptiness of the container and applies
-effectively the operation only if the container is not empty. It provides an expressive API but the emptiness
-checking involves a slight performance penalty since each result is wrapped and then unwrapped for the subsequent
-method invocation.
-
-TypeScript is a language compiling to JavaScript supporting classes, modules and soft typing. Because TypeScript is
-a superset of JavaScript, any valid JavaScript program is a valid TypeScript program. Developers can progressively
-leverage TypeScript’s features to turn their JavaScript sources into large, modular and more robust code bases.
-However, TypeScript does not address problems specific to RIAs development like DOM fragments definition or
-asynchronous programming.
-
-The Roy programming language is a statically typed functional programming language compiling to JavaScript. It
-features global type inference, algebraic data types, pattern matching and a sequencing notation similar to Scala’s
-\code{for} notation.
-
-GWT compiles Java code to JavaScript. It exposes the browser’s APIs through type safe Java APIs and simplifies DOM
-fragments definition. However, it still uses a callback-driven programming models for asynchronous computations.
-
-Kotlin is a recent language written by JetBrains, that targets both the JVM and JavaScript. It has libraries
-addressing, at least partially, Web programming concerns and supports advanced programming language features such as
-mixins, type inference, variance annotations for type parameters and pattern matching.
-
-Opa is a language and a full stack designed for Web programming. It features static typing with global
-type inference, modules and pattern matching. The language provides a syntactic sugar for writing HTML fragments in
-the code. The server and client parts of an application can be written in the Opa language, the system can decide
-whether a code fragment will run on server-side, client-side or both.
-
-Dart is a language designed specifically for Web programming. It has been designed to have a familiar syntax for
-JavaScript developers and supports soft typing, classes and a module system. It also provides a high-level concurrent
-programming system similar to an actor system. Finally, Dart code can be executed on both client and server sides.
-
-Concerning the client-server code sharing feature, because our solution is the only one that gives full control on
-the generated code on both client and server for a given piece of shared code, we considered that our support was
-better than other approaches.
-
-\section{Conclusion and perspectives}
+\section{Conclusion and Perspectives}
 \label{discussion}

### 2012-12-14 papers/ecoop2013/repairing.tex
Weekend

@@
 \textsuperscript{\textdaggerdbl}Zenexity, Paris, France: \texttt{jrf@zenexity.com}\\
-\textsuperscript{\textdagger}EPFL, Lausanne, Switzerland: \texttt{\{first\}.\{last\}@epfl.ch}}
+\textsuperscript{\textdagger}EPFL, Lausanne, Switzerland: \texttt{tiark.rompf@epfl.ch}}
 
@@
 client-side code listens to user interface events (such as mouse clicks), triggers the appropriate action
-asynchronously on the server using AJAX~\cite{Garrett05_Ajax} and updates the application’s state and the DOM content
+asynchronously on the server using AJAX~\cite{Garrett05_Ajax} and updates the application’s state and the page content
 with a new DOM fragment computed from the response data~\cite{Farrell07_RIA,Busch09_StateOfArtRIA}.
@@
 the same language in both sides, which, in the case of the Web, means using JavaScript to write the whole
-application. However this language has several drawbacks making it hardly suitable for large code base (\eg{} no
+application. However this language has several drawbacks making it hardly suitable for large code bases (\eg{} no
 static typing, no module system, verbose syntax, \etc). Some other in-browser execution environments give the
@@
 not be available on all devices having a Web browser: for instance there is no way to execute Flash objects within an
-Apple smart phone), the page content can't naturally be referenced by search engines, and the content is not
+Apple smart phone), the page content cannot naturally be referenced by search engines, and the content is not
 structured in URLs that users can bookmark or share.
@@
 
-This paper continues their work and presents safe and expressive DSL constructs for RIAs development. We increase the
-expressivity of JavaScript native APIs with no runtime overhead and expose them through statically typed APIs. We
-make asynchronous programming simpler by raising asynchronous computations to first-class citizen. We leverage the
-Scala type system to bring type-directed language features such as \emph{ad hoc} polymorphism and type coercion to
-the client-side.
+This paper continues their work and presents several safe and expressive DSL constructs for RIAs development. We
+increase the expressivity of JavaScript native APIs with no runtime overhead and expose them through statically typed
+APIs. We make asynchronous programming simpler by raising asynchronous computations to first-class citizens. We
+leverage the Scala type system to bring type-directed language features such as \emph{ad hoc} polymorphism and type
+coercion to the client-side.
 
@@
 
-The next section introduces a simple application example showing challenges of RIA development, section
-\ref{solution} presents our DSL tackling these challenges, section \ref{validation} validates our solution, section
-\ref{related} presents related works and compares it to related work and section \ref{discussion} concludes.
+The next section introduces a simple application example showing the main challenges of RIA development, section
+\ref{solution} presents our DSL tackling these challenges, section \ref{validation} evaluates our solution, section
+\ref{related} compares our solution to related works and section \ref{discussion} concludes.
 
@@
 mispell it. (2) The callback passed as a second parameter takes a parameter \code{e} whose fields vary according to
-the listened event but developers have no way to check that the fields they’re using are indeed defined on the event
+the listened event, but developers have no way to check that the fields they’re using are indeed defined on the event
 they’re listening to. For instance, in our case we use the \code{wheelDeltaY} event field that is defined only on the
@@
 Another issue with callback-driven programming arises when several dependent asynchronous computations are executed
-sequentially. Consider for example the following listing performing three consecutive Ajax requests:
-
+sequentially. Consider for example the code of listing \ref{async-callback-hell} performing three consecutive Ajax
+requests. Notice that the code is getting deeper toward the right. This problem is often referred to as the
+“callback hell”~\cite{McKenna_Roy}. This code is also hard to reason about: when will the second \code{console.log}
+statement be executed? Before or after \code{baz} has been fetched?
+
+\begin{figure}
 \begin{lstlisting}[language=JavaScript,label=async-callback-hell,caption=Sequential asynchronous calls]
@@
 \end{lstlisting}
-
-Notice that the code is getting deeper toward the right, this problem is often referred to as the
-“callback-hell”~\cite{McKenna_Roy}. This code is also hard to reason about: when will the second \code{console.log}
-statement be executed? Before or after \code{baz} has been fetched?
+\end{figure}
 
@@
 \end{lstlisting}
+% 
+% \begin{lstlisting}
+% var showMap = function (map) {
+% 
+% var vertices = function (vs) {
+%   return vs.map(function (v) {
+%     var g = document.createElementNS('g', 'svgns');
+%     g.setAttribute('class', 'vertex');
+%     g.setAttribute('transform', "translate(" + v.x + "," + v.y + ")");
+%     var rect = document.createElementNS('rect', 'svgns');
+%     rect.setAttribute('width', v.width);
+%     rect.setAttribute('height', v.height);
+%     var f = document.createElement('foreignObject');
+%     f.setAttribute('x', 0);
+%     f.setAttribute('y', 0);
+%     f.setAttribute('width', v.width);
+%     f.setAttribute('height', v.height);
+%     var p = document.createElement('p');
+%     p.setAttribute('style', 'height: '+v.height+'px;');
+%     p.appendChild(document.createTextNode(v.content));
+%     f.appendChild(p)
+%     g.appendChild(rect);
+%     g.appendChild(f);
+%     return g
+%   });
+% };
+% 
+% var edges = function (es) {
+%   return es.map(function (e) {
+%     var x1 = e.orig.x + e.orig.width / 2;
+%     var y1 = e.orig.y + e.orig.height / 2;
+%     var x2 = e.end.x + e.end.with / 2;
+%     var y2 = e.end.y + e.end.height / 2;
+%     var g = document.createElementNS('g', 'svgns');
+%     g.setAttribute('class', 'edge');
+%     var line = document.createElement('line', 'svg');
+%     line.setAttribute('x1', x1);
+%     line.setAttribute('x2', x2);
+%     line.setAttribute('x3', x3);
+%     line.setAttribute('x4', x4);
+%     return g;
+%   });
+% };
+% 
+% return vertices(map.vertices).concat(edges(map.edges));
+% 
+% };
+% \end{lstlisting}
 
@@
 The above code may be more readable, however this implementation is wrong: if we call it with a content containing
-brackets (\eg{} \code{"<foo>"}) they won’t be escaped and will produce a \code{foo} tag nested in the \code{text} tag,
-which is not the intended behavior. So, although this way reads slightly better it’s not less error prone.
+brackets (\eg{} \code{"<foo>"}), they won’t be escaped and will produce a \code{foo} tag nested in the \code{text}
+tag, which is not the intended behavior. So, although this way reads slightly better it’s not less error prone.
 
@@
 structure and allowing the insertion of dynamic expressions in a safe way. Some template engines can be used on both
-client-side and server-side, however none has a practical expression language (dynamic content can only be provided
-as a map of key-value pairs and no operation can be done on values) or is statically typed.
+client-side and server-side, however none is statically typed or has a practical expression language: dynamic content
+can only be provided as a map of key-value pairs (rather than typed values) and cannot be processed for
+presentational purposes from within the template.
 
@@
 
-In the previous section we showed the three main tasks performed by the JavaScript code in RIAs: events handling,
+In the previous section we presented the three main tasks performed by the JavaScript code in RIAs: event handling,
 asynchronous programming and DOM manipulation. We noticed the following issues:
 
-\subsubsection{Lack of static hecks}
+\subsubsection{Lack of static checks}
 
@@
 The lack of static checks is even harder to tackle in the case of code shared between client and server sides. If we
-write the expression \code{xs.size == 0} in a template, how should it be translated to the client-side environment
+write the expression \code{xs.size == 0}, how should it be translated to the client-side environment
 (JavaScript) and to the server-side environment (\eg{} the Java Virtual Machine)? Several solutions are valid,
@@
 \code{xs.length === 0} if \code{xs} refers to a collection (in JavaScript the property to get the size of a
-collection is named \code{length}). We need to define a proper expression language and to define how it should be
+collection is named \code{length}). So we need to define a proper expression language and to define how it should be
 translated for the client-side environment and the server-side environment, but such a task requires a high effort
@@
 notation). Defining a language requires a high effort, especially to build development tools for the language so we
-chose to define our language as a compiled embedded DSL in Scala\footnote{The code is available at
-\href{http://github.com/js-scala}{http://github.com/js-scala}}.
+chose to define our language as a compiled embedded DSL in Scala.
 
@@
 representation of the program and the second stage transforms this intermediate representation into executable code
-(figure \ref{lms-diagram}). The embedding lets us reuse the host language infrastructure (compiler, development
-tools) and type system, and staging gives the opportunity to generate efficient code from expressive code by applying
-domain specific optimizations.
+(figure \ref{lms-diagram}).
 
@@
 to read a property that is not defined on the type of the handled event. The difficulty comes from the fact that the
-type of the event passed to the handler varies with the event name. For instance in the listing \ref{event-js}, the
-handler processses \code{MouseWheelEvent} values because it listens to \code{mousewheel} events. Values of type
+type of the event passed to the handler varies with the event name. For instance, in listing \ref{event-js} the
+handler processes \code{MouseWheelEvent} values because it listens to \code{mousewheel} events. Values of type
 \code{MouseWheelEvent} have a property \code{wheelDeltaY} but that’s not the case of an event data of type
@@
 class EventDef {
-  type Type
-}
-
-def on(ev: EventDef)(handler: Rep[ev.Type] => Rep[Unit]): Rep[Unit]
+  type Data
+}
+
+def on(ev: EventDef)(handler: Rep[ev.Data] => Rep[Unit]): Rep[Unit]
 \end{lstlisting}
@@
 (We renamed \code{addEventListener} to \code{on}, for the sake of brevity). An \code{EventDef} value represents an
-event that carries its corresponding event data type in its \code{Type} type member. The \code{on} method takes a
-parameter \code{ev} of type \code{EventDef} and a \code{handler} parameter whose type refers to the \code{Type}
+event that carries its corresponding event data type in its \code{Data} type member. The \code{on} method takes a
+parameter \code{ev} of type \code{EventDef} and a \code{handler} parameter whose type refers to the \code{Data}
 member of the \code{ev} parameter, so the type of the handler depends on the \code{ev} value. We can then define the
@@
 \begin{lstlisting}
-object KeyUp extends EventDef { type Type = KeyboardEvent }
-object MouseWheel extends EventDef { type Type = MouseWheelEvent }
+object KeyUp extends EventDef { type Data = KeyboardEvent }
+object MouseWheel extends EventDef { type Data = MouseWheelEvent }
 \end{lstlisting}
@@
 
+The previous work on js-scala \cite{Kossakowski12_JsDESL} showed how to make asynchronous programming more
+convenient by making asynchronous calls looking like synchronous calls (\ie{} returning a value instead of taking a
+callback as parameter). This work was based on the Scala continuations compiler plugin. We claim that, although this
+programming model removes the “callback hell”, it can make the code hard to reason about because there is no explicit
+distinction between synchronous and asynchronous parts.
+
 We propose a DSL that explicitly reflects the asynchronous nature of computations in their types and provides methods
-turning them into first-class citizens. The DSL is monadic so we can solve the callback-hell problem thanks to the
+turning them into first-class citizens. The DSL is monadic so we can solve the callback hell problem thanks to the
 Scala \code{for} notation.
 
-For instance, listing \ref{async-first-class} shows how the listing \ref{async-js} can be
-re-written with our DSL. The \code{createVertex} now returns an asynchronous value instead of taking a callback as
-parameter. Then, the \code{for} expression allows us to use the vertex, when available, and to insert it on the user
+For instance, listing \ref{async-first-class} shows how the listing \ref{async-js} can be re-written using our DSL.
+The \code{createVertex} function now returns an asynchronous value instead of taking a callback as
+parameter. Then, the \code{for} expression allows us to get the vertex, when available, and to insert it on the user
 interface. By making the \code{createVertex} function return an asynchronous value instead of taking a callback as
@@
 
-The listing \ref{async-no-callback} translates the “callback-hell” example (listing \ref{async-callback-hell}) using
+The listing \ref{async-no-callback} translates the callback hell example (listing \ref{async-callback-hell}) using
 our DSL. The \code{for} notation can intuitively be thought of as a sequencing notation: whenever the response of the
@@
 
-In this section we show how we can define a template engine as an embedded DSL, with minimal effort. This template
+In this section we show how we can define a template engine as an embedded DSL with minimal effort. This template
 engine is statically typed and able to insert dynamic content in a safe way. It provides a powerful expression
@@
 So the only remaining work consists in defining the DSL vocabulary to define DOM nodes. We provide a \code{tag}
-function to define a tag and a \code{text} function to define a text node. The following listing uses our DSL and
-generates a code equivalent to listing \ref{dom-api}:
-
+function to define a tag and a \code{text} function to define a text node.
+
+\begin{figure}
 \begin{lstlisting}[label=forest-hello,caption=DOM definition DSL]
@@
 \end{lstlisting}
-
-The readability has been highly improved: nesting tags is just like nesting code blocks, HTML entities are
+\end{figure}
+
+Listing \ref{forest-hello} uses our DSL and generates a code equivalent to listing \ref{dom-api}. The readability has
+been highly improved: nesting tags is just like nesting code blocks, HTML entities are
 automatically escaped in text nodes, developers have the full computational power of Scala to inject dynamic data and
@@
 Our DSL is equivalent to a template engine with Scala as the expression language. Making it usable on both server and
-client sides is surprisingly as simple as defining another code generator for the DSL, producing Scala code.
+client sides was surprisingly as simple as defining another code generator for the DSL, producing Scala code.
 
@@
 
-Because of the dynamically typed nature of JavaScript, when calling a function there is no proper way to call a
+Because of the dynamically typed nature of JavaScript, when calling a function there is no proper way to select a
 specialized implementation according to the function’s parameters types. JavaScript is only able to dispatch
@@
 
-In summary, the compiled embedded approach may require more work to support custom data types but gives control on
-the way this data type will be encoded for each target platform (\eg{} the JavaScript code generator uses native
+In summary, the compiled embedded approach may require more work to define custom data types but gives control on
+the way these data types will be encoded for each target platform (\eg{} the JavaScript code generator uses native
 arrays to encode the lists of the list manipulation DSL).
@@
 
-Furthermore, a code generator can be tweaked by defining a sub-trait specializing some of its methods. This is useful
-to handle cross-browser incompatibilities. For instance we have written code generators specializing the output for
-some statement nodes in order to handle Internet Explorer incompatibilities.
+Furthermore, a code generator can be tweaked by defining a sub-trait specializing some of its methods. This feature
+can be used to handle cross-browser incompatibilities. For instance we have written code generators specializing the
+output for some statement nodes in order to handle Internet Explorer incompatibilities.
 
@@
 
-This section presents relevant implementation details of the js-scala DSL.
+This section presents relevant implementation details of the js-scala DSL\footnote{The complete implementation is
+available at \href{http://github.com/js-scala}{http://github.com/js-scala}}.
 
@@
 
-The \emph{ad hoc} polymorphism feature of our DSL is directly borrowed from Scala’s typeclasses but the type coercion
-feature required to define a type-level DSL to express constraints. Numeric operations take an implicit value of
-type \code{(A \~ B) ~> C}. This type is actually a syntactic sugar for the type \code{\~>[\~[A, B], C]} (Scala
-supports an infix notation for type constructors of arity two). The \code{\~\>} type also takes two value parameters,
-saying how to obtain a value of the target type for each operand. Here is the definition of the \code{\~\>} type:
+The \emph{ad hoc} polymorphism feature of our DSL is directly borrowed from Scala’s typeclasses encoding but the type
+coercion feature required to define a type-level DSL to express constraints. Numeric operations take an implicit
+value of type \code{(A \~ B) ~> C}. This type is actually a syntactic sugar for the type \code{\~>[\~[A, B], C]}
+(Scala supports an infix notation for type constructors of arity two). The \code{\~\>} type also takes two value
+parameters, saying how to obtain a value of the target type for each operand. Here is the definition of the
+\code{\~\>} type:
 
@@
 trait OptionOps { this: Base =>
-  implicit def repOptionToOps[A](o: Rep[Option[A]]) = new OptionOpsCls(o)
+  implicit def repOptionToOps[A](o: Rep[Option[A]]) =
+    new OptionOpsCls(o)
   class OptionOpsCls[+A](o: Rep[Option[A]]) {
     def isEmpty = option_isEmpty(o)
-    def foreach(f: Rep[A] => Rep[Unit]) = option_foreach(o, f)
+    def foreach(f: Rep[A] => Rep[Unit]) =
+      option_foreach(o, f)
   }
@@
   def option_isEmpty[A](o: Rep[Option[A]]): Rep[Boolean]
-  def option_foreach[A](o: Rep[Option[A]], f: Rep[A] => Rep[Unit]): Rep[Unit]
+  def option_foreach[A](o: Rep[Option[A]],
+      f: Rep[A] => Rep[Unit]): Rep[Unit]
 }
@@
   def option_isEmpty[A](o: Exp[Option[A]]) = OptionIsEmpty(o)
-  def option_foreach[A](o: Exp[Option[A]], f: Exp[A] => Exp[Unit]) = {
+  def option_foreach[A](o: Exp[Option[A]],
+      f: Exp[A] => Exp[Unit]) = {
     val a = fresh[A]
-    val block = reifyEffects(f(a))
-    reflectEffect(OptionForeach(o, a, block), summarizeEffects(block).star)
+    val b = reifyEffects(f(a))
+    reflectEffect(OptionForeach(o, a, b), summarizeEffects(b).star)
   }
@@
   case class OptionIsEmpty[A](o: Exp[Option[A]]) extends Def[Boolean]
-  case class OptionForeach[A](o: Exp[Option[A]], a: Sym[A], block: Block[Unit]) extends Def[Unit]
+  case class OptionForeach[A](o: Exp[Option[A]],
+      a: Sym[A], block: Block[Unit]) extends Def[Unit]
 }
@@
 
-We limited the DSL to two operations for the sake of brevity. The \code{OptionOps} trait defines the user facing DSL,
-it enriches \code{Rep[Option[_]]} values by providing an implicit conversion to \code{OptionOpsCls} that defines the
-\code{isEmpty} and \code{foreach} methods that respectively tests the emptiness of a value and derefences a value if
-not empty. These methods provide a convenient syntax to use \code{Rep[Option[_]]} values and delegate to the
-\code{option_isEmpty} and \code{option_foreach} operations, respectively, that are left abstract. The
-\code{OptionOpsExp} trait implements the \code{option_isEmpty} and \code{option_foreach} methods.
-The \code{option_isEmpty} implementation simply returns a statement node of type \code{ForeachIsEmpty}, but the
-\code{option_foreach} implementation is more subtle because it has to track the dependencies of its function
-parameter \code{f} (for instance, side-effects performed within the \code{f} function must be reflected by the
-returned \code{OptionForeach} node).
-
-The principle is the same for all DSLs: first, provide a trait with the DSL vocabulary, then implement each DSL
-construct to return a reified intermediate representation of itself. However, sometimes you have the opportunity to
-perform some optimizations at the evaluation level such as constants folding. For instance, the DOM definition DSL
+We limited the DSL to two operations (\code{foreach} and \code{isEmpty}) for the sake of brevity. The
+\code{OptionOps} trait defines the user facing DSL, it enriches \code{Rep[Option[_]]} values by providing an implicit
+conversion to \code{OptionOpsCls} that defines the \code{isEmpty} and \code{foreach} methods that respectively tests
+the emptiness of a value and derefences a value if not empty. These methods provide a convenient syntax to use
+\code{Rep[Option[_]]} values and delegate to the \code{option_isEmpty} and \code{option_foreach} operations,
+respectively, that are left abstract. The \code{OptionOpsExp} trait implements the \code{option_isEmpty} and
+\code{option_foreach} methods. The \code{option_isEmpty} implementation simply returns a statement node of type
+\code{OptionIsEmpty}, but the \code{option_foreach} implementation is more subtle because it has to track the
+dependencies of its function parameter \code{f} (for instance, side-effects performed within the \code{f} function
+must be reflected by the returned \code{OptionForeach} node).
+
+The principle is the same for all DSLs: first, provide an abstract trait with the DSL vocabulary, then implement each
+DSL construct to return a reified intermediate representation of itself. Sometimes you have the opportunity
+to perform some optimizations at the evaluation level such as constants folding. For instance, the DOM definition DSL
 builds trees consisting of a root node and eventually several children. If these children are known at evaluation
@@
 trait ForestExp extends Forest with EffectExp { this: ListOpsExp =>
-  def forest_tag(name: String, attrs: Map[String, Exp[String]], children: Exp[List[Node]]) = {
-    reflectEffect {
-      children match {
+  def forest_tag(name: String, attrs: Map[String, Exp[String]],
+      children: Exp[List[Node]]) = children match {
     case Def(ListNew(children)) =>
@@
   }
-    }
-  }
-  case class Tag(name: String, children: Either[List[Exp[Node]], Exp[List[Node]]], attrs: Map[String, Exp[String]]) extends Def[Node]
+  case class Tag(name: String,
+      children: Either[List[Exp[Node]], Exp[List[Node]]],
+      attrs: Map[String, Exp[String]]) extends Def[Node]
 }
@@
 
-Code generation consists in traversing the statement nodes procudes by the program evaluation according to their
+Code generation consists in traversing the statement nodes produced by the program evaluation according to their
 dependencies and to emit the code corresponding to each statement. LMS already sorts the statements graph so DSL
@@
 generates an expression testing if the value is \code{null}, in the case of the \code{OptionForeach} node, it wraps
-the code block using the value within a \code{if} checking that the value is not \code{null}.
+the code block dereferencing the value within a \code{if} checking that the value is not \code{null}.
 
@@
 trait JSGenOptionOps extends JSGenEffect {
-  val IR: OptionOpsExp
+  val IR: EfffectExp with OptionOpsExp
   import IR._
-  override def emitNode(sym: Sym[Any], rhs: Def[Any]) = rhs match {
+  override def emitNode(sym: Sym[Any], rhs: Def[Any]) =
+    rhs match {
       case OptionIsEmpty(o) =>
@@
         stream.println("}")
-    case _ super.emitNode(sym, rhs)
+      case _ =>
+        super.emitNode(sym, rhs)
     }
@@
 Our goal was to define, with a minimal effort, a language for RIAs development. The main challange was to provide
-language constructs both safer and more expressive programming without abstraction penalty. We focused on the main
+programming language constructs both safer and more expressive without abstraction penalty. We focused on the main
 concerns of RIA development: event handling, asynchronous programming and DOM fragments definition, as well as on
@@
 
-Using our DSL, we were able to completely rewrite the client-side part of our mind mapping application, and divided
-by two the length of some parts of code (DOM definition and \code{null} reference handling). The size of code parts dealing with
-events did not change but gained type safety. Finally, code handling asynchronous computations were easier to
-modularize. An interesting side-effect of the design of our DSL as a compiled embedded DSL is the ability to reuse
-some code between both client and server sides (DOM definition). (TODO More on this consequence)
+Using our DSL, we were able to completely rewrite the client-side part of our mind mapping application and to
+integrate it in a mainstream Web framework for the server-side part. The number of
+lines of code needed to write the DOM fragments reduced by 38\%. The size of code parts dealing with events did not
+change but gained type safety. Finally, code handling asynchronous computations were easier to modularize.
+
+An interesting side-effect of the design of our DSL as a compiled embedded DSL is the ability to reuse some code
+between both client and server sides. We were able to reuse DOM fragments definition on both sides and to build two
+variants of the application with minimal effort. The first variant renders HTML from the server-side and the second
+variant renders DOM fragments from the client-side. The flexibility to decide later where to run the logic (on
+client-side or on server-side) gives more agility in the development because we can easily adapt the application. For
+instance we can choose to perform the rendering on server-side if the application targets smartphones with a small
+processing speed or to perform the rendering on client-side, to preserve the server CPU usage, if the application
+targets desktop computers.
 
 All these benefits came with no runtime overhead, excepted for the asynchronous programming DSL that creates, for
-each asynchronous computation, a \code{Promise} value registering the callbacks to call when the value is be
-available.
+each asynchronous computation, a \code{Promise} value registering the callbacks to call when the value is available.
 
@@
 sites\footnote{\href{http://trends.builtwith.com/javascript}{http://trends.builtwith.com/javascript}} aims to
-simplify RIA development. Unsurprisingly, it supports, at least partially, the RIA specific features of the table but
-suffers from the JavaScript language limitations. jQuery handles the \code{null} references problem by wrapping each
-query result in a container so before each further method call it tests the emptiness of the container and applies
-effectively the operation only if the container is not empty. It provides an expressive API but the emptiness
-checking involves a slight performance penalty since each result is wrapped and then unwrapped for the subsequent
-method invocation.
-
-TypeScript is a language compiling to JavaScript supporting classes, modules and soft typing. Because TypeScript is
-a superset of JavaScript, any valid JavaScript program is a valid TypeScript program. Developers can progressively
-leverage TypeScript’s features to turn their JavaScript sources into large, modular and more robust code bases.
-However, TypeScript does not address problems specific to RIAs development like DOM fragments definition or
-asynchronous programming.
-
-The Roy programming language is a statically typed functional programming language compiling to JavaScript. It
-features global type inference, algebraic data types, pattern matching and a sequencing notation similar to Scala’s
-\code{for} notation.
-
-GWT compiles Java code to JavaScript. It exposes the browser’s APIs through type safe Java APIs and simplifies DOM
-fragments definition. However, it still uses a callback-driven programming models for asynchronous computations.
-
-Kotlin is a recent language written by JetBrains, that targets both the JVM and JavaScript. It has libraries
-addressing, at least partially, Web programming concerns and supports advanced programming language features such as
-mixins, type inference, variance annotations for type parameters and pattern matching.
-
-Opa is a language and a full stack designed for Web programming. It features static typing with global
-type inference, modules and pattern matching. The language provides a syntactic sugar for writing HTML fragments in
-the code. The server and client parts of an application can be written in the Opa language, the system can decide
-whether a code fragment will run on server-side, client-side or both.
-
-Dart is a language designed specifically for Web programming. It has been designed to have a familiar syntax for
-JavaScript developers and supports soft typing, classes and a module system. It also provides a high-level concurrent
-programming system similar to an actor system. Finally, Dart code can be executed on both client and server sides.
-
-(TODO Some words about the CPS-based DSL for asynchronous programming of the previous paper)
-
-Concerning the client-server code sharing feature, our solution is the only one that gives full control on the
-generated code on both client and server for a given piece of shared code.
+simplify RIA development. It handles the \code{null} references problem by wrapping each query result in a container
+so before each further method call it tests the emptiness of the container and applies effectively the operation only
+if the container is not empty. It provides an expressive API but the emptiness checking involves a slight performance
+penalty since each result is wrapped and then unwrapped for the subsequent method invocation. jQuery also has an API
+to turn asynchronous computations into first-class citizens. However, jQuery is just a library and thus cannot bring
+type safety to JavaScript programs.
+
+TypeScript\footnote{\href{http://www.typescriptlang.org/}{http://www.typescriptlang.org/}} is a language compiling to
+JavaScript supporting classes, modules and soft typing. Because TypeScript is a superset of JavaScript, any valid
+JavaScript program is a valid TypeScript program. Developers can progressively leverage TypeScript’s features to turn
+their JavaScript sources into large, modular and more robust code bases. However, TypeScript does not address
+problems specific to RIAs development like DOM fragments definition or asynchronous programming.
+
+The Roy~\cite{McKenna_Roy} programming language is a statically typed functional programming language compiling to
+JavaScript. It features global type inference, algebraic data types, pattern matching and a sequencing notation
+similar to Scala’s \code{for} notation. Roy does not address concerns such as event handling and DOM fragments
+definition.
+
+GWT~\cite{Chaganti07_GWT} compiles Java code to JavaScript. It exposes the browser’s APIs through type safe Java APIs
+and simplifies DOM fragments definition. However, it still uses a callback-driven programming model for asynchronous
+computations.
+
+Kotlin\footnote{\href{http://kotlin.jetbrains.org/}{http://kotlin.jetbrains.org/}} is a recent language written by
+JetBrains, that targets both the JVM and JavaScript. It has libraries addressing partially Web programming concerns
+and supports advanced programming language features such as mixins, type inference, variance annotations for type
+parameters and pattern matching.
+
+Opa\footnote{\href{http://opalang.org/}{http://opalang.org/}} is a language and a full stack designed for Web
+programming. It features static typing with global type inference, modules and pattern matching. The language
+provides a syntactic sugar for writing HTML fragments in the code. The server and client parts of an application can
+be written in the Opa language, the system can decide whether a code fragment will run on server-side, client-side or
+both. However, the Opa language is not extensible, so there is no way for developers to customize the compilation
+process of their abstractions on client and server sides.
+
+Dart~\cite{Griffith11_Dart} is a language designed specifically for Web programming. It has been designed to have a
+familiar syntax for JavaScript developers and supports soft typing, classes and a module system. It also provides a
+high-level concurrent programming system similar to an actor system. Finally, Dart code can be executed on both
+client and server sides but the compiler is not extensible so developers cannot customize the compilation process of
+a given abstraction on client and server sides.
 
@@
 \hline
-Easiness for defining libraries & * & *** & *** & *** & *** & *** & *** & * \\
+Easiness for defining libraries & ** & *** & *** & *** & *** & *** & *** & * \\
 \hline
@@
 
-Writing RIAs is challenging because the JavaScript language and the Web browser’s native APIs have several
-inadequacies, making the code fragile, hard to reason about and to maintain.
-
-We improved the js-scala DSL by adding first-class support for RIA development concerns such as events handling,
+Writing RIAs is challenging because they require a lot of logic on the client-side but the JavaScript language and
+the Web browser’s native APIs make the code fragile and hard to reason about and to maintain.
+
+We improved the js-scala DSL by adding first-class support for RIA concerns such as events handling,
 DOM fragments definition and asynchronous programming. Our work provides a safer and more expressive language than
-JavaScript and its native APIs to write RIAs.
-
-Our approach’s distinguishing support of code sharing between server and clients let us consider the exploration of
-this feature more deeply in our future work, in order to factor out an increasing part of code between client and
-servers in RIAs.
+JavaScript and its native APIs to write RIAs but has no runtime overhead.
+
+Future work will focus on the ability to choose where to execute a piece of code (on server-side or on client-side)
+in order to be able to delay this choice until the runtime.
 

### 2012-12-17 src/main/scala/scala/js/JSDom.scala
Make some values Serializable

@@ -36,10 +36,10 @@ trait JSDom { this: Base =>
   trait MouseWheelEvent extends MouseEvent
   def infix_wheelDeltaY(e: Rep[MouseWheelEvent]): Rep[Double]
 
-  object MouseWheel extends EventName[MouseWheelEvent]("mousewheel")
-  object MouseDown extends EventName[MouseEvent]("mousedown")
-  object MouseMove extends EventName[MouseEvent]("mousemove")
-  object MouseUp extends EventName[MouseEvent]("mouseup")
+  case object MouseWheel extends EventName[MouseWheelEvent]("mousewheel")
+  case object MouseDown extends EventName[MouseEvent]("mousedown")
+  case object MouseMove extends EventName[MouseEvent]("mousemove")
+  case object MouseUp extends EventName[MouseEvent]("mouseup")
 
   trait Window extends EventTarget
 
@@ -83,7 +83,7 @@ trait JSDomExp extends JSDom with EffectExp {
   def infix_setAttribute(e: Exp[Element], name: Exp[String], value: Exp[Any]) = reflectEffect(ElementSetAttribute(e, name, value))
   def infix_tagName(e: Exp[Element]) = ElementTagName(e)
   def history_replaceState(h: Exp[History], state: Exp[_], title: Exp[String], url: Exp[String]) = reflectEffect(HistoryReplaceState(h, state, title, url))
-  object window extends Exp[Window]
+  case object window extends Exp[Window]
 
   // FIXME We can’t yet use dependent types on constructors parameters, see https://issues.scala-lang.org/browse/SI-5712 so at this point we lost the information that event.Type =:= A
   case class EventTargetOn[A](t: Exp[EventTarget], event: EventDef, capture: Exp[Boolean], e: Sym[A], handler: Block[Unit]) extends Def[Unit]
@@ -92,8 +92,8 @@ trait JSDomExp extends JSDom with EffectExp {
   case class MouseEventOffsetX(e: Exp[MouseEvent]) extends Def[Double]
   case class MouseEventOffsetY(e: Exp[MouseEvent]) extends Def[Double]
   case class MouseWheelEventDeltaY(e: Exp[MouseWheelEvent]) extends Def[Double]
-  object WindowDocument extends Exp[Document]
-  object WindowHistory extends Exp[History]
+  case object WindowDocument extends Exp[Document]
+  case object WindowHistory extends Exp[History]
[…]

### 2012-12-18 papers/ecoop2013/repairing.tex
Time to dance

@@
 
-\begin{lstlisting}
+\begin{lstlisting}[language=JavaScript]
 var createVertex = function (text) {
@@
 For instance the following call:
-\begin{lstlisting}
+\begin{lstlisting}[language=JavaScript]
 vertexDom({
@@
 
-The lack of static checks is even harder to tackle in the case of code shared between client and server sides. If we
-write the expression \code{xs.size == 0}, how should it be translated to the client-side environment
+The lack of static checks is even harder to tackle in the case of code shared between client and server sides. If one
+writes the expression \code{xs.size == 0}, how should it be translated to the client-side environment
 (JavaScript) and to the server-side environment (\eg{} the Java Virtual Machine)? Several solutions are valid,
-depending on the meaning of the expression: in JavaScript we could translate it as \code{xs.size === 0} or
+depending on the meaning of the expression: in JavaScript it could be translated as \code{xs.size === 0} if \code{xs}
+refers to an object, or as
 \code{xs.length === 0} if \code{xs} refers to a collection (in JavaScript the property to get the size of a
@@
 values and returns the intermediate representation of an addition. Finally, the \code{inc} function returns the
-intermediate representation of a computation yielding the number following the value of the parameter \code{x}. You
-can get a \code{T} value from a \code{Rep[T]} value by generating code from the intermediate representation and
+intermediate representation of a computation yielding the number following the value of the parameter \code{x}.
+
+You can get a \code{T} value from a \code{Rep[T]} value by generating code from the intermediate representation and
 compiling it:
@@
 to read a property that is not defined on the type of the handled event. The difficulty comes from the fact that the
-type of the event passed to the handler varies with the event name. For instance, in listing \ref{event-js} the
+type of data passed to the handler varies with the event name. For instance, in listing \ref{event-js} the
 handler processes \code{MouseWheelEvent} values because it listens to \code{mousewheel} events. Values of type
@@
 events. We want to provide a single polymorphic method similar to the native API. To achieve that, we need to encode
-the dependency relation between an event name and its corresponding event type at the type-level. This dependency can
+the dependency relation between an event name and its corresponding data at the type-level. This dependency can
 naturally and elegantly be encoded using dependent method types~\cite{Oliveira10_Typeclasses}:
@@
 
-The following Scala listing is equivalent to the JavaScript listing \ref{event-js} and is completely type safe: if
+The following Scala listing generates code equivalent to listing \ref{event-js} and is completely type safe: if
 the user misspells the event name or tries to use an undefined property on an event his program won’t compile.
@@
 programming model removes the “callback hell”, it can make the code hard to reason about because there is no explicit
-distinction between synchronous and asynchronous parts.
+distinction between synchronous and asynchronous computations.
 
@@
 
-The listing \ref{async-no-callback} translates the callback hell example (listing \ref{async-callback-hell}) using
+Listing \ref{async-no-callback} translates the callback hell example (listing \ref{async-callback-hell}) using
 our DSL. The \code{for} notation can intuitively be thought of as a sequencing notation: whenever the response of the
-first Ajax request is available, the next statement will be executed, and so on. There is no nested callbacks and the
+first Ajax request is available, the next statement is executed, and so on. There is no nested callbacks and the
 order of execution is directly reflected by the order of statements.
@@
 \begin{lstlisting}[language=JavaScript,caption=Generated JavaScript code,label=async-parallel-2]
-var x1 = new Promise();
+var x1 = { cs: [] };
 AjaxGet("/foo", x1);
-var x2 = new Promise();
+var x2 = { cs: [] };
 AjaxGet("/bar", x2);
-x1.onComplete(function (foo) {
- x2.onComplete(function (bar) {
+foreach(x1, function (foo) {
+ foreach(x2, function (bar) {
   var x3 = foo + bar;
@@
 \ref{async-parallel-2} shows the generated JavaScript code (we renamed some identifiers for the sake of readability).
-The generated \code{Promise} constructor code has been omitted, it creates an object holding a list of callbacks to
-call when the asynchronous value is completed. The \code{AjaxGet} function code has also been omitted, it creates a
-\code{XMLHttpRequest} object that sends an HTTP request and completes its promise parameter with the response, when
-received.
+The \code{x1} and \code{x2} values encode the asynchronous results of the Ajax requests, they contain an array of
+callbacks (\code{cs}, initially empty) to call when the result is available. The \code{AjaxGet} function code has
+been omitted, it creates a \code{XMLHttpRequest} object that sends a HTTP request and completes the value of its
+last parameter with the response, when received. Completing an asynchronous value means defining a \code{v} field
+on it containing the actual value, and calling all pending callbacks that are in the \code{cs} field. The
+\code{foreach} code has also been omitted, it registers a callback on an asynchronous value or calls the callback
+directly with the value if this one is already available.
 
@@
 
-However, the above code does not handle operands coercion yet. To do so, we want to express that an expression
+The above code allows to add two values of any numeric type (\code{Int}, \code{Rational}, \etc) using the \code{+}
+operator, but it does not handle operands coercion yet. To do so, we want to express that an expression
 \code{a + b} is valid either if \code{a} and \code{b} have the same numeric type, or if one of them has a numeric
@@
 
-The listing \ref{type-coercion} can be rewritten as shown in listing \ref{type-coercion-solution} with our DSL.
-The \code{+} operator can be safely used to add rational and floating-point numbers and type promotion is
-automatically handled by the type system. It is worth noting that the \code{Rational} data type can be written
-independently of the type coercion system.
-
+\begin{figure}
 \begin{lstlisting}[label=type-coercion-solution,caption=Type-coercion automatically handled by our DSL]
@@
 \end{lstlisting}
+\end{figure}
+
+Listing \ref{type-coercion} can be rewritten as shown in listing \ref{type-coercion-solution} with our DSL.
+The \code{+} operator can be safely used to add rational and floating-point numbers and type promotion is
+automatically handled by the type system. It is worth noting that the \code{Rational} data type can be written
+independently of the type coercion system.
 
@@
 program’s code is generated from this intermediate representation, with the opportunity to generate efficient code
-by applying domain specific optimizations. We think it’s worth considering a third step, which occurs before the
+by applying domain specific optimizations. We think it’s worth considering a third step that occurs before the
 program evaluation: its compilation. This step affects DSLs design because their embedded nature let them reuse the
@@
 All these benefits came with no runtime overhead, excepted for the asynchronous programming DSL that creates, for
-each asynchronous computation, a \code{Promise} value registering the callbacks to call when the value is available.
+each asynchronous computation, a reified asynchronous value registering the callbacks to call when the value is
+available.
 

### 2012-12-18 papers/ecoop2013/repairing.tex
Commit rapide

@@
 \begin{abstract}
-Rich Internet Applications involve even more code on the client-side, dealing with DOM manipulation, event handling
-and asynchronous calls to the server. Writing and maintaining large JavaScript code bases is challenging because this
-language has no static typing and no module system. We investigate further the definition of JavaScript as a Scala
-compiled embedded DSL using Lightweight Modular Staging. We define safe and expressive abstractions for DOM
-manipulation, event handling and asynchronous programming that translate to efficient JavaScript code using native
-APIs. We use both type-level information and staging to introduce high-level abstractions with few or no overhead on
-the generated JavaScript code.
+Rich Internet Applications rely on ever more code on the client-side, 
+typically written in JavaScript, dealing
+with DOM manipulation, event handling and asynchronous calls to the 
+server. Beyond well documented JavaScript programming intrinsic issues,
+the hard wired distribution of an application business logic across 
+clients and servers hampers its flexibility and maintainability. This 
+paper presents a Scala compiled
+embedded DSL for Web programming trying to address these shortcomings. 
+Close to JavaScript, it provides type-safe
+and expressive abstractions for DOM manipulation, event handling and
+asynchronous programming. Code written using the DSL can be compiled to both
+server and client sides. Built on top of the Lightweight Modular Staging
+framework, this DSL uses both type-level information and staging to  introduce
+high-level abstractions that translate to efficient JavaScript and Scala 
+code using native APIs.
 \end{abstract}
@@
 
-This paper continues their work and presents several safe and expressive DSL constructs for RIAs development. We
-increase the expressivity of JavaScript native APIs with no runtime overhead and expose them through statically typed
-APIs. We make asynchronous programming simpler by raising asynchronous computations to first-class citizens. We
-leverage the Scala type system to bring type-directed language features such as \emph{ad hoc} polymorphism and type
-coercion to the client-side.
+Based on this work, this paper introduces several type-safe and expressive DSL constructs for RIAs development. This
+DSL increases the expressivity of JavaScript native APIs with no runtime overhead and exposes them through statically
+typed APIs. It makes asynchronous programming simpler by raising asynchronous computations to first-class citizens.
+It leverages the Scala type system to bring type-directed language features such as \emph{ad hoc} polymorphism and
+type coercion to the client-side. Finally, code written using the DSL can be compiled to both server and client
+sides.
 
@@
 Dynamic typing may give programmers more flexibility but can make it harder for little scripts to grow into mature
-and robust code bases and to perform refactorings. For instance, in the case of RIAs, a mispell in an event name may
-break the program behavior without anyone getting a sensible error message.
+and robust code bases and to perform refactorings~\cite{Meijer04_StaticDynamic}. For instance, in the case of RIAs, a
+mispell in an event name may break the program behavior without anyone getting a sensible error message.
 
@@
 
-\subsubsection{Sharing type safe code between clients and servers is hard to achieve}
+\subsubsection{Sharing type-safe code between clients and servers is hard to achieve}
 
@@
 
-The following Scala listing generates code equivalent to listing \ref{event-js} and is completely type safe: if
+The following Scala listing generates code equivalent to listing \ref{event-js} and is completely type-safe: if
 the user misspells the event name or tries to use an undefined property on an event his program won’t compile.
@@
 
-The \code{\~} type is just an alias to the \code{Args} type to get the infix notation. The \code{Args} type
+The \code{\~} type is just an alias for the \code{Args} type to get the infix notation. The \code{Args} type
 aggregates the types of the operands (\code{Lhs} and \code{Rhs}) involved in a binary operation.
@@
 
-We limited the DSL to two operations (\code{foreach} and \code{isEmpty}) for the sake of brevity. The
+We limited the listing to two operations (\code{foreach} and \code{isEmpty}) for the sake of brevity. The
 \code{OptionOps} trait defines the user facing DSL, it enriches \code{Rep[Option[_]]} values by providing an implicit
@@
 
-Our goal was to define, with a minimal effort, a language for RIAs development. The main challange was to provide
-programming language constructs both safer and more expressive without abstraction penalty. We focused on the main
-concerns of RIA development: event handling, asynchronous programming and DOM fragments definition, as well as on
-more general programming concerns: \code{null} reference handling and type safety.
+Our goal was to define, with a minimal effort, a language for RIAs development. The main challenge was to provide
+programming language constructs that are both safer and more expressive without performance penalty. We focused on
+the main concerns of RIA development: event handling, asynchronous programming and DOM fragments definition, as well
+as on more general programming concerns: \code{null} reference handling and type safety.
 
@@
 simplify RIA development. It handles the \code{null} references problem by wrapping each query result in a container
-so before each further method call it tests the emptiness of the container and applies effectively the operation only
+so before each further method call it tests the emptiness of the container and applies the operation effectively only
 if the container is not empty. It provides an expressive API but the emptiness checking involves a slight performance
@@
 
-GWT~\cite{Chaganti07_GWT} compiles Java code to JavaScript. It exposes the browser’s APIs through type safe Java APIs
+GWT~\cite{Chaganti07_GWT} compiles Java code to JavaScript. It exposes the browser’s APIs through type-safe Java APIs
 and simplifies DOM fragments definition. However, it still uses a callback-driven programming model for asynchronous
@@
 
-Table \ref{comparison} gives a synthetic view of the comparison of our work with other approaches. Lines give
-features desired to write large Web applications and colums list mainstream tools for Web programming. Cells show the
-support level of each feature for each tool.
+Table \ref{comparison} gives a synthetic view of the comparison of our work with other mainstream approaches. Lines
+give features desired to write large Web applications and colums list mainstream tools for Web programming. Cells
+show the support level of each feature for each tool.
 
@@
 
+Several model-driven methodologies for creating web applications have been proposed, including
+\cite{schwabe1996systematic,lima2003modeling,ceri2000web,koch2001authoring,pastor2003oows,valverde2007mda,vdovjak2003engineering,kraus2007model,nunes2006rapid,brambilla2008designing,valderas2007transformational,van2006hera,Groenewegen08_WebDSL}.
+Many of them follow the standard top-down MDE vision in providing abstractions and code generators, nevertheless
+these cannot be easily customized or generate only application skeletons.
+
 \section{Conclusion and Perspectives}

### 2012-12-19 papers/ecoop2013/repairing.tex
morning commit

@@
 \end{tabular}
-\caption{Support level of Web programming features by mainstream tools. Zero star means that the feature is not supported
+\caption{Support of Web programming features by mainstream tools. Zero star means that the feature is not supported
 at all, three stars mean a strong support.}
@@
 Many of them follow the standard top-down MDE vision in providing abstractions and code generators, nevertheless
-these cannot be easily customized or generate only application skeletons.
+these cannot be easily customized or they generate only application skeletons.
 

### 2012-12-19 src/main/scala/scala/js/JSDom.scala
Workaround for SI-6135

@@ -6,9 +6,10 @@ trait JSDom { this: Base =>
 
   trait EventTarget
   implicit class EventTargetOps(t: Rep[EventTarget]) {
-    def on(event: EventDef, capture: Rep[Boolean] = unit(false))(handler: Rep[event.Type] => Rep[Unit])(implicit m: Manifest[event.Type]) = eventtarget_on(t, event, capture)(handler)
+    def on(event: EventDef, capture: Rep[Boolean] = unit(false))(handler: Rep[event.Type] => Rep[Unit])(implicit m: Manifest[event.Type]) = eventtarget_on(t, new EventName[event.Type](event.name), capture, handler)
   }
-  def eventtarget_on(t: Rep[EventTarget], event: EventDef, capture: Rep[Boolean])(handler: Rep[event.Type] => Rep[Unit])(implicit m: Manifest[event.Type]): Rep[Unit]
+  //def eventtarget_on(t: Rep[EventTarget], event: EventDef, capture: Rep[Boolean])(handler: Rep[event.Type] => Rep[Unit])(implicit m: Manifest[event.Type]): Rep[Unit]
+  def eventtarget_on[A : Manifest](t: Rep[EventTarget], e: EventName[A], capture: Rep[Boolean], handler: Rep[A] => Rep[Unit]): Rep[Unit]
 
   trait Event
   def infix_target(e: Rep[Event]): Rep[EventTarget]
@@ -67,10 +68,15 @@ trait JSDom { this: Base =>
 }
 
 trait JSDomExp extends JSDom with EffectExp {
-  def eventtarget_on(t: Exp[EventTarget], event: EventDef, capture: Exp[Boolean])(handler: Exp[event.Type] => Exp[Unit])(implicit m: Manifest[event.Type]) = {
+  /*def eventtarget_on(t: Exp[EventTarget], event: EventDef, capture: Exp[Boolean])(handler: Exp[event.Type] => Exp[Unit])(implicit m: Manifest[event.Type]) = {
     val e = fresh[event.Type]
     val block = reifyEffects(handler(e))
     reflectEffect(EventTargetOn(t, event, capture, e, block))
+  }*/
+  def eventtarget_on[A : Manifest](t: Exp[EventTarget], event: EventName[A], capture: Exp[Boolean], handler: Exp[A] => Exp[Unit]) = {
+    val e = fresh[A]
+    val b = reifyEffects(handler(e))
[…]

### 2012-12-20 papers/ecoop2013/repairing.tex
Some words on the asynchronous programming DSL implementation

@@
 \begin{abstract}
-Rich Internet Applications rely on ever more code on the client-side, 
-typically written in JavaScript, dealing
-with DOM manipulation, event handling and asynchronous calls to the 
-server. Beyond well documented JavaScript programming intrinsic issues,
-the hard wired distribution of an application business logic across 
-clients and servers hampers its flexibility and maintainability. This 
-paper presents a Scala compiled
-embedded DSL for Web programming trying to address these shortcomings. 
-Close to JavaScript, it provides type-safe
-and expressive abstractions for DOM manipulation, event handling and
-asynchronous programming. Code written using the DSL can be compiled to both
-server and client sides. Built on top of the Lightweight Modular Staging
-framework, this DSL uses both type-level information and staging to  introduce
-high-level abstractions that translate to efficient JavaScript and Scala 
-code using native APIs.
+Rich Internet Applications rely on ever more code on the client-side, typically written in JavaScript, dealing with
+DOM manipulation, event handling and asynchronous calls to the server. Beyond well documented JavaScript programming
+intrinsic issues, the hard wired distribution of an application business logic across clients and servers hampers its
+flexibility and maintainability. This paper presents a Scala compiled embedded DSL for Web programming trying to
+address these shortcomings. Close to JavaScript, it provides type-safe and expressive abstractions for DOM
+manipulation, event handling and asynchronous programming. Code written using the DSL can be compiled to both server
+and client sides. Built on top of the Lightweight Modular Staging framework, this DSL uses both type-level information
+and staging to  introduce high-level abstractions that translate to efficient JavaScript and Scala code using native
+APIs.
 \end{abstract}
@@
 
-Asynchronous values can be transformed using the \code{map} and \code{flatMap} methods, turning them into first-class
-citizens: functions can take them as parameters and return them.
+Asynchronous values can be used and transformed using the \code{foreach}, \code{map} and \code{flatMap} methods, turning them into first-class
+citizens: functions can take them as parameters and return them. Consequently, it is easier to write concurrent code:
+listing \ref{async-parallel-1} shows how to run two parallel asynchronous computations and to do something with their
+results when both are available. Writing equivalent code in pure JavaScript is tedious because it requires that the
+callback passed to each asynchronous computation checks if the other has been completed before, and to store the
+value, when availalbe, in a shared place (so both callbacks can get the other’s value).
 
 \begin{figure}
-\begin{multicols}{2}
 \begin{lstlisting}[caption=Parallel computations in Scala,label=async-parallel-1]
@@
 \end{lstlisting}
-\vfill
-\columnbreak
-\begin{lstlisting}[language=JavaScript,caption=Generated JavaScript code,label=async-parallel-2]
-var x1 = { cs: [] };
-AjaxGet("/foo", x1);
-var x2 = { cs: [] };
-AjaxGet("/bar", x2);
-foreach(x1, function (foo) {
- foreach(x2, function (bar) {
-  var x3 = foo + bar;
-  console.log(x3);
- });
-});
-\end{lstlisting}
-\end{multicols}
 \end{figure}
-
-Listing \ref{async-parallel-1} defines two asynchronous computations running in parallel and listing
-\ref{async-parallel-2} shows the generated JavaScript code (we renamed some identifiers for the sake of readability).
-The \code{x1} and \code{x2} values encode the asynchronous results of the Ajax requests, they contain an array of
-callbacks (\code{cs}, initially empty) to call when the result is available. The \code{AjaxGet} function code has
-been omitted, it creates a \code{XMLHttpRequest} object that sends a HTTP request and completes the value of its
-last parameter with the response, when received. Completing an asynchronous value means defining a \code{v} field
-on it containing the actual value, and calling all pending callbacks that are in the \code{cs} field. The
-\code{foreach} code has also been omitted, it registers a callback on an asynchronous value or calls the callback
-directly with the value if this one is already available.
 
@@
 \end{figure}
+
+\subsubsection{Taking advantage of the three transformation steps}
+
+The asynchronous programming DSL makes use of a combination of the three transformation steps described above.
+Asynchronous values are reified in values with a \code{put} and a \code{get} operations. The \code{put} operation
+sets the value of the asynchronous value and the \code{get} operation uses the asynchronous value when available.
+All other operations (\code{foreach}, \code{map} and \code{flatMap}) are implemented in terms of these two primitives.
+
+\begin{figure}
+\begin{lstlisting}[language=JavaScript,caption={Generated JavaScript code for listing \protect\ref{async-parallel-1}},label=async-parallel-2]
+var x1 = { v: null, cs: [] };
+AjaxGet("/foo", x1);
+var x2 = { v: null, cs: [] };
+AjaxGet("/bar", x2);
+var Get = function (f, callback) {
+  if (f.v === null) f.cs.push(callback)
+  else callback(f.v)
+};
+Get(x1, function (foo) {
+ Get(x2, function (bar) {
+  var x3 = foo + bar;
+  console.log(x3);
+ });
+});
+\end{lstlisting}
+\end{figure}
+
+Listing \ref{async-parallel-2} shows the generated JavaScript code for the Scala listing \ref{async-parallel-1} (we
+renamed some identifiers for the sake of readability). The \code{x1} and \code{x2} values reify the asynchronous
+results of the Ajax requests, they contain a field \code{v} holding the result value (initialized to \code{null}) and an
+array of callbacks \code{cs} (initially empty) to call when the result is available. The \code{AjaxGet} function code
+has been omitted, it creates a \code{XMLHttpRequest} object that sends a HTTP request and completes the value of its
+last parameter with the response, when received, using the \code{put} operation. The implementation of \code{put} sets
+the \code{v} field to the value and calls all the pending \code{cs} callbacks. The \code{Get} function shows the
+implementation of the \code{get} primitive. It takes two parameters, an asynchronous value and a callback. If the value
+is not yet available it adds the callback on the pending callback list, otherwise it calls it with the value as
+parameter.
+
+The transformation process of code using the asynchronous programming DSL can be summarized as follows. First, the
+compilation step desugars the \code{for} notation to calls to \code{foreach}, \code{map} and \code{flatMap}. These
+methods are implemented in terms of the \code{put} and \code{get} primitives, so they evaluate to an intermediate
+representation containing \code{Put} and \code{Get} nodes. Finally, these nodes are translated into the code actually
+implementing the asynchronous values reification.
 

### 2012-12-21 papers/ecoop2013/repairing.tex
packing and adding some words about Links

@@
 \begin{abstract}
+\vspace*{-0.7cm}
 Rich Internet Applications rely on ever more code on the client-side, typically written in JavaScript, dealing with
@@
 \end{abstract}
-
+\vspace*{-0.3cm}
 {\bf Keywords:} Web, programming languages, embedded DSL, type-level programming, staging, Scala
@@
 \section{Introduction}
-
+\vspace*{-0.2cm}
 The Web is an appealing platform on which to write applications (\aka{} Rich Internet Applications, RIA) because it
@@
 
-Opa\footnote{\href{http://opalang.org/}{http://opalang.org/}} is a language and a full stack designed for Web
-programming. It features static typing with global type inference, modules and pattern matching. The language
-provides a syntactic sugar for writing HTML fragments in the code. The server and client parts of an application can
-be written in the Opa language, the system can decide whether a code fragment will run on server-side, client-side or
-both. However, the Opa language is not extensible, so there is no way for developers to customize the compilation
-process of their abstractions on client and server sides.
+Opa\footnote{\href{http://opalang.org/}{http://opalang.org/}} and Links~\cite{Cooper07_Links} are languages designed
+for Web programming. They feature static typing with global type inference, modules and pattern matching. These
+languages provide a syntactic sugar for writing HTML fragments directly in the code. The server and client parts of
+an application are written in a single language and the system can decide whether a code fragment will run on
+server-side, client-side or both. However, these languages are not extensible, so there is no way for developers to
+customize the compilation process of their abstractions on client and server sides. Furthermore, hiding the code
+execution distribution gives programmers less control on communication error handling~\cite{Guerraoui99_Distributed}.
 

### 2013-04-25 papers/gpce2013/gpce2013.tex
Introduction Draft

\documentclass[preprint]{sigplanconf}

% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

\usepackage{amsmath}
\usepackage[utf8]{inputenc}
\usepackage[unicode=true,hidelinks]{hyperref}

\newcommand{\ie}{\emph{i.e.}}
\newcommand{\eg}{\emph{e.g.}}
\newcommand{\cf}{\emph{cf.~}}
\newcommand{\etal}{\emph{et al.~}}
\newcommand{\etc}{\emph{etc.}}
\newcommand{\aka}{\emph{a.k.a.}}

\begin{document}

\conferenceinfo{GPCE '13}{October 27--28, 2013, Indianapolis, Indiana, USA} 
\copyrightyear{2013}
\copyrightdata{[to be supplied]}

\titlebanner{banner above paper title}        % These are ignored unless
\preprintfooter{short description of paper}   % 'preprint' option specified.

\title{Title Text}
\subtitle{Subtitle Text, if any}

\authorinfo{Julien Richard-Foy\and Olivier Barais\and Jean-Marc Jézéquel}
           {IRISA, Université de Rennes 1}
           {\{firstname\}.\{lastname\}@irisa.fr}

\maketitle

\begin{abstract}
This is the text of the abstract.
\end{abstract}

\category{CR-number}{subcategory}{third-level}

\terms
term1, term2

\keywords
keyword1, keyword2

\section{Introduction}

(TODO more context)

Writing large Web applications is known to be
challenging~\cite{Mikkonen08_SpaghettiJs,Preciado05_RIAMethodologyNecessity} mainly because the business logic is
distributed into heterogeneous client-side and server-side
environments~\cite{Echeverria09_RIA,Kuuskeri09_PartitioningClientServer}. This makes it difficult to share code
between client-side and server-side (FIXME more on the motivation for code sharing?).

Using the same programming language on both server-side and client-side could reduce this gap, however the JavaScript
language -- which is currently the only action language natively supported by all Web browsers -- has several
drawbacks making it hardly suitable for large code bases (\eg no static typing, no module system, verbose syntax,
\etc). An increasing number of programming languages support generating JavaScript (\eg GWT,
SharpKit\footnote{\href{http://sharpkit.net}{http://sharpkit.net}}, Dart,
Kotlin\footnote{\href{http://kotlin.jetbrains.org/}{http://kotlin.jetbrains.org/}},
ClojureScript~\cite{McGranaghan11_ClojureScript}, Fay\footnote{\href{http://fay-lang.org/}{http://fay-lang.org/}},
Haxe~\cite{Cannasse08_HaXe} or Opa\footnote{\href{http://opalang.org/}{http://opalang.org/}}), increasing the panel
of programming languages available to write both the client-side and the server-side code of a Web application.

Having a full featured, cutting edge, programming language that can be run on both client-side and server-side can
help developers to write more maintainable code, however by abstracting over the differences of the client-side and
the server-side environments, the code may suffer from performance issues. Performance is a primary concern in Web
applications, because they are supposed to run on a broad range of devices, from the powerful desktop personal
computer to the less powerful smartphone. “every 100~ms delay costs 1\% of sales”, said Amazon in 2006.

For instance, the boundaries of the code parts to emit on client-side are less visible when you share code between
client-side and server-side so transitive dependencies may pull a lot of code, causing a high download overhead.
Moreover, generating efficient code for heterogeneous platforms is hard to achieve in an extensible way. (TODO Give a
concrete example for each problem)

On one hand, for engineering reasons, developers want to write Web applications using a single programming language,
abstracting over the target platforms differences. But on the other hand, for performance reasons, they want to keep
control on the way their code is compiled to each target platform. How to solve this dilemma?

Lightweight Modular Staging~\cite{Rompf12_LMSThesis} is a framework for defining deeply embedded DSLs in Scala. It
has been used to define high-performance DSLs for parallel computing~\cite{Brown11_Parallel} and can be used to
generate JavaScript code~\cite{Kossakowski12_JsDESL}.

On top of this previous work, this paper presents several new APIs (?) that can be shared between client-side and
server-side code and/or that are efficiently translated on each target platform. More precisely, our contributions
are:

\begin{itemize}
 \item Type-directed ad-hoc polymorphism on client-side without runtime dynamic dispatch logic;
 \item Use monads without extra container object creation;
 \item Ability to define DOM fragments using a common language for server-side and client-side, but that generates
code using standard APIs on both server-side and client-side ;
 \item An API for searching in the DOM, that exposes a single entry point but that generates code potentially using
more optimized native APIs
\end{itemize}

The remainder of this paper is organized as follows. The next section introduces existing approaches for defining
cross-compiling languages. Section \ref{contribution} presents our contribution. Section \ref{validation} evaluates
our contribution. Section \ref{discussion} concludes.

\section{Related Works}

\subsection{Fat Languages}

\subsection{Thin Languages}

\subsection{Deeply Embedded Languages}

\section{High-Level Abstractions Generating Efficient (and Heterogeneous) Code}
\label{contribution}

\subsection{Ad-Hoc Polymorphism}

\subsection{Monads Sequencing}

\subsection{DOM Fragments}

\subsection{Selectors}

\section{Implementation}

\section{Evaluation}
\label{validation}

\subsection{Real World Application}

Chooze.

\subsection{Several Implementations}

\subsubsection{Vanilla JavaScript}

\subsubsection{jQuery}

\subsubsection{GWT}

\subsubsection{SharpKit}

\subsubsection{Js-Scala}

\subsection{Benchmarks, Code Metrics}

\section{Conclusion, Future Work}
\label{discussion}

%\appendix
%\section{Appendix Title}
%
%This is the text of the appendix, if you need one.
%
\acks

Acknowledgments, if needed.

\bibliographystyle{abbrvnat}
\bibliography{biblio}
%\begin{thebibliography}{}
%\softraggedright
%
%\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
%P. Q. Smith, and X. Y. Jones. ...reference text...
%
%\end{thebibliography}
%
\end{document}

### 2013-04-25 papers/sle2013/sle2013.tex
sle directory

--- 2012-07-13 papers/ecoop2012/template.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%% file typeinst.tex %%%%%%%%%%%%%%%%%%%%%%%%%
-%
-% This is the LaTeX source for the instructions to authors using
-% the LaTeX document class 'llncs.cls' for contributions to
-% the Lecture Notes in Computer Sciences series.
-% http://www.springer.com/lncs       Springer Heidelberg 2006/05/04
-%
-% It may be used as a template for your own input - copy it
-% to a new file with a new name and use it as the basis
-% for your article.
-%
-% NB: the document class 'llncs' has its own and detailed documentation, see
-% ftp://ftp.springer.de/data/pubftp/pub/tex/latex/llncs/latex2e/llncsdoc.pdf
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \documentclass[runningheads,a4paper]{llncs}
@@
 \usepackage{url}
+\urldef{\mailsa}\path|{alfred.hofmann, ursula.barth, ingrid.haas, frank.holzwarth,|
+\urldef{\mailsb}\path|anna.kramer, leonie.kunz, christine.reiss, nicole.sator,|
+\urldef{\mailsc}\path|erika.siebert-cole, peter.strasser, lncs}@springer.com|    
 \newcommand{\keywords}[1]{\par\addvspace\baselineskip
 \noindent\keywordname\enspace\ignorespaces#1}
-
-\input{macros}
 
@@
 
-\title{JavaScript as an embedded DSL in Scala}
+% first the title is needed
+\title{Lecture Notes in Computer Science:\\Authors' Instructions
+for the Preparation\\of Camera-Ready
+Contributions\\to LNCS/LNAI/LNBI Proceedings}
+
 % a short form should be given in case it is too long for the running head
-%\titlerunning{}
+\titlerunning{Lecture Notes in Computer Science: Authors' Instructions}
 
@@
 %
-\author{Grzegorz Kossakowski \and Nada Amin \and Tiark Rompf \and Martin Odersky}
+\author{Alfred Hofmann%
+\thanks{Please note that the LNCS Editorial assumes that all authors have used
+the western naming convention, with given names preceding surnames. This determines
+the structure of the names in the running heads and the author index.}%
+\and Ursula Barth\and Ingrid Haas\and Frank Holzwarth\and\\
+Anna Kramer\and Leonie Kunz\and Christine Rei\ss\and\\
+Nicole Sator\and Erika Siebert-Cole\and Peter Stra\ss er}
 %
-\authorrunning{JavaScript as an embedded DSL in Scala}
+\authorrunning{Lecture Notes in Computer Science: Authors' Instructions}
 % (feature abused for this document to repeat the title also on left hand pages)
@@
 % unless you accept that it will be published
-\institute{Ecole Polytechnique F\'ed\'erale de Lausanne,\\
-\email{\{first.last\}@epfl.ch}}
-%\mailsa\\
-%\mailsb\\
-%\url{http://lamp.epfl.ch}}
+\institute{Springer-Verlag, Computer Science Editorial,\\
+Tiergartenstr. 17, 69121 Heidelberg, Germany\\
+\mailsa\\
+\mailsb\\
+\mailsc\\
+\url{http://www.springer.com/lncs}}
 
@@
 
-%\toctitle{}
-%\tocauthor{}
+\toctitle{Lecture Notes in Computer Science}
+\tocauthor{Authors' Instructions}
 \maketitle
@@
 \begin{abstract}
-% =>70 & <=150 words
-\note{TODO}
-
-\keywords{JavaScript, Scala, DSL, programming languages}
+The abstract should summarize the contents of the paper and should
+contain at least 70 and at most 150 words. It should be written using the
+\emph{abstract} environment.
+\keywords{We would like to encourage you to list your keywords within
+the abstract section}
 \end{abstract}
 
-We use Lightweight Modular Staging (LMS) to create a DSL in Scala which generates JavaScript code.
-
-Our approach addresses many problems encountered in web development in an elegant, coherent and lightweight way. Programming directly in JavaScript include the following problems:
-\begin{itemize}
-\item lack of static typing, which makes it difficult to maintain, refactor and develop large web applications,
-\item impedance mismatch between client and server code,
-\item lack of modularity and abstractions,
-\item difficult to reason about control structures that arise from high-performance asynchronous callback-driven programming style.
-\end{itemize}
-Frameworks such as GWT and Links which generates JavaScript solve some of the issues, but introduce new problems, including:
-\begin{itemize}
-\item lack of integration with existing JavaScript libraries,
-\item difficult to extend with new APIs, let alone new language primitives,
-\item little control over generated code which makes performance-tuning very difficult,
-\item no first-class status for generated code, which makes it impossible to compose and abstract over snippets.
-\end{itemize}
-
-Our approach is lightweight: By embedding our DSL as a library in Scala, we exploit the features of the host language such as expressive static typing, easy creation of new control structures, and even partial CPS transformation.
-
-\section{Introduction to LMS}
-- virtualization
-- rep types
-- IR construction
-- code generation
-advantage: can have abstractions that only exist at staging-time
-
-\section{Gradual Typing}
-- Scala Dynamic
-- use of external APIs
-- snippets with a simple example showing evolution from dynamic to typed (Canvas example which will become Koch example in the next section)
-
-\section{Trivial Embedding}
-- DOM Canvas Koch example: generate both client-side code or desktop widget
-
-\section{Modularity Interlude}
-- modularity at many levels
-   e.g. dynamic feature is just a trait
-   users specify external APIs with traits
-   interpretation of rep also done via traits (generating IR nodes vs trivial embedding)
-- example of user adding a language primitive: e.g. for-comprehension
-- mix and match features by composing traits (e.g. everything available on main thread 'JS trait', vs no-DOM manipualtion in worker threads 'WorkerJS trait)
-
-\section{Reification of Objects}
-- JS Literals: intuitive translation to JS as useful JSON-like objects, static typing
-- mice example: shows use of JSLiterals + API gluing (create abstractions that you need, not necessarily exactly matching the underlying JS API)
-- reification of traits: mutability, inheritance
-
-\section{Case Study: CPS transformation to simply asynchronous web development}
-
-\clearpage
 \section{Introduction}
@@
 \centering
-%\includegraphics[height=6.2cm]{eijkel2}
+\includegraphics[height=6.2cm]{eijkel2}
 \caption{One kernel at $x_s$ (\emph{dotted kernel}) or two kernels at

### 2013-05-03 papers/gpce2013/gpce2013.tex
Slight change in the intro

@@
 
-(TODO more context)
+Rich Internet Applications have interesting properties: they require no installation or deployment on
+clients and enables large scale collaborative experiences. However, writing large Web applications is known to be
+difficult~\cite{Mikkonen08_SpaghettiJs,Preciado05_RIAMethodologyNecessity}. One challenge comes from the fact
+that the business logic is scattered into heterogeneous client-side and server-side
+environments~\cite{Echeverria09_RIA,Kuuskeri09_PartitioningClientServer}. For instance, HTML fragments may be
+built from the server-side when a page is requested by a client, but they may also be built from the client-side to
+perform an incremental update subsequent to a user action. How could developers write HTML fragment definitions once
+and render them on both client-side and server-side? The more interactive the application is, the more logic needs to
+be duplicated between the server-side and the client-side.
 
-Writing large Web applications is known to be
-challenging~\cite{Mikkonen08_SpaghettiJs,Preciado05_RIAMethodologyNecessity} mainly because the business logic is
-distributed into heterogeneous client-side and server-side
-environments~\cite{Echeverria09_RIA,Kuuskeri09_PartitioningClientServer}. This makes it difficult to share code
-between client-side and server-side (FIXME more on the motivation for code sharing?).
-
-Using the same programming language on both server-side and client-side could reduce this gap, however the JavaScript
-language -- which is currently the only action language natively supported by all Web browsers -- has several
-drawbacks making it hardly suitable for large code bases (\eg no static typing, no module system, verbose syntax,
-\etc). An increasing number of programming languages support generating JavaScript (\eg GWT,
-SharpKit\footnote{\href{http://sharpkit.net}{http://sharpkit.net}}, Dart,
+Using the same programming language on both server-side and client-side could reduce the gap between client-side and
+server-side environments, however the JavaScript language -- which is currently the only action language natively
+supported by all Web browsers -- has several drawbacks making it hardly suitable for large code bases (\eg no static
+typing, no module system, verbose syntax, \etc). An increasing number of programming languages support generating
+JavaScript (\eg GWT, SharpKit\footnote{\href{http://sharpkit.net}{http://sharpkit.net}}, Dart,
 Kotlin\footnote{\href{http://kotlin.jetbrains.org/}{http://kotlin.jetbrains.org/}},
@@
 
-Having a full featured, cutting edge, programming language that can be run on both client-side and server-side can
-help developers to write more maintainable code, however by abstracting over the differences of the client-side and
-the server-side environments, the code may suffer from performance issues. Performance is a primary concern in Web
+Having a full featured, cutting edge, programming language that runs on both client-side and server-side can help
+developers to write more maintainable code, however by abstracting over the differences of the client-side and the
+server-side environments, the code may suffer from performance issues. Performance is a primary concern in Web
 applications, because they are supposed to run on a broad range of devices, from the powerful desktop personal
-computer to the less powerful smartphone. “every 100~ms delay costs 1\% of sales”, said Amazon in 2006.
+computer to the less powerful smartphone. “Every 100~ms delay costs 1\% of sales”, said Amazon in 2006.
 

### 2013-05-06 papers/gpce2013/gpce2013.tex
intro

@@
 \title{Title Text}
-\subtitle{Subtitle Text, if any}
+\subtitle{You don’t have to trade abstraction for control}
 
@@
 
-Rich Internet Applications have interesting properties: they require no installation or deployment on
-clients and enables large scale collaborative experiences. However, writing large Web applications is known to be
+Web applications are attractive because they require no installation or deployment on clients and enable large scale
+collaborative experiences. However, writing large Web applications is known to be
 difficult~\cite{Mikkonen08_SpaghettiJs,Preciado05_RIAMethodologyNecessity}. One challenge comes from the fact
 that the business logic is scattered into heterogeneous client-side and server-side
-environments~\cite{Echeverria09_RIA,Kuuskeri09_PartitioningClientServer}. For instance, HTML fragments may be
-built from the server-side when a page is requested by a client, but they may also be built from the client-side to
-perform an incremental update subsequent to a user action. How could developers write HTML fragment definitions once
-and render them on both client-side and server-side? The more interactive the application is, the more logic needs to
-be duplicated between the server-side and the client-side.
+environments~\cite{Echeverria09_RIA,Kuuskeri09_PartitioningClientServer}. This gives less flexibility in the
+engineering process and requires a higher maintainance effort: once you decided to implement a feature on
+client-side, changing your *mind* means completely rewriting the feature on server-side (and \emph{vice versa}).
+Even worse, logic parts that need to run on both client-side and server-side are duplicated. For instance HTML
+fragments may be built from the server-side when a page is requested by a client, but they may also be built from the
+client-side to perform an incremental update subsequent to an user action. How could developers write HTML fragment
+definitions once and render them on both client-side and server-side? The more interactive the application is, the
+more logic needs to be duplicated between the server-side and the client-side (explain why?).
 
-Using the same programming language on both server-side and client-side could reduce the gap between client-side and
-server-side environments, however the JavaScript language -- which is currently the only action language natively
-supported by all Web browsers -- has several drawbacks making it hardly suitable for large code bases (\eg no static
-typing, no module system, verbose syntax, \etc). An increasing number of programming languages support generating
-JavaScript (\eg GWT, SharpKit\footnote{\href{http://sharpkit.net}{http://sharpkit.net}}, Dart,
-Kotlin\footnote{\href{http://kotlin.jetbrains.org/}{http://kotlin.jetbrains.org/}},
+Using the same programming language on both server-side and client-side can improve the software engineering
+process by enabling code reuse between both sides. Incidentally, the JavaScript language -- which is currently the
+only action language natively supported by almost all Web clients -- can be used on server-side, and an increasing
+number of programming languages or compiler backends can generate JavaScript code (\eg
+Java/GWT~\cite{Chaganti07_GWT}, SharpKit\footnote{\href{http://sharpkit.net}{http://sharpkit.net}},
+Dart~\cite{Griffith11_Dart}, Kotlin\footnote{\href{http://kotlin.jetbrains.org/}{http://kotlin.jetbrains.org/}},
 ClojureScript~\cite{McGranaghan11_ClojureScript}, Fay\footnote{\href{http://fay-lang.org/}{http://fay-lang.org/}},
-Haxe~\cite{Cannasse08_HaXe} or Opa\footnote{\href{http://opalang.org/}{http://opalang.org/}}), increasing the panel
-of programming languages available to write both the client-side and the server-side code of a Web application.
+Haxe~\cite{Cannasse08_HaXe} or Opa\footnote{\href{http://opalang.org/}{http://opalang.org/}}).
 
-Having a full featured, cutting edge, programming language that runs on both client-side and server-side can help
-developers to write more maintainable code, however by abstracting over the differences of the client-side and the
-server-side environments, the code may suffer from performance issues. Performance is a primary concern in Web
-applications, because they are supposed to run on a broad range of devices, from the powerful desktop personal
-computer to the less powerful smartphone. “Every 100~ms delay costs 1\% of sales”, said Amazon in 2006.
+However, this engineering comfort may come at the price of an inefficient runtime: abstracting over platform
+differences often means restricting to a subset of common features and losing opportunities to perform platform
+specific optimizations. Performance is a primary concern in Web applications, because they are expected to run on a
+broad range of devices, from the powerful desktop personal computer to the less powerful smartphone. “Every 100~ms
+delay costs 1\% of sales”, said Amazon in 2006. For instance, because the boundaries of the code sent to the client
+are less visible when you share code between client-side and server-side, transitive dependencies may pull a lot of
+code on the client, causing a high download overhead. Moreover, generating efficient code for heterogeneous platforms
+is hard to achieve in an extensible way: the translation of common abstractions like collections into their native
+counterpart (JavaScript arrays on client-side and standard library’s collections on server-side) may be hardcoded in
+the compiler, but that would not scale to handle all the abstractions a complete application may use (\eg HTML
+fragment definitions, form validation rules, or even some business data type that may be represented differently for
+performance reasons).
 
-For instance, the boundaries of the code parts to emit on client-side are less visible when you share code between
-client-side and server-side so transitive dependencies may pull a lot of code, causing a high download overhead.
-Moreover, generating efficient code for heterogeneous platforms is hard to achieve in an extensible way. (TODO Give a
-concrete example for each problem)
-
-On one hand, for engineering reasons, developers want to write Web applications using a single programming language,
-abstracting over the target platforms differences. But on the other hand, for performance reasons, they want to keep
+On one hand, for engineering reasons, developers want to write Web applications using a single language, abstracting
+over the target platforms differences. But on the other hand, for performance reasons, they want to keep
 control on the way their code is compiled to each target platform. How to solve this dilemma?
 
-Lightweight Modular Staging~\cite{Rompf12_LMSThesis} is a framework for defining deeply embedded DSLs in Scala. It
-has been used to define high-performance DSLs for parallel computing~\cite{Brown11_Parallel} and can be used to
-generate JavaScript code~\cite{Kossakowski12_JsDESL}.
+Compiled domain specific embedded languages~\cite{Elliott2003_Compiling} allow the definition of domain specific
+languages (DSLs) as libraries on top of a host language, and to compile them to a target platform. The deep embedding
+gives the opportunity to control the code generation scheme for a given abstraction and target platform.
 
-On top of this previous work, this paper presents several new APIs (?) that can be shared between client-side and
-server-side code and/or that are efficiently translated on each target platform. More precisely, our contributions
-are:
+This paper presents such a DSL allowing developers to write Web applications in a single language which code
+fragments can be shared between client and server sides, and which is efficiently compiled to each side. More
+precisely, we demonstrate the following features:
 
@@
  \item Type-directed ad-hoc polymorphism on client-side without runtime dynamic dispatch logic;
- \item Use monads without extra container object creation;
+ \item Usage of monads without extra container object creation;
  \item Ability to define DOM fragments using a common language for server-side and client-side, but that generates
-code using standard APIs on both server-side and client-side ;
+code using standard APIs on both server-side and client-side;
  \item An API for searching in the DOM, that exposes a single entry point but that generates code potentially using
-more optimized native APIs
+more optimized native APIs.
 \end{itemize}
@@
 
-\section{Related Works}
+\section{Related Work}
 
@@
 \subsection{Deeply Embedded Languages}
+
+Lightweight Modular Staging~\cite{Rompf12_LMSThesis} is a framework for defining deeply embedded DSLs in Scala. It
+has been used to define high-performance DSLs for parallel computing~\cite{Brown11_Parallel} and can be used to
+generate JavaScript code~\cite{Kossakowski12_JsDESL}.
 
@@
 
-\section{Implementation}
+\section{Implementation?}
 

### 2013-05-13 papers/gpce2013/gpce2013.tex
pwet

@@
 \usepackage[unicode=true,hidelinks]{hyperref}
+
+\input{macros}
 
@@
 
-Web applications are attractive because they require no installation or deployment on clients and enable large scale
-collaborative experiences. However, writing large Web applications is known to be
+Web applications are attractive because they require no installation or deployment steps on clients and enable large
+scale collaborative experiences. However, writing large Web applications is known to be
 difficult~\cite{Mikkonen08_SpaghettiJs,Preciado05_RIAMethodologyNecessity}. One challenge comes from the fact
@@
 client-side, changing your *mind* means completely rewriting the feature on server-side (and \emph{vice versa}).
-Even worse, logic parts that need to run on both client-side and server-side are duplicated. For instance HTML
+Even worse, logic parts that need to run on both client-side and server-side are duplicated. For instance, HTML
 fragments may be built from the server-side when a page is requested by a client, but they may also be built from the
@@
 process by enabling code reuse between both sides. Incidentally, the JavaScript language -- which is currently the
-only action language natively supported by almost all Web clients -- can be used on server-side, and an increasing
+action language natively supported by most of Web clients -- can be used on server-side, and an increasing
 number of programming languages or compiler backends can generate JavaScript code (\eg
@@
 
-This paper presents such a DSL allowing developers to write Web applications in a single language which code
-fragments can be shared between client and server sides, and which is efficiently compiled to each side. More
-precisely, we demonstrate the following features:
+This paper presents such a compiled embedded DSL allowing developers to write Web applications in a single language
+which code fragments can be shared between client and server sides, and which is efficiently compiled to each side.
+More precisely, we demonstrate the following features:
 
@@
 
+Because of the dynamically typed nature of JavaScript, when calling a function there is no proper way to select a
+specialized implementation according to the function’s parameters types. JavaScript is only able to dispatch
+according to a method receiver prototype, \eg{} if one writes \code{foo.bar()} the JavaScript runtime will look into
+the prototype of the \code{foo} object for a property named \code{bar} and will call it. So, the only way to achieve
+\emph{ad hoc} polymorphism on JavaScript objects consists in defining the polymorphic function on the prototypes of
+the objects. However, modifying existing object prototypes is considered bad
+practice~\cite{Zakas12_MaintainableJs}. Another way could consist in manually coding the dispatch logic, by
+registering supported data types at the beginning of the program execution, as described in section 2.4.3
+of~\cite{Abelson83_SICP}, but this solution is painful for developers and incurs a performance overhead.
+
+We propose to achieve \emph{ad hoc} polymorphism using
+typeclasses~\cite{Wadler89_AdhocPolymorphism,Odersky06_Typeclasses,Oliveira10_Typeclasses} so that it supports
+retroactive extension without modifying objects prototypes because it is type-directed: the dispatch happens at
+compile-time rather than at runtime.
+
+\begin{figure}
+\begin{lstlisting}[label=polymorphism,caption=Ad hoc polymorphism using typeclasses]
+// Interface
+case class Show[A](show: Rep[A => Node])
+
+// Polymorphic function
+def listWidget[A : Show](items: Rep[List[A]]): Rep[Node] = {
+  tag("ul")(
+    for (item <- items) yield {
+      tag("li")(implicitly[Show[A]].show(item))
+    }
+  )
+}
+
+// Type `User`
+type User = Record { val name: String; val age: Int }
+// Implementation of Show for a User
+implicit val showUser = Show[User] { user =>
+  tag("span", "class"->"user")(
+    text(user.name + "(" + user.age + " years)")
+  )
+}
+
+// Type `Article`
+type Article = Record { val name: String; val price: Double }
+// Implementation of Show for an Article
+implicit val showArticle = Show[Article] { article =>
+  tag("span")(
+    text(article.name),
+    tag("strong")(text(article.price + " Euros"))
+  )
+}
+
+// Main program
+def main(users: Rep[List[User]], articles: Rep[List[Article]]) = {
+  document.body.append(listWidget(users))
+  document.body.append(listWidget(articles))
+}
+\end{lstlisting}
+\end{figure}
+
+Listing \ref{polymorphism} demonstrates how to define a polymorphic \code{listWidget} function that returns a DOM
+tree containing the representation of a list of items. The \code{Show[A]} typeclass defines how to produce a DOM tree
+for a value of type \code{A}. It is used by the \code{listWidget} function to get the DOM fragments of the list
+items. The listing shows how to reuse the same \code{listWidget} function to show a list of users and a list of
+articles.
+
 \subsection{Monads Sequencing}
+
+The previous work on js-scala \cite{Kossakowski12_JsDESL} showed how to make asynchronous programming more
+convenient by making asynchronous calls looking like synchronous calls (\ie{} returning a value instead of taking a
+callback as parameter). This work was based on the Scala continuations compiler plugin. We claim that, although this
+programming model removes the “callback hell”, it can make the code hard to reason about because there is no explicit
+distinction between synchronous and asynchronous computations.
+
+We propose a DSL that explicitly reflects the asynchronous nature of computations in their types and provides methods
+turning them into first-class citizens. The DSL is monadic so we can solve the callback hell problem thanks to the
+Scala \code{for} notation.
+
+For instance, listing \ref{async-first-class} shows how the listing \ref{async-js} can be re-written using our DSL.
+The \code{createVertex} function now returns an asynchronous value instead of taking a callback as
+parameter. Then, the \code{for} expression allows us to get the vertex, when available, and to insert it on the user
+interface. By making the \code{createVertex} function return an asynchronous value instead of taking a callback as
+a parameter, the code is easier to modularize into loosely coupled parts.
+
+\begin{figure}
+\centering
+\begin{lstlisting}[caption=Asynchronous values are first class citizen,label=async-first-class]
+def createVertex(text: Rep[String]): Rep[Future[Vertex]] =
+  Ajax.post[Vertex]("/create", new Record { val content = text })
+
+for (vertex <- createVertex("Hello, World!")) {
+  addVertex(vertex)
+}
+\end{lstlisting}
+\end{figure}
+
+\begin{figure}
+\centering
+\begin{lstlisting}[caption=No callback hell,label=async-no-callback]
+for {
+  foo <- Ajax.get(fooUrl())
+  bar <- Ajax.get(barUrl(foo))
+  _ <- future(println("bar = " + bar))
+  baz <- Ajax.get(bazUrl(bar))
+} println(foo + bar + baz)
+\end{lstlisting}
+\end{figure}
+
+Listing \ref{async-no-callback} translates the callback hell example (listing \ref{async-callback-hell}) using
+our DSL. The \code{for} notation can intuitively be thought of as a sequencing notation: whenever the response of the
+first Ajax request is available, the next statement is executed, and so on. There is no nested callbacks and the
+order of execution is directly reflected by the order of statements.
+
+An asynchronous value of type \code{Rep[A]} is modelled by a value of type \code{Rep[Future[A]]}. Because Scala’s
+\code{for} notation is just syntactic sugar for methods \code{foreach}, \code{map} and \code{flatMap}, we are
+able to define our DSL by just defining these methods on \code{Rep[Future[A]]} values, with the following semantic:
+\begin{itemize}
+\item \code{foreach(f: Rep[A] => Rep[Unit]): Rep[Unit]}, eventually does something with the value when available~;
+\item \code{map(f: Rep[A] => Rep[B]): Rep[Future[B]]}, eventually transforms the value when available~;
+\item \code{flatMap(f: Rep[A] => Rep[Future[B]]): Rep[Future[B]]} also eventually transforms the value when
+available~;
+\end{itemize}
+
+Asynchronous values can be used and transformed using the \code{foreach}, \code{map} and \code{flatMap} methods,
+turning them into first-class
+citizens: functions can take them as parameters and return them. Consequently, it is easier to write concurrent code:
+listing \ref{async-parallel-1} shows how to run two parallel asynchronous computations and to do something with their
+results when both are available. Writing equivalent code in pure JavaScript is tedious because it requires that the
+callback passed to each asynchronous computation checks if the other has been completed before, and to store the
+value, when availalbe, in a shared place (so both callbacks can get the other’s value).
+
+\begin{figure}
+\begin{lstlisting}[caption=Parallel computations in Scala,label=async-parallel-1]
+val fooAsync = Ajax.get("/foo")
+val barAsync = Ajax.get("/bar")
+for {
+  foo <- fooAsync
+  bar <- barAsync
+} println(foo + bar)
+\end{lstlisting}
+\end{figure}
+
+As an illustration of the staging mechanism, we present a simple DSL to handle null references. This DSL provides an
+abstraction at the stage-level that is removed by optimization during the code generation.
+
+Null references are a known source of problems in programming languages~\cite{Hoare09_Null,Nanda09_Null}. For
+example, consider the following typical JavaScript code finding a particular widget in the page and a then particular
+button in the widget:
+
+\begin{lstlisting}[language=JavaScript,label=null-unsafe,caption=Unsafe code]
+var loginWidget = document.querySelector("div.login");
+var loginButton = loginWidget.querySelector("button.submit");
+loginButton.addEventListener("click", function (e) { ... });
+\end{lstlisting}
+
+The native \code{querySelector} method returns \code{null} if no node matched the given selector in the document. If
+we run the above code in a page where the widget is not present, it will throw an error and stop further JavaScript
+execution. We can write defensive code to handle \code{null} cases, but it leads to very cumbersome code:
+
+\begin{lstlisting}[language=JavaScript,label=null-defensive,caption=Defensive programming to handle null references]
+var loginWidget = document.querySelector("div.login");
+if (loginWidget !== null) {
+  var loginButton = loginWidget.querySelector("button.submit");
+  if (loginButton !== null) {
+    loginButton.addEventListener("click", function (e) { ... });
+  }
+}
+\end{lstlisting}
+
+We want to define a DSL that has both the safety and performance of listing \ref{null-defensive} but the
+expressiveness of listing \ref{null-unsafe}. We can get safety by wrapping potentially null values of type
+\code{Rep[A]} in a container of type \code{Rep[Option[A]]} requiring explicit dereferencing, we can get
+expressiveness by using the Scala \code{for} notation for dereferencing, and finally we can get performance by
+generating code that does not actually wraps values in a container but instead checks if they are \code{null} or not
+when dereferenced. The wrapping container exists only at the stage-level and is removed during the code generation.
+Here is a Scala listing that uses our DSL (implementation details are given in section \ref{implementation}):
+
+\begin{lstlisting}
+for {
+  loginWidget <- document.find("div.login")
+  loginButton <- loginWidget.find("submit.button")
+} loginButton.on(Click) { e => ... }
+\end{lstlisting}
+
+The evaluation of the above listing produces a graph of statements from which JavaScript code equivalent to
+listing \ref{null-defensive} is generated.
 
@@
 
+In this section we show how we can define a template engine as an embedded DSL with minimal effort. This template
+engine is statically typed and able to insert dynamic content in a safe way. It provides a powerful expression
+language, requires no extra compilation step and can be used on both client-side and server-side.
+
+Because the template engine is defined as en embedded DSL, we can reuse Scala’s constructs:
+
+\begin{itemize}
+\item a function taking some parameters and returning a DOM fragment directly models a template taking parameters and
+returning a DOM fragment~;
+\item the type system typechecks template definitions and template calls~;
+\item the Scala language itself is the expression language~;
+\item compiling a template is the same as compiling user code.
+\end{itemize}
+
+So the only remaining work consists in defining the DSL vocabulary to define DOM nodes. We provide a \code{tag}
+function to define a tag and a \code{text} function to define a text node.
+
+\begin{figure}
+\begin{lstlisting}[label=forest-hello,caption=DOM definition DSL]
+def vertexDom(v: Rep[Vertex]) =
+    tag("g", "class"->"vertex",
+             "transform"->("translate("+v.x+","+v.y+")"))(
+        tag("rect", "width"->v.width, "height"->v.height)(),
+        tag("text", "width"->v.width, "height"->v.height)(
+            text(v.content)
+        )
+    )
+\end{lstlisting}
+\end{figure}
+
+Listing \ref{forest-hello} uses our DSL and generates a code equivalent to listing \ref{dom-api}. The readability has
+been highly improved: nesting tags is just like nesting code blocks, HTML entities are
+automatically escaped in text nodes, developers have the full computational power of Scala to inject dynamic data and
+DOM fragments definitions are written using functions so they compose just as functions compose. These benefits come
+with no performance loss because the DSL generates code building DOM fragments by using the native JavaScript API.
+
+\subsubsection{Reuse the DOM definition DSL on server-side}
+
+Our DSL is equivalent to a template engine with Scala as the expression language. Making it usable on both server and
+client sides was surprisingly as simple as defining another code generator for the DSL, producing Scala code.
+
+For instance, the template written in listing \ref{forest-hello} produces the following Scala code usable on
+server-side (the generated code for client-side is roughly equivalent to listing \ref{dom-api}):
+
+\begin{lstlisting}
+def vertexDom(v: Vertex) = {
+  val x0 =
+    <text width="{v.width}" height="{v.height}">
+      {v.content}
+    </text>
+  val x1 = <rect width="{v.width}" height="{v.height}" />
+  val x2 =
+    <g class="vertex" transform="translate({v.x},{v.y})">
+      {List(x0, x1)}
+    </g>
+  x2
+}
+\end{lstlisting}
+
+We are able to tackle the code sharing issues described in section \ref{assessment} because of the embdedded nature
+of our DSLs: dynamic content of templates is written using embedded DSLs too, so their translation into JavaScript
+and Scala is managed by their respective code generators.
+
 \subsection{Selectors}
+
+querySelectorAll, getElementsByTagName, getElementsByClassName, getElementById, querySelector
 

### 2013-05-13 papers/sle2013/sle2013.tex
pwet

@@
 
+\usepackage[utf8]{inputenc}
+\usepackage[unicode=true,hidelinks]{hyperref}
+
 \usepackage{url}
-\urldef{\mailsa}\path|{alfred.hofmann, ursula.barth, ingrid.haas, frank.holzwarth,|
-\urldef{\mailsb}\path|anna.kramer, leonie.kunz, christine.reiss, nicole.sator,|
-\urldef{\mailsc}\path|erika.siebert-cole, peter.strasser, lncs}@springer.com|    
 \newcommand{\keywords}[1]{\par\addvspace\baselineskip
@@
 
+\newcommand{\ie}{\emph{i.e.}}
+\newcommand{\eg}{\emph{e.g.}}
+\newcommand{\cf}{\emph{cf.~}}
+\newcommand{\etal}{\emph{et al.~}}
+\newcommand{\etc}{\emph{etc.}}
+\newcommand{\aka}{\emph{a.k.a.}}
+
 \begin{document}
 
-\mainmatter  % start of an individual contribution
+\mainmatter
 
-% first the title is needed
-\title{Lecture Notes in Computer Science:\\Authors' Instructions
-for the Preparation\\of Camera-Ready
-Contributions\\to LNCS/LNAI/LNBI Proceedings}
+\title{FFI Is Not Enough. Need Dependent Types}
+\titlerunning{Dependent Types Rock}
 
-% a short form should be given in case it is too long for the running head
-\titlerunning{Lecture Notes in Computer Science: Authors' Instructions}
+\author{Julien Richard-Foy \and Olivier Barais\and Jean-Marc Jézéquel}
+\authorrunning{Julien Richard-Foy \emph{et. al.}}
 
-% the name(s) of the author(s) follow(s) next
-%
-% NB: Chinese authors should write their first names(s) in front of
-% their surnames. This ensures that the names appear correctly in
-% the running heads and the author index.
-%
-\author{Alfred Hofmann%
-\thanks{Please note that the LNCS Editorial assumes that all authors have used
-the western naming convention, with given names preceding surnames. This determines
-the structure of the names in the running heads and the author index.}%
-\and Ursula Barth\and Ingrid Haas\and Frank Holzwarth\and\\
-Anna Kramer\and Leonie Kunz\and Christine Rei\ss\and\\
-Nicole Sator\and Erika Siebert-Cole\and Peter Stra\ss er}
-%
-\authorrunning{Lecture Notes in Computer Science: Authors' Instructions}
-% (feature abused for this document to repeat the title also on left hand pages)
+\institute{IRISA, Université de Rennes 1, France. \texttt{\{first\}.\{last\}@irisa.fr}}
 
-% the affiliations are given next; don't give your e-mail address
-% unless you accept that it will be published
-\institute{Springer-Verlag, Computer Science Editorial,\\
-Tiergartenstr. 17, 69121 Heidelberg, Germany\\
-\mailsa\\
-\mailsb\\
-\mailsc\\
-\url{http://www.springer.com/lncs}}
-
-%
-% NB: a more complex sample for affiliations and the mapping to the
-% corresponding authors can be found in the file "llncs.dem"
-% (search for the string "\mainmatter" where a contribution starts).
-% "llncs.dem" accompanies the document class "llncs.cls".
-%
-
-\toctitle{Lecture Notes in Computer Science}
-\tocauthor{Authors' Instructions}
+%\toctitle{Lecture Notes in Computer Science}
+%\tocauthor{Authors' Instructions}
 \maketitle
@@
 
-You are strongly encouraged to use \LaTeXe{} for the
-preparation of your camera-ready manuscript together with the
-corresponding Springer class file \verb+llncs.cls+. Only if you use
-\LaTeXe{} can hyperlinks be generated in the online version
-of your manuscript.
+Web applications are attractive because they require no installation or deployment step on clients and enable large
+scale collaborative experiences. However, writing large Web applications is known to be
+difficult~\cite{Mikkonen08_SpaghettiJs,Preciado05_RIAMethodologyNecessity}. One challenge comes from the fact that
+the JavaScript programming language -- which is currently the only action language natively supported by almost all
+Web clients -- lacks of constructs making large code bases maintainable (\eg static typing, first-class modules).
 
-The \LaTeX{} source of this instruction file for \LaTeX{} users may be
-used as a template. This is
-located in the ``authors'' subdirectory in
-\url{ftp://ftp.springer.de/pub/tex/latex/llncs/latex2e/instruct/} and
-entitled \texttt{typeinst.tex}. There is a separate package for Word 
-users. Kindly send the final and checked source
-and PDF files of your paper to the Contact Volume Editor. This is
-usually one of the organizers of the conference. You should make sure
-that the \LaTeX{} and the PDF files are identical and correct and that
-only one version of your paper is sent. It is not possible to update
-files at a later stage. Please note that we do not need the printed
-paper.
+One solution consists in considering JavaScript as an assembly language and generating JavaScript from compilers
+of full-featured and cutting-edge programming languages. Incidentally, an increasing number of programming languages
+or compiler backends can generate JavaScript code (\eg Java/GWT~\cite{Chaganti07_GWT},
+SharpKit\footnote{\href{http://sharpkit.net}{http://sharpkit.net}}, Dart~\cite{Griffith11_Dart},
+Kotlin\footnote{\href{http://kotlin.jetbrains.org/}{http://kotlin.jetbrains.org/}},
+ClojureScript~\cite{McGranaghan11_ClojureScript}, Fay\footnote{\href{http://fay-lang.org/}{http://fay-lang.org/}},
+Haxe~\cite{Cannasse08_HaXe}, Opa\footnote{\href{http://opalang.org/}{http://opalang.org/}}).
 
-We would like to draw your attention to the fact that it is not possible
-to modify a paper in any way, once it has been published. This applies
-to both the printed book and the online version of the publication.
-Every detail, including the order of the names of the authors, should
-be checked before the paper is sent to the Volume Editors.
+However, compiling to JavaScript is not enough. Developers also need the Web browser programming environment: they
+need to interact with the Web page, to build DOM fragments, to listen to user events, \etc. A Foreign Function
+Interface mechanism could be used to make browser’s APIs available to the developers. However, JavaScript APIs are
+not statically typed and make a heavy use of overloading, making them hard to expose in a statically typed language.
+Indeed, most of the existing statically typed languages compiling to JavaScript either lose control or type safety
+when they expose Web browser’s APIs. How to give developers the same level of control as if they were using the
+native Web APIs, but in a statically typed and convenient way?
 
-\subsection{Checking the PDF File}
+In this paper we present several ways to integrate Web browser’s APIs as statically typed APIs that are safe and
+give developers the same control level as if they were using the native APIs. We could achieve this
+by using advanced features of type systems like dependent types and functional dependencies.
 
-Kindly assure that the Contact Volume Editor is given the name and email
-address of the contact author for your paper. The Contact Volume Editor
-uses these details to compile a list for our production department at
-SPS in India. Once the files have been worked upon, SPS sends a copy of
-the final pdf of each paper to its contact author. The contact author is
-asked to check through the final pdf to make sure that no errors have
-crept in during the transfer or preparation of the files. This should
-not be seen as an opportunity to update or copyedit the papers, which is
-not possible due to time constraints. Only errors introduced during the
-preparation of the files will be corrected.
+\section{Motivating Example}
 
-This round of checking takes place about two weeks after the files have
-been sent to the Editorial by the Contact Volume Editor, i.e., roughly
-seven weeks before the start of the conference for conference
-proceedings, or seven weeks before the volume leaves the printer's, for
-post-proceedings. If SPS does not receive a reply from a particular
-contact author, within the timeframe given, then it is presumed that the
-author has found no errors in the paper. The tight publication schedule
-of LNCS does not allow SPS to send reminders or search for alternative
-email addresses on the Internet.
+Typical tasks involved in Web applications.
 
-In some cases, it is the Contact Volume Editor that checks all the final
-pdfs. In such cases, the authors are not involved in the checking phase.
+Why is it difficult to type Web browser’s APIs?
 
-\subsection{Additional Information Required by the Volume Editor}
+\section{Lightweight Modular Staging}
 
-If you have more than one surname, please make sure that the Volume Editor
-knows how you are to be listed in the author index.
+\section{Contribution}
 
-\subsection{Copyright Forms}
+\subsection{Events}
 
-The copyright form may be downloaded from the ``For Authors"
-(Information for LNCS Authors) section of the LNCS Website:
-\texttt{www.springer.com/lncs}. Please send your signed copyright form
-to the Contact Volume Editor, either as a scanned pdf or by fax or by
-courier. One author may sign on behalf of all of the other authors of a
-particular paper. Digital signatures are acceptable.
+Path-dependent types to abstract over an event name and its data type.
 
-\section{Paper Preparation}
+\subsection{Selectors}
 
-Springer provides you with a complete integrated \LaTeX{} document class
-(\texttt{llncs.cls}) for multi-author books such as those in the LNCS
-series. Papers not complying with the LNCS style will be reformatted.
-This can lead to an increase in the overall number of pages. We would
-therefore urge you not to squash your paper.
+\begin{itemize}
+ \item Less type annotations on DOM queries, less chance to write nonsense casts
+ \item Inference-driving macros help inferring more specific types
+\end{itemize}
 
-Please always cancel any superfluous definitions that are
-not actually used in your text. If you do not, these may conflict with
-the definitions of the macro package, causing changes in the structure
-of the text and leading to numerous mistakes in the proofs.
+\subsection{DOM}
 
-If you wonder what \LaTeX{} is and where it can be obtained, see the
-``\textit{LaTeX project site}'' (\url{http://www.latex-project.org})
-and especially the webpage ``\textit{How to get it}''
-(\url{http://www.latex-project.org/ftp.html}) respectively.
+?
 
-When you use \LaTeX\ together with our document class file,
-\texttt{llncs.cls},
-your text is typeset automatically in Computer Modern Roman (CM) fonts.
-Please do
-\emph{not} change the preset fonts. If you have to use fonts other
-than the preset fonts, kindly submit these with your files.
+\section{Evaluation}
 
-Please use the commands \verb+\label+ and \verb+\ref+ for
-cross-references and the commands \verb+\bibitem+ and \verb+\cite+ for
-references to the bibliography, to enable us to create hyperlinks at
-these places.
+\subsection{Events}
 
-For preparing your figures electronically and integrating them into
-your source file we recommend using the standard \LaTeX{} \verb+graphics+ or
-\verb+graphicx+ package. These provide the \verb+\includegraphics+ command.
-In general, please refrain from using the \verb+\special+ command.
+Other languages either provide lose information about the data type of the listened event (Dart) or give no way to
+abstract over an event (GWT, Kotlin).
 
-Remember to submit any further style files and
-fonts you have used together with your source files.
-
-\subsubsection{Headings.}
-
-Headings should be capitalized
-(i.e., nouns, verbs, and all other words
-except articles, prepositions, and conjunctions should be set with an
-initial capital) and should,
-with the exception of the title, be aligned to the left.
-Words joined by a hyphen are subject to a special rule. If the first
-word can stand alone, the second word should be capitalized.
-
-Here are some examples of headings: ``Criteria to Disprove
-Context-Freeness of Collage Language", ``On Correcting the Intrusion of
-Tracing Non-deterministic Programs by Software", ``A User-Friendly and
-Extendable Data Distribution System", ``Multi-flip Networks:
-Parallelizing GenSAT", ``Self-determinations of Man".
-
-\subsubsection{Lemmas, Propositions, and Theorems.}
-
-The numbers accorded to lemmas, propositions, and theorems, etc. should
-appear in consecutive order, starting with Lemma 1, and not, for
-example, with Lemma 11.
-
-\subsection{Figures}
-
-For \LaTeX\ users, we recommend using the \emph{graphics} or \emph{graphicx}
-package and the \verb+\includegraphics+ command.
-
-Please check that the lines in line drawings are not
-interrupted and are of a constant width. Grids and details within the
-figures must be clearly legible and may not be written one on top of
-the other. Line drawings should have a resolution of at least 800 dpi
-(preferably 1200 dpi). The lettering in figures should have a height of
-2~mm (10-point type). Figures should be numbered and should have a
-caption which should always be positioned \emph{under} the figures, in
-contrast to the caption belonging to a table, which should always appear
-\emph{above} the table; this is simply achieved as matter of sequence in
-your source.
-
-Please center the figures or your tabular material by using the \verb+\centering+
-declaration. Short captions are centered by default between the margins
-and typeset in 9-point type (Fig.~\ref{fig:example} shows an example).
-The distance between text and figure is preset to be about 8~mm, the
-distance between figure and caption about 6~mm.
-
-To ensure that the reproduction of your illustrations is of a reasonable
-quality, we advise against the use of shading. The contrast should be as
-pronounced as possible.
-
-If screenshots are necessary, please make sure that you are happy with
-the print quality before you send the files.
-\begin{figure}
-\centering
-\includegraphics[height=6.2cm]{eijkel2}
-\caption{One kernel at $x_s$ (\emph{dotted kernel}) or two kernels at
-$x_i$ and $x_j$ (\textit{left and right}) lead to the same summed estimate
-at $x_s$. This shows a figure consisting of different types of
-lines. Elements of the figure described in the caption should be set in
-italics, in parentheses, as shown in this sample caption.}
-\label{fig:example}
-\end{figure}
-
-Please define figures (and tables) as floating objects. Please avoid
-using optional location parameters like ``\verb+[h]+" for ``here".
-
-\paragraph{Remark 1.}
-
-In the printed volumes, illustrations are generally black and white
-(halftones), and only in exceptional cases, and if the author is
-prepared to cover the extra cost for color reproduction, are colored
-pictures accepted. Colored pictures are welcome in the electronic
-version free of charge. If you send colored figures that are to be
-printed in black and white, please make sure that they really are
-legible in black and white. Some colors as well as the contrast of
-converted colors show up very poorly when printed in black and white.
-
-\subsection{Formulas}
-
-Displayed equations or formulas are centered and set on a separate
-line (with an extra line or halfline space above and below). Displayed
-expressions should be numbered for reference. The numbers should be
-consecutive within each section or within the contribution,
-with numbers enclosed in parentheses and set on the right margin --
-which is the default if you use the \emph{equation} environment, e.g.,
-\begin{equation}
-  \psi (u) = \int_{o}^{T} \left[\frac{1}{2}
-  \left(\Lambda_{o}^{-1} u,u\right) + N^{\ast} (-u)\right] dt \;  .
-\end{equation}
-
-Equations should be punctuated in the same way as ordinary
-text but with a small space before the end punctuation mark.
-
-\subsection{Footnotes}
-
-The superscript numeral used to refer to a footnote appears in the text
-either directly after the word to be discussed or -- in relation to a
-phrase or a sentence -- following the punctuation sign (comma,
-semicolon, or period). Footnotes should appear at the bottom of
-the
-normal text area, with a line of about 2~cm set
-immediately above them.\footnote{The footnote numeral is set flush left
-and the text follows with the usual word spacing.}
-
-\subsection{Program Code}
-
-Program listings or program commands in the text are normally set in
-typewriter font, e.g., CMTT10 or Courier.
-
-\medskip
-
-\noindent
-{\it Example of a Computer Program}
-\begin{verbatim}
-program Inflation (Output)
-  {Assuming annual inflation rates of 7%, 8%, and 10%,...
-   years};
-   const
-     MaxYears = 10;
-   var
-     Year: 0..MaxYears;
-     Factor1, Factor2, Factor3: Real;
-   begin
-     Year := 0;
-     Factor1 := 1.0; Factor2 := 1.0; Factor3 := 1.0;
-     WriteLn('Year  7% 8% 10%'); WriteLn;
-     repeat
-       Year := Year + 1;
-       Factor1 := Factor1 * 1.07;
-       Factor2 := Factor2 * 1.08;
-       Factor3 := Factor3 * 1.10;
-       WriteLn(Year:5,Factor1:7:3,Factor2:7:3,Factor3:7:3)
-     until Year = MaxYears
-end.
-\end{verbatim}
-%
-\noindent
-{\small (Example from Jensen K., Wirth N. (1991) Pascal user manual and
-report. Springer, New York)}
-
-\subsection{Citations}
-
-For citations in the text please use
-square brackets and consecutive numbers: \cite{jour}, \cite{lncschap},
-\cite{proceeding1} -- provided automatically
-by \LaTeX 's \verb|\cite| \dots\verb|\bibitem| mechanism.
-
-\subsection{Page Numbering and Running Heads}
-
-There is no need to include page numbers. If your paper title is too
-long to serve as a running head, it will be shortened. Your suggestion
-as to how to shorten it would be most welcome.
-
-\section{LNCS Online}
-
-The online version of the volume will be available in LNCS Online.
-Members of institutes subscribing to the Lecture Notes in Computer
-Science series have access to all the pdfs of all the online
-publications. Non-subscribers can only read as far as the abstracts. If
-they try to go beyond this point, they are automatically asked, whether
-they would like to order the pdf, and are given instructions as to how
-to do so.
-
-Please note that, if your email address is given in your paper,
-it will also be included in the meta data of the online version.
-
-\section{BibTeX Entries}
-
-The correct BibTeX entries for the Lecture Notes in Computer Science
-volumes can be found at the following Website shortly after the
-publication of the book:
-\url{http://www.informatik.uni-trier.de/~ley/db/journals/lncs.html}
-
-\subsubsection*{Acknowledgments.} The heading should be treated as a
-subsubsection heading and should not be assigned a number.
-
-\section{The References Section}\label{references}
-
-In order to permit cross referencing within LNCS-Online, and eventually
-between different publishers and their online databases, LNCS will,
-from now on, be standardizing the format of the references. This new
-feature will increase the visibility of publications and facilitate
-academic research considerably. Please base your references on the
-examples below. References that don't adhere to this style will be
-reformatted by Springer. You should therefore check your references
-thoroughly when you receive the final pdf of your paper.
-The reference section must be complete. You may not omit references.
-Instructions as to where to find a fuller version of the references are
-not permissible.
-
-We only accept references written using the latin alphabet. If the title
-of the book you are referring to is in Russian or Chinese, then please write
-(in Russian) or (in Chinese) at the end of the transcript or translation
-of the title.
-
-The following section shows a sample reference list with entries for
-journal articles \cite{jour}, an LNCS chapter \cite{lncschap}, a book
-\cite{book}, proceedings without editors \cite{proceeding1} and
-\cite{proceeding2}, as well as a URL \cite{url}.
-Please note that proceedings published in LNCS are not cited with their
-full titles, but with their acronyms!
+\section{Conclusion and Perspectives}
 
@@
 
-\section*{Appendix: Springer-Author Discount}
-
-LNCS authors are entitled to a 33.3\% discount off all Springer
-publications. Before placing an order, the author should send an email, 
-giving full details of his or her Springer publication,
-to \url{orders-HD-individuals@springer.com} to obtain a so-called token. This token is a
-number, which must be entered when placing an order via the Internet, in
-order to obtain the discount.
-
-\section{Checklist of Items to be Sent to Volume Editors}
-Here is a checklist of everything the volume editor requires from you:
-
-\begin{itemize}
-\settowidth{\leftmargin}{{\Large$\square$}}\advance\leftmargin\labelsep
-\itemsep8pt\relax
-\renewcommand\labelitemi{{\lower1.5pt\hbox{\Large$\square$}}}
-
-\item The final \LaTeX{} source files
-\item A final PDF file
-\item A copyright form, signed by one author on behalf of all of the
-authors of the paper.
-\item A readme giving the name and email address of the
-corresponding author.
-\end{itemize}
 \end{document}

### 2013-05-16 src/main/scala/scala/js/JSMaps.scala
small fixes and improvments

@@ -9,14 +9,37 @@ import scala.virtualization.lms.common._
  */
 trait JSMaps extends Base {
 
-  object Map {
-    def apply[K:Manifest,V:Manifest]() = map_new[K,V]()
+  object JSMap {
+    def apply[K:JSKey,V:Manifest]() = map_new[K,V]()
   }
 
-  implicit def repMapToMapOps[K:Manifest,V:Manifest](m: Rep[Map[K,V]]) = new mapOpsCls(m)
+  /* only some types are authorized as keys to avoid problems in javascript
+   * authorized keys are of type:
+   *  - String
+   *  - Char
+   *  - numeric types (Int, Long, Float, Double, Byte, ...)
+   *  - Boolean
+   */
+  abstract class JSKey[A:Manifest] extends Manifest[A] {
+    def runtimeClass = implicitly[Manifest[A]].runtimeClass
+  }
+  object JSKey {
+    implicit val stringKey = new JSKey[String] {}
+    implicit val charKey = new JSKey[Char] {}
+    implicit val intKey = new JSKey[Int] {}
+    implicit val longKey = new JSKey[Long] {}
+    implicit val floatKey = new JSKey[Float] {}
+    implicit val doubleKey = new JSKey[Double] {}
+    implicit val boolKey = new JSKey[Boolean] {}
+  }
 
-  class mapOpsCls[K:Manifest,V:Manifest](m: Rep[Map[K,V]]) {
+  implicit def repMapToMapOps[K:JSKey,V:Manifest](m: Rep[Map[K,V]]) = new mapOpsCls(m)
+
+  class mapOpsCls[K:JSKey,V:Manifest](m: Rep[Map[K,V]]) {
+    val keyType = implicitly[Manifest[K]]
+    val valueType = implicitly[Manifest[V]]
     def apply(k: Rep[K]) = map_apply(m, k)
+    def get(k: Rep[K]) = map_get(m, k)
     def update(k: Rep[K], v: Rep[V]) = map_update(m,k,v)
     def contains(k: Rep[K]) = map_contains(m, k)
     def size = map_size(m)
@@ -28,52 +51,59 @@ trait JSMaps extends Base {
[…]

### 2013-05-16 src/test/scala/scala/js/TestMaps.scala

@@ -7,7 +7,7 @@ import java.io.FileOutputStream
 
 trait MapsProg { this: JS with JSMaps =>
   def test1(): Rep[Map[String, Int]] = {
-    Map[String, Int]()
+    JSMap[String, Int]()
   }
 
   def test2(map: Rep[Map[String, Int]]): Rep[Int] = {
@@ -16,6 +16,14 @@ trait MapsProg { this: JS with JSMaps =>
 
   def test3(map: Rep[Map[Int, String]]): Rep[String] = {
     map(3)
+  }
+
+  def test2b(map: Rep[Map[String, Int]]): Rep[Option[Int]] = {
+    map.get("test")
+  }
+
+  def test3b(map: Rep[Map[Int, String]]): Rep[Option[String]] = {
+    map.get(3)
   }
 
   def test4(map: Rep[Map[String, Int]]): Rep[Unit] = {
@@ -94,6 +102,18 @@ class TestMaps extends FileDiffSuite {
 
     }
     assertFileEqualsCheck(prefix+"access")
+  }
+
+  def testFieldAccessOpt = {
+    withOutFile(prefix+"access-opt") {
+      new MapsProg with JSExp with JSMapsExp { self =>
+        val codegen = new JSGen with JSGenMaps { val IR: self.type = self }
+        codegen.emitSource(test2 _, "test2b", new PrintWriter(System.out))
+        codegen.emitSource(test3 _, "test3b", new PrintWriter(System.out))
+      }
+
+    }
+    assertFileEqualsCheck(prefix+"access-opt")
   }
 
   def testFieldUpdate = {

### 2013-05-17 papers/gpce2013/gpce2013.tex
Figures
we

@@
 \usepackage[unicode=true,hidelinks]{hyperref}
+\usepackage{graphicx}
 
@@
 client-side, changing your *mind* means completely rewriting the feature on server-side (and \emph{vice versa}).
-Even worse, logic parts that need to run on both client-side and server-side are duplicated. For instance, HTML
-fragments may be built from the server-side when a page is requested by a client, but they may also be built from the
-client-side to perform an incremental update subsequent to an user action. How could developers write HTML fragment
-definitions once and render them on both client-side and server-side? The more interactive the application is, the
-more logic needs to be duplicated between the server-side and the client-side (explain why?).
+Even worse, logic parts that need to run on both client-side and server-side have to be duplicated. For instance,
+HTML fragments may be built from the server-side when a page is requested by a client, but they may also be built
+from the client-side to perform an incremental update subsequent to an user action. How could developers write HTML
+fragment definitions once and render them on both client-side and server-side? The more interactive the application
+is, the more logic needs to be duplicated between the server-side and the client-side (explain why?).
 
@@
 
+Developers express programs using a language that is translated into a form executable on a target platform. This
+section presents different approaches for defining cross-platform languages.
+
 \subsection{Fat Languages}
+
+One way consists in hard-coding, in the language’s compiler, the transformation to each target platform. Figure
+\ref{fat-lang} depicts this process. In order to support a feature related to a specific domain, the whole compiler
+pipeline (parser, code generator, \etc) may have to be adapted. This approach gives \emph{fat} languages because
+a lot of concepts are defined at the language level. Examples of such languages are Links~\cite{Cooper07_Links},
+Opa and Dart~\cite{Griffith11_Dart}. These languages are difficult to extend because each concept is defined in the
+compiler, and modifying a compiler requires a high effort. Furthermore, these languages also require to support
+common abstraction and composition mechanisms, as general purpose languages do, in order to make it easier to write
+real programs. So they need to be at least general purpose languages. It means that for every problem you have to
+rewrite a full-featured programming language before addressing the concepts specific to the problem domain. This
+approach does not scale.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=3cm]{fat.pdf}
+  \caption{Fat language}
+  \label{fat-lang}
+\end{figure}
+
+\subsection{Domain Specific Languages}
+
+Problem: interoperability.
 
@@
 
+Alternatively, one can define concepts relative to a specific domain as libraries on top of a thin general purpose
+language. Figure \ref{thin-lang} depicts this approach. The general purpose language is used as a host language that
+does not need to be modified if a new concept is introduced, because concepts are defined as libraries. However,
+this approach gives no opportunity to translate a concept efficiently, according to the host platform
+characteristics.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=3cm]{thin.pdf}
+  \caption{Thin language}
+  \label{thin-lang}
+\end{figure}
+
 \subsection{Deeply Embedded Languages}
+
+Figure \ref{dedsl} shows a last approach that tries to get the benifits of the two previous approaches, without
+their shortcomings.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=3cm]{lms.pdf}
+  \caption{Deeply embedded languages}
+  \label{dedsl}
+\end{figure}
 
@@
 
-\subsection{Ad-Hoc Polymorphism}
-
-Because of the dynamically typed nature of JavaScript, when calling a function there is no proper way to select a
-specialized implementation according to the function’s parameters types. JavaScript is only able to dispatch
-according to a method receiver prototype, \eg{} if one writes \code{foo.bar()} the JavaScript runtime will look into
-the prototype of the \code{foo} object for a property named \code{bar} and will call it. So, the only way to achieve
-\emph{ad hoc} polymorphism on JavaScript objects consists in defining the polymorphic function on the prototypes of
-the objects. However, modifying existing object prototypes is considered bad
-practice~\cite{Zakas12_MaintainableJs}. Another way could consist in manually coding the dispatch logic, by
-registering supported data types at the beginning of the program execution, as described in section 2.4.3
-of~\cite{Abelson83_SICP}, but this solution is painful for developers and incurs a performance overhead.
-
-We propose to achieve \emph{ad hoc} polymorphism using
-typeclasses~\cite{Wadler89_AdhocPolymorphism,Odersky06_Typeclasses,Oliveira10_Typeclasses} so that it supports
-retroactive extension without modifying objects prototypes because it is type-directed: the dispatch happens at
-compile-time rather than at runtime.
-
-\begin{figure}
-\begin{lstlisting}[label=polymorphism,caption=Ad hoc polymorphism using typeclasses]
-// Interface
-case class Show[A](show: Rep[A => Node])
-
-// Polymorphic function
-def listWidget[A : Show](items: Rep[List[A]]): Rep[Node] = {
-  tag("ul")(
-    for (item <- items) yield {
-      tag("li")(implicitly[Show[A]].show(item))
-    }
-  )
-}
-
-// Type `User`
-type User = Record { val name: String; val age: Int }
-// Implementation of Show for a User
-implicit val showUser = Show[User] { user =>
-  tag("span", "class"->"user")(
-    text(user.name + "(" + user.age + " years)")
-  )
-}
-
-// Type `Article`
-type Article = Record { val name: String; val price: Double }
-// Implementation of Show for an Article
-implicit val showArticle = Show[Article] { article =>
-  tag("span")(
-    text(article.name),
-    tag("strong")(text(article.price + " Euros"))
-  )
-}
-
-// Main program
-def main(users: Rep[List[User]], articles: Rep[List[Article]]) = {
-  document.body.append(listWidget(users))
-  document.body.append(listWidget(articles))
-}
-\end{lstlisting}
-\end{figure}
-
-Listing \ref{polymorphism} demonstrates how to define a polymorphic \code{listWidget} function that returns a DOM
-tree containing the representation of a list of items. The \code{Show[A]} typeclass defines how to produce a DOM tree
-for a value of type \code{A}. It is used by the \code{listWidget} function to get the DOM fragments of the list
-items. The listing shows how to reuse the same \code{listWidget} function to show a list of users and a list of
-articles.
-
 \subsection{Monads Sequencing}
-
-The previous work on js-scala \cite{Kossakowski12_JsDESL} showed how to make asynchronous programming more
-convenient by making asynchronous calls looking like synchronous calls (\ie{} returning a value instead of taking a
-callback as parameter). This work was based on the Scala continuations compiler plugin. We claim that, although this
-programming model removes the “callback hell”, it can make the code hard to reason about because there is no explicit
-distinction between synchronous and asynchronous computations.
-
-We propose a DSL that explicitly reflects the asynchronous nature of computations in their types and provides methods
-turning them into first-class citizens. The DSL is monadic so we can solve the callback hell problem thanks to the
-Scala \code{for} notation.
-
-For instance, listing \ref{async-first-class} shows how the listing \ref{async-js} can be re-written using our DSL.
-The \code{createVertex} function now returns an asynchronous value instead of taking a callback as
-parameter. Then, the \code{for} expression allows us to get the vertex, when available, and to insert it on the user
-interface. By making the \code{createVertex} function return an asynchronous value instead of taking a callback as
-a parameter, the code is easier to modularize into loosely coupled parts.
-
-\begin{figure}
-\centering
-\begin{lstlisting}[caption=Asynchronous values are first class citizen,label=async-first-class]
-def createVertex(text: Rep[String]): Rep[Future[Vertex]] =
-  Ajax.post[Vertex]("/create", new Record { val content = text })
-
-for (vertex <- createVertex("Hello, World!")) {
-  addVertex(vertex)
-}
-\end{lstlisting}
-\end{figure}
-
-\begin{figure}
-\centering
-\begin{lstlisting}[caption=No callback hell,label=async-no-callback]
-for {
-  foo <- Ajax.get(fooUrl())
-  bar <- Ajax.get(barUrl(foo))
-  _ <- future(println("bar = " + bar))
-  baz <- Ajax.get(bazUrl(bar))
-} println(foo + bar + baz)
-\end{lstlisting}
-\end{figure}
-
-Listing \ref{async-no-callback} translates the callback hell example (listing \ref{async-callback-hell}) using
-our DSL. The \code{for} notation can intuitively be thought of as a sequencing notation: whenever the response of the
-first Ajax request is available, the next statement is executed, and so on. There is no nested callbacks and the
-order of execution is directly reflected by the order of statements.
-
-An asynchronous value of type \code{Rep[A]} is modelled by a value of type \code{Rep[Future[A]]}. Because Scala’s
-\code{for} notation is just syntactic sugar for methods \code{foreach}, \code{map} and \code{flatMap}, we are
-able to define our DSL by just defining these methods on \code{Rep[Future[A]]} values, with the following semantic:
-\begin{itemize}
-\item \code{foreach(f: Rep[A] => Rep[Unit]): Rep[Unit]}, eventually does something with the value when available~;
-\item \code{map(f: Rep[A] => Rep[B]): Rep[Future[B]]}, eventually transforms the value when available~;
-\item \code{flatMap(f: Rep[A] => Rep[Future[B]]): Rep[Future[B]]} also eventually transforms the value when
-available~;
-\end{itemize}
-
-Asynchronous values can be used and transformed using the \code{foreach}, \code{map} and \code{flatMap} methods,
-turning them into first-class
-citizens: functions can take them as parameters and return them. Consequently, it is easier to write concurrent code:
-listing \ref{async-parallel-1} shows how to run two parallel asynchronous computations and to do something with their
-results when both are available. Writing equivalent code in pure JavaScript is tedious because it requires that the
-callback passed to each asynchronous computation checks if the other has been completed before, and to store the
-value, when availalbe, in a shared place (so both callbacks can get the other’s value).
-
-\begin{figure}
-\begin{lstlisting}[caption=Parallel computations in Scala,label=async-parallel-1]
-val fooAsync = Ajax.get("/foo")
-val barAsync = Ajax.get("/bar")
-for {
-  foo <- fooAsync
-  bar <- barAsync
-} println(foo + bar)
-\end{lstlisting}
-\end{figure}
 
@@
 Null references are a known source of problems in programming languages~\cite{Hoare09_Null,Nanda09_Null}. For
-example, consider the following typical JavaScript code finding a particular widget in the page and a then particular
+example, consider the following typical JavaScript code finding a particular widget in the page and then a particular
 button in the widget:
@@
 \begin{lstlisting}[language=JavaScript,label=null-unsafe,caption=Unsafe code]
-var loginWidget = document.querySelector("div.login");
-var loginButton = loginWidget.querySelector("button.submit");
-loginButton.addEventListener("click", function (e) { ... });
+var loginWidget =
+  document.querySelector("div.login");
+var loginButton =
+  loginWidget.querySelector("button.submit");
+loginButton.addEventListener("click", handler);
 \end{lstlisting}
@@
 \begin{lstlisting}[language=JavaScript,label=null-defensive,caption=Defensive programming to handle null references]
-var loginWidget = document.querySelector("div.login");
+var loginWidget =
+  document.querySelector("div.login");
 if (loginWidget !== null) {
-  var loginButton = loginWidget.querySelector("button.submit");
+  var loginButton =
+    loginWidget.querySelector("button.submit");
   if (loginButton !== null) {
-    loginButton.addEventListener("click", function (e) { ... });
+    loginButton.
+      addEventListener("click", handler);
   }
@@
 
-So the only remaining work consists in defining the DSL vocabulary to define DOM nodes. We provide a \code{tag}
-function to define a tag and a \code{text} function to define a text node.
+So the only remaining work consists in defining the DSL vocabulary to define DOM nodes. We provide a \code{el}
+function to define a tag and any \code{String} value is considered to be a text node.
 
@@
 def vertexDom(v: Rep[Vertex]) =
-    tag("g", "class"->"vertex",
+    el("g", "class"->"vertex",
             "transform"->("translate("+v.x+","+v.y+")"))(
-        tag("rect", "width"->v.width, "height"->v.height)(),
-        tag("text", "width"->v.width, "height"->v.height)(
-            text(v.content)
+        el("rect", "width"->v.width, "height"->v.height)(),
+        el("text", "width"->v.width, "height"->v.height)(
+            v.content
         )
@@
 
+\subsection{Ad-Hoc Polymorphism}
+
+Because of the dynamically typed nature of JavaScript, when calling a function there is no proper way to select a
+specialized implementation according to the function’s parameters types. JavaScript is only able to dispatch
+according to a method receiver prototype, \eg{} if one writes \code{foo.bar()} the JavaScript runtime will look into
+the prototype of the \code{foo} object for a property named \code{bar} and will call it. So, the only way to achieve
+\emph{ad hoc} polymorphism on JavaScript objects consists in defining the polymorphic function on the prototypes of
+the objects. However, modifying existing object prototypes is considered bad
+practice~\cite{Zakas12_MaintainableJs}. Another way could consist in manually coding the dispatch logic, by
+registering supported data types at the beginning of the program execution, as described in section 2.4.3
+of~\cite{Abelson83_SICP}, but this solution is painful for developers and incurs a performance overhead.
+
+We propose to achieve \emph{ad hoc} polymorphism using
+typeclasses~\cite{Wadler89_AdhocPolymorphism,Odersky06_Typeclasses,Oliveira10_Typeclasses} so that it supports
+retroactive extension without modifying objects prototypes. The dispatch logic is type-directed and performed by the
+compiler, so there is no runtime overhead.
+
+\begin{figure}
+\begin{lstlisting}[label=polymorphism,caption=Ad hoc polymorphism using typeclasses]
+// Interface
+case class Show[A](show: Rep[A => Node])
+
+// Polymorphic function
+def listWidget[A](items: Rep[List[A]])
+      (implicit A: Show[A]): Rep[Node] =
+  el("ul")(
+    for (item <- items) yield {
+      el("li")(A.show(item))
+    }
+  )
+
+// Type `User`
+type User = Record {
+  val name: String
+  val age: Int
+}
+// Implementation of Show for a User
+implicit val showUser = Show[User] { user =>
+  el("span", "class"->"user")(
+    user.name + "(" + user.age + " years)"
+  )
+}
+
+// Main program
+def main(users: Rep[List[User]]) = {
+  document.body.append(listWidget(users))
+}
+\end{lstlisting}
+\end{figure}
+
+Listing \ref{polymorphism} demonstrates how to define a polymorphic \code{listWidget} function that returns a DOM
+tree containing the representation of a list of items. The \code{Show[A]} typeclass defines how to produce a DOM tree
+for a value of type \code{A}. It is used by the \code{listWidget} function to get the DOM fragments of the list
+items. The listing shows how to reuse the same \code{listWidget} function to show a list of users and a list of
+articles.
+
 \subsection{Selectors}
@@
 
+See http://github.com/js-scala/js-scala.
+
 \section{Evaluation}
 \label{validation}
+
+We implemented several applications using js-scala. We also have written several implementations of a complete
+application, using different approaches to write the client and server sides, and compared the amount of code
+written, the runtime performances and the ability to modularize the code and to maintain it.
 

### 2013-05-17 src/main/scala/scala/js/JSMaps.scala
small fixes
allow for nicer use in for-comprehensions for `foreach` and `filter`
operators. Some test added on filter followed by foreach

@@ -33,11 +33,7 @@ trait JSMaps extends Base {
     implicit val boolKey = new JSKey[Boolean] {}
   }
 
-  implicit def repMapToMapOps[K:JSKey,V:Manifest](m: Rep[Map[K,V]]) = new mapOpsCls(m)
-
-  class mapOpsCls[K:JSKey,V:Manifest](m: Rep[Map[K,V]]) {
-    val keyType = implicitly[Manifest[K]]
-    val valueType = implicitly[Manifest[V]]
+  implicit class mapOpsCls[K:JSKey,V:Manifest](m: Rep[Map[K,V]]) {
     def apply(k: Rep[K]) = map_apply(m, k)
     def get(k: Rep[K]) = map_get(m, k)
     def update(k: Rep[K], v: Rep[V]) = map_update(m,k,v)
@@ -46,9 +42,9 @@ trait JSMaps extends Base {
     def values = map_values(m)
     def clear() = map_clear(m)
     def keys = map_keys(m)
-    def foreach(block: Rep[(K,V)] => Rep[Unit]) = map_foreach(m, block)
+    def foreach(block: ((Rep[K], Rep[V])) => Rep[Unit]) = map_foreach(m, block)
     def mapValues[U:Manifest](block: Rep[V] => Rep[U]) = map_mapValues(m, block)
-    def filter(block: Rep[(K,V)] => Rep[Boolean]) = map_filter(m, block)
+    def filter(block: ((Rep[K], Rep[V])) => Rep[Boolean]) = map_filter(m, block)
   }
 
   def map_new[K:JSKey,V:Manifest]() : Rep[Map[K,V]]
@@ -60,9 +56,9 @@ trait JSMaps extends Base {
   def map_values[K:JSKey,V:Manifest](m: Rep[Map[K,V]]): Rep[Array[V]]
   def map_clear[K:JSKey,V:Manifest](m: Rep[Map[K,V]]): Rep[Unit]
   def map_keys[K:JSKey,V:Manifest](m: Rep[Map[K,V]]): Rep[Array[K]]
-  def map_foreach[K:JSKey,V:Manifest](m: Rep[Map[K,V]], block: Rep[(K,V)] => Rep[Unit]): Rep[Unit]
+  def map_foreach[K:JSKey,V:Manifest](m: Rep[Map[K,V]], block: ((Rep[K], Rep[V])) => Rep[Unit]): Rep[Unit]
[…]

### 2013-05-17 src/test/scala/scala/js/TestMaps.scala

@@ -59,8 +59,9 @@ trait MapsProg { this: JS with JSMaps =>
   }
 
   def test12(map: Rep[Map[String, Int]]): Rep[Unit] = {
-    map.foreach { pair =>
-      map(pair._1) = pair._2 + 1
+    map.foreach {
+      case (key, value) =>
+        map(key) = value + 1
     }
   }
 
@@ -71,10 +72,18 @@ trait MapsProg { this: JS with JSMaps =>
   }
 
   def test14(map: Rep[Map[String, Int]]): Rep[Map[String, Int]] = {
-    map.filter { pair =>
-      pair._2 > 10
+    map.filter {
+      case (_, value) =>
+        value > 10
     }
   }
+
+  def test15(map: Rep[Map[String, Int]]): Rep[Unit] = {
+    for((key, value) <- map) {
+      ()
+    }
+  }
+
 }
 
 class TestMaps extends FileDiffSuite {
@@ -215,5 +224,16 @@ class TestMaps extends FileDiffSuite {
 
     }
     assertFileEqualsCheck(prefix+"filter")
+  }
+
+  def testFilterForeach = {
+    withOutFile(prefix+"filter-foreach") {
+      new MapsProg with JSExp with JSMapsExp { self =>
+        val codegen = new JSGen with JSGenMaps { val IR: self.type = self }
+        codegen.emitSource(test15 _, "test15", new PrintWriter(System.out))
+      }
+
+    }
+    assertFileEqualsCheck(prefix+"filter-foreach")
   }
 }

### 2013-05-21 src/main/scala/scala/js/JSDom.scala
optimize selector api

@@ -172,15 +172,74 @@ trait JSDomExp extends JSDom with EffectExp with JSFunctionsExp with OptionOpsEx
   def infix_document(w: Exp[Window]) = WindowDocument
   def infix_history(w: Exp[Window]) = WindowHistory
 
-  // TODO generate a getElementById if possible
-  def selector_find[A : Selectable : Manifest](s: Exp[SelectorApi], selector: Exp[String]) =
-    reflectEffect(SelectorFind[A](s, selector))
-  // TODO generate a getElementsByClassName or getElementsByTagName if possible
-  def selector_findAll[A : Selectable : Manifest](s: Exp[SelectorApi], selector: Exp[String]) = {
-    val ns = reflectEffect(SelectorFindAll[A](s, selector))
-    val toArray: Exp[NodeList => List[A]] = NodeListToArray[A]()
-    toArray(ns)
+  def selector_find[A : Selectable : Manifest](s: Exp[SelectorApi], selector: Exp[String]) = {
+   
+    //the Regex to recover the ID
+    val Id = "#((-?[A-Za-z0-9_]+)+)".r
+    
+    //pattern matching on the selector
+    selector match {
+      //if it is a constant (which is normally always the case here)
+      case Const(selectorString) => 
+        //we check the selector's value
+        selectorString.trim match {
+          //if it's a '#' we are searching a ID
+          case Id(id, _) => 
+            //we call SelectorGetElementById (with reflect effect) for this ID
+            reflectEffect(SelectorGetElementById[A](s, unit(id)))
+          //in other cases
+          case _ =>
+            //we call by default SelectorFind for the selector
+            reflectEffect(SelectorFind[A](s, selector))           
+        }
+      //if it's a variable, we call SelectorFind with the selector
+      case _ => 
+        reflectEffect(SelectorFind[A](s, selector))
+    }  
   }
+    
+  def selector_findAll[A : Selectable : Manifest](s: Exp[SelectorApi], selector: Exp[String]) = {
+    
+    //the Regex to recover the class
+    val Classe = "\\.((-?[A-Za-z0-9_]+)+)".r
+    //the Regex to recover the tag
+    val Tag = "((-?[A-Za-z0-9_]+)+)".r
+    
+    //function for transform a nodeList to a list
+    val toArray: Exp[NodeList => List[A]] = NodeListToArray[A]() 
+    
+    //pattern matching on the selector
+    selector match {
[…]

### 2013-05-21 src/test/scala/scala/js/TestDom.scala

@@ -21,8 +21,18 @@ class TestDom extends FileDiffSuite {
       }
 
       def selectors() = {
-        val bar = document.find(unit(".bar"))
-        log(bar: Rep[Option[Element]])
+        val bar = document.findAll(unit("babar"))
+        log(bar: Rep[List[Element]])
+        val bar2 = document.findAll(unit(".babar"))
+        log(bar2: Rep[List[Element]])
+        val foo = document.find(unit(" #foo-1-0    "))
+        log(foo: Rep[Option[Element]])
+        val foo2 = document.find(unit(" #---    "))
+        log(foo2: Rep[Option[Element]])
+        val foo3 = document.find(unit(" #-a_a    "))
+        log(foo3: Rep[Option[Element]])
+        val foo4 = document.find(unit(" #bbaa    "))
+        log(foo4: Rep[Option[Element]])
         val input = document.find[Input](unit("input"))
         log(input: Rep[Option[Input]])
       }

### 2013-05-21 papers/gpce2013/gpce2013.tex
i++

@@
 
-\title{Title Text}
+\title{How to build an homogeneous language for heterogeneous platforms}
 \subtitle{You don’t have to trade abstraction for control}
@@
 environments~\cite{Echeverria09_RIA,Kuuskeri09_PartitioningClientServer}. This gives less flexibility in the
-engineering process and requires a higher maintainance effort: once you decided to implement a feature on
-client-side, changing your *mind* means completely rewriting the feature on server-side (and \emph{vice versa}).
-Even worse, logic parts that need to run on both client-side and server-side have to be duplicated. For instance,
-HTML fragments may be built from the server-side when a page is requested by a client, but they may also be built
-from the client-side to perform an incremental update subsequent to an user action. How could developers write HTML
-fragment definitions once and render them on both client-side and server-side? The more interactive the application
-is, the more logic needs to be duplicated between the server-side and the client-side (explain why?).
+engineering process and requires a higher maintenance effort: if a piece of logic was implemented on
+client-side and it turns out that it needs to be implemented on server-side instead, the code can not be reused and
+the feature needs to be completely rewritten (and \emph{vice versa}). Even worse, logic parts that run on both
+client-side and server-side need to be duplicated. For instance, HTML fragments may be built from the server-side
+when a page is requested by a client, but they may also be built from the client-side to perform an incremental
+update subsequent to an user action. How could developers write HTML fragment definitions once and render them on
+both client-side and server-side? The more interactive the application is, the more logic needs to be duplicated
+between the server-side and the client-side (explain why?).
 
@@
 action language natively supported by most of Web clients -- can be used on server-side, and an increasing
-number of programming languages or compiler backends can generate JavaScript code (\eg
+number of programming languages or compiler back-ends can generate JavaScript code (\eg
 Java/GWT~\cite{Chaganti07_GWT}, SharpKit\footnote{\href{http://sharpkit.net}{http://sharpkit.net}},
@@
 
-However, this engineering comfort may come at the price of an inefficient runtime: abstracting over platform
-differences often means restricting to a subset of common features and losing opportunities to perform platform
-specific optimizations. Performance is a primary concern in Web applications, because they are expected to run on a
-broad range of devices, from the powerful desktop personal computer to the less powerful smartphone. “Every 100~ms
-delay costs 1\% of sales”, said Amazon in 2006. For instance, because the boundaries of the code sent to the client
-are less visible when you share code between client-side and server-side, transitive dependencies may pull a lot of
-code on the client, causing a high download overhead. Moreover, generating efficient code for heterogeneous platforms
-is hard to achieve in an extensible way: the translation of common abstractions like collections into their native
-counterpart (JavaScript arrays on client-side and standard library’s collections on server-side) may be hardcoded in
-the compiler, but that would not scale to handle all the abstractions a complete application may use (\eg HTML
-fragment definitions, form validation rules, or even some business data type that may be represented differently for
-performance reasons).
+However, using the same programming language is not enough because the client and server programming environments
+are not the same. For instance, DOM fragments can be defined on client-side using the standard DOM API, but this
+API does not exist on server-side. How to define a common vocabulary for such concepts? And how to make it leverage
+the native APIs, when possible, for performance reasons?
+
+Performance is a primary concern in Web applications, because they are expected to run on a broad range of devices,
+from the powerful desktop personal computer to the less powerful smart phone. “Every 100~ms delay costs 1\% of
+sales”, said Amazon in 2006.
+
+For instance, because the boundaries of the code sent to the client are less visible when you share code between
+client-side and server-side, transitive dependencies may pull a lot of code on the client, causing a high download
+overhead. Moreover, generating efficient code for heterogeneous platforms is hard to achieve in an extensible way:
+the translation of common abstractions like collections into their native counterpart (JavaScript arrays on
+client-side and standard library's collections on server-side) may be hard coded in the compiler, but that would not
+scale to handle all the abstractions a complete application may use (\eg HTML fragment definitions, form validation
+rules, or even some business data type that may be represented differently for performance reasons).
 
@@
 
+(More on the results metrics)
+
 The remainder of this paper is organized as follows. The next section introduces existing approaches for defining
@@
 
-One way consists in hard-coding, in the language’s compiler, the transformation to each target platform. Figure
+One way consists in hard-coding, in the language's compiler, the transformation to each target platform. Figure
 \ref{fat-lang} depicts this process. In order to support a feature related to a specific domain, the whole compiler
@@
 
-Figure \ref{dedsl} shows a last approach that tries to get the benifits of the two previous approaches, without
+Figure \ref{dedsl} shows a last approach that tries to get the benefits of the previous two approaches, without
 their shortcomings.
@@
 returning a DOM fragment~;
-\item the type system typechecks template definitions and template calls~;
+\item the type system type-checks template definitions and template calls~;
 \item the Scala language itself is the expression language~;

### 2013-05-22 src/main/scala/scala/js/JSDom.scala
optimize selector api (best comments)

@@ -179,20 +179,17 @@ trait JSDomExp extends JSDom with EffectExp with JSFunctionsExp with OptionOpsEx
     
     //pattern matching on the selector
     selector match {
-      //if it is a constant (which is normally always the case here)
+      //if it is a constant
       case Const(selectorString) => 
-        //we check the selector's value
         selectorString.trim match {
-          //if it's a '#' we are searching a ID
+          //if the first charactere is a '#' we are searching a ID
           case Id(id, _) => 
-            //we call SelectorGetElementById (with reflect effect) for this ID
             reflectEffect(SelectorGetElementById[A](s, unit(id)))
           //in other cases
           case _ =>
-            //we call by default SelectorFind for the selector
             reflectEffect(SelectorFind[A](s, selector))           
         }
-      //if it's a variable, we call SelectorFind with the selector
+      //if it's a variable
       case _ => 
         reflectEffect(SelectorFind[A](s, selector))
     }  
@@ -209,35 +206,26 @@ trait JSDomExp extends JSDom with EffectExp with JSFunctionsExp with OptionOpsEx
     val toArray: Exp[NodeList => List[A]] = NodeListToArray[A]() 
     
     //pattern matching on the selector
-    selector match {
-      //if it is a constant (which is normally always the case here)
+    val result:Exp[NodeList] = selector match {
+      //if it is a constant 
       case Const(selectorString) => 
-        //we check the selector's value
         selectorString.trim match {
-          //if it's a '.' we are searching a class
+          //if the first charactere is a '.' we are searching a class
           case Classe(classe, _) => 
-            //we call SelectorGetElementsByClassName (with reflect effect) for this class name
-            val ns = reflectEffect(SelectorGetElementsByClassName[A](s, unit(classe)))
-            //we transform the result in a list
-            toArray(ns) 
+            reflectEffect(SelectorGetElementsByClassName[A](s, unit(classe)))            
           //if it's a character chain we are searching a tag
           case Tag(tag, _) => 
-            //we call SelectorGetElementsByTagName (with reflect effect) for this tag name
-            val ns = reflectEffect(SelectorGetElementsByTagName[A](s, unit(tag)))
-            //we transform the result in a list
-            toArray(ns) 
+            reflectEffect(SelectorGetElementsByTagName[A](s, unit(tag)))
           //in other cases
           case _ =>
-            //we call by default SelectorFindAll for the selector 
[…]

### 2013-05-22 src/test/scala/scala/js/TestDom.scala

@@ -21,18 +21,22 @@ class TestDom extends FileDiffSuite {
       }
 
       def selectors() = {
-        val bar = document.findAll(unit("babar"))
-        log(bar: Rep[List[Element]])
-        val bar2 = document.findAll(unit(".babar"))
-        log(bar2: Rep[List[Element]])
-        val foo = document.find(unit(" #foo-1-0    "))
-        log(foo: Rep[Option[Element]])
-        val foo2 = document.find(unit(" #---    "))
-        log(foo2: Rep[Option[Element]])
-        val foo3 = document.find(unit(" #-a_a    "))
-        log(foo3: Rep[Option[Element]])
-        val foo4 = document.find(unit(" #bbaa    "))
-        log(foo4: Rep[Option[Element]])
+        val tag = document.findAll(unit("babar"))
+        log(tag: Rep[List[Element]])
+        val classe = document.findAll(unit(".babar"))
+        log(classe: Rep[List[Element]])
+        val tag2 = document.findAll(unit("foo #bar baz"))
+        log(tag2: Rep[List[Element]])
+        val tag3 = document.findAll(unit("foo .bar baz"))
+        log(tag3: Rep[List[Element]])
+        val id = document.find(unit(" #foo-1-0    "))
+        log(id: Rep[Option[Element]])
+        val error = document.find(unit(" #---    "))
+        log(error: Rep[Option[Element]])
+        val id2 = document.find(unit(" #-a_a    "))
+        log(id2: Rep[Option[Element]])
+        val id3 = document.find(unit(" #bbaa    "))
+        log(id3: Rep[Option[Element]])
         val input = document.find[Input](unit("input"))
         log(input: Rep[Option[Input]])
       }

### 2013-05-23 papers/gpce2013/gpce2013.tex
Taking shape

@@
 \usepackage{amsmath}
+\usepackage[T1]{fontenc}
 \usepackage[utf8]{inputenc}
@@
 
+(Mention the problem of using client-side libraries to get high-level abstractions: double runtime overhead --
+download and execution)
+
 Using the same programming language on both server-side and client-side can improve the software engineering
@@
 
-This paper presents such a compiled embedded DSL allowing developers to write Web applications in a single language
-which code fragments can be shared between client and server sides, and which is efficiently compiled to each side.
-More precisely, we demonstrate the following features:
+\texttt{js-scala}~\cite{Kossakowski12_JsDESL} is a compiled embedded DSL defined in Scala that generates JavaScript
+code. In this paper we present additional contributions on top of this DSL. We use staging (explain) to generate
+efficient code for typical abstractions used for Web programming and to generate efficient specialized code for both
+client and server sides for shared abstractions. More precisely, we demonstrate the following features:
 
 \begin{itemize}
- \item Type-directed ad-hoc polymorphism on client-side without runtime dynamic dispatch logic;
+ \item An API for searching in the DOM, that exposes a single entry point but that generates code potentially using
+more optimized native APIs;
  \item Usage of monads without extra container object creation;
@@
 code using standard APIs on both server-side and client-side;
- \item An API for searching in the DOM, that exposes a single entry point but that generates code potentially using
-more optimized native APIs.
+ \item Type-directed ad-hoc polymorphism on client-side without runtime dynamic dispatch logic.
+ \item (Modules?)
 \end{itemize}
@@
 
+\subsection{Selectors}
+
+In a Web application, the user interface is defined by a HTML document which can be manipulated by the JavaScript
+code to be updated. One typical operation consists in searching some “interesting” element in the document, in order
+to extract its content, replace it or listen to user events fired on this element. The standard API provides several
+function to search elements in a Web document according to their characteristics such as their attribute values.
+Figure~\ref{selectors-api} (wrong reference number) summarizes the available functions and their differences.
+
+\begin{figure}
+\label{selectors-api}
+\begin{tabular}{| l | p{3cm} |}
+\hline
+Function & Description \\
+\hline
+\code{querySelector(s)} & First element matching the CSS selector \code{s} \\
+\hline
+\code{getElementById(i)} & Element which attribute \code{id} equals to \code{i} \\
+\hline
+\code{querySelectorAll(s)} & All nodes matching the CSS selector \code{s} \\
+\hline
+\code{getElementsByTagName(n)} & All elements of type \code{n} \\
+\hline
+\code{getElementsByClassName(c)} & All elements which \code{class} attribute equals to \code{c} \\
+\hline
+\end{tabular}
+\caption{Standard selectors API}
+\end{figure}
+
+The \code{querySelector} and \code{querySelectorAll} are the most general functions, the others handle special
+cases. For the developer it is not convenient to have to master several functions performing similar tasks. In fact,
+most JavaScript developers use the jQuery library (ref and stats) that provides just one function to search for
+elements. Listing~\ref{jquery-vanilla-api} shows two equivalent snippets of code, the first one using the native
+APIs (a) and the second one using jQuery (b).
+
+\begin{figure}
+\begin{lstlisting}[language=JavaScript]
+function getWords() {
+  var form = document.getElementById('add-user');
+  var sections =
+    form.getElementsByTagName('fieldset');
+  var results = [];
+  for (var i = 0 ; i < sections.length ; i++) {
+    var words = sections[i]
+      .getElementsByClassName('word');
+    results[i] = words;
+  }
+  return results
+}
+\end{lstlisting}
+\caption{(a)}
+\begin{lstlisting}
+function getWords() {
+  var form = $('#add-user');
+  var sections = $('fieldset', form);
+  return sections.map(function () {
+    return $('.word', this)
+  })
+}
+\end{lstlisting}
+\caption{(b)}
+\label{jquery-vanilla-api}
+\end{figure}
+
+jQuery provides an API that is simpler to master because it has less functions, but by doing so it can not benefit
+from the performance of the browser’s implementation of specialized search functions (\code{getElementById},
+\code{getElementsByTagName} and \code{getElementsByClassName}).
+
+Using LMS we are able to provide a simple API (like jQuery) and to generate code using specialized functions (and
+hence benefit from their performances), by analyzing the staged program. Listing~\ref{js-scala-selectors} shows how
+to implement listing~\ref{jquery-vanilla-api} using js-scala. This listing generates a JavaScript program similar to
+listing~\ref{jquery-vanilla-api} (a).
+
+\begin{figure}
+\label{js-scala-selectors}
+\begin{lstlisting}
+def getWords() = {
+  val form = document.find("#add-user")
+  val sections = form.findAll("fieldset")
+  sections map (_.findAll(".word"))
+}
+\end{lstlisting}
+\caption{Selectors in Js-Scala}
+\end{figure}
+
+Writing the program with js-scala instead of using a JavaScript library such as jQuery is interesting because the
+abstraction for searching elements exist only in the initial program source code, not in the final JavaScript
+program. It gives two advantages: (1) the execution of the final program is more performant because of the use of
+specialized APIs, (2) the final program’s size is smaller because it does not need to include jQuery.
+
 \subsection{Monads Sequencing}
 
-As an illustration of the staging mechanism, we present a simple DSL to handle null references. This DSL provides an
-abstraction at the stage-level that is removed by optimization during the code generation.
+As another illustration of the staging mechanism, we present a simple DSL to handle null references. This DSL
+provides an abstraction at the stage-level that is removed by optimization during the code generation.
 
@@
 
+\begin{figure}
+\label{dom-api}
+\begin{lstlisting}[language=JavaScript]
+var articleUi = function (article) {
+  var div = document.createElement('div');
+  div.setAttribute('class', 'article');
+  var span = document.createElement('span');
+  var name =
+    document.createTextNode(article.name + ': ');
+  span.appendChild(name);
+  div.appendChild(span);
+  var strong = document.createElement('strong');
+  var price = document.createTextNode(article.price);
+  strong.appendChild(price);
+  div.appendChild(strong);
+  return div
+};
+\end{lstlisting}
+
+\begin{lstlisting}[language=HTML]
+<div class=article>
+  <span>French wine: </span>
+  <strong>10</strong>
+</div>
+\end{lstlisting}
+\caption{DOM}
+\end{figure}
+
 In this section we show how we can define a template engine as an embedded DSL with minimal effort. This template
@@
 \begin{figure}
-\begin{lstlisting}[label=forest-hello,caption=DOM definition DSL]
-def vertexDom(v: Rep[Vertex]) =
-    el("g", "class"->"vertex",
-            "transform"->("translate("+v.x+","+v.y+")"))(
-        el("rect", "width"->v.width, "height"->v.height)(),
-        el("text", "width"->v.width, "height"->v.height)(
-            v.content
-        )
+\begin{lstlisting}[label=forest,caption=DOM definition DSL]
+def articleUi(article: Rep[Article]) =
+    el('div, 'class -> 'article)(
+        el('span)(article.name + ": "),
+        el('strong)(article.price)
     )
@@
 
-Listing \ref{forest-hello} uses our DSL and generates a code equivalent to listing \ref{dom-api}. The readability has
+Listing \ref{forest} uses our DSL and generates a code equivalent to listing \ref{dom-api}. The readability has
 been highly improved: nesting tags is just like nesting code blocks, HTML entities are
@@
 
-\subsubsection{Reuse the DOM definition DSL on server-side}
-
 Our DSL is equivalent to a template engine with Scala as the expression language. Making it usable on both server and
@@
 
-For instance, the template written in listing \ref{forest-hello} produces the following Scala code usable on
-server-side (the generated code for client-side is roughly equivalent to listing \ref{dom-api}):
+For instance, the template written in listing \ref{forest} produces the following Scala code usable on
+server-side:
 
 \begin{lstlisting}
-def vertexDom(v: Vertex) = {
-  val x0 =
-    <text width="{v.width}" height="{v.height}">
-      {v.content}
-    </text>
-  val x1 = <rect width="{v.width}" height="{v.height}" />
+def articleUi(article: Article) = {
+  val x0 = <span>{ article.name + ": " }</span>
+  val x1 = <strong>{ article.price }</strong>
   val x2 =
-    <g class="vertex" transform="translate({v.x},{v.y})">
+    <div class="article">
       {List(x0, x1)}
-    </g>
+    </div>
   x2
@@
 
-We are able to tackle the code sharing issues described in section \ref{assessment} because of the embdedded nature
+We are able to tackle the code sharing issues described in the introduction because of the embdedded nature
 of our DSLs: dynamic content of templates is written using embedded DSLs too, so their translation into JavaScript
@@
 
+\section{Implementation}
+\label{implementation}
+
+See http://github.com/js-scala/js-scala.
+
 \subsection{Selectors}
 
-querySelectorAll, getElementsByTagName, getElementsByClassName, getElementById, querySelector
+\begin{figure}
+\label{selector-impl}
+\begin{lstlisting}
+def find(receiver: Exp[SelectorApi],
+         selector: Exp[String]) =
+  getConstIdentifier(selector) match {
+    case Some(id) if receiver == document =>
+      DocumentGetElementById(Const(id))
+    case _ =>
+      SelectorFind(receiver, selector)
+  }
+\end{lstlisting}
+\caption{Selectors optimization}
+\end{figure}
 
-\section{Implementation?}
+\subsection{DOM Fragments}
 
-See http://github.com/js-scala/js-scala.
+\subsection{Ad-Hoc Polymorphism}
 

### 2013-05-23 src/main/scala/scala/js/InScala.scala
- Ignore tests that do not pass and do not matter for now - Fix some warnings

@@ -3,6 +3,7 @@ package scala.js
 import scala.virtualization.lms.common._
 import scala.reflect.SourceContext
 
+// TODO Get rid of this trait
 trait InScala extends Base {
   type Rep[+T] = T
 

### 2013-05-23 src/test/scala/scala/js/FileDiffSuite.scala

@@ -6,7 +6,7 @@ import org.scalatest._
 
 trait FileDiffSuite extends Suite {
   
-  def withOutFile(name: String)(func: => Unit): Unit = {
+  def withOutFile(name: String)(func: => Unit) {
     val file = new File(name)
     file.getParentFile.mkdirs()
     withOutput(new PrintStream(new FileOutputStream(file)))(func)
@@ -16,7 +16,7 @@ trait FileDiffSuite extends Suite {
     withOutput(new PrintStream(bstream))(func)
     bstream.toString
   }
-  def withOutput(out: PrintStream)(func: => Unit): Unit = {
+  def withOutput(out: PrintStream)(func: => Unit) {
     val oldStdOut = System.out
     val oldStdErr = System.err
     try {
@@ -38,8 +38,8 @@ trait FileDiffSuite extends Suite {
     fis.close()
     new String(buf)
   }
-  def assertFileEqualsCheck(name: String): Unit = {
-    expect(readFile(name+".check")){readFile(name)}
+  def assertFileEqualsCheck(name: String) {
+    expectResult(readFile(name+".check")){readFile(name)}
     new File(name) delete ()
   }
 }

### 2013-05-23 src/test/scala/scala/js/TestDynamic.scala

@@ -13,7 +13,7 @@ trait DynamicProg { this: DynamicBase =>
 
 trait AllocProg { this: JS =>
   def test(x: Rep[Any]): Rep[Any] = {
-    val f = fun { (y : Rep[Any]) => new Record { val a = 42/*newDynamic("Foo")()*/ } }
+    val f = fun { (y : Rep[Any]) => new Record { val a = (newDynamic("Foo")(): Rep[Any]) } }
     f.apply(x)
   }
 }

### 2013-05-23 src/test/scala/scala/js/TestLiteral.scala

@@ -4,6 +4,7 @@ import scala.virtualization.lms.common._
 
 import java.io.PrintWriter
 import java.io.FileOutputStream
+import org.scalatest.Ignore
 
 trait LiteralProg { this: Structs with LiftNumeric with NumericOps =>
   def test(x: Rep[Double]): Rep[Double] = {
@@ -31,8 +32,9 @@ trait LiteralFunProg { this: JS =>
 
 class TestLiteral extends FileDiffSuite {
   val prefix = "test-out/"
-  
-  def testLiteral = {
+
+  // Ignore these tests until self references are handled by structs
+  @Ignore def testLiteral = {
     withOutFile(prefix+"literal") {
       new LiteralProg with StructExp with LiftNumeric with NumericOpsExpOpt { self =>
         val codegen = new JSNestedCodegen with JSGenStruct with JSGenNumericOps { val IR: self.type = self }
@@ -42,7 +44,7 @@ class TestLiteral extends FileDiffSuite {
     assertFileEqualsCheck(prefix+"literal")
   }
 
-  def testLiteralFun = {
+  @Ignore def testLiteralFun = {
     withOutFile(prefix+"literalfun") {
       new LiteralFunProg with JSExp { self =>
         val codegen = new JSGen { val IR: self.type = self }

### 2013-05-23 src/test/scala/scala/js/TestRegExps.scala

@@ -46,21 +46,21 @@ class TestRegExps extends FileDiffSuite {
 
   def testScala = {
     new RegExpsProgInScala { self =>
-      expect(true){testTest("123")}
-      expect(true){testTest("x123y")}
-      expect(false){testTest("xy")}
+      expectResult(true){testTest("123")}
+      expectResult(true){testTest("x123y")}
+      expectResult(false){testTest("xy")}
 
-      expect(true){testTestWhole("123")}
-      expect(false){testTestWhole("x123y")}
-      expect(false){testTestWhole("xy")}
+      expectResult(true){testTestWhole("123")}
+      expectResult(false){testTestWhole("x123y")}
+      expectResult(false){testTestWhole("xy")}
 
-      expect(0){testSearch("123")}
-      expect(1){testSearch("x123y")}
-      expect(-1){testSearch("xy")}
+      expectResult(0){testSearch("123")}
+      expectResult(1){testSearch("x123y")}
+      expectResult(-1){testSearch("xy")}
 
-      expect(true){testSearchMatch("123")}
-      expect(true){testSearchMatch("x123y")}
-      expect(false){testSearchMatch("xy")}
+      expectResult(true){testSearchMatch("123")}
+      expectResult(true){testSearchMatch("x123y")}
+      expectResult(false){testSearchMatch("xy")}
     }
   }
 }

### 2013-05-23 src/test/scala/scala/js/TestTraits.scala

@@ -4,6 +4,7 @@ import scala.virtualization.lms.common._
 
 import java.io.PrintWriter
 import java.io.FileOutputStream
+import org.scalatest.Ignore
 
 trait TraitsProg { this: JS with JSTraits =>
   trait Foo {
@@ -90,11 +91,12 @@ class TestTraits extends FileDiffSuite {
     assertFileEqualsCheck(prefix+"traits-double-extends")
   }
 
-  def testTraitsInScala = {
+  // VariablesInScala is not implemented, so this test should be ignored
+  @Ignore def testTraitsInScala = {
     new TraitsProgInScala { self =>
-      expect(9){test(3)}
-      expect(10){testExtends(3)}
-      expect(3){testDoubleExtends(0)}
+      expectResult(9){test(3)}
+      expectResult(10){testExtends(3)}
+      expectResult(3){testDoubleExtends(0)}
     }
   }
 }

### 2013-05-24 papers/gpce2013/gpce2013.tex
Some changes

@@
            {IRISA, Université de Rennes 1}
-           {\{firstname\}.\{lastname\}@irisa.fr}
+           {\{first\}.\{last\}@irisa.fr}
 
@@
 \begin{abstract}
-This is the text of the abstract.
+Writing code partitioned between heterogeneous client-side and server-side makes the engineering process of Web
+applications more tedious. We use a deep embedded DSL to define a common language abstracting over platform
+differences for the developers point of view, but generating efficient code leveraging the target platforms
+characteristics and giving them control on the amount of code produced on client-side.
 \end{abstract}
@@
 
+\begin{figure}
+  \centering
+  \includegraphics[width=8cm]{langs.pdf}
+  \caption{Language engineering processes}
+  \label{langs}
+\end{figure}
+
 \subsection{Fat Languages}
@@
 One way consists in hard-coding, in the language's compiler, the transformation to each target platform. Figure
-\ref{fat-lang} depicts this process. In order to support a feature related to a specific domain, the whole compiler
-pipeline (parser, code generator, \etc) may have to be adapted. This approach gives \emph{fat} languages because
-a lot of concepts are defined at the language level. Examples of such languages are Links~\cite{Cooper07_Links},
-Opa and Dart~\cite{Griffith11_Dart}. These languages are difficult to extend because each concept is defined in the
-compiler, and modifying a compiler requires a high effort. Furthermore, these languages also require to support
-common abstraction and composition mechanisms, as general purpose languages do, in order to make it easier to write
-real programs. So they need to be at least general purpose languages. It means that for every problem you have to
-rewrite a full-featured programming language before addressing the concepts specific to the problem domain. This
-approach does not scale.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=3cm]{fat.pdf}
-  \caption{Fat language}
-  \label{fat-lang}
-\end{figure}
+\ref{langs} (left) depicts this process. In order to support a feature related to a specific domain, the whole
+compiler pipeline (parser, code generator, \etc) may have to be adapted. This approach gives \emph{fat} languages
+because a lot of concepts are defined at the language level. Examples of such languages are
+Links~\cite{Cooper07_Links}, Opa and Dart~\cite{Griffith11_Dart}. These languages are difficult to extend because
+each concept is defined in the compiler, and modifying a compiler requires a high effort. Furthermore, these
+languages also require to support common abstraction and composition mechanisms, as general purpose languages do, in
+order to make it easier to write real programs. So they need to be at least general purpose languages. It means that
+for every problem you have to rewrite a full-featured programming language before addressing the concepts specific to
+the problem domain. This approach does not scale.
 
@@
 Alternatively, one can define concepts relative to a specific domain as libraries on top of a thin general purpose
-language. Figure \ref{thin-lang} depicts this approach. The general purpose language is used as a host language that
-does not need to be modified if a new concept is introduced, because concepts are defined as libraries. However,
+language. Figure \ref{langs} (middle) depicts this approach. The general purpose language is used as a host language
+that does not need to be modified if a new concept is introduced, because concepts are defined as libraries. However,
 this approach gives no opportunity to translate a concept efficiently, according to the host platform
@@
 
-\begin{figure}
-  \centering
-  \includegraphics[width=3cm]{thin.pdf}
-  \caption{Thin language}
-  \label{thin-lang}
-\end{figure}
-
 \subsection{Deeply Embedded Languages}
 
-Figure \ref{dedsl} shows a last approach that tries to get the benefits of the previous two approaches, without
-their shortcomings.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=3cm]{lms.pdf}
-  \caption{Deeply embedded languages}
-  \label{dedsl}
-\end{figure}
+Figure \ref{langs} (right) shows a last approach that tries to get the benefits of the previous two approaches,
+without their shortcomings.
 
@@
 
-We are able to tackle the code sharing issues described in the introduction because of the embdedded nature
-of our DSLs: dynamic content of templates is written using embedded DSLs too, so their translation into JavaScript
-and Scala is managed by their respective code generators.
+We are able to tackle the code sharing issues described in the introduction (more details?) because of the embdedded
+nature of our DSLs: dynamic content of templates is written using embedded DSLs too, so their translation into
+JavaScript and Scala is managed by their respective code generators.
 
@@
 
+\subsection{Null references}
+
+Code generation consists in traversing the statement nodes produced by the program evaluation according to their
+dependencies and to emit the code corresponding to each statement. LMS already sorts the statements graph so DSL
+authors just need to say how to emit code for each statement node of their DSL. Listing \ref{option-codegen} shows
+such a code generator for the \code{null} reference handling DSL. The \code{emitNode} method handles
+\code{OptionIsEmpty} and \code{OptionForeach} nodes. In the case of the \code{OptionIsEmpty} node, it simply
+generates an expression testing if the value is \code{null}, in the case of the \code{OptionForeach} node, it wraps
+the code block dereferencing the value within a \code{if} checking that the value is not \code{null}.
+
+\begin{figure}
+\begin{lstlisting}[caption=Null reference handling DSL code generator,label=option-codegen]
+trait JSGenOptionOps extends JSGenEffect {
+  val IR: EfffectExp with OptionOpsExp
+  import IR._
+  override def emitNode(sym: Sym[Any], rhs: Def[Any]) =
+    rhs match {
+      case OptionIsEmpty(o) =>
+        emitValDef(sym, quote(o) + " === null")
+      case OptionForeach(o, a, b) =>
+        stream.println("if (" + quote(o) + " !== null) {")
+        emitValDef(a, quote(o))
+        emitBlock(b)
+        stream.println("}")
+      case _ =>
+        super.emitNode(sym, rhs)
+    }
+}
+\end{lstlisting}
+\end{figure}
+
 \subsection{DOM Fragments}
@@
 
-Chooze.
+Chooze~\footnote{\href{http://chooze.herokuapp.com}{http://chooze.herokuapp.com}} is an existing complete
+application for making polls. It allows users to create a poll, define the choice alternatives, share the poll, vote
+and look at the results. It contains some JavaScript code to handle the dynamic behavior of the application:
+double-posting prevention, dynamic form update and rich interaction with the document.
 
-\subsection{Several Implementations}
-
-\subsubsection{Vanilla JavaScript}
-
-\subsubsection{jQuery}
-
-\subsubsection{GWT}
-
-\subsubsection{SharpKit}
-
-\subsubsection{Js-Scala}
+The application was initially written using jQuery. We rewrote it using several technologies for the client-side
+part: plain JavaScript (without third-party library), js-scala, GWT and SharpKit.
 
 \subsection{Benchmarks, Code Metrics}
+
+Our goal is to evalute the level of abstraction provided by each solution and their performances. We took the
+number of lines of code as a measure of the level of abstraction and we wrote micro-benchmarks to evaluate the
+runtime performances of a solution. We also measured the size of the data sent to the client-side. We also measured
+the ability to share code between client and server sides.
 

### 2013-05-24 src/main/scala/scala/js/JSDom.scala
optimize nodeList with functions size(), foreach() and filter()

@@ -72,8 +72,20 @@ trait JSDom extends Base {
   }
   // Note that selector_find[A](…)(implicit ev: A <:< Element) was simpler but sometimes didn’t make the type inferencer happy
   def selector_find[A : Selectable : Manifest](s: Rep[SelectorApi], selector: Rep[String]): Rep[Option[A]]
-  def selector_findAll[A : Selectable : Manifest](s: Rep[SelectorApi], selector: Rep[String]): Rep[List[A]]
+  def selector_findAll[A : Selectable : Manifest](s: Rep[SelectorApi], selector: Rep[String]): Rep[NodeList[A]]
 
+  class NodeList[A : Selectable]
+  implicit class NodeListOps[A : Manifest](ns: Rep[NodeList[A]]){
+    def size = nodeList_size[A](ns)
+    def filter(f:Rep[A] => Rep[Boolean]) = nodeList_filter[A](ns,f)
+    def foreach(f:Rep[A] => Rep[Unit]) = nodeList_foreach[A](ns,f)
+  }
+  def nodeList_size[A : Manifest](s: Rep[NodeList[A]]): Rep[Int]
+  def nodeList_filter[A : Manifest](ns: Rep[NodeList[A]],f: Rep[A] => Rep[Boolean]): Rep[List[A]]
+  def nodeList_foreach[A : Manifest](ns: Rep[NodeList[A]],f: Rep[A] => Rep[Unit]) : Rep[Unit]
+  
+
+  
   trait Document extends SelectorApi with EventTarget
 
   trait Element extends EventTarget with SelectorApi
@@ -172,6 +184,22 @@ trait JSDomExp extends JSDom with EffectExp with JSFunctionsExp with OptionOpsEx
   def infix_document(w: Exp[Window]) = WindowDocument
   def infix_history(w: Exp[Window]) = WindowHistory
 
+  def nodeList_size[A : Manifest](ns: Exp[NodeList[A]]) = {
+    reflectEffect(NodeListSize[A](ns))
+  }
+  
+  def nodeList_filter[A : Manifest](ns: Exp[NodeList[A]],f: Exp[A] => Exp[Boolean]) = {
+    val it = fresh[A]
+    val block = reifyEffects(f(it))
+    reflectEffect(NodeListFilter(ns, it, block), summarizeEffects(block).star)
+  }
+  
[…]

### 2013-05-24 src/test/scala/scala/js/TestDom.scala

@@ -21,14 +21,14 @@ class TestDom extends FileDiffSuite {
       }
 
       def selectors() = {
-        val tag = document.findAll(unit("babar"))
+        /*val tag = document.findAll(unit("babar"))
         log(tag: Rep[List[Element]])
         val classe = document.findAll(unit(".babar"))
         log(classe: Rep[List[Element]])
         val tag2 = document.findAll(unit("foo #bar baz"))
         log(tag2: Rep[List[Element]])
         val tag3 = document.findAll(unit("foo .bar baz"))
-        log(tag3: Rep[List[Element]])
+        log(tag3: Rep[List[Element]])*/
         val id = document.find(unit(" #foo-1-0    "))
         log(id: Rep[Option[Element]])
         val error = document.find(unit(" #---    "))

### 2013-05-24 src/test/scala/scala/js/TestNodeList.scala

@@ -1,0 +1,40 @@
+package scala.js
+
+import scala.virtualization.lms.common._
+import java.io.PrintWriter
+import scala.xml.Document
+
+class TestNodeList  extends FileDiffSuite {
+  val prefix = "test-out/"
+  
+  trait DSL extends Base with JSDom with JSDebug with Equal
+  trait DSLExp extends DSL with JSDomExp with JSDebugExp with EqualExp 
+  trait DSLJSGen extends JSGenEffect with JSGenDom with JSGenDebug with JSGenEqual { val IR: DSLExp }
+
+  def testOn() {
+
+    trait Prog extends DSL {
+
+      def testSize(ns: Rep[NodeList[Element]]) = ns.size
+      def testFilter(ns: Rep[NodeList[Element]]) = ns.filter(n => n == 2)
+      def testForeach(ns: Rep[NodeList[Element]]) = ns.foreach(n => log(n))
+  
+      def nodelist() = {
+        val tag = document.findAll(unit("babar"))
+        log(tag)
+        val size = tag.size
+        log(size)
+        val filter = testFilter(tag)
+        log(filter)
+        val foreach = testForeach(tag)
+        log(foreach)
+      }
+    }
+    withOutFile(prefix+"nodeList") {
+      val prog = new Prog with DSLExp
+      val codegen = new DSLJSGen { val IR: prog.type = prog }
+      codegen.emitSource0(prog.nodelist _, "nodeList", new PrintWriter(System.out))
+    }
+    assertFileEqualsCheck(prefix+"nodeList")
+  }
+}

### 2013-05-24 src/main/scala/scala/js/JSDom.scala
enrich nodeList with functions size(), foreach() and filter()

@@ -76,9 +76,9 @@ trait JSDom extends Base {
 
   class NodeList[A : Selectable]
   implicit class NodeListOps[A : Manifest](ns: Rep[NodeList[A]]){
-    def size = nodeList_size[A](ns)
-    def filter(f:Rep[A] => Rep[Boolean]) = nodeList_filter[A](ns,f)
-    def foreach(f:Rep[A] => Rep[Unit]) = nodeList_foreach[A](ns,f)
+    def size = nodeList_size(ns)
+    def filter(f:Rep[A] => Rep[Boolean]) = nodeList_filter(ns,f)
+    def foreach(f:Rep[A] => Rep[Unit]) = nodeList_foreach(ns,f)
   }
   def nodeList_size[A : Manifest](s: Rep[NodeList[A]]): Rep[Int]
   def nodeList_filter[A : Manifest](ns: Rep[NodeList[A]],f: Rep[A] => Rep[Boolean]): Rep[List[A]]
@@ -185,7 +185,7 @@ trait JSDomExp extends JSDom with EffectExp with JSFunctionsExp with OptionOpsEx
   def infix_history(w: Exp[Window]) = WindowHistory
 
   def nodeList_size[A : Manifest](ns: Exp[NodeList[A]]) = {
-    reflectEffect(NodeListSize[A](ns))
+    NodeListSize[A](ns)
   }
   
   def nodeList_filter[A : Manifest](ns: Exp[NodeList[A]],f: Exp[A] => Exp[Boolean]) = {

### 2013-05-24 src/test/scala/scala/js/TestDom.scala

@@ -21,24 +21,15 @@ class TestDom extends FileDiffSuite {
       }
 
       def selectors() = {
-        /*val tag = document.findAll(unit("babar"))
-        log(tag: Rep[List[Element]])
-        val classe = document.findAll(unit(".babar"))
-        log(classe: Rep[List[Element]])
-        val tag2 = document.findAll(unit("foo #bar baz"))
-        log(tag2: Rep[List[Element]])
-        val tag3 = document.findAll(unit("foo .bar baz"))
-        log(tag3: Rep[List[Element]])*/
-        val id = document.find(unit(" #foo-1-0    "))
-        log(id: Rep[Option[Element]])
-        val error = document.find(unit(" #---    "))
-        log(error: Rep[Option[Element]])
-        val id2 = document.find(unit(" #-a_a    "))
-        log(id2: Rep[Option[Element]])
-        val id3 = document.find(unit(" #bbaa    "))
-        log(id3: Rep[Option[Element]])
-        val input = document.find[Input](unit("input"))
-        log(input: Rep[Option[Input]])
+        document.findAll(unit("babar"))
+        document.findAll(unit(".babar"))
+        document.findAll(unit("foo #bar baz"))
+        document.findAll(unit("foo .bar baz"))
+        document.find(unit(" #foo-1-0    "))
+        document.find(unit(" #---    "))
+        document.find(unit(" #-a_a    "))
+        document.find(unit(" #bbaa    "))
+        document.find[Input](unit("input"))
       }
     }
     withOutFile(prefix+"dom") {

### 2013-05-24 src/test/scala/scala/js/TestNodeList.scala

@@ -19,21 +19,18 @@ class TestNodeList  extends FileDiffSuite {
       def testFilter(ns: Rep[NodeList[Element]]) = ns.filter(n => n == 2)
       def testForeach(ns: Rep[NodeList[Element]]) = ns.foreach(n => log(n))
   
-      def nodelist() = {
-        val tag = document.findAll(unit("babar"))
-        log(tag)
-        val size = tag.size
-        log(size)
-        val filter = testFilter(tag)
+      def nodelist(ns: Rep[NodeList[Element]]) = {
+        val size = ns.size
+        log(ns.size)
+        val filter = testFilter(ns)
         log(filter)
-        val foreach = testForeach(tag)
-        log(foreach)
+        testForeach(ns)
       }
     }
     withOutFile(prefix+"nodeList") {
       val prog = new Prog with DSLExp
       val codegen = new DSLJSGen { val IR: prog.type = prog }
-      codegen.emitSource0(prog.nodelist _, "nodeList", new PrintWriter(System.out))
+      codegen.emitSource(prog.nodelist _, "nodeList", new PrintWriter(System.out))
     }
     assertFileEqualsCheck(prefix+"nodeList")
   }

### 2013-05-24 papers/gpce2013/gpce2013.tex
plop

@@
 Writing code partitioned between heterogeneous client-side and server-side makes the engineering process of Web
-applications more tedious. We use a deep embedded DSL to define a common language abstracting over platform
-differences for the developers point of view, but generating efficient code leveraging the target platforms
-characteristics and giving them control on the amount of code produced on client-side.
+applications tedious. We use a deep embedded DSL to define a common language abstracting over platform differences
+for the developers point of view, but generating efficient code leveraging the target platforms characteristics and
+giving them control on the amount of code produced on client-side.
 \end{abstract}
@@
 environments~\cite{Echeverria09_RIA,Kuuskeri09_PartitioningClientServer}. This gives less flexibility in the
-engineering process and requires a higher maintenance effort: if a piece of logic was implemented on
-client-side and it turns out that it needs to be implemented on server-side instead, the code can not be reused and
-the feature needs to be completely rewritten (and \emph{vice versa}). Even worse, logic parts that run on both
-client-side and server-side need to be duplicated. For instance, HTML fragments may be built from the server-side
-when a page is requested by a client, but they may also be built from the client-side to perform an incremental
-update subsequent to an user action. How could developers write HTML fragment definitions once and render them on
-both client-side and server-side? The more interactive the application is, the more logic needs to be duplicated
-between the server-side and the client-side (explain why?).
+engineering process and requires a higher maintenance effort: if a piece of logic is implemented on client-side and
+finally needs to be implemented on server-side instead, the code can not be reused and the feature needs to be
+completely rewritten (and \emph{vice versa}). Even worse, logic parts that run on both client-side and server-side
+need to be duplicated. For instance, HTML fragments may be built from the server-side when a page is requested by a
+client, but they may also be built from the client-side to perform an incremental update subsequent to an user
+action. How could developers write HTML fragment definitions once and render them on both client-side and
+server-side?
 
-(Mention the problem of using client-side libraries to get high-level abstractions: double runtime overhead --
-download and execution)
+The more interactive the application is, the more logic needs to be duplicated between the server-side
+and the client-side (explain why?), and the higher is the amount of client-side code. Developers use libraries and
+frameworks to get higher-level abstractions on client-side, making their code easier to reason about and to
+maintain. Unfortunately, JavaScript -- which is currently the most supported the action language on Web clients
+-- has no static linking system allowing to pick just the functions actually used by a program, forcing clients to
+download the whole library even though only a few functions are actually used.
 
 Using the same programming language on both server-side and client-side can improve the software engineering
-process by enabling code reuse between both sides. Incidentally, the JavaScript language -- which is currently the
-action language natively supported by most of Web clients -- can be used on server-side, and an increasing
-number of programming languages or compiler back-ends can generate JavaScript code (\eg
+process by enabling code reuse between both sides. Incidentally, the JavaScript language can be used on server-side,
+and an increasing number of programming languages or compiler back-ends can generate JavaScript code (\eg
 Java/GWT~\cite{Chaganti07_GWT}, SharpKit\footnote{\href{http://sharpkit.net}{http://sharpkit.net}},
@@
 are not the same. For instance, DOM fragments can be defined on client-side using the standard DOM API, but this
-API does not exist on server-side. How to define a common vocabulary for such concepts? And how to make it leverage
-the native APIs, when possible, for performance reasons?
+API does not exist on server-side. How to define a common vocabulary for such concepts? And how to make the
+executable code leverage the native APIs, when possible, for performance reasons?
 
@@
 
-\texttt{js-scala}~\cite{Kossakowski12_JsDESL} is a compiled embedded DSL defined in Scala that generates JavaScript
-code. In this paper we present additional contributions on top of this DSL. We use staging (explain) to generate
-efficient code for typical abstractions used for Web programming and to generate efficient specialized code for both
-client and server sides for shared abstractions. More precisely, we demonstrate the following features:
+\texttt{js-scala}~\cite{Kossakowski12_JsDESL} is such a compiled embedded DSL defined in Scala that generates
+JavaScript code, making it possible to write the client-side code of Web applications using JavaScript. This paper
+enriches js-scala in order to solve the problems described above. We use staging (explain) to generate efficient code
+for typical abstractions used for Web programming and to generate efficient specialized code for both client and
+server sides for shared abstractions. More precisely, we demonstrate the following features:
 
@@
 code using standard APIs on both server-side and client-side;
- \item Type-directed ad-hoc polymorphism on client-side without runtime dynamic dispatch logic.
- \item (Modules?)
+ \item Type-directed ad-hoc polymorphism on client-side without runtime dynamic dispatch logic;
+ \item Static linking.
 \end{itemize}
 
-(More on the results metrics)
+Compared to libraries offering the same level of abstraction, we observe that js-scala produces significantly smaller
+executables with better runtime performances.
 

### 2013-05-30 papers/gpce2013/gpce2013.tex
abstract

@@
 \begin{abstract}
-Writing code partitioned between heterogeneous client-side and server-side makes the engineering process of Web
-applications tedious. We use a deep embedded DSL to define a common language abstracting over platform differences
-for the developers point of view, but generating efficient code leveraging the target platforms characteristics and
-giving them control on the amount of code produced on client-side.
+Writing large Web applications is known to be difficult. One challenge comes from the fact that the application's
+logic is scattered into heterogeneous clients and servers, making it difficult to share code between both sides or to
+move code from one side to the other. Another challenge is performance: while Web applications rely on ever more
+code on the client-side, they may run on smart phones with little hardware capabilities. These two challenges raise
+the following problem: how to benefit from high-level languages and libraries making code complexity easier to manage
+and abstracting over the clients and servers differences without trading this engineering comfort for performance?
+This article presents high-level abstractions defined as deep embedded DSLs in Scala, that can (1) generate efficient
+code leveraging the target platform characteristics, (2) be shared between client and server code. Our DSL gives a
+performance / expressiveness ratio about two times higher than other approaches.
 \end{abstract}
@@
 and the client-side (explain why?), and the higher is the amount of client-side code. Developers use libraries and
-frameworks to get higher-level abstractions on client-side, making their code easier to reason about and to
+frameworks to get high-level abstractions on client-side, making their code easier to reason about and to
 maintain. Unfortunately, JavaScript -- which is currently the most supported the action language on Web clients
@@
  \item Type-directed ad-hoc polymorphism on client-side without runtime dynamic dispatch logic;
- \item Static linking.
+ \item Static linking?
 \end{itemize}
@@
 
-Problem: interoperability.
+Another approach consists in defining several small languages, each one being specific to a given set of concerns,
+and then to combine all the source artifacts written with these language into one executable program. This approach
+makes it difficult to make the separate languages inter-operate with each other.
 
@@
 The application was initially written using jQuery. We rewrote it using several technologies for the client-side
-part: plain JavaScript (without third-party library), js-scala, GWT and SharpKit.
+part: plain JavaScript (without third-party library), js-scala, GWT and HaXe.
+
+\paragraph*{Vanilla JavaScript} Low-level code.
+
+\paragraph*{jQuery} High-level code.
+
+\paragraph*{js-scala} High-level code. HTML fragment definition reused between server and client sides.
+
+\paragraph*{HaXe} Low-level code.
+
+\paragraph*{GWT} High-level code?
 
@@
 Our goal is to evalute the level of abstraction provided by each solution and their performances. We took the
-number of lines of code as a measure of the level of abstraction and we wrote micro-benchmarks to evaluate the
-runtime performances of a solution. We also measured the size of the data sent to the client-side. We also measured
-the ability to share code between client and server sides.
+number of lines of code as a measure of the level of abstraction. We also measured the size of the data sent to the
+client-side. We also measured the ability to share code between client and server sides.
+
+Rather than doing a micro-benchmark focusing on just one abstraction, we performed a global benchmark that is more
+likely to reflect the real performances of the application: we simulated user actions on a Web page (2000 clicks on
+buttons, triggering a dynamic update of the page and involving the use of the Option monad, the selectors API and
+the HTML fragment definition API) and measured the time it took complete them. We only measured the execution time of
+client-side code execution. The tests were run on a DELL Latitude E6430 laptop with 8 GB of RAM, on the Google
+Chrome v27 Web browser.
+
+\begin{figure}
+\label{benchmark}
+\centering
+\includegraphics[width=8cm]{chooze.png}
+\caption{Benchmarks on a real application}
+\end{figure}
 

### 2013-05-30 src/main/scala/scala/js/exp/dom/DomExp.scala
#9 Implement selectors optimizations into SelectorOpsExpOpt trait

@@ -3,3 +3,5 @@ package scala.js.exp.dom
 import scala.js.language.dom.Dom
 
 trait DomExp extends Dom with CoreExp with ElementOpsExp with EventOpsExp with NodeListOpsExp with BrowserExp with SelectorOpsExp
+
+trait DomExpOpt extends DomExp with SelectorOpsExpOpt

### 2013-05-30 src/main/scala/scala/js/exp/dom/SelectorOpsExp.scala

@@ -1,66 +1,61 @@
 package scala.js.exp.dom
 
-import scala.js.language.dom.{Browser, ElementOps, SelectorOps}
+import scala.js.language.dom.SelectorOps
 import scala.virtualization.lms.common.EffectExp
 
-trait SelectorOpsExp extends SelectorOps with EffectExp with ElementOps with Browser {
+trait SelectorOpsExp extends SelectorOps with EffectExp with ElementOpsExp {
 
-  def selector_find[A : Selectable : Manifest](s: Exp[Selector], selector: Exp[String]) = {
-         
-    //the Regex to recover the ID
-    val Id = "#((-?[A-Za-z0-9_]+)+)".r
-      
-    //pattern matching on the selector
-    selector match {
-      //if it is a constant
-      case Const(selectorString) => 
-        selectorString.trim match {
-          //if the first charactere is a '#' we are searching a ID
-          case Id(id, _) if s == document => 
-            reflectEffect(SelectorGetElementById[A](s, unit(id)))
-          //in other cases
-          case _ =>
-            reflectEffect(SelectorFind[A](s, selector))           
-        }
-      //if it's a variable
-      case _ => 
-        reflectEffect(SelectorFind[A](s, selector))
-    }  
-    
-  }
-    
-  def selector_findAll[A : Selectable : Manifest](s: Exp[Selector], selector: Exp[String]) = {
-    
-    //the Regex to recover the class
-    val Classe = "\\.((-?[A-Za-z0-9_]+)+)".r
-    //the Regex to recover the tag
-    val Tag = "((-?[A-Za-z0-9_]+)+)".r
-    
-    //pattern matching on the selector
-    selector match {
-      //if it is a constant 
-      case Const(selectorString) => 
-        selectorString.trim match {
-          //if the first charactere is a '.' we are searching a class
-          case Classe(classe, _) => 
-            reflectEffect(SelectorGetElementsByClassName[A](s, unit(classe)))            
-          //if it's a character chain we are searching a tag
-          case Tag(tag, _) => 
-            reflectEffect(SelectorGetElementsByTagName[A](s, unit(tag)))
-          //in other cases
-          case _ =>
-            reflectEffect(SelectorFindAll[A](s, selector))  
-        }
-      //if it's a variable
-      case _ => 
-         reflectEffect(SelectorFindAll[A](s, selector))
-    }
-  }
-  
[…]

### 2013-05-30 src/test/scala/scala/js/TestDom.scala

@@ -3,18 +3,17 @@ package scala.js
 import scala.virtualization.lms.common._
 import java.io.PrintWriter
 import language.Debug
-import scala.js.exp.DebugExp
-import scala.js.gen.js.GenEffect
-import scala.js.gen.js.GenDebug
-import scala.js.language.dom.Dom
-import scala.js.exp.dom.DomExp
-import scala.js.gen.js.dom.GenDom
+import language.dom.Dom
+import exp.DebugExp
+import exp.dom.DomExpOpt
+import gen.js.{GenEffect, GenDebug}
+import gen.js.dom.GenDom
 
 class TestDom extends FileDiffSuite {
   val prefix = "test-out/"
 
   trait DSL extends Base with Dom with Debug
-  trait DSLExp extends DSL with DomExp with DebugExp
+  trait DSLExp extends DSL with DomExpOpt with DebugExp
   trait DSLJSGen extends GenEffect with GenDom with GenDebug { val IR: DSLExp }
 
   def testOn() {

### 2013-05-30 src/test/scala/scala/js/TestArrays.scala
Clean some imports

@@ -3,12 +3,9 @@ package scala.js
 import scala.virtualization.lms.common._
 import java.io.PrintWriter
 import java.io.FileOutputStream
-import scala.js.language.Arrays
-import scala.js.exp.ArraysExp
-import scala.js.gen.js.GenArrays
-import scala.js.gen.js.GenNumericOps
-import scala.js.gen.js.GenEqual
-import scala.js.gen.js.GenArraysLegacy
+import language.Arrays
+import exp.ArraysExp
+import gen.js.{GenArrays, GenNumericOps, GenEqual, GenArraysLegacy}
 
 trait ArraysProg { this: LiftNumeric with NumericOps with Equal with Arrays =>
   def test1(x: Rep[Int]): Rep[Int] = {

### 2013-05-30 src/test/scala/scala/js/TestCPS.scala

@@ -3,23 +3,9 @@ package scala.js
 import scala.virtualization.lms.common._
 import scala.util.continuations._
 import java.io.PrintWriter
-import java.io.FileOutputStream
-import scala.js.language.JSLib
-import scala.js.language.JS
-import scala.js.language.CPS
-import scala.js.language.Casts
-import scala.js.language.Ajax
-import scala.js.language.Debug
-import scala.js.exp.AjaxExp
-import scala.js.exp.CPSExp
-import scala.js.exp.DebugExp
-import scala.js.exp.JSLibExp
-import scala.js.exp.JSExp
-import scala.js.gen.js.GenJSLib
-import scala.js.gen.js.GenAjax
-import scala.js.gen.js.GenCPS
-import scala.js.gen.js.GenJS
-import scala.js.gen.js.GenDebug
+import language.{JSLib, JS, CPS, Casts, Ajax, Debug}
+import exp.{AjaxExp, CPSExp, DebugExp, JSLibExp, JSExp}
+import gen.js.{GenJSLib, GenAjax, GenCPS, GenJS, GenDebug}
 
 trait CPSProg { this: JS with Debug with JSLib with CPS with Ajax with Casts =>
 

### 2013-05-30 src/test/scala/scala/js/TestCasts.scala

@@ -3,11 +3,9 @@ package scala.js
 import scala.virtualization.lms.common._
 import java.io.PrintWriter
 import java.io.FileOutputStream
-import scala.js.language.Casts
-import scala.js.language.JS
-import scala.js.exp.CastsCheckedExp
-import scala.js.gen.js.GenCastChecked
-import scala.js.exp.JSExp
+import language.{JS, Casts}
+import exp.{JSExp, CastsCheckedExp}
+import gen.js.GenCastChecked
 
 trait RefinementCastProg { this: Casts =>
   def test(x: Rep[Any]): Rep[Any] = {

### 2013-05-30 src/test/scala/scala/js/TestConditional.scala

@@ -3,10 +3,7 @@ package scala.js
 import scala.virtualization.lms.common._
 import java.io.PrintWriter
 import java.io.FileOutputStream
-import scala.js.gen.js.GenEffect
-import scala.js.gen.js.GenNumericOps
-import scala.js.gen.js.GenEqual
-import scala.js.gen.js.GenIfThenElse
+import gen.js.{GenEffect, GenNumericOps, GenEqual, GenIfThenElse}
 
 trait Print extends Base {
   implicit def unit(s: String): Rep[String]

### 2013-05-30 src/test/scala/scala/js/TestDynamic.scala

@@ -3,12 +3,9 @@ package scala.js
 import scala.virtualization.lms.common._
 import java.io.PrintWriter
 import java.io.FileOutputStream
-import scala.js.language.Dynamics
-import scala.js.language.JS
-import scala.js.exp.DynamicsExp
-import scala.js.gen.js.GenDynamics
-import scala.js.exp.JSExp
-import scala.js.gen.js.GenJS
+import language.{JS, Dynamics}
+import exp.{JSExp, DynamicsExp}
+import gen.js.{GenJS, GenDynamics}
 
 trait DynamicProg { this: Dynamics =>
   def test(x: Rep[Any]): Rep[Any] = {

### 2013-05-30 src/test/scala/scala/js/TestList.scala

@@ -2,9 +2,7 @@ package scala.js
 
 import scala.virtualization.lms.common._
 import java.io.PrintWriter
-import scala.js.gen.js.GenNumericOps
-import scala.js.gen.js.GenListOps
-import scala.js.gen.js.GenEffect
+import gen.js.{GenNumericOps, GenListOps, GenEffect}
 
 class TestList extends FileDiffSuite {
 

### 2013-05-30 src/test/scala/scala/js/TestLiteral.scala

@@ -4,12 +4,9 @@ import scala.virtualization.lms.common._
 import java.io.PrintWriter
 import java.io.FileOutputStream
 import org.scalatest.Ignore
-import scala.js.language.JS
-import scala.js.gen.js.NestedCodegen
-import scala.js.gen.js.GenStruct
-import scala.js.gen.js.GenNumericOps
-import scala.js.exp.JSExp
-import scala.js.gen.js.GenJS
+import language.JS
+import exp.JSExp
+import gen.js.{NestedCodegen, GenStruct, GenNumericOps, GenJS}
 
 trait LiteralProg { this: Structs with LiftNumeric with NumericOps =>
   def test(x: Rep[Double]): Rep[Double] = {

### 2013-05-30 src/test/scala/scala/js/TestMaps.scala

@@ -3,12 +3,9 @@ package scala.js
 import scala.virtualization.lms.common._
 import java.io.PrintWriter
 import java.io.FileOutputStream
-import scala.js.language.JSMaps
-import scala.js.language.JS
-import scala.js.exp.JSExp
-import scala.js.gen.js.GenJSMaps
-import scala.js.exp.JSMapsExp
-import scala.js.gen.js.GenJS
+import language.{JS, JSMaps}
+import exp.{JSExp, JSMapsExp}
+import gen.js.{GenJS, GenJSMaps}
 
 trait MapsProg { this: JS with JSMaps =>
   def test1(): Rep[Map[String, Int]] = {

### 2013-05-30 src/test/scala/scala/js/TestNodeList.scala

@@ -3,14 +3,12 @@ package scala.js
 import scala.virtualization.lms.common._
 import java.io.PrintWriter
 import scala.xml.Document
-import scala.js.language.Debug
-import scala.js.exp.DebugExp
-import scala.js.gen.js.GenEqual
-import scala.js.gen.js.GenDebug
-import scala.js.gen.js.GenEffect
-import scala.js.language.dom.Dom
-import scala.js.exp.dom.DomExp
-import scala.js.gen.js.dom.GenDom
+import language.Debug
+import language.dom.Dom
+import exp.DebugExp
+import exp.dom.DomExp
+import gen.js.{GenEqual, GenDebug, GenEffect}
+import gen.js.dom.GenDom
 
 class TestNodeList  extends FileDiffSuite {
   val prefix = "test-out/"

### 2013-05-30 src/test/scala/scala/js/TestOpt.scala

@@ -3,9 +3,9 @@ package scala.js
 import scala.virtualization.lms.common._
 import java.io.PrintWriter
 import java.io.FileOutputStream
-import scala.js.language.JS
-import scala.js.exp.JSExp
-import scala.js.gen.js.GenJSOpt
+import language.JS
+import exp.JSExp
+import gen.js.GenJSOpt
 
 trait OptProg { this: JS =>
   def test(x: Rep[Any]): Rep[Any] = {

### 2013-05-30 src/test/scala/scala/js/TestOptionOps.scala

@@ -2,11 +2,9 @@ package scala.js
 
 import scala.virtualization.lms.common._
 import java.io.PrintWriter
-import scala.js.language.OptionOps
-import scala.js.exp.OptionOpsExp
-import scala.js.gen.js.GenEffect
-import scala.js.gen.js.GenNumericOps
-import scala.js.gen.js.GenOptionOps
+import language.OptionOps
+import exp.OptionOpsExp
+import gen.js.{GenEffect, GenNumericOps, GenOptionOps}
 
 class TestOptionOps extends FileDiffSuite {
   val prefix = "test-out/"

### 2013-05-30 src/test/scala/scala/js/TestProxy.scala

@@ -3,12 +3,9 @@ package scala.js
 import scala.virtualization.lms.common._
 import java.io.PrintWriter
 import java.io.FileOutputStream
-import scala.js.language.Proxy
-import scala.js.exp.ProxyExp
-import scala.js.language.JS
-import scala.js.exp.JSExp
-import scala.js.gen.js.GenJS
-import scala.js.gen.js.GenProxy
+import language.{JS, Proxy}
+import exp.{JSExp, ProxyExp}
+import gen.js.{GenJS, GenProxy}
 
 trait TestProxyDummy extends Proxy {
   trait Dummy {

### 2013-05-30 src/test/scala/scala/js/TestStruct.scala

@@ -2,10 +2,8 @@ package scala.js
 
 import scala.virtualization.lms.common._
 import java.io.PrintWriter
-import scala.js.exp.JSLiteralExp
-import scala.js.gen.js.GenJSLiteral
-import scala.js.gen.js.NestedCodegen
-import scala.js.gen.js.GenStruct
+import exp.JSLiteralExp
+import gen.js.{GenJSLiteral, NestedCodegen, GenStruct}
 
 trait PersonDSL extends Base {
   type Person

### 2013-05-30 src/test/scala/scala/js/TestTraits.scala

@@ -4,12 +4,9 @@ import scala.virtualization.lms.common._
 import java.io.PrintWriter
 import java.io.FileOutputStream
 import org.scalatest.Ignore
-import scala.js.language.JS
-import scala.js.language.Traits
-import scala.js.exp.TraitsExp
-import scala.js.exp.JSExp
-import scala.js.gen.js.GenJS
-import scala.js.gen.js.GenTraits
+import language.{JS, Traits}
+import exp.{JSExp, TraitsExp}
+import gen.js.{GenJS, GenTraits}
 
 trait TraitsProg { this: JS with Traits =>
   trait Foo {

### 2013-05-30 src/test/scala/scala/js/TestTuple.scala

@@ -3,9 +3,9 @@ package scala.js
 import scala.virtualization.lms.common._
 import java.io.PrintWriter
 import java.io.FileOutputStream
-import scala.js.gen.js.{GenTupleOps, GenNumericOps, NestedCodegen, GenFunctions, GenJS}
-import scala.js.exp.JSExp
-import scala.js.language.JS
+import gen.js.{GenTupleOps, GenNumericOps, NestedCodegen, GenFunctions, GenJS}
+import exp.JSExp
+import language.JS
 
 trait TwoArgsProg { this: NumericOps =>
   def test(x: Rep[Double], y: Rep[Double]): Rep[Double] = {

### 2013-05-30 src/test/scala/scala/js/TestUnit.scala

@@ -2,9 +2,9 @@ package scala.js
 
 import java.io.PrintWriter
 
-import scala.js.exp.JSExp
-import scala.js.gen.js.GenJS
-import scala.js.language.JS
+import language.JS
+import exp.JSExp
+import gen.js.GenJS
 
 trait UnitProg { this: JS =>
   def test(x: Rep[Any]): Rep[Any] = {

### 2013-05-30 src/main/scala/scala/js/exp/JSExp.scala
Add DomExpOpt to JSExpOpt

@@ -1,8 +1,9 @@
 package scala.js.exp
 
 import scala.js.language.JS
+import scala.js.exp.dom.DomExpOpt
 
 trait JSExp extends JS with JsScalaExp with DynamicsExp with ArraysExp
   with RegExpsExp with OptionOpsExp
 
-trait JSExpOpt extends JSExp with JsScalaExpOpt
+trait JSExpOpt extends JSExp with JsScalaExpOpt with DomExpOpt

### 2013-05-30 papers/gpce2013/gpce2013.tex
taking shape

@@
 frameworks to get high-level abstractions on client-side, making their code easier to reason about and to
-maintain. Unfortunately, JavaScript -- which is currently the most supported the action language on Web clients
--- has no static linking system allowing to pick just the functions actually used by a program, forcing clients to
-download the whole library even though only a few functions are actually used.
+maintain, but also making their code run less efficiently (abstraction penalty).
 
-Using the same programming language on both server-side and client-side can improve the software engineering
-process by enabling code reuse between both sides. Incidentally, the JavaScript language can be used on server-side,
-and an increasing number of programming languages or compiler back-ends can generate JavaScript code (\eg
-Java/GWT~\cite{Chaganti07_GWT}, SharpKit\footnote{\href{http://sharpkit.net}{http://sharpkit.net}},
-Dart~\cite{Griffith11_Dart}, Kotlin\footnote{\href{http://kotlin.jetbrains.org/}{http://kotlin.jetbrains.org/}},
+Using the same programming language on both server-side and client-side could improve the software engineering
+process by enabling code reuse between both sides. Incidentally, the JavaScript language -- which is currently the
+most supported the action language on Web clients -- can be used on server-side, and an increasing number of
+programming languages or compiler back-ends can generate JavaScript code (\eg Java/GWT~\cite{Chaganti07_GWT},
+SharpKit\footnote{\href{http://sharpkit.net}{http://sharpkit.net}}, Dart~\cite{Griffith11_Dart},
+Kotlin\footnote{\href{http://kotlin.jetbrains.org/}{http://kotlin.jetbrains.org/}},
 ClojureScript~\cite{McGranaghan11_ClojureScript}, Fay\footnote{\href{http://fay-lang.org/}{http://fay-lang.org/}},
@@
 
-For instance, because the boundaries of the code sent to the client are less visible when you share code between
-client-side and server-side, transitive dependencies may pull a lot of code on the client, causing a high download
-overhead. Moreover, generating efficient code for heterogeneous platforms is hard to achieve in an extensible way:
-the translation of common abstractions like collections into their native counterpart (JavaScript arrays on
-client-side and standard library's collections on server-side) may be hard coded in the compiler, but that would not
-scale to handle all the abstractions a complete application may use (\eg HTML fragment definitions, form validation
-rules, or even some business data type that may be represented differently for performance reasons).
+Generating efficient code for heterogeneous platforms is hard to achieve in an extensible way: the translation of
+common abstractions like collections into their native counterpart (JavaScript arrays on client-side and standard
+library's collections on server-side) may be hard coded in the compiler, but that would not scale to handle all the
+abstractions a complete application may use (\eg HTML fragment definitions, form validation rules, or even some
+business data type that may be represented differently for performance reasons).
 
-On one hand, for engineering reasons, developers want to write Web applications using a single language, abstracting
-over the target platforms differences. But on the other hand, for performance reasons, they want to keep
+On one hand, for engineering reasons, developers want to write Web applications using a single high-level language,
+abstracting over the target platforms differences. But on the other hand, for performance reasons, they want to keep
 control on the way their code is compiled to each target platform. How to solve this dilemma?
@@
 
-\texttt{js-scala}~\cite{Kossakowski12_JsDESL} is such a compiled embedded DSL defined in Scala that generates
-JavaScript code, making it possible to write the client-side code of Web applications using JavaScript. This paper
-enriches js-scala in order to solve the problems described above. We use staging (explain) to generate efficient code
-for typical abstractions used for Web programming and to generate efficient specialized code for both client and
+\texttt{js-scala} is such a compiled embedded DSL defined in Scala that generates JavaScript code, making it possible
+to write the client-side code of Web applications using JavaScript~\cite{Kossakowski12_JsDESL}. This paper
+enriches js-scala in order to solve the problems described above. We use staging (explain?) to generate efficient
+code for typical abstractions used for Web programming and to generate efficient specialized code for both client and
 server sides for shared abstractions. More precisely, we demonstrate the following features:
@@
 code using standard APIs on both server-side and client-side;
- \item Type-directed ad-hoc polymorphism on client-side without runtime dynamic dispatch logic;
- \item Static linking?
+ \item Type-directed ad-hoc polymorphism on client-side without runtime dynamic dispatch logic.
 \end{itemize}
@@
 
-\section{Related Work}
+\section{Related Work (?)}
 
-Developers express programs using a language that is translated into a form executable on a target platform. This
-section presents different approaches for defining cross-platform languages.
+This section presents different approaches for defining cross-platform programming languages.
 
 \begin{figure}
-  \centering
-  \includegraphics[width=8cm]{langs.pdf}
+\begin{center}
+\includegraphics[width=6cm]{langs.pdf}
+\end{center}
 \caption{Language engineering processes}
@@
 
-\subsection{Fat Languages}
+\paragraph{Fat Languages}
 
-One way consists in hard-coding, in the language's compiler, the transformation to each target platform. Figure
-\ref{langs} (left) depicts this process. In order to support a feature related to a specific domain, the whole
-compiler pipeline (parser, code generator, \etc) may have to be adapted. This approach gives \emph{fat} languages
-because a lot of concepts are defined at the language level. Examples of such languages are
-Links~\cite{Cooper07_Links}, Opa and Dart~\cite{Griffith11_Dart}. These languages are difficult to extend because
-each concept is defined in the compiler, and modifying a compiler requires a high effort. Furthermore, these
-languages also require to support common abstraction and composition mechanisms, as general purpose languages do, in
-order to make it easier to write real programs. So they need to be at least general purpose languages. It means that
-for every problem you have to rewrite a full-featured programming language before addressing the concepts specific to
-the problem domain. This approach does not scale.
+The first approach for defining a cross-platform language consists in hard-coding, in the compiler, the
+code generation scheme of each language feature to each target platform. Figure \ref{langs} (a) depicts this
+process. In order to support a feature related to a specific domain, the whole compiler pipeline (parser, code
+generator, \etc) may have to be adapted. This approach gives \emph{fat} languages because a lot of concepts are
+defined at the language level: general programming concepts such as naming, functions, classes, and more specific
+concepts such as HTML fragment definition. Examples of such languages are Links~\cite{Cooper07_Links}, Opa and
+Dart~\cite{Griffith11_Dart}. These languages are difficult to extend because each concept is defined in the compiler,
+and modifying a compiler requires a high effort. Furthermore, these languages also require to support common
+programming abstraction and composition mechanisms, as general purpose languages do. So they usually try to re-invent
+the features of general purpose languages. We argue that this approach for defining programming languages is
+difficult to scale: for every problem you have to rewrite a full-featured programming language before addressing the
+concepts specific to the problem domain.
 
-\subsection{Domain Specific Languages}
+\paragraph{Domain Specific Languages}
 
-Another approach consists in defining several small languages, each one being specific to a given set of concerns,
-and then to combine all the source artifacts written with these language into one executable program. This approach
-makes it difficult to make the separate languages inter-operate with each other.
+Another approach consists in defining several small independent languages, each one focusing on concerns specific to
+a given problem domain, and then to combine all the source artifacts written with these language into one executable
+program (ref), as shown in figure~\ref{langs} (b). Defining such a language requires a minimal effort compared to the
+previous approach. On the other hand, it is difficult to have interoperability between different DSLs (ref).
+(add refs to DSL-based approaches to define Web programming environments)
 
-\subsection{Thin Languages}
+\paragraph{Thin Languages}
 
 Alternatively, one can define concepts relative to a specific domain as libraries on top of a thin general purpose
-language. Figure \ref{langs} (middle) depicts this approach. The general purpose language is used as a host language
-that does not need to be modified if a new concept is introduced, because concepts are defined as libraries. However,
-this approach gives no opportunity to translate a concept efficiently, according to the host platform
-characteristics.
+language (this is also referred to as \emph{embedded} domain specific languages). Figure \ref{langs} (c) depicts this
+approach. The general purpose language is used as a host language and does not need to be modified if a new concept
+is introduced, because concepts are defined as pure libraries. However, this approach gives no opportunity to
+translate a concept efficiently according to the host platform characteristics. Examples of languages following this
+approach are Java/GWT, Kotlin, HaXe and SharpKit.
 
-\subsection{Deeply Embedded Languages}
+\paragraph{Deeply Embedded Languages}
 
-Figure \ref{langs} (right) shows a last approach that tries to get the benefits of the previous two approaches,
-without their shortcomings.
+The last approach, shown in figure \ref{langs} (d), can be seen as a middle-ground between the two previous
+approaches: DSLs are embedded in a host language but use a code generation process. (benefits and limitations of
+this approach)
 
@@
 
-\section{High-Level Abstractions Generating Efficient (and Heterogeneous) Code}
+\section{High-Level Abstractions For Client-Side Code}
 \label{contribution}
@@
 
-In a Web application, the user interface is defined by a HTML document which can be manipulated by the JavaScript
-code to be updated. One typical operation consists in searching some “interesting” element in the document, in order
-to extract its content, replace it or listen to user events fired on this element. The standard API provides several
-function to search elements in a Web document according to their characteristics such as their attribute values.
-Figure~\ref{selectors-api} (wrong reference number) summarizes the available functions and their differences.
+In a Web application, the user interface is defined by a HTML document that can be updated by the JavaScript code.
+A typical operation consists in searching some “interesting” element in the document, in order to extract its
+content, replace it or listen to user events triggered on it. The standard API provides several functions to search
+elements in a HTML document according to their name or attribute values. Figure~\ref{selectors-api} summarizes the
+available functions and their differences.
 
 \begin{figure}
-\label{selectors-api}
+\begin{center}
 \begin{tabular}{| l | p{3cm} |}
@@
 \hline
-\code{querySelectorAll(s)} & All nodes matching the CSS selector \code{s} \\
+\code{querySelectorAll(s)} & All elements matching the CSS selector \code{s} \\
 \hline
@@
 \hline
-\code{getElementsByClassName(c)} & All elements which \code{class} attribute equals to \code{c} \\
+\code{getElementsByClassName(c)} & All elements which \code{class} attribute contains \code{c} \\
 \hline
 \end{tabular}
+\end{center}
 \caption{Standard selectors API}
+\label{selectors-api}
 \end{figure}
@@
 cases. For the developer it is not convenient to have to master several functions performing similar tasks. In fact,
-most JavaScript developers use the jQuery library (ref and stats) that provides just one function to search for
-elements. Listing~\ref{jquery-vanilla-api} shows two equivalent snippets of code, the first one using the native
-APIs (a) and the second one using jQuery (b).
+most JavaScript developers use the jQuery
+library~\cite{Bibeault08_jQuery}\footnote{According to the following document:
+\href{http://trends.builtwith.com/javascript}{ http://trends.builtwith.com/javascript}, jQuery is used by more than
+40\% of the top million sites} that provides only one function to search for elements.
+Listings \ref{vanilla-selectors} and \ref{jquery-selectors} show two equivalent JavaScript programs, the first one
+using the native APIs and the second one using jQuery.
 
 \begin{figure}
-\begin{lstlisting}[language=JavaScript]
+\begin{lstlisting}[language=JavaScript,label=vanilla-selectors,caption=Selectors in plain JavaScript]
 function getWords() {
@@
 \end{lstlisting}
-\caption{(a)}
-\begin{lstlisting}
+\end{figure}
+
+\begin{figure}
+\begin{lstlisting}[language=JavaScript,label=jquery-selectors,caption=Selectors in jQuery]
 function getWords() {
@@
 \end{lstlisting}
-\caption{(b)}
-\label{jquery-vanilla-api}
 \end{figure}
@@
 
-Using LMS we are able to provide a simple API (like jQuery) and to generate code using specialized functions (and
-hence benefit from their performances), by analyzing the staged program. Listing~\ref{js-scala-selectors} shows how
-to implement listing~\ref{jquery-vanilla-api} using js-scala. This listing generates a JavaScript program similar to
-listing~\ref{jquery-vanilla-api} (a).
-
 \begin{figure}
-\label{js-scala-selectors}
-\begin{lstlisting}
+\begin{lstlisting}[label=js-scala-selectors,caption=Selectors in js-scala]
 def getWords() = {
@@
 \end{lstlisting}
-\caption{Selectors in Js-Scala}
 \end{figure}
 
-Writing the program with js-scala instead of using a JavaScript library such as jQuery is interesting because the
-abstraction for searching elements exist only in the initial program source code, not in the final JavaScript
-program. It gives two advantages: (1) the execution of the final program is more performant because of the use of
-specialized APIs, (2) the final program’s size is smaller because it does not need to include jQuery.
+Listing~\ref{js-scala-selectors} shows how to implement listing~\ref{jquery-selectors} using js-scala. We provide two
+functions for searching elements: \code{find} to find the first element matching a selector and \code{findAll} to
+find all the matching elements. During staging these functions analyze the selector that is passed as parameter and,
+when possible, produce code using the specialized API, otherwise they produce code using \code{querySelector} and
+\code{querySelectorAll}. Thanks to this optimizations, listing~\ref{js-scala-selectors} generates a JavaScript
+program identical to listing~\ref{vanilla-selectors}. Writing the program with js-scala instead of using a JavaScript
+library such as jQuery is interesting because the abstraction for searching elements exist only in the initial
+program source code, not in the final JavaScript program. It gives two advantages: (1) the execution of the final
+program is more performant because of the use of specialized APIs, (2) the final program’s size is smaller because it
+does not need to include jQuery.
 
@@
 Null references are a known source of problems in programming languages~\cite{Hoare09_Null,Nanda09_Null}. For
-example, consider the following typical JavaScript code finding a particular widget in the page and then a particular
-button in the widget:
+example, consider listing \ref{null-unsafe} finding a particular widget in the page and then a particular
+button in the widget. The native \code{querySelector} method returns \code{null} if no node matched the given
+selector in the document. If we run this code in a page where the widget is not present, it will throw an error
+and stop further JavaScript execution. Listing \ref{null-defensive} shows defensive code to handle \code{null}
+cases, but it leads to very cumbersome code.
 
+\begin{figure}
 \begin{lstlisting}[language=JavaScript,label=null-unsafe,caption=Unsafe code]
@@
 \end{lstlisting}
+\end{figure}
 
-The native \code{querySelector} method returns \code{null} if no node matched the given selector in the document. If
-we run the above code in a page where the widget is not present, it will throw an error and stop further JavaScript
-execution. We can write defensive code to handle \code{null} cases, but it leads to very cumbersome code:
-
+\begin{figure}
 \begin{lstlisting}[language=JavaScript,label=null-defensive,caption=Defensive programming to handle null references]
@@
 \end{lstlisting}
+\end{figure}
 
@@
 when dereferenced. The wrapping container exists only at the stage-level and is removed during the code generation.
-Here is a Scala listing that uses our DSL (implementation details are given in section \ref{implementation}):
+Listing \ref{null-js-scala} shows how to write the same program using our DSL (implementation details are
+given in section \ref{implementation}). The evaluation of this listing produces a graph of statements from which
+JavaScript code equivalent to listing \ref{null-defensive} is generated.
 
-\begin{lstlisting}
+\begin{figure}
+\begin{lstlisting}[label=null-js-scala,caption=Handling null references in js-scala]
 for {
@@
 \end{lstlisting}
+\end{figure}
 
-The evaluation of the above listing produces a graph of statements from which JavaScript code equivalent to
-listing \ref{null-defensive} is generated.
-
-\subsection{DOM Fragments}
+\section{High-Level Abstractions Shared on Clients and Servers}
 
 \begin{figure}
-\label{dom-api}
-\begin{lstlisting}[language=JavaScript]
+\begin{lstlisting}[language=JavaScript,caption=JavaScript DOM API,label=dom-api]
 var articleUi = function (article) {
@@
 \end{lstlisting}
+\end{figure}
 
-\begin{lstlisting}[language=HTML]
+\begin{figure}
+\begin{lstlisting}[language=HTML,caption=HTML,label=html]
 <div class=article>
@@
 \end{lstlisting}
-\caption{DOM}
 \end{figure}
@@
 
-\subsection{Ad-Hoc Polymorphism}
-
-Because of the dynamically typed nature of JavaScript, when calling a function there is no proper way to select a
-specialized implementation according to the function’s parameters types. JavaScript is only able to dispatch
-according to a method receiver prototype, \eg{} if one writes \code{foo.bar()} the JavaScript runtime will look into
-the prototype of the \code{foo} object for a property named \code{bar} and will call it. So, the only way to achieve
-\emph{ad hoc} polymorphism on JavaScript objects consists in defining the polymorphic function on the prototypes of
-the objects. However, modifying existing object prototypes is considered bad
-practice~\cite{Zakas12_MaintainableJs}. Another way could consist in manually coding the dispatch logic, by
-registering supported data types at the beginning of the program execution, as described in section 2.4.3
-of~\cite{Abelson83_SICP}, but this solution is painful for developers and incurs a performance overhead.
-
-We propose to achieve \emph{ad hoc} polymorphism using
-typeclasses~\cite{Wadler89_AdhocPolymorphism,Odersky06_Typeclasses,Oliveira10_Typeclasses} so that it supports
-retroactive extension without modifying objects prototypes. The dispatch logic is type-directed and performed by the
-compiler, so there is no runtime overhead.
-
-\begin{figure}
-\begin{lstlisting}[label=polymorphism,caption=Ad hoc polymorphism using typeclasses]
-// Interface
-case class Show[A](show: Rep[A => Node])
-
-// Polymorphic function
-def listWidget[A](items: Rep[List[A]])
-      (implicit A: Show[A]): Rep[Node] =
-  el("ul")(
-    for (item <- items) yield {
-      el("li")(A.show(item))
-    }
-  )
-
-// Type `User`
-type User = Record {
-  val name: String
-  val age: Int
-}
-// Implementation of Show for a User
-implicit val showUser = Show[User] { user =>
-  el("span", "class"->"user")(
-    user.name + "(" + user.age + " years)"
-  )
-}
-
-// Main program
-def main(users: Rep[List[User]]) = {
-  document.body.append(listWidget(users))
-}
-\end{lstlisting}
-\end{figure}
-
-Listing \ref{polymorphism} demonstrates how to define a polymorphic \code{listWidget} function that returns a DOM
-tree containing the representation of a list of items. The \code{Show[A]} typeclass defines how to produce a DOM tree
-for a value of type \code{A}. It is used by the \code{listWidget} function to get the DOM fragments of the list
-items. The listing shows how to reuse the same \code{listWidget} function to show a list of users and a list of
-articles.
+% \subsection{Ad-Hoc Polymorphism}
+% 
+% Because of the dynamically typed nature of JavaScript, when calling a function there is no proper way to select a
+% specialized implementation according to the function’s parameters types. JavaScript is only able to dispatch
+% according to a method receiver prototype, \eg{} if one writes \code{foo.bar()} the JavaScript runtime will look into
+% the prototype of the \code{foo} object for a property named \code{bar} and will call it. So, the only way to achieve
+% \emph{ad hoc} polymorphism on JavaScript objects consists in defining the polymorphic function on the prototypes of
+% the objects. However, modifying existing object prototypes is considered bad
+% practice~\cite{Zakas12_MaintainableJs}. Another way could consist in manually coding the dispatch logic, by
+% registering supported data types at the beginning of the program execution, as described in section 2.4.3
+% of~\cite{Abelson83_SICP}, but this solution is painful for developers and incurs a performance overhead.
+% 
+% We propose to achieve \emph{ad hoc} polymorphism using
+% typeclasses~\cite{Wadler89_AdhocPolymorphism,Odersky06_Typeclasses,Oliveira10_Typeclasses} so that it supports
+% retroactive extension without modifying objects prototypes. The dispatch logic is type-directed and performed by the
+% compiler, so there is no runtime overhead.
+% 
+% \begin{figure}
+% \begin{lstlisting}[label=polymorphism,caption=Ad hoc polymorphism using typeclasses]
+% // Interface
+% case class Show[A](show: Rep[A => Node])
+% 
+% // Polymorphic function
+% def listWidget[A](items: Rep[List[A]])
+%       (implicit A: Show[A]): Rep[Node] =
+%   el("ul")(
+%     for (item <- items) yield {
+%       el("li")(A.show(item))
+%     }
+%   )
+% 
+% // Type `User`
+% type User = Record {
+%   val name: String
+%   val age: Int
+% }
+% // Implementation of Show for a User
+% implicit val showUser = Show[User] { user =>
+%   el("span", "class"->"user")(
+%     user.name + "(" + user.age + " years)"
+%   )
+% }
+% 
+% // Main program
+% def main(users: Rep[List[User]]) = {
+%   document.body.append(listWidget(users))
+% }
+% \end{lstlisting}
+% \end{figure}
+% 
+% Listing \ref{polymorphism} demonstrates how to define a polymorphic \code{listWidget} function that returns a DOM
+% tree containing the representation of a list of items. The \code{Show[A]} typeclass defines how to produce a DOM tree
+% for a value of type \code{A}. It is used by the \code{listWidget} function to get the DOM fragments of the list
+% items. The listing shows how to reuse the same \code{listWidget} function to show a list of users and a list of
+% articles.
 
@@
 
-See http://github.com/js-scala/js-scala.
+\footnote{The code is available at \href{http://github.com/js-scala}{http://github.com/js-scala}}
 
@@
 \begin{figure}
-\label{selector-impl}
-\begin{lstlisting}
-def find(receiver: Exp[SelectorApi],
+\begin{lstlisting}[label=selector-impl,caption=Selectors optimization]
+def find(receiver: Exp[Selector],
          selector: Exp[String]) =
@@
 \end{lstlisting}
-\caption{Selectors optimization}
 \end{figure}
+
+Listing \ref{selector-impl}
 
@@
 trait JSGenOptionOps extends JSGenEffect {
-  val IR: EfffectExp with OptionOpsExp
+  val IR: OptionOpsExp
   import IR._
@@
       case OptionIsEmpty(o) =>
-        emitValDef(sym, quote(o) + " === null")
+        emitValDef(sym, q" $o === null")
       case OptionForeach(o, a, b) =>
-        stream.println("if (" + quote(o) + " !== null) {")
+        stream.println(q"if ($o !== null) {")
         emitValDef(a, quote(o))
@@
 
-\subsection{Ad-Hoc Polymorphism}
-
 \section{Evaluation}
@@
 application for making polls. It allows users to create a poll, define the choice alternatives, share the poll, vote
-and look at the results. It contains some JavaScript code to handle the dynamic behavior of the application:
+and look at the results. It contains JavaScript code to handle the dynamic behavior of the application:
 double-posting prevention, dynamic form update and rich interaction with the document.
@@
 client-side code execution. The tests were run on a DELL Latitude E6430 laptop with 8 GB of RAM, on the Google
-Chrome v27 Web browser.
+Chrome v27 Web browser. Figure \ref{benchmark} shows the benchmark results.
 
 \begin{figure}
-\label{benchmark}
 \centering
@@
 \caption{Benchmarks on a real application}
+\label{benchmark}
 \end{figure}
 
-\section{Conclusion, Future Work}
+\section{Conclusion}
 \label{discussion}
+
+We implemented a high-level language abstracting over client and server heterogeneity but producing efficient code.
 

### 2013-05-31 papers/gpce2013/gpce2013.tex
abstract + intro

@@
 This article presents high-level abstractions defined as deep embedded DSLs in Scala, that can (1) generate efficient
-code leveraging the target platform characteristics, (2) be shared between client and server code. Our DSL gives a
+code leveraging the target platform characteristics, (2) be shared between client and server code. Our DSLs have a
 performance / expressiveness ratio about two times higher than other approaches.
@@
 
-\category{CR-number}{subcategory}{third-level}
+\category{D.3.3}{Programming Languages}{Language Constructs and Features}
 
-\terms
-term1, term2
+\terms Languages, Software Engineering
 
-\keywords
-keyword1, keyword2
+\keywords Heterogeneous Code Generation, Domain-specific languages
 
@@
 The more interactive the application is, the more logic needs to be duplicated between the server-side
-and the client-side (explain why?), and the higher is the amount of client-side code. Developers use libraries and
-frameworks to get high-level abstractions on client-side, making their code easier to reason about and to
-maintain, but also making their code run less efficiently (abstraction penalty).
+and the client-side, and the higher is the amount of client-side code. Developers use libraries and frameworks to get
+high-level abstractions on client-side, making their code easier to reason about and to maintain, but also making
+their code run less efficiently (abstraction penalty).
 
@@
 to write the client-side code of Web applications using JavaScript~\cite{Kossakowski12_JsDESL}. This paper
-enriches js-scala in order to solve the problems described above. We use staging (explain?) to generate efficient
-code for typical abstractions used for Web programming and to generate efficient specialized code for both client and
-server sides for shared abstractions. More precisely, we demonstrate the following features:
+enriches \texttt{js-scala} in order to solve the problem described above. We use staging~\cite{Jorring1986_Staging}
+to:
 
 \begin{itemize}
- \item An API for searching in the DOM, that exposes a single entry point but that generates code potentially using
-more optimized native APIs;
- \item Usage of monads without extra container object creation;
- \item Ability to define DOM fragments using a common language for server-side and client-side, but that generates
-code using standard APIs on both server-side and client-side;
- \item Type-directed ad-hoc polymorphism on client-side without runtime dynamic dispatch logic.
+ \item generate efficient code for typical abstractions used in Web programming;
+ \item generate specialized code for both client and server sides for shared abstractions.
 \end{itemize}
 
-Compared to libraries offering the same level of abstraction, we observe that js-scala produces significantly smaller
-executables with better runtime performances.
+Though the code written in \texttt{js-scala} is high-level and can be shared between clients and
+servers, it has the same runtime performances as hand-tune low-level JavaScript code. We observed a performance /
+expressiveness ratio two times higher than using other approaches.
 
 The remainder of this paper is organized as follows. The next section introduces existing approaches for defining
-cross-compiling languages. Section \ref{contribution} presents our contribution. Section \ref{validation} evaluates
-our contribution. Section \ref{discussion} concludes.
+cross-compiling languages. Sections \ref{contribution} and \ref{contribution-shared} present our contribution.
+Section \ref{implementation} gives implementation details. Section \ref{validation} evaluates the contribution.
+Section \ref{discussion} concludes.
 
-\section{Related Work (?)}
+\section{Background}
 
@@
 is introduced, because concepts are defined as pure libraries. However, this approach gives no opportunity to
-translate a concept efficiently according to the host platform characteristics. Examples of languages following this
-approach are Java/GWT, Kotlin, HaXe and SharpKit.
+translate a concept efficiently according to the target platform characteristics. Examples of languages following
+this approach are Java/GWT, Kotlin, HaXe and SharpKit.
 
@@
 
-\section{High-Level Abstractions For Client-Side Code}
+\section{High-Level Abstractions for Client-Side Code}
 \label{contribution}
@@
 \section{High-Level Abstractions Shared on Clients and Servers}
+\label{contribution-shared}
 

### 2013-05-31 papers/sle2013/sle2013.tex
sle abstract

@@
 
+\include{macros}
+
 \newcommand{\ie}{\emph{i.e.}}
@@
 
-\title{FFI Is Not Enough. Need Dependent Types}
+\title{FFI Is Not Enough. We Need Dependent Types}
 \titlerunning{Dependent Types Rock}
@@
 \begin{abstract}
-The abstract should summarize the contents of the paper and should
-contain at least 70 and at most 150 words. It should be written using the
-\emph{abstract} environment.
-\keywords{We would like to encourage you to list your keywords within
-the abstract section}
+JavaScript sucks. It is a known fact. Several other programming languages can target JavaScript as a back-end,
+alleviating developers from the burden of writing and maintaining JavaScript code. The Web browser APIs,
+which are needed to interact with a Web page, are designed for JavaScript, making it challenging to expose them in a
+statically typed language. Indeed, existing statically typed languages either loose control or loose type safety.
+How to give users the same level of control as if they were using the native Web APIs, but in a statically typed and
+convenient way? This article shows how dependent type systems can help.
+\keywords{Dependent Types}
 \end{abstract}
@@
 
-One solution consists in considering JavaScript as an assembly language and generating JavaScript from compilers
-of full-featured and cutting-edge programming languages. Incidentally, an increasing number of programming languages
-or compiler backends can generate JavaScript code (\eg Java/GWT~\cite{Chaganti07_GWT},
+One solution consists in considering JavaScript as an assembly language\footnote{\cf
+\href{http://asmjs.org/}{http://asmjs.org/}} and generating JavaScript from compilers of full-featured and
+cutting-edge programming languages. Incidentally, an increasing number of programming languages or compiler backends
+can generate JavaScript code (\eg Java/GWT~\cite{Chaganti07_GWT},
 SharpKit\footnote{\href{http://sharpkit.net}{http://sharpkit.net}}, Dart~\cite{Griffith11_Dart},
@@
 not statically typed and make a heavy use of overloading, making them hard to expose in a statically typed language.
-Indeed, most of the existing statically typed languages compiling to JavaScript either lose control or type safety
-when they expose Web browser’s APIs. How to give developers the same level of control as if they were using the
-native Web APIs, but in a statically typed and convenient way?
+
+Indeed, existing statically typed languages compiling to JavaScript often expose weaker types than they should. For
+instance, the function \code{createElement} is polymorphic in its return type: it can return a \code{DivElement} as
+well as an \code{InputElement}, among others, but the Dart, Fay, SharpKit and Kotlin APIs return the super-type of
+all the possible values, namely the \code{Element} type. As a consequence, developers need to explicitly down-cast
+the value they get, which is a tedious and error prone task.
+
+Some other languages try to workaround this problem by using overloading instead of polymorphism. For instance, HaXe
+provides functions \code{createDivElement}, \code{createInputElement}, which return a \code{DivElement} and an
+\code{InputElement}, respectively. Besides requiring a higher effort to implement, this solution also reduces the
+control level of users: by being statically resolved, the element type can not anymore be passed as a parameter.
+
+It turns out that most of the existing statically typed languages compiling to JavaScript either loose control or
+loose type safety when they expose Web browser’s APIs. How to give developers the same level of control as if they
+were using the native Web APIs, but in a statically typed and convenient way?
 
 In this paper we present several ways to integrate Web browser’s APIs as statically typed APIs that are safe and
-give developers the same control level as if they were using the native APIs. We could achieve this
-by using advanced features of type systems like dependent types and functional dependencies.
+give developers the same control level as if they were using the native APIs. We can achieve this by using advanced
+features of type systems like dependent types and functional dependencies.
 
@@
 
-Other languages either provide lose information about the data type of the listened event (Dart) or give no way to
+Other languages either provide loose information about the data type of the listened event (Dart) or give no way to
 abstract over an event (GWT, Kotlin).

### 2013-06-03 papers/gpce2013/gpce2013.tex
dull

@@
 the features of general purpose languages. We argue that this approach for defining programming languages is
-difficult to scale: for every problem you have to rewrite a full-featured programming language before addressing the
+difficult to scale: for every problem you have to rewrite a full-featured programming language besides addressing the
 concepts specific to the problem domain.
 
-\paragraph{Domain Specific Languages}
+\paragraph{Domain-Specific Languages}
 
-Another approach consists in defining several small independent languages, each one focusing on concerns specific to
-a given problem domain, and then to combine all the source artifacts written with these language into one executable
-program (ref), as shown in figure~\ref{langs} (b). Defining such a language requires a minimal effort compared to the
-previous approach. On the other hand, it is difficult to have interoperability between different DSLs (ref).
-(add refs to DSL-based approaches to define Web programming environments)
+Another approach consists in defining several independent domain-specific languages~\cite{Van00_DSL}, each one
+focusing on concerns specific to a given problem domain, and then to combine all the source artifacts written with
+these language into one executable program, as shown in figure~\ref{langs} (b). Defining such a language requires a
+minimal effort compared to the previous approach. On the other hand, it is difficult to have interoperability between
+DSLs (ref). \cite{Groenewegen08_WebDSL} shows such a domain-specific lanugage for defining Web applications.
 
@@
 
-As another illustration of the staging mechanism, we present a simple DSL to handle null references. This DSL
-provides an abstraction at the stage-level that is removed by optimization during the code generation.
-
 Null references are a known source of problems in programming languages~\cite{Hoare09_Null,Nanda09_Null}. For
@@
   loginButton <- loginWidget.find("submit.button")
-} loginButton.on(Click) { e => ... }
+} loginButton.on(Click)(handler)
 \end{lstlisting}

### 2013-06-04 src/main/scala/scala/js/exp/OptionOpsExp.scala
Slightly better code generation for options

@@ -6,62 +6,68 @@ import scala.virtualization.lms.common.EffectExp
 trait OptionOpsExp extends OptionOps with EffectExp {
 
   case object none extends Exp[None.type]
-  case class OptionSome[A](a: Rep[A]) extends Def[Option[A]]
-  def some[A : Manifest](a: Rep[A]) = OptionSome(a)
+  case class OptionSome[A](a: Exp[A]) extends Def[Option[A]]
+  def some[A : Manifest](a: Exp[A]) = OptionSome(a)
 
-  implicit class OptionOpsCls[+A : Manifest](o: Rep[Option[A]]) extends OptionOpsBase[A] {
-    def foreach(f: Exp[A] => Exp[Unit]) = {
-      val a = fresh[A]
+  implicit class OptionOpsCls[+A : Manifest](o: Exp[Option[A]]) extends OptionOpsBase[A] {
+
+    private def sym = o match {
+      case Sym(n) => (Sym[A](n), false)
+      case _ => (fresh[A], true)
+    }
+
+    def foreach(f: Exp[A] => Exp[Unit]) = { // TODO Optimize constants
+      val (a, bound) = sym
       val block = reifyEffects(f(a))
-      reflectEffect(OptionForeach(o, a, block), summarizeEffects(block).star)
+      reflectEffect(OptionForeach(o, a, block, bound), summarizeEffects(block).star)
     }
     def map[B : Manifest](f: Exp[A] => Exp[B]) = {
-      val a = fresh[A]
+      val (a, bound) = sym
       val block = reifyEffects(f(a))
-      reflectEffect(OptionMap(o, a, block), summarizeEffects(block).star)
+      reflectEffect(OptionMap(o, a, block, bound), summarizeEffects(block).star)
     }
     def flatMap[B : Manifest](f: Exp[A] => Exp[Option[B]]) = {
-      val a = fresh[A]
+      val (a, bound) = sym
       val block = reifyEffects(f(a))
-      reflectEffect(OptionFlatMap(o, a, block), summarizeEffects(block).star)
+      reflectEffect(OptionFlatMap(o, a, block, bound), summarizeEffects(block).star)
     }
     def isEmpty = OptionIsEmpty(o)
-    def fold[B : Manifest](none: => Rep[B], some: Rep[A] => Rep[B]) = {
-      val a = fresh[A]
[…]

### 2013-06-04 src/main/scala/scala/js/gen/js/GenOptionOps.scala

@@ -1,38 +1,39 @@
 package scala.js.gen.js
 
 import scala.js.exp.OptionOpsExp
+import scala.js.gen.QuoteGen
 
-trait GenOptionOps extends GenEffect {
+trait GenOptionOps extends GenEffect with QuoteGen {
   val IR: OptionOpsExp
   import IR._
 
   override def emitNode(sym: Sym[Any], rhs: Def[Any]) = rhs match {
-    case OptionForeach(o, a, block) =>
+    case OptionForeach(o, a, block, bound) =>
       stream.println("if (" + quote(o) + " !== null) {")
-      emitValDef(a, quote(o)) // Ouin
+      if (bound) emitValDef(a, quote(o))
       emitBlock(block)
       stream.println("}")
       emitValDef(sym, "undefined")
-    case OptionMap(o, a, block) =>
-      emitValDef(a, quote(o))
-      stream.println("if (" + quote(a) + " !== null) {")
+    case OptionMap(o, a, block, bound) =>
+      emitValDef(sym, "null")
+      stream.println(q"if ($o !== null) {")
+      if (bound) emitValDef(a, quote(o))
       emitBlock(block)
-      stream.println(quote(a) + " = " + quote(getBlockResult(block)) + ";")
+      emitAssignment(quote(sym), quote(getBlockResult(block)))
       stream.println("}")
-      emitValDef(sym, quote(a))
-    case OptionFlatMap(o, a, block) =>
-      emitValDef(a, quote(o))
-      stream.println("if (" + quote(a) + " !== null) {")
+    case OptionFlatMap(o, a, block, bound) =>
+      emitValDef(sym, "null")
+      stream.println(q"if ($o !== null) {")
+      if (bound) emitValDef(a, quote(o))
       emitBlock(block)
-      stream.println(quote(a) + " = " + quote(getBlockResult(block)) + ";")
+      emitAssignment(quote(sym), quote(getBlockResult(block)))
       stream.println("}")
-      emitValDef(sym, quote(a))
     case OptionIsEmpty(o) =>
       emitValDef(sym, quote(o) + " === null")
-    case OptionFold(o, a, n, s) =>
+    case OptionFold(o, a, n, s, bound) =>
       emitValDef(sym, "undefined")
-      emitValDef(a, quote(o))
-      stream.println("if (" + quote(a) + " !== null) {")
+      stream.println(q"if ($o !== null) {")
+      if (bound) emitValDef(a, quote(o))
[…]

### 2013-06-05 src/main/scala/scala/js/language/OptionOps.scala
option-getorelse made with fold (#3)

@@ -16,5 +16,6 @@ trait OptionOps extends Base {
     def flatMap[B : Manifest](f: Rep[A] => Rep[Option[B]]): Rep[Option[B]]
     def isEmpty: Rep[Boolean]
     def fold[B : Manifest](none: => Rep[B], some: Rep[A] => Rep[B]): Rep[B]
+    def getOrElse[B >: A : Manifest](default: => Rep[B]): Rep[B]
   }
 }

### 2013-06-05 papers/gpce2013/gpce2013.tex
Better abstract and intro. WIP contrib.

@@
 This article presents high-level abstractions defined as deep embedded DSLs in Scala, that can (1) generate efficient
-code leveraging the target platform characteristics, (2) be shared between client and server code. Our DSLs have a
-performance / expressiveness ratio about two times higher than other approaches.
+code leveraging the target platform characteristics, (2) be shared between client and server code. Code written with
+our DSL has a high level of abstraction but runs as fast as hand tuned low-level JavaScript code.
 \end{abstract}
@@
 
-\keywords Heterogeneous Code Generation, Domain-specific languages
+\keywords Heterogeneous code generation, Domain-specific languages
 
@@
 The more interactive the application is, the more logic needs to be duplicated between the server-side
-and the client-side, and the higher is the amount of client-side code. Developers use libraries and frameworks to get
-high-level abstractions on client-side, making their code easier to reason about and to maintain, but also making
-their code run less efficiently (abstraction penalty).
+and the client-side, and the higher is the complexity of the client-side code. Developers use libraries and
+frameworks to get high-level abstractions on client-side, making their code easier to reason about and to maintain,
+but also making their code run less efficiently (due to \emph{abstraction penalty}).
+
+Performance is a primary concern in Web applications, because they are expected to run on a broad range of devices,
+from the powerful desktop personal computer to the less powerful smart phone. “Every 100~ms delay costs 1\% of
+sales”, said Amazon in 2006.
 
@@
 
-Performance is a primary concern in Web applications, because they are expected to run on a broad range of devices,
-from the powerful desktop personal computer to the less powerful smart phone. “Every 100~ms delay costs 1\% of
-sales”, said Amazon in 2006.
-
 Generating efficient code for heterogeneous platforms is hard to achieve in an extensible way: the translation of
@@
 abstractions a complete application may use (\eg HTML fragment definitions, form validation rules, or even some
-business data type that may be represented differently for performance reasons).
+business data type that may be represented differently).
 
 On one hand, for engineering reasons, developers want to write Web applications using a single high-level language,
-abstracting over the target platforms differences. But on the other hand, for performance reasons, they want to keep
-control on the way their code is compiled to each target platform. How to solve this dilemma?
+abstracting over the target platforms differences and reducing code complexity. But on the other hand, for
+performance reasons, they want to keep control on the way their code is compiled to each target platform. How to
+solve this dilemma?
 
-Compiled domain specific embedded languages~\cite{Elliott2003_Compiling} allow the definition of domain specific
+Compiled domain-specific embedded languages~\cite{Elliott2003_Compiling} allow the definition of domain-specific
 languages (DSLs) as libraries on top of a host language, and to compile them to a target platform. The deep embedding
@@
 
-\texttt{js-scala} is such a compiled embedded DSL defined in Scala that generates JavaScript code, making it possible
-to write the client-side code of Web applications using JavaScript~\cite{Kossakowski12_JsDESL}. This paper
-enriches \texttt{js-scala} in order to solve the problem described above. We use staging~\cite{Jorring1986_Staging}
-to:
+Kossakowski \etal introduced \emph{js-scala}, a compiled embedded DSL defined in Scala that generates JavaScript
+code, making it possible to write the client-side code of Web applications using
+JavaScript~\cite{Kossakowski12_JsDESL}. However, Kossakowski \etal did not address the engineering dilemma described
+above. This paper enriches \emph{js-scala} with the following contributions:
 
 \begin{itemize}
- \item generate efficient code for typical abstractions used in Web programming;
- \item generate specialized code for both client and server sides for shared abstractions.
+ \item we define high-level abstractions typically used in Web programming generating low-level code leveraging
+native Web browsers APIs;
+ \item in the case of abstractions shared between servers and clients, we specialize the code generation in order to
+leverage the target platform native environments.
 \end{itemize}
 
-Though the code written in \texttt{js-scala} is high-level and can be shared between clients and
-servers, it has the same runtime performances as hand-tune low-level JavaScript code. We observed a performance /
-expressiveness ratio two times higher than using other approaches.
+We validate our approach with a case study implemented with various candidate technologies and discuss the relative
+pro and cons of them. Though the code written in our DSL is high-level and can be shared between clients and servers,
+it has the same runtime performances as hand-tune low-level JavaScript code.
 
@@
 minimal effort compared to the previous approach. On the other hand, it is difficult to have interoperability between
-DSLs (ref). \cite{Groenewegen08_WebDSL} shows such a domain-specific lanugage for defining Web applications.
+DSLs (ref). \cite{Groenewegen08_WebDSL} shows such a domain-specific language for defining Web applications.
 
@@
 
-Lightweight Modular Staging~\cite{Rompf12_LMSThesis} is a framework for defining deeply embedded DSLs in Scala. It
-has been used to define high-performance DSLs for parallel computing~\cite{Brown11_Parallel} and can be used to
-generate JavaScript code~\cite{Kossakowski12_JsDESL}.
+Lightweight Modular Staging~\cite{Rompf12_LMSThesis} is a framework for defining deeply embedded DSLs in Scala using
+staging~\cite{Jorring1986_Staging}. It has been used to define high-performance DSLs for parallel
+computing~\cite{Brown11_Parallel}.
 
-\section{High-Level Abstractions for Client-Side Code}
+\section{High-Level Abstraction for Client-Side Code: Selectors}
 \label{contribution}
-
-\subsection{Selectors}
 
@@
 A typical operation consists in searching some “interesting” element in the document, in order to extract its
-content, replace it or listen to user events triggered on it. The standard API provides several functions to search
-elements in a HTML document according to their name or attribute values. Figure~\ref{selectors-api} summarizes the
-available functions and their differences.
+content, replace it or listen to user events triggered on it (such as a mouse click). The standard API provides
+several functions to search elements in a HTML document according to their name or attribute values.
+Figure~\ref{selectors-api} summarizes the available functions and their differences.
 
@@
 
-The \code{querySelector} and \code{querySelectorAll} are the most general functions, the others handle special
+The \code{querySelector} and \code{querySelectorAll} are the most general functions while the others handle special
 cases. For the developer it is not convenient to have to master several functions performing similar tasks. In fact,
@@
 \href{http://trends.builtwith.com/javascript}{ http://trends.builtwith.com/javascript}, jQuery is used by more than
-40\% of the top million sites} that provides only one function to search for elements.
-Listings \ref{vanilla-selectors} and \ref{jquery-selectors} show two equivalent JavaScript programs, the first one
-using the native APIs and the second one using jQuery.
+40\% of the top million sites} that provides only one high-level function to search for elements.
+Listings \ref{vanilla-selectors} and \ref{jquery-selectors} show two equivalent JavaScript programs performing
+element searches, the first one using the native APIs and the second one using jQuery.
 
@@
 
-Listing~\ref{js-scala-selectors} shows how to implement listing~\ref{jquery-selectors} using js-scala. We provide two
-functions for searching elements: \code{find} to find the first element matching a selector and \code{findAll} to
-find all the matching elements. During staging these functions analyze the selector that is passed as parameter and,
-when possible, produce code using the specialized API, otherwise they produce code using \code{querySelector} and
-\code{querySelectorAll}. Thanks to this optimizations, listing~\ref{js-scala-selectors} generates a JavaScript
-program identical to listing~\ref{vanilla-selectors}. Writing the program with js-scala instead of using a JavaScript
-library such as jQuery is interesting because the abstraction for searching elements exist only in the initial
-program source code, not in the final JavaScript program. It gives two advantages: (1) the execution of the final
-program is more performant because of the use of specialized APIs, (2) the final program’s size is smaller because it
-does not need to include jQuery.
+Listing~\ref{js-scala-selectors} shows how to implement listing~\ref{jquery-selectors} using \emph{js-scala}. We
+provide two functions for searching elements: \code{find} to find the first element matching a selector and
+\code{findAll} to find all the matching elements. During staging these functions try to analyze the selector that is
+passed as parameter and, when appropriate, produce code using the specialized API, otherwise they produce code using
+\code{querySelector} and \code{querySelectorAll}. As a result, listing~\ref{js-scala-selectors} generates a
+JavaScript program identical to listing~\ref{vanilla-selectors}: the high-level abstraction (the \code{find} and
+\code{findAll} functions) exist only in the initial program, not in the final JavaScript program. It gives two
+advantages: (1) the execution of the final program is more performant because of the use of specialized APIs, (2) the
+final program’s size is smaller because it does not need to include jQuery. (Move this last sentence in the
+discussion)
+
+\section{High-Level Abstractions Shared on Clients and Servers}
+\label{contribution-shared}
 
@@
 selector in the document. If we run this code in a page where the widget is not present, it will throw an error
-and stop further JavaScript execution. Listing \ref{null-defensive} shows defensive code to handle \code{null}
-cases, but it leads to very cumbersome code.
+and stop further JavaScript execution. Defensive code can handle \code{null} reference, but leads to very cumbersome
+code, as shown in listing \ref{null-defensive}.
 
@@
 
+Some programming languages encode optional values with a monad (\eg \code{Maybe} in Haskell and \code{Option} in
+Scala). In that case, sequencing over the monad encodes optional value dereferencing. If the language supports a
+convenient syntax for monad sequencing, it brings a convenient syntax for optional value dereferencing, alleviating
+developers from the burden of defensive programming.
+
+Listing \ref{null-js-scala} implements in \emph{js-scala} a program equivalent to listing \ref{null-defensive}. The
+\code{for} notation is used to sequence computations over optional values encoded with a monad, allowing a high
+level of expressiveness. (To be continued, mention \code{Rep[Option[A]]}, \code{Rep[Some[A]]}, \code{Rep[None.type]})
+
 We want to define a DSL that has both the safety and performance of listing \ref{null-defensive} but the
@@
 
-\section{High-Level Abstractions Shared on Clients and Servers}
-\label{contribution-shared}
+\subsection{DOM Fragments Definition}
 

### 2013-06-06 papers/gpce2013/gpce2013.tex
Progress on sections 2 and 3. Still no title.

@@
 \begin{itemize}
- \item we define high-level abstractions typically used in Web programming generating low-level code leveraging
+ \item we define high-level abstractions typically used in Web programming, that generate low-level code leveraging
 native Web browsers APIs;
@@
 pro and cons of them. Though the code written in our DSL is high-level and can be shared between clients and servers,
-it has the same runtime performances as hand-tune low-level JavaScript code.
+it has the same runtime performances as hand-tuned low-level JavaScript code.
 
@@
 
-Lightweight Modular Staging~\cite{Rompf12_LMSThesis} is a framework for defining deeply embedded DSLs in Scala using
-staging~\cite{Jorring1986_Staging}. It has been used to define high-performance DSLs for parallel
-computing~\cite{Brown11_Parallel}.
+Lightweight Modular Staging~\cite{Rompf12_LMSThesis} (LMS) is a framework for defining deeply embedded DSLs in Scala
+using staging~\cite{Jorring1986_Staging}. It has been used to define high-performance DSLs for parallel
+computing~\cite{Brown11_Parallel}. (Todo. Mention \code{Rep[_]})
 
@@
 selector in the document. If we run this code in a page where the widget is not present, it will throw an error
-and stop further JavaScript execution. Defensive code can handle \code{null} reference, but leads to very cumbersome
-code, as shown in listing \ref{null-defensive}.
+and stop further JavaScript execution. Defensive code can be written to handle \code{null} references, but leads to
+very cumbersome code, as shown in listing \ref{null-defensive}.
 
@@
 Listing \ref{null-js-scala} implements in \emph{js-scala} a program equivalent to listing \ref{null-defensive}. The
-\code{for} notation is used to sequence computations over optional values encoded with a monad, allowing a high
-level of expressiveness. (To be continued, mention \code{Rep[Option[A]]}, \code{Rep[Some[A]]}, \code{Rep[None.type]})
-
-We want to define a DSL that has both the safety and performance of listing \ref{null-defensive} but the
-expressiveness of listing \ref{null-unsafe}. We can get safety by wrapping potentially null values of type
-\code{Rep[A]} in a container of type \code{Rep[Option[A]]} requiring explicit dereferencing, we can get
-expressiveness by using the Scala \code{for} notation for dereferencing, and finally we can get performance by
-generating code that does not actually wraps values in a container but instead checks if they are \code{null} or not
-when dereferenced. The wrapping container exists only at the stage-level and is removed during the code generation.
-Listing \ref{null-js-scala} shows how to write the same program using our DSL (implementation details are
-given in section \ref{implementation}). The evaluation of this listing produces a graph of statements from which
-JavaScript code equivalent to listing \ref{null-defensive} is generated.
+\code{for} notation is used to sequence computations over optional values encoded with a monad. The \code{find}
+function returns a \code{Rep[Option[Element]]} value, that can either be a \code{Rep[Some[Element]]} (if an element
+was found) or a \code{Rep[None.type]} (if no element was found). The \code{for} expression contains a sequence
+of statements that are executed in order, as long as the previous statement returned a \code{Rep[Some[Element]]}
+value.
 
@@
 
+Such a monadic API brings both safety and expressiveness to developers manipulating optional values but usually
+involves the creation of an extra container object holding the optional value. In our case, we expose
+the monadic API at the first stage but we generate code that does not wrap values in a container object but instead
+checks if they are \code{null} or not when dereferenced. So the extra container object exists only at the
+stage-level and is removed during code generation: listing \ref{null-js-scala} produces a code equivalent to listing
+\ref{null-defensive}.
+
+Last but not least, we leverage the modular architecture of LMS to define two code generators for this API: one
+targeting JavaScript and the other targeting Scala. So the same abstraction is efficiently translated on both server
+and client sides.
+
 \subsection{DOM Fragments Definition}
+
+This section shows how we define an abstraction shared between clients and servers, as in the previous section, but
+that has different native counterparts on client and server sides. The challenge is to define an API providing a
+common vocabulary that generates code using the target platform native APIs.
 
@@
 \begin{figure}
-\begin{lstlisting}[language=HTML,caption=HTML,label=html]
-<div class=article>
-  <span>French wine: </span>
-  <strong>10</strong>
+\begin{lstlisting}[caption=Scala XML API,label=scala-xml-api]
+def articleUi(article: Article) =
+  <div class="article">
+    <span>{ article.name + ": " }</span>
+    <strong>{ article.price }</strong>
   </div>
@@
 
-In this section we show how we can define a template engine as an embedded DSL with minimal effort. This template
-engine is statically typed and able to insert dynamic content in a safe way. It provides a powerful expression
-language, requires no extra compilation step and can be used on both client-side and server-side.
+A common task in Web applications consists in computing HTML fragments representing a part of the page content. This
+task can be performed either from the server-side (to initially respond to a request) or from the client-side (to
+update the current page). As an example, listing \ref{dom-api} defines a JavaScript function \code{articleUi} that
+builds a DOM tree containing an article description, and listing \ref{scala-xml-api} shows how one could implement
+a similar function on server-side using the standard Scala XML library. The reader may notice that the client-side
+and server-side API are very different and that the client-side API is very low-level and inconvenient to use.
 
-Because the template engine is defined as en embedded DSL, we can reuse Scala’s constructs:
-
-\begin{itemize}
-\item a function taking some parameters and returning a DOM fragment directly models a template taking parameters and
-returning a DOM fragment~;
-\item the type system type-checks template definitions and template calls~;
-\item the Scala language itself is the expression language~;
-\item compiling a template is the same as compiling user code.
-\end{itemize}
-
-So the only remaining work consists in defining the DSL vocabulary to define DOM nodes. We provide a \code{el}
-function to define a tag and any \code{String} value is considered to be a text node.
+Instead, we provide a common high-level DSL for defining HTML fragments and we make this DSL generate code leveraging
+native environments. Listing \ref{forest} shows how to implement our example with our DSL. The \code{el} function
+defines an HTML element, eventually containing attributes and children elements. Any children of an element that is
+not an element itself is converted into a text node. The children elements of an element can also be obtained
+dynamically from a collection, as shown in listing \ref{forest-loops}.
 
@@
 
-Listing \ref{forest} uses our DSL and generates a code equivalent to listing \ref{dom-api}. The readability has
-been highly improved: nesting tags is just like nesting code blocks, HTML entities are
-automatically escaped in text nodes, developers have the full computational power of Scala to inject dynamic data and
-DOM fragments definitions are written using functions so they compose just as functions compose. These benefits come
-with no performance loss because the DSL generates code building DOM fragments by using the native JavaScript API.
+\begin{figure}
+\begin{lstlisting}[label=forest-loops,caption=Using loops]
+def articlesUi(articles: Rep[Seq[Article]]) =
+    el('ul)(
+        for (article <- articles)
+        yield el('li)(articleUi(article))
+    )
+\end{lstlisting}
+\end{figure}
 
-Our DSL is equivalent to a template engine with Scala as the expression language. Making it usable on both server and
-client sides was surprisingly as simple as defining another code generator for the DSL, producing Scala code.
-
-For instance, the template written in listing \ref{forest} produces the following Scala code usable on
-server-side:
-
-\begin{lstlisting}
-def articleUi(article: Article) = {
-  val x0 = <span>{ article.name + ": " }</span>
-  val x1 = <strong>{ article.price }</strong>
-  val x2 =
-    <div class="article">
-      {List(x0, x1)}
-    </div>
-  x2
-}
-\end{lstlisting}
-
-We are able to tackle the code sharing issues described in the introduction (more details?) because of the embdedded
-nature of our DSLs: dynamic content of templates is written using embedded DSLs too, so their translation into
-JavaScript and Scala is managed by their respective code generators.
+The \code{el} function returns an \code{Element} IR node that is a tree composed of other \code{Element} nodes and
+\code{Text} nodes. This tree is traversed by the code generators to produce code building an equivalent DOM tree on
+client-side and code building an equivalent XML fragment on server-side. When the children elements of an element are
+constant values (as in listing \ref{forest}) rather than dynamically obtained (as in listing \ref{forest-loops}), the
+code generators inline the loop that adds children to their parent, for better performance. As a result, listing
+\ref{forest} generates a code equivalent to listing \ref{dom-api} on client-side and equivalent to
+\ref{scala-xml-api} on server-side.
 

### 2013-06-06 papers/sle2013/sle2013.tex
Fix sle abstract

@@
 
-\title{FFI Is Not Enough. We Need Dependent Types}
+\title{FFI Is Not Enough. We Need Dependent Types.}
 \titlerunning{Dependent Types Rock}
@@
 \begin{abstract}
-JavaScript sucks. It is a known fact. Several other programming languages can target JavaScript as a back-end,
-alleviating developers from the burden of writing and maintaining JavaScript code. The Web browser APIs,
-which are needed to interact with a Web page, are designed for JavaScript, making it challenging to expose them in a
-statically typed language. Indeed, existing statically typed languages either loose control or loose type safety.
-How to give users the same level of control as if they were using the native Web APIs, but in a statically typed and
-convenient way? This article shows how dependent type systems can help.
+Several programming languages can target JavaScript as a back-end, giving developers programming language features
+that are absent from JavaScript, such as static typing. However, the Web browser APIs, which are needed to interact
+with a Web page, are designed for JavaScript, making it challenging to expose them in a statically typed language.
+Indeed, existing statically typed languages exposing Web browser APIs either break type safety or give developers
+less control than if they were using JavaScript. This article shows how to expose Web browser APIs in Scala in a
+type safe way while keeping the same level of control as with native APIs. We use advanced type system features such
+as dependent types and functional dependencies.
 \keywords{Dependent Types}

### 2013-06-07 papers/gpce2013/gpce2013.tex
eod

@@
 This article presents high-level abstractions defined as deep embedded DSLs in Scala, that can (1) generate efficient
-code leveraging the target platform characteristics, (2) be shared between client and server code. Code written with
+code leveraging the target platform characteristics, (2) be shared between clients and servers. Code written with
 our DSL has a high level of abstraction but runs as fast as hand tuned low-level JavaScript code.
@@
 JavaScript~\cite{Kossakowski12_JsDESL}. However, Kossakowski \etal did not address the engineering dilemma described
-above. This paper enriches \emph{js-scala} with the following contributions:
+above. This paper enriches \emph{js-scala} with the following contributions\footnote{The code is available at
+\href{http://github.com/js-scala}{http://github.com/js-scala}}:
 
 \begin{itemize}
- \item we define high-level abstractions typically used in Web programming, that generate low-level code leveraging
+ \item we define high-level abstractions, typically used in Web programming, that generate low-level code leveraging
 native Web browsers APIs;
@@
 cross-compiling languages. Sections \ref{contribution} and \ref{contribution-shared} present our contribution.
-Section \ref{implementation} gives implementation details. Section \ref{validation} evaluates the contribution.
-Section \ref{discussion} concludes.
+Section \ref{validation} compares our solution to common approaches. Section \ref{discussion} concludes.
 
@@
 Alternatively, one can define concepts relative to a specific domain as libraries on top of a thin general purpose
-language (this is also referred to as \emph{embedded} domain specific languages). Figure \ref{langs} (c) depicts this
+language (this is also referred to as \emph{embedded} domain-specific languages). Figure \ref{langs} (c) depicts this
 approach. The general purpose language is used as a host language and does not need to be modified if a new concept
@@
 
+Code generation consists in traversing the statement nodes produced by the program evaluation according to their
+dependencies and to emit the code corresponding to each statement. LMS already sorts the statements graph so DSL
+authors just need to say how to emit code for each statement node of their DSL.
+
 \section{High-Level Abstraction for Client-Side Code: Selectors}
@@
 A typical operation consists in searching some “interesting” element in the document, in order to extract its
-content, replace it or listen to user events triggered on it (such as a mouse click). The standard API provides
+content, replace it or listen to user events triggered on it (such as mouse clicks). The standard API provides
 several functions to search elements in a HTML document according to their name or attribute values.
@@
 most JavaScript developers use the jQuery
-library~\cite{Bibeault08_jQuery}\footnote{According to the following document:
-\href{http://trends.builtwith.com/javascript}{ http://trends.builtwith.com/javascript}, jQuery is used by more than
+library~\cite{Bibeault08_jQuery}\footnote{According to \href{http://trends.builtwith.com/javascript}{
+http://trends.builtwith.com/javascript}, jQuery is used by more than
 40\% of the top million sites} that provides only one high-level function to search for elements.
@@
 \begin{figure}
-\begin{lstlisting}[language=JavaScript,label=vanilla-selectors,caption=Selectors in plain JavaScript]
+\begin{lstlisting}[language=JavaScript,label=vanilla-selectors,caption=Searching elements in plain JavaScript]
 function getWords() {
@@
 \begin{figure}
-\begin{lstlisting}[language=JavaScript,label=jquery-selectors,caption=Selectors in jQuery]
+\begin{lstlisting}[language=JavaScript,label=jquery-selectors,caption=Searching elements in jQuery]
 function getWords() {
@@
 \begin{figure}
-\begin{lstlisting}[label=js-scala-selectors,caption=Selectors in js-scala]
+\begin{lstlisting}[label=js-scala-selectors,caption=Searching elements in js-scala]
 def getWords() = {
@@
 advantages: (1) the execution of the final program is more performant because of the use of specialized APIs, (2) the
-final program’s size is smaller because it does not need to include jQuery. (Move this last sentence in the
+final program’s size is smaller because it does not need to include jQuery. (Todo. Move this last sentence in the
 discussion)
+
+\begin{figure}
+\begin{lstlisting}[label=selector-impl,caption=Selectors optimization]
+def find(receiver: Exp[Selector],
+         selector: Exp[String]) =
+  getConstIdCss(selector) match {
+    case Some(id) if receiver == document =>
+      DocumentGetElementById(Const(id))
+    case _ =>
+      SelectorFind(receiver, selector)
+  }
+\end{lstlisting}
+\end{figure}
+
+Listing \ref{selector-impl} shows the implementation of the \code{find} function producing different IR node
+according to the selector passed as parameter. The \code{getConstIdCss} function analyzes the selector: if it
+is a constant \code{String} value containing a CSS ID selector, it returns the value of the identifier. If the
+\code{find} function is applied to the \code{document} and to an ID selector, it returns a
+\code{DocumentGetElementById} IR node (that is translated to a \code{document.getElementById} call by the code
+generator), otherwise it returns a \code{SelectorFind} IR node (that is translated to a \code{querySelector}
+call). (FIXME Add a ref to the spec of CSS selectors?)
+
+The same applies to the implementation of \code{findAll}: the selector passed as parameter is analyzed and the
+function returns a \code{SelectorGetElementsByClassName} in case of a CSS class name selector, a
+\code{SelectorGetElementsByTagName} in case of a CSS tag name selector, and a \code{SelectorFindAll} otherwise.
 
@@
 Such a monadic API brings both safety and expressiveness to developers manipulating optional values but usually
-involves the creation of an extra container object holding the optional value. In our case, we expose
-the monadic API at the first stage but we generate code that does not wrap values in a container object but instead
-checks if they are \code{null} or not when dereferenced. So the extra container object exists only at the
-stage-level and is removed during code generation: listing \ref{null-js-scala} produces a code equivalent to listing
-\ref{null-defensive}.
+involves the creation of an extra container object holding the optional value. In our case, we expose the monadic API
+at the first stage but we generate code that does not wrap values in container objects but instead checks if they are
+\code{null} or not when dereferenced. So the extra container object exists only at the stage-level and is removed
+during code generation: listing \ref{null-js-scala} produces a code equivalent to listing \ref{null-defensive}.
+
+\begin{figure}
+\begin{lstlisting}[caption=JavaScript code generator for null references handling DSL,label=option-codegen]
+override def emitNode(sym: Sym[Any], rhs: Def[Any]) =
+  rhs match {
+    case OptionIsEmpty(o) =>
+      emitValDef(sym, q" $o === null")
+    case OptionForeach(o, b) =>
+      stream.println(q"if ($o !== null) {")
+      emitBlock(b)
+      stream.println("}")
+    case _ =>
+      super.emitNode(sym, rhs)
+  }
+\end{lstlisting}
+\end{figure}
+
+Listing \ref{option-codegen} shows the JavaScript code generator for methods \code{isEmpty} (that checks if the
+optional value contains a value) and \code{foreach} (that is called when the \code{for} notation is used, as in
+listing \ref{null-js-scala}). The \code{emitNode} method handles \code{OptionIsEmpty} and \code{OptionForeach} nodes
+returned by the implementations of \code{isEmpty} and \code{foreach}, respectively. In the case of the
+\code{OptionIsEmpty} node, it simply generates an expression testing if the value is \code{null}. In the case of the
+\code{OptionForeach} node, it wraps the code block dereferencing the value within a \code{if} checking that the value
+is not \code{null}.
 
@@
 a similar function on server-side using the standard Scala XML library. The reader may notice that the client-side
-and server-side API are very different and that the client-side API is very low-level and inconvenient to use.
+and server-side APIs are very different and that the client-side API is very low-level and inconvenient to use.
 
@@
 \code{Text} nodes. This tree is traversed by the code generators to produce code building an equivalent DOM tree on
-client-side and code building an equivalent XML fragment on server-side. When the children elements of an element are
-constant values (as in listing \ref{forest}) rather than dynamically obtained (as in listing \ref{forest-loops}), the
-code generators inline the loop that adds children to their parent, for better performance. As a result, listing
+client-side and code building an equivalent XML fragment on server-side. When the children of an element are constant
+values (as in listing \ref{forest}) rather than dynamically computed (as in listing \ref{forest-loops}), the code
+generators inline the loop that adds children to their parent, for better performance. As a result, listing
 \ref{forest} generates a code equivalent to listing \ref{dom-api} on client-side and equivalent to
 \ref{scala-xml-api} on server-side.
+
+\begin{figure}
+\begin{lstlisting}[language=JavaScript,label=js-gen-forest,caption=JavaScript code generator for the DOM fragment
+definition DSL]
+case Tag(name, children, attrs) =>
+  emitValDef(sym, q"document.createElement('$name')")
+  for ((n, v) <- attrs) {
+    stream.println(q"$sym.setAttribute('$n', $v);")
+  }
+  children match {
+    case Left(children) =>
+      for (child <- children) {
+        stream.println(q"$sym.appendChild($child);")
+      }
+    case Right(children) =>
+      val x = fresh[Int]
+      stream.println(q"for (var $x = 0; $x < $children.length; $x++) {")
+      stream.println(q"$sym.appendChild($children[$x]);")
+      stream.println("}")
+  }
+case Text(content) =>
+  emitValDef(sym, q"document.createTextNode($content)")
+\end{lstlisting}
+\end{figure}
+
+\begin{figure}
+\begin{lstlisting}[label=scala-gen-forest,caption=Scala code generator for the DOM fragment definition DSL]
+case Tag(name, children, attrs) =>
+  val attrsFormatted =
+    (for ((name, value) <- attrs)
+     yield q" $name={ $value }").mkString
+  children match {
+    case Left(children) =>
+      if (children.isEmpty) {
+        emitValDef(sym, q"<$name$attrsFormatted />")
+      } else {
+        emitValDef(sym,
+          q"<name$attrsFormatted>{ ${children.map(quote)} }</$name>"
+        )
+      }
+    case Right(children) =>
+      emitValDef(sym, q"<$name$attrsFormatted>{ $children }</$name>")
+  }
+case Text(content) =>
+  emitValDef(sym, q"{xml.Text(content)}")
+\end{lstlisting}
+\end{figure}
+
+Listings \ref{js-gen-forest} and \ref{scala-gen-forest} show the relevant parts of the code generators for this DSL.
+They basically follow the same pattern: they visit \code{Tag} and \code{Text} IR nodes and produce the corresponding
+elements in the target language. (TODO More details)
 
@@
 
-\section{Implementation}
-\label{implementation}
-
-\footnote{The code is available at \href{http://github.com/js-scala}{http://github.com/js-scala}}
-
-\subsection{Selectors}
-
-\begin{figure}
-\begin{lstlisting}[label=selector-impl,caption=Selectors optimization]
-def find(receiver: Exp[Selector],
-         selector: Exp[String]) =
-  getConstIdentifier(selector) match {
-    case Some(id) if receiver == document =>
-      DocumentGetElementById(Const(id))
-    case _ =>
-      SelectorFind(receiver, selector)
-  }
-\end{lstlisting}
-\end{figure}
-
-Listing \ref{selector-impl}
-
-\subsection{Null references}
-
-Code generation consists in traversing the statement nodes produced by the program evaluation according to their
-dependencies and to emit the code corresponding to each statement. LMS already sorts the statements graph so DSL
-authors just need to say how to emit code for each statement node of their DSL. Listing \ref{option-codegen} shows
-such a code generator for the \code{null} reference handling DSL. The \code{emitNode} method handles
-\code{OptionIsEmpty} and \code{OptionForeach} nodes. In the case of the \code{OptionIsEmpty} node, it simply
-generates an expression testing if the value is \code{null}, in the case of the \code{OptionForeach} node, it wraps
-the code block dereferencing the value within a \code{if} checking that the value is not \code{null}.
-
-\begin{figure}
-\begin{lstlisting}[caption=Null reference handling DSL code generator,label=option-codegen]
-trait JSGenOptionOps extends JSGenEffect {
-  val IR: OptionOpsExp
-  import IR._
-  override def emitNode(sym: Sym[Any], rhs: Def[Any]) =
-    rhs match {
-      case OptionIsEmpty(o) =>
-        emitValDef(sym, q" $o === null")
-      case OptionForeach(o, a, b) =>
-        stream.println(q"if ($o !== null) {")
-        emitValDef(a, quote(o))
-        emitBlock(b)
-        stream.println("}")
-      case _ =>
-        super.emitNode(sym, rhs)
-    }
-}
-\end{lstlisting}
-\end{figure}
-
-\subsection{DOM Fragments}
-
 \section{Evaluation}
@@
 
-We implemented several applications using js-scala. We also have written several implementations of a complete
-application, using different approaches to write the client and server sides, and compared the amount of code
-written, the runtime performances and the ability to modularize the code and to maintain it.
+We measured the runtime performances of the DSL for optional values using a micro-benchmark.
+
+We also have written several implementations of a complete application using different approaches for the client and
+server sides, and compared the amount of code written, the runtime performances and the ability to modularize the
+code and to maintain it.
+
+\subsection{Micro-Benchmark}
 

### 2013-06-07 papers/sle2013/sle2013.tex
eod

@@
 
-\title{FFI Is Not Enough. We Need Dependent Types.}
+\title{Using Dependent Types to Build Safe and Concise JavaScript Foreign Function Interfaces}
 \titlerunning{Dependent Types Rock}
@@
 less control than if they were using JavaScript. This article shows how to expose Web browser APIs in Scala in a
-type safe way while keeping the same level of control as with native APIs. We use advanced type system features such
-as dependent types and functional dependencies.
-\keywords{Dependent Types}
+type safe way while keeping the same level of control as with native APIs by using dependent types and functional
+dependencies. We validate this approach in designing safe and concise foreign function interfaces between Scala and
+JavaScript for DOM events handling and DOM manipulation. We compared this approach to common frameworks such as GWT,
+Fay, Kotlin, Dart and SharpKit.
+\keywords{Dependent Types, JavaScript, Scala, Foreign Function Interface}
 \end{abstract}

### 2013-06-07 papers/sle2013/sle2013.tex
sle code examples

@@
 
-\section{Motivating Example}
-
-Typical tasks involved in Web applications.
+\section{Background Material}
 
 Why is it difficult to type Web browser’s APIs?
+
+\subsection{DOM Creation}
+
+\begin{lstlisting}[language=JavaScript]
+var div = document.createElement('div');
+var input = document.createElement('input');
+\end{lstlisting}
+
+\code{div} has type \code{DivElement} while \code{input} has type \code{InputElement}. These types have different
+methods. As stated in the introduction, most statically typed language have an API that returns an \code{Element},
+which is the least upper bound of the possible types returned by \code{createElement}, forcing users to explicitly
+down-cast the result to the expected type, thus loosing type safety.
+
+Alternatively, some languages use overloading.
+
+Consider the following function creating an element and giving it a class name:
+
+\begin{lstlisting}[language=JavaScript]
+var create = function (elementType, className) {
+  val el = document.createElement(elementType);
+  el.className = className;
+  return el
+};
+create('input', 'field').focus();
+create('img', 'figure').src = 'http://google.com/logo.png';
+\end{lstlisting}
+
+How to type check the above code? The problem is that the return type of the \code{create} function depends on its
+first parameter.
+
+Note that a possible solution in Java could be the following:
+
+\begin{lstlisting}[language=Java]
+class ElementName<E> {}
+ElementName<InputElement> Input = new ElementName<InputElement>();
+ElementName<ImageElement> Img = new ElementName<ImageElement>();
+
+<E> E create(ElementName<E> type, String className) {
+  ...
+}
+
+create(Input, "field").focus();
+create(Img, "figure").setSrc("http://google.com/logo.png");
+\end{lstlisting}
+
+However, the type parameter \code{E} has to be filled at usage site. If the compiler can infer it, that’s fine.
+
+\subsection{DOM Events}
+
+A similar issue applies to the DOM events API.
+
+\begin{lstlisting}[language=Java]
+class EventName<E> {}
+EventName<ClickEvent> Click = new EventName<ClickEvent>();
+
+<E> Stream<E> on(EventName<E> type, Callback<E>) {
+  ...
+}
+
+Callback<ClickEvent> incrementCounter = new Callback<ClickEvent>() {
+  @Override
+  void execute(ClickEvent event) {
+    ...
+  }
+}
+
+on(Click, incrementCounter);
+\end{lstlisting}
+
+Again, the type parameter \code{E} has to be filled at usage site.
+
+\subsection{Selectors}
+
+The problem with selectors is slightly different.
+
+\begin{lstlisting}[language=JavaScript]
+var input = document.querySelector('input');
+var el = document.querySelector('#content');
+\end{lstlisting}
+
+\code{input} can not have another type than \code{InputElement}, but \code{el} can be an element of any type (depending
+on which element has the id \code{content}). Return type can not \emph{always} be inferred from the parameters.
+In this case, down-casting is impossible to avoid, but how to make it optional, and how to make it less risky?
+
+We want to be able to write the following code:
+
+\begin{lstlisting}
+val input = document.find(Input) // input has type InputElement
+val el = document.querySelector("#content") // el has type Element
+val img = document.querySelector[Img](".figure") // note that the type Img is explicitly applied in this case but was automatically inferred in previous cases
+val error = document.querySelector[Int]("div") // Does not compile: querySelector can not return a number
+\end{lstlisting}
 

### 2013-06-10 src/main/scala/scala/js/exp/OptionOpsExp.scala
Disable optimization in Option because it seems to break depepndencies

@@ -12,7 +12,7 @@ trait OptionOpsExp extends OptionOps with EffectExp {
   implicit class OptionOpsCls[+A : Manifest](o: Exp[Option[A]]) extends OptionOpsBase[A] {
 
     private def sym = o match {
-      case Sym(n) => (Sym[A](n), false)
+      //case Sym(n) => (Sym[A](n), false)
       case _ => (fresh[A], true)
     }
 
@@ -48,10 +48,10 @@ trait OptionOpsExp extends OptionOps with EffectExp {
   case class OptionFold[A, B](o: Exp[Option[A]], a: Sym[A], noneBlock: Block[B], someBlock: Block[B], bound: Boolean) extends Def[B]
 
   override def syms(e: Any) = e match {
-    case OptionForeach(o, _, block, _) => syms(o) ++ syms(block)
-    case OptionMap(o, _, block, _) => syms(o) ++ syms(block)
-    case OptionFlatMap(o, _, block, _) => syms(o) ++ syms(block)
-    case OptionFold(o, _, b1, b2, _) => syms(o) ++ syms(b1) ++ syms(b2)
+    case OptionForeach(o, a, block, b) => (if (!b) syms(a) else Nil) ++ syms(o) ++ syms(block)
+    case OptionMap(o, a, block, b) => (if (!b) syms(a) else Nil) ++ syms(o) ++ syms(block)
+    case OptionFlatMap(o, a, block, b) => (if (!b) syms(a) else Nil) ++ syms(o) ++ syms(block)
+    case OptionFold(o, a, b1, b2, b) => (if (!b) syms(a) else Nil) ++ syms(o) ++ syms(b1) ++ syms(b2)
     case _ => super.syms(e)
   }
 
@@ -64,10 +64,10 @@ trait OptionOpsExp extends OptionOps with EffectExp {
   }
 
   override def symsFreq(e: Any) = e match {
-    case OptionForeach(o, _, block, _) => freqNormal(o) ++ freqCold(block)
-    case OptionMap(o, _, block, _) => freqNormal(o) ++ freqCold(block)
-    case OptionFlatMap(o, _, block, _) => freqNormal(o) ++ freqCold(block)
-    case OptionFold(o, _, b1, b2, _) => freqNormal(o) ++ freqCold(b1) ++ freqCold(b2)
[…]

### 2013-06-10 src/test/scala/scala/js/TestOptionOps.scala

@@ -4,7 +4,7 @@ import scala.virtualization.lms.common._
 import java.io.PrintWriter
 import language.OptionOps
 import exp.OptionOpsExp
-import gen.js.{GenEffect, GenNumericOps, GenOptionOps}
+import gen.js.{GenEffect, GenNumericOps, GenOptionOps, GenEqual, GenIfThenElse, GenPrimitiveOps}
 
 class TestOptionOps extends FileDiffSuite {
   val prefix = "test-out/"
@@ -56,5 +56,24 @@ class TestOptionOps extends FileDiffSuite {
       codegen.emitSource(prog.main, "test", new PrintWriter(System.out))
     }
     assertFileEqualsCheck(prefix+"option-getorelse")
+  }
+
+  def testDependencies() {
+    trait Prog extends DSL with TupledFunctions with IfThenElse {
+      lazy val rec: Rep[((Int => Option[Int], Int => Boolean, Int)) => Option[Int]] = fun { (f: Rep[Int => Option[Int]], p: Rep[Int => Boolean], x: Rep[Int]) =>
+        for {
+          y <- f(x)
+          z <- if (p(y)) some(y) else rec(f, p, y)
+        } yield z
+        //if (p(x)) some(x) else rec(f, p, x)
+      }
+      def main(f: Rep[Int => Option[Int]], p: Rep[Int => Boolean], x: Rep[Int]) = rec(f, p, x)
+    }
+    withOutFile(prefix + "option-deps") {
+      val prog = new Prog with DSLExp with TupledFunctionsRecursiveExp with IfThenElseExp
+      val codegen = new DSLJSGen with gen.js.GenFunctions with GenericGenUnboxedTupleAccess with GenIfThenElse { val IR: prog.type = prog }
+      codegen.emitSource3(prog.main, "test", new PrintWriter(System.out))
+    }
+    assertFileEqualsCheck(prefix + "option-deps")
   }
 }

### 2013-06-10 src/main/scala/scala/js/exp/dom/ElementOpsExp.scala
Minor tweak on Element#closest and Element#prev

@@ -20,12 +20,18 @@ trait ElementOpsExp extends ElementOps with EffectExp with TupledFunctionsExp wi
   def element_focus(e: Exp[Element]) = reflectEffect(ElementFocus(e))
   def element_closest_reified: Exp[((Element, Element => Boolean)) => Option[Element]] =
     fun { (e: Exp[Element], p: Exp[Element => Boolean]) =>
-      e.parentNode.fold(none, parent => if (p(parent)) some(parent) else element_closest_reified(parent, p))
+      for {
+        parent <- e.parentNode
+        closest <- if (p(parent)) some(parent) else element_closest_reified(parent, p)
+      } yield closest
     }
   def element_closest(e: Exp[Element], p: Exp[Element] => Exp[Boolean]) = element_closest_reified(e, p)
   def element_prev_reified: Exp[((Element, Element => Boolean)) => Option[Element]] =
     fun { (e: Exp[Element], p: Exp[Element => Boolean]) =>
-      e.previousSibling.fold(none, sib => if (p(sib)) some(sib) else element_prev_reified(sib, p))
+      for {
+        sib <- e.previousSibling
+        prev <- if (p(sib)) some(sib) else element_prev_reified(sib, p)
+      } yield prev
     }
   def element_prev(e: Exp[Element], p: Exp[Element] => Exp[Boolean]) = element_prev_reified(e, p)
   def element_remove(e: Exp[Element]) = fun { e: Exp[Element] =>

### 2013-06-10 papers/gpce2013/gpce2013.tex
figures

@@
 
+\begin{figure}
+\centering
+\includegraphics[width=8cm]{microbenchmark.png}
+\caption{Micro-benchmark on the optional values abstraction}
+\label{micro-benchmark}
+\end{figure}
+
+Figure \ref{micro-benchmark} shows the micro-benchmark results. We measured the runtime performance, the number of
+lines of code (as an indication of the abstraction level), and finally the ratio of runtime performance and lines of
+code. \emph{js-scala} is between 3 to 100 times faster than other approaches on our benchmark. Combined with that
+fact that it provides high-level abstractions, it has a performance / lines of code ratio more than 4 times higher
+than other approaches.
+
 \subsection{Real World Application}

### 2013-06-11 src/main/scala/scala/js/language/JSLiteral.scala
upgrade to 2.10.2-RC1

@@ -3,7 +3,7 @@ package scala.js.language
 import scala.virtualization.lms.common.Base
 
 trait JSLiteral extends Base with EmbeddedControls {
-  trait JSLiteral extends Struct[Rep]
+  trait JSLiteral extends Struct
   def __new[T](args: (String, Boolean, Rep[T] => Rep[_])*): Rep[T] =
     newJSLiteral(args.map(x => (x._1, x._3).asInstanceOf[(String, Rep[JSLiteral] => Rep[_])]): _*).asInstanceOf[Rep[T]]
   def newJSLiteral(args: (String, Rep[JSLiteral] => Rep[_])*): Rep[JSLiteral]

### 2013-06-11 papers/gpce2013/gpce2013.tex

@@
 
+\subsection{Language Engineering Processes}
+
 This section presents different approaches for defining cross-platform programming languages.
@@
 The last approach, shown in figure \ref{langs} (d), can be seen as a middle-ground between the two previous
-approaches: DSLs are embedded in a host language but use a code generation process. (benefits and limitations of
-this approach)
+approaches: DSLs are embedded in a host language but use a code generation process. The embedding makes it easy to
+combine several DSLs and makes the language implementation easier because the host language infrastructure is
+reused. On the other hand, the syntax of the DSLs is limited by the syntax flexibility of the host language. The code
+generation process is specific to each DSL and gives the opportunity to perform domain-specific optimizations.
 
-Lightweight Modular Staging~\cite{Rompf12_LMSThesis} (LMS) is a framework for defining deeply embedded DSLs in Scala
-using staging~\cite{Jorring1986_Staging}. It has been used to define high-performance DSLs for parallel
-computing~\cite{Brown11_Parallel}. (Todo. Mention \code{Rep[_]})
+(Summary explaining why we choose the deep embedded DSLs approach)
 
-Code generation consists in traversing the statement nodes produced by the program evaluation according to their
-dependencies and to emit the code corresponding to each statement. LMS already sorts the statements graph so DSL
-authors just need to say how to emit code for each statement node of their DSL.
+\subsection{Lightweight Modular Staging}
+
+Lightweight Modular Staging~\cite{Rompf12_LMSThesis, Rompf12_LMS} (LMS) is a framework for defining deeply embedded
+DSLs in Scala. It has been used to define high-performance DSLs for parallel computing~\cite{Brown11_Parallel} and to
+define JavaScript as an embedded DSL in Scala~\cite{Kossakowski12_JsDESL}.
+
+LMS is based on staging~\cite{Jorring1986_Staging}: a program using LMS is a regular Scala program that evaluates
+to an intermediate representation (IR) of a final program. A code generator can traverse this IR to produce the final
+program code. Expressions evaluated in the initial program and those evaluated in the final program are distinguished
+by their type: a \code{Rep[Int]} value in the initial program is a staged computation that generates code evaluating
+to an \code{Int} value in the final program. An \code{Int} computation in the initial program is evaluated during the
+initial program evaluation and becomes a constant in the final program.
+
+Defining a DSL with LMS consists in the following steps:
+
+\begin{itemize}
+ \item writing a Scala module providing the DSL vocabulary as an abstract API,
+ \item implementing the API in terms of IR nodes,
+ \item defining a code generator visiting IR nodes and generating the corresponding code.
+\end{itemize}
 
@@
 \begin{lstlisting}[label=selector-impl,caption=Selectors optimization]
-def find(receiver: Exp[Selector],
-         selector: Exp[String]) =
+def find(receiver: Rep[Selector],
+         selector: Rep[String]) =
   getConstIdCss(selector) match {
@@
 server sides, and compared the amount of code written, the runtime performances and the ability to modularize the
-code and to maintain it.
+code.
 
@@
 lines of code (as an indication of the abstraction level), and finally the ratio of runtime performance and lines of
-code. \emph{js-scala} is between 3 to 100 times faster than other approaches on our benchmark. Combined with that
+code. \emph{js-scala} is between 3 to 100 times faster than other approaches on our benchmark. Combined with the
 fact that it provides high-level abstractions, it has a performance / lines of code ratio more than 4 times higher
@@
 The application was initially written using jQuery. We rewrote it using several technologies for the client-side
-part: plain JavaScript (without third-party library), js-scala, GWT and HaXe.
+part: vanilla JavaScript (low-level code without third-party library), js-scala, GWT and HaXe. In each case we tried
+to write the application in an idiomatic way.
 
@@
 
-Our goal is to evalute the level of abstraction provided by each solution and their performances. We took the
-number of lines of code as a measure of the level of abstraction. We also measured the size of the data sent to the
-client-side. We also measured the ability to share code between client and server sides.
+Our goal is to evaluate the level of abstraction provided by each solution and their performances. We took the
+number of lines of code as a measure of the level of abstraction. We also measured the ability to share code between
+client and server sides.
 
-Rather than doing a micro-benchmark focusing on just one abstraction, we performed a global benchmark that is more
-likely to reflect the real performances of the application: we simulated user actions on a Web page (2000 clicks on
-buttons, triggering a dynamic update of the page and involving the use of the Option monad, the selectors API and
-the HTML fragment definition API) and measured the time it took complete them. We only measured the execution time of
-client-side code execution. The tests were run on a DELL Latitude E6430 laptop with 8 GB of RAM, on the Google
-Chrome v27 Web browser. Figure \ref{benchmark} shows the benchmark results.
+We simulated user actions on a Web page (2000 clicks on buttons, triggering a dynamic update of the page and
+involving the use of the Option monad, the selectors API and the HTML fragment definition API) and measured the time
+it took complete them. We only measured the execution time of client-side code execution. The tests were run on a
+DELL Latitude E6430 laptop with 8 GB of RAM, on the Google Chrome v27 Web browser. Figure \ref{benchmark} shows the
+benchmark results.
 
@@
 \end{figure}
+
+The runtime performances of the vanilla JavaScript, HaXe and \emph{js-scala} versions are similar (though the
+\emph{js-scala} version is slightly slower by 6\%). It is worth noting that the vanilla JavaScript and the HaXe
+versions use low-level code compared to \emph{js-scala}, as shown in the second part of the figure (lines of code):
+the \emph{js-scala} version needs only 74 lines of code while the vanilla JavaScript version needs 116 lines of code
+(57\% bigger) and the HaXe version needs 148 lines of code (100\% bigger). The jQuery JavaScript version, of which
+code is high-level (54 lines of code, 27\% less than \emph{js-scala}) runs 10 times slower than the \emph{js-scala}
+version.
+
+The last part of the figure compares the runtime performance / lines of code ratio. \emph{js-scala} shows the best
+score, being 1.48 times better than the vanilla JavaScript version, 1.88 times better than the HaXe version,
+3.77 times better than the GWT version and 7.82 times better than the jQuery JavaScript version.
 

### 2013-06-11 papers/gpce2013/gpce2013.tex
gpce draft

@@
 
-\title{How to build an homogeneous language for heterogeneous platforms}
-\subtitle{You don’t have to trade abstraction for control}
+\title{Efficient High-Level Abstractions for Web Programming}
+%\subtitle{You don’t have to trade abstraction for control}
 
@@
 This article presents high-level abstractions defined as deep embedded DSLs in Scala, that can (1) generate efficient
-code leveraging the target platform characteristics, (2) be shared between clients and servers. Code written with
-our DSL has a high level of abstraction but runs as fast as hand tuned low-level JavaScript code.
+code leveraging the target platform characteristics, (2) be shared between clients and servers. Though code written
+with our DSL has a high level of abstraction our benchmark on a real world application reports that it runs as fast
+as hand tuned low-level JavaScript code.
 \end{abstract}
@@
 
-\keywords Heterogeneous code generation, Domain-specific languages
+\keywords Heterogeneous code generation, Domain-specific languages, Scala, Web
 
@@
 environments~\cite{Echeverria09_RIA,Kuuskeri09_PartitioningClientServer}. This gives less flexibility in the
-engineering process and requires a higher maintenance effort: if a piece of logic is implemented on client-side and
-finally needs to be implemented on server-side instead, the code can not be reused and the feature needs to be
-completely rewritten (and \emph{vice versa}). Even worse, logic parts that run on both client-side and server-side
-need to be duplicated. For instance, HTML fragments may be built from the server-side when a page is requested by a
-client, but they may also be built from the client-side to perform an incremental update subsequent to an user
-action. How could developers write HTML fragment definitions once and render them on both client-side and
-server-side?
+engineering process and requires a higher maintenance effort: there is no way to move a piece of code targeting
+the server-side to target the client-side, the code has to be rewritten. Even worse, logic parts that run on both
+client-side and server-side need to be duplicated. For instance, HTML fragments may be built from the server-side
+when a page is requested by a client, but they may also be built from the client-side to perform an incremental
+update subsequent to an user action. How could developers write HTML fragment definitions once and render them on
+both client-side and server-side?
 
-The more interactive the application is, the more logic needs to be duplicated between the server-side
-and the client-side, and the higher is the complexity of the client-side code. Developers use libraries and
-frameworks to get high-level abstractions on client-side, making their code easier to reason about and to maintain,
-but also making their code run less efficiently (due to \emph{abstraction penalty}).
+The more interactive the application is, the more logic needs to be duplicated between the server-side and the
+client-side, and the higher is the complexity of the client-side code. Developers use libraries and frameworks to get
+high-level abstractions on client-side, making their code easier to reason about and to maintain, but also making
+their code run less efficiently (due to \emph{abstraction penalty}).
 
@@
 common abstractions like collections into their native counterpart (JavaScript arrays on client-side and standard
-library's collections on server-side) may be hard coded in the compiler, but that would not scale to handle all the
-abstractions a complete application may use (\eg HTML fragment definitions, form validation rules, or even some
-business data type that may be represented differently).
+library's collections on server-side) may be hard-coded in the compiler, but that approach would not scale to handle
+all the abstractions a complete application may use (\eg HTML fragment definitions, form validation rules, or even
+some business data type that may be represented differently).
 
@@
 JavaScript~\cite{Kossakowski12_JsDESL}. However, Kossakowski \etal did not address the engineering dilemma described
-above. This paper enriches \emph{js-scala} with the following contributions\footnote{The code is available at
+above. This paper enriches js-scala with the following contributions\footnote{The code is available at
 \href{http://github.com/js-scala}{http://github.com/js-scala}}:
@@
 pro and cons of them. Though the code written in our DSL is high-level and can be shared between clients and servers,
-it has the same runtime performances as hand-tuned low-level JavaScript code.
+it has the same runtime performances on client-side as hand-tuned low-level JavaScript code.
 
 The remainder of this paper is organized as follows. The next section introduces existing approaches for defining
-cross-compiling languages. Sections \ref{contribution} and \ref{contribution-shared} present our contribution.
-Section \ref{validation} compares our solution to common approaches. Section \ref{discussion} concludes.
+cross-compiling languages and presents the framework we used to define our DSLs. Sections \ref{contribution} and
+\ref{contribution-shared} present our contribution. Section \ref{validation} compares our solution to common
+approaches. Section \ref{discussion} concludes.
 
@@
 generator, \etc) may have to be adapted. This approach gives \emph{fat} languages because a lot of concepts are
-defined at the language level: general programming concepts such as naming, functions, classes, and more specific
-concepts such as HTML fragment definition. Examples of such languages are Links~\cite{Cooper07_Links}, Opa and
-Dart~\cite{Griffith11_Dart}. These languages are difficult to extend because each concept is defined in the compiler,
-and modifying a compiler requires a high effort. Furthermore, these languages also require to support common
-programming abstraction and composition mechanisms, as general purpose languages do. So they usually try to re-invent
-the features of general purpose languages. We argue that this approach for defining programming languages is
-difficult to scale: for every problem you have to rewrite a full-featured programming language besides addressing the
-concepts specific to the problem domain.
+defined at the language level: general programming concepts such as naming, functions, classes, as well as more
+domain-specific concepts such as HTML fragment definition. Examples of such languages for Web programming are
+Links~\cite{Cooper07_Links}, Opa and Dart~\cite{Griffith11_Dart}. These languages are difficult to extend because
+each concept is defined in the compiler, and modifying a compiler requires a high effort. Furthermore, these
+languages also require to support common programming abstraction and composition mechanisms, as general purpose
+languages do. So they usually try to re-invent the features of general purpose languages, that’s why we argue that
+this approach for defining programming languages is difficult to scale: for every specific problem you would have to
+rewrite a full-featured programming language besides addressing the concepts specific to the problem domain.
 
@@
 focusing on concerns specific to a given problem domain, and then to combine all the source artifacts written with
-these language into one executable program, as shown in figure~\ref{langs} (b). Defining such a language requires a
-minimal effort compared to the previous approach. On the other hand, it is difficult to have interoperability between
-DSLs (ref). \cite{Groenewegen08_WebDSL} shows such a domain-specific language for defining Web applications.
+these language into one executable program, as shown in figure~\ref{langs} (b). Defining such languages requires a
+minimal effort compared to the previous approach because each language has a limited set of features. On the other
+hand, it is difficult to have interoperability between DSLs (missing ref). \cite{Groenewegen08_WebDSL} gave an
+example of such a domain-specific language for defining Web applications.
 
@@
 
-Alternatively, one can define concepts relative to a specific domain as libraries on top of a thin general purpose
-language (this is also referred to as \emph{embedded} domain-specific languages). Figure \ref{langs} (c) depicts this
-approach. The general purpose language is used as a host language and does not need to be modified if a new concept
-is introduced, because concepts are defined as pure libraries. However, this approach gives no opportunity to
-translate a concept efficiently according to the target platform characteristics. Examples of languages following
-this approach are Java/GWT, Kotlin, HaXe and SharpKit.
+Alternatively, one can define concepts relative to a specific domain as a library on top of a thin general purpose
+language (it is also referred to as a domain-specific \emph{embedded} language~\cite{Hudak96_DSEL}). Figure
+\ref{langs} (c) depicts this approach. The general purpose language is used as a host language and does not need to
+be modified if a new concept is introduced, because concepts are defined as pure libraries. On the other hand, the
+syntax of the DSLs is limited by the syntax flexibility of the host language. Furthermore, this approach gives no
+opportunity to translate a concept efficiently according to the target platform characteristics because concepts are
+defined as libraries and are translated by the compiler, which has no domain-specific knowledge. Examples of
+languages following this approach are Java/GWT, Kotlin, HaXe and SharpKit.
 
@@
 The last approach, shown in figure \ref{langs} (d), can be seen as a middle-ground between the two previous
-approaches: DSLs are embedded in a host language but use a code generation process. The embedding makes it easy to
-combine several DSLs and makes the language implementation easier because the host language infrastructure is
-reused. On the other hand, the syntax of the DSLs is limited by the syntax flexibility of the host language. The code
-generation process is specific to each DSL and gives the opportunity to perform domain-specific optimizations.
+approaches: DSLs are embedded in a host language but use a code generation process. This approach share the same
+benefits and limits as embedded DSLs for defining language units. The code generation process is specific to each DSL
+and gives the opportunity to perform domain-specific optimizations. In other words deeply embedded DSLs bring
+domain-specific knowledge to the compiler.
 
@@
 LMS is based on staging~\cite{Jorring1986_Staging}: a program using LMS is a regular Scala program that evaluates
-to an intermediate representation (IR) of a final program. A code generator can traverse this IR to produce the final
-program code. Expressions evaluated in the initial program and those evaluated in the final program are distinguished
-by their type: a \code{Rep[Int]} value in the initial program is a staged computation that generates code evaluating
-to an \code{Int} value in the final program. An \code{Int} computation in the initial program is evaluated during the
-initial program evaluation and becomes a constant in the final program.
+to an intermediate representation (IR) of a final program. This IR is a graph of expressions that can be traversed
+by code generators to produce the final program code. Expressions evaluated in the initial program and those
+evaluated in the final program (namely, staged expressions) are distinguished by their type: a \code{Rep[Int]} value
+in the initial program is a staged expression that generates code evaluating to an \code{Int} value in the final
+program. An \code{Int} computation in the initial program is evaluated during the initial program evaluation and
+becomes a constant in the final program.
 
@@
 
-Listing~\ref{js-scala-selectors} shows how to implement listing~\ref{jquery-selectors} using \emph{js-scala}. We
+Listing~\ref{js-scala-selectors} shows how to implement listing~\ref{jquery-selectors} using js-scala. We
 provide two functions for searching elements: \code{find} to find the first element matching a selector and
-\code{findAll} to find all the matching elements. During staging these functions try to analyze the selector that is
-passed as parameter and, when appropriate, produce code using the specialized API, otherwise they produce code using
-\code{querySelector} and \code{querySelectorAll}. As a result, listing~\ref{js-scala-selectors} generates a
-JavaScript program identical to listing~\ref{vanilla-selectors}: the high-level abstraction (the \code{find} and
-\code{findAll} functions) exist only in the initial program, not in the final JavaScript program. It gives two
-advantages: (1) the execution of the final program is more performant because of the use of specialized APIs, (2) the
-final program’s size is smaller because it does not need to include jQuery. (Todo. Move this last sentence in the
-discussion)
+\code{findAll} to find all the matching elements. During the first evaluation step, these functions try to analyze
+the selector that is passed as parameter and, when appropriate, produce code using the specialized  API, otherwise
+they produce  code using \code{querySelector} and \code{querySelectorAll}. As a result,
+listing~\ref{js-scala-selectors} generates a JavaScript program identical to listing~\ref{vanilla-selectors}: the
+high-level abstraction (the \code{find} and \code{findAll} functions) exist only in the initial program, not in the
+final JavaScript program.
 
@@
 
-Listing \ref{selector-impl} shows the implementation of the \code{find} function producing different IR node
+Listing \ref{selector-impl} shows the implementation of the \code{find} function producing different IR nodes
 according to the selector passed as parameter. The \code{getConstIdCss} function analyzes the selector: if it
@@
 example, consider listing \ref{null-unsafe} finding a particular widget in the page and then a particular
-button in the widget. The native \code{querySelector} method returns \code{null} if no node matched the given
+button within the widget. The native \code{querySelector} method returns \code{null} if no node matched the given
 selector in the document. If we run this code in a page where the widget is not present, it will throw an error
@@
 
-Listing \ref{null-js-scala} implements in \emph{js-scala} a program equivalent to listing \ref{null-defensive}. The
-\code{for} notation is used to sequence computations over optional values encoded with a monad. The \code{find}
-function returns a \code{Rep[Option[Element]]} value, that can either be a \code{Rep[Some[Element]]} (if an element
-was found) or a \code{Rep[None.type]} (if no element was found). The \code{for} expression contains a sequence
-of statements that are executed in order, as long as the previous statement returned a \code{Rep[Some[Element]]}
-value.
+Listing \ref{null-js-scala} implements in js-scala a program equivalent to listing \ref{null-defensive}. The
+\code{for} notation is used to sequence computations over optional values that are encoded with a monad. The
+\code{find} function returns a \code{Rep[Option[Element]]} value, that can either be a \code{Rep[Some[Element]]} (if
+an element was found) or a \code{Rep[None.type]} (if no element was found). The \code{for} expression contains a
+sequence of statements that are executed in order, as long as the previous statement returned a
+\code{Rep[Some[Element]]} value.
 
@@
 Such a monadic API brings both safety and expressiveness to developers manipulating optional values but usually
-involves the creation of an extra container object holding the optional value. In our case, we expose the monadic API
-at the first stage but we generate code that does not wrap values in container objects but instead checks if they are
-\code{null} or not when dereferenced. So the extra container object exists only at the stage-level and is removed
-during code generation: listing \ref{null-js-scala} produces a code equivalent to listing \ref{null-defensive}.
+involves the creation of an extra container object holding the optional value. In our case, the monadic API is
+used in the initial program but generates code that does not wrap values in container objects but instead checks if
+they are \code{null} or not when dereferenced. So the extra container object exists only in the initial program and
+is removed during code generation: listing \ref{null-js-scala} produces a code equivalent to listing
+\ref{null-defensive}.
 
@@
 
-Last but not least, we leverage the modular architecture of LMS to define two code generators for this API: one
-targeting JavaScript and the other targeting Scala. So the same abstraction is efficiently translated on both server
-and client sides.
+Last but not least, we make this abstraction available on server-side by writing a code generator similar to the
+JavaScript code generator, but targeting Scala (FIXME Show the code generator?). So the same abstraction is
+efficiently translated on both server and client sides.
 
@@
 They basically follow the same pattern: they visit \code{Tag} and \code{Text} IR nodes and produce the corresponding
-elements in the target language. (TODO More details)
+elements in the target language. (FIXME More details)
 
@@
 
+Our goal was to evaluate the level of abstraction provided by each solution and their performances. We took the
+number of lines of code as a measure of the level of abstraction. We also measured the ability to share code between
+client and server sides.
+
+The tests were run on a DELL Latitude E6430 laptop with 8 GB of RAM, on the Google Chrome v27 Web browser.
+
 \subsection{Micro-Benchmark}
+
+We reimplemented the optional value abstraction and a same program using it in the following languages: JavaScript,
+HaXe and Java.
 
@@
 
-Figure \ref{micro-benchmark} shows the micro-benchmark results. We measured the runtime performance, the number of
-lines of code (as an indication of the abstraction level), and finally the ratio of runtime performance and lines of
-code. \emph{js-scala} is between 3 to 100 times faster than other approaches on our benchmark. Combined with the
-fact that it provides high-level abstractions, it has a performance / lines of code ratio more than 4 times higher
-than other approaches.
+Figure \ref{micro-benchmark} shows the micro-benchmark results. Js-scala is between 3 to 100 times faster than
+other approaches on our benchmark. Combined with the fact that it provides high-level abstractions, it has a
+performance / lines of code ratio more than 4 times higher than other approaches.
 
@@
 The application was initially written using jQuery. We rewrote it using several technologies for the client-side
-part: vanilla JavaScript (low-level code without third-party library), js-scala, GWT and HaXe. In each case we tried
-to write the application in an idiomatic way.
+part: vanilla JavaScript (low-level code without third-party library), js-scala, GWT and HaXe. In each case we
+tried to write the application in an idiomatic way\footnote{Source code is available at
+\href{http://github.com/julienrf/chooze}{http://github.com/julienrf/chooze}}.
 
-\paragraph*{Vanilla JavaScript} Low-level code.
+\subsubsection{Performance}
 
-\paragraph*{jQuery} High-level code.
-
-\paragraph*{js-scala} High-level code. HTML fragment definition reused between server and client sides.
-
-\paragraph*{HaXe} Low-level code.
-
-\paragraph*{GWT} High-level code?
-
-\subsection{Benchmarks, Code Metrics}
-
-Our goal is to evaluate the level of abstraction provided by each solution and their performances. We took the
-number of lines of code as a measure of the level of abstraction. We also measured the ability to share code between
-client and server sides.
-
-We simulated user actions on a Web page (2000 clicks on buttons, triggering a dynamic update of the page and
-involving the use of the Option monad, the selectors API and the HTML fragment definition API) and measured the time
-it took complete them. We only measured the execution time of client-side code execution. The tests were run on a
-DELL Latitude E6430 laptop with 8 GB of RAM, on the Google Chrome v27 Web browser. Figure \ref{benchmark} shows the
-benchmark results.
+The benchmark code simulates user actions on a Web page (2000 clicks on buttons, triggering a dynamic update of
+the page and involving the use of the optional value monad, the selectors API and the HTML fragment definition API).
+Figure \ref{benchmark} shows the benchmark results.
 
@@
 
-The runtime performances of the vanilla JavaScript, HaXe and \emph{js-scala} versions are similar (though the
-\emph{js-scala} version is slightly slower by 6\%). It is worth noting that the vanilla JavaScript and the HaXe
-versions use low-level code compared to \emph{js-scala}, as shown in the second part of the figure (lines of code):
-the \emph{js-scala} version needs only 74 lines of code while the vanilla JavaScript version needs 116 lines of code
-(57\% bigger) and the HaXe version needs 148 lines of code (100\% bigger). The jQuery JavaScript version, of which
-code is high-level (54 lines of code, 27\% less than \emph{js-scala}) runs 10 times slower than the \emph{js-scala}
-version.
+The runtime performances of the vanilla JavaScript, HaXe and js-scala versions are similar (though the js-scala
+version is slightly slower by 6\%). It is worth noting that the vanilla JavaScript and the HaXe versions use
+low-level code compared to js-scala, as shown in the second part of the figure (lines of code): the js-scala version
+needs only 74 lines of code while the vanilla JavaScript version needs 116 lines of code (57\% bigger) and the HaXe
+version needs 148 lines of code (100\% bigger). The jQuery JavaScript version, of which code is high-level (54 lines
+of code, 27\% less than js-scala) runs 10 times slower than the js-scala version.
 
-The last part of the figure compares the runtime performance / lines of code ratio. \emph{js-scala} shows the best
-score, being 1.48 times better than the vanilla JavaScript version, 1.88 times better than the HaXe version,
-3.77 times better than the GWT version and 7.82 times better than the jQuery JavaScript version.
+The last part of the figure compares the runtime performance / lines of code ratio. Js-scala shows the best score,
+being 1.48 times better than the vanilla JavaScript version, 1.88 times better than the HaXe version, 3.45 times
+better than the GWT version and 7.82 times better than the jQuery JavaScript version.
+
+\subsubsection{Code Reuse}
+
+We were able to share HTML DOM fragment definitions between server-side and client-side in js-scala.
 
@@
 We implemented a high-level language abstracting over client and server heterogeneity but producing efficient code.
+Generated code size?
 
@@
 
-Acknowledgments, if needed.
+This work was funded by Zenexity.
 

### 2013-07-15 core/src/main/scala/scala/js/gen/js/GenModuleAMD.scala
GenModule modified GenModuleAMD created GenModuleScala created 3 test for the third module 6 file *.check (2 files for 1 test)

@@ -4,7 +4,7 @@ import java.io.PrintWriter
 import scala.virtualization.lms.internal.GenericCodegen
 
 
-  trait GenModule extends Codegen {
+  trait GenModuleAMD extends Codegen {
     def module: Module
     abstract sealed class Element
     case class Module(nes: (String, Element)*) extends Element
@@ -32,10 +32,12 @@ import scala.virtualization.lms.internal.GenericCodegen
       val args = List(s1, s2, s3)
       Function(args, body)
     }
-    def emitModule(name: String, out: java.io.PrintWriter) {
-      out.println(s"""var "$name" = {""")
+    def emitModuleAMD(out: java.io.PrintWriter) {
+      out.println("define([],function(){")
+      out.println("return{")
       emitModule(module, out)
-      out.println("};")
+      out.println("}")
+      out.println("});")
       out.close()
     }
     protected def emitElement(m: Element, out: java.io.PrintWriter) {

### 2013-07-15 core/src/main/scala/scala/js/gen/scala/GenModule.scala

@@ -1,10 +1,9 @@
-package scala.js.gen.js
+package scala.js.gen.scala
 
 import java.io.PrintWriter
-import scala.virtualization.lms.internal.GenericCodegen
+import scala.virtualization.lms.common.ScalaGenBase
 
-
-  trait GenModule extends Codegen {
+  trait GenModule extends ScalaGenBase {
     def module: Module
     abstract sealed class Element
     case class Module(nes: (String, Element)*) extends Element
@@ -33,32 +32,30 @@ import scala.virtualization.lms.internal.GenericCodegen
       Function(args, body)
     }
     def emitModule(name: String, out: java.io.PrintWriter) {
-      out.println(s"""var "$name" = {""")
+      out.println(s"""object $name {""")
       emitModule(module, out)
       out.println("};")
       out.close()
     }
     protected def emitElement(m: Element, out: java.io.PrintWriter) {
       m match {
-        case f @ Function(arg, body) => emitSource(arg, body,"",out)(f.Manifest)
-        case m: Module =>
-          out.println("{")
-          emitModule(m, out)
-          out.print("}")
+        case f @ Function(arg, body) => emitSource(arg, body,"" ,out)(f.Manifest)
+        case m: Module => emitModule(m, out)
       }
     }
     protected def emitModule(mod: Module, out: java.io.PrintWriter) {
       mod.nes match {
         case (name, el) +: nes => 
           def emitEl(n: String, e: Element) = {
-            out.print(s""""$n": """)
+            out.print(s"""object $n {""")
             emitElement(e, out)
           }
           emitEl(name, el)
           for ((name, el) <- nes) {
-            out.println(",")
+            out.println("},")
             emitEl(name, el)
           }
+          out.println("}")
         case _ =>
       }
     }

### 2013-07-15 core/src/test/scala/scala/js/TestModuleAMD.scala

@@ -1,10 +1,10 @@
 package scala.js
 
 import java.io.PrintWriter
-import scala.js.gen.js.{GenEffect, GenNumericOps, GenModule}
+import scala.js.gen.js.{GenEffect, GenNumericOps, GenModuleAMD}
 import scala.virtualization.lms.common.{NumericOps, NumericOpsExp}
 
-class TestModule extends FileDiffSuite {
+class TestModuleAMD extends FileDiffSuite {
 
   trait Prog extends NumericOps {
 
@@ -22,8 +22,8 @@ class TestModule extends FileDiffSuite {
     val prog = new Prog with NumericOpsExp 
 
     val prefix = "test-out/"
-    withOutFile(prefix + "module") {
-      val genModule = new GenEffect with GenNumericOps with GenModule {
+    withOutFile(prefix + "moduleAMD") {
+      val genModuleAMD = new GenEffect with GenNumericOps with GenModuleAMD {
         val IR: prog.type = prog
         val module = Module(
             ("pack" -> Module("fun1" -> fun(prog.id _))),
@@ -32,23 +32,23 @@ class TestModule extends FileDiffSuite {
             ("fun4" -> fun(prog.third _))
         )
       }
-      genModule.emitModule("Module 1",new PrintWriter(System.out))
+      genModuleAMD.emitModuleAMD(new PrintWriter(System.out))
     }
-    assertFileEqualsCheck(prefix + "module")
+    assertFileEqualsCheck(prefix + "moduleAMD")
   }
   
   def testEmitEmptyModule() = {
     val prog = new Prog with NumericOpsExp 
 
     val prefix = "test-out/"
-    withOutFile(prefix + "module-empty") {
-      val genModule = new GenEffect with GenNumericOps with GenModule {
+    withOutFile(prefix + "moduleAMD-empty") {
+      val genModuleAMD = new GenEffect with GenNumericOps with GenModuleAMD {
         val IR: prog.type = prog
         val module = Module()
       }
-      genModule.emitModule("Module 1",new PrintWriter(System.out))
+      genModuleAMD.emitModuleAMD(new PrintWriter(System.out))
     }
-    assertFileEqualsCheck(prefix + "module-empty")
+    assertFileEqualsCheck(prefix + "moduleAMD-empty")
     
   }
 }

### 2013-07-15 core/src/test/scala/scala/js/TestModuleScala.scala

@@ -1,10 +1,10 @@
 package scala.js
 
 import java.io.PrintWriter
-import scala.js.gen.js.{GenEffect, GenNumericOps, GenModule}
-import scala.virtualization.lms.common.{NumericOps, NumericOpsExp}
+import scala.js.gen.scala.GenModule
+import scala.virtualization.lms.common.{NumericOps, NumericOpsExp, ScalaGenEffect, ScalaGenNumericOps}
 
-class TestModule extends FileDiffSuite {
+class TestModuleScala extends FileDiffSuite {
 
   trait Prog extends NumericOps {
 
@@ -22,8 +22,8 @@ class TestModule extends FileDiffSuite {
     val prog = new Prog with NumericOpsExp 
 
     val prefix = "test-out/"
-    withOutFile(prefix + "module") {
-      val genModule = new GenEffect with GenNumericOps with GenModule {
+    withOutFile(prefix + "module-scala") {
+      val genModule = new ScalaGenEffect with ScalaGenNumericOps with GenModule {
         val IR: prog.type = prog
         val module = Module(
             ("pack" -> Module("fun1" -> fun(prog.id _))),
@@ -34,21 +34,21 @@ class TestModule extends FileDiffSuite {
       }
       genModule.emitModule("Module 1",new PrintWriter(System.out))
     }
-    assertFileEqualsCheck(prefix + "module")
+    assertFileEqualsCheck(prefix + "module-scala")
   }
   
   def testEmitEmptyModule() = {
     val prog = new Prog with NumericOpsExp 
 
     val prefix = "test-out/"
-    withOutFile(prefix + "module-empty") {
-      val genModule = new GenEffect with GenNumericOps with GenModule {
+    withOutFile(prefix + "module-empty-scala") {
+      val genModule = new ScalaGenEffect with GenModule {
         val IR: prog.type = prog
         val module = Module()
       }
       genModule.emitModule("Module 1",new PrintWriter(System.out))
     }
-    assertFileEqualsCheck(prefix + "module-empty")
+    assertFileEqualsCheck(prefix + "module-empty-scala")
     
   }
 }

### 2013-07-15 examples/src/main/scala/scala/js/examples/HelloWorld.scala
Move examples in an sbt sub project

@@ -1,5 +1,4 @@
-package scala.js
-package examples
+package scala.js.examples
 
 import scala.js.language.JsScala
 import scala.js.exp.JsScalaExp

### 2013-07-15 project/Build.scala

@@ -11,14 +11,14 @@ object BuildSettings {
   )
 }
 
-object MyBuild extends Build {
+object JsScalaBuild extends Build {
   import BuildSettings._
 
   lazy val root = Project(
     "root",
-    file("core"),
+    file("."),
     settings = buildSettings
-  ) aggregate (macros, core)
+  ) aggregate (macros, core, examples)
 
   lazy val macros = Project(
     "macros",
@@ -38,7 +38,7 @@ object MyBuild extends Build {
     "core",
     file("core"),
     settings = buildSettings ++ Seq(
-      scalacOptions ++= Seq("-deprecation", "-unchecked", "-Xexperimental", "-P:continuations:enable", "-Yvirtualize", "-language:dynamics", "-Ymacro-debug-lite"),
+      scalacOptions ++= Seq("-deprecation", "-unchecked", "-Xexperimental", "-P:continuations:enable", "-Yvirtualize", "-language:dynamics"/*, "-Ymacro-debug-lite"*/),
       name := "js-scala",
       //Our tests are not threadsafe so disabling parallel execution for now
       parallelExecution in Test := false,
@@ -54,6 +54,17 @@ object MyBuild extends Build {
       libraryDependencies <<= (scalaVersion, libraryDependencies) { (ver, deps) => deps :+ compilerPlugin("org.scala-lang.virtualized.plugins" % "continuations" % ver)}
     )
   ) dependsOn (macros)
+
+  lazy val examples = Project(
+    "examples",
+    file("examples"),
+    settings = buildSettings ++ Seq(
+      autoCompilerPlugins := true,
+      libraryDependencies <<= (scalaVersion, libraryDependencies) { (ver, deps) => deps :+ compilerPlugin("org.scala-lang.virtualized.plugins" % "continuations" % ver)},
+      scalacOptions ++= Seq("-deprecation", "-unchecked", "-Xexperimental", "-P:continuations:enable", "-Yvirtualize", "-language:dynamics")
+    )
+  ) dependsOn (core)
+
 }
 
 

### 2013-07-15 papers/gpce2013/gpce2013.tex
update papers

@@
 \usepackage{graphicx}
+\usepackage{moresize}
 
@@
 
-\titlebanner{banner above paper title}        % These are ignored unless
-\preprintfooter{short description of paper}   % 'preprint' option specified.
+%\titlebanner{banner above paper title}        % These are ignored unless
+%\preprintfooter{short description of paper}   % 'preprint' option specified.
 
@@
 \begin{abstract}
-Writing large Web applications is known to be difficult. One challenge comes from the fact that the application's
-logic is scattered into heterogeneous clients and servers, making it difficult to share code between both sides or to
-move code from one side to the other. Another challenge is performance: while Web applications rely on ever more
-code on the client-side, they may run on smart phones with little hardware capabilities. These two challenges raise
-the following problem: how to benefit from high-level languages and libraries making code complexity easier to manage
-and abstracting over the clients and servers differences without trading this engineering comfort for performance?
-This article presents high-level abstractions defined as deep embedded DSLs in Scala, that can (1) generate efficient
-code leveraging the target platform characteristics, (2) be shared between clients and servers. Though code written
-with our DSL has a high level of abstraction our benchmark on a real world application reports that it runs as fast
-as hand tuned low-level JavaScript code.
+%\vspace{0.2cm}
+
+Writing large Web applications is known to be difficult. One challenge comes from the fact that the
+application's logic is scattered into heterogeneous clients and servers, making it difficult to
+share code between both sides or to move code from one side to the other. Another challenge is
+performance: while Web applications rely on ever more code on the client-side, they may run on smart
+phones with limited hardware capabilities. These two challenges raise the following problem: how to
+benefit from high-level languages and libraries making code complexity easier to manage and
+abstracting over the clients and servers differences without trading this engineering comfort for
+performance? This article presents high-level abstractions defined as deep embedded DSLs in Scala,
+that can (1) generate efficient code leveraging the target platform characteristics, (2) be shared
+between clients and servers. Though code written with our DSL has a high level of abstraction our
+benchmark on a real world application reports that it runs as fast as hand tuned low-level
+JavaScript code.
 \end{abstract}
@@
 
-Web applications are attractive because they require no installation or deployment steps on clients and enable large
-scale collaborative experiences. However, writing large Web applications is known to be
-difficult~\cite{Mikkonen08_SpaghettiJs,Preciado05_RIAMethodologyNecessity}. One challenge comes from the fact
-that the business logic is scattered into heterogeneous client-side and server-side
-environments~\cite{Echeverria09_RIA,Kuuskeri09_PartitioningClientServer}. This gives less flexibility in the
-engineering process and requires a higher maintenance effort: there is no way to move a piece of code targeting
-the server-side to target the client-side, the code has to be rewritten. Even worse, logic parts that run on both
-client-side and server-side need to be duplicated. For instance, HTML fragments may be built from the server-side
-when a page is requested by a client, but they may also be built from the client-side to perform an incremental
-update subsequent to an user action. How could developers write HTML fragment definitions once and render them on
-both client-side and server-side?
+%\vspace{0.2cm}
 
-The more interactive the application is, the more logic needs to be duplicated between the server-side and the
-client-side, and the higher is the complexity of the client-side code. Developers use libraries and frameworks to get
-high-level abstractions on client-side, making their code easier to reason about and to maintain, but also making
-their code run less efficiently (due to \emph{abstraction penalty}).
+Web applications are attractive because they require no installation or deployment steps on clients
+and enable large scale collaborative experiences. However, writing large Web applications is known
+to be difficult~\cite{Mikkonen08_SpaghettiJs,Preciado05_RIAMethodologyNecessity}. One challenge
+comes from the fact that the business logic is scattered into heterogeneous client-side and
+server-side environments~\cite{Echeverria09_RIA,Kuuskeri09_PartitioningClientServer}. This gives
+less flexibility in the engineering process and requires a higher maintenance effort: there is no
+way to move a piece of code targeting the server-side to target the client-side -- the code has to
+be rewritten. Even worse, logic parts that run on both client-side and server-side need to be
+duplicated. For instance, HTML fragments may be built from the server-side when a page is requested
+by a client, but they may also be built from the client-side to perform an incremental update
+subsequent to a user action. How could developers write HTML fragment definitions once and render
+them on both client-side and server-side?
 
-Performance is a primary concern in Web applications, because they are expected to run on a broad range of devices,
-from the powerful desktop personal computer to the less powerful smart phone. “Every 100~ms delay costs 1\% of
-sales”, said Amazon in 2006.
+The more interactive the application is, the more logic needs to be duplicated between the
+server-side and the client-side, and the higher is the complexity of the client-side code.
+Developers can use libraries and frameworks to get high-level abstractions on client-side, making
+their code easier to reason about and to maintain, but also making their code run less efficiently
+due to \emph{abstraction penalty}.
 
-Using the same programming language on both server-side and client-side could improve the software engineering
-process by enabling code reuse between both sides. Incidentally, the JavaScript language -- which is currently the
-most supported the action language on Web clients -- can be used on server-side, and an increasing number of
-programming languages or compiler back-ends can generate JavaScript code (\eg Java/GWT~\cite{Chaganti07_GWT},
+Performance is a primary concern in many Web applications, because they are expected to run on a
+broad range of devices, from the powerful desktop personal computer to the less powerful smart
+phone~\cite{Souders08_Perf,Huang10_Perf}.
+
+Using the same programming language on both server-side and client-side could improve the software
+engineering process by enabling code reuse between both sides. Incidentally, the JavaScript language
+-- which is currently the most supported action language on Web clients -- can be used on
+server-side. Conversely, an increasing number of programming languages or compiler back-ends can
+generate JavaScript code (\eg{} Java/GWT~\cite{Chaganti07_GWT},
 SharpKit\footnote{\href{http://sharpkit.net}{http://sharpkit.net}}, Dart~\cite{Griffith11_Dart},
 Kotlin\footnote{\href{http://kotlin.jetbrains.org/}{http://kotlin.jetbrains.org/}},
-ClojureScript~\cite{McGranaghan11_ClojureScript}, Fay\footnote{\href{http://fay-lang.org/}{http://fay-lang.org/}},
+ClojureScript~\cite{McGranaghan11_ClojureScript},
+Fay\footnote{\href{http://fay-lang.org/}{http://fay-lang.org/}},
 Haxe~\cite{Cannasse08_HaXe} or Opa\footnote{\href{http://opalang.org/}{http://opalang.org/}}).
 
-However, using the same programming language is not enough because the client and server programming environments
-are not the same. For instance, DOM fragments can be defined on client-side using the standard DOM API, but this
-API does not exist on server-side. How to define a common vocabulary for such concepts? And how to make the
-executable code leverage the native APIs, when possible, for performance reasons?
+However, using the same programming language is not enough because the client and server programming
+environments are not the same. For instance, DOM fragments can be defined on client-side using the
+standard DOM API, but this API does not exist on server-side. How to define a common vocabulary for
+such concepts? And how to make the executable code leverage the native APIs, when possible, for
+performance reasons?
 
-Generating efficient code for heterogeneous platforms is hard to achieve in an extensible way: the translation of
-common abstractions like collections into their native counterpart (JavaScript arrays on client-side and standard
-library's collections on server-side) may be hard-coded in the compiler, but that approach would not scale to handle
-all the abstractions a complete application may use (\eg HTML fragment definitions, form validation rules, or even
-some business data type that may be represented differently).
+Generating efficient code for heterogeneous platforms is hard to achieve in an extensible way: the
+translation of common abstractions like collections into their native counterpart (JavaScript arrays
+on client-side and standard library's collections on server-side) may be hard-coded in the compiler,
+but that approach would not scale to handle all the abstractions a complete application may use
+(\eg{} HTML fragment definitions, form validation rules, or even some business data type that may be
+represented differently).
 
-On one hand, for engineering reasons, developers want to write Web applications using a single high-level language,
-abstracting over the target platforms differences and reducing code complexity. But on the other hand, for
-performance reasons, they want to keep control on the way their code is compiled to each target platform. How to
-solve this dilemma?
+On one hand, for engineering reasons, developers want to write Web applications using a single
+high-level language, abstracting over the target platforms differences and reducing code complexity.
+But on the other hand, for performance reasons, they want to keep control on the way their code is
+compiled to each target platform. We propose to solve this dilemma by providing high-level
+abstractions in compiled domain-specific embedded languages (DSELs)~\cite{Hudak96_DSEL,
+Elliott2003_Compiling}. Compiled DSELs allow the definition of domain-specific languages (DSLs) as
+libraries on top of a host language, and to compile them to a target platform. Their deep embedding
+gives the opportunity to control the code generation scheme for a given abstraction and target
+platform.
 
-Compiled domain-specific embedded languages~\cite{Elliott2003_Compiling} allow the definition of domain-specific
-languages (DSLs) as libraries on top of a host language, and to compile them to a target platform. The deep embedding
-gives the opportunity to control the code generation scheme for a given abstraction and target platform.
-
-Kossakowski \etal introduced \emph{js-scala}, a compiled embedded DSL defined in Scala that generates JavaScript
-code, making it possible to write the client-side code of Web applications using
-JavaScript~\cite{Kossakowski12_JsDESL}. However, Kossakowski \etal did not address the engineering dilemma described
-above. This paper enriches js-scala with the following contributions\footnote{The code is available at
+Kossakowski \etal introduced \emph{js-scala}, a compiled embedded DSL defined in Scala that
+generates JavaScript code, making it possible to write the client-side code of Web applications
+using JavaScript~\cite{Kossakowski12_JsDESL}. However, Kossakowski \etal did not address the
+engineering dilemma described above. This paper enriches js-scala with the following
+contributions\footnote{The code is available at
 \href{http://github.com/js-scala}{http://github.com/js-scala}}:
@@
 \begin{itemize}
- \item we define high-level abstractions, typically used in Web programming, that generate low-level code leveraging
-native Web browsers APIs;
- \item in the case of abstractions shared between servers and clients, we specialize the code generation in order to
-leverage the target platform native environments.
+ \item we define high-level abstractions, typically used in Web programming, that generate low-level
+code leveraging native Web browser APIs;
+ \item in the case of abstractions shared between servers and clients, we specialize the code
+generation in order to leverage the target platform native environments.
 \end{itemize}
 
-We validate our approach with a case study implemented with various candidate technologies and discuss the relative
-pro and cons of them. Though the code written in our DSL is high-level and can be shared between clients and servers,
-it has the same runtime performances on client-side as hand-tuned low-level JavaScript code.
+We validate our approach with a case study implemented with various candidate technologies and
+discuss the relative pro and cons of them. Though the code written in our DSL is high-level and can
+be shared between clients and servers, it has the same runtime performances on client-side as
+hand-tuned low-level JavaScript code.
 
-The remainder of this paper is organized as follows. The next section introduces existing approaches for defining
-cross-compiling languages and presents the framework we used to define our DSLs. Sections \ref{contribution} and
-\ref{contribution-shared} present our contribution. Section \ref{validation} compares our solution to common
-approaches. Section \ref{discussion} concludes.
+The remainder of this paper is organized as follows. The next section overviews the
+existing approaches defining high-level languages for Web programming. Section \ref{sec:lms}
+presents the framework we used to define our DSLs. Section \ref{sec:contribution} presents our
+contribution. Section \ref{sec:validation} compares our solution to common approaches. Section
+\ref{sec:discussion} discusses our results and section \ref{sec:conclusion} concludes.
 
-\section{Background}
+\section{Related Work}
+\label{sec:related-work}
 
-\subsection{Language Engineering Processes}
+%\vspace{0.1cm}
 
-This section presents different approaches for defining cross-platform programming languages.
+We classified existing approaches providing high-level abstractions for Web programming in four
+categories, as shown in figure \ref{fig:langs}.
 
-\begin{figure}
+\begin{figure}[htb]
 \begin{center}
-\includegraphics[width=6cm]{langs.pdf}
+\includegraphics[width=7.5cm]{langs.pdf}
 \end{center}
 \caption{Language engineering processes}
-\label{langs}
+\vspace{-0.5cm}
+\label{fig:langs}
 \end{figure}
@@
 
-The first approach for defining a cross-platform language consists in hard-coding, in the compiler, the
-code generation scheme of each language feature to each target platform. Figure \ref{langs} (a) depicts this
-process. In order to support a feature related to a specific domain, the whole compiler pipeline (parser, code
-generator, \etc) may have to be adapted. This approach gives \emph{fat} languages because a lot of concepts are
-defined at the language level: general programming concepts such as naming, functions, classes, as well as more
-domain-specific concepts such as HTML fragment definition. Examples of such languages for Web programming are
-Links~\cite{Cooper07_Links}, Opa and Dart~\cite{Griffith11_Dart}. These languages are difficult to extend because
-each concept is defined in the compiler, and modifying a compiler requires a high effort. Furthermore, these
-languages also require to support common programming abstraction and composition mechanisms, as general purpose
-languages do. So they usually try to re-invent the features of general purpose languages, that’s why we argue that
-this approach for defining programming languages is difficult to scale: for every specific problem you would have to
-rewrite a full-featured programming language besides addressing the concepts specific to the problem domain.
+\enlargethispage{0.3cm}
+%\vspace{0.1cm}
+
+The first approach for defining a cross-platform language consists in hard-coding, in the compiler,
+the code generation scheme of each language feature to each target platform. Figure \ref{fig:langs}
+(a) depicts this process. In order to support a feature related to a specific domain, the whole
+compiler pipeline (parser, code generator, \etc) may have to be adapted. This approach gives
+\emph{fat} languages because a lot of concepts are defined at the language level: general
+programming concepts such as naming, functions, classes, as well as more domain-specific concepts
+such as HTML fragment definition. Thus, implementing a fat language may require a high effort.
+Examples of such languages for Web programming are Links~\cite{Cooper07_Links}, Opa and
+Dart~\cite{Griffith11_Dart}.
 
@@
 
-Another approach consists in defining several independent domain-specific languages~\cite{Van00_DSL}, each one
-focusing on concerns specific to a given problem domain, and then to combine all the source artifacts written with
-these language into one executable program, as shown in figure~\ref{langs} (b). Defining such languages requires a
-minimal effort compared to the previous approach because each language has a limited set of features. On the other
-hand, it is difficult to have interoperability between DSLs (missing ref). \cite{Groenewegen08_WebDSL} gave an
-example of such a domain-specific language for defining Web applications.
+%\vspace{0.1cm}
+
+Another approach consists in defining several independent domain-specific
+languages~\cite{Van00_DSL}, each one focusing on concerns specific to a given problem domain, and
+then to combine all the source artifacts written with these language into one executable program, as
+shown in figure~\ref{fig:langs} (b). Defining such languages requires a minimal effort compared to
+the previous approach because each language has a limited set of features. On the other hand, it is
+difficult to have interoperability between DSLs. \cite{Groenewegen08_WebDSL} gave an example of such
+a domain-specific language for defining Web applications.
 
@@
 
-Alternatively, one can define concepts relative to a specific domain as a library on top of a thin general purpose
-language (it is also referred to as a domain-specific \emph{embedded} language~\cite{Hudak96_DSEL}). Figure
-\ref{langs} (c) depicts this approach. The general purpose language is used as a host language and does not need to
-be modified if a new concept is introduced, because concepts are defined as pure libraries. On the other hand, the
-syntax of the DSLs is limited by the syntax flexibility of the host language. Furthermore, this approach gives no
-opportunity to translate a concept efficiently according to the target platform characteristics because concepts are
-defined as libraries and are translated by the compiler, which has no domain-specific knowledge. Examples of
-languages following this approach are Java/GWT, Kotlin, HaXe and SharpKit.
+\enlargethispage{0.3cm}
+
+%\vspace{0.1cm}
+
+Alternatively, one can define concepts relative to a specific domain as a library on top of a thin
+general purpose language (it is also referred to as a domain-specific \emph{embedded}
+language~\cite{Hudak96_DSEL}). Figure \ref{fig:langs} (c) depicts this approach. Defining such a
+library requires minimal effort (though the syntax of the DSL is limited by the syntax flexibility
+of the host language) and several DSLs can interoperate freely within the host language. However,
+this approach gives no opportunity to efficiently translate a concept according to the target
+platform characteristics because the compiler has no domain-specific knowledge (though some
+compilers hard-code the translation of some common abstractions such as arrays to leverage the
+target platform characteristics). Examples of languages following this approach are Java/GWT,
+Kotlin, HaXe and SharpKit.
 
@@
 
-The last approach, shown in figure \ref{langs} (d), can be seen as a middle-ground between the two previous
-approaches: DSLs are embedded in a host language but use a code generation process. This approach share the same
-benefits and limits as embedded DSLs for defining language units. The code generation process is specific to each DSL
-and gives the opportunity to perform domain-specific optimizations. In other words deeply embedded DSLs bring
-domain-specific knowledge to the compiler.
+%\vspace{0.1cm}
+The last approach, shown in figure \ref{fig:langs} (d), can be seen as a middle-ground between the
+two previous approaches: DSLs are embedded in a host language but use a code generation process.
+This approach shares the same benefits and limitations as embedded DSLs for defining language units.
+However, the code generation process is specific to each DSL and gives the opportunity to perform
+domain-specific optimizations. In other words deeply embedded DSLs bring domain-specific knowledge
+to the compiler. Js-scala~\cite{Kossakowski12_JsDESL} is an example of deeply embedded DSL in Scala
+for Web programming. It makes it possible to produce JavaScript programs from Scala code that uses
+basic language concepts like arrays and control structures (\code{if} and \code{while}) as well as
+mechanisms specific to the Scala compiler, like delimited continuations to handle asynchronous
+computations. Though the initial authors of js-scala showed how to use the native browser APIs in an
+untyped fashion, they did not show how to build high-level abstractions for Web programming
+generating efficient code leveraging the browser APIs and did not show how to share such
+abstractions between server and client sides (they achieved code sharing, yet on the server-side
+their DSL was not compiled but interpreted). Our paper addresses these limitations.
 
-(Summary explaining why we choose the deep embedded DSLs approach)
+\section{Lightweight Modular Staging}
+\label{sec:lms}
 
-\subsection{Lightweight Modular Staging}
+This section gives background material on the framework used to define js-scala.
 
-Lightweight Modular Staging~\cite{Rompf12_LMSThesis, Rompf12_LMS} (LMS) is a framework for defining deeply embedded
-DSLs in Scala. It has been used to define high-performance DSLs for parallel computing~\cite{Brown11_Parallel} and to
-define JavaScript as an embedded DSL in Scala~\cite{Kossakowski12_JsDESL}.
+Lightweight Modular Staging~\cite{Rompf12_LMSThesis, Rompf12_LMS} (LMS) is a framework for defining
+deeply embedded DSLs in Scala. It has been used to define high-performance DSLs for parallel
+computing~\cite{Brown11_Parallel} and to define JavaScript as an embedded DSL in
+Scala~\cite{Kossakowski12_JsDESL}.
 
-LMS is based on staging~\cite{Jorring1986_Staging}: a program using LMS is a regular Scala program that evaluates
-to an intermediate representation (IR) of a final program. This IR is a graph of expressions that can be traversed
-by code generators to produce the final program code. Expressions evaluated in the initial program and those
-evaluated in the final program (namely, staged expressions) are distinguished by their type: a \code{Rep[Int]} value
-in the initial program is a staged expression that generates code evaluating to an \code{Int} value in the final
-program. An \code{Int} computation in the initial program is evaluated during the initial program evaluation and
+LMS is based on staging~\cite{Jorring1986_Staging}: a program using LMS is a regular Scala program
+that evaluates to an intermediate representation (IR) of a final program. This IR is a graph of
+expressions that can be traversed by code generators to produce the final program code. Expressions
+evaluated in the initial program and those evaluated in the final program (namely, staged
+expressions) are distinguished by their type: a \code{Rep[Int]} value in the initial program is a
+staged expression that generates code evaluating to an \code{Int} value in the final program. An
+\code{Int} computation in the initial program is evaluated during the initial program evaluation and
 becomes a constant in the final program.
@@
 
-\section{High-Level Abstraction for Client-Side Code: Selectors}
-\label{contribution}
+\section{Efficient High-Level Abstractions for Web Programming}
+\label{sec:contribution}
 
-In a Web application, the user interface is defined by a HTML document that can be updated by the JavaScript code.
-A typical operation consists in searching some “interesting” element in the document, in order to extract its
-content, replace it or listen to user events triggered on it (such as mouse clicks). The standard API provides
-several functions to search elements in a HTML document according to their name or attribute values.
-Figure~\ref{selectors-api} summarizes the available functions and their differences.
+This section presents some tasks typically performed in Web applications, either on client-side or
+server-side or on both, generalizes them in terms of high-level abstractions, and shows how they are
+implemented in js-scala to generate efficient code.
 
-\begin{figure}
+\subsection{Selectors API}
+\label{sec:selectors}
+
+In a Web application, the user interface is defined by a HTML document that can be updated by the
+JavaScript code. A typical operation consists in searching some “interesting” element in the
+document, in order to extract its content, replace it or listen to user events triggered on it (such
+as mouse clicks). The standard API provides several functions to search elements in a HTML document
+according to their name or attribute values. Figure~\ref{selectors-api} summarizes the available
+functions and their differences.
+
+\begin{figure}[htb]
 \begin{center}
@@
 \end{center}
-\caption{Standard selectors API}
+\caption{Standard selectors API. The \code{querySelector} and \code{querySelectorAll} are the most
+general functions while the others handle special cases.}
 \label{selectors-api}
@@
 
-The \code{querySelector} and \code{querySelectorAll} are the most general functions while the others handle special
-cases. For the developer it is not convenient to have to master several functions performing similar tasks. In fact,
-most JavaScript developers use the jQuery
-library~\cite{Bibeault08_jQuery}\footnote{According to \href{http://trends.builtwith.com/javascript}{
-http://trends.builtwith.com/javascript}, jQuery is used by more than
-40\% of the top million sites} that provides only one high-level function to search for elements.
-Listings \ref{vanilla-selectors} and \ref{jquery-selectors} show two equivalent JavaScript programs performing
-element searches, the first one using the native APIs and the second one using jQuery.
-
-\begin{figure}
-\begin{lstlisting}[language=JavaScript,label=vanilla-selectors,caption=Searching elements in plain JavaScript]
+\begin{figure}[htb]
+\begin{lstlisting}[language=JavaScript,label=vanilla-selectors,caption=Searching elements using the
+native selectors API]
 function getWords() {
@@
 
-\begin{figure}
+Listing \ref{vanilla-selectors} gives an example of use of various functions from the native
+selectors API to retrieve a list of input fields within a form. The existence of several
+specialized functions in the API makes it possible to write efficient code, but forces users to
+think at a low abstraction level.
+
+\begin{figure}[htb]
 \begin{lstlisting}[language=JavaScript,label=jquery-selectors,caption=Searching elements in jQuery]
@@
 
-jQuery provides an API that is simpler to master because it has less functions, but by doing so it can not benefit
-from the performance of the browser’s implementation of specialized search functions (\code{getElementById},
-\code{getElementsByTagName} and \code{getElementsByClassName}).
+A high-level abstraction for searching elements in a document could be just one function
+finding all elements matching a given CSS selector. In fact, most JavaScript
+developers\footnote{According to
+\href{http://trends.builtwith.com/javascript}{http://trends.builtwith.com/javascript}, jQuery is
+used by more than 40\% of the top million sites.} use the jQuery library~\cite{Bibeault08_jQuery}
+that actually provides only one function to search for elements. Listing \ref{jquery-selectors}
+shows an equivalent JavaScript program as listing \ref{vanilla-selectors}, but using jQuery. The
+code is both shorter and simpler, thanks to its higher-level of abstraction. jQuery provides an API
+that is simpler to master because it has less functions, but this benefit comes at the price of a
+decrease in runtime performances.
 
-\begin{figure}
+Instead, we propose a solution that has a high-level API but generates JavaScript code using the
+specialized native API, when possible, in order to get both engineering comfort and performances.
+We achieve this by analyzing, during the first evaluation step, the selector that is
+passed as parameter and, when appropriate, by producing JavaScript code using the specialized API,
+and otherwise producing code using \code{querySelector} and \code{querySelectorAll}.
+
+\begin{figure}[htb]
+\begin{lstlisting}[label=selector-impl,caption=Selectors optimization]
+def find(selector: Rep[String]) =
+  getConstIdCss(selector) match {
+    case Some(id) if receiver == document =>
+      DocumentGetElementById(Const(id))
+    case _ =>
+      SelectorFind(receiver, selector)
+  }
+\end{lstlisting}
+\end{figure}
+
+Our API has two functions: \code{find} to find the first element matching a selector and
+\code{findAll} to find all the matching elements. Listing \ref{selector-impl} gives the
+implementation of the \code{find} function. It is a Scala function that returns an IR node
+representing the JavaScript computation that will search the element in the final program. The
+\code{getConstIdCss} function analyzes the selector: if it is a constant \code{String} value
+containing a CSS ID selector, it returns the value of the identifier. So, if the \code{find}
+function is applied to the \code{document} and to an ID selector, it returns a
+\code{DocumentGetElementById} IR node (that is translated to a \code{document.getElementById} call
+by the code generator), otherwise it returns a \code{SelectorFind} IR node (that is translated to a
+\code{querySelector} call).
+
+The same applies to the implementation of \code{findAll}: the selector passed as parameter is
+analyzed and the function returns a \code{SelectorGetElementsByClassName} in case of a CSS class
+name selector, a \code{SelectorGetElementsByTagName} in case of a CSS tag name selector, and a
+\code{SelectorFindAll} otherwise.
+
+\begin{figure}[htb]
 \begin{lstlisting}[label=js-scala-selectors,caption=Searching elements in js-scala]
@@
 
-Listing~\ref{js-scala-selectors} shows how to implement listing~\ref{jquery-selectors} using js-scala. We
-provide two functions for searching elements: \code{find} to find the first element matching a selector and
-\code{findAll} to find all the matching elements. During the first evaluation step, these functions try to analyze
-the selector that is passed as parameter and, when appropriate, produce code using the specialized  API, otherwise
-they produce  code using \code{querySelector} and \code{querySelectorAll}. As a result,
-listing~\ref{js-scala-selectors} generates a JavaScript program identical to listing~\ref{vanilla-selectors}: the
-high-level abstraction (the \code{find} and \code{findAll} functions) exist only in the initial program, not in the
-final JavaScript program.
-
-\begin{figure}
-\begin{lstlisting}[label=selector-impl,caption=Selectors optimization]
-def find(receiver: Rep[Selector],
-         selector: Rep[String]) =
-  getConstIdCss(selector) match {
-    case Some(id) if receiver == document =>
-      DocumentGetElementById(Const(id))
-    case _ =>
-      SelectorFind(receiver, selector)
-  }
-\end{lstlisting}
+\begin{figure}[htb]
+\begin{center}
+\includegraphics[width=5cm]{ir.pdf}
+\end{center}
+\caption{Intermediate representations returned by the evaluation of (a)
+\code{document.find("\#add-user button")} and (b) \code{document.find("\#add-user")}}
+\label{fig:ir}
 \end{figure}
 
-Listing \ref{selector-impl} shows the implementation of the \code{find} function producing different IR nodes
-according to the selector passed as parameter. The \code{getConstIdCss} function analyzes the selector: if it
-is a constant \code{String} value containing a CSS ID selector, it returns the value of the identifier. If the
-\code{find} function is applied to the \code{document} and to an ID selector, it returns a
-\code{DocumentGetElementById} IR node (that is translated to a \code{document.getElementById} call by the code
-generator), otherwise it returns a \code{SelectorFind} IR node (that is translated to a \code{querySelector}
-call). (FIXME Add a ref to the spec of CSS selectors?)
+\vspace{1 cm}
 
-The same applies to the implementation of \code{findAll}: the selector passed as parameter is analyzed and the
-function returns a \code{SelectorGetElementsByClassName} in case of a CSS class name selector, a
-\code{SelectorGetElementsByTagName} in case of a CSS tag name selector, and a \code{SelectorFindAll} otherwise.
+Figure \ref{fig:ir} shows the IRs returned by the evaluation of \small{
+$document\\.find("\#add-user button")$} \normalsize{and}  \small{$ document.find("\#add-user")$}.
 
-\section{High-Level Abstractions Shared on Clients and Servers}
-\label{contribution-shared}
+\normalsize{}
+
+\sloppy
+In the former case, the selector is parsed and does not match an ID selector (it is a composite
+selector matching button elements within the element having the \code{add-user} id), so a
+\code{SelectorFind} node is returned, then translated into a call to the general
+\code{querySelector} function. In the latter case, the selector matches an ID selector so a
+\code{DocumentGetElementById} node is returned, then translated into a call to the specialized
+\code{getElementById} function.
+
+Finally, listing~\ref{js-scala-selectors} shows how to implement listing~\ref{jquery-selectors} in
+Scala using js-scala. The code has the same abstraction level as with jQuery, however it
+generates a JavaScript program identical to listing~\ref{vanilla-selectors}: the high-level
+abstractions (the \code{find} and \code{findAll} functions) exist only in the initial program, not
+in the final JavaScript program.
 
 \subsection{Monads Sequencing}
+\label{sec:option}
 
-Null references are a known source of problems in programming languages~\cite{Hoare09_Null,Nanda09_Null}. For
-example, consider listing \ref{null-unsafe} finding a particular widget in the page and then a particular
-button within the widget. The native \code{querySelector} method returns \code{null} if no node matched the given
-selector in the document. If we run this code in a page where the widget is not present, it will throw an error
-and stop further JavaScript execution. Defensive code can be written to handle \code{null} references, but leads to
-very cumbersome code, as shown in listing \ref{null-defensive}.
+This section presents an abstraction to handle \code{null} references and shows how this
+abstraction can be shared between client and server code.
 
-\begin{figure}
+\code{null} references are a known source of problems in programming
+languages~\cite{Hoare09_Null,Nanda09_Null}. For example, consider listing \ref{null-unsafe} finding
+a particular widget in the page and then a particular button within the widget. The native
+\code{querySelector} method returns \code{null} if no node matched the given selector in the
+document. If we run this code in a page where the widget is not present, it will throw an error
+and stop further JavaScript execution. Defensive code can be written to handle \code{null}
+references, but leads to very cumbersome code, as shown in listing \ref{null-defensive}.
+
+\begin{figure}[htb]
 \begin{lstlisting}[language=JavaScript,label=null-unsafe,caption=Unsafe code]
@@
 
-\begin{figure}
-\begin{lstlisting}[language=JavaScript,label=null-defensive,caption=Defensive programming to handle null references]
+\begin{figure}[htb]
+\begin{lstlisting}[language=JavaScript,label=null-defensive,caption=Defensive programming to handle
+null references]
 var loginWidget =
@@
 
-Some programming languages encode optional values with a monad (\eg \code{Maybe} in Haskell and \code{Option} in
-Scala). In that case, sequencing over the monad encodes optional value dereferencing. If the language supports a
-convenient syntax for monad sequencing, it brings a convenient syntax for optional value dereferencing, alleviating
-developers from the burden of defensive programming.
+Some programming languages encode optional values with a monad (\eg{} \code{Maybe} in Haskell and
+\code{Option} in Scala). In that case, sequencing over the monad encodes optional value
+dereferencing. If the language supports a convenient syntax for monad sequencing, it brings a
+convenient syntax for optional value dereferencing, alleviating developers from the burden of
+defensive programming.
 
-Listing \ref{null-js-scala} implements in js-scala a program equivalent to listing \ref{null-defensive}. The
-\code{for} notation is used to sequence computations over optional values that are encoded with a monad. The
-\code{find} function returns a \code{Rep[Option[Element]]} value, that can either be a \code{Rep[Some[Element]]} (if
-an element was found) or a \code{Rep[None.type]} (if no element was found). The \code{for} expression contains a
-sequence of statements that are executed in order, as long as the previous statement returned a
-\code{Rep[Some[Element]]} value.
+In our DSL, we encode an optional value of type \code{Rep[A]} using a \code{Rep[Option[A]]} value,
+which can either be a \code{Rep[Some[A]]} (if there is a value) or a \code{Rep[None.type]} (if
+there is no value). An optional value can be dereferenced using the \code{for} notation, as shown
+in listing \ref{null-js-scala}, that implements in js-scala a program equivalent to listing
+\ref{null-defensive}. The \code{find} function returns a \code{Rep[Option[Element]]}. The \code{for}
+expression contains a sequence of statements that are executed in order, as long as the previous
+statement returned a \code{Rep[Some[Element]]} value.
 
-\begin{figure}
+\begin{figure}[htb]
 \begin{lstlisting}[label=null-js-scala,caption=Handling null references in js-scala]
@@
 
-Such a monadic API brings both safety and expressiveness to developers manipulating optional values but usually
-involves the creation of an extra container object holding the optional value. In our case, the monadic API is
-used in the initial program but generates code that does not wrap values in container objects but instead checks if
-they are \code{null} or not when dereferenced. So the extra container object exists only in the initial program and
-is removed during code generation: listing \ref{null-js-scala} produces a code equivalent to listing
-\ref{null-defensive}.
+Such a monadic API brings both safety and expressiveness to developers manipulating optional values
+but usually involves the creation of an extra container object holding the optional value. In our
+case, the monadic API is used in the initial program but generates code that does not wrap values in
+container objects but instead checks if they are \code{null} or not when dereferenced. So the extra
+container object exists only in the initial program and is removed during code generation: listing
+\ref{null-js-scala} produces a code equivalent to listing~\ref{null-defensive}.
 
-\begin{figure}
-\begin{lstlisting}[caption=JavaScript code generator for null references handling DSL,label=option-codegen]
+\begin{figure}[htb]
+\begin{lstlisting}[caption=JavaScript code generator for null references handling
+DSL,label=option-codegen,captionpos=b]
 override def emitNode(sym: Sym[Any], rhs: Def[Any]) =
@@
 
-Listing \ref{option-codegen} shows the JavaScript code generator for methods \code{isEmpty} (that checks if the
-optional value contains a value) and \code{foreach} (that is called when the \code{for} notation is used, as in
-listing \ref{null-js-scala}). The \code{emitNode} method handles \code{OptionIsEmpty} and \code{OptionForeach} nodes
-returned by the implementations of \code{isEmpty} and \code{foreach}, respectively. In the case of the
-\code{OptionIsEmpty} node, it simply generates an expression testing if the value is \code{null}. In the case of the
-\code{OptionForeach} node, it wraps the code block dereferencing the value within a \code{if} checking that the value
-is not \code{null}.
+Listing \ref{option-codegen} shows the JavaScript code generator for methods \code{isEmpty} (that
+checks if the optional value contains a value) and \code{foreach} (that is called when the
+\code{for} notation is used, as in listing \ref{null-js-scala}). The \code{emitNode} method handles
+\code{OptionIsEmpty} and \code{OptionForeach} nodes returned by the implementations of
+\code{isEmpty} and \code{foreach}, respectively. In the case of the \code{OptionIsEmpty} node, it
+simply generates an expression testing if the value is \code{null}. In the case of the
+\code{OptionForeach} node, it wraps the code block dereferencing the value within a \code{if}
+checking that the value is not \code{null}.
 
-Last but not least, we make this abstraction available on server-side by writing a code generator similar to the
-JavaScript code generator, but targeting Scala (FIXME Show the code generator?). So the same abstraction is
-efficiently translated on both server and client sides.
+The IR nodes are not tied to the JavaScript code generator, so we are able to make this abstraction
+available on server-side by writing a code generator similar to the JavaScript code generator, but
+targeting Scala. So the same abstraction is efficiently translated on both server and client sides.
 
 \subsection{DOM Fragments Definition}
+\label{sec:forest}
 
-This section shows how we define an abstraction shared between clients and servers, as in the previous section, but
-that has different native counterparts on client and server sides. The challenge is to define an API providing a
-common vocabulary that generates code using the target platform native APIs.
+This section shows how we define an abstraction shared between clients and servers, as in the
+previous section, but that has different native counterparts on client and server sides. The
+challenge is to define an API providing a common vocabulary that generates code using the target
+platform native APIs.
 
-\begin{figure}
+\begin{figure}[htb]
 \begin{lstlisting}[language=JavaScript,caption=JavaScript DOM API,label=dom-api]
@@
 
-\begin{figure}
+\begin{figure}[htb]
 \begin{lstlisting}[caption=Scala XML API,label=scala-xml-api]
@@
 
-A common task in Web applications consists in computing HTML fragments representing a part of the page content. This
-task can be performed either from the server-side (to initially respond to a request) or from the client-side (to
-update the current page). As an example, listing \ref{dom-api} defines a JavaScript function \code{articleUi} that
-builds a DOM tree containing an article description, and listing \ref{scala-xml-api} shows how one could implement
-a similar function on server-side using the standard Scala XML library. The reader may notice that the client-side
-and server-side APIs are very different and that the client-side API is very low-level and inconvenient to use.
+A common task in Web applications consists in computing HTML fragments representing a part of the
+page content. This task can be performed either from the server-side (to initially respond to a
+request) or from the client-side (to update the current page). As an example, listing \ref{dom-api}
+defines a JavaScript function \code{articleUi} that builds a DOM tree containing an article
+description, and listing \ref{scala-xml-api} shows how one could implement a similar function on
+server-side using the standard Scala XML library. The reader may notice that the client-side
+and server-side APIs are very different and that the client-side API is very low-level and
+inconvenient to use.
 
-Instead, we provide a common high-level DSL for defining HTML fragments and we make this DSL generate code leveraging
-native environments. Listing \ref{forest} shows how to implement our example with our DSL. The \code{el} function
-defines an HTML element, eventually containing attributes and children elements. Any children of an element that is
-not an element itself is converted into a text node. The children elements of an element can also be obtained
-dynamically from a collection, as shown in listing \ref{forest-loops}.
+Our first step consists in capturing, in a high-level API, the concepts common to the JavaScript and
+Scala APIs. Though they are different, both APIs define HTML elements with attributes and content.
+We propose to have a function \code{el} to define an HTML element, eventually containing attributes
+and children elements. Any children of an element that is not an element itself is converted into a
+text node. Listing \ref{forest} shows how to implement our example with our DSL. The children
+elements of an element can also be obtained dynamically from a collection, as shown in listing
+\ref{forest-loops}.
 
-\begin{figure}
+\begin{figure}[htb]
 \begin{lstlisting}[label=forest,caption=DOM definition DSL]
@@
 
-\begin{figure}
+\begin{figure}[htb]
 \begin{lstlisting}[label=forest-loops,caption=Using loops]
@@
 
-The \code{el} function returns an \code{Element} IR node that is a tree composed of other \code{Element} nodes and
-\code{Text} nodes. This tree is traversed by the code generators to produce code building an equivalent DOM tree on
-client-side and code building an equivalent XML fragment on server-side. When the children of an element are constant
-values (as in listing \ref{forest}) rather than dynamically computed (as in listing \ref{forest-loops}), the code
-generators inline the loop that adds children to their parent, for better performance. As a result, listing
-\ref{forest} generates a code equivalent to listing \ref{dom-api} on client-side and equivalent to
-\ref{scala-xml-api} on server-side.
+The \code{el} function returns an \code{Element} IR node that is a tree composed of other
+\code{Element} and \code{Text} nodes. The JavaScript and Scala code generators traverse this tree
+and produce code building an equivalent DOM tree and XML fragment, respectively. When the children
+of an element are constant values (as in listing \ref{forest}) rather than dynamically computed (as
+in listing \ref{forest-loops}), the code generators unroll the loop that adds children to their
+parent, for better performance. As a result, listing \ref{forest} generates a code equivalent to
+listing \ref{dom-api} on client-side and equivalent to listing \ref{scala-xml-api} on server-side.
 
-\begin{figure}
-\begin{lstlisting}[language=JavaScript,label=js-gen-forest,caption=JavaScript code generator for the DOM fragment
-definition DSL]
+\begin{figure*}
+\begin{lstlisting}[language=JavaScript,label=js-gen-forest,caption=JavaScript code generator for the
+DOM fragment definition DSL]
 case Tag(name, children, attrs) =>
@@
 \end{lstlisting}
-\end{figure}
+\end{figure*}
 
-\begin{figure}
-\begin{lstlisting}[label=scala-gen-forest,caption=Scala code generator for the DOM fragment definition DSL]
+\begin{figure*}
+\begin{lstlisting}[label=scala-gen-forest,caption=Scala code generator for the DOM fragment
+definition DSL]
 case Tag(name, children, attrs) =>
@@
 \end{lstlisting}
+\end{figure*}
+
+Listings \ref{js-gen-forest} and \ref{scala-gen-forest} show the relevant parts of the code
+generators for this DSL. They basically follow the same pattern: they visit \code{Tag} and
+\code{Text} IR nodes and produce the corresponding elements in the target language.
+
+\section{Evaluation}
+\label{sec:validation}
+
+Our goal is to evaluate the level of abstraction provided by our solution and its performances, by
+comparing it with common approaches. We take the number of lines of code as an inverse approximation
+of the level of abstraction. We also evaluate the ability to share code between client and server
+sides.
+
+We realized two micro-benchmarks involving programs using the selectors DSL and the
+optional value DSL, and we benchmarked a real world program. In each case we have written several
+implementations of the program, using plain JavaScript, Java/GWT, HaXe and js-scala (in each case we
+tried to write the application in an idiomatic way). The performance benchmarks measured the
+execution time of the generated JavaScript code. The tests were executed on a DELL Latitude E6430
+laptop with 8 GB of RAM, on the Google Chrome v27 Web browser.
+
+All our charts show three kinds of measures: the first group is the speed execution in operations
+per second (higher is better), the second group is the number of lines of code (lower is better) and
+the last group is the execution speed to number of lines of code ratio (higher is better). We
+normalized the values so the three groups can be shown within a same chart without scale issue.
+
+\subsection{Micro-Benchmarks}
+
+The micro-benchmarks measure the performance of our implementation of the selectors and optional
+value abstractions\footnote{The source code of the benchmarks is available at
+\href{https://github.com/js-scala/js-scala/tree/master/papers/gpce2013/benchmarks}{
+https://github.com/js-scala/js-scala/tree/master/papers/gpce2013/benchmarks}}.
+
+\subsubsection{Selectors}
+
+We could not implement this abstraction in GWT or HaXe as efficiently as we did in js-scala because
+it relies on the staging mechanism: the best we could do in GWT or Haxe is to expose the native
+high-level API (\code{querySelector} and \code{querySelectorAll}). So we directly compared the
+execution time of the JavaScript code generated by listing \ref{js-scala-selectors} with a
+JavaScript program equivalent to listing \ref{vanilla-selectors} but using the high-level native
+API (\code{querySelector} and \code{querySelectorAll}) instead. The code was executed in a Web page
+containing a few elements: 4 \code{fieldset} elements, each containing 0 to 2 elements with class
+\code{word}.
+
+\begin{figure}
+\centering
+\includegraphics[width=8cm]{selectors-benchmark.png}
+\caption{Micro-benchmark on the selectors abstraction}
+\label{fig:selectors-benchmark}
 \end{figure}
 
-Listings \ref{js-gen-forest} and \ref{scala-gen-forest} show the relevant parts of the code generators for this DSL.
-They basically follow the same pattern: they visit \code{Tag} and \code{Text} IR nodes and produce the corresponding
-elements in the target language. (FIXME More details)
+Figure \ref{fig:selectors-benchmark} shows the benchmark results. The JavaScript-opt version is
+listing \ref{vanilla-selectors}, which uses low-level native APIs, the JavaScript version is the
+equivalent listing using the high-level native API, and the jQuery version is listing
+\ref{jquery-selectors}. The js-scala version is slightly slower than the JavaScript-opt (by 14\%),
+but is 2.88 times faster than the JavaScript version, and 28.6 times faster than the jQuery
+version. Finally, the js-scala version has a performance to lines of code ratio more than 1.72
+times higher than others.
 
-% \subsection{Ad-Hoc Polymorphism}
-% 
-% Because of the dynamically typed nature of JavaScript, when calling a function there is no proper way to select a
-% specialized implementation according to the function’s parameters types. JavaScript is only able to dispatch
-% according to a method receiver prototype, \eg{} if one writes \code{foo.bar()} the JavaScript runtime will look into
-% the prototype of the \code{foo} object for a property named \code{bar} and will call it. So, the only way to achieve
-% \emph{ad hoc} polymorphism on JavaScript objects consists in defining the polymorphic function on the prototypes of
-% the objects. However, modifying existing object prototypes is considered bad
-% practice~\cite{Zakas12_MaintainableJs}. Another way could consist in manually coding the dispatch logic, by
-% registering supported data types at the beginning of the program execution, as described in section 2.4.3
-% of~\cite{Abelson83_SICP}, but this solution is painful for developers and incurs a performance overhead.
-% 
-% We propose to achieve \emph{ad hoc} polymorphism using
-% typeclasses~\cite{Wadler89_AdhocPolymorphism,Odersky06_Typeclasses,Oliveira10_Typeclasses} so that it supports
-% retroactive extension without modifying objects prototypes. The dispatch logic is type-directed and performed by the
-% compiler, so there is no runtime overhead.
-% 
-% \begin{figure}
-% \begin{lstlisting}[label=polymorphism,caption=Ad hoc polymorphism using typeclasses]
-% // Interface
-% case class Show[A](show: Rep[A => Node])
-% 
-% // Polymorphic function
-% def listWidget[A](items: Rep[List[A]])
-%       (implicit A: Show[A]): Rep[Node] =
-%   el("ul")(
-%     for (item <- items) yield {
-%       el("li")(A.show(item))
-%     }
-%   )
-% 
-% // Type `User`
-% type User = Record {
-%   val name: String
-%   val age: Int
-% }
-% // Implementation of Show for a User
-% implicit val showUser = Show[User] { user =>
-%   el("span", "class"->"user")(
-%     user.name + "(" + user.age + " years)"
-%   )
-% }
-% 
-% // Main program
-% def main(users: Rep[List[User]]) = {
-%   document.body.append(listWidget(users))
-% }
-% \end{lstlisting}
-% \end{figure}
-% 
-% Listing \ref{polymorphism} demonstrates how to define a polymorphic \code{listWidget} function that returns a DOM
-% tree containing the representation of a list of items. The \code{Show[A]} typeclass defines how to produce a DOM tree
-% for a value of type \code{A}. It is used by the \code{listWidget} function to get the DOM fragments of the list
-% items. The listing shows how to reuse the same \code{listWidget} function to show a list of users and a list of
-% articles.
+\subsubsection{Optional Value}
 
-\section{Evaluation}
-\label{validation}
+We reimplemented the optional value abstraction in plain JavaScript, Java and HaXe and wrote a small
+program manipulating optional values. Listing \ref{lst:option-benchmark} shows the js-scala
+version of this program. The \code{maybe} function is a function partially defined on
+\code{Int} values.
 
-We measured the runtime performances of the DSL for optional values using a micro-benchmark.
+\begin{figure}
+\begin{lstlisting}[label=lst:option-benchmark,caption=Micro-benchmark code for the optional values
+abstraction]
+val maybe = fun { (x: Rep[Int]) =>
+  some(x + 1)
+}
 
-We also have written several implementations of a complete application using different approaches for the client and
-server sides, and compared the amount of code written, the runtime performances and the ability to modularize the
-code.
-
-Our goal was to evaluate the level of abstraction provided by each solution and their performances. We took the
-number of lines of code as a measure of the level of abstraction. We also measured the ability to share code between
-client and server sides.
-
-The tests were run on a DELL Latitude E6430 laptop with 8 GB of RAM, on the Google Chrome v27 Web browser.
-
-\subsection{Micro-Benchmark}
-
-We reimplemented the optional value abstraction and a same program using it in the following languages: JavaScript,
-HaXe and Java.
+def benchmark = for {
+  a <- maybe(0)
+  b <- maybe(a)
+  c <- maybe(b)
+  d <- maybe(c)
+} yield d
+\end{lstlisting}
+\end{figure}
 
@@
 
-Figure \ref{micro-benchmark} shows the micro-benchmark results. Js-scala is between 3 to 100 times faster than
-other approaches on our benchmark. Combined with the fact that it provides high-level abstractions, it has a
-performance / lines of code ratio more than 4 times higher than other approaches.
+Figure \ref{micro-benchmark} shows the benchmark results. The js-scala version of the program runs
+between 3 to 10 times faster than other approaches. This version also takes less lines of code than
+others (this result is almost due to the special \code{for} notation, that has no equivalent in
+other benchmarked languages). Finally, the js-scala program has a performance to lines of code ratio
+more than 4 times higher than others.
 
@@
 
-Chooze~\footnote{\href{http://chooze.herokuapp.com}{http://chooze.herokuapp.com}} is an existing complete
-application for making polls. It allows users to create a poll, define the choice alternatives, share the poll, vote
-and look at the results. It contains JavaScript code to handle the dynamic behavior of the application:
-double-posting prevention, dynamic form update and rich interaction with the document.
+Chooze~\footnote{Source code is available at
+\href{http://github.com/julienrf/chooze}{http://github.com/julienrf/chooze}} is an existing
+complete application for making polls. It allows users to create a poll, define the choice
+alternatives, share the poll, vote and look at the results. It contains JavaScript code to handle
+the dynamic behavior of the application: double-posting prevention, dynamic form update and rich
+interaction with the document.
 
-The application was initially written using jQuery. We rewrote it using several technologies for the client-side
-part: vanilla JavaScript (low-level code without third-party library), js-scala, GWT and HaXe. In each case we
-tried to write the application in an idiomatic way\footnote{Source code is available at
-\href{http://github.com/julienrf/chooze}{http://github.com/julienrf/chooze}}.
+The application was initially written using jQuery. We rewrote it in vanilla JavaScript (low-level
+hand-tuned code without third-party library), js-scala, GWT and HaXe.
 
@@
 
-The benchmark code simulates user actions on a Web page (2000 clicks on buttons, triggering a dynamic update of
-the page and involving the use of the optional value monad, the selectors API and the HTML fragment definition API).
-Figure \ref{benchmark} shows the benchmark results.
+The benchmark code simulates user actions on a Web page (2000 clicks on buttons, triggering a
+dynamic update of the page and involving the use of the optional value monad, the selectors API and
+the HTML fragment definition API). Figure \ref{benchmark} shows the benchmark results.
 
@@
 
-The runtime performances of the vanilla JavaScript, HaXe and js-scala versions are similar (though the js-scala
-version is slightly slower by 6\%). It is worth noting that the vanilla JavaScript and the HaXe versions use
-low-level code compared to js-scala, as shown in the second part of the figure (lines of code): the js-scala version
-needs only 74 lines of code while the vanilla JavaScript version needs 116 lines of code (57\% bigger) and the HaXe
-version needs 148 lines of code (100\% bigger). The jQuery JavaScript version, of which code is high-level (54 lines
-of code, 27\% less than js-scala) runs 10 times slower than the js-scala version.
+The runtime performances of the vanilla JavaScript, HaXe and js-scala versions are similar (though
+the js-scala version is slightly slower by 6\%). It is worth noting that the vanilla JavaScript and
+the HaXe versions use low-level code compared to js-scala, as shown in the middle of the figure
+(lines of code): the js-scala version needs only 74 lines of code while the vanilla JavaScript
+version needs 116 lines of code (57\% bigger) and the HaXe version needs 148 lines of code (100\%
+bigger). The jQuery JavaScript version, which code is high-level (54 lines of code, 27\% less
+than js-scala) runs 10 times slower than the js-scala version.
 
-The last part of the figure compares the runtime performance / lines of code ratio. Js-scala shows the best score,
-being 1.48 times better than the vanilla JavaScript version, 1.88 times better than the HaXe version, 3.45 times
-better than the GWT version and 7.82 times better than the jQuery JavaScript version.
+The last part of the figure compares the runtime performance to lines of code ratio. Js-scala shows
+the best score, being 1.48 times better than the vanilla JavaScript version, 1.88 times better than
+the HaXe version, 3.45 times better than the GWT version and 7.82 times better than the jQuery
+JavaScript version.
 
@@
 
-We were able to share HTML DOM fragment definitions between server-side and client-side in js-scala.
+We were able to share some DOM fragment definitions between server-side and client-side only in
+the js-scala version. In the GWT version we don’t have choice: dynamic DOM fragments are always
+built only on client-side (a practice that makes it more difficult to make the pages content
+crawlable by search engines and may increase the initial display time). In the other versions the
+code for building the DOM fragment is duplicated between client and server sides, representing 20
+lines of JavaScript code (17\% of the total) and 15 lines of HTML (5\% of the total) in the
+JavaScript version, and 19 lines of HaXe code (13\% of the total) and 15 lines of HTML (5\%
+of the total) in the HaXe version. In the js-scala version the DOM fragment definitions shared
+between clients and servers represent 22 lines of Scala code (30\% of the total) and save 15 lines
+of HTML (5\% of the total).
+
+\subsubsection{Threats to Validity}
+
+Our goal was to put the runtime performances in perspective with the level of abstraction. We are
+aware that the indicator we chose as an inverse approximation of the abstraction level, the number
+of lines of code, is not scientifically established and may be subject to discussion. However, we
+think it is a reasonable approximation in our case because all the candidate languages we use have
+a similar syntax, inherited from the C programming language.
+
+Another weakness of our validation may come from the fact that our application does not make a
+heavy use of client-side code and thus may not be representative of the way large Web applications
+are written. However, we think that a richer application would have more parts of code susceptible
+to be shared between client and server sides, thus giving even better results on the code reuse
+statistics.
+
+Finally, the GWT version may not have been written in an as idiomatic as possible way. 
+Indeed, we mainly catch the events directly on the HTML DOM, as we do in JavaScript, without
+reusing all the GWT widgets. We do not build the application as a blank page with a set of
+widgets. However, this way of developing using GWT has no impact on the performance and a minor
+impact on the number of lines of code.
+
+\section{Discussion}
+\label{sec:discussion}
+
+We implemented our solution as compiled embedded DSLs in Scala. Generating code from our DSLs is a
+two steps process: an initial Scala program first evaluates to an intermediate representation of
+the final program that is traversed by code generators to produce the final JavaScript code.
+Domain-specific optimizations can happen during the IR construction (as shown in section
+\ref{sec:selectors}) or during the code generation (as shown in section \ref{sec:option}).
+
+An important consequence of the implementation as compiled embedded DSLs is that defining a DSL
+that can be shared between server and client sides requires a low effort: compiled embedded DSLs are
+simply defined as libraries but let developers specialize the generated code according to each
+target platform (as shown in section \ref{sec:forest}).
+
+In other words, the compiled embedded DSL approach gives us a way to exploit the Scala host language
+to define high-level language units that integrate seamlessly together and bring domain-specific
+knowledge to the code generation scheme to produce efficient code for client and server sides.
+
+These characteristics allowed us to capture some Web programming patterns as high-level
+abstractions, making the code of our application simpler to reason about and making some parts of
+the code reusable between client and server sides, while keeping execution performances on
+client-side as high as if we used hand-tuned low-level JavaScript code.
 
 \section{Conclusion}
-\label{discussion}
+\label{sec:conclusion}
 
-We implemented a high-level language abstracting over client and server heterogeneity but producing efficient code.
-Generated code size?
+High-level abstractions for Web programming, which are useful to decrease the complexity of the code
+and to abstract over the differences between the client and server environments, must be implemented
+in a way to efficiently run on hardware with limited capabilities.
+
+In this paper we showed how to leverage staging to implement high-level abstractions for Web
+programming that are efficiently compiled for heterogeneous platforms such as Web clients and
+servers that differ in their technical API. We also showed how these abstractions can be shared
+between client and server sides.
+
+Our two kinds of benchmarks, (i) micro-benchmarks to evaluate one abstraction and (ii) a benchmark
+on a real application using these abstractions, show performances similar to hand-optimized
+low-level code.
+
+In a future work we may investigate more coarse-grained optimizations like smart DOM updates
+minimizing the number of browser reflows.
 
@@
 %
-\acks
-
-This work was funded by Zenexity.
+%\acks
+%
+%This work was funded by Zenexity.
 

### 2013-07-15 papers/sle2013/sle2013.tex
fx events examples
update papers

--- 2013-06-07 papers/sle2013/sle2013.tex
@@
 
-\title{Using Dependent Types to Build Safe and Concise JavaScript Foreign Function Interfaces}
-\titlerunning{Dependent Types Rock}
+\title{Using Path-Dependent Types to Build Safe JavaScript Foreign Function Interfaces}
+%\titlerunning{Dependent Types Rock}
 
@@
 \begin{abstract}
-Several programming languages can target JavaScript as a back-end, giving developers programming language features
-that are absent from JavaScript, such as static typing. However, the Web browser APIs, which are needed to interact
-with a Web page, are designed for JavaScript, making it challenging to expose them in a statically typed language.
-Indeed, existing statically typed languages exposing Web browser APIs either break type safety or give developers
-less control than if they were using JavaScript. This article shows how to expose Web browser APIs in Scala in a
-type safe way while keeping the same level of control as with native APIs by using dependent types and functional
-dependencies. We validate this approach in designing safe and concise foreign function interfaces between Scala and
-JavaScript for DOM events handling and DOM manipulation. We compared this approach to common frameworks such as GWT,
-Fay, Kotlin, Dart and SharpKit.
-\keywords{Dependent Types, JavaScript, Scala, Foreign Function Interface}
+Several programming languages can target JavaScript as a back-end, giving developers programming
+language features that are absent from JavaScript, such as static typing. However, the Web browser
+APIs, which are needed to interact with a Web page, are designed for JavaScript, making it
+challenging to expose them in a statically typed language. Indeed, existing statically typed
+languages exposing Web browser APIs either break type safety or give developers less control than if
+they were using JavaScript. This article shows how to expose Web browser APIs in Scala in a type
+safe way while keeping the same level of control as with native APIs by using path-dependent types
+and functional dependencies. We validate this approach in designing safe and concise foreign
+function interfaces between Scala and JavaScript for DOM events handling and DOM manipulation. We
+compared this approach to common frameworks such as GWT, Fay, Kotlin, Dart and SharpKit.
+\keywords{Path-Dependent Types, JavaScript, Scala, Foreign Function Interface}
 \end{abstract}
@@
 
-Web applications are attractive because they require no installation or deployment step on clients and enable large
-scale collaborative experiences. However, writing large Web applications is known to be
-difficult~\cite{Mikkonen08_SpaghettiJs,Preciado05_RIAMethodologyNecessity}. One challenge comes from the fact that
-the JavaScript programming language -- which is currently the only action language natively supported by almost all
-Web clients -- lacks of constructs making large code bases maintainable (\eg static typing, first-class modules).
+Web applications are attractive because they require no installation or deployment step on clients
+and enable large scale collaborative experiences. Besides, now they can be executed
+efficiently~\cite{Gal:2009:TJT:1542476.1542528} on top of modern web-browsers. However, writing
+large Web applications is known to be
+difficult~\cite{Mikkonen08_SpaghettiJs,Preciado05_RIAMethodologyNecessity}. One challenge comes from
+the fact that the JavaScript programming language -- which is currently the only action language
+natively supported by almost all Web clients -- lacks of constructs making large code bases
+maintainable (\eg static typing, first-class modules).
 
 One solution consists in considering JavaScript as an assembly language\footnote{\cf
-\href{http://asmjs.org/}{http://asmjs.org/}} and generating JavaScript from compilers of full-featured and
-cutting-edge programming languages. Incidentally, an increasing number of programming languages or compiler backends
-can generate JavaScript code (\eg Java/GWT~\cite{Chaganti07_GWT},
+\href{http://asmjs.org/}{http://asmjs.org/}} and generating JavaScript from compilers of
+full-featured and cutting-edge programming languages. Incidentally, an increasing number of
+programming languages or compiler back-ends can generate JavaScript code (\eg
+Java/GWT~\cite{Chaganti07_GWT},
 SharpKit\footnote{\href{http://sharpkit.net}{http://sharpkit.net}}, Dart~\cite{Griffith11_Dart},
 Kotlin\footnote{\href{http://kotlin.jetbrains.org/}{http://kotlin.jetbrains.org/}},
-ClojureScript~\cite{McGranaghan11_ClojureScript}, Fay\footnote{\href{http://fay-lang.org/}{http://fay-lang.org/}},
-Haxe~\cite{Cannasse08_HaXe}, Opa\footnote{\href{http://opalang.org/}{http://opalang.org/}}).
+ClojureScript~\cite{McGranaghan11_ClojureScript},
+Fay\footnote{\href{http://fay-lang.org/}{http://fay-lang.org/}}, Haxe~\cite{Cannasse08_HaXe},
+Opa\footnote{\href{http://opalang.org/}{http://opalang.org/}}). However, compiling to JavaScript is
+not enough. Developers also need the Web browser programming environment: they need to interact with
+the Web page, to build DOM fragments, to listen to user events, \etc. A Foreign Function
+Interface (FFI) mechanism could be used to make browser’s APIs available to the developers. However,
+JavaScript APIs are not statically typed and make a heavy use of overloading, making them hard to
+expose in a statically typed language.
 
-However, compiling to JavaScript is not enough. Developers also need the Web browser programming environment: they
-need to interact with the Web page, to build DOM fragments, to listen to user events, \etc. A Foreign Function
-Interface mechanism could be used to make browser’s APIs available to the developers. However, JavaScript APIs are
-not statically typed and make a heavy use of overloading, making them hard to expose in a statically typed language.
+Indeed, existing statically typed languages compiling to JavaScript often expose weaker types than
+they should. For instance, the function \code{createElement} is polymorphic in its return type: it
+can return a \code{DivElement} as well as an \code{InputElement}, among others, but the Dart, Fay,
+SharpKit and Kotlin APIs return the super-type of all the possible values, namely the \code{Element}
+type. As a consequence, developers need to explicitly down-cast the value they get, which is a
+tedious and error prone task.
 
-Indeed, existing statically typed languages compiling to JavaScript often expose weaker types than they should. For
-instance, the function \code{createElement} is polymorphic in its return type: it can return a \code{DivElement} as
-well as an \code{InputElement}, among others, but the Dart, Fay, SharpKit and Kotlin APIs return the super-type of
-all the possible values, namely the \code{Element} type. As a consequence, developers need to explicitly down-cast
-the value they get, which is a tedious and error prone task.
+Some other languages try to workaround this problem by using overloading instead of polymorphism.
+For instance, HaXe provides functions \code{createDivElement}, \code{createInputElement}, which
+return a \code{DivElement} and an \code{InputElement}, respectively. Besides requiring a higher
+effort to implement, this solution also reduces the control level of users: by being statically
+resolved, the element type can not anymore be passed as a parameter.
 
-Some other languages try to workaround this problem by using overloading instead of polymorphism. For instance, HaXe
-provides functions \code{createDivElement}, \code{createInputElement}, which return a \code{DivElement} and an
-\code{InputElement}, respectively. Besides requiring a higher effort to implement, this solution also reduces the
-control level of users: by being statically resolved, the element type can not anymore be passed as a parameter.
+It turns out that most of the existing statically typed languages compiling to JavaScript either
+loose control or loose type safety when they expose Web browser’s APIs. How to give developers the
+same level of control as if they were using the native Web APIs, but in a statically typed and
+convenient way?
 
-It turns out that most of the existing statically typed languages compiling to JavaScript either loose control or
-loose type safety when they expose Web browser’s APIs. How to give developers the same level of control as if they
-were using the native Web APIs, but in a statically typed and convenient way?
+In this paper we present several ways to integrate Web browser’s APIs as statically typed APIs that
+are safe and give developers the same control level as if they were using the native APIs. We
+achieve this by using advanced features of type systems like dependent types and functional
+dependencies. We validate this approach in designing safe FFI between Scala and JavaScript for DOM
+events handling and DOM manipulation. We compared this approach to common frameworks such as GWT,
+Fay, Kotlin, Dart and SharpKit.
 
-In this paper we present several ways to integrate Web browser’s APIs as statically typed APIs that are safe and
-give developers the same control level as if they were using the native APIs. We can achieve this by using advanced
-features of type systems like dependent types and functional dependencies.
+The outline of this paper is the following.
+Section~\ref{sec:background} presents several motivating example and gives a background about
+path-dependent type.
+Section~\ref{sec:jsscala} presents %...
+Section~\ref{sec:evaluation} validates our ideas through the design of FFI for DOM events handling
+and DOM manipulation in Scala and compare these FFIs w.r.t. the existing FFIs for these features
+within frameworks such as GWT, Fay, Kotlin, Dart and SharpKit.
+Section~\ref{sec:related} discusses related work, Section~\ref{sec:conclusion} concludes this paper
+and highlights  future work.
 
-\section{Motivating Example}
+\section{Background Material}
 
-Typical tasks involved in Web applications.
+\subsection{Foreign Function Interface}
 
-Why is it difficult to type Web browser’s APIs?
+A FFI is a mechanism that allows a programmer to link one programming language program to programs
+written in another language. Such mechanisms exist in lots of languages: Ada, Java, Haskell, Lisp or
+GWT. Despite the name, FFIs are not necessarily restricted to function calls; many FFIs permit
+method calls on objects; and some even permit migration of non-trivial datatypes and/or objects
+across the language boundary. Nevertheless, the challenging tasks come when you want to: (i) manage
+shared garbage collection policies, (ii) handle the conversion of complex datatype to map from one
+environment to another and (iii) handle the design of safe foreign function interfaces regarding
+typing system constraints that can differ from one environment to another. This last issue is the
+one which is targeted in this paper for the design of FFI between a statically typed language such
+as Scala and a dynamically typed language such as JavaScript.
 
-\section{Lightweight Modular Staging}
+\subsection{Path-Dependent Type Overview}
+
+A type system provides us a near mathematical proof that programs that compile successfully obey a
+set of rules and restrictions and, as such, are bug-free in that respect. For example, after
+compiling a Java program, developer knows that it will not do a division of a number with a
+string and that, if our method foo receives an integer as argument, it will not be called anywhere
+with an argument of any type other than an integer.
+Among all the type theory,  dependent types is a term that describes using values as types, the type depends on a value.
+This means you can encode some properties of an object in its type.
+Once you do
+that, it allows a type checker (a compiler) to verify that the object (type) is used in a suitable
+way. Scala has the concept of abstract types: types
+that, just like abstract methods, must be defined by subclasses.
+Objects can have a type as members.
+On top of this abstract type, Scala defines a concept of path-dependent type.
+As the term path-dependent type says, the type depends on the path: in general, different paths give
+rise to different types.
+In the example of Figure~\ref{pathdependenttypeexample}\footnote{highly inspired from https://github.com/milessabin/shapeless\#sized-types}, we show how the size of the list can be used as a type information.
+We define a function row that convert a list of element into a printed html array, we define a function html that takes a list of size N as a first parameter and a list of list of size N as a second parameter.
+We use the path dependent type to create these type lists: \emph{Sized}.
+Consequently if we use this function \emph{html} with a list of three elements (\textit{headers}) and a list of list of 3 elements (\textit{rows}), the program compiles and runs correctly.
+If we use  this function \emph{html} with a list of four elements (\textit{extendedHeaders}) and a list of list of 3 elements (\textit{rows}), it does not compiled.
+
+\begin{figure}
+\begin{lstlisting}[label=pathdependenttypeexample,caption=Path dependent type example]
+//functions definition
+def row(cols : Seq[String]) 
+	= cols.mkString("<td>", "</td> <td>", "</td>")
+def html[N <: Nat]
+	(hdrs : Sized[Seq[String], N],
+	rows : List[Sized[Seq[String], N]]) 
+		= row(hdrs) :: rows.map(row(_))
+
+//Example1 that compiles
+val headers = Sized("Title", "Studio", "Release year")
+val rows = List(  
+	Sized("Ice Age: Dawn of the Dinosaurs", "Fox","2009"),
+	Sized("Finding Nemo","Disney Pixar","2003")
+)
+// headers and rows statically known to have the name number of columns
+val formatted = html(headers, rows)                        // Compiles
+
+//Sample 2 that does not compile
+// extendedHeaders has the wrong number of columns for rows
+val extendedHeaders = Sized("Title", "Author", "Release year", "Runtime")
+val badFormatted = html(extendedHeaders, rows)          
+\end{lstlisting}
+\end{figure}
+
+\begin{figure}
+\begin{lstlisting}[label=path-dependent-type-example-2,caption=Path-dependent type example]
+def html(
+    hdrs: Sized[Seq[String]],
+    rows: List[Sized[Seq[String]]])
+    (implicit evidence: hdrs.Size =:= rows.Size) = ...
+\end{lstlisting}
+\end{figure}
+
+More advanced type systems, such as the one featured in Scala, increase our confidence in the
+systems we develop and help us reduce the number of bugs before deployment. Path-dependent types and
+dependent method types play a crucial role for attempts to encode information into types that is
+typically only known at runtime. It is important to note that these important validations don't
+require extra effort from developers (no need to manually write tests or to run third-party tools and have no runtime performance impact), which is always a key concern when developing online
+systems. The goal of this paper is to highlight the benefits of putting the Scala type system for
+building efficient web applications.
+
+% 
+Dynamic typing may give programmers more flexibility but can make it harder for little scripts to grow into mature
+and robust code bases and to perform refactorings~\cite{Meijer04_StaticDynamic}. For instance, in the case of RIAs, a
+mispell in an event name may break the program behavior without anyone getting a sensible error message.
+Let us trying to come back on some examples of Web API that can no be easily typed correctly without the use of advanced type system.
+The following subsection shows some examples for which FFIs can not be designed easily.
+
+%TODO do the link and describe the problem. 
+%Why is it difficult to type Web browser’s APIs?
+
+\subsection{Motivating Examples}
+
+\subsubsection{DOM Creation.}
+
+The first motivating example is the definition of a FFI to wrap the JavaScript \textit{createElement} function.
+
+\begin{lstlisting}[language=JavaScript]
+var div = document.createElement('div');
+var input = document.createElement('input');
+\end{lstlisting}
+
+\code{div} has type \code{DivElement} while \code{input} has type \code{InputElement}. These types
+have different members. As stated in the introduction, most statically typed language have an API
+that returns an \code{Element}, which is the least upper bound of the possible types returned by
+\code{createElement}, forcing users to explicitly down-cast the result to the expected type, thus
+loosing type safety.
+
+Alternatively, some languages use overloading. Consider the following function creating an element and giving it a class name:
+
+\begin{lstlisting}[language=JavaScript]
+var create = function (elementType, className) {
+  val el = document.createElement(elementType);
+  el.className = className;
+  return el
+};
+create('input', 'field').focus();
+create('img', 'figure').src = 'http://google.com/logo.png';
+\end{lstlisting}
+
+How to type check the above code? The problem is that the return type of the \code{create} function
+depends on its first parameter.
+
+Note that a possible solution in Java could be the use of Generic Types as in the followinglisting:
+
+\begin{lstlisting}[language=Java,label=dom-generics]
+class ElementName<E> {}
+ElementName<InputElement> Input = new ElementName<InputElement>();
+ElementName<ImageElement> Img = new ElementName<ImageElement>();
+
+<E> E createElement(ElementName<E> name) {
+  ...
+}
+
+<E> E create(ElementName<E> name, String className) {
+  E el = createElement(name);
+  el.className = className;
+  return el;
+}
+
+create(Input, "field").focus();
+create(Img, "figure").setSrc("http://google.com/logo.png");
+\end{lstlisting}
+
+However, the type parameter \code{E} has to be filled at use-site. If the compiler could infer it,
+that’s better for the conciseness.
+
+\subsubsection{DOM Events.}
+
+A similar issue applies to the DOM events API. Indeed, RIAs distinguish themselves from classic Web
+pages by a higher degree of interactivity. In our example we want to let users zoom in and out a map
+using their mouse wheel. To implement this feature we need to attach an event handler for the
+\code{mousewheel} event on the DOM element containing the map. The following listing shows how to
+attach such an event handler using the native JavaScript API:
+
+\begin{lstlisting}[language=JavaScript,label=event-js,caption=Native JavaScript API to handle
+events]
+mapElement.addEventListener("mousewheel", function (e) {
+  scale = Math.round(scale + e.wheelDeltaY / 16);
+  // ... update the user interface with the new scale
+});
+\end{lstlisting}
+
+The event handler is simply a function taking the event data as a parameter. In the above example we
+use the \code{wheelDeltaY} property of the event to update the scale of the map.
+
+We claim this code is fragile for two reasons. (1) The name of the event is passed as a string, so
+it is easy to mispell it. (2) The callback passed as a second parameter takes a parameter \code{e}
+whose fields vary according to the listened event, but developers have no way to check that the
+fields they’re using are indeed defined on the event they’re listening to. For instance, in our case
+we use the \code{wheelDeltaY} event field that is defined only on the \code{mousewheel} event.
+
+We want to bring safety to the event handling API so that developers cannot mispell an event name
+and cannot attempt to read a property that is not defined on the type of the handled event. The
+difficulty comes from the fact that the type of data passed to the handler varies with the event
+name. For instance, in listing \ref{event-js} the handler processes \code{MouseWheelEvent} values
+because it listens to \code{mousewheel} events. Values of type \code{MouseWheelEvent} have a
+property \code{wheelDeltaY} but that’s not the case of an event data of type \code{KeyboardEvent},
+for example.
+
+A possible solution could be to define a distinct method for each event instead of the single
+\code{addEventListener} method, so each method takes a handler with the according event type:
+
+\begin{lstlisting}
+window.onKeyUp { e: Rep[KeyboardEvent] =>
+  println(e.key)
+}
+window.onMouseWheel { e: Rep[MouseWheelEvent] =>
+  println(e.wheelDeltaY)
+}
+\end{lstlisting}
+
+In the above listing, the \code{onKeyUp} method attaches an event handler for the \code{keyup} event
+that uses values of type \code{KeyboardEvent}, and the \code{onMouseWheel} method does the same for
+\code{mousewheel} events that use values of type \code{MouseWheelEvent}.
+
+However, with this solution the event name can not anymore be a parameter of the program, making it
+impossible to abstract over the task of adding an event listener, though this is easy to write in
+JavaScript (using curryfication) and useful to decouple the actions of adding an event
+listener and reacting to the events:
+
+\begin{lstlisting}[language=JavaScript,label=react-js]
+var events = function (name) {
+  return function (callback) {
+    window.addEventListener(name, callback);
+  }
+};
+
+var clicks = events('click');
+clicks(function (event) { console.log(event); });
+clicks(function (event) { alert(event); });
+\end{lstlisting}
+
+As in the previous section, this problem can be solved using Java generics:
+
+\begin{lstlisting}
+class EventName<EventData> {}
+EventName<MouseEvent> Click = new EventName<EventData> {};
+
+<EventData> void addEventListener(EventName<EventData> name, Function<EventData, Void> callback) {
+    ...
+}
+\end{lstlisting}
+
+\subsubsection{Selectors.}
+
+The last motivating example refers to DOM selectors. The problem with selectors is slightly different.
+
+\begin{lstlisting}[language=JavaScript]
+var input = document.querySelector('input');
+var el = document.querySelector('#content');
+\end{lstlisting}
+
+\code{input} can not have another type than \code{InputElement}, but \code{el} can be an element of
+any type (depending on which element has the id \code{content}). Return type can not \emph{always}
+be inferred from the parameters. In this case, down-casting is impossible to avoid, but how to make
+it optional, and how to make it less risky?
+
+We want to be able to write the following code:
+
+\begin{lstlisting}
+val input = document.find(Input) // input has type InputElement
+val el = document.find("#content") // el has type Element
+val img = document.find[Img](".figure") // note that the type Img is 
+                                        // explicitly applied in this case
+                                        // but was automatically inferred 
+                                        // in previous cases
+val error = document.find[Int]("div") // Does not compile: querySelector 
+                                      // can not return a number
+\end{lstlisting}
+
+These three motivating examples highlights some functions often used in a web application developped in JavaScript and discuss the desired FFI.
+
+%\subsection{dependent types}
+
+\section{Js-scala: A DSL for web programming design using LMS}
+
+JavaScript libraries could address some of the challenges presented above, but addressing type safety and
+asynchronous programming issues require language modifications (\eg{} to bring a type system and a sequencing
+notation). Defining a language requires a high effort, especially to build development tools for the language so we
+chose to define our language as a compiled embedded DSL in Scala.
+
+%\subsection{Introduction to Lightweight Modular Staging}
+%\label{intro-lms}
+
+\subsection{LMS Overview}
+
+We use the Lightweight Modular Staging framework (LMS) to define our compiled embedded DSL. The main idea is that a
+program written using a DSL is evaluated in two stages (or steps): the first stage builds an intermediate
+representation of the program and the second stage transforms this intermediate representation into executable code
+(figure \ref{lms-diagram}).
+
+\begin{figure}
+  \centering
+  \includegraphics[width=7cm]{lms.pdf}
+  \caption{Compilation of a program using LMS. An initial Scala program using embedded DSLs (on the left) evaluates
+  to an intermediate representation from which the final program’s code is generated (on the right).}
+  \label{lms-diagram}
+\end{figure}
+
+The bindings between stages are type-directed: a value of type \code{Rep[Int]} in the first stage will yield a value
+of type \code{Int} in the second stage. If you consider the following code:
+\begin{lstlisting}
+val inc: Rep[Int] => Rep[Int] =
+  x => x + 1
+\end{lstlisting}
+The function looks like
+a regular Scala function excepted that its parameter type and its return type are wrapped in the \code{Rep[T]} type
+constructor that denotes intermediate representations. The \code{+} operator has been defined on \code{Rep[Int]}
+values and returns the intermediate representation of an addition. Finally, the \code{inc} function returns the
+intermediate representation of a computation yielding the number following the value of the parameter \code{x}.
+
+You can get a \code{T} value from a \code{Rep[T]} value by generating code from the intermediate representation and
+compiling it:
+\begin{lstlisting}
+val compiledInc: Int => Int =
+  compile(inc)
+\end{lstlisting}
+The \code{compile} function takes a staged program of type \code{Rep[A] => Rep[B]} and returns a final program of
+type \code{A => B}.
+
+The intermediate representation implementation is hidden for users but DSLs authors have to provide the corresponding
+intermediate representation of each construct of their language. For that purpose, LMS comes with an extensible
+intermediate representation implementation defining computations as a graph of statements. In the case of the
+\code{inc} function, this graph contains a \code{Plus} node applied on a \code{Sym} node (the \code{x} parameter) and
+a \code{Const} node (the literal value \code{1}).
+
+Then, the code generation process consists in sorting this graph according to expressions dependencies and to emit
+the code corresponding to each node. The listings \ref{codegen-js} and \ref{codegen-scala} show the JavaScript and
+Scala generated code for the \code{inc} function:
+\begin{multicols}{2}
+\lstinputlisting[language=JavaScript, caption=JavaScript code generation, label=codegen-js]{inc.js}
+\lstinputlisting[caption=Scala code generation, label=codegen-scala]{inc.scala}
+\end{multicols}
+
+Defining a DSL consists in three steps, each defining:
+
+\begin{itemize}
+\item The concrete syntax: an abstract API manipulating \code{Rep[_]} values~;
+\item The intermediate representation: an implementation of the concrete syntax in terms of statement nodes~;
+\item A code generator for the intermediate representation: a \emph{pretty-printer} for each DSL statement node.
+\end{itemize}
+
+\subsubsection{Example: null references}
+
+As an illustration of the staging mechanism, we present a simple DSL to handle null references. This DSL provides an
+abstraction at the stage-level that is removed by optimization during the code generation.
+
+Null references are a known source of problems in programming languages~\cite{Hoare09_Null,Nanda09_Null}. For
+example, consider the following typical JavaScript code finding a particular widget in the page and a then particular
+button in the widget:
+
+\begin{lstlisting}[language=JavaScript,label=null-unsafe,caption=Unsafe code]
+var loginWidget = document.querySelector("div.login");
+var loginButton = loginWidget.querySelector("button.submit");
+loginButton.addEventListener("click", function (e) { ... });
+\end{lstlisting}
+
+The native \code{querySelector} method returns \code{null} if no node matched the given selector in the document. If
+we run the above code in a page where the widget is not present, it will throw an error and stop further JavaScript
+execution. We can write defensive code to handle \code{null} cases, but it leads to very cumbersome code:
+
+\begin{lstlisting}[language=JavaScript,label=null-defensive,caption=Defensive programming to handle null references]
+var loginWidget = document.querySelector("div.login");
+if (loginWidget !== null) {
+  var loginButton = loginWidget.querySelector("button.submit");
+  if (loginButton !== null) {
+    loginButton.addEventListener("click", function (e) { ... });
+  }
+}
+\end{lstlisting}
+
+Ja-scala is a DSL that has both the safety and performance of listing \ref{null-defensive} but the
+expressiveness of listing \ref{null-unsafe}. We can get safety by wrapping potentially null values of type
+\code{Rep[A]} in a container of type \code{Rep[Option[A]]} requiring explicit dereferencing, we can get
+expressiveness by using the Scala \code{for} notation for dereferencing, and finally we can get performance by
+generating code that does not actually wraps values in a container but instead checks if they are \code{null} or not
+when dereferenced. The wrapping container exists only at the stage-level and is removed during the code generation.
+Here is a Scala listing that uses our DSL (implementation details are given in section \ref{implementation}):
+
+\begin{lstlisting}
+for {
+  loginWidget <- document.find("div.login")
+  loginButton <- loginWidget.find("submit.button")
+} loginButton.on(Click) { e => ... }
+\end{lstlisting}
+
+The evaluation of the above listing produces a graph of statements from which JavaScript code equivalent to
+listing \ref{null-defensive} is generated.
 
@@
 
+\subsection{DOM}
+
+As seen in the introduction, the return type of the function creating DOM elements depends on
+the value of its first parameter. Listing \ref{dom-generics} showed a statically typed interface
+for the \code{createElement} function using type parameters. However, as shown in the listing,
+functions taking an \code{ElementName} parameter also have to take a type parameter (\eg{} function
+\code{create}). Using type parameters may lead to an explosion of use of type parameters in
+the code. Instead, we propose to use path-dependent types so the dependency between the
+element name and the function return type is more self-contained (no need to fill a type parameter
+at use-site):
+
+\begin{lstlisting}
+class ElementName {
+  type Element
+}
+
+def createElement(name: ElementName): name.Element = ...
+\end{lstlisting}
+
+We can rewrite the \code{create} function as follows:
+
+\begin{lstlisting}
+def create(name: ElementName, className: String) = {
+  val el = createElement(name)
+  el.className = className
+  el
+}
+\end{lstlisting}
+
+The \code{create} function type signature is not polluted with an extra type parameter.
+
 \subsection{Events}
 
-Path-dependent types to abstract over an event name and its data type.
+We can apply the same pattern to the \code{addEventListener} function (renamed to \code{on}):
+
+\begin{lstlisting}
+class EventName {
+  type Data
+}
+object Click extends EventDef { type Data = MouseEvent }
+
+def on(name: EventName)(handler: Rep[name.Data] => Rep[Unit]): Rep[Unit]
+
+val clicks = on(Click)
+clicks(event => println(event.offsetX))
+\end{lstlisting}
+
+An \code{EventName} value represents an event that carries its corresponding event data type in its
+\code{Data} type member. The \code{on} method takes a parameter \code{ev} of type \code{EventDef}
+and a \code{handler} parameter whose type refers to the \code{Data} member of the \code{name}
+parameter, so the type of the handler depends on the \code{name} value.
+
+The following Scala listing generates code equivalent to listing \ref{event-js} and is completely
+type-safe: if the user misspells the event name or tries to use an undefined property on an event
+his program won’t compile.
+
+\begin{lstlisting}
+window.on(MouseWheel) { e =>
+  scale = Math.round(scale + e.wheelDeltaY / 16)
+}
+\end{lstlisting}
+
+We can write a statically typed version of listing \ref{lst:react-js} as follows:
+
+\begin{lstlisting}
+val clicks = on(Click)
+clicks(event => println(event))
+clicks(event => alert(event))
+\end{lstlisting}
 
@@
 
-\begin{itemize}
- \item Less type annotations on DOM queries, less chance to write nonsense casts
- \item Inference-driving macros help inferring more specific types
-\end{itemize}
+As presented in the introduction, the JavaScript \code{querySelector} function is polymorphic in
+its return type but this return type can not \emph{always} be inferred from its parameters. A
+pragmatic solution to this problem would be to return by default the most general type, to handle
+cases where a more specialized return type can be safely inferred without requiring an explicit
+annotation from users, and finally to handle cases where users have to down-cast the return type
+while forbidding to cast to an impossible return type.
 
-\subsection{DOM}
+First, we start by making the return type bound to the most general return type:
 
-?
+\begin{lstlisting}
+def find[A <: Element](selector: Rep[String]): Rep[A]
+\end{lstlisting}
+
+When calling this function, the \code{A} type parameter can be filled at use-site but can only be
+filled with a type that is a subtype of \code{Element}, so we forbid cases like the following:
+
+\begin{lstlisting}
+find[Int](".title") // Does not compile
+\end{lstlisting}
+
+However, if the \code{A} type parameter is not explicitly filled by the user, the Scala compiler
+infers the greatest lower bound for \code{A}: \code{Nothing}.
+
+\begin{lstlisting}
+find(".title") // Returns Rep[Nothing]...
+\end{lstlisting}
+
+We can fix that by replacing the bounded quantification by an implicit parameter \emph{witnessing}
+that \code{A} can be the result of searching elements in the DOM:
+
+\begin{lstlisting}
+def find[A](selector: Rep[String])(implicit ev: Findable[A]): Rep[A]
+
+class Findable[A]
+object Findable {
+  implicit val default: Findable[Element] = new Findable[Element]
+  implicit def elementSub[A <: Element]: Findable[A] = new Findable[A]
+}
+\end{lstlisting}
+
+If users try to fill \code{A} with \code{Int} they get a compile error (implicit not found).
+
+So far, our definition has the most general return type by default and allows users to explicitly
+ask to down-cast the result to a subtype of \code{Element}, but our definition is still unable to
+automatically infer a more specialized return type when possible.
+
+We solve this problem by making the \code{selector} parameter polymorphic too and by replacing the
+previous implicit parameter by another one fixing the return type of the function according to the
+selector parameter type:
+
+\begin{lstlisting}
+def find[A, B](selector: A)(implicit evidence: A Selects B): B
+
+class Selects[A, B]
+object Selects {
+  // Case of selector being a String
+  implicit val string: String Selects Element = new Selects[String, Element]
+  implicit def elementSub[A <: Element]: String Selects A = new Selects[String, A]
+  // Case of selector being an element name (cf below)
+  implicit def elementName[A <: Element]: ElementName[A] Selects A =
+    new Selects[ElementName[A], A]
+
+  type ElementNameAlias[El] = ElementName { type Element = El }
+  implicit def elementName: ElementNameAlias[El] Selects El
+}
+
+class ElementName[A <: Element]
+val Div: ElementName[DivElement] = new ElementName[DivElement]
+
+class ElementName {
+  type Element
+}
+val Div: ElementName { type Element = DivElement } = new ElementName { type Element = DivElement }
+\end{lstlisting}
+
+Now users can write the following:
+
+\begin{lstlisting}
+find(".foo") // returns an Element
+find[String, InputElement](".foo") // returns an InputElement
+find[String, Int](".foo") // does not compile
+find(Div) // returns a DivElement
+find[ElementName[DivElement], Int](Div) // does not compile
+\end{lstlisting}
+
+However, now that the \code{find} function takes two type parameters, users have to supply both
+if they want to explicitly down-cast the return type of the query. We can solve this problem by
+creating an intermediate class:
+
+\begin{lstlisting}
+trait FindBuilder[B] {
+  def apply[A](selector: A)(implicit evidence: A Selects B): B
+}
+
+def find[B]: FindBuilder[B]
+\end{lstlisting}
+
+Now users can write the following:
+
+\begin{lstlisting}
+find(".foo") // returns an Element
+find[InputElement](".foo") // returns an InputElement
+find[Int](".foo") // does not compile
+find(Div) // returns a DivElement
+find[Int](Div) // does not compile
+\end{lstlisting}
 
@@
 
-Other languages either provide loose information about the data type of the listened event (Dart) or give no way to
-abstract over an event (GWT, Kotlin).
+Other languages either provide loose information about the data type of the listened event (Dart) or
+give no way to abstract over an event (GWT, Kotlin).
+
+The most popular JavaScript library, jQuery~\cite{Bibeault08_jQuery}, used by more than 40\% of the top million
+sites\footnote{\href{http://trends.builtwith.com/javascript}{http://trends.builtwith.com/javascript}} aims to
+simplify RIA development. It handles the \code{null} references problem by wrapping each query result in a container
+so before each further method call it tests the emptiness of the container and applies the operation effectively only
+if the container is not empty. It provides an expressive API but the emptiness checking involves a slight performance
+penalty since each result is wrapped and then unwrapped for the subsequent method invocation. jQuery also has an API
+to turn asynchronous computations into first-class citizens. However, jQuery is just a library and thus cannot bring
+type safety to JavaScript programs.
+
+TypeScript\footnote{\href{http://www.typescriptlang.org/}{http://www.typescriptlang.org/}} is a language compiling to
+JavaScript supporting classes, modules and soft typing. Because TypeScript is a superset of JavaScript, any valid
+JavaScript program is a valid TypeScript program. Developers can progressively leverage TypeScript’s features to turn
+their JavaScript sources into large, modular and more robust code bases. However, TypeScript does not provide advanced type feature to deal with DOM management.
+
+The Roy~\cite{McKenna_Roy} programming language is a statically typed functional programming language compiling to
+JavaScript. It features global type inference, algebraic data types, pattern matching and a sequencing notation
+similar to Scala’s \code{for} notation. Roy does not address concerns such as event handling and DOM fragments
+definition. It has no specific feature to deal with correctly typed FFI.
+
+GWT~\cite{Chaganti07_GWT} compiles Java code to JavaScript. It exposes the browser’s APIs through type-safe Java APIs
+and simplifies DOM fragments definition.
+However, to provide correct type management, GWT provides a verbose API to deal with DOM fragment definition and event handling.
+New DOM Api can not be easily managed in a concise and safe way as the Java programming language does not provide advanced type feature.
+
+Kotlin\footnote{\href{http://kotlin.jetbrains.org/}{http://kotlin.jetbrains.org/}} is a recent language written by
+JetBrains, that targets both the JVM and JavaScript. It has libraries addressing partially Web programming concerns
+and supports advanced programming language features such as mixins, type inference, variance annotations for type
+parameters and pattern matching. In the specific problem of the definition of concise and safe FFIs it has the same issue than GWT.
+
+Opa\footnote{\href{http://opalang.org/}{http://opalang.org/}} and Links~\cite{Cooper07_Links} are languages designed
+for Web programming. They feature static typing with global type inference, modules and pattern matching. These
+languages provide a syntactic sugar for writing HTML fragments directly in the code. The server and client parts of
+an application are written in a single language and the system can decide whether a code fragment will run on
+server-side, client-side or both. However, these languages are not extensible, so there is no way for developers to
+customize the compilation process of their abstractions on client and server sides. Furthermore, hiding the code
+execution distribution gives programmers less control on communication error handling~\cite{Guerraoui99_Distributed}.
+
+Dart~\cite{Griffith11_Dart} is a language designed specifically for Web programming. It has been designed to have a
+familiar syntax for JavaScript developers and supports soft typing, classes and a module system. It also provides a
+high-level concurrent programming system similar to an actor system. Finally, Dart code can be executed on both
+client and server sides but the compiler is not extensible so developers cannot customize the compilation process of
+a given abstraction on client and server sides.
+
+Table \ref{comparison} gives a synthetic view of the comparison of our work with other mainstream approaches. Lines
+give features desired to write large Web applications and colums list mainstream tools for Web programming. Cells
+show the support level of each feature for each tool.
+
+TODO{Fay}
+
+\begin{table}
+\centering
+\begin{tabular}{| l | c | c | c | c | c | c | c | c |}
+\hline
+& jQuery & TypeScript & Roy & GWT & Kotlin & Opa & Dart & js-scala \\
+\hline
+Type safe event handling & & & & *** & & & & *** \\
+\hline
+DOM definition & * & & & *** & ** & *** & * & *** \\
+\hline
+Static typing & & * & ** & ** & ** & ** & * & ** \\
+\hline
+Easiness for defining libraries & ** & *** & *** & *** & *** & *** & *** & * \\
+\hline
+\end{tabular}
+\caption{Support of concise and safe FFI. Zero star means that the feature is not supported
+at all, three stars mean a strong support.}
+\label{comparison}
+\end{table}
+
+Several model-driven methodologies for creating web applications have been proposed, including
+\cite{schwabe1996systematic,lima2003modeling,ceri2000web,koch2001authoring,pastor2003oows,valverde2007mda,vdovjak2003engineering,kraus2007model,nunes2006rapid,brambilla2008designing,valderas2007transformational,van2006hera,Groenewegen08_WebDSL}.
+Many of them follow the standard top-down MDE vision in providing abstractions and code generators, nevertheless
+these cannot be easily customized or they generate only application skeletons.
 
@@
 
-\begin{thebibliography}{4}
-
-\bibitem{jour} Smith, T.F., Waterman, M.S.: Identification of Common Molecular
-Subsequences. J. Mol. Biol. 147, 195--197 (1981)
-
-\bibitem{lncschap} May, P., Ehrlich, H.C., Steinke, T.: ZIB Structure Prediction Pipeline:
-Composing a Complex Biological Workflow through Web Services. In: Nagel,
-W.E., Walter, W.V., Lehner, W. (eds.) Euro-Par 2006. LNCS, vol. 4128,
-pp. 1148--1158. Springer, Heidelberg (2006)
-
-\bibitem{book} Foster, I., Kesselman, C.: The Grid: Blueprint for a New Computing
-Infrastructure. Morgan Kaufmann, San Francisco (1999)
-
-\bibitem{proceeding1} Czajkowski, K., Fitzgerald, S., Foster, I., Kesselman, C.: Grid
-Information Services for Distributed Resource Sharing. In: 10th IEEE
-International Symposium on High Performance Distributed Computing, pp.
-181--184. IEEE Press, New York (2001)
-
-\bibitem{proceeding2} Foster, I., Kesselman, C., Nick, J., Tuecke, S.: The Physiology of the
-Grid: an Open Grid Services Architecture for Distributed Systems
-Integration. Technical report, Global Grid Forum (2002)
-
-\bibitem{url} National Center for Biotechnology Information, \url{http://www.ncbi.nlm.nih.gov}
-
-\end{thebibliography}
+\bibliographystyle{plain}
+\bibliography{biblio}
 

### 2013-07-15 papers/sle2013/inc.js

@@ -1,0 +1,4 @@
+var inc = function(x0) {
+  var x1 = x0 + 1;
+  return x1;
+};

### 2013-07-15 project/Build.scala
Do not publish sub-projects artifacts

@@ -17,7 +17,9 @@ object JsScalaBuild extends Build {
   lazy val root = Project(
     "root",
     file("."),
-    settings = buildSettings
+    settings = buildSettings ++ Seq(
+      publishLocal := ()
+    )
   ) aggregate (macros, core, examples)
 
   lazy val macros = Project(
@@ -30,7 +32,8 @@ object JsScalaBuild extends Build {
       libraryDependencies ++= Seq(
         "org.scalatest" %% "scalatest" % "2.0.M5b" % "test",
         "EPFL" %% "lms" % "0.3-SNAPSHOT"
-      )
+      ),
+      publishLocal := ()
     )
   )
 
@@ -39,7 +42,11 @@ object JsScalaBuild extends Build {
     file("core"),
     settings = buildSettings ++ Seq(
       scalacOptions ++= Seq("-deprecation", "-unchecked", "-Xexperimental", "-P:continuations:enable", "-Yvirtualize", "-language:dynamics"/*, "-Ymacro-debug-lite"*/),
+
       name := "js-scala",
+      mappings in (Compile, packageBin) <++= mappings in (macros, Compile, packageBin),
+      mappings in (Compile, packageSrc) <++= mappings in (macros, Compile, packageSrc),
+
       //Our tests are not threadsafe so disabling parallel execution for now
       parallelExecution in Test := false,
       // disable publishing of main docs
@@ -59,6 +66,7 @@ object JsScalaBuild extends Build {
     "examples",
     file("examples"),
     settings = buildSettings ++ Seq(
+      publishLocal := (),
       autoCompilerPlugins := true,
       libraryDependencies <<= (scalaVersion, libraryDependencies) { (ver, deps) => deps :+ compilerPlugin("org.scala-lang.virtualized.plugins" % "continuations" % ver)},
       scalacOptions ++= Seq("-deprecation", "-unchecked", "-Xexperimental", "-P:continuations:enable", "-Yvirtualize", "-language:dynamics")

### 2013-07-16 project/Build.scala
Fix scala version to use with macros

@@ -26,7 +26,7 @@ object JsScalaBuild extends Build {
     "macros",
     file("macros"),
     settings = buildSettings ++ Seq(
-      scalaVersion := "2.10.2-SNAPSHOT",
+      scalaVersion := "2.10.3-SNAPSHOT",
       scalaOrganization := "org.scala-lang.macro-paradise",
       libraryDependencies <+= (scalaVersion)("org.scala-lang.macro-paradise" % "scala-reflect" % _),
       libraryDependencies ++= Seq(

### 2013-09-02 papers/gpce2013/gpce2013.tex
typos

@@
 benefit from high-level languages and libraries making code complexity easier to manage and
-abstracting over the clients and servers differences without trading this engineering comfort for
+abstracting over the clients and servers differences without trading this ease of engineering for
 performance? This article presents high-level abstractions defined as deep embedded DSLs in Scala,
 that can (1) generate efficient code leveraging the target platform characteristics, (2) be shared
-between clients and servers. Though code written with our DSL has a high level of abstraction our
+between clients and servers. Though code written with our DSL has a high level of abstraction, our
 benchmark on a real world application reports that it runs as fast as hand tuned low-level
@@
 generates JavaScript code, making it possible to write the client-side code of Web applications
-using JavaScript~\cite{Kossakowski12_JsDESL}. However, Kossakowski \etal did not address the
+using Scala~\cite{Kossakowski12_JsDESL}. However, Kossakowski \etal did not address the
 engineering dilemma described above. This paper enriches js-scala with the following
@@
 discuss the relative pro and cons of them. Though the code written in our DSL is high-level and can
-be shared between clients and servers, it has the same runtime performances on client-side as
+be shared between clients and servers, it has the same runtime performance on client-side as
 hand-tuned low-level JavaScript code.
@@
 the previous approach because each language has a limited set of features. On the other hand, it is
-difficult to have interoperability between DSLs. \cite{Groenewegen08_WebDSL} gave an example of such
+difficult to have interoperability between DSLs. \cite{Visser07} gave an example of such
 a domain-specific language for defining Web applications.
@@
 shows an equivalent JavaScript program as listing \ref{vanilla-selectors}, but using jQuery. The
-code is both shorter and simpler, thanks to its higher-level of abstraction. jQuery provides an API
-that is simpler to master because it has less functions, but this benefit comes at the price of a
-decrease in runtime performances.
+code is both shorter and simpler, thanks to its higher level of abstraction. jQuery provides an API
+that is simpler to master because it has fewer functions, but this benefit comes at the price of a
+decrease in runtime performance.
 
 Instead, we propose a solution that has a high-level API but generates JavaScript code using the
-specialized native API, when possible, in order to get both engineering comfort and performances.
+specialized native API, when possible, in order to get both ease of engineering and performance.
 We achieve this by analyzing, during the first evaluation step, the selector that is
@@
 
-Our goal is to evaluate the level of abstraction provided by our solution and its performances, by
+Our goal is to evaluate the level of abstraction provided by our solution and its performance, by
 comparing it with common approaches. We take the number of lines of code as an inverse approximation
@@
 
-The runtime performances of the vanilla JavaScript, HaXe and js-scala versions are similar (though
+The runtime performance of the vanilla JavaScript, HaXe and js-scala versions are similar (though
 the js-scala version is slightly slower by 6\%). It is worth noting that the vanilla JavaScript and
@@
 We were able to share some DOM fragment definitions between server-side and client-side only in
-the js-scala version. In the GWT version we don’t have choice: dynamic DOM fragments are always
+the js-scala version. In the GWT version we don’t have a choice: dynamic DOM fragments are always
 built only on client-side (a practice that makes it more difficult to make the pages content
@@
 
-Our goal was to put the runtime performances in perspective with the level of abstraction. We are
+Our goal was to put the runtime performance in perspective with the level of abstraction. We are
 aware that the indicator we chose as an inverse approximation of the abstraction level, the number
@@
 We implemented our solution as compiled embedded DSLs in Scala. Generating code from our DSLs is a
-two steps process: an initial Scala program first evaluates to an intermediate representation of
+two step process: an initial Scala program first evaluates to an intermediate representation of
 the final program that is traversed by code generators to produce the final JavaScript code.
@@
 abstractions, making the code of our application simpler to reason about and making some parts of
-the code reusable between client and server sides, while keeping execution performances on
+the code reusable between client and server sides, while keeping execution performance on
 client-side as high as if we used hand-tuned low-level JavaScript code.
@@
 Our two kinds of benchmarks, (i) micro-benchmarks to evaluate one abstraction and (ii) a benchmark
-on a real application using these abstractions, show performances similar to hand-optimized
+on a real application using these abstractions, show performance similar to hand-optimized
 low-level code.

### 2013-09-02 papers/gpce2013/gpce2013.tex
Add info on the size of the Chooze application
Capitalization

@@
 We classified existing approaches providing high-level abstractions for Web programming in four
-categories, as shown in figure \ref{fig:langs}.
+categories, as shown in Figure \ref{fig:langs}.
 
@@
 then to combine all the source artifacts written with these language into one executable program, as
-shown in figure~\ref{fig:langs} (b). Defining such languages requires a minimal effort compared to
+shown in Figure~\ref{fig:langs} (b). Defining such languages requires a minimal effort compared to
 the previous approach because each language has a limited set of features. On the other hand, it is
@@
 %\vspace{0.1cm}
-The last approach, shown in figure \ref{fig:langs} (d), can be seen as a middle-ground between the
+The last approach, shown in Figure \ref{fig:langs} (d), can be seen as a middle-ground between the
 two previous approaches: DSLs are embedded in a host language but use a code generation process.
@@
 that actually provides only one function to search for elements. Listing \ref{jquery-selectors}
-shows an equivalent JavaScript program as listing \ref{vanilla-selectors}, but using jQuery. The
+shows an equivalent JavaScript program as Listing \ref{vanilla-selectors}, but using jQuery. The
 code is both shorter and simpler, thanks to its higher level of abstraction. jQuery provides an API
@@
 
-Finally, listing~\ref{js-scala-selectors} shows how to implement listing~\ref{jquery-selectors} in
+Finally, Listing~\ref{js-scala-selectors} shows how to implement Listing~\ref{jquery-selectors} in
 Scala using js-scala. The code has the same abstraction level as with jQuery, however it
-generates a JavaScript program identical to listing~\ref{vanilla-selectors}: the high-level
+generates a JavaScript program identical to Listing~\ref{vanilla-selectors}: the high-level
 abstractions (the \code{find} and \code{findAll} functions) exist only in the initial program, not
@@
 \code{null} references are a known source of problems in programming
-languages~\cite{Hoare09_Null,Nanda09_Null}. For example, consider listing \ref{null-unsafe} finding
+languages~\cite{Hoare09_Null,Nanda09_Null}. For example, consider Listing \ref{null-unsafe} finding
 a particular widget in the page and then a particular button within the widget. The native
@@
 and stop further JavaScript execution. Defensive code can be written to handle \code{null}
-references, but leads to very cumbersome code, as shown in listing \ref{null-defensive}.
+references, but leads to very cumbersome code, as shown in Listing \ref{null-defensive}.
 
@@
 there is no value). An optional value can be dereferenced using the \code{for} notation, as shown
-in listing \ref{null-js-scala}, that implements in js-scala a program equivalent to listing
+in Listing \ref{null-js-scala}, that implements in js-scala a program equivalent to Listing
 \ref{null-defensive}. The \code{find} function returns a \code{Rep[Option[Element]]}. The \code{for}
@@
 container objects but instead checks if they are \code{null} or not when dereferenced. So the extra
-container object exists only in the initial program and is removed during code generation: listing
-\ref{null-js-scala} produces a code equivalent to listing~\ref{null-defensive}.
+container object exists only in the initial program and is removed during code generation: Listing
+\ref{null-js-scala} produces a code equivalent to Listing~\ref{null-defensive}.
 
@@
 checks if the optional value contains a value) and \code{foreach} (that is called when the
-\code{for} notation is used, as in listing \ref{null-js-scala}). The \code{emitNode} method handles
+\code{for} notation is used, as in Listing \ref{null-js-scala}). The \code{emitNode} method handles
 \code{OptionIsEmpty} and \code{OptionForeach} nodes returned by the implementations of
@@
 page content. This task can be performed either from the server-side (to initially respond to a
-request) or from the client-side (to update the current page). As an example, listing \ref{dom-api}
+request) or from the client-side (to update the current page). As an example, Listing \ref{dom-api}
 defines a JavaScript function \code{articleUi} that builds a DOM tree containing an article
-description, and listing \ref{scala-xml-api} shows how one could implement a similar function on
+description, and Listing \ref{scala-xml-api} shows how one could implement a similar function on
 server-side using the standard Scala XML library. The reader may notice that the client-side
@@
 text node. Listing \ref{forest} shows how to implement our example with our DSL. The children
-elements of an element can also be obtained dynamically from a collection, as shown in listing
+elements of an element can also be obtained dynamically from a collection, as shown in Listing
 \ref{forest-loops}.
@@
 and produce code building an equivalent DOM tree and XML fragment, respectively. When the children
-of an element are constant values (as in listing \ref{forest}) rather than dynamically computed (as
-in listing \ref{forest-loops}), the code generators unroll the loop that adds children to their
-parent, for better performance. As a result, listing \ref{forest} generates a code equivalent to
-listing \ref{dom-api} on client-side and equivalent to listing \ref{scala-xml-api} on server-side.
+of an element are constant values (as in Listing \ref{forest}) rather than dynamically computed (as
+in Listing \ref{forest-loops}), the code generators unroll the loop that adds children to their
+parent, for better performance. As a result, Listing \ref{forest} generates a code equivalent to
+Listing \ref{dom-api} on client-side and equivalent to Listing \ref{scala-xml-api} on server-side.
 
@@
 high-level API (\code{querySelector} and \code{querySelectorAll}). So we directly compared the
-execution time of the JavaScript code generated by listing \ref{js-scala-selectors} with a
-JavaScript program equivalent to listing \ref{vanilla-selectors} but using the high-level native
+execution time of the JavaScript code generated by Listing \ref{js-scala-selectors} with a
+JavaScript program equivalent to Listing \ref{vanilla-selectors} but using the high-level native
 API (\code{querySelector} and \code{querySelectorAll}) instead. The code was executed in a Web page
@@
 Figure \ref{fig:selectors-benchmark} shows the benchmark results. The JavaScript-opt version is
-listing \ref{vanilla-selectors}, which uses low-level native APIs, the JavaScript version is the
-equivalent listing using the high-level native API, and the jQuery version is listing
+Listing \ref{vanilla-selectors}, which uses low-level native APIs, the JavaScript version is the
+equivalent listing using the high-level native API, and the jQuery version is Listing
 \ref{jquery-selectors}. The js-scala version is slightly slower than the JavaScript-opt (by 14\%),
@@
 the dynamic behavior of the application: double-posting prevention, dynamic form update and rich
-interaction with the document.
+interaction with the document. The size of the whole application (server and client sides) is about one thousand of lines of code.
 

### 2013-09-02 papers/gpce2013/gpce2013.tex
Explain the selector listing. Mention JavaScript libraries in the related works

@@
 target platform characteristics). Examples of languages following this approach are Java/GWT,
-Kotlin, HaXe and SharpKit.
+Kotlin, HaXe and SharpKit. Libraries written in JavaScript (\eg{} jQuery~\cite{Bibeault08_jQuery}) also match this category though most of them do not support both client and server sides.
 
@@
 Listing \ref{vanilla-selectors} gives an example of use of various functions from the native
-selectors API to retrieve a list of input fields within a form. The existence of several
-specialized functions in the API makes it possible to write efficient code, but forces users to
-think at a low abstraction level.
+selectors API to retrieve a list of input fields within a form.
+The \code{getWords} function first finds in the document the HTML element with id \code{add-user}, then collects all its \code{fieldset} children elements, and for each one returns the list of its children elements having class \code{word}.
+The existence of several specialized functions in the API makes it possible to write efficient code, but forces users to think at a low abstraction level.
 
@@
 \href{http://trends.builtwith.com/javascript}{http://trends.builtwith.com/javascript}, jQuery is
-used by more than 40\% of the top million sites.} use the jQuery library~\cite{Bibeault08_jQuery}
+used by more than 40\% of the top million sites.} use the jQuery library
 that actually provides only one function to search for elements. Listing \ref{jquery-selectors}

### 2013-09-02 core/src/main/scala/scala/js/gen/js/GenAdts.scala
adt

@@ -36,5 +36,3 @@ trait GenAdts extends GenBase with GenFunctions with GenEqual {
        
   }
 }
-
-// emitValDef(sym, "Function("+quote(obj)+", "+fs.map(quote).mkString(",")+"){ ["+fs.map(quote).mkString(",")+"]["+quote(obj)+".$variant]("+quote(obj)+") }")

### 2013-09-02 core/src/test/scala/scala/js/TestAdt.scala

@@ -72,7 +72,7 @@ class TestAdt extends FileDiffSuite {
         val europe = Woman(unit("Europe"), unit(36), zeus, minos)
         
         val Man = adt[Man]
-        val asterion  = Man(unit("Astérion"), unit(42), europe)
+        val asterion  = Man(unit("Asterion"), unit(42), europe)
         
         log(asterion.wife)
         
@@ -99,4 +99,4 @@ class TestAdt extends FileDiffSuite {
      
   }
 }
-             
+            

### 2013-09-02 macros/src/main/scala/scala/js/macroimpl/Adt.scala

@@ -138,7 +138,7 @@ object Adts {
 
         val paramsFold = for((param, symbol) <- allSubClasses) yield q"val $symbol: (Rep[$param] => Rep[A])"
 
-        val paramsFoldLambda = for((param, symbol) <- allSubClasses) yield q"doLambda($symbol)"  //ok (?)  si val = doLambda, erreur de type, à montrer à Olivier
+        val paramsFoldLambda = for((param, symbol) <- allSubClasses) yield q"doLambda($symbol)"
 
         val paramsFoldName = for(param <- paramsFoldLambda) yield q"$param.asInstanceOf[Rep[${U.tpe} => A]]"
 

### 2013-09-02 project/Build.scala

@@ -41,7 +41,7 @@ object JsScalaBuild extends Build {
     "core",
     file("core"),
     settings = buildSettings ++ Seq(
-      scalacOptions ++= Seq("-deprecation", "-unchecked", "-Xexperimental", "-P:continuations:enable", "-Yvirtualize", "-language:dynamics", "-Ymacro-debug-lite"),
+      scalacOptions ++= Seq("-deprecation", "-unchecked", "-Xexperimental", "-P:continuations:enable", "-Yvirtualize", "-language:dynamics"/*, "-Ymacro-debug-lite"*/),
 
       name := "js-scala",
       mappings in (Compile, packageBin) <++= mappings in (macros, Compile, packageBin),

### 2013-09-03 papers/gpce2013/gpce2013.tex
Clearer separation with ECOOP12 in the related works. More precisions in the dom creation section

@@
 basic language concepts like arrays and control structures (\code{if} and \code{while}) as well as
-mechanisms specific to the Scala compiler, like delimited continuations to handle asynchronous
-computations. Though the initial authors of js-scala showed how to use the native browser APIs in an
-untyped fashion, they did not show how to build high-level abstractions for Web programming
-generating efficient code leveraging the browser APIs and did not show how to share such
-abstractions between server and client sides (they achieved code sharing, yet on the server-side
-their DSL was not compiled but interpreted). Our paper addresses these limitations.
+mechanisms specific to the Scala compiler like delimited continuations to handle asynchronous
+computations.
+Paper \cite{Kossakowski12_JsDESL} presented the implementation of js-scala using staging, but did not discuss any specific optimization and did not consider performance issues of this approach.
+In this paper, we show how js-scala has been extended to support a set of specific optimizations allowing our high-level abstractions for Web programming to be efficiently compiled on heterogeneous platforms.
 
@@
 \begin{figure}[htb]
-\begin{lstlisting}[language=JavaScript,label=jquery-selectors,caption=Searching elements in jQuery]
+\begin{lstlisting}[language=JavaScript,label=jquery-selectors,caption=Searching elements using jQuery]
 function getWords() {
@@
 \begin{figure}[htb]
-\begin{lstlisting}[language=JavaScript,caption=JavaScript DOM API,label=dom-api]
+\begin{lstlisting}[language=JavaScript,caption=JavaScript DOM creation native API,label=dom-api]
 var articleUi = function (article) {
@@
 defines a JavaScript function \code{articleUi} that builds a DOM tree containing an article
-description, and Listing \ref{scala-xml-api} shows how one could implement a similar function on
-server-side using the standard Scala XML library. The reader may notice that the client-side
-and server-side APIs are very different and that the client-side API is very low-level and
-inconvenient to use.
+description.
+Listing \ref{scala-xml-api} shows how one could implement a similar function on server-side using the standard Scala XML library.
+The reader may notice that the client-side and server-side APIs are very different and that the client-side native API is very low-level and inconvenient to use.
+We could use a library on client-side to get a higher level API for DOM fragment creation, but that would decrease the runtime performance.
+Instead, we want to define a high-level API that compiles to code as efficient as if it was written using the native APIs, on both platforms.
 
@@
 the dynamic behavior of the application: double-posting prevention, dynamic form update and rich
-interaction with the document. The size of the whole application (server and client sides) is about one thousand of lines of code.
+interaction with the document. The size of the whole application (server and client sides) is about one thousand lines of code.
 

### 2013-09-03 papers/gpce2013/gpce2013.tex
Reword the introduction to make the contribution clearer

@@
 generates JavaScript code, making it possible to write the client-side code of Web applications
-using Scala~\cite{Kossakowski12_JsDESL}. However, Kossakowski \etal did not address the
-engineering dilemma described above. This paper enriches js-scala with the following
-contributions\footnote{The code is available at
-\href{http://github.com/js-scala}{http://github.com/js-scala}}:
+using Scala~\cite{Kossakowski12_JsDESL}.
+However, the authors did not discuss any specific optimization and did not consider performance issues of their approach.
+Our paper shows how js-scala has been extended to support a set of specific optimizations allowing our high-level abstractions for Web programming to be efficiently compiled on both client and server sides\footnote{The code is available at \href{http://github.com/js-scala}{http://github.com/js-scala}}.
 
-\begin{itemize}
- \item we define high-level abstractions, typically used in Web programming, that generate low-level
-code leveraging native Web browser APIs;
- \item in the case of abstractions shared between servers and clients, we specialize the code
-generation in order to leverage the target platform native environments.
-\end{itemize}
+% \begin{itemize}
+%  \item we define high-level abstractions, typically used in Web programming, that generate low-level code leveraging native Web browser APIs;
+%  \item in the case of abstractions shared between servers and clients, we specialize the code generation in order to leverage the target platform native environments.
+% \end{itemize}
 
@@
 We could use a library on client-side to get a higher level API for DOM fragment creation, but that would decrease the runtime performance.
-Instead, we want to define a high-level API that compiles to code as efficient as if it was written using the native APIs, on both platforms.
+Instead, we want to define a high-level API that compiles to code as efficient as if it was written using the native APIs on both platforms.
 

### 2013-09-03 core/src/test/scala/scala/js/TestAdts.scala
Code style

@@ -7,7 +7,7 @@ import gen.js.{GenEffect, GenDebug, GenAdts, GenListOps}
 import language.{Debug, Adts}
 import exp.{DebugExp, AdtsExp}
 
-class TestAdt extends FileDiffSuite {
+class TestAdts extends FileDiffSuite {
 
   def testAdt() {
 
@@ -76,14 +76,13 @@ class TestAdt extends FileDiffSuite {
         
         log(asterion.wife)
         
-        
-        def hello(p: Rep[Person]) = p.Fold(
-            (m: Rep[Mutant]) => log(m.name),
-            (g: Rep[God]) => log(g.name),
-            (d: Rep[Devil]) => log(d.name),
-            (m: Rep[Man]) => log(m.name),
-            (w: Rep[Woman]) => log(w.name)
-            )
+        def hello(p: Rep[Person]) = p.fold(
+          (m: Rep[Mutant]) => log(m.name),
+          (g: Rep[God]) => log(g.name),
+          (d: Rep[Devil]) => log(d.name),
+          (m: Rep[Man]) => log(m.name),
+          (w: Rep[Woman]) => log(w.name)
+        )
         
         log(hello(minos))
         

### 2013-09-03 macros/src/main/scala/scala/js/macroimpl/Adts.scala

@@ -30,40 +30,34 @@ object Adts {
     }
 
     /** @return The members of the type `tpe` */
-    def listMembers(tpe: Type): List[Member] = {
-      val fields = tpe.typeSymbol.typeSignature.declarations.toList.collect{ case x: TermSymbol if x.isVal && x.isCaseAccessor => Member(x) }
-      fields
-    }
+    def listMembers(tpe: Type): List[Member] =
+      tpe.typeSymbol.typeSignature.declarations.toList.collect { case x: TermSymbol if x.isVal && x.isCaseAccessor => Member(x) }
 
     def getVariant(it: Iterator[Symbol], s: Symbol, i: Int, find: Boolean): (Int, Boolean) = {
-          val result = if(!it.hasNext){
-              (i, false)
+      val result = if (!it.hasNext) {
+        (i, false)
+      } else {
+        val sc = it.next
+        val k = if (s == sc) {
+          (i, true)
+        } else {
+          val nextI: (Int, Boolean) = if (sc.asClass.isTrait) {
+            getVariant(sc.asClass.knownDirectSubclasses.toList.init.iterator, s, i, false)
+          } else {
+            (i+1, false)
+          }
 
-            }else{
-              val sc = it.next
+          if (nextI._2) {
+            nextI
+          } else {
+            getVariant(it, s, nextI._1, false)
+          }
 
-              val k = if (s == sc){
-                (i, true)
-
-                }else{
-                  val nextI: (Int, Boolean) = if (sc.asClass.isTrait){
-                      getVariant(sc.asClass.knownDirectSubclasses.toList.init.iterator, s, i, false)
-
-                    }else{
-                      (i+1, false)
-                    }
-
-                  if(nextI._2){
-                    nextI
-                  }else{
-                    getVariant(it, s, nextI._1, false)
-                  }
-
-                }
-              k
-            }
-          result
         }
+        k
+      }
+      result
+    }
 
 
     def ops[U: c.WeakTypeTag](obj: c.Expr[U]) = {
@@ -118,23 +112,14 @@ object Adts {
         val s = c.weakTypeOf[U].typeSymbol.asClass
         s.typeSignature // SI-7046
 
-        val subClasses = s.knownDirectSubclasses
-        
-        def getSubClasses (subClasses: List[Symbol], list: List[Symbol]): List[Symbol] = {
-            if(subClasses.isEmpty){
[…]

### 2013-09-04 core/src/main/scala/scala/js/exp/AdtsExp.scala
Fix fold implementation. Add some test cases. Add comments when implementation is wrong.

@@ -1,9 +1,9 @@
 package scala.js.exp
 
-import scala.virtualization.lms.common.{BaseExp, EqualExp, TupledFunctionsExp}
+import scala.virtualization.lms.common.{TupledFunctionsExp, BaseExp}
 import scala.js.language.Adts
 
-trait AdtsExp extends BaseExp with Adts with TupledFunctionsExp with EqualExp {
+trait AdtsExp extends Adts with BaseExp with TupledFunctionsExp {
   
   def adt_construct[A : Manifest](fields: (String, Exp[_])*) = {
     AdtApply[A](fields)

### 2013-09-04 core/src/main/scala/scala/js/gen/js/GenAdts.scala

@@ -2,7 +2,7 @@ package scala.js.gen.js
 
 import scala.js.exp.AdtsExp
 
-trait GenAdts extends GenBase with GenFunctions with GenEqual {
+trait GenAdts extends GenBase with GenFunctions {
   val IR: AdtsExp
   import IR._
 
@@ -29,7 +29,7 @@ trait GenAdts extends GenBase with GenFunctions with GenEqual {
         emitValDef(sym, quote(obj) + "==" + quote(bis))
       }
     case AdtFold(obj, fs) => {
-      emitValDef(sym, "function("+quote(obj)+", "+fs.map(quote).mkString(",")+"){ ["+fs.map(quote).mkString(",")+"]["+quote(obj)+".$variant]("+quote(obj)+"); }")
+      emitValDef(sym, "["+fs.map(quote).mkString(",")+"]["+quote(obj)+".$variant]("+quote(obj)+")")
     }
     case _ => 
       super.emitNode(sym, rhs)

### 2013-09-04 core/src/main/scala/scala/js/language/Adts.scala

@@ -1,7 +1,7 @@
 package scala.js.language
 
 import scala.language.experimental.macros
-import scala.virtualization.lms.common.{Base, Equal}
+import scala.virtualization.lms.common.Functions
 
 /**
  * Reifies case classes as staged adts
@@ -20,7 +20,7 @@ import scala.virtualization.lms.common.{Base, Equal}
  *     Point(p1.x + p2.x, p1.y + p2.y)
  * }}}
  */
-trait Adts extends Base with Equal {
+trait Adts extends Functions {
 
   type Adt = scala.js.macroimpl.Adts.Adt
 
@@ -38,7 +38,7 @@ trait Adts extends Base with Equal {
    * 
    * @return a staged smart constructor for the data type T
    */
-  def adt[T] = macro scala.js.macroimpl.Adts.adt[T]
+  def adt[T <: Adt] = macro scala.js.macroimpl.Adts.adt[T]
   
   /**
    * {{{
@@ -51,6 +51,6 @@ trait Adts extends Base with Equal {
    *
    * @return an object with staged members for the type T
    */
-  def adtOps[T](o: Rep[T]) = macro scala.js.macroimpl.Adts.ops[T]
+  def adtOps[T <: Adt](o: Rep[T]) = macro scala.js.macroimpl.Adts.ops[T]
 
 }

### 2013-09-04 core/src/test/scala/scala/js/TestAdts.scala

@@ -9,12 +9,174 @@ import exp.{DebugExp, AdtsExp}
 
 class TestAdts extends FileDiffSuite {
 
+  val prefix = "test-out/"
+
+  def testProducts(): Unit = {
+
+    trait Prog extends Adts {
+
+      case class Product(x: Int, s: String) extends Adt
+      case class NestedProduct(p: Product, b: Boolean) extends Adt
+
+      // Smart constructors
+      object C {
+        val Product = adt[Product]
+        val NestedProduct = adt[NestedProduct]
+      }
+
+      // Methods
+      implicit def ProductOps(p: Rep[Product]) = adtOps(p)
+      implicit def NestedProductOps(n: Rep[NestedProduct]) = adtOps(n)
+
+      def construction(x: Rep[Int], s: Rep[String], b: Rep[Boolean]): Rep[NestedProduct] = {
+        val p = C.Product(x, s)
+        C.NestedProduct(p, b)
+      }
+
+      def memberSelection(n: Rep[NestedProduct]) = n.p
+
+      def equal(n1: Rep[NestedProduct], n2: Rep[NestedProduct]) = n1 === n2
+
+      def copy(n: Rep[NestedProduct], b: Rep[Boolean]) = n.copy(p = n.p, b = b)
+
+    }
+
+    withOutFile(prefix + "adt/product") {
+      val prog = new Prog with AdtsExp
+      val gen = new GenAdts { val IR: prog.type = prog }
+      val out = new PrintWriter(System.out)
+      gen.emitSource3(prog.construction, "construction", out)
+      gen.emitSource(prog.memberSelection, "selection", out)
+      out.println("ERROR Return type of member selection is not correct")
+      gen.emitSource2(prog.equal, "equal", out)
+      out.println("ERROR Return type of === is not correct")
+      gen.emitSource2(prog.copy, "copy", out)
+    }
+    assertFileEqualsCheck(prefix + "adt/product")
+  }
+
+  def testSums(): Unit = {
+
+    trait Prog extends Adts {
+
+      sealed trait CoProduct extends Adt
+      case class Left(x: Int) extends CoProduct
+      case class Right(s: String) extends CoProduct
+
+      object C {
+        val Left = adt[Left]
+        val Right = adt[Right]
+      }
+
+      implicit def CoProductOps(c: Rep[CoProduct]) = adtOps(c)
[…]

### 2013-09-04 macros/src/main/scala/scala/js/macroimpl/Adts.scala

@@ -8,11 +8,11 @@ object Adts {
   
   trait Adt
 
-  def adt[U: c.WeakTypeTag](c: Context) =
+  def adt[U <: Adt : c.WeakTypeTag](c: Context) =
     c.Expr[Any](new Generator[c.type](c).construct[U])
 
 
-  def ops[U: c.WeakTypeTag](c: Context)(o: c.Expr[U]) =
+  def ops[U : c.WeakTypeTag](c: Context)(o: c.Expr[U]) =
     c.Expr[Any](new Generator[c.type](c).ops(o))
 
 

### 2013-09-09 papers/gpce2013/gpce2013.tex
mention coffeescript, mention micro-benchmarks, update the figures

@@
 We validate our approach with a case study implemented with various candidate technologies and
-discuss the relative pro and cons of them. Though the code written in our DSL is high-level and can
+discuss the relative pro and cons of them. We also measured the individual impact of each of our optimizations using micro-benchmarks. Though the code written in our DSL is high-level and can
 be shared between clients and servers, it has the same runtime performance on client-side as
@@
 programming concepts such as naming, functions, classes, as well as more domain-specific concepts
-such as HTML fragment definition. Thus, implementing a fat language may require a high effort.
-Examples of such languages for Web programming are Links~\cite{Cooper07_Links}, Opa and
-Dart~\cite{Griffith11_Dart}.
+such as HTML fragment definition. Thus, implementing a fat language may require a high effort and adding support for these languages in development environments may require a even higher effort.
+Examples of such languages for Web programming are Links~\cite{Cooper07_Links}, Opa, Dart~\cite{Griffith11_Dart}.
 
@@
 and stop further JavaScript execution. Defensive code can be written to handle \code{null}
-references, but leads to very cumbersome code, as shown in Listing \ref{null-defensive}.
+references, but leads to very cumbersome code, as shown in Listing \ref{null-defensive}.\footnote{However, one could alleviate the syntax burden by using a language such as CoffeeScript~\cite{Ashkenas2011_CoffeeScript}, that suppports a special notation for optional values dereferencing and desugars directly to JavaScript.}
 

### 2013-09-10 papers/gpce2013/gpce2013.tex
more precise indications to find the code of chooze

@@
 Chooze~\footnote{Source code is available at
-\href{http://github.com/julienrf/chooze}{http://github.com/julienrf/chooze}} is an existing
+\href{http://github.com/julienrf/chooze}{http://github.com/julienrf/chooze}, under the branches \code{vanilla}, \code{jquery}, \code{gwt}, \code{haxe} and \code{js-scala}} is an existing
 complete application for making polls. It allows users to create a poll, define the choice

### 2013-09-11 papers/gpce2013/gpce2013.tex
abstract reworded

@@
 abstracting over the clients and servers differences without trading this ease of engineering for
-performance? This article presents high-level abstractions defined as deep embedded DSLs in Scala,
-that can (1) generate efficient code leveraging the target platform characteristics, (2) be shared
-between clients and servers. Though code written with our DSL has a high level of abstraction, our
+performance?
+This article presents high-level abstractions defined as deep embedded DSLs in Scala that can generate efficient code leveraging the characteristics of both client and server environments.
+We compare performance on client-side against other candidate technologies and against hand written low-level JavaScript code.
+Though code written with our DSL has a high level of abstraction, our
 benchmark on a real world application reports that it runs as fast as hand tuned low-level

### 2013-11-04 papers/gpce2013/gpce19c-foy.tex
update paper

--- 2013-09-11 papers/gpce2013/gpce2013.tex
@@
-\documentclass[preprint]{sigplanconf}
+\documentclass{sigplanconf}
 
@@
 
-\input{macros}
+\usepackage{listings}
+%\usepackage[scaled]{luximono}
+
+%\usepackage{natbib}
+
+% ----- begin macros
+
+\lstdefinelanguage{Scala}%
+{morekeywords={abstract,%
+  case,catch,char,class,%
+  def,else,extends,final,for,%
+  if,import,implicit,%
+  match,module,%
+  new,null,%
+  object,override,%
+  package,private,protected,public,%
+  for,public,return,super,%
+  this,throw,trait,try,type,%
+  val,var,%
+  with,%
+  yield,%
+  lazy%
+  },%
+  sensitive,%
+  morecomment=[l]//,%
+  morecomment=[s]{/*}{*/},%
+  morestring=[b]",%
+  morestring=[b]',%
+  showstringspaces=false%
+}[keywords,comments,strings]%
+
+\lstdefinelanguage{JavaScript}%
+{morekeywords={for, var, attributes, class, classend, do, else, empty, endif, endwhile, fail, function,
+functionend, if, implements, in, inherit, inout, not, of, operations, out, return,
+then, types, while, use},%
+  sensitive,%
+  morecomment=[l]//,%
+  morecomment=[s]{/*}{*/},%
+  morestring=[b]",%
+  morestring=[b]',%
+  showstringspaces=false%
+}[keywords,comments,strings]%
+
+\lstset{language=Scala,%
+  mathescape=false,%
+%  columns=[c]fixed,%
+  aboveskip=\smallskipamount,
+  belowskip=\smallskipamount,
+%  basewidth={0.5em, 0.4em},%
+  basicstyle=\footnotesize,%
+  captionpos=b,
+  keywordstyle=\bfseries%\sffamily\bfseries,%
+%  keywordstyle=\sffamily\bfseries,%
+%  xleftmargin=0.5cm
+}
+
+\newcommand{\commentstyle}[1]{\slseries{#1}}
+\newcommand{\keywordstyle}[1]{\bfseries{#1}}
+
+\lstnewenvironment{slisting}{\lstset{language=Scala}}{}
+
+\newcommand{\code}[1]{\lstinline[language=Scala,columns=fixed,basicstyle=\footnotesize]|#1|}
+
+\def\changemargin#1#2{\list{}{\rightmargin#2\leftmargin#1}\item[]}
+\let\endchangemargin=\endlist
 
@@
 
-\conferenceinfo{GPCE '13}{October 27--28, 2013, Indianapolis, Indiana, USA} 
-\copyrightyear{2013}
-\copyrightdata{[to be supplied]}
+\conferenceinfo{GPCE~'13}{October 27--28, 2013, Indianapolis, Indiana, USA} 
+\copyrightyear{2013} \copyrightdata{978-1-4503-2373-4/13/10} \doi{2517208.2517227}
 
@@
 
-\terms Languages, Software Engineering
+%\terms Languages, Software Engineering
 
@@
 \begin{center}
-\includegraphics[width=7.5cm]{langs.pdf}
+\includegraphics[width=7.5cm]{gpce19c-foy-langs.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=5cm]{ir.pdf}
+\includegraphics[width=5cm]{gpce19c-foy-ir.pdf}
 \end{center}
@@
 \centering
-\includegraphics[width=8cm]{selectors-benchmark.png}
+\includegraphics[width=8cm]{gpce19c-foy-selectors-benchmark.png}
 \caption{Micro-benchmark on the selectors abstraction}
@@
 \centering
-\includegraphics[width=8cm]{microbenchmark.png}
+\includegraphics[width=8cm]{gpce19c-foy-microbenchmark.png}
 \caption{Micro-benchmark on the optional values abstraction}
@@
 \centering
-\includegraphics[width=8cm]{chooze.png}
+\includegraphics[width=8cm]{gpce19c-foy-chooze.png}
 \caption{Benchmarks on a real application}
@@
 \bibliographystyle{abbrvnat}
-\bibliography{biblio}
+\begin{thebibliography}{24}
+\providecommand{\natexlab}[1]{#1}
+\providecommand{\url}[1]{\texttt{#1}}
+\expandafter\ifx\csname urlstyle\endcsname\relax
+  \providecommand{\doi}[1]{doi: #1}\else
+  \providecommand{\doi}{doi: \begingroup \urlstyle{rm}\Url}\fi
+
+\bibitem[Ashkenas(2011)]{Ashkenas2011_CoffeeScript}
+J.~Ashkenas.
+\newblock Coffeescript, 2011.
+
+\bibitem[Bibeault and Kats(2008)]{Bibeault08_jQuery}
+B.~Bibeault and Y.~Kats.
+\newblock \emph{jQuery in Action}.
+\newblock Dreamtech Press, 2008.
+
+\bibitem[Brown et~al.(2011)Brown, Sujeeth, Lee, Rompf, Chafi, Odersky, and
+  Olukotun]{Brown11_Parallel}
+K.~J. Brown, A.~K. Sujeeth, H.~J. Lee, T.~Rompf, H.~Chafi, M.~Odersky, and
+  K.~Olukotun.
+\newblock A heterogeneous parallel framework for domain-specific languages.
+\newblock In \emph{Parallel Architectures and Compilation Techniques (PACT),
+  2011 International Conference on}, pages 89--100. IEEE, 2011.
+
+\bibitem[Cannasse(2008)]{Cannasse08_HaXe}
+N.~Cannasse.
+\newblock Using haxe.
+\newblock \emph{The Essential Guide to Open Source Flash Development}, pages
+  227--244, 2008.
+
+\bibitem[Chaganti(2007)]{Chaganti07_GWT}
+P.~Chaganti.
+\newblock \emph{Google Web Toolkit: GWT Java Ajax Programming}.
+\newblock Packt Pub Limited, 2007.
+
+\bibitem[Cooper et~al.(2007)Cooper, Lindley, Wadler, and
+  Yallop]{Cooper07_Links}
+E.~Cooper, S.~Lindley, P.~Wadler, and J.~Yallop.
+\newblock Links: Web programming without tiers.
+\newblock In \emph{Formal Methods for Components and Objects}, pages 266--296.
+  Springer, 2007.
+
+\bibitem[Elliott et~al.(2003)Elliott, Finne, and
+  De~Moor]{Elliott2003_Compiling}
+C.~Elliott, S.~Finne, and O.~De~Moor.
+\newblock Compiling embedded languages.
+\newblock \emph{Journal of Functional Programming}, 13\penalty0 (3):\penalty0
+  455--481, 2003.
+
+\bibitem[Griffith(2011)]{Griffith11_Dart}
+R.~Griffith.
+\newblock The dart programming language for non-programmers-overview.
+\newblock 2011.
+
+\bibitem[Hoare(2009)]{Hoare09_Null}
+T.~Hoare.
+\newblock Null references: The billion dollar mistake.
+\newblock \emph{Presentation at QCon London}, 2009.
+
+\bibitem[Huang et~al.(2010)Huang, Xu, Tiwana, Mao, Zhang, and
+  Bahl]{Huang10_Perf}
+J.~Huang, Q.~Xu, B.~Tiwana, Z.~M. Mao, M.~Zhang, and P.~Bahl.
+\newblock Anatomizing application performance differences on smartphones.
+\newblock In \emph{Proceedings of the 8th international conference on Mobile
+  systems, applications, and services}, pages 165--178. ACM, 2010.
+
+\bibitem[Hudak(1996)]{Hudak96_DSEL}
+P.~Hudak.
+\newblock Building domain-specific embedded languages.
+\newblock \emph{ACM Computing Surveys}, 28, 1996.
+\newblock URL \url{http://dsel.ps}.
+
+\bibitem[J{\o}rring and Scherlis(1986)]{Jorring1986_Staging}
+U.~J{\o}rring and W.~L. Scherlis.
+\newblock Compilers and staging transformations.
+\newblock In \emph{Proceedings of the 13th ACM SIGACT-SIGPLAN symposium on
+  Principles of programming languages}, pages 86--96. ACM, 1986.
+
+\bibitem[Kossakowski et~al.(2012)Kossakowski, Amin, Rompf, and
+  Odersky]{Kossakowski12_JsDESL}
+G.~Kossakowski, N.~Amin, T.~Rompf, and M.~Odersky.
+\newblock Java{S}cript as an {E}mbedded {DSL}.
+\newblock In J.~Noble, editor, \emph{{ECOOP} 2012 -- {O}bject-{O}riented
+  {P}rogramming}, volume 7313 of \emph{Lecture Notes in Computer Science},
+  pages 409--434, Berlin, Heidelberg, 2012. Springer Berlin Heidelberg.
+\newblock \doi{10.1007/978-3-642-31057-7_19}.
+\newblock URL \url{https://github.com/js-scala/js-scala/}.
+
+\bibitem[Kuuskeri and Mikkonen(2009)]{Kuuskeri09_PartitioningClientServer}
+J.~Kuuskeri and T.~Mikkonen.
+\newblock Partitioning web applications between the server and the client.
+\newblock In \emph{Proceedings of the 2009 ACM symposium on Applied Computing},
+  SAC '09, pages 647--652, New York, NY, USA, 2009. ACM.
+\newblock ISBN 978-1-60558-166-8.
+\newblock \doi{10.1145/1529282.1529416}.
+\newblock URL \url{http://doi.acm.org/10.1145/1529282.1529416}.
+
+\bibitem[McGranaghan(2011)]{McGranaghan11_ClojureScript}
+M.~McGranaghan.
+\newblock Clojurescript: Functional programming for javascript platforms.
+\newblock \emph{Internet Computing, IEEE}, 15\penalty0 (6):\penalty0 97--102,
+  2011.
+
+\bibitem[Mikkonen and Taivalsaari(2008)]{Mikkonen08_SpaghettiJs}
+T.~Mikkonen and A.~Taivalsaari.
+\newblock Web applications - spaghetti code for the 21st century.
+\newblock In \emph{Proceedings of the 2008 Sixth International Conference on
+  Software Engineering Research, Management and Applications}, pages 319--328,
+  Washington, DC, USA, 2008. IEEE Computer Society.
+\newblock ISBN 978-0-7695-3302-5.
+\newblock \doi{10.1109/SERA.2008.16}.
+\newblock URL \url{http://dl.acm.org/citation.cfm?id=1443226.1444030}.
+
+\bibitem[Nanda and Sinha(2009)]{Nanda09_Null}
+M.~Nanda and S.~Sinha.
+\newblock Accurate interprocedural null-dereference analysis for java.
+\newblock In \emph{Software Engineering, 2009. ICSE 2009. IEEE 31st
+  International Conference on}, pages 133--143. IEEE, 2009.
+
+\bibitem[Preciado et~al.(2005)Preciado, Trigueros, S{\'a}nchez-Figueroa, and
+  Comai]{Preciado05_RIAMethodologyNecessity}
+J.~C. Preciado, M.~L. Trigueros, F.~S{\'a}nchez-Figueroa, and S.~Comai.
+\newblock Necessity of methodologies to model rich internet applications.
+\newblock In \emph{WSE}, pages 7--13. IEEE Computer Society, 2005.
+\newblock ISBN 0-7695-2470-2.
+
+\bibitem[Rodr{\'\i}guez-Echeverr{\'\i}a(2009)]{Echeverria09_RIA}
+R.~Rodr{\'\i}guez-Echeverr{\'\i}a.
+\newblock Ria: more than a nice face.
+\newblock In \emph{Proceedings of the Doctolral Consortium of the International
+  Conference on Web Engineering}, volume 484. CEUR-WS.org, 2009.
+
+\bibitem[Rompf(2012)]{Rompf12_LMSThesis}
+T.~Rompf.
+\newblock \emph{Lightweight Modular Staging and Embedded Compilers: Abstraction
+  without Regret for High-Level High-Performance Programming}.
+\newblock PhD thesis, {\'E}COLE POLYTECHNIQUE F{\'E}D{\'E}RALE DE LAUSANNE,
+  2012.
+
+%\vfill\eject
+
+\bibitem[Rompf et~al.(2012)Rompf, Sujeeth, Amin, Brown, Jovanovic, Lee,
+  Jonnalagedda, Olukotun, and Odersky]{Rompf12_LMS}
+T.~Rompf, A.~Sujeeth, N.~Amin, K.~Brown, V.~Jovanovic, H.~Lee, M.~Jonnalagedda,
+  K.~Olukotun, and M.~Odersky.
+\newblock Optimizing {D}ata {S}tructures in {H}igh-{L}evel {P}rograms: {N}ew
+  {D}irections for {E}xtensible {C}ompilers based on {S}taging.
+\newblock Technical report, 2012.
+
+\bibitem[Souders(2008)]{Souders08_Perf}
+S.~Souders.
+\newblock High-performance web sites.
+\newblock \emph{Communications of the ACM}, 51\penalty0 (12):\penalty0 36--41,
+  2008.
+
+\bibitem[Van~Deursen et~al.(2000)Van~Deursen, Klint, and Visser]{Van00_DSL}
+A.~Van~Deursen, P.~Klint, and J.~Visser.
+\newblock Domain-specific languages: an annotated bibliography.
+\newblock \emph{ACM Sigplan Notices}, 35\penalty0 (6):\penalty0 26--36, 2000.
+
+\bibitem[Visser(2007)]{Visser07}
+E.~Visser.
+\newblock {WebDSL}: {A} case study in domain-specific language engineering.
+\newblock In R.~Lämmel, J.~Visser, and J.~Saraiva, editors, \emph{Generative
+  and Transformational Techniques in Software Engineering II, International
+  Summer School, {GTTSE} 2007}, volume 5235 of \emph{Lecture Notes in Computer
+  Science}, pages 291--373, Braga, Portugal, 2007. Springer.
+\newblock ISBN 978-3-540-88642-6.
+\newblock \doi{http://dx.doi.org/10.1007/978-3-540-88643-3_7}.
+
+\end{thebibliography}
 %\begin{thebibliography}{}

### 2013-12-30 papers/icwe14/icwe14.tex
icwe14 wip

\documentclass{llncs}

\usepackage[utf8]{inputenc}
\usepackage[unicode=true,hidelinks]{hyperref}
\usepackage{listings}
\usepackage{lmodern}

\lstdefinelanguage{Scala}%
{morekeywords={abstract,%
  case,catch,char,class,%
  def,else,extends,final,for,%
  if,import,implicit,%
  match,module,%
  new,null,%
  object,override,%
  package,private,protected,public,%
  for,public,return,super,%
  this,throw,trait,try,type,%
  val,var,%
  with,%
  yield,%
  lazy%
  },%
  sensitive,%
  morecomment=[l]//,%
  morecomment=[s]{/*}{*/},%
  morestring=[b]",%
  morestring=[b]',%
  showstringspaces=false%
}[keywords,comments,strings]%

\lstdefinelanguage{JavaScript}%
{morekeywords={for, var, attributes, class, classend, do, else, empty, endif, endwhile, fail, function,
functionend, if, implements, in, inherit, inout, not, of, operations, out, return,
then, types, while, use},%
  sensitive,%
  morecomment=[l]//,%
  morecomment=[s]{/*}{*/},%
  morestring=[b]",%
  morestring=[b]',%
  showstringspaces=false%
}[keywords,comments,strings]%

\lstset{language=Scala,%
%  mathescape=false,%
%  columns=[c]fixed,%
%  aboveskip=\smallskipamount,
%  belowskip=\smallskipamount,
%  basewidth={0.5em, 0.4em},%
  basicstyle=\small\ttfamily,
  captionpos=b,
  keywordstyle=\bfseries
%  xleftmargin=0.5cm
}

\newcommand{\jscode}[1]{\lstinline[language=JavaScript]|#1|}
\newcommand{\scalacode}[1]{\lstinline[language=Scala]|#1|}
%\newcommand{\code}[1]{\texttt{#1}}

\begin{document}

 \title{Using Path-Dependent Types to Build Type-Safe JavaScript Foreign Function Interfaces}

 \author{Julien Richard-Foy \and Olivier Barais}

 \institute{IRISA, Universite de Rennes, France}

 %\email{\{first\}.\{last\}@irisa.fr}

 \maketitle

\begin{abstract}

\end{abstract}

\section{Introduction}

We recently observed the emergence of several statically typed programming languages compiling to JavaScript (\emph{e.g.} Java/GWT~\cite{Kereki09_GWT}, Dart~\cite{Griffith11_Dart}, TypeScript~\cite{fenton2012typescript}, Kotlin\footnote{\href{http://kotlin.jetbrains.org}{http://kotlin.jetbrains.org}}, Opa\footnote{\href{http://opalang.org}{http://opalang.org}}, SharpKit\footnote{\href{http://sharpkit.net}{http://sharpkit.net}}, Haxe~\cite{Cannasse08_HaXe}, Idris~\cite{Brady13_Idris}, Elm~\cite{czaplicki2012elm}). Developers can use them to write the client-side part of their Web applications, relying on a \emph{foreign function interface} mechanism to call the Web browser native API.

However we observe that, despite these languages are statically typed, their integration of the browser API either is not type safe or gives less expression power to developers. Indeed, integrating an API designed for a dynamically typed language into a statically typed language can be challenging. For instance, the \jscode{createElement} function return type depends on the value of the \jscode{String} parameter it is passed.

This paper reviews some common functions of the browser API, identifies those which are not well encoded in statically typed programming languages and shows new ways to encode them with more type safety while keeping the native API expression power. We find that type parameters are sufficient to achieve this goal and that \emph{path-dependent types} provide an even more convenient encoding for the end developers.

The remainder of the paper is organized as follows. The next section reviews the most common functions of the browser API and how they are integrated in statically typed languages. Section \ref{sec-contribution} shows ways to improve their integration. Section \ref{sec-validation} validates our contribution. Section \ref{sec-related} discusses some related works and section \ref{sec-conclusion} concludes.

\section{Background}
\label{sec-background}

This section reviews the most commonly used browser functions and presents the way they are integrated in GWT, Dart, TypeScript, Kotlin, Opa, SharpKit, Haxe, Idris and Elm.

\subsection{The browser API and its integration in statically typed languages}

The client-side part of the code of a Web application essentially reacts to user events (\emph{e.g.} mouse clicks), triggers actions and updates the document (DOM) according to their effect. Table~\ref{table-dom-api} lists the main functions supported by the Web browsers according to the Mozilla Developer Network\footnote{\href{https://developer.mozilla.org/en-US/docs/DOM/DOM\_Reference/Introduction}{https://developer.mozilla.org/en-US/docs/DOM/DOM\_Reference/Introduction}} (we removed the functions that can trivially be encoded in a static type system).

\begin{table}
 \centering
 \begin{tabular}{ll}
  \hline
  Name & Description \\
  \hline
  \jscode{getElementsByTagName(name)} & Find elements by their tag name \\
  \jscode{getElementById(id)} & Find an element by its \jscode{id} attribute \\
  \jscode{createElement(name)} & Create an element \\
  \jscode{target.addEventListener(name, listener)} & React to events \\
  \hline
 \end{tabular}

 \label{table-dom-api}
 \caption{Web browsers main functions}
\end{table}

The main challenge comes from the fact that these functions are polymorphic in their return type or in their parameters.

For instance, functions \jscode{getElementsByTagName(name)}, \jscode{getElementById(id)} and \jscode{createElement(name)} can return values of type \jscode{DivElement} or \jscode{InputElement} or any other subtype of \jscode{Element} (their least upper bound). The interface of \jscode{Element} is more general and provides less features than its subtypes, so it is important for the encoding of these functions in a statically typed language to be as precise as possible.

In the case of \jscode{getElementById(id)}, the \jscode{id} parameter gives no clue on the possible type of the searched element so it is hard to infer more precisely the return type of this function. Hence, most implementations define a return type of \jscode{Element}. The drawback of this approach is that developers may get a value with a type weaker than they need and may require to explicitly downcast it, thus loosing type safety.

However, in the case of \jscode{getElementsByTagName(name)} and \jscode{createElement(name)}, there is exactly one possible return type for each value of the \jscode{name} parameter: \emph{e.g.} \jscode{getElementsByTagName('input')} always returns a list of \jscode{InputElement} and \jscode{createElement('div')} always returns a \jscode{DivElement}. This characteristic makes it possible to encode each of these functions by defining as many parameterless functions as there are possible tag names, where each function fixes the initial \jscode{name} parameter to be one of the possible values and exposes the corresponding specialized return type. Listing~\ref{lst-create-element-overload} illustrates such an encoding in the Java language.

\begin{figure}
\begin{lstlisting}[label=lst-create-element-overload,language=Java,caption={Encoding of the \jscode{createElement(name)} function in Java using parameterless functions fixing the initial \jscode{name} parameter value}]
DivElement createDivElement();
InputElement createInputElement();
FormElement createFormElement();
...
\end{lstlisting}
\end{figure}

The case of \jscode{target.addEventListener(name, listener)} is a bit different. The \jscode{name} parameter defines the event to listen to and the \jscode{listener} parameter the function to call back each time such an event occurs. Instead of being polymorphic in its return type, it is polymorphic in its \jscode{listener} parameter. Nevertheless, a similar property as above holds: there is exactly one possible type for the \jscode{listener} parameter for each value of the \jscode{name} parameter. For instance, a listener of \jscode{'click'} events is a function taking a \jscode{MouseEvent} parameter, a listener of \jscode{'keydown'} events is a function taking a \jscode{KeyboardEvent} parameter, and so on. The same pattern as above (defining a set of functions fixing the \jscode{name} parameter value) can be used to encode this function in a statically typed language, but an alternative encoding could be to define one function taking one parameter carrying both the information of the event name and the event listener. Listing~\ref{lst-add-event-listener-one-param} shows such an encoding in Java.

\begin{figure}
\begin{lstlisting}[label=lst-add-event-listener-one-param,language=Java,caption={Encoding of the \jscode{target.addEventListener(name, listener)} function in Java using one paremeter carrying both the information of the event name and the event listener}]
// --- API Definition
interface EventTarget {
  void addEventListener(EventListener listener);
}

interface EventListener {}

interface ClickEventListener extends EventListener {
  void onClick(ClickEvent event);
}

interface KeyDownEventListener extends EventListener {
  void onKeyDown(KeyboardEvent event);
}

// --- Usage
void logClicks(EventTarget target) {
  target.addEventListener(new ClickEventListener {
    public void onClick(ClickEvent event) {
      console.log(event.button);
    }
  });
}
\end{lstlisting}
\end{figure}

Table~\ref{table-existing-encodings} summarizes, for each statically typed language, which approach is used to encode each function of the browser API.

\begin{table}
\centering
\begin{tabular}{|l|c|c|c|c|}
\hline  & \jscode{getElementById} & \jscode{getElementsByTagName} & \jscode{createElement} & \jscode{addEventListener} \\
\hline Java & lub & lub & comb & plop \\
\hline Dart & lub & lub \\
\hline TypeScript \\
\hline Haxe & lub & lub \\
\hline Opa \\
\hline Kotlin & lub & lub \\
\hline SharpKit & lub & lub \\
\hline Idris \\
\hline Elm \\
\hline
\end{tabular}
\label{table-existing-encodings}
\caption{Summary of the encodings used by existing statically typed languages}
\end{table}

\subsection{Limitations of existing encoding approaches}

We distinguished three approaches to integrate the challenging parts of the browser API into statically typed languages. This section compares these approaches in terms of type safety, expression power and convenience to use from the end developer point of view.

The first approach, consisting in using the least upper bound of all the possible types is not type safe because it sometimes requires developers to explicitly downcast values to their expected specialized type.

The second approach, consisting in defining as many functions as there are possible return types of a function, is completely type safe but can lead to combinatorial explosion in some situations. Consider for instance listing~\ref{lst-js-comb} defining a JavaScript function that both creates an element and registers an event listener when a given event occurs on it. Note that the event listener is passed both the event and the element. Encoding such a function in a statically typed language using this approach would require to create one function for each combination of tag name and event name.

\begin{figure}
\begin{lstlisting}[label=lst-js-comb,language=JavaScript]
var createAndListenTo = function (tagName, eventName, listener) {
  var element = document.createElement(tagName);
  element.addEventListener(eventName, function (event) {
    listener(event, element);
  });
};
\end{lstlisting}
\end{figure}

The third approach, consisting in combining two parameters into one parameter carrying all the required information, is type safe too, but reduces the expression power because it forbids developers to partially apply the function by supplying only one parameter. Consider for instance listing~\ref{lst-js-react} that defines a function partially applying the \jscode{addEventListener} function\footnote{This pattern is often used by functional reactive programming libraries like Rx.js~\cite{liberty2011reactive}}. Such a function is impossible to encode using this approach, in a statically typed language.

\begin{figure}
\begin{lstlisting}[label=lst-js-react,language=JavaScript]
var observe = function (target, name) {
  return function (listener) {
    target.addEventListener(name, listener);
  }
};
\end{lstlisting}
\end{figure}

In summary, browser API integration by existing statically typed languages compiling to JavaScript is either not type safe or not as modular or expressive as the underlying JavaScript API.

\section{Contribution}
\label{sec-contribution}

In this section we show how we can encode the functions presented previously, in a type safe and convenient way for developers while keeping the same expression power.

We first propose a solution in the Java mainstream language, using \emph{generics}, and show how we can improve it using \emph{path-dependent types}, in Scala.

\subsection{Parametric Polymorphism}

In all the cases where a type \scalacode{T} involved in a function depends on the value of a parameter \scalacode{p} of this function (all the aforementionned functions are in this case, excepted \jscode{getElementById}), we can encode this relationship in the type system using type parameters as follows:

\begin{enumerate}
 \item Create a parametrized class \scalacode{P<U>}
 \item Set the type of \scalacode{p} to \scalacode{P<U>}
 \item Use type \scalacode{U} instead of type \scalacode{T}
\end{enumerate}

Listing~\ref{lst-generics-dom} shows this approach applied to the \jscode{createElement} function: we created a type \scalacode{ElementName<E>}, the \scalacode{name} parameter is not a \scalacode{String} but a \scalacode{ElementName<E>}, and the function returns an \scalacode{E} instead of an \scalacode{Element}. The \scalacode{ElementName<E>} type encodes the relationship between the name of an element and the type of this element. For instance, we created a value \scalacode{Input} of type \scalacode{ElementName<InputElement>}. The last two lines shows how this API is used by end developers: by passing the \scalacode{Input} value as parameter to the function, it fixes the type parameter \scalacode{E} to \scalacode{InputElement} so the returned value has the most possible precise type.

\begin{figure}
\begin{lstlisting}[label=lst-generics-dom,language=java]
// --- API Definition
class ElementName<E> {}

<E> E createElement(ElementName<E> name);

final ElementName<InputElement> Input = new ElementName<InputElement>();
final ElementName<ImageElement> Img = new ElementName<ImageElement>();

// --- Usage
InputElement input = createElement(Input);
ImageElement img = createElement(Img);
\end{lstlisting}
\end{figure}

\scalacode{getElementsByTagName} can be encoded in very similar way, and listing~\ref{lst-generics-events} shows the encoding of the \jscode{addEventListener} function.

\begin{figure}
\begin{lstlisting}[label=lst-generics-events,language=java]
// --- API Definition
class EventName<E> {}

interface EventTarget {
  <E> void addEventListener(EventName<E> name, Function<E, Void> callback);
}

final EventName<MouseEvent> Click = new EventName<MouseEvent>();

// --- Usage
ButtonElement btn = createElement(Button);
btn.addEventListener(Click, e -> console.log(e.button))
\end{lstlisting}
\end{figure}

\subsection{Path-Dependent Types}

\begin{figure}
\begin{lstlisting}[label=lst-dt-dom]
trait ElementName {
  type Element
}

def createElement(name: ElementName): name.Element = ...
\end{lstlisting}
\end{figure}

\begin{figure}
\begin{lstlisting}[label=lst-dt-events]
trait EventName {
  type Data
}
object Click extends EventDef { type Data = MouseEvent }

def on(name: EventName)(handler: Rep[name.Data] => Rep[Unit]): Rep[Unit]

val clicks = on(Click)
clicks(event => println(event.offsetX))
\end{lstlisting}
\end{figure}

\section{Validation}
\label{sec-validation}

We show that our encodings support the challenging functions defined in listings \ref{lst-js-comb} and \ref{lst-js-react}.

Listing \ref{lst-generics-comb} and \ref{lst-generics-react} show how these functions can be implemented. They are basically a direct translation from JavaScript to Java.

\begin{figure}
\begin{lstlisting}[label=lst-generics-comb,language=java]
<A, B> void createAndListenTo(
    ElementName<A> tagName,
    EventName<B> eventName,
    Function2<A, B, Void> listener) {
  A element = document.createElement(tagName);
  element.addEventListener(eventName, event -> listener.apply(event, element));
}
\end{lstlisting}
\end{figure}

\begin{figure}
\begin{lstlisting}[label=lst-generics-react,language=java]
<A> Function<Function<A, Void>, Void> observe(EventTarget target, EventName<A> name) {
  return listener -> {
    target.addEventListener(name, listener);
  }
}
\end{lstlisting}
\end{figure}

\subsection{Limitations}

We do not help with \jscode{getElementById}.

The name of an element or the name of an event can not anymore be a value resulting of the concatenation of Strings.

\section{Related Works}
\label{sec-related}

Ravi Chugh \emph{et. al.}~\cite{Chugh12_DJS} showed how to make a subset of JavaScript statically typed using a dependent type system. They require complex type annotations to be written by developers.

\section{Conclusion}
\label{sec-conclusion}

\bibliographystyle{plain}
\bibliography{references.bib}

\end{document}

### 2014-01-02 papers/icwe14/icwe14.tex
minor update

@@
 \begin{abstract}
-
+Encoding dynamically typed APIs in statically typed languages can be challenging. We present new ways to encode them in order to preserve type safety and to keep the same expression power.
 \end{abstract}
@@
 For instance, a listener of \jscode{'click'} events is a function taking a \jscode{MouseEvent} parameter, a listener of \jscode{'keydown'} events is a function taking a \jscode{KeyboardEvent} parameter, and so on.
-The same pattern as above (defining a set of functions fixing the \jscode{name} parameter value) can be used to encode this function in a statically typed language, but an alternative encoding could be to define one function taking one parameter carrying both the information of the event name and the event listener.
-Listing~\ref{lst-add-event-listener-one-param} shows such an encoding in Java.
+The same pattern as above (defining a set of functions fixing the \jscode{name} parameter value) can be used to encode this function in a statically typed language, but an alternative encoding could be to define one function taking one parameter carrying both the information of the event name and the
+event listener. Listing~\ref{lst-add-event-listener-one-param} shows such an encoding in Java.
 
@@
 Listing~\ref{lst-generics-dom} shows this approach applied to the \jscode{createElement} function: we created a type \scalacode{ElementName<E>}, the \scalacode{name} parameter is not a \scalacode{String} but a \scalacode{ElementName<E>}, and the function returns an \scalacode{E} instead of an \scalacode{Element}.
-The \scalacode{ElementName<E>} type encodes the relationship between the name of an element and the type of this element.
+The \scalacode{ElementName<E>} type encodes the relationship between the name of an element and the type of this element\footnote{The type parameter \scalacode{E} is also called a \emph{phantom type}~\cite{leijen1999domain} because \scalacode{ElementName} values never hold a \scalacode{E} value}.
 For instance, we created a value \scalacode{Input} of type \scalacode{ElementName<InputElement>}.
@@
 
-\subsection{Path-Dependent Types}
-
-\begin{figure}
-\begin{lstlisting}[label=lst-dt-dom]
-trait ElementName {
-  type Element
-}
-
-def createElement(name: ElementName): name.Element = ...
-\end{lstlisting}
-\end{figure}
-
-\begin{figure}
-\begin{lstlisting}[label=lst-dt-events]
-trait EventName {
-  type Data
-}
-object Click extends EventDef { type Data = MouseEvent }
-
-def on(name: EventName)(handler: Rep[name.Data] => Rep[Unit]): Rep[Unit]
-
-val clicks = on(Click)
-clicks(event => println(event.offsetX))
-\end{lstlisting}
-\end{figure}
-
-\section{Validation}
-\label{sec-validation}
-
 We show that our encodings support the challenging functions defined in listings \ref{lst-js-comb} and \ref{lst-js-react}.
@@
 
+Our encoding makes it possible to implement exactly the same functions as we are able to implement in plain JavaScript, but with type safety.
+
+However, every function taking an element name or an event name as parameter must have type parameters too, making its type signature harder to read.
+
+\subsection{Path-Dependent Types}
+
+This section shows how we can remove the extra type parameters needed in the previous section by using \emph{path-dependent types}.
+Essentially, the idea is that, instead of using a type parameter, we use a type member.
+Listings \ref{lst-dt-dom} and \ref{lst-dt-events} show this encoding in Scala.
+The return type of the \scalacode{createElement} function is \scalacode{name.Element}: it refers to the \scalacode{Element} type member of its \scalacode{name} parameter.
+
+\begin{figure}
+\begin{lstlisting}[label=lst-dt-dom]
+trait ElementName {
+  type Element
+}
+object Div extends ElementName { type Element = DivElement }
+object Input extends ElementName { type Element = InputElement }
+
+def createElement(name: ElementName): name.Element
+\end{lstlisting}
+\end{figure}
+
+\begin{figure}
+\begin{lstlisting}[label=lst-dt-events]
+trait EventName {
+  type Event
+}
+object Click extends EventDef { type Event = MouseEvent }
+
+trait EventTarget {
+  def addEventListener(name: EventName)(handler: name.Event => Unit): Unit
+}
+\end{lstlisting}
+\end{figure}
+
+Implementing listings \ref{lst-js-comb} and \ref{lst-js-react} using this encoding is straightforward, as shown by listings \ref{lst-dt-comb} and \ref{lst-dt-react}, respectively.
+
+\begin{figure}
+\begin{lstlisting}[label=lst-dt-comb]
+def createAndListenTo(tagName: ElementName, eventName: EventName, listener: (eventName.Event, tagName.Element) => Unit) = {
+  val element = document.createElement(tagName)
+  element.addEventListener(eventName)(event => listener(event, element))
+}
+\end{lstlisting}
+\end{figure}
+
+\begin{figure}
+\begin{lstlisting}[label=lst-dt-react]
+def observe(target: EventTarget, name: EventName) =
+  (listener: (name.Event => Unit)) => target.addEventListener(name)(listener)
+\end{lstlisting}
+\end{figure}
+
+With this encoding, the functions using event names or element names are not anymore cluttered with type parameters.
+
+\section{Validation}
+\label{sec-validation}
+
+\subsection{Implementation in js-scala}
+
+We implemented our pattern in js-scala~\cite{Kossakowski12_JsDESL}, a Scala library providing JavaScript code generators\footnote{Source code is available at \href{http://github.com/js-scala}{http://github.com/js-scala}}.
+
 \subsection{Limitations}
@@
 
+We presented two ways to encode dynamically typed browser functions in mainstream statically typed languages like Java and Scala, using \emph{generics} or path-dependent types.
+Our encoding gives more type safety than existing solutions, while keeping the same expression power and modularity level as the native API.
+
 \bibliographystyle{plain}

### 2014-01-10 papers/icwe14/icwe14.tex
blabla

@@
 
-However we observe that, despite these languages are statically typed, their integration of the browser API either is not type safe or gives less expression power to developers.
+However we observe that, despite these languages are statically typed, their integration of the browser API either is not type safe or gives less expression power to the developers.
 Indeed, integrating an API designed for a dynamically typed language into a statically typed language can be challenging.
-For instance, the \jscode{createElement} function return type depends on the value of the \jscode{String} parameter it is passed.
+For instance, the \jscode{createElement(name)} function return type depends on the value of its \jscode{name} parameter: \jscode{createElement('div')} returns a \jscode{DivElement}, \jscode{createElement('input')} returns an \jscode{InputElement}, \emph{etc}.
+Existing languages expose this function by making it return an \jscode{Element} (the least upper bound of the types of all the possible returned values), though a more precise type could be known at compile-time.
+Another way to expose this function consists in exposing several functions, each one fixing the value of the \jscode{name} parameter along with its return type.
+The first solution is not type safe and the second solution does not scale.
 

### 2014-01-13 papers/icwe14/icwe14.tex
up

--- 2014-01-02 papers/icwe14/icwe14.tex
@@
 \begin{abstract}
-Encoding dynamically typed APIs in statically typed languages can be challenging. We present new ways to encode them in order to preserve type safety and to keep the same expression power.
+Encoding dynamically typed APIs in statically typed languages can be challenging.
+Incidentally, existing statically typed languages compiling to JavaScript expose its native API in ways that either are not type safe or give less expression power to developers.
+We present two new ways to encode the challenging parts of the Web browser API such that type safety and expression power are preserved.
+Our first encoding relies on type parameters and can be implemented in most mainstream languages but drags phantom types up to the use sites.
+The second encoding does not suffer from this inconvenience but requires dependent types.
 \end{abstract}
@@
 
-However we observe that, despite these languages are statically typed, their integration of the browser API either is not type safe or gives less expression power to developers.
+However we observe that, despite these languages are statically typed, their integration of the browser API either is not type safe or gives less expression power.
 Indeed, integrating an API designed for a dynamically typed language into a statically typed language can be challenging.
-For instance, the \jscode{createElement} function return type depends on the value of the \jscode{String} parameter it is passed.
+For instance, the \jscode{createElement} function return type depends on the value of its parameter: \jscode{createElement('div')} returns a \jscode{DivElement}, \jscode{createElement('input')} returns an \jscode{InputElement}, \emph{etc}.
 
-This paper reviews some common functions of the browser API, identifies those which are not well encoded in statically typed programming languages and shows new ways to encode them with more type safety while keeping the native API expression power.
+Existing languages expose this function by making it return an \jscode{Element}, the least upper bound of the types of all the possible returned values, thus loosing type information and requiring users to explicitly downcast the returned value to its expected more precise type.
+Another way to expose this function consists in exposing several functions, each one fixing the value of the parameter along with its return type: \jscode{createDivElement}, \jscode{createInputElement}, \emph{etc} would be parameterless functions returning a \jscode{DivElement} and an \jscode{InputElement}, respectively.
+This encoding forces to hard-code the name of the element to create: it can not anymore be a parameter.
+The first solution is not type safe and the second solution reduces the expression power.
+
+This paper reviews some common functions of the browser API, identifies the patterns that are difficult to encode in static type systems and shows new ways to encode them such that type safety and expression power are preserved.
 We find that type parameters are sufficient to achieve this goal and that \emph{path-dependent types} provide an even more convenient encoding for the end developers.
@@
 Listing~\ref{lst-create-element-overload} illustrates such an encoding in the Java language.
+Note that the listing gives only the type signature of the functions visible to the end developers.
+The implementation of the functions is not relevant for our purpose.
 
@@
 
-Table~\ref{table-existing-encodings} summarizes, for each statically typed language, which approach is used to encode each function of the browser API.
+Table~\ref{table-existing-encodings} summarizes, for each studied statically typed language, which approach is used to encode each function of the browser API.
 
@@
 We distinguished three approaches to integrate the challenging parts of the browser API into statically typed languages.
-This section compares these approaches in terms of type safety, expression power and convenience to use from the end developer point of view.
+This section compares these approaches in terms of type safety and expression power.
 
@@
 
-The second approach, consisting in defining as many functions as there are possible return types of a function, is completely type safe but can lead to combinatorial explosion in some situations.
-Consider for instance listing~\ref{lst-js-comb} defining a JavaScript function that both creates an element and registers an event listener when a given event occurs on it.
-Note that the event listener is passed both the event and the element.
-Encoding such a function in a statically typed language using this approach would require to create one function for each combination of tag name and event name.
+The second approach, consisting in defining as many functions as there are possible return types of the encoded function, is type safe but leads to a less general API: each function fixes a parameter value of the encoded function, hence being less general.
+The limits of this approach are better illustrated when one tries to combine several functions.
+Consider for instance listing~\ref{lst-js-comb} defining a JavaScript function that both find elements and registers an event listener when a given event occurs on them.
+Note that the event listener is passed both the event and the element: its type depends on both the tag name and the event name.
+This function could not be implemented if the general \jscode{getElementsByTagName} and \jscode{addEventListener} functions were not available.
+The best that could be done would be to create one function for each combination of tag name and event name.
 
 \begin{figure}
-\begin{lstlisting}[label=lst-js-comb,language=JavaScript]
-var createAndListenTo = function (tagName, eventName, listener) {
-  var element = document.createElement(tagName);
+\begin{lstlisting}[label=lst-js-comb,language=JavaScript,caption={Combination of use of \jscode{getElementsByTagName} and \jscode{addEventListener}}]
+var findAndListenTo = function (tagName, eventName, listener) {
+  document.getElementsByTagName(tagName).forEach(function (element) {
     element.addEventListener(eventName, function (event) {
       listener(event, element);
+    });
   });
@@
 \begin{figure}
-\begin{lstlisting}[label=lst-js-react,language=JavaScript]
+\begin{lstlisting}[label=lst-js-react,language=JavaScript,caption={Partial application of \jscode{addEventListener} parameters}]
 var observe = function (target, name) {
@@
 
-In summary, browser API integration by existing statically typed languages compiling to JavaScript is either not type safe or not as modular or expressive as the underlying JavaScript API.
+[FIXME En fait les deux dernieres approches sont deux cas particuliers d'une meme approche, plus generale, qui consiste a fixer la valeur d'un parametre. Je regroupe ou pas ?]
+
+In summary, browser API integration by existing statically typed languages compiling to JavaScript is either not type safe or not as expressive as the underlying JavaScript API.
 
@@
 
-In this section we show how we can encode the functions presented previously, in a type safe and convenient way for developers while keeping the same expression power.
+In this section we show how we can encode the functions presented previously, in a type safe way while keeping the same expression power.
 
@@
 
-Listing~\ref{lst-generics-dom} shows this approach applied to the \jscode{createElement} function: we created a type \scalacode{ElementName<E>}, the \scalacode{name} parameter is not a \scalacode{String} but a \scalacode{ElementName<E>}, and the function returns an \scalacode{E} instead of an \scalacode{Element}.
+Listing~\ref{lst-generics-dom} shows this approach applied to the \jscode{createElement} function which return type depends on its \jscode{name} parameter value: a type \scalacode{ElementName<E>} has been created, the type of the \scalacode{name} parameter has been set to \scalacode{ElementName<E>} instead of \scalacode{String}, and the return type of the function is \scalacode{E} instead of \scalacode{Element}.
 The \scalacode{ElementName<E>} type encodes the relationship between the name of an element and the type of this element\footnote{The type parameter \scalacode{E} is also called a \emph{phantom type}~\cite{leijen1999domain} because \scalacode{ElementName} values never hold a \scalacode{E} value}.
 For instance, we created a value \scalacode{Input} of type \scalacode{ElementName<InputElement>}.
-The last two lines shows how this API is used by end developers: by passing the \scalacode{Input} value as parameter to the function, it fixes the type parameter \scalacode{E} to \scalacode{InputElement} so the returned value has the most possible precise type.
+The last two lines shows how this API is used by end developers: passing the \scalacode{Input} value as parameter to the function fixes the type parameter \scalacode{E} to \scalacode{InputElement} so the returned value has the most possible precise type.
 
 \begin{figure}
-\begin{lstlisting}[label=lst-generics-dom,language=java]
+\begin{lstlisting}[label=lst-generics-dom,language=java,caption={Encoding of the \jscode{createElement} function using type parameters}]
 // --- API Definition
@@
 \begin{figure}
-\begin{lstlisting}[label=lst-generics-events,language=java]
+\begin{lstlisting}[label=lst-generics-events,language=java,caption={Encoding of the \jscode{addEventListener} function using type parameters}]
 // --- API Definition
@@
 \begin{figure}
-\begin{lstlisting}[label=lst-generics-comb,language=java]
-<A, B> void createAndListenTo(
+\begin{lstlisting}[label=lst-generics-comb,language=java,caption={Combination of \scalacode{getElementsByTagName} and \scalacode{addEventListener} functions encoded using type parameters}]
+<A, B> void findAndListenTo(
     ElementName<A> tagName,
@@
     Function2<A, B, Void> listener) {
-  A element = document.createElement(tagName);
+  for (A element : document.getElementsByTagName(tagName)) {
     element.addEventListener(eventName, event -> listener.apply(event, element));
+  }
 }
@@
 \begin{figure}
-\begin{lstlisting}[label=lst-generics-react,language=java]
+\begin{lstlisting}[label=lst-generics-react,language=java,caption={Partial application of \scalacode{addEventListener} encoded with type parameters}]
 <A> Function<Function<A, Void>, Void> observe(EventTarget target, EventName<A> name) {
@@
 
-Our encoding makes it possible to implement exactly the same functions as we are able to implement in plain JavaScript, but with type safety.
+Our encoding is type safe and gives as much expression power as the native API since it makes it possible to implement exactly the same functions as we are able to implement in plain JavaScript.
 
-However, every function taking an element name or an event name as parameter must have type parameters too, making its type signature harder to read.
+However, every function taking an element name or an event name as parameter has its type signature cluttered with phantom types (extra type parameters).
 
@@
 \begin{figure}
-\begin{lstlisting}[label=lst-dt-dom]
+\begin{lstlisting}[label=lst-dt-dom,caption={Encoding of \jscode{createElement} using path-dependent types}]
 trait ElementName {
@@
 \begin{figure}
-\begin{lstlisting}[label=lst-dt-events]
+\begin{lstlisting}[label=lst-dt-events,caption={Encoding of \jscode{addEventListener} using path-dependent types}]
 trait EventName {
@@
 \begin{figure}
-\begin{lstlisting}[label=lst-dt-comb]
-def createAndListenTo(tagName: ElementName, eventName: EventName, listener: (eventName.Event, tagName.Element) => Unit) = {
-  val element = document.createElement(tagName)
+\begin{lstlisting}[label=lst-dt-comb,caption={Combination of \scalacode{getElementsByTagName} and \scalacode{addEventListener} using path-dependent types}]
+def findAndListenTo(tagName: ElementName, eventName: EventName, listener: (eventName.Event, tagName.Element) => Unit) = {
+  for (element <- document.getElementsByTagName(tagName) {
     element.addEventListener(eventName)(event => listener(event, element))
+  }
 }
@@
 \begin{figure}
-\begin{lstlisting}[label=lst-dt-react]
+\begin{lstlisting}[label=lst-dt-react,caption={Partial application of \scalacode{addEventListener} using path-dependent types}]
 def observe(target: EventTarget, name: EventName) =
@@
 
-With this encoding, the functions using event names or element names are not anymore cluttered with type parameters.
+With this encoding, the functions using event names or element names are not anymore cluttered with phantom types, and type safety is still preserved.
 
@@
 
-We presented two ways to encode dynamically typed browser functions in mainstream statically typed languages like Java and Scala, using \emph{generics} or path-dependent types.
-Our encoding gives more type safety than existing solutions, while keeping the same expression power and modularity level as the native API.
+We presented two ways to encode dynamically typed browser functions in mainstream statically typed languages like Java and Scala, using type parameters or path-dependent types.
+Our encoding gives more type safety than existing solutions, while keeping the same expression power as the native API.
+
+We argue that, if industry shippers want to write their Web applications in statically typed languages, dependent types are going to be the most desired feature of these languages.
 

### 2014-01-31 papers/icwe14/icwe14.tex
i++

@@
 
+% listing numbering
+\renewcommand{\thelstlisting}{\arabic{lstlisting}}
+
  \title{Using Path-Dependent Types to Build Type-Safe JavaScript Foreign Function Interfaces}
@@
 Incidentally, existing statically typed languages compiling to JavaScript expose its native API in ways that either are not type safe or give less expression power to developers.
-We present two new ways to encode the challenging parts of the Web browser API such that type safety and expression power are preserved.
-Our first encoding relies on type parameters and can be implemented in most mainstream languages but drags phantom types up to the use sites.
+We present two new ways to encode the challenging parts of the Web browser API such that both type safety and expression power are preserved.
+Our first encoding relies on type parameters and can be implemented in most mainstream languages but drags phantom types up to the usage sites.
 The second encoding does not suffer from this inconvenience but requires dependent types.
@@
 
-We recently observed the emergence of several statically typed programming languages compiling to JavaScript (\emph{e.g.} Java/GWT~\cite{Kereki09_GWT}, Dart~\cite{Griffith11_Dart}, TypeScript~\cite{fenton2012typescript}, Kotlin\footnote{\href{http://kotlin.jetbrains.org}{http://kotlin.jetbrains.org}}, Opa\footnote{\href{http://opalang.org}{http://opalang.org}}, SharpKit\footnote{\href{http://sharpkit.net}{http://sharpkit.net}}, Haxe~\cite{Cannasse08_HaXe}, Idris~\cite{Brady13_Idris}, Elm~\cite{czaplicki2012elm}).
+We recently observed the emergence of several statically typed programming languages compiling to JavaScript (\emph{e.g.} Java/GWT~\cite{Kereki09_GWT}, Dart~\cite{Griffith11_Dart}, TypeScript~\cite{fenton2012typescript}, Kotlin\footnote{\href{http://kotlin.jetbrains.org}{http://kotlin.jetbrains.org}}, Opa\footnote{\href{http://opalang.org}{http://opalang.org}}, SharpKit\footnote{\href{http://sharpkit.net}{http://sharpkit.net}}, Haxe~\cite{Cannasse08_HaXe}, Scala~\cite{Doeraene13_ScalaJs}, Idris~\cite{Brady13_Idris}, Elm~\cite{czaplicki2012elm}).
 Developers can use them to write the client-side part of their Web applications, relying on a \emph{foreign function interface} mechanism to call the Web browser native API.
@@
 
-Existing languages expose this function by making it return an \jscode{Element}, the least upper bound of the types of all the possible returned values, thus loosing type information and requiring users to explicitly downcast the returned value to its expected more precise type.
-Another way to expose this function consists in exposing several functions, each one fixing the value of the parameter along with its return type: \jscode{createDivElement}, \jscode{createInputElement}, \emph{etc} would be parameterless functions returning a \jscode{DivElement} and an \jscode{InputElement}, respectively.
+Most of the aforementionned languages expose this function by making it return an \jscode{Element}, the least upper bound of the types of all the possible returned values, thus loosing type information and requiring users to explicitly downcast the returned value to its expected more precise type.
+Another way to expose this function consists in exposing several functions, each one fixing the value of the parameter along with its return type: \jscode{createDivElement}, \jscode{createInputElement}, \emph{etc.} are parameterless functions returning a \jscode{DivElement} and an \jscode{InputElement}, respectively.
 This encoding forces to hard-code the name of the element to create: it can not anymore be a parameter.
-The first solution is not type safe and the second solution reduces the expression power.
+In summary, the first solution is not type safe and the second solution reduces the expression power.
 
-This paper reviews some common functions of the browser API, identifies the patterns that are difficult to encode in static type systems and shows new ways to encode them such that type safety and expression power are preserved.
+This paper reviews some common functions of the browser API, identifies the patterns that are difficult to encode in static type systems and shows new ways to encode them such that both type safety and expression power are preserved.
 We find that type parameters are sufficient to achieve this goal and that \emph{path-dependent types} provide an even more convenient encoding for the end developers.
@@
 
-This section reviews the most commonly used browser functions and presents the way they are integrated in GWT, Dart, TypeScript, Kotlin, Opa, SharpKit, Haxe, Idris and Elm.
+This section reviews the most commonly used browser functions and presents the different integration strategies currently used by statically typed programming languages like GWT, Dart, TypeScript, Kotlin, Opa, SharpKit, Haxe, Scala, Idris or Elm.
 
@@
 The client-side part of the code of a Web application essentially reacts to user events (\emph{e.g.} mouse clicks), triggers actions and updates the document (DOM) according to their effect.
-Table~\ref{table-dom-api} lists the main functions supported by the Web browsers according to the Mozilla Developer Network\footnote{\href{https://developer.mozilla.org/en-US/docs/DOM/DOM\_Reference/Introduction}{https://developer.mozilla.org/en-US/docs/DOM/DOM\_Reference/Introduction}} (we removed the functions that can trivially be encoded in a static type system).
+Table~\ref{table-dom-api} lists the main functions supported by Web browsers according to the Mozilla Developer Network\footnote{\href{https://developer.mozilla.org/en-US/docs/DOM/DOM\_Reference/Introduction}{https://developer.mozilla.org/en-US/docs/DOM/DOM\_Reference/Introduction}} (we removed the functions that can trivially be encoded in a static type system).
 
@@
  \label{table-dom-api}
- \caption{Web browsers main functions}
+ \caption{Web browsers main functions that are challenging to encode in a static type system}
 \end{table}
 
-The main challenge comes from the fact that these functions are polymorphic in their return type or in their parameters.
+To illustrate the problems caused by these functions, we present a simple JavaScript program using them and show how it can be implemented in statically typed programming languages according to the different strategies used to encode these functions.
+Listing \ref{lst-problem} shows the initial JavaScript code of the program.
+It defines a function \jscode{slideshow} that creates a slide show from an array of image URLs.
+The function returns an image element displaying the first image of the slide show, and each time a user clicks on it the next image is displayed.
 
-For instance, functions \jscode{getElementsByTagName(name)}, \jscode{getElementById(id)} and \jscode{createElement(name)} can return values of type \jscode{DivElement} or \jscode{InputElement} or any other subtype of \jscode{Element} (their least upper bound).
-The interface of \jscode{Element} is more general and provides less features than its subtypes, so it is important for the encoding of these functions in a statically typed language to be as precise as possible.
+\begin{figure}
+\begin{lstlisting}[label=lst-problem,language=JavaScript,caption=JavaScript function creating a slide show from an array of image URLs]
+function slideshow(sources) {
+  var img = document.createElement('img');
+  var current = 0;
+  img.src = sources[current];
+  img.addEventListener('click', function (event) {
+    if (event.button == 0) {
+      current = (current + 1) % (sources.length - 1);
+      img.src = sources[current];
+    }
+  });
+  return img
+}
+\end{lstlisting}
+\end{figure}
+
+The most common way to encode the DOM API in statically typed languages is to follow the standard interface specifications of HTML~\cite{Raggett99_HTML} and DOM~\cite{w3c2004document}.
+
+The main challenge comes from the fact that these function return types or parameter types are often too general.
+Indeed, functions \jscode{getElementsByTagName(name)}, \jscode{getElementById(id)} and \jscode{createElement(name)} can return values of type \jscode{DivElement} or \jscode{InputElement} or any other subtype of \jscode{Element} (their least upper bound).
+The interface of \jscode{Element} is more general and provides less features than its subtypes.
+For instance, the \jscode{ImageElement} type (representing images) has a \jscode{src} property that does not exist at the \jscode{Element} level.
+Similarly, the \jscode{MouseEvent} type has a \jscode{button} property that does not exist at the (more general) \jscode{Event} level, used by the function \jscode{addEventListener}.
+
+\begin{figure}
+\begin{lstlisting}[label=lst-problem-cast,language=Scala,caption={Scala implementation of \texttt{slideshow} using the standard HTML and DOM API}]
+def slideshow(sources: Array[String]): ImageElement = {
+  val img =
+    document.createElement("img").asInstanceOf[ImageElement]
+  var current = 0
+  img.src = sources(current)
+  img.addEventListener("click", event => {
+    if (event.asInstanceOf[MouseEvent].button == 0) {
+      current = (current + 1) % (sources.size - 1)
+      img.src = sources(current)
+    }
+  })
+  img
+}
+\end{lstlisting}
+\end{figure}
+
+Listing \ref{lst-problem-cast} shows a Scala implementation of the \jscode{slideshow} program using an API following the standard specifications of HTML and DOM.
+The listing contains two type casts, making it possible to use the \scalacode{src} property on the \scalacode{img} value and the \scalacode{button} property on the \scalacode{event} value, respectively.
+
+These type casts make the code more fragile and less convenient to read and write. That's why some statically typed languages attempt to provide an API preserving as precise types as possible.
 
@@
 Hence, most implementations define a return type of \jscode{Element}.
-The drawback of this approach is that developers may get a value with a type weaker than they need and may require to explicitly downcast it, thus loosing type safety.
 
@@
 This characteristic makes it possible to encode each of these functions by defining as many parameterless functions as there are possible tag names, where each function fixes the initial \jscode{name} parameter to be one of the possible values and exposes the corresponding specialized return type.
-Listing~\ref{lst-create-element-overload} illustrates such an encoding in the Java language.
-Note that the listing gives only the type signature of the functions visible to the end developers.
-The implementation of the functions is not relevant for our purpose.
-
-\begin{figure}
-\begin{lstlisting}[label=lst-create-element-overload,language=Java,caption={Encoding of the \jscode{createElement(name)} function in Java using parameterless functions fixing the initial \jscode{name} parameter value}]
-DivElement createDivElement();
-InputElement createInputElement();
-FormElement createFormElement();
-...
-\end{lstlisting}
-\end{figure}
 
@@
 For instance, a listener of \jscode{'click'} events is a function taking a \jscode{MouseEvent} parameter, a listener of \jscode{'keydown'} events is a function taking a \jscode{KeyboardEvent} parameter, and so on.
-The same pattern as above (defining a set of functions fixing the \jscode{name} parameter value) can be used to encode this function in a statically typed language, but an alternative encoding could be to define one function taking one parameter carrying both the information of the event name and the
-event listener. Listing~\ref{lst-add-event-listener-one-param} shows such an encoding in Java.
+The same pattern as above (defining a set of functions fixing the \jscode{name} parameter value) can be used to encode this function in statically typed languages.
 
 \begin{figure}
-\begin{lstlisting}[label=lst-add-event-listener-one-param,language=Java,caption={Encoding of the \jscode{target.addEventListener(name, listener)} function in Java using one paremeter carrying both the information of the event name and the event listener}]
-// --- API Definition
-interface EventTarget {
-  void addEventListener(EventListener listener);
+\begin{lstlisting}[label=lst-problem-comb,language=Scala,caption={Scala implementation of \texttt{slideshow} using specialized functions}]
+def slideshow(sources: Array[String]): ImageElement = {
+  val img = document.createImageElement()
+  var current = 0
+  img.src = sources(current)
+  img.addClickEventListener { event =>
+    if (event.button == 0) {
+      current = (current + 1) % (sources.size - 1)
+      img.src = sources(current)
     }
-
-interface EventListener {}
-
-interface ClickEventListener extends EventListener {
-  void onClick(ClickEvent event);
   }
-
-interface KeyDownEventListener extends EventListener {
-  void onKeyDown(KeyboardEvent event);
-}
-
-// --- Usage
-void logClicks(EventTarget target) {
-  target.addEventListener(new ClickEventListener {
-    public void onClick(ClickEvent event) {
-      console.log(event.button);
-    }
-  });
+  img
 }
@@
 
-Table~\ref{table-existing-encodings} summarizes, for each studied statically typed language, which approach is used to encode each function of the browser API.
+Listing~\ref{lst-problem-comb} shows what would our \scalacode{slideshow} implementation look like using such an encoding.
+The \scalacode{createImageElement} function takes no parameter and returns a value of type \scalacode{ImageElement}, and the \scalacode{addClickEventListener} function takes as parameter a function that takes a \scalacode{MouseEvent} value as parameter, ruling the need of type casts out.
 
-\begin{table}
-\centering
-\begin{tabular}{|l|c|c|c|c|}
-\hline  & \jscode{getElementById} & \jscode{getElementsByTagName} & \jscode{createElement} & \jscode{addEventListener} \\
-\hline Java & lub & lub & comb & plop \\
-\hline Dart & lub & lub \\
-\hline TypeScript \\
-\hline Haxe & lub & lub \\
-\hline Opa \\
-\hline Kotlin & lub & lub \\
-\hline SharpKit & lub & lub \\
-\hline Idris \\
-\hline Elm \\
-\hline
-\end{tabular}
-\label{table-existing-encodings}
-\caption{Summary of the encodings used by existing statically typed languages}
-\end{table}
+In the case of \jscode{addEventListener} we also encountered a slight variation of the encoding, consisting of defining one general function taking one parameter carrying both the 
+information of the event name and the event listener.
+
+\begin{figure}
+\begin{lstlisting}[label=lst-problem-curry,language=Scala,caption={Implementation of \texttt{slideshow} using a general \texttt{addEventListener} function taking one parameter containing both the event name and the even listener}]
+img.addEventListener(ClickEventListener { event =>
+  // ...
+})
+\end{lstlisting}
+\end{figure}
+
+Listing~\ref{lst-problem-curry} shows the relevant changes in our program if we use this encoding.
+The \scalacode{addEventListener} function takes one parameter, a \scalacode{ClickEventListener}, carrying both the name of the event and the event listener code.
+
+[TODO Donner quand meme des chiffres un peu detailles sur quel langage utilise quelle strategie]
+
+% Table~\ref{table-existing-encodings} summarizes, for each studied statically typed language, which approach is used to encode each function of the browser API.
+% 
+% \begin{table}
+% \centering
+% \begin{tabular}{|l|c|c|c|c|}
+% \hline  & \jscode{getElementById} & \jscode{getElementsByTagName} & \jscode{createElement} & \jscode{addEventListener} \\
+% \hline Java & lub & lub & comb & plop \\
+% \hline Dart & lub & lub \\
+% \hline TypeScript \\
+% \hline Haxe & lub & lub \\
+% \hline Opa \\
+% \hline Kotlin & lub & lub \\
+% \hline SharpKit & lub & lub \\
+% \hline Idris \\
+% \hline Elm \\
+% \hline
+% \end{tabular}
+% \label{table-existing-encodings}
+% \caption{Summary of the encodings used by existing statically typed languages}
+% \end{table}
 
 \subsection{Limitations of existing encoding approaches}
+\label{sec-limitations}
 
@@
 
-The first approach, consisting in using the least upper bound of all the possible types is not type safe because it sometimes requires developers to explicitly downcast values to their expected specialized type.
+The first approach, consisting in using the least upper bound of all the possible types has the same expression power as the native browser API, but is not type safe because it sometimes requires developers to explicitly downcast values to their expected specialized type.
 
@@
 The limits of this approach are better illustrated when one tries to combine several functions.
-Consider for instance listing~\ref{lst-js-comb} defining a JavaScript function that both find elements and registers an event listener when a given event occurs on them.
+Consider for instance listing~\ref{lst-js-comb} defining a JavaScript function \jscode{findAndListenTo} that both find elements and registers an event listener when a given event occurs on them.
 Note that the event listener is passed both the event and the element: its type depends on both the tag name and the event name.
-This function could not be implemented if the general \jscode{getElementsByTagName} and \jscode{addEventListener} functions were not available.
-The best that could be done would be to create one function for each combination of tag name and event name.
+This function can not be implemented if the general functions \jscode{getElementsByTagName} and \jscode{addEventListener} are not available.
+The best that could be done would be to create one function for each combination of tag name and event name, leading to an explosion of the number of functions to implement.
+Thus, this approach gives less expression power than the native browser API.
 
@@
 \begin{lstlisting}[label=lst-js-comb,language=JavaScript,caption={Combination of use of \jscode{getElementsByTagName} and \jscode{addEventListener}}]
-var findAndListenTo = function (tagName, eventName, listener) {
-  document.getElementsByTagName(tagName).forEach(function (element) {
+function findAndListenTo(tagName, eventName, listener) {
+  var elements = document.getElementsByTagName(tagName);
+  elements.forEach(function (element) {
     element.addEventListener(eventName, function (event) {
@@
 The third approach, consisting in combining two parameters into one parameter carrying all the required information, is type safe too, but reduces the expression power because it forbids developers to partially apply the function by supplying only one parameter.
-Consider for instance listing~\ref{lst-js-react} that defines a function partially applying the \jscode{addEventListener} function\footnote{This pattern is often used by functional reactive programming libraries like Rx.js~\cite{liberty2011reactive}}.
-Such a function is impossible to encode using this approach, in a statically typed language.
+Consider for instance listing~\ref{lst-js-react} that defines a function \jscode{observe} partially applying the \jscode{addEventListener} function\footnote{This pattern is often used by functional reactive programming libraries like Rx.js~\cite{liberty2011reactive}}.
+Such a function can not be implemented with this approach.
+Thus, this one gives less expression power than the native browser API.
 
@@
 \begin{lstlisting}[label=lst-js-react,language=JavaScript,caption={Partial application of \jscode{addEventListener} parameters}]
-var observe = function (target, name) {
+function observe(target, name) {
   return function (listener) {
@@
 
-[FIXME En fait les deux dernieres approches sont deux cas particuliers d'une meme approche, plus generale, qui consiste a fixer la valeur d'un parametre. Je regroupe ou pas ?]
-
-In summary, browser API integration by existing statically typed languages compiling to JavaScript is either not type safe or not as expressive as the underlying JavaScript API.
+In summary, the current integration of the browser API by statically typed languages compiling to JavaScript is either not type safe or not as expressive as the underlying JavaScript API.
 
@@
 
-In this section we show how we can encode the functions presented previously, in a type safe way while keeping the same expression power.
+In this section we show how we can encode the challenging main functions of the DOM API in a type safe way while keeping the same expression power.
 
@@
 
-In all the cases where a type \scalacode{T} involved in a function depends on the value of a parameter \scalacode{p} of this function (all the aforementionned functions are in this case, excepted \jscode{getElementById}), we can encode this relationship in the type system using type parameters as follows:
+In all the cases where a type \scalacode{T} involved in a function depends on the value of a parameter \scalacode{p} of this function (all the aforementionned functions of the DOM API are in this case), we can encode this relationship in the type system using type parameters as follows:
 
 \begin{enumerate}
- \item Create a parametrized class \scalacode{P<U>}
+ \item Define a parametrized class \scalacode{P<U>}
  \item Set the type of \scalacode{p} to \scalacode{P<U>}
  \item Use type \scalacode{U} instead of type \scalacode{T}
+ \item Define as many values of type \scalacode{P<U>} as there are possible values for \scalacode{p}, each one fixing its \scalacode{U} type parameter to the corresponding more precise type
 \end{enumerate}
@@
 For instance, we created a value \scalacode{Input} of type \scalacode{ElementName<InputElement>}.
-The last two lines shows how this API is used by end developers: passing the \scalacode{Input} value as parameter to the function fixes the type parameter \scalacode{E} to \scalacode{InputElement} so the returned value has the most possible precise type.
 
@@
 \begin{lstlisting}[label=lst-generics-dom,language=java,caption={Encoding of the \jscode{createElement} function using type parameters}]
-// --- API Definition
 class ElementName<E> {}
@@
 
-final ElementName<InputElement> Input = new ElementName<InputElement>();
-final ElementName<ImageElement> Img = new ElementName<ImageElement>();
-
-// --- Usage
-InputElement input = createElement(Input);
-ImageElement img = createElement(Img);
+final ElementName<InputElement> Input = new ElementName<>();
+final ElementName<ImageElement> Img = new ElementName<>();
+// etc. for each possible element name
 \end{lstlisting}
@@
 
-\scalacode{getElementsByTagName} can be encoded in very similar way, and listing~\ref{lst-generics-events} shows the encoding of the \jscode{addEventListener} function.
+\scalacode{getElementsByTagName} can be straightforwardly encoded in a similar way, and listing~\ref{lst-generics-events} shows the encoding of the \jscode{addEventListener} function.
+Our listing assumes the definition of a \scalacode{Function<A, B>} type representing a function from \scalacode{A} to \scalacode{B}.
 
@@
 \begin{lstlisting}[label=lst-generics-events,language=java,caption={Encoding of the \jscode{addEventListener} function using type parameters}]
-// --- API Definition
 class EventName<E> {}
@@
 interface EventTarget {
-  <E> void addEventListener(EventName<E> name, Function<E, Void> callback);
+  <E> void addEventListener(
+          EventName<E> name, Function<E, Void> callback);
 }
 
-final EventName<MouseEvent> Click = new EventName<MouseEvent>();
-
-// --- Usage
-ButtonElement btn = createElement(Button);
-btn.addEventListener(Click, e -> console.log(e.button))
+final EventName<MouseEvent> Click = new EventName<>();
+final EventName<KeyboardEvent> KeyUp = new EventName<>();
+// etc. for each possible event name
 \end{lstlisting}
@@
 
-We show that our encodings support the challenging functions defined in listings \ref{lst-js-comb} and \ref{lst-js-react}.
+\begin{figure}
+\begin{lstlisting}[label=lst-slideshow-generics,language=Scala,caption={Scala implementation of the \texttt{slideshow} function using generics}]
+def slideshow(sources: Array[String]) {
+  val img = document.createElement(Img)
+  var current = 0
+  img.src = sources(current)
+  img.addEventListener(Click, event => {
+    if (event.button == 0) {
+      current = (current + 1) % (sources.length - 1)
+      img.src = sources(current)
+    }
+  })
+  img
+}
+\end{lstlisting}
+\end{figure}
 
-Listing \ref{lst-generics-comb} and \ref{lst-generics-react} show how these functions can be implemented. They are basically a direct translation from JavaScript to Java.
+Listing \ref{lst-slideshow-generics} illustrate the usage of such an encoding by implementing our \texttt{slideshow} program in Scala.
+Passing the \scalacode{Img} value as a parameter to the \scalacode{createElement} function fixes its \scalacode{E} type parameter to \scalacode{ImageElement} so the returned value has the most possible precise type and the \scalacode{src} property can be used on it.
+Similarly, passing the \scalacode{Click} value to the \scalacode{addEventListener} function fixes its \scalacode{E} type parameter to \scalacode{MouseEvent}, so the event listener has the most possible precise type and the \scalacode{button} property can be used on the \scalacode{event} parameter.
 
@@
   for (A element : document.getElementsByTagName(tagName)) {
-    element.addEventListener(eventName, event -> listener.apply(event, element));
+    element.addEventListener(
+         eventName,
+         event -> listener.apply(event, element));
   }
@@
 \begin{lstlisting}[label=lst-generics-react,language=java,caption={Partial application of \scalacode{addEventListener} encoded with type parameters}]
-<A> Function<Function<A, Void>, Void> observe(EventTarget target, EventName<A> name) {
+<A> Function<Function<A, Void>, Void> observe(
+      EventTarget target, EventName<A> name) {
   return listener -> {
@@
 
+Finally, listing \ref{lst-generics-comb} and \ref{lst-generics-react} show how the challenging functions of section \ref{sec-limitations}, \jscode{findAndListenTo} and \jscode{observe}, can be implemented with our encoding.
+They are basically a direct translation from JavaScript to Java.
+
 Our encoding is type safe and gives as much expression power as the native API since it makes it possible to implement exactly the same functions as we are able to implement in plain JavaScript.
 
-However, every function taking an element name or an event name as parameter has its type signature cluttered with phantom types (extra type parameters).
+However, every function taking an element name or an event name as parameter has its type signature cluttered with phantom types (extra type parameters): the \scalacode{observe} function takes a phantom type parameter \scalacode{A} and the \scalacode{findAndListenTo} function takes two phantom type parameters, \scalacode{A} and \scalacode{B}.
 
@@
 This section shows how we can remove the extra type parameters needed in the previous section by using \emph{path-dependent types}.
-Essentially, the idea is that, instead of using a type parameter, we use a type member.
-Listings \ref{lst-dt-dom} and \ref{lst-dt-events} show this encoding in Scala.
-The return type of the \scalacode{createElement} function is \scalacode{name.Element}: it refers to the \scalacode{Element} type member of its \scalacode{name} parameter.
+Essentially, the idea is that, instead of using a type parameter, we use a \emph{type member}.
+Listings \ref{lst-dt-dom} and \ref{lst-dt-events} show this encoding in Scala for \jscode{createElement} and \jscode{addEventListener}, respectively.
 
@@
 }
-object Div extends ElementName { type Element = DivElement }
-object Input extends ElementName { type Element = InputElement }
 
 def createElement(name: ElementName): name.Element
+
+object Div extends ElementName {
+  type Element = DivElement
+}
+object Input extends ElementName {
+  type Element = InputElement
+}
+// etc. for each possible element name
 \end{lstlisting}
 \end{figure}
+
+Now, the \scalacode{ElementName} type has no type parameter but a type member \scalacode{Element}.
+The return type of the \scalacode{createElement} function is \scalacode{name.Element}: it refers to the \scalacode{Element} type member of its \scalacode{name} parameter.
+The \scalacode{Div} and \scalacode{Input} values illustrate how their corresponding element type is fixed.
 
@@
 }
-object Click extends EventDef { type Event = MouseEvent }
+
+object Click extends EventName { type Event = MouseEvent }
 
 trait EventTarget {
-  def addEventListener(name: EventName)(handler: name.Event => Unit): Unit
+  def addEventListener(name: EventName)
+                      (handler: name.Event => Unit): Unit
 }
@@
 
-Implementing listings \ref{lst-js-comb} and \ref{lst-js-react} using this encoding is straightforward, as shown by listings \ref{lst-dt-comb} and \ref{lst-dt-react}, respectively.
+[FIXME un peu plus de details sur le fonctionnement des path-dependent types]
+
+The implementation of the \scalacode{slideshow} function with this encoding is exactly the same as with the previous approach using generics.
+However, functions \jscode{findAndListenTo} and \jscode{observe} can be implemented more straightforwardly, as shown by listings \ref{lst-dt-comb} and \ref{lst-dt-react}, respectively.
 
@@
 \begin{lstlisting}[label=lst-dt-comb,caption={Combination of \scalacode{getElementsByTagName} and \scalacode{addEventListener} using path-dependent types}]
-def findAndListenTo(tagName: ElementName, eventName: EventName, listener: (eventName.Event, tagName.Element) => Unit) = {
-  for (element <- document.getElementsByTagName(tagName) {
-    element.addEventListener(eventName)(event => listener(event, element))
+def findAndListenTo(eltName: ElementName, evtName: EventName)
+     (listener: (evtName.Event, eltName.Element) => Unit) = {
+  for (element <- document.getElementsByTagName(eltName) {
+    element.addEventListener(evtName) { event =>
+      listener(event, element)
+    }
   }
@@
 def observe(target: EventTarget, name: EventName) =
-  (listener: (name.Event => Unit)) => target.addEventListener(name)(listener)
+  (listener: (name.Event => Unit)) => {
+    target.addEventListener(name)(listener)
+  }
 \end{lstlisting}
@@
 
-We implemented our pattern in js-scala~\cite{Kossakowski12_JsDESL}, a Scala library providing JavaScript code generators\footnote{Source code is available at \href{http://github.com/js-scala}{http://github.com/js-scala}}.
+We implemented our encoding in js-scala~\cite{Kossakowski12_JsDESL}, a Scala library providing composable JavaScript code generators\footnote{Source code is available at \href{http://github.com/js-scala}{http://github.com/js-scala}}.
+On top of that we implemented various samples, including non trivial ones like a realtime chat application and a poll application.
+
+Our encoding ascribes types as precise as possible but is still expressive enough to implement the challenging \jscode{findAndListenTo} and \jscode{observe} functions that were impossible to implement with other approaches.
 
@@
 
-We do not help with \jscode{getElementById}.
+Our encodings only work with cases where a polymorphic type can be fixed by a value.
+In our examples, the only one that is not in this case is \jscode{getElementById}.
+Therefore we are not able to type this function more accurately (achieving this would require to support the DOM tree itself in the type system as in~\cite{Lerner13_TypedJQuery}).
 
-The name of an element or the name of an event can not anymore be a value resulting of the concatenation of Strings.
+Our solution is actually slightly less expressive than the JavaScript API: indeed, the value representing the name of an event or an element is not anymore a \jscode{String}, so it can not anymore be the result of a \jscode{String} manipulation, like \emph{e.g.} a concatenation.
+Fortunately, this case is uncommon.
 
@@
 Ravi Chugh \emph{et.
-al.}~\cite{Chugh12_DJS} showed how to make a subset of JavaScript statically typed using a dependent type system.
+al.} showed how to make a subset of JavaScript statically typed using a dependent type system~\cite{Chugh12_DJS}.
 They require complex type annotations to be written by developers.
+
+Sebastien Doreane proposed a way to integrate JavaScript APIs in Scala~\cite{Doeraene13_ScalaJs}.
+His approach allows developers to seamlessly use JavaScript APIs from statically typed Scala code.
+However, he did not use the features of Scala's type system, as we did, to ascribe more precise types when possible.
+
+TypeScript supports overloading on constant values.
+This solution is type safe and as expressive as the native API but has limited applicability (overload resolution requires parameters to be constant values, so the \jscode{findAndListenTo} function would be weakly typed with this approach).
 
@@
 We presented two ways to encode dynamically typed browser functions in mainstream statically typed languages like Java and Scala, using type parameters or path-dependent types.
-Our encoding gives more type safety than existing solutions, while keeping the same expression power as the native API.
+Our encodings give more type safety than existing solutions, while keeping the same expression power as the native API.
 

### 2014-02-04 papers/icwe14/icwe14.tex
++i

@@
 \begin{abstract}
-Encoding dynamically typed APIs in statically typed languages can be challenging.
-Incidentally, existing statically typed languages compiling to JavaScript expose its native API in ways that either are not type safe or give less expression power to developers.
-We present two new ways to encode the challenging parts of the Web browser API such that both type safety and expression power are preserved.
+Statically typed programming languages compiling to JavaScript make it worth considering the benefits of static typing to write Web applications.
+Nevertheless, at some point these languages need to expose the Web browser dynamically typed native API, which is a challenging task.
+Indeed, we observe that existing statically typed languages compiling to JavaScript expose the browser API in ways that either are not type safe or give less expression power and therefore force developers to pick only one of these two properties.
+This article presents new ways to encode the challenging parts of the Web browser API in static type systems such that both type safety and expression power are preserved.
 Our first encoding relies on type parameters and can be implemented in most mainstream languages but drags phantom types up to the usage sites.
@@
 We recently observed the emergence of several statically typed programming languages compiling to JavaScript (\emph{e.g.} Java/GWT~\cite{Kereki09_GWT}, Dart~\cite{Griffith11_Dart}, TypeScript~\cite{fenton2012typescript}, Kotlin\footnote{\href{http://kotlin.jetbrains.org}{http://kotlin.jetbrains.org}}, Opa\footnote{\href{http://opalang.org}{http://opalang.org}}, SharpKit\footnote{\href{http://sharpkit.net}{http://sharpkit.net}}, Haxe~\cite{Cannasse08_HaXe}, Scala~\cite{Doeraene13_ScalaJs}, Idris~\cite{Brady13_Idris}, Elm~\cite{czaplicki2012elm}).
-Developers can use them to write the client-side part of their Web applications, relying on a \emph{foreign function interface} mechanism to call the Web browser native API.
+These languages make it possible to benefit from static typing (\emph{e.g.} better refactoring support in IDE, earlier error detection) to write Web applications, usually written in the dynamically typed language JavaScript.
+Nevertheless, at some point developers need a way to interface with the underlying Web browser dynamically typed native API using a \emph{foreign function interface} mechanism.
 
-However we observe that, despite these languages are statically typed, their integration of the browser API either is not type safe or gives less expression power.
+We observe that, despite these languages are statically typed, their integration of the browser API either is not type safe or gives less expression power.
 Indeed, integrating an API designed for a dynamically typed language into a statically typed language can be challenging.
@@
 The next section reviews the most common functions of the browser API and how they are integrated in statically typed languages.
-Section \ref{sec-contribution} shows ways to improve their integration.
+Section \ref{sec-contribution} shows two ways to improve their integration such that type safety and expressivity are preserved.
 Section \ref{sec-validation} validates our contribution.
@@
 Listing \ref{lst-problem-cast} shows a Scala implementation of the \jscode{slideshow} program using an API following the standard specifications of HTML and DOM.
-The listing contains two type casts, making it possible to use the \scalacode{src} property on the \scalacode{img} value and the \scalacode{button} property on the \scalacode{event} value, respectively.
+The listing contains two type casts, needed to use the \scalacode{src} property on the \scalacode{img} value and the \scalacode{button} property on the \scalacode{event} value, respectively.
 
@@
 The best that could be done would be to create one function for each combination of tag name and event name, leading to an explosion of the number of functions to implement.
-Thus, this approach gives less expression power than the native browser API.
+Thus, this approach gives less
+expression power than the native browser API.
 
@@
 
-We first propose a solution in the Java mainstream language, using \emph{generics}, and show how we can improve it using \emph{path-dependent types}, in Scala.
+Our listings use the Scala language, though our first solution could be implemented in any language with basic type parameters support, such as Java's \emph{generics}.
+Our second solution is an improvement of the first one, using \emph{path-dependent types}.
 
@@
 \begin{enumerate}
- \item Define a parametrized class \scalacode{P<U>}
- \item Set the type of \scalacode{p} to \scalacode{P<U>}
+ \item Define a parametrized class \scalacode{P[U]}
+ \item Set the type of \scalacode{p} to \scalacode{P[U]}
  \item Use type \scalacode{U} instead of type \scalacode{T}
- \item Define as many values of type \scalacode{P<U>} as there are possible values for \scalacode{p}, each one fixing its \scalacode{U} type parameter to the corresponding more precise type
+ \item Define as many values of type \scalacode{P[U]} as there are possible values for \scalacode{p}, each one fixing its \scalacode{U} type parameter to the corresponding more precise type
 \end{enumerate}
 
-Listing~\ref{lst-generics-dom} shows this approach applied to the \jscode{createElement} function which return type depends on its \jscode{name} parameter value: a type \scalacode{ElementName<E>} has been created, the type of the \scalacode{name} parameter has been set to \scalacode{ElementName<E>} instead of \scalacode{String}, and the return type of the function is \scalacode{E} instead of \scalacode{Element}.
-The \scalacode{ElementName<E>} type encodes the relationship between the name of an element and the type of this element\footnote{The type parameter \scalacode{E} is also called a \emph{phantom type}~\cite{leijen1999domain} because \scalacode{ElementName} values never hold a \scalacode{E} value}.
-For instance, we created a value \scalacode{Input} of type \scalacode{ElementName<InputElement>}.
+\begin{figure}
+\begin{lstlisting}[label=lst-generics-dom,language=Scala,caption={Encoding of the \jscode{createElement} function using type parameters}]
+class ElementName[E]
 
-\begin{figure}
-\begin{lstlisting}[label=lst-generics-dom,language=java,caption={Encoding of the \jscode{createElement} function using type parameters}]
-class ElementName<E> {}
+trait Document {
+  def createElement[E](name: ElementName[E]): E
+  def getElementsByTagName[E](name: ElementName[E]): Array[E]
+}
 
-<E> E createElement(ElementName<E> name);
-
-final ElementName<InputElement> Input = new ElementName<>();
-final ElementName<ImageElement> Img = new ElementName<>();
+val Input = new ElementName[InputElement]
+val Img = new ElementName[ImageElement]
 // etc. for each possible element name
@@
 
-\scalacode{getElementsByTagName} can be straightforwardly encoded in a similar way, and listing~\ref{lst-generics-events} shows the encoding of the \jscode{addEventListener} function.
-Our listing assumes the definition of a \scalacode{Function<A, B>} type representing a function from \scalacode{A} to \scalacode{B}.
+Listing~\ref{lst-generics-dom} shows this approach applied to the \jscode{createElement} and \jscode{getElementsByTagName} functions which return type depends on their \jscode{name} parameter value: a type \scalacode{ElementName[E]} has been created, the type of the \scalacode{name} parameter has been set to \scalacode{ElementName[E]} instead of \scalacode{String}, and the return type of the function is \scalacode{E} instead of \scalacode{Element} (\scalacode{Array[E]} instead of \scalacode{Array[Element]}, in the case of \scalacode{getElementsByTagName}).
+The \scalacode{ElementName[E]} type encodes the relationship between the name of an element and the type of this element\footnote{The type parameter \scalacode{E} is also called a \emph{phantom type}~\cite{leijen1999domain} because \scalacode{ElementName} values never hold a \scalacode{E} value}.
+For instance, we created a value \scalacode{Input} of type \scalacode{ElementName[InputElement]}.
 
 \begin{figure}
-\begin{lstlisting}[label=lst-generics-events,language=java,caption={Encoding of the \jscode{addEventListener} function using type parameters}]
-class EventName<E> {}
+\begin{lstlisting}[label=lst-generics-events,language=Scala,caption={Encoding of the \jscode{addEventListener} function using type parameters}]
+class EventName[E]
 
-interface EventTarget {
-  <E> void addEventListener(
-          EventName<E> name, Function<E, Void> callback);
+trait EventTarget {
+  def addEventListener[E](
+          name: EventName[E], callback: E => Unit): Unit
 }
 
-final EventName<MouseEvent> Click = new EventName<>();
-final EventName<KeyboardEvent> KeyUp = new EventName<>();
+val Click = new EventName[MouseEvent]
+val KeyUp = new EventName[KeyboardEvent]
 // etc. for each possible event name
@@
 \end{figure}
+
+Listing~\ref{lst-generics-events} shows the encoding of the \jscode{addEventListener} function.
+The \scalacode{EventName[E]} type represents the name of an event which type is \scalacode{E}.
+For instance, \scalacode{Click} is a value of type \scalacode{EventName[MouseEvent]}: when a user adds an event listener to the \scalacode{Click} event, it fixes to \scalacode{MouseEvent} the type parameter \scalacode{E} of the \scalacode{callback} function passed to \scalacode{addEventListener}.
 
@@
 
-Listing \ref{lst-slideshow-generics} illustrate the usage of such an encoding by implementing our \texttt{slideshow} program in Scala.
+Listing \ref{lst-slideshow-generics} illustrates the usage of such an encoding by implementing our \texttt{slideshow} program in Scala.
 Passing the \scalacode{Img} value as a parameter to the \scalacode{createElement} function fixes its \scalacode{E} type parameter to \scalacode{ImageElement} so the returned value has the most possible precise type and the \scalacode{src} property can be used on it.
 Similarly, passing the \scalacode{Click} value to the \scalacode{addEventListener} function fixes its \scalacode{E} type parameter to \scalacode{MouseEvent}, so the event listener has the most possible precise type and the \scalacode{button} property can be used on the \scalacode{event} parameter.
+It is worth noting that there is no type cast in the listing because the browser API is exposed in a way that preserves enough type information.
 
 \begin{figure}
-\begin{lstlisting}[label=lst-generics-comb,language=java,caption={Combination of \scalacode{getElementsByTagName} and \scalacode{addEventListener} functions encoded using type parameters}]
-<A, B> void findAndListenTo(
-    ElementName<A> tagName,
-    EventName<B> eventName,
-    Function2<A, B, Void> listener) {
-  for (A element : document.getElementsByTagName(tagName)) {
-    element.addEventListener(
-         eventName,
-         event -> listener.apply(event, element));
+\begin{lstlisting}[label=lst-generics-comb,language=Scala,caption={Combination of \scalacode{getElementsByTagName} and \scalacode{addEventListener} functions encoded using type parameters}]
+def findAndListenTo[A, B](
+      tagName: ElementName[A],
+      eventName: EventName[B],
+      listener: (A, B) => Unit) = {
+  for (element <- document.getElementsByTagName(tagName)) {
+    element.addEventListener(eventName, event => {
+      listener(event, element)
+    })
   }
@@
 \begin{figure}
-\begin{lstlisting}[label=lst-generics-react,language=java,caption={Partial application of \scalacode{addEventListener} encoded with type parameters}]
-<A> Function<Function<A, Void>, Void> observe(
-      EventTarget target, EventName<A> name) {
-  return listener -> {
-    target.addEventListener(name, listener);
+\begin{lstlisting}[label=lst-generics-react,language=Scala,caption={Partial application of \scalacode{addEventListener} encoded with type parameters}]
+def observe[A](target: EventTarget, name: EventName[A]) = {
+  (listener: A => Unit) => {
+    target.addEventListener(name, listener)
   }
@@
 Finally, listing \ref{lst-generics-comb} and \ref{lst-generics-react} show how the challenging functions of section \ref{sec-limitations}, \jscode{findAndListenTo} and \jscode{observe}, can be implemented with our encoding.
-They are basically a direct translation from JavaScript to Java.
-
-Our encoding is type safe and gives as much expression power as the native API since it makes it possible to implement exactly the same functions as we are able to implement in plain JavaScript.
+They are basically a direct translation from JavaScript syntax to Scala syntax, with additional type annotations.
+Our encoding is type safe and gives as much expression power as the native API since it is possible to implement exactly the same functions as we are able to implement in plain JavaScript.
 
@@
 
-This section shows how we can remove the extra type parameters needed in the previous section by using \emph{path-dependent types}.
-Essentially, the idea is that, instead of using a type parameter, we use a \emph{type member}.
-Listings \ref{lst-dt-dom} and \ref{lst-dt-events} show this encoding in Scala for \jscode{createElement} and \jscode{addEventListener}, respectively.
+This section shows how we can remove the extra type parameters needed in the previous section by using \emph{path-dependent types}~\cite{Odersky03_vObj}.
+Essentially, the idea is to model type parameters using \emph{type members}, as suggested in~\cite{odersky2005scalable}.
 
@@
 
+trait Document {
   def createElement(name: ElementName): name.Element
+  def getElementsByTagName(
+        name: ElementName): Array[name.Element]
+}
 
@@
 
-Now, the \scalacode{ElementName} type has no type parameter but a type member \scalacode{Element}.
-The return type of the \scalacode{createElement} function is \scalacode{name.Element}: it refers to the \scalacode{Element} type member of its \scalacode{name} parameter.
-The \scalacode{Div} and \scalacode{Input} values illustrate how their corresponding element type is fixed.
-
 \begin{figure}
@@
   def addEventListener(name: EventName)
-                      (handler: name.Event => Unit): Unit
+                      (callback: name.Event => Unit): Unit
 }
@@
 
-[FIXME un peu plus de details sur le fonctionnement des path-dependent types]
-
-The implementation of the \scalacode{slideshow} function with this encoding is exactly the same as with the previous approach using generics.
-However, functions \jscode{findAndListenTo} and \jscode{observe} can be implemented more straightforwardly, as shown by listings \ref{lst-dt-comb} and \ref{lst-dt-react}, respectively.
+Listings \ref{lst-dt-dom} and \ref{lst-dt-events} show this encoding in Scala for \jscode{createElement}, \jscode{getElementsByTagName} and \jscode{addEventListener}.
+Now, the \scalacode{ElementName} type has no type parameter but a type member \scalacode{Element}.
+The return type of the \scalacode{createElement} function is \scalacode{name.Element}: it refers to the \scalacode{Element} type member of its \scalacode{name} parameter.
+The \scalacode{Div} and \scalacode{Input} values illustrate how their corresponding element type is fixed: if one writes \scalacode{createElement(Input)}, the return type is the \scalacode{Element} type member of the \scalacode{Input} value, namely \scalacode{InputElement}.
+The same idea applies to \scalacode{EventName} and \scalacode{addEventListener}: the name of the event fixes the type of the callback.
 
@@
 
+The implementation of the \scalacode{slideshow} function with this encoding is exactly the same as with the previous approach using generics.
+However, functions \jscode{findAndListenTo} and \jscode{observe} can be implemented more straightforwardly, as shown by listings \ref{lst-dt-comb} and \ref{lst-dt-react}, respectively.
+
 With this encoding, the functions using event names or element names are not anymore cluttered with phantom types, and type safety is still preserved.
@@
 On top of that we implemented various samples, including non trivial ones like a realtime chat application and a poll application.
+
+Mesurer les gains.
 
@@
 
+Christian Hofer \emph{et. al.} leveraged the fact that type parameters can be modeled with type members~\cite{hofer2008polymorphic}.
+
 \section{Conclusion}

### 2014-02-06 papers/icwe14/icwe14.tex
almost finished

@@
 
- \title{Using Path-Dependent Types to Build Type-Safe JavaScript Foreign Function Interfaces}
+ \title{Using Path-Dependent Types to Build Type Safe JavaScript Foreign Function Interfaces}
 
@@
 Nevertheless, at some point these languages need to expose the Web browser dynamically typed native API, which is a challenging task.
-Indeed, we observe that existing statically typed languages compiling to JavaScript expose the browser API in ways that either are not type safe or give less expression power and therefore force developers to pick only one of these two properties.
+Indeed, we observe that existing statically typed languages compiling to JavaScript expose the browser API in ways that either are not type safe or give less expression power, and therefore force developers to pick only one of these two properties.
 This article presents new ways to encode the challenging parts of the Web browser API in static type systems such that both type safety and expression power are preserved.
@@
 Most of the aforementionned languages expose this function by making it return an \jscode{Element}, the least upper bound of the types of all the possible returned values, thus loosing type information and requiring users to explicitly downcast the returned value to its expected more precise type.
-Another way to expose this function consists in exposing several functions, each one fixing the value of the parameter along with its return type: \jscode{createDivElement}, \jscode{createInputElement}, \emph{etc.} are parameterless functions returning a \jscode{DivElement} and an \jscode{InputElement}, respectively.
+Another way to expose this function consists in exposing several functions, each one fixing the value of the initial parameter along with its return type: \jscode{createDivElement}, \jscode{createInputElement}, \emph{etc.} are parameterless functions returning a \jscode{DivElement} and an \jscode{InputElement}, respectively.
 This encoding forces to hard-code the name of the element to create: it can not anymore be a parameter.
@@
 Section \ref{sec-contribution} shows two ways to improve their integration such that type safety and expressivity are preserved.
-Section \ref{sec-validation} validates our contribution.
+Section \ref{sec-validation} validates our contribution and discusses its limits.
 Section \ref{sec-related} discusses some related works and section \ref{sec-conclusion} concludes.
@@
 
-This section reviews the most commonly used browser functions and presents the different integration strategies currently used by statically typed programming languages like GWT, Dart, TypeScript, Kotlin, Opa, SharpKit, Haxe, Scala, Idris or Elm.
+This section reviews the most commonly used browser functions and presents the different integration strategies currently used by the statically typed programming languages GWT, Dart, TypeScript, Kotlin, Opa, SharpKit, Haxe, Scala, Idris and Elm.
+
+All languages support a foreign function interface mechanism, allowing developers to write JavaScript expressions from their programs.
+This mechanism is generally untyped and error prone, that's why most languages (TypeScript, Kotlin, Opa, SharpKit, Haxe, Scala and Elm) support a way to define \emph{external} typed interfaces.
+Most of them also expose the browser API through this way, as described in the next section.
 
@@
 
-To illustrate the problems caused by these functions, we present a simple JavaScript program using them and show how it can be implemented in statically typed programming languages according to the different strategies used to encode these functions.
+To illustrate the challenges raised by these functions, we present a simple JavaScript program using them and show how it can be implemented in statically typed programming languages according to the different strategies used to encode these functions.
 Listing \ref{lst-problem} shows the initial JavaScript code of the program.
 It defines a function \jscode{slideshow} that creates a slide show from an array of image URLs.
-The function returns an image element displaying the first image of the slide show, and each time a user clicks on it the next image is displayed.
+The function returns an image element displaying the first image of the slide show, and each time a user clicks on it with the mouse left button the next image is displayed.
 
@@
 However, in the case of \jscode{getElementsByTagName(name)} and \jscode{createElement(name)}, there is exactly one possible return type for each value of the \jscode{name} parameter: \emph{e.g.} \jscode{getElementsByTagName('input')} always returns a list of \jscode{InputElement} and \jscode{createElement('div')} always returns a \jscode{DivElement}.
-This characteristic makes it possible to encode each of these functions by defining as many parameterless functions as there are possible tag names, where each function fixes the initial \jscode{name} parameter to be one of the possible values and exposes the corresponding specialized return type.
+This characteristic makes it possible to encode these two functions by defining as many parameterless functions as there are possible tag names, where each function fixes the initial \jscode{name} parameter to be one of the possible values and exposes the corresponding specialized return type.
 
@@
 Listing~\ref{lst-problem-comb} shows what would our \scalacode{slideshow} implementation look like using such an encoding.
+There are two modifications compared to listing \ref{lst-problem-cast}: we use \scalacode{document.createImageElement} instead of \scalacode{document.createElement}, and we use \scalacode{img.addClickEventListener} instead of \scalacode{img.addEventListener}.
+
 The \scalacode{createImageElement} function takes no parameter and returns a value of type \scalacode{ImageElement}, and the \scalacode{addClickEventListener} function takes as parameter a function that takes a \scalacode{MouseEvent} value as parameter, ruling the need of type casts out.
 
-In the case of \jscode{addEventListener} we also encountered a slight variation of the encoding, consisting of defining one general function taking one parameter carrying both the 
+In the case of the \jscode{addEventListener} function we also encountered a slight variation of the encoding, consisting of defining one general function taking one parameter carrying both the 
 information of the event name and the event listener.
@@
 
-[TODO Donner quand meme des chiffres un peu detailles sur quel langage utilise quelle strategie]
+Most of the studied languages expose the browser API following the standard specification, but some of them (GWT, Dart, HaXe and Elm) define a modified API getting rid of (or at least reducing) the need for downcasting, following the approaches described above.
 
@@
 The limits of this approach are better illustrated when one tries to combine several functions.
-Consider for instance listing~\ref{lst-js-comb} defining a JavaScript function \jscode{findAndListenTo} that both find elements and registers an event listener when a given event occurs on them.
+Consider for instance listing~\ref{lst-js-comb} defining a JavaScript function \jscode{findAndListenTo} that both finds elements and registers an event listener when a given event occurs on them.
 Note that the event listener is passed both the event and the element: its type depends on both the tag name and the event name.
@@
   });
-};
+}
 \end{lstlisting}
@@
 The third approach, consisting in combining two parameters into one parameter carrying all the required information, is type safe too, but reduces the expression power because it forbids developers to partially apply the function by supplying only one parameter.
-Consider for instance listing~\ref{lst-js-react} that defines a function \jscode{observe} partially applying the \jscode{addEventListener} function\footnote{This pattern is often used by functional reactive programming libraries like Rx.js~\cite{liberty2011reactive}}.
-Such a function can not be implemented with this approach.
+Consider for instance listing~\ref{lst-js-react} that defines a function \jscode{observe} partially applying the \jscode{addEventListener} function\footnote{The code of this function has been taken (and simplified) from the existing functional reactive programming libraries Rx.js~\cite{liberty2011reactive} and Bacon.js~(\href{http://baconjs.github.io/}{http://baconjs.github.io/}).}.
+Such a function can not be implemented with this approach because the name of the event and the code of the listener can not be decoupled.
 Thus, this one gives less expression power than the native browser API.
@@
   }
-};
+}
 \end{lstlisting}
@@
 In summary, the current integration of the browser API by statically typed languages compiling to JavaScript is either not type safe or not as expressive as the underlying JavaScript API.
+Indeed, we showed that our simple \jscode{slideshow} program requires type casts if the browser API is exposed according to the standard specification.
+We are able to get rid of type casts on this program by using modified browser APIs, but we presented two functions that we were not able to implement using these APIs, showing that they give less expression power than the native API.
+
+This article aims to answer the following questions: is it possible to expose the browser API in statically typed languages in a way that both reduces the need for type casts and preserves the same expression power?
+What typing mechanisms do we need to achieve this?
+Would it be convenient to use by end developers?
 
@@
 
-Listing~\ref{lst-generics-dom} shows this approach applied to the \jscode{createElement} and \jscode{getElementsByTagName} functions which return type depends on their \jscode{name} parameter value: a type \scalacode{ElementName[E]} has been created, the type of the \scalacode{name} parameter has been set to \scalacode{ElementName[E]} instead of \scalacode{String}, and the return type of the function is \scalacode{E} instead of \scalacode{Element} (\scalacode{Array[E]} instead of \scalacode{Array[Element]}, in the case of \scalacode{getElementsByTagName}).
+Listing~\ref{lst-generics-dom} shows this approach applied to the \jscode{createElement} and \jscode{getElementsByTagName} functions which return type depends on their \jscode{name} parameter value: a type \scalacode{ElementName[E]} has been created, the type of the \scalacode{name} parameter has been set to \scalacode{ElementName[E]} instead of \scalacode{String}, and the return type of the function is \scalacode{E} instead of \scalacode{Element} (or \scalacode{Array[E]} instead of \scalacode{Array[Element]}, in the case of \scalacode{getElementsByTagName}).
 The \scalacode{ElementName[E]} type encodes the relationship between the name of an element and the type of this element\footnote{The type parameter \scalacode{E} is also called a \emph{phantom type}~\cite{leijen1999domain} because \scalacode{ElementName} values never hold a \scalacode{E} value}.
@@
 
-Listing \ref{lst-slideshow-generics} illustrates the usage of such an encoding by implementing our \texttt{slideshow} program in Scala.
+Listing \ref{lst-slideshow-generics} illustrates the usage of such an encoding by implementing our \texttt{slideshow} program presented in the introduction.
 Passing the \scalacode{Img} value as a parameter to the \scalacode{createElement} function fixes its \scalacode{E} type parameter to \scalacode{ImageElement} so the returned value has the most possible precise type and the \scalacode{src} property can be used on it.
@@
 
+Usually, in object oriented programming languages, a type \scalacode{T} can have abstract methods.
+In Scala it can also have abstract type members.
+Such a type \scalacode{T} is abstract and its concrete subtypes have to define their type members (like abstract methods have to be implemented).
+
 \begin{figure}
@@
 
-Listings \ref{lst-dt-dom} and \ref{lst-dt-events} show this encoding in Scala for \jscode{createElement}, \jscode{getElementsByTagName} and \jscode{addEventListener}.
+Listings \ref{lst-dt-dom} and \ref{lst-dt-events} show an encoding of \jscode{createElement}, \jscode{getElementsByTagName} and \jscode{addEventListener} in Scala using type members.
 Now, the \scalacode{ElementName} type has no type parameter but a type member \scalacode{Element}.
@@
 
-Mesurer les gains.
+We showed in this paper that our encoding ascribes types as precise as possible (our \scalacode{slideshow} program is free of type casts) while being expressive enough to implement the challenging \jscode{findAndListenTo} and \jscode{observe} functions that were impossible to implement with other approaches.
 
-Our encoding ascribes types as precise as possible but is still expressive enough to implement the challenging \jscode{findAndListenTo} and \jscode{observe} functions that were impossible to implement with other approaches.
+\subsection{Convenience to use for end developers}
+
+Statically typed languages can be criticized for the verbosity of the information they add compared to dynamically typed languages~\cite{meijer2004static}.
+In our case, what is the price to pay to get accurate type ascriptions?
+
+The first encoding, using type parameters, can be implemented in most programming languages because it only requires a basic support of type parameters (for instance Java, Dart, TypeScript, Kotlin, HaXe, Opa, Idris and Elm can implement it).
+However this encoding leads to cluttered type signatures and forces functions parametrized by event or element names to also take phantom type parameters.
+
+However, the second encoding, using type members, leads to type signatures that are not more verbose than those of the standard specifications of the HTML and DOM APIs, so we argue that there is no price to pay.
+This encoding can only be implemented in language supporting type members or dependent types (Scala and Idris), though.
 
@@
 
-Ravi Chugh \emph{et.
-al.} showed how to make a subset of JavaScript statically typed using a dependent type system~\cite{Chugh12_DJS}.
-They require complex type annotations to be written by developers.
+The idea of using dependent types to type JavaScript has already been explored by Ravi Chugh \emph{et.
+al.}~\cite{Chugh12_DJS}.
+They showed how to make a subset of JavaScript statically typed using a dependent type system.
+However, their solution requires complex and verbose type annotations to be written by developers.
 
@@
 
-TypeScript supports overloading on constant values.
+TypeScript supports overloading on constant values: the type of the expression \jscode{createElement("div")} is statically resolved to \jscode{DivElement} by the constant parameter value \jscode{"div"}.
 This solution is type safe and as expressive as the native API but has limited applicability (overload resolution requires parameters to be constant values, so the \jscode{findAndListenTo} function would be weakly typed with this approach).
-
-Christian Hofer \emph{et. al.} leveraged the fact that type parameters can be modeled with type members~\cite{hofer2008polymorphic}.
 

### 2014-02-12 papers/icwe14/icwe14.tex
jmj

@@
 We recently observed the emergence of several statically typed programming languages compiling to JavaScript (\emph{e.g.} Java/GWT~\cite{Kereki09_GWT}, Dart~\cite{Griffith11_Dart}, TypeScript~\cite{fenton2012typescript}, Kotlin\footnote{\href{http://kotlin.jetbrains.org}{http://kotlin.jetbrains.org}}, Opa\footnote{\href{http://opalang.org}{http://opalang.org}}, SharpKit\footnote{\href{http://sharpkit.net}{http://sharpkit.net}}, Haxe~\cite{Cannasse08_HaXe}, Scala~\cite{Doeraene13_ScalaJs}, Idris~\cite{Brady13_Idris}, Elm~\cite{czaplicki2012elm}).
-These languages make it possible to benefit from static typing (\emph{e.g.} better refactoring support in IDE, earlier error detection) to write Web applications, usually written in the dynamically typed language JavaScript.
+Though dynamic typing has its pros and supporters, these statically typed languages show that there is also a community interested in benefiting from static typing features (\emph{e.g.} better refactoring support in IDE, earlier error detection) to write Web applications.
 Nevertheless, at some point developers need a way to interface with the underlying Web browser dynamically typed native API using a \emph{foreign function interface} mechanism.
@@
 We distinguished three approaches to integrate the challenging parts of the browser API into statically typed languages.
-This section compares these approaches in terms of type safety and expression power.
+This section shows that each approach favours either type safety or expression power but none provides both type safety \emph{and} the same expression power as the native browser API.
+We consider that an API requiring users to do type casts is not type safe and that an API making it impossible implement a function that can yet be implemented using the native API gives less expression power.
 
@@
 
-Our listings use the Scala language, though our first solution could be implemented in any language with basic type parameters support, such as Java's \emph{generics}.
+The listings in the paper use the Scala language, though our first solution could be implemented in any language with basic type parameters support, such as Java's \emph{generics}\footnote{Java versions of all the Scala listings (excepted those using type members) are available online at \href{http://github.com/js-scala/js-scala/wiki/ICWE'14}{http://github.com/js-scala/js-scala/wiki/ICWE'14}}.
 Our second solution is an improvement of the first one, using \emph{path-dependent types}.
@@
 Similarly, passing the \scalacode{Click} value to the \scalacode{addEventListener} function fixes its \scalacode{E} type parameter to \scalacode{MouseEvent}, so the event listener has the most possible precise type and the \scalacode{button} property can be used on the \scalacode{event} parameter.
-It is worth noting that there is no type cast in the listing because the browser API is exposed in a way that preserves enough type information.
+
+It is worth noting that this code is actually exactly the same as in listing \ref{lst-problem-cast} excepted that type casts are not anymore required because the browser API is exposed in a way that preserves enough type information.
+Our way to encode the browser API is more type safe, but is it as expressive as the native API?
 
@@
 
-Finally, listing \ref{lst-generics-comb} and \ref{lst-generics-react} show how the challenging functions of section \ref{sec-limitations}, \jscode{findAndListenTo} and \jscode{observe}, can be implemented with our encoding.
+Listings \ref{lst-generics-comb} and \ref{lst-generics-react} show how the challenging functions of section \ref{sec-limitations}, \jscode{findAndListenTo} and \jscode{observe}, can be implemented with our encoding.
 They are basically a direct translation from JavaScript syntax to Scala syntax, with additional type annotations.
-Our encoding is type safe and gives as much expression power as the native API since it is possible to implement exactly the same functions as we are able to implement in plain JavaScript.
+
+In summary, our encoding is type safe and gives as much expression power as the native API since it is possible to implement exactly the same functions as we are able to implement in plain JavaScript.
 
 However, every function taking an element name or an event name as parameter has its type signature cluttered with phantom types (extra type parameters): the \scalacode{observe} function takes a phantom type parameter \scalacode{A} and the \scalacode{findAndListenTo} function takes two phantom type parameters, \scalacode{A} and \scalacode{B}.
+These extra type parameters are redundant with their corresponding value parameters and they make type signatures harder to read and reason about.
 
@@
 
-Usually, in object oriented programming languages, a type \scalacode{T} can have abstract methods.
-In Scala it can also have abstract type members.
-Such a type \scalacode{T} is abstract and its concrete subtypes have to define their type members (like abstract methods have to be implemented).
+Programming languages generally support two means of abstraction: parametrization and abstract members.
+For instance Java supports parametrization for values (method parameters) and types (\emph{generics}), and member abstraction for values (abstract methods).
+Scala also supports member abstraction for types through type members~\cite{cremet2006core,Odersky03_vObj}.
+An abstract type member of a class is an inner abstract type that can be used to qualify values.
+Subclasses can implement and override their methods, and similarly they can define or refine their type members.
+A concrete subclass must provide a concrete implementation of its type members.
+Outside of the class, type members can be referred to using a type selection on an instance of the class: the type designator \scalacode{p.C} refers to the \scalacode{C} type member of the value \scalacode{p} and expands to the \scalacode{C} type member implementation of the singleton type of \scalacode{p}.
 

### 2014-02-13 papers/icwe14/icwe14.tex
jmj changes

@@
 
- \author{Julien Richard-Foy \and Olivier Barais}
+ \author{Julien Richard-Foy \and Olivier Barais \and Jean-Marc J\'ez\'equel}
 
@@
 \begin{abstract}
-Statically typed programming languages compiling to JavaScript make it worth considering the benefits of static typing to write Web applications.
-Nevertheless, at some point these languages need to expose the Web browser dynamically typed native API, which is a challenging task.
-Indeed, we observe that existing statically typed languages compiling to JavaScript expose the browser API in ways that either are not type safe or give less expression power, and therefore force developers to pick only one of these two properties.
-This article presents new ways to encode the challenging parts of the Web browser API in static type systems such that both type safety and expression power are preserved.
+The popularity of statically typed programming languages compiling to JavaScript show that there exists a fringe of the programmer population interested in leveraging the benefits of static typing to write Web applications.
+To be of any use, these languages need to statically expose the Web browser dynamically typed native API, which seems to be a contradiction in terms.
+Indeed, we observe that existing statically typed languages compiling to JavaScript expose the browser API in ways that either are not type safe, or when they are, typically over constrain the programmers.
+This article presents new ways to encode the challenging parts of the Web browser API in static type systems such that both type safety and expressive power are preserved.
 Our first encoding relies on type parameters and can be implemented in most mainstream languages but drags phantom types up to the usage sites.
-The second encoding does not suffer from this inconvenience but requires dependent types.
+The second encoding does not suffer from this inconvenience but requires the support of dependent types in the language.
 \end{abstract}
@@
 We recently observed the emergence of several statically typed programming languages compiling to JavaScript (\emph{e.g.} Java/GWT~\cite{Kereki09_GWT}, Dart~\cite{Griffith11_Dart}, TypeScript~\cite{fenton2012typescript}, Kotlin\footnote{\href{http://kotlin.jetbrains.org}{http://kotlin.jetbrains.org}}, Opa\footnote{\href{http://opalang.org}{http://opalang.org}}, SharpKit\footnote{\href{http://sharpkit.net}{http://sharpkit.net}}, Haxe~\cite{Cannasse08_HaXe}, Scala~\cite{Doeraene13_ScalaJs}, Idris~\cite{Brady13_Idris}, Elm~\cite{czaplicki2012elm}).
-Though dynamic typing has its pros and supporters, these statically typed languages show that there is also a community interested in benefiting from static typing features (\emph{e.g.} better refactoring support in IDE, earlier error detection) to write Web applications.
+Though dynamic typing has its merits and supporters, the mere existence of these statically typed languages shows that there is also a community interested in benefiting from static typing features (allowing \emph{e.g.} better refactoring support in IDE, earlier error detection, etc.) to write Web applications.
 Nevertheless, at some point developers need a way to interface with the underlying Web browser dynamically typed native API using a \emph{foreign function interface} mechanism.
 
-We observe that, despite these languages are statically typed, their integration of the browser API either is not type safe or gives less expression power.
+We observe that, despite these languages are statically typed, their integration of the browser API either is not type safe or over constrain the programmers.
 Indeed, integrating an API designed for a dynamically typed language into a statically typed language can be challenging.
@@
 
-Most of the aforementionned languages expose this function by making it return an \jscode{Element}, the least upper bound of the types of all the possible returned values, thus loosing type information and requiring users to explicitly downcast the returned value to its expected more precise type.
+Most of the aforementionned languages expose this function by making it return an \jscode{Element}, the least upper bound of the types of all the possible returned values, thus loosing type information and requiring users to explicitly downcast the returned value to its expected, more precise type.
 Another way to expose this function consists in exposing several functions, each one fixing the value of the initial parameter along with its return type: \jscode{createDivElement}, \jscode{createInputElement}, \emph{etc.} are parameterless functions returning a \jscode{DivElement} and an \jscode{InputElement}, respectively.
-This encoding forces to hard-code the name of the element to create: it can not anymore be a parameter.
-In summary, the first solution is not type safe and the second solution reduces the expression power.
+This encoding forces to hard-code the name of the to-be created element: it cannot anymore be a parameter.
+In summary, the first solution is not type safe and the second solution reduces the expressive power of the API.
 
-This paper reviews some common functions of the browser API, identifies the patterns that are difficult to encode in static type systems and shows new ways to encode them such that both type safety and expression power are preserved.
-We find that type parameters are sufficient to achieve this goal and that \emph{path-dependent types} provide an even more convenient encoding for the end developers.
+This paper reviews some common functions of the browser API, identifies the patterns that are difficult to encode in static type systems and shows new ways to encode them in such a way that both type safety and expression power are preserved.
+We show that type parameters are sufficient to achieve this goal and that \emph{path-dependent types} provide an even more convenient encoding of the browser API.
 
 The remainder of the paper is organized as follows.
-The next section reviews the most common functions of the browser API and how they are integrated in statically typed languages.
+The next section reviews the most common functions of the browser API and how they are typically integrated into statically typed languages.
 Section \ref{sec-contribution} shows two ways to improve their integration such that type safety and expressivity are preserved.
@@
 
-All languages support a foreign function interface mechanism, allowing developers to write JavaScript expressions from their programs.
-This mechanism is generally untyped and error prone, that's why most languages (TypeScript, Kotlin, Opa, SharpKit, Haxe, Scala and Elm) support a way to define \emph{external} typed interfaces.
-Most of them also expose the browser API through this way, as described in the next section.
+All these languages support a foreign function interface mechanism, allowing developers to write JavaScript expressions from their programs.
+Since this mechanism is generally untyped and error prone, most languages (TypeScript, Kotlin, Opa, SharpKit, Haxe, Scala and Elm) support a way to define \emph{external} typed interfaces.
+Most of them also expose the browser API this way, as described in the next section.
 
@@
 The client-side part of the code of a Web application essentially reacts to user events (\emph{e.g.} mouse clicks), triggers actions and updates the document (DOM) according to their effect.
-Table~\ref{table-dom-api} lists the main functions supported by Web browsers according to the Mozilla Developer Network\footnote{\href{https://developer.mozilla.org/en-US/docs/DOM/DOM\_Reference/Introduction}{https://developer.mozilla.org/en-US/docs/DOM/DOM\_Reference/Introduction}} (we removed the functions that can trivially be encoded in a static type system).
+Table~\ref{table-dom-api} lists the main functions supported by Web browsers according to the Mozilla Developer Network\footnote{\href{https://developer.mozilla.org/en-US/docs/DOM/DOM\_Reference/Introduction}{https://developer.mozilla.org/en-US/docs/DOM/DOM\_Reference/Introduction}} (we omit the functions that can trivially be encoded in a static type system).
 
@@
 
-The main challenge comes from the fact that these function return types or parameter types are often too general.
+The main challenge comes from the fact that the parameter types and return types of these function are often too general.
 Indeed, functions \jscode{getElementsByTagName(name)}, \jscode{getElementById(id)} and \jscode{createElement(name)} can return values of type \jscode{DivElement} or \jscode{InputElement} or any other subtype of \jscode{Element} (their least upper bound).
@@
 
-These type casts make the code more fragile and less convenient to read and write. That's why some statically typed languages attempt to provide an API preserving as precise types as possible.
+These type casts make the code more fragile and less convenient to read and write. That's why some statically typed languages attempt to provide an API preserving types as precisly as possible.
 
-In the case of \jscode{getElementById(id)}, the \jscode{id} parameter gives no clue on the possible type of the searched element so it is hard to infer more precisely the return type of this function.
+Of course, in the case of \jscode{getElementById(id)}, the \jscode{id} parameter does not give any clue on the possible type of the searched element, so it is hard to more precisely infer the return type of this function.
 Hence, most implementations define a return type of \jscode{Element}.
@@
 However, in the case of \jscode{getElementsByTagName(name)} and \jscode{createElement(name)}, there is exactly one possible return type for each value of the \jscode{name} parameter: \emph{e.g.} \jscode{getElementsByTagName('input')} always returns a list of \jscode{InputElement} and \jscode{createElement('div')} always returns a \jscode{DivElement}.
-This characteristic makes it possible to encode these two functions by defining as many parameterless functions as there are possible tag names, where each function fixes the initial \jscode{name} parameter to be one of the possible values and exposes the corresponding specialized return type.
+This feature makes it possible to encode these two functions by defining as many parameterless functions as there are possible tag names, where each function fixes the initial \jscode{name} parameter to be one of the possible values and exposes the corresponding specialized return type.
 
 The case of \jscode{target.addEventListener(name, listener)} is a bit different.
-The \jscode{name} parameter defines the event to listen to and the \jscode{listener} parameter the function to call back each time such an event occurs.
+The \jscode{name} parameter defines the event to listen to while the \jscode{listener} parameter identifies the function to call back each time such an event occurs.
 Instead of being polymorphic in its return type, it is polymorphic in its \jscode{listener} parameter.
@@
 
-The \scalacode{createImageElement} function takes no parameter and returns a value of type \scalacode{ImageElement}, and the \scalacode{addClickEventListener} function takes as parameter a function that takes a \scalacode{MouseEvent} value as parameter, ruling the need of type casts out.
+The \scalacode{createImageElement} function takes no parameter and returns a value of type \scalacode{ImageElement}, and the \scalacode{addClickEventListener} function takes as parameter a function that takes a \scalacode{MouseEvent} value as parameter, ruling out the need for type casts.
 
-In the case of the \jscode{addEventListener} function we also encountered a slight variation of the encoding, consisting of defining one general function taking one parameter carrying both the 
+In the case of the \jscode{addEventListener} function we also encountered a slight variation of the encoding, consisting in defining one general function taking one parameter carrying both the 
 information of the event name and the event listener.
@@
 This section shows that each approach favours either type safety or expression power but none provides both type safety \emph{and} the same expression power as the native browser API.
-We consider that an API requiring users to do type casts is not type safe and that an API making it impossible implement a function that can yet be implemented using the native API gives less expression power.
+We indeed  consider that an API requiring users to do type casts is not type safe, while an API making it impossible to implement a function that can readily be implemented using the native API gives less expressive power to the programmer.
 
@@
 Note that the event listener is passed both the event and the element: its type depends on both the tag name and the event name.
-This function can not be implemented if the general functions \jscode{getElementsByTagName} and \jscode{addEventListener} are not available.
+This function cannot be implemented if the general functions \jscode{getElementsByTagName} and \jscode{addEventListener} are not available.
 The best that could be done would be to create one function for each combination of tag name and event name, leading to an explosion of the number of functions to implement.
 Thus, this approach gives less
-expression power than the native browser API.
+expressive power than the native browser API.
 
@@
 Consider for instance listing~\ref{lst-js-react} that defines a function \jscode{observe} partially applying the \jscode{addEventListener} function\footnote{The code of this function has been taken (and simplified) from the existing functional reactive programming libraries Rx.js~\cite{liberty2011reactive} and Bacon.js~(\href{http://baconjs.github.io/}{http://baconjs.github.io/}).}.
-Such a function can not be implemented with this approach because the name of the event and the code of the listener can not be decoupled.
-Thus, this one gives less expression power than the native browser API.
+Such a function cannot be implemented with this approach because the name of the event and the code of the listener cannot be decoupled.
+Thus, this one gives less expressive power than the native browser API.
 
@@
 What typing mechanisms do we need to achieve this?
-Would it be convenient to use by end developers?
+Would it be convenient to be used by end developers?
 
@@
 
-In this section we show how we can encode the challenging main functions of the DOM API in a type safe way while keeping the same expression power.
+In this section we show how we can encode the challenging main functions of the DOM API in a type safe way while keeping the same expressive power.
 
-The listings in the paper use the Scala language, though our first solution could be implemented in any language with basic type parameters support, such as Java's \emph{generics}\footnote{Java versions of all the Scala listings (excepted those using type members) are available online at \href{http://github.com/js-scala/js-scala/wiki/ICWE'14}{http://github.com/js-scala/js-scala/wiki/ICWE'14}}.
-Our second solution is an improvement of the first one, using \emph{path-dependent types}.
+The listings in this paper use the Scala language, though our first solution could be implemented in any language with basic type parameters support, such as Java's \emph{generics}\footnote{For a lack of space, we do not present them here but all Java versions of all the Scala listings (excepted those using type members) are available online at \href{http://github.com/js-scala/js-scala/wiki/ICWE'14}{http://github.com/js-scala/js-scala/wiki/ICWE'14}}.
+Our second solution is an improvement over the first one, using \emph{path-dependent types}.
 
@@
 
-We showed in this paper that our encoding ascribes types as precise as possible (our \scalacode{slideshow} program is free of type casts) while being expressive enough to implement the challenging \jscode{findAndListenTo} and \jscode{observe} functions that were impossible to implement with other approaches.
+We have shown in this paper that our encoding leverages types as precisely as possible (our \scalacode{slideshow} program is free of type casts) while being expressive enough to implement the challenging \jscode{findAndListenTo} and \jscode{observe} functions that were impossible to implement with other approaches.
 
-\subsection{Convenience to use for end developers}
+\subsection{Convenience for end developers}
 
-Statically typed languages can be criticized for the verbosity of the information they add compared to dynamically typed languages~\cite{meijer2004static}.
+Statically typed languages are often criticized for the verbosity of the information they add compared to dynamically typed languages~\cite{meijer2004static}.
 In our case, what is the price to pay to get accurate type ascriptions?
@@
 However, the second encoding, using type members, leads to type signatures that are not more verbose than those of the standard specifications of the HTML and DOM APIs, so we argue that there is no price to pay.
-This encoding can only be implemented in language supporting type members or dependent types (Scala and Idris), though.
+However, this encoding can only be implemented in language supporting type members or dependent types (Scala and Idris).
 
@@
 
-Our solution is actually slightly less expressive than the JavaScript API: indeed, the value representing the name of an event or an element is not anymore a \jscode{String}, so it can not anymore be the result of a \jscode{String} manipulation, like \emph{e.g.} a concatenation.
+Our solution is actually slightly less expressive than the JavaScript API: indeed, the value representing the name of an event or an element is not anymore a \jscode{String}, so it cannot anymore be the result of a \jscode{String} manipulation, like \emph{e.g.} a concatenation.
 Fortunately, this case is uncommon.
@@
 His approach allows developers to seamlessly use JavaScript APIs from statically typed Scala code.
-However, he did not use the features of Scala's type system, as we did, to ascribe more precise types when possible.
+However, he did not use the features of Scala's type system, as we did, to leverage more precise types when possible.
 

### 2014-02-13 papers/icwe14/icwe14.tex
numbers
tweaks

@@
 
-\usepackage[utf8]{inputenc}
 \usepackage[unicode=true,hidelinks]{hyperref}
@@
 
- \institute{IRISA, Universite de Rennes, France}
+ \institute{IRISA, Universit\'e de Rennes, France}
 
@@
 \begin{abstract}
-The popularity of statically typed programming languages compiling to JavaScript show that there exists a fringe of the programmer population interested in leveraging the benefits of static typing to write Web applications.
+The popularity of statically typed programming languages compiling to JavaScript shows that there exists a fringe of the programmer population interested in leveraging the benefits of static typing to write Web applications.
 To be of any use, these languages need to statically expose the Web browser dynamically typed native API, which seems to be a contradiction in terms.
@@
 Though dynamic typing has its merits and supporters, the mere existence of these statically typed languages shows that there is also a community interested in benefiting from static typing features (allowing \emph{e.g.} better refactoring support in IDE, earlier error detection, etc.) to write Web applications.
-Nevertheless, at some point developers need a way to interface with the underlying Web browser dynamically typed native API using a \emph{foreign function interface} mechanism.
+Nevertheless, at some point developers need a way to interface with the underlying Web browser dynamically typed native API using a \emph{foreign function
+interface} mechanism.
 
@@
 
-This paper reviews some common functions of the browser API, identifies the patterns that are difficult to encode in static type systems and shows new ways to encode them in such a way that both type safety and expression power are preserved.
+This paper reviews some common functions of the browser API, identifies the patterns that are difficult to encode in static type systems and shows new ways to encode them in such a way that both type safety and expressive power are preserved.
 We show that type parameters are sufficient to achieve this goal and that \emph{path-dependent types} provide an even more convenient encoding of the browser API.
@@
 
-The main challenge comes from the fact that the parameter types and return types of these function are often too general.
+The main challenge comes from the fact that the parameter types and return types of these functions are often too general.
 Indeed, functions \jscode{getElementsByTagName(name)}, \jscode{getElementById(id)} and \jscode{createElement(name)} can return values of type \jscode{DivElement} or \jscode{InputElement} or any other subtype of \jscode{Element} (their least upper bound).
@@
 We distinguished three approaches to integrate the challenging parts of the browser API into statically typed languages.
-This section shows that each approach favours either type safety or expression power but none provides both type safety \emph{and} the same expression power as the native browser API.
+This section shows that each approach favours either type safety or expressive power but none provides both type safety \emph{and} the same expressive power as the native browser API.
 We indeed  consider that an API requiring users to do type casts is not type safe, while an API making it impossible to implement a function that can readily be implemented using the native API gives less expressive power to the programmer.
 
-The first approach, consisting in using the least upper bound of all the possible types has the same expression power as the native browser API, but is not type safe because it sometimes requires developers to explicitly downcast values to their expected specialized type.
+The first approach, consisting in using the least upper bound of all the possible types has the same expressive power as the native browser API, but is not type safe because it sometimes requires developers to explicitly downcast values to their expected specialized type.
 
@@
 
-The third approach, consisting in combining two parameters into one parameter carrying all the required information, is type safe too, but reduces the expression power because it forbids developers to partially apply the function by supplying only one parameter.
+The third approach, consisting in combining two parameters into one parameter carrying all the required information, is type safe too, but reduces the expressive power because it forbids developers to partially apply the function by supplying only one parameter.
 Consider for instance listing~\ref{lst-js-react} that defines a function \jscode{observe} partially applying the \jscode{addEventListener} function\footnote{The code of this function has been taken (and simplified) from the existing functional reactive programming libraries Rx.js~\cite{liberty2011reactive} and Bacon.js~(\href{http://baconjs.github.io/}{http://baconjs.github.io/}).}.
@@
 Indeed, we showed that our simple \jscode{slideshow} program requires type casts if the browser API is exposed according to the standard specification.
-We are able to get rid of type casts on this program by using modified browser APIs, but we presented two functions that we were not able to implement using these APIs, showing that they give less expression power than the native API.
+We are able to get rid of type casts on this program by using modified browser APIs, but we presented two functions that we were not able to implement using these APIs, showing that they give less expressive power than the native API.
 
-This article aims to answer the following questions: is it possible to expose the browser API in statically typed languages in a way that both reduces the need for type casts and preserves the same expression power?
+This article aims to answer the following questions: is it possible to expose the browser API in statically typed languages in a way that both reduces the need for type casts and preserves the same expressive power?
 What typing mechanisms do we need to achieve this?
@@
 
-In summary, our encoding is type safe and gives as much expression power as the native API since it is possible to implement exactly the same functions as we are able to implement in plain JavaScript.
+In summary, our encoding is type safe and gives as much expressive power as the native API since it is possible to implement exactly the same functions as we are able to implement in plain JavaScript.
 
@@
 
+For comparison, our single function \scalacode{addEventListener} is as type safe as the about thirty functions defined in GWT for event handling while giving more expressive power to users.
+
 \subsection{Convenience for end developers}
@@
 We presented two ways to encode dynamically typed browser functions in mainstream statically typed languages like Java and Scala, using type parameters or path-dependent types.
-Our encodings give more type safety than existing solutions, while keeping the same expression power as the native API.
+Our encodings give more type safety than existing solutions, while keeping the same expressive power as the native API.
 

### 2014-02-24 papers/icwe14/icwe14.tex
j-1

@@
 The next section reviews the most common functions of the browser API and how they are typically integrated into statically typed languages.
-Section \ref{sec-contribution} shows two ways to improve their integration such that type safety and expressivity are preserved.
+Section \ref{sec-contribution} shows two ways to improve their integration such that type safety and expressiveness are preserved.
 Section \ref{sec-validation} validates our contribution and discusses its limits.
@@
 expressive power than the native browser API.
+Moreover, we find that defining many functions for the same task (creating a DOM element or listening to an event) clutters the API documentation: functions serving other purposes are hidden by these \emph{same-purpose-functions}.
 
@@
 
-For comparison, our single function \scalacode{addEventListener} is as type safe as the about thirty functions defined in GWT for event handling while giving more expressive power to users.
+\subsection{API clarity}
+
+We mentioned in the background section that a common drawback of existing approaches to bring more type safety was the multiplication of functions having the same purpose, making the API documentation harder to read.
+
+Our encoding preserves a one to one mapping with browser API functions whereas existing approaches often have more than 30 functions for a same purpose.
+For instance, the \jscode{createElement} function is mapped by 31 specialized functions in GWT and 62 in Dart, the \jscode{addEventListener} is mapped by 32 specialized functions in GWT and 49 in Dart.
 
@@
 TypeScript supports overloading on constant values: the type of the expression \jscode{createElement("div")} is statically resolved to \jscode{DivElement} by the constant parameter value \jscode{"div"}.
-This solution is type safe and as expressive as the native API but has limited applicability (overload resolution requires parameters to be constant values, so the \jscode{findAndListenTo} function would be weakly typed with this approach).
+This solution is type safe, as expressive and as easy to learn as the native API because its functions have a one to one mapping.
+However, this kind of overloading has limited applicability because overload resolution requires parameters to be constant values: indeed, the \jscode{findAndListenTo} function would be weakly typed with this approach.
 
@@
 
+Having a statically typed programming language compiling to JavaScript is not enough to leverage static typing in Web applications.
+The native browser API has to be exposed in a statically typed way, but this is not an easy task.
+
 We presented two ways to encode dynamically typed browser functions in mainstream statically typed languages like Java and Scala, using type parameters or path-dependent types.
-Our encodings give more type safety than existing solutions, while keeping the same expressive power as the native API.
+Our encodings give more type safety than existing solutions while keeping the same expressive power as the native API.
 
-We argue that, if industry shippers want to write their Web applications in statically typed languages, dependent types are going to be the most desired feature of these languages.
+We feel that parametric polymorphism and, even more, dependent types are precious type system features for languages aiming to bring static typing to Web applications.
+
+\paragraph{Acknowledgements} The authors would like to thank Thomas Degueule.
 

### 2014-02-25 papers/icwe14/icwe14.tex
typos

@@
 Section \ref{sec-validation} validates our contribution and discusses its limits.
-Section \ref{sec-related} discusses some related works and section \ref{sec-conclusion} concludes.
+Section \ref{sec-related} discusses some related works and Section \ref{sec-conclusion} concludes.
 
@@
 
-These type casts make the code more fragile and less convenient to read and write. That's why some statically typed languages attempt to provide an API preserving types as precisly as possible.
+These type casts make the code more fragile and less convenient to read and write. That's why some statically typed languages attempt to provide an API preserving types as precisely as possible.
 
 Of course, in the case of \jscode{getElementById(id)}, the \jscode{id} parameter does not give any clue on the possible type of the searched element, so it is hard to more precisely infer the return type of this function.
-Hence, most implementations define a return type of \jscode{Element}.
+Hence, most implementations use \jscode{Element} as their return type.
 
@@
 Listing~\ref{lst-problem-comb} shows what would our \scalacode{slideshow} implementation look like using such an encoding.
-There are two modifications compared to listing \ref{lst-problem-cast}: we use \scalacode{document.createImageElement} instead of \scalacode{document.createElement}, and we use \scalacode{img.addClickEventListener} instead of \scalacode{img.addEventListener}.
+There are two modifications compared to Listing \ref{lst-problem-cast}: we use \scalacode{document.createImageElement} instead of \scalacode{document.createElement}, and we use \scalacode{img.addClickEventListener} instead of \scalacode{img.addEventListener}.
 
@@
 The limits of this approach are better illustrated when one tries to combine several functions.
-Consider for instance listing~\ref{lst-js-comb} defining a JavaScript function \jscode{findAndListenTo} that both finds elements and registers an event listener when a given event occurs on them.
+Consider for instance Listing~\ref{lst-js-comb} defining a JavaScript function \jscode{findAndListenTo} that both finds elements and registers an event listener when a given event occurs on them.
 Note that the event listener is passed both the event and the element: its type depends on both the tag name and the event name.
@@
 The third approach, consisting in combining two parameters into one parameter carrying all the required information, is type safe too, but reduces the expressive power because it forbids developers to partially apply the function by supplying only one parameter.
-Consider for instance listing~\ref{lst-js-react} that defines a function \jscode{observe} partially applying the \jscode{addEventListener} function\footnote{The code of this function has been taken (and simplified) from the existing functional reactive programming libraries Rx.js~\cite{liberty2011reactive} and Bacon.js~(\href{http://baconjs.github.io/}{http://baconjs.github.io/}).}.
+Consider for instance Listing~\ref{lst-js-react} that defines a function \jscode{observe} partially applying the \jscode{addEventListener} function\footnote{The code of this function has been taken (and simplified) from the existing functional reactive programming libraries Rx.js~\cite{liberty2011reactive} and Bacon.js~(\href{http://baconjs.github.io/}{http://baconjs.github.io/}).}.
 Such a function cannot be implemented with this approach because the name of the event and the code of the listener cannot be decoupled.
@@
 \begin{enumerate}
- \item Define a parametrized class \scalacode{P[U]}
+ \item Define a parameterized class \scalacode{P[U]}
  \item Set the type of \scalacode{p} to \scalacode{P[U]}
@@
 
-It is worth noting that this code is actually exactly the same as in listing \ref{lst-problem-cast} excepted that type casts are not anymore required because the browser API is exposed in a way that preserves enough type information.
+It is worth noting that this code is actually exactly the same as in Listing~\ref{lst-problem-cast} excepted that type casts are not anymore required because the browser API is exposed in a way that preserves enough type information.
 Our way to encode the browser API is more type safe, but is it as expressive as the native API?
@@
 
-Listings \ref{lst-generics-comb} and \ref{lst-generics-react} show how the challenging functions of section \ref{sec-limitations}, \jscode{findAndListenTo} and \jscode{observe}, can be implemented with our encoding.
+Listings \ref{lst-generics-comb} and \ref{lst-generics-react} show how the challenging functions of Section \ref{sec-limitations}, \jscode{findAndListenTo} and \jscode{observe}, can be implemented with our encoding.
 They are basically a direct translation from JavaScript syntax to Scala syntax, with additional type annotations.
@@
 
-Programming languages generally support two means of abstraction: parametrization and abstract members.
-For instance Java supports parametrization for values (method parameters) and types (\emph{generics}), and member abstraction for values (abstract methods).
+Programming languages generally support two means of abstraction: parameterization and abstract members.
+For instance Java supports parameterization for values (method parameters) and types (\emph{generics}), and member abstraction for values (abstract methods).
 Scala also supports member abstraction for types through type members~\cite{cremet2006core,Odersky03_vObj}.
@@
 The first encoding, using type parameters, can be implemented in most programming languages because it only requires a basic support of type parameters (for instance Java, Dart, TypeScript, Kotlin, HaXe, Opa, Idris and Elm can implement it).
-However this encoding leads to cluttered type signatures and forces functions parametrized by event or element names to also take phantom type parameters.
+However this encoding leads to cluttered type signatures and forces functions parameterized by event or element names to also take phantom type parameters.
 

### 2014-03-31 papers/gdrgpl2014/LLNCStmpl.tex
initial version

% LLNCStmpl.tex
% Template file to use for LLNCS papers prepared in LaTeX
%websites for more information: http://www.springer.com
%http://www.springer.com/lncs

\documentclass{llncs}
%Use this line instead if you want to use running heads (i.e. headers on each page):
%\documentclass[runningheads]{llncs}

\begin{document}
\title{Title Goes Here}

%If you're using runningheads you can add an abreviated title for the running head on odd pages using the following
%\titlerunning{abreviated title goes here}
%and an alternative title for the table of contents:
%\toctitle{table of contents title}

\subtitle{Subtitle Goes Here}

%For a single author
%\author{Author Name}

%For multiple authors:
\author{First Author Name\inst{1} \and Second Author Name\inst{2}}

%If using runnningheads you can abbreviate the author name on even pages:
%\authorrunning{abbreviated author name}
%and you can change the author name in the table of contents
%\tocauthor{enhanced author name}

%For a single institute
%\institute{Institute Name \email{email address}}

% If authors are from different institutes 
\institute{First Institute Name \email{email address} \and Second Institute Name\thanks{Thank you to...} \email{email address}}

%to remove your email just remove '\email{email address}'
% you can also remove the thanks footnote by removing '\thanks{Thank you to...}'

\maketitle

\begin{abstract}
abstract text goes here - Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
\end{abstract}

\section{First Section}

\subsection{First Subsection}
\subsubsection{First Subsubsection}
Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.

\subsubsection{Second Subsubsection}
Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.

%Example of a numbered list
\begin{enumerate}
\item First item in list
\begin{enumerate}
\item First sub item
\item Second sub item
\end{enumerate}
\item Second item in list
\end{enumerate}

\subsection{Second Subsection}
Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud\cite{reference1} exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

\begin{proposition}
Insert proposition here
\end{proposition}

\begin{proof}
Insert proof here
\end{proof}

%The bibliography, done here without a bib file
%This is the old BibTeX style for use with llncs.cls
\bibliographystyle{splncs}

%Alternative bibliography styles:
%the following does the same as above except with alphabetic sorting
%\bibliographystyle{splncs_srt}
%the following is the current LNCS BibTex with alphabetic sorting
%\bibliographystyle{splncs03}
%If you want to use a different BibTex style include [oribibl] in the document class line

\begin{thebibliography}{1}
%add each reference in here like this:
\bibitem[RE1]{reference1}
Author:
Article/Book:
Other info: (date) page numbers.
\end{thebibliography}

\end{document}

### 2014-03-31 papers/gdrgpl2014/gdr-foy.tex
initial version

--- 2013-09-11 papers/gpce2013/gpce2013.tex
@@
-\documentclass[preprint]{sigplanconf}
+\documentclass{llncs}
 
@@
 
-\conferenceinfo{GPCE '13}{October 27--28, 2013, Indianapolis, Indiana, USA} 
-\copyrightyear{2013}
-\copyrightdata{[to be supplied]}
+%\conferenceinfo{GPCE~'13}{October 27--28, 2013, Indianapolis, Indiana, USA} 
+%\copyrightyear{2013} \copyrightdata{978-1-4503-2373-4/13/10} \doi{2517208.2517227}
 
@@
 
-\authorinfo{Julien Richard-Foy\and Olivier Barais\and Jean-Marc Jézéquel}
-           {IRISA, Université de Rennes 1}
-           {\{first\}.\{last\}@irisa.fr}
+\author{Richard-Foy\and Olivier Barais\and Jean-Marc Jézéquel}
+
+\institute{ IRISA, Université de Rennes 1
+           {\email{\{first\}.\{last\}@irisa.fr}}}
 
@@
 
-\category{D.3.3}{Programming Languages}{Language Constructs and Features}
+%\category{D.3.3}{Programming Languages}{Language Constructs and Features}
 
-\terms Languages, Software Engineering
+%\terms Languages, Software Engineering
 
@@
 compiled to each target platform. We propose to solve this dilemma by providing high-level
-abstractions in compiled domain-specific embedded languages (DSELs)~\cite{Hudak96_DSEL,
-Elliott2003_Compiling}. Compiled DSELs allow the definition of domain-specific languages (DSLs) as
+abstractions in compiled domain-specific embedded languages (DSELs)~\cite{Hudak96_DSEL,Elliott2003_Compiling}. Compiled DSELs allow the definition of domain-specific languages (DSLs) as
 libraries on top of a host language, and to compile them to a target platform. Their deep embedding
@@
 
-Lightweight Modular Staging~\cite{Rompf12_LMSThesis, Rompf12_LMS} (LMS) is a framework for defining
+Lightweight Modular Staging~\cite{Rompf12_LMSThesis,Rompf12_LMS} (LMS) is a framework for defining
 deeply embedded DSLs in Scala. It has been used to define high-performance DSLs for parallel
@@
 
+\subsection{LMS technical overview}
+
+In LMS, a DSL is split into two parts, its
+interface and its implementation. Both parts can be assembled from
+components in the form of Scala traits. DSL programs are written in
+terms of the DSL interface only, without knowledge of the
+implementation.
+
+Part of each DSL interface is an abstract type constructor
+\code{Rep[_]} that is used to wrap types in the DSL programs. The DSL
+implementation provides a concrete instantiation of \code{Rep} as IR
+nodes. When the DSL program is staged, it produces an intermediate
+representation (IR), from which the final code can be generated. In
+the DSL program, wrapped types such as \code{Rep[Int]} represent
+staged computations while expressions of plain unwrapped types
+(\code{Int}, \code{Bool}, etc.) are evaluated at staging time as
+in~\cite{finally-tagless,polymorphic-embedding}.
+
+Consider the difference between these two programs:
+\begin{lstlisting}
+def prog1(b: Bool, x: Rep[Int]) = if (b) x else x+1
+def prog2(b: Rep[Bool], x: Rep[Int]) = if (b) x else x+1
+\end{lstlisting}
+
+The only difference in these two programs is the type of the parameter
+\code{b}, illustrating that staging is purely type-driven with no
+syntactic overhead as the body of the programs are identical.
+
+In \code{prog1}, \code{b} is a simple boolean, so it must be provided
+at staging time, and the \code{if} is evaluated at staging time. For
+example, \code{prog1(true, x)} evaluates to \code{x}. In \code{prog2},
+\code{b} is a staged value, representing a computation which yields a
+boolean. So \code{prog2(b, x)} evaluates to an IR node for the \code{if}:
+{\tt\small If(b, x, Plus(x, Const(1)))}.
+
+For \code{prog2}, notice that the \code{if} got transformed into an IR
+node. To achieve this, LMS uses
+Scala-Virtualized~\cite{scala-virtualized}, a suite of minimal
+extensions to the regular Scala compiler, in which control
+structures such as \code{if} can be reified into method calls, so that
+alternative implementations can be provided. In our case, we provide an
+implementation of \code{if} that constructs an IR node instead of
+acting as a conditional. In addition, the \code{+} operation is
+overloaded to act on both staged and unstaged expressions. This is
+achieved by an implicit conversion from \code{Rep[Int]} to a class
+\code{IntOps}, which defines a \code{+} method that creates an IR node
+\code{Plus} when executed. Both of \code{Plus}'s arguments must be
+staged. We use an implicit conversion to stage constants when needed
+by creating a \code{Const} IR node.
+
+\subsection{Example: a DSL program and its generated JavaScript code}
+
+The following DSL snippet creates an array representing a table of
+multiplications:
+\begin{lstlisting}
+def test(n: Rep[Int]): Rep[Array[Int]] =
+  for (i <- range(0, n); j <- range(0, n)) yield i*j
+\end{lstlisting}
+
+Here is the JavaScript code generated for this snippet:
+\begin{lstlisting}[language=JavaScript]
+function test(x0) {
+  var x6 = []
+  for(var x1=0;x1<x0;x1++){
+    var x4 = []
+    for(var x2=0;x2<x0;x2++){
+      var x3 = x1 * x2
+      x4[x2]=x3
+    }
+    x6.splice.apply(x6, [x6.length,0].concat(x4))
+  }
+  return x6
+}
+\end{lstlisting}
+
+The generated code resembles single-assignment form. The nested
+\code{for}-loop is desugared into a \code{flatMap} which generates the
+nested \code{for}-loop and the \code{splice} pattern concatenating the
+inner \code{x4} arrays into one \code{x6} array in the JavaScript
+code.\footnote{Obviously, the generated code can be optimized further.}
+
+\subsection{Walkthrough: defining a DSL component}\label{sec:defineDslComponent}
+
+To conclude the introduction to LMS, we show how to add a component
+for logging in a DSL, generating JavaScript code which calls
+\code{console.log}.
+
+We start by defining the interface:
+\begin{lstlisting}
+trait Debug extends Base {
+  def log(msg: Rep[String]): Rep[Unit]
+}
+\end{lstlisting}
+
+The \code{Base} trait is part of the core LMS framework and provides
+the abstract type constructor \code{Rep}.
+
+Now, we define the implementation:
+\begin{lstlisting}
+trait DebugExp extends Debug with EffectExp {
+  case class Log(msg: Exp[String]) extends Def[Unit]
+  def log(msg: Exp[String]): Exp[Unit] = reflectEffect(Log(msg))
+}
+\end{lstlisting}
+
+The \code{EffectExp} trait is part of the core LMS framework. It
+inherits from \code{BaseExp} which instantiates \code{Rep} as
+\code{Exp}. \code{Exp} represents an IR via two subclasses:
+\code{Const} for constants and \code{Sym} for named values defining a
+\code{Def}. \code{Def} is the base class for all IR nodes. In our
+\code{DebugExp} trait, we extend \code{Def} to support a new IR node:
+\code{Log}.
+
+IR nodes are defined as \code{Def}s but they are never referenced
+explicitly as such. Instead each \code{Def} has a corresponding symbol
+(an instance of \code{Sym}). IR nodes refer to each other using their
+symbols. This is why, in the code shown, the \code{msg} parameter is
+of type \code{Exp} (not \code{Def}). The method \code{log} 
+returns an \code{Exp}. Calling \code{reflectEffect} is what creates
+this symbol from the \code{Def}.
+
+In general, the framework provides an implicit conversion from
+\code{Def} to \code{Exp}, which performs common subexpression
+elimination by re-using the same symbol for identical definitions. We
+do not use the automatic conversion here, because \code{log} is a 
+side-effecting operation, and we do not want 
+to (re)move any such calls even if their message
+is the same.
+
+The framework schedules the code generation from the graph of
+\code{Exp}s and their dependencies through \code{Def}s. It chooses
+which \code{Sym}/\code{Def} pairs to emit and in which order. To
+implement code generation to JavaScript for our logging IR node, we
+simply override \code{emitNode} to handle \code{Log}:
+\begin{lstlisting}
+trait JSGenDebug extends JSGenEffect {
+  val IR: DebugExp
+  import IR._
+  override def emitNode(sym: Sym[Any], rhs: Def[Any])(
+    implicit stream: PrintWriter) = rhs match {
+      case Log(s) => emitValDef(sym, "console.log(" + quote(s) + ")")
+      case _ => super.emitNode(sym, rhs)
+    }
+}
+\end{lstlisting}
+Notice that in order to compose nicely with other traits, the
+overridden method just handles the case it knows and delegates to
+other traits, via \code{super}, the emitting of nodes it doesn't know
+about.
+
 \section{Efficient High-Level Abstractions for Web Programming}
@@
 In a Web application, the user interface is defined by a HTML document that can be updated by the
-JavaScript code. A typical operation consists in searching some “interesting” element in the
+JavaScript code. A typical operation consists in searching some "interesting" element in the
 document, in order to extract its content, replace it or listen to user events triggered on it (such
@@
 
-\bibliographystyle{abbrvnat}
+\bibliographystyle{splncs}
+%\bibliographystyle{plain}
 \bibliography{biblio}
-%\begin{thebibliography}{}
-%\softraggedright
-%
-%\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
-%P. Q. Smith, and X. Y. Jones. ...reference text...
-%
-%\end{thebibliography}
-%
+
 \end{document}

### 2014-03-31 papers/gdrgpl2014/gdr-foy.tex
clean things up

--- 2013-09-11 papers/gpce2013/gpce2013.tex
@@
 \usepackage{graphicx}
-\usepackage{moresize}
+%\usepackage{moresize}
 
-\input{macros}
+\usepackage{listings}
+%\usepackage[scaled]{luximono}
+
+%\usepackage{natbib}
+
+% ----- begin macros
+
+\lstdefinelanguage{Scala}%
+{morekeywords={abstract,%
+  case,catch,char,class,%
+  def,else,extends,final,for,%
+  if,import,implicit,%
+  match,module,%
+  new,null,%
+  object,override,%
+  package,private,protected,public,%
+  for,public,return,super,%
+  this,throw,trait,try,type,%
+  val,var,%
+  with,%
+  yield,%
+  lazy%
+  },%
+  sensitive,%
+  morecomment=[l]//,%
+  morecomment=[s]{/*}{*/},%
+  morestring=[b]",%
+  morestring=[b]',%
+  showstringspaces=false%
+}[keywords,comments,strings]%
+
+\lstdefinelanguage{JavaScript}%
+{morekeywords={for, var, attributes, class, classend, do, else, empty, endif, endwhile, fail, function,
+functionend, if, implements, in, inherit, inout, not, of, operations, out, return,
+then, types, while, use},%
+  sensitive,%
+  morecomment=[l]//,%
+  morecomment=[s]{/*}{*/},%
+  morestring=[b]",%
+  morestring=[b]',%
+  showstringspaces=false%
+}[keywords,comments,strings]%
+
+\lstset{language=Scala,%
+  mathescape=false,%
+%  columns=[c]fixed,%
+  aboveskip=\smallskipamount,
+  belowskip=\smallskipamount,
+%  basewidth={0.5em, 0.4em},%
+  basicstyle=\footnotesize,%
+  captionpos=b,
+  keywordstyle=\bfseries%\sffamily\bfseries,%
+%  keywordstyle=\sffamily\bfseries,%
+%  xleftmargin=0.5cm
+}
+
+\newcommand{\commentstyle}[1]{\slseries{#1}}
+\newcommand{\keywordstyle}[1]{\bfseries{#1}}
+
+\lstnewenvironment{slisting}{\lstset{language=Scala}}{}
+
+\newcommand{\code}[1]{\lstinline[language=Scala,columns=fixed,basicstyle=\footnotesize]|#1|}
+
+\def\changemargin#1#2{\list{}{\rightmargin#2\leftmargin#1}\item[]}
+\let\endchangemargin=\endlist
 
@@
 
+\subsection{Walkthrough: defining a DSL component}\label{sec:defineDslComponent}
+
+For readers who are new to LMS, this section shows how to implement a language unit for logging.
+
+We start by defining the language unit vocabulary in an abstract trait:
+\begin{lstlisting}
+trait Debug extends Base {
+  def log(msg: Rep[String]): Rep[Unit]
+}
+\end{lstlisting}
+
+The \code{Base} trait is part of the core LMS framework and provides
+the abstract type constructor \code{Rep}.
+
+The trait contains just one method, \code{log}, representing the action of appending a message to the log.
+
+Then we define the language unit implementation:
+\begin{lstlisting}
+trait DebugExp extends Debug with EffectExp {
+  case class Log(msg: Exp[String]) extends Def[Unit]
+  def log(msg: Exp[String]): Exp[Unit] = reflectEffect(Log(msg))
+}
+\end{lstlisting}
+
+The \code{EffectExp} trait is part of the core LMS framework. It
+inherits from \code{BaseExp} which aliases \code{Rep} to
+\code{Exp}. \code{Exp} represents an IR via two subclasses:
+\code{Const} for constants and \code{Sym} for symbols referring to a composite expression. Composite expressions are represented by type
+\code{Def}.
+
+In our
+\code{DebugExp} trait, we extend \code{Def} to define a new IR node:
+\code{Log}. This node represents a statement appending a message to the log.
+
+An LMS program returns a value of type \code{Rep[_]}, a graph of IR nodes, representing the staged program.
+This graph is traversed and a code generator is invoked to transform each node into the corresponding code in the target language.
+To
+implement code generation to JavaScript for our logging IR node, we
+simply override \code{emitNode} to handle \code{Log}:
+\begin{lstlisting}
+trait JSGenDebug extends JSGenEffect {
+  val IR: DebugExp
+  import IR._
+  override def emitNode(sym: Sym[Any], rhs: Def[Any])(
+    implicit stream: PrintWriter) = rhs match {
+      case Log(s) => emitValDef(sym, q"console.log($s)")
+      case _ => super.emitNode(sym, rhs)
+    }
+}
+\end{lstlisting}
+
+We chose to implement the logging action by using the standard \code{console.log} Web browser's function.
+
 \section{Efficient High-Level Abstractions for Web Programming}

### 2014-04-07 papers/icwe14/icwe14.tex
minor rewording

@@
 
-We observe that, despite these languages are statically typed, their integration of the browser API either is not type safe or over constrain the programmers.
+We observe that, even though these languages are statically typed, their integration of the browser API either is not type safe or over constrain the programmers.
 Indeed, integrating an API designed for a dynamically typed language into a statically typed language can be challenging.
@@
 His approach allows developers to seamlessly use JavaScript APIs from statically typed Scala code.
-However, he did not use the features of Scala's type system, as we did, to leverage more precise types when possible.
+However, his work does not expose types as precise as ours (\emph{e.g.} in their encoding the return type of \jscode{createElement} is always \jscode{Element}).
 
